### 2016-04-07 paper.tex
Initial commit

\documentclass[11pt]{article}

\title{Architectures for worklfow integration in science gateways}

%\authors{Tristan Glatard, Marc-Etienne Rousseau, Pierre Rioux, Samir
%Das, Natacha Beck, Reza Adalat, Pierre Bellec, Pierre-Olivier Quirion,
%S\'ilvia D. Olabarriaga, Sorina Camarasu-Pop, Alan C. Evans}
% Rafael? Ewa? Peter Kacsuk? Johan? Satra? Najma?

\begin{document}

\maketitle

\section{Introduction}

Goal: review and compare the different architectures used to integrate
workflows in science gateways, based on practical experience.  A
workflow engine is defined as a software that submits jobs to a
cluster, grid or cloud. Workflows may be expressed in any language,
including scripts. Several workflow engines may be combined in the
same science gateway.

Context: The question of integrating workflow engines in science gateways can
be seen at various levels, corresponding to various definitions of
workflows. One level is the SHIWA level, where it was considered that
workflow engines are aware of the DCI (and 'D' is important because of
data transfers, proxies, etc). Another level is to remove 'D' from the
definition: a workflow engine becomes a program that submits jobs,
potentially only to local clusters. It opens a whole new class of
workflow engines that we used to consider as "applications". For
instance, in neuroinformatics: Nipype, PSOM, but also FSL through the
fslsub tool. A workflow engine is not supposed to be aware of the science gateway.

Without loss of generality, several of the examples presented in the
paper will be taken from medical image analysis, in particular
neuroimaging.

SG: a web portal with some features to handle DCIs (data transfers,
job submission, auth, etc). An ecosystem of services to access
distributed infrastructures. Examples: CBRAIN, VIP, NSG, P-GRADE, etc

Infrastructure: computing and storage resources. Includes schedulers, catalogues, etc.

Workflow: a process that submits tasks to the infrastructure. FSL
tools, say Feat, can be considered as workflows when they are
configured to submit tasks to clusters using fslsub.

A workflow consists of activities.

Contribution of the paper:
\begin{itemize}
\item We evaluate architectures based on our experience with existing systems
\item We propose a new architecture
\end{itemize}

We are talking only about concrete workflows with a configuration (see terstyansky et al 2014 "enabling scientific workflow sharing...")

\section{Architectures}

\subsection{Tight integration}

\subsection{Service invocation}

Or as a component.

= Examples
* VIP
* Silvia's gateway, starting from VBrowser

P-Grade: http://onlinelibrary.wiley.com/doi/10.1002/cpe.1654/full

= Description 
*  Workflow engine is deployed as a (Web) service called by the
science gateway. Probably the most adopted architecture.

= Evaluation
+ Easy to implement in the SG (just call the service)
- Load-balancing between services is difficult
(refer to VIP's experience).
- Need to wrap workflow engines in
services.
- Need to know where to submit each workflow, e.g. when
multiple workflows are involved.
- The science gateway has no
knowledge of the detailed progress of the workflow, in particular
job-level information (statuses, stdout, stderr). If we wanted to do
that, specific feedback channels have to be developed, which requires more work (see VIP).

\subsection{Sub-tasking}

Examples:
* CBRAIN and Niak, CBRAIN and FSL, CBRAIN and Nipype if time permits

+ Simplified parallelization
- Needs extension on the workflow engine side. Simple if workflow engine already supports task submission.

In  this model, the task of the SG is only to submit command line tools and command line tools submitted by these command line tools. The architecture is really upsided-down (was: workflow engine on top, jobs below. Is: job on top, workflow engine below). And this is what it makes it so powerful.

Sub-tasking also requires dependencies between tasks.

Can support various models such as pilot jobs, fsl sub, etc.

\subsection{Agent model}

= Examples
* SHIWA Pool

= Evaluation
+ Scale-up and load balancing
+ Multi-language
- Needs a complete framework (the pool). Most likely third-party software as it's a lot of work to implement.

\subsection{Nested workflows}

=  Examples
* SHIWA CGI, but SHIWA CGI is more complex because it enables meta workflows. In general, if we wanted only to use an embedded workflow, we don't need the meta-workflow for that. 
* Maybe other tool integrations in CBRAIN or VIP. For instance FSL or any tool that can be parallelized by this is not exploited.

Refer to the recent paper from Kacsuk.

Command line tool can run anwywhere.
- Parallelization is tricky because command line tool can run anywhere, on any cluster. Sometimes, it is solved by wrapping workflow engine in another worfklow engine, to iterate, e.g. on subjects. That's what we do in VIP. But this is not ideal.
+ No extension required on the SG side to support a workflow engine, as long as it can wrap command line tools.

\subsection{Workflow import}

= Examples
* IWIR (SHIWA FGI). Language conversion.
* first Niak+CBRAIN integration?

= Evaluation
* To interpret the language, you need either a generic interpreter or a conversion tool. Quite complex and hard to  maintain.

\section{Evaluation}

\subsection{Criteria}

Complexity of the architecture (therefore development effort, robustness, maintainability, etc):
  * On the SG side
  * On the workflow engine side (what is required to wrap it).
Could be measured by counting the number of required functions,
e.g. submit job, call a workflow service, etc. Ponderated by the
"complexity" of the features between 1 (easy) and 3 (tricky). Look at
other types of measures.

Provided features:
  * Fine-grained workflow monitoring (debugging, progress report)
  * Scalability, load-balancing between engines
  * Support for multiple engines in the same gateway. Allows to reuse several different workflows in a single gateway but less complex than meta workflows. But do we really need meta workflows?
  * Support for meta workflows

"A Formal Approach to Support Interoperability in Scientific
Meta-workflows" (reviewed for IWSG and JoGC) has a formal model to
evaluate CGI and FGI.

"four major approaches for workflow interoperability include a,b,c,d" (see Terstyanzky et al, "Enabling scientific workflow sharing through CGI...", FGCS 2014) -> read this ref.

http://www.oreilly.com/programming/free/files/software-architecture-patterns.pdf

\subsection{Evaluation}

\section{Discussion}

\section{Conclusion}

\section{Acknowledgments}

FLI-IAM, Labex PRIMES, Ludmer Centre

\bibliographystyle{plain} 
\bibliography{biblio}

\end{document}

### 2016-04-08 paper.tex
Added figures, updates in paper.tex

@@
-\documentclass[11pt]{article}
+\documentclass[preprint,3p,twocolumn]{elsarticle}
 
-\title{Architectures for worklfow integration in science gateways}
+\usepackage{svg}
+
+\newcommand{\todo}[1]{\color{blue} #1 \color{black}}
+\newcommand{\note}[1]{\color{blue}\textbf{Note: }\textit{#1} \color{black}}
+
+\begin{document}
+\begin{frontmatter}
+
+%% Title, authors and addresses
+
+%% use the tnoteref command within \title for footnotes;
+%% use the tnotetext command for theassociated footnote;
+%% use the fnref command within \author or \address for footnotes;
+%% use the fntext command for theassociated footnote;
+%% use the corref command within \author for corresponding author footnotes;
+%% use the cortext command for theassociated footnote;
+%% use the ead command for the email address,
+%% and the form \ead[url] for the home page:
+%% \title{Title\tnoteref{label1}}
+%% \tnotetext[label1]{}
+%% \author{Name\corref{cor1}\fnref{label2}}
+%% \ead{email address}
+%% \ead[url]{home page}
+%% \fntext[label2]{}
+%% \cortext[cor1]{}
+%% \address{Address\fnref{label3}}
+%% \fntext[label3]{}
+
+\title{Architectures for workflow integration in science gateways}
+
+%% use optional labels to link authors explicitly to addresses:
+%% \author[label1,label2]{}
+%% \address[label1]{}
+%% \address[label2]{}
+
+\author{}
+
+\address{}
+
+\begin{abstract}
+%% Text of abstract
+
+\end{abstract}
+
+\begin{keyword}
+%% keywords here, in the form: keyword \sep keyword
+
+%% PACS codes here, in the form: \PACS code \sep code
+
+%% MSC codes here, in the form: \MSC code \sep code
+%% or \MSC[2008] code \sep code (2000 is the default)
+
+\end{keyword}
+
+\end{frontmatter}
 
@@
 
-\begin{document}
+\journal{Future Generation Computer Systems}
 
@@
 
-Goal: review and compare the different architectures used to integrate
-workflows in science gateways, based on practical experience.  A
-workflow engine is defined as a software that submits jobs to a
-cluster, grid or cloud. Workflows may be expressed in any language,
-including scripts. Several workflow engines may be combined in the
-same science gateway.
-
-Context: The question of integrating workflow engines in science gateways can
-be seen at various levels, corresponding to various definitions of
-workflows. One level is the SHIWA level, where it was considered that
-workflow engines are aware of the DCI (and 'D' is important because of
-data transfers, proxies, etc). Another level is to remove 'D' from the
-definition: a workflow engine becomes a program that submits jobs,
-potentially only to local clusters. It opens a whole new class of
-workflow engines that we used to consider as "applications". For
-instance, in neuroinformatics: Nipype, PSOM, but also FSL through the
-fslsub tool. A workflow engine is not supposed to be aware of the science gateway.
-
-Without loss of generality, several of the examples presented in the
+\paragraph{Context} The question of integrating workflow engines in
+science gateways can be seen at various levels, corresponding to
+various definitions of workflows. One level is the SHIWA level, where
+it was considered that workflow engines are aware of the DCI (and 'D'
+is important because of data transfers, proxies, etc). Another level
+is to remove 'D' from the definition: a workflow engine becomes a
+program that submits jobs, potentially only to local clusters. It
+opens a whole new class of workflow engines that we used to consider
+as "applications". For instance, in neuroinformatics: Nipype, PSOM,
+but also FSL through the fslsub tool. A workflow engine is not
+supposed to be aware of the science gateway.  A wide-array of workflow
+engines are available with specificies coming from the application
+domain, available tools, etc Their integration in science gateways
+becomes critical. Several of the examples presented in the
 paper will be taken from medical image analysis, in particular
@@
 
-SG: a web portal with some features to handle DCIs (data transfers,
-job submission, auth, etc). An ecosystem of services to access
-distributed infrastructures. Examples: CBRAIN, VIP, NSG, P-GRADE, etc
+\paragraph{Goal} this paper reviews and compares the architectures to
+integrate workflow engines in science gateways.
 
-Infrastructure: computing and storage resources. Includes schedulers, catalogues, etc.
-
-Workflow: a process that submits tasks to the infrastructure. FSL
-tools, say Feat, can be considered as workflows when they are
-configured to submit tasks to clusters using fslsub.
-
-A workflow consists of activities.
-
-Contribution of the paper:
+\paragraph{Contributions}
 \begin{itemize}
@@
 
-We are talking only about concrete workflows with a configuration (see terstyansky et al 2014 "enabling scientific workflow sharing...")
+\section{Workflow engines and science gateways}
+
+\subsection{Workflow engines}
+
+In the last decade, the e-Science workflow community has developed
+high-level workflow systems to help developers access distributed
+infrastructures such as grids and web services, resulting in tools
+such as Askalon, Hyperflow, MOTEUR, Pegasus, Swift, Taverna, Triana,
+VizTrails, WS-PGRADE/gUSE, etc. Such workflow engines usually describe
+applications in a specific language that offers operators for parallel
+computing, visual description and edition, links with domain-specific
+tool repositories, etc. Descriptions and experiments conducted with
+e-Science workflow engines were published in journals such as Future
+Generation Computer Systems and conference venues such as
+WORKS.
+
+At the same time, specific toolboxes were emerging in various
+scientific domains to facilitate interactions between different
+software components. In neuroimaging for instance, tools such as
+Nipype, PSOM, SPM and FSL provide abstractions and functions to define
+processes that handle the data flow between other processes. Soon,
+such tools were interfaced to computing systems, in particular
+clusters: they were extended to create cluster tasks, handle their
+dependencies, and execute them on clusters. For instance, FSL can
+launch tasks on SGE through its \texttt{fsl\_sub} tool, Nipype
+\todo{...}, PSOM \todo{...}, and SPM \todo{...}. Although distributed
+infrastructures are hardly handled, such domain-specific tools have to
+be called workflow engines. They represent a tremendous opportunity
+for science gateways to leverage existing tools and
+applications. Whether these should be combined with e-Science workflow
+engines is an open question.
+
+A workflow engine is defined as a software that submits jobs to a
+cluster, grid or cloud. Workflows may be expressed in any language,
+including scripts. Several workflow engines may be combined in the
+same science gateway. Workflow: a process that submits tasks to the
+infrastructure. FSL tools, say Feat, can be considered as workflows
+when they are configured to submit tasks to clusters using fslsub. A
+workflow consists of activities. We are talking only about concrete
+workflows with a configuration (see terstyansky et al 2014 "enabling
+scientific workflow sharing...")
+
+\subsection{Science gateways}
+
+Science Gateway: An ecosystem, usually accessible through a web
+  portal, that provides tools to access distributed
+  infrastructures. Tools usually help users manage data transfers,
+  task execution and authentication on multiple computing and storage
+  locations. Examples: CBRAIN, VIP, NSG, etc
+
+\subsection{Infrastructures}
+
+\todo{Maybe this is not relevant}
 
@@
 
+We define architectures from their main components and the interactions
+between these components.
+
+A workflow engine is not supposed to be aware of the science gateway (?).
+
+\subsection{Common components}
+
+Infrastructure: computing and storage resources. Includes
+  schedulers, catalogues, etc. Can be grid, cloud or even
+  cluster. Some workflow engines may assume a shared file system
+  usually available only on clusters.
+
+  Workflow engine: executes workflows, i.e. process their dependencies
+  and create computing tasks. Might transfer data too.
+
+\subsection{Interactions}
+
+a.
+
+d. Sub-tasking: science gateway knows the sub-tasks of a
+task. Therefore, it can kill all sub-tasks of a task when a task is killed.
+
+g. Workflow integration results in an interface, for instance a web form,
+where users can enter the parameters of the workflow to be
+executed. Integrating a workflow is not the same process as
+integrating a workflow engine. Integrating a workflow engine results
+in an interface that allows developers to enter the parameters of the
+workflow engine.
+
+Only workflow execution is described. 
+Workflow integration is not covered.
+
 \subsection{Tight integration}
+
+\begin{figure}
+\centering
+\def\svgwidth{0.5\columnwidth}
+\input{images/tight.pdf_tex}
+\label{archi:tight}
+\caption{Tight integration}
+\end{figure}
 
@@
 
+\begin{figure}
+\centering
+\def\svgwidth{0.5\columnwidth}
+\input{images/service.pdf_tex}
+\label{archi:service}
+\caption{Service integration}
+\end{figure}
+
 \subsection{Sub-tasking}
@@
 Can support various models such as pilot jobs, fsl sub, etc.
+
+No global vision of the workflow.
+
+Science gateway has to make sure that the machine on which a workflow
+task executes is reliable as several sub-tasks will depend on the
+success of this task. Or workflow engine task has to be recoverable.
+
+Performance: there could be a deadlock when there is contention. Also,
+the workflow task has to be reliable (for instance, proper walltime
+estimation), which is difficult because these tasks are
+long. Restarting a workflow task may mean that all the tasks in the
+workflow have to be restarted, depending on the capability of the
+workflow engine. This complexifies task scheduling (a system could
+ignore these issues but they may backfire).
 
@@
 
+These architectures can be combined (give examples).
+
 \section{Conclusion}
@@
 
-\bibliographystyle{plain} 
+\bibliographystyle{elsarticle-num} 
 \bibliography{biblio}
@@
 \end{document}
+\endinput

### 2016-04-08 images/export.sh

@@ -1,0 +1,9 @@
+#!/bin/bash
+
+for name in tight service
+do
+  echo "Exporting ${name}.svg"
+  inkscape -D -z --file=images/${name}.svg --export-pdf=images/${name}.pdf --export-latex
+  sed s,${name}\.pdf,images/${name}\.pdf,g images/${name}.pdf_tex > images/${name}.pdf_tex.new
+  \mv images/${name}.pdf_tex.new images/${name}.pdf_tex
+done

### 2016-04-11 paper.tex
Updates in section 3

@@
 \usepackage{svg}
+\usepackage{url}
+\usepackage{xspace}
 
-\newcommand{\todo}[1]{\color{blue} #1 \color{black}}
+\newcommand{\todo}[1]{\color{blue}\xspace\emph{#1}\xspace\color{black}}
 \newcommand{\note}[1]{\color{blue}\textbf{Note: }\textit{#1} \color{black}}
@@
 
+c. Data movements can be triggered by the user, to upload input data
+or download processed data, or by the workflow engine, to transfer
+data across the infrastructure.
+
 d. Sub-tasking: science gateway knows the sub-tasks of a
@@
 
+See Figure~\ref{archi:tight}. The workflow engine is tightly
+integrated with the science gateway, which means that it shares code,
+framework components, libraries, etc. For instance, the workflow
+engine might be a portlet in a Liferay portal, a controller in a Ruby
+on Rails application, and so on. The workflow engine and the science
+gateway usually share a database where application, users and other
+resources are stored. In this model, task control and data control are
+initiated from the science gateway. This is the model adopted in the
+Catania Science Gateway Framework~\cite{Ardizzone2012} (see specific
+documentation on
+workflows\footnote{\url{http://bit.ly/1oQrzvQ}})
+and LONI Pipeline Environment~\cite{dinov2009efficient}.\todo{Add better justification}
+
 \begin{figure}
@@
 \input{images/tight.pdf_tex}
+\caption{Tight integration}
 \label{archi:tight}
-\caption{Tight integration}
 \end{figure}
@@
 
-Or as a component.
+See Figure~\ref{archi:service}. The workflow engine is available
+externally to the science gateway. The science gateway controls it
+through a specific interaction that might be implemented as a
+web-service call (e.g. RESTful or SOAP), as a command-line or as any
+other method that separates the workflow engine from the science
+gateway. The workflow engine might be invoked either as a black box
+that completely masks the infrastructure and workflow activities, or
+as a grey box that allows for some interaction with them.  The
+workflow engine is responsible for controlling the tasks on the
+infrastructure, and performing the required data transfers to execute
+them. User data is usually managed through the science gateway,
+although it might as well be delivered by the workflow engine directly
+to the user.
 
-= Examples
-* VIP
-* Silvia's gateway, starting from VBrowser
-
-P-Grade: http://onlinelibrary.wiley.com/doi/10.1002/cpe.1654/full
-
-= Description 
-*  Workflow engine is deployed as a (Web) service called by the
-science gateway. Probably the most adopted architecture.
-
-= Evaluation
-+ Easy to implement in the SG (just call the service)
-- Load-balancing between services is difficult
-(refer to VIP's experience).
-- Need to wrap workflow engines in
-services.
-- Need to know where to submit each workflow, e.g. when
-multiple workflows are involved.
-- The science gateway has no
-knowledge of the detailed progress of the workflow, in particular
-job-level information (statuses, stdout, stderr). If we wanted to do
-that, specific feedback channels have to be developed, which requires more work (see VIP).
-
+This architecture is largely adopted, in systems such as Apache
+Airvata~\cite{marru2011apache}, Vine
+Toolkit~\cite{DBLP:journals/scpe/SzejnfeldDKKKKLPTWDNW10}, Virtual
+Imaging Platform~\cite{GLAT-13}, the WS-PGRADE/gUSE
+framework~\cite{Kacsuk2012} and the numerous science gateway instances
+that use it~\cite{kacsuk2014science}.
 \begin{figure}
@@
 \input{images/service.pdf_tex}
+\caption{Service integration}
 \label{archi:service}
-\caption{Service integration}
 \end{figure}
@@
 
-Examples:
-* CBRAIN and Niak, CBRAIN and FSL, CBRAIN and Nipype if time permits
+See Figure~\ref{archi:sub-task}. The science gateway is responsible to
+execute the tasks on the infrastructure, dealing with heterogeneous
+batch managers and meta-schedulers. The workflow engine is a
+particular task that can submit sub-tasks to the science gateway. The
+workflow engine keeps track of the dependencies between the sub-tasks
+but their execution is delegated to the science gateway. The science
+gateway may implement mechanisms to deal with task dependencies such
+as basic dependency lists as available in most batch managers. It has
+no global vision of the workflow. The fact that tasks are submitted
+through the science gateway does not restrict the range of possible
+execution models. Pilot jobs, for instance, can still be used.
 
-+ Simplified parallelization
-- Needs extension on the workflow engine side. Simple if workflow engine already supports task submission.
+This model is implemented in CBRAIN~\cite{SHER-14} where it is used to
+integrate the PSOM workflow engine and the FSL toolkit through its
+\texttt{fsl\_sub} tool. In particular, the CBRAIN-PSOM integration is
+described in~\cite{GLAT-16} and uses an agent computing model (a.k.a pilot
+jobs).
+\begin{figure}
+\centering
+\def\svgwidth{0.5\columnwidth}
+\input{images/sub-task.pdf_tex}
+\caption{Sub-tasking}
+\label{archi:sub-task}
+\end{figure}
 
-In  this model, the task of the SG is only to submit command line tools and command line tools submitted by these command line tools.
-The architecture is really upsided-down (was: workflow engine on top, jobs below.
-Is: job on top, workflow engine below).
-And this is what it makes it so powerful.
+\subsection{Pool model}
 
-Sub-tasking also requires dependencies between tasks.
-
-Can support various models such as pilot jobs, fsl sub, etc.
-
-No global vision of the workflow.
-
-Science gateway has to make sure that the machine on which a workflow
-task executes is reliable as several sub-tasks will depend on the
-success of this task. Or workflow engine task has to be recoverable.
-
-Performance: there could be a deadlock when there is contention. Also,
-the workflow task has to be reliable (for instance, proper walltime
-estimation), which is difficult because these tasks are
-long. Restarting a workflow task may mean that all the tasks in the
-workflow have to be restarted, depending on the capability of the
-workflow engine. This complexifies task scheduling (a system could
-ignore these issues but they may backfire).
-
-\subsection{Agent model}
-
-= Examples
-* SHIWA Pool
-
-= Evaluation
-+ Scale-up and load balancing
-+ Multi-language
-- Needs a complete framework (the pool). Most likely third-party software as it's a lot of work to implement.
+See Figure~\ref{archi:agent}. Workflows are submitted by the science
+gateway to a pool to which agents connect asynchronously to retrieve
+and execute workflows (interaction \texttt{f}). Agents may be started according to various
+policies, for instance to ensure load balancing. Agents may wrap
+different types of workflow engines. This model was implemented in the
+SHIWA pool~\cite{ROGE-13}.
+\begin{figure}
+\centering
+\def\svgwidth{0.8\columnwidth}
+\input{images/agent.pdf_tex}
+\caption{Pool model}
+\label{archi:agent}
+\end{figure}
 
@@
 
+\paragraph{Service}
++ Easy to implement in the SG (just call the service)
+- Load-balancing between services is difficult
+(refer to VIP's experience).
+- Need to wrap workflow engines in
+services.
+- Need to know where to submit each workflow, e.g. when
+multiple workflows are involved.
+- The science gateway has no
+knowledge of the detailed progress of the workflow, in particular
+job-level information (statuses, stdout, stderr). If we wanted to do
+that, specific feedback channels have to be developed, which requires more work (see VIP).
+
+\paragraph{Sub-task}
+
+Science gateway has to make sure that the machine on which a workflow
+task executes is reliable as several sub-tasks will depend on the
+success of this task. Or workflow engine task has to be recoverable.
+
+Performance: there could be a deadlock when there is contention. Also,
+the workflow task has to be reliable (for instance, proper walltime
+estimation), which is difficult because these tasks are
+long. Restarting a workflow task may mean that all the tasks in the
+workflow have to be restarted, depending on the capability of the
+workflow engine. This complexifies task scheduling (a system could
+ignore these issues but they may backfire).
+
+\paragraph{Pool}
+
+= Evaluation
++ Scale-up and load balancing
++ Multi-language
+- Needs a complete framework (the pool). Most likely third-party software as it's a lot of work to implement.
+
 \section{Discussion}

### 2016-04-11 images/export.sh

@@ -1,6 +1,6 @@
 #!/bin/bash
 
-for name in tight service
+for name in tight service sub-task agent
 do
   echo "Exporting ${name}.svg"
   inkscape -D -z --file=images/${name}.svg --export-pdf=images/${name}.pdf --export-latex

### 2016-04-11 images/export.sh
Updates in image export script

@@ -1,6 +1,6 @@
 #!/bin/bash
 
-for name in tight service sub-task agent
+for name in tight service sub-task agent nested-1 nested-2 import-1 import-2
 do
   echo "Exporting ${name}.svg"
   inkscape -D -z --file=images/${name}.svg --export-pdf=images/${name}.pdf --export-latex

### 2016-04-11 paper.tex
Updates in section 3

@@
 
+Multi-engine, multi-language.
+
 \subsection{Infrastructures}
@@
 A workflow engine is not supposed to be aware of the science gateway (?).
+
+%http://ieeexplore.ieee.org/stamp/stamp.jsp?tp=&arnumber=4782949
 
@@
 See Figure~\ref{archi:tight}. The workflow engine is tightly
-integrated with the science gateway, which means that it shares code,
-framework components, libraries, etc. For instance, the workflow
-engine might be a portlet in a Liferay portal, a controller in a Ruby
-on Rails application, and so on. The workflow engine and the science
-gateway usually share a database where application, users and other
-resources are stored. In this model, task control and data control are
-initiated from the science gateway. This is the model adopted in the
-Catania Science Gateway Framework~\cite{Ardizzone2012} (see specific
-documentation on
-workflows\footnote{\url{http://bit.ly/1oQrzvQ}})
-and LONI Pipeline Environment~\cite{dinov2009efficient}.\todo{Add better justification}
+integrated with the science gateway, which means that it is deployed
+on the same machine and potentially shares code, libraries and other
+software components with the science gateway. For instance, the
+workflow engine might be a portlet in a Liferay portal, a controller
+in a Ruby on Rails application, and so on. The workflow engine and the
+science gateway usually share a database where application, users and
+other resources are stored. In this model, task control and data
+control are initiated from the science gateway. This is the model
+adopted in the Catania Science Gateway Framework~\cite{Ardizzone2012}
+(see specific documentation on
+workflows\footnote{\url{http://bit.ly/1oQrzvQ}}),
+CIPRES~\cite{miller2010creating} and LONI Pipeline
+Environment~\cite{dinov2009efficient}.\todo{Add better justification}
 
@@
 
-=  Examples
-* SHIWA CGI, but SHIWA CGI is more complex because it enables meta workflows. In general, if we wanted only to use an embedded workflow, we don't need the meta-workflow for that. 
-* Maybe other tool integrations in CBRAIN or VIP. For instance FSL or any tool that can be parallelized by this is not exploited.
+See Figure~\ref{archi:nested}. A nested workflow is an abstract
+architectural pattern where an activity of a master workflow is itself
+a workflow that is executed by another engine. Nested workflows can be
+instantiated in the various architectures described previously. We
+focus on instantiation with the service invocation model as this is
+the most used architecture. The master workflow is also called
+meta-workflow. The nested workflow might be described with a different
+language than the one used to describe the master workflow.
 
-Refer to the recent paper from Kacsuk.
+Nested workflows have long been available in workflow engines, for
+instance in the Taverna workbench~\cite{oinn2004taverna}. This model
+was used by the SHIWA Simulation Platform to implement Coarse-Grained
+workflow interoperability~\cite{terstyanszky2014enabling}, i.e. to
+integrate various workflow engines in a consistent platform. Nested
+workflows are also often used implicitly to wrap scripts as
+black-boxes while these scripts actually correspond to complete workflows.
 
-Command line tool can run anwywhere.
-- Parallelization is tricky because command line tool can run anywhere, on any cluster.
-Sometimes, it is solved by wrapping workflow engine in another worfklow engine, to iterate, e.g.
-on subjects.
-That's what we do in VIP.
-But this is not ideal.
-+ No extension required on the SG side to support a workflow engine, as long as it can wrap command line tools.
+\begin{figure}
+\centering
+\def\svgwidth{0.48\columnwidth}
+\input{images/nested-1.pdf_tex}
+\def\svgwidth{0.48\columnwidth}
+\input{images/nested-2.pdf_tex}
+\caption{Nested workflows. Left: abstract model. Right: instantiation with service invocation.}
+\label{archi:nested}
+\end{figure}
 
 \subsection{Workflow import}
+
+See Figure~\ref{archi:import}.
 
@@
 * To interpret the language, you need either a generic interpreter or a conversion tool. Quite complex and hard to  maintain.
+
+\begin{figure}
+\centering
+\def\svgwidth{0.48\columnwidth}
+\input{images/import-1.pdf_tex}
+\def\svgwidth{0.48\columnwidth}
+\input{images/import-2.pdf_tex}
+\caption{Workflow import. Left: abstract model. Right: instantiation with service invocation.}
+\label{archi:import}
+\end{figure}
 

### 2016-04-11 paper.tex
Updates in section 3 (first complete draft)

@@
 \usepackage{xspace}
+\usepackage{enumitem}
+\usepackage{hyperref}
 
@@
 
+  Workflow engine: executes workflows, i.e. process their dependencies
+  and create computing tasks. Might transfer data too. A workflow
+  engine is not supposed to be aware of the science gateway.
+
 \subsection{Science gateways}
@@
 
+The infrastructure consists of the computing and storage resources
+involved in the workflow execution, and the software services used to
+access these resources. Infrastructures can be servers, clusters,
+grids or clouds. Some workflow engines may assume specific
+characteristics about the infrastructure, such as the presence of a
+shared file system between the computing nodes.
+
 \section{Architectures}
 
-We define architectures from their main components and the interactions
-between these components.
-
-A workflow engine is not supposed to be aware of the science gateway (?).
+Architectures are defined below from their main components and the
+interactions between these components.
 
@@
 
-\subsection{Common components}
-
-Infrastructure: computing and storage resources. Includes
-  schedulers, catalogues, etc. Can be grid, cloud or even
-  cluster. Some workflow engines may assume a shared file system
-  usually available only on clusters.
-
-  Workflow engine: executes workflows, i.e. process their dependencies
-  and create computing tasks. Might transfer data too.
-
 \subsection{Interactions}
 
-a.
+The interactions described below are labeled consistently with the
+notations used on Figures~\ref{archi:tight}-\ref{archi:import}.
 
-c. Data movements can be triggered by the user, to upload input data
-or download processed data, or by the workflow engine, to transfer
-data across the infrastructure.
+\begin{enumerate}[leftmargin=0cm,itemindent=0.6cm,label=\texttt{(\alph*)}]
 
-d. Sub-tasking: science gateway knows the sub-tasks of a
-task. Therefore, it can kill all sub-tasks of a task when a task is killed.
-
-g. Workflow integration results in an interface, for instance a web form,
-where users can enter the parameters of the workflow to be
-executed. Integrating a workflow is not the same process as
-integrating a workflow engine. Integrating a workflow engine results
-in an interface that allows developers to enter the parameters of the
-workflow engine.
-
-Only workflow execution is described. 
-Workflow integration is not covered.
+\item Workflow integration: consists in adding a new workflow to the
+  science gateway so that users can execute it. It is conducted by
+  science gateway administrators, developers or users, and it results
+  in an interface, for instance a web form, where users can enter the
+  parameters of the workflow to be executed. Integrating a workflow is
+  not the same process as integrating a workflow engine.
+\item Task control: operations to manage tasks on the infrastructure,
+  including: submission, monitoring, termination, deletion, etc.
+\item Data control: operations to manage data on
+  the infrastructure, such as: upload, download, deletion, browsing,
+  replication, caching, etc. Data movements can be triggered by the
+  user, to upload input data or download processed data - interaction
+  \texttt{(c1)}, or by the workflow engine, to transfer data across the
+  infrastructure - interaction \texttt{(c2)}. \todo{add c1 and c2 to the
+    Figures}
+\item Workflow control: operations to execute a workflow, including:
+  submission, monitoring, termination, etc. Workflow control can be
+  coarse-grained (a.k.a. black box) or fine-grained (white box). In a
+  coarse-grained model, workflow activities are masked. 
+\item Sub-task control: operations used by tasks to submit sub-tasks
+  on the infrastructure, including: submission, monitoring,
+  termination, deletion, etc.
+\item Pool-agent: specific to the architecture described in~\ref{sec:pool}.
+\item Workflow conversion: translation from one workflow language to
+  another one.
+\end{enumerate}
 
@@
 gateway may implement mechanisms to deal with task dependencies such
-as basic dependency lists as available in most batch managers. It has
-no global vision of the workflow. The fact that tasks are submitted
-through the science gateway does not restrict the range of possible
-execution models. Pilot jobs, for instance, can still be used.
+as basic dependency lists as available in most batch
+managers. Although it has no global vision of the workflow, it can
+keep track of the sub-tasks submitted by a given task, for instance to
+be able to cancel them when the task is canceled. The fact that tasks
+are submitted through the science gateway does not restrict the range
+of possible execution models: pilot jobs, for instance, can still be
+used.
 
@@
 \subsection{Pool model}
-
+\label{sec:pool}
 See Figure~\ref{archi:agent}. Workflows are submitted by the science
@@
 
-See Figure~\ref{archi:import}.
-
-= Examples
-* IWIR (SHIWA FGI). Language conversion.
-* first Niak+CBRAIN integration?
-
-= Evaluation
-* To interpret the language, you need either a generic interpreter or a conversion tool. Quite complex and hard to  maintain.
+See Figure~\ref{archi:import}. This is an abstract model that we
+instantiate with the service invocation architecture for
+consistency. Workflows are integrated in the science gateway through
+format conversion from a native format to the science gateway
+format. This was implemented in the SHIWA Simulation Platform through
+the IWIR language that provided a common language for portability
+across grid workflow
+systems~\cite{plankensteiner-prodan-etal:2013}.
 

### 2016-04-12 paper.tex
Updates in paper.tex

@@
 
+\usepackage[x11names,dvipsnames,table]{xcolor} %for use in color links
 \usepackage{svg}
@@
 \usepackage{hyperref}
+\usepackage{multirow}
 
@@
 \begin{document}
+
 \begin{frontmatter}
@@
 
+\todo{Introduce the notion of architectural diagram with an abstract figure.}
+
 %http://ieeexplore.ieee.org/stamp/stamp.jsp?tp=&arnumber=4782949
@@
   replication, caching, etc. Data movements can be triggered by the
-  user, to upload input data or download processed data - interaction
-  \texttt{(c1)}, or by the workflow engine, to transfer data across the
-  infrastructure - interaction \texttt{(c2)}. \todo{add c1 and c2 to the
+  user, to upload input data or download processed data -- interaction
+  \texttt{$c_1$}, or by the workflow engine, to transfer data across the
+  infrastructure -- interaction \texttt{$c_2$}. \todo{add c1 and c2 to the
     Figures}
@@
 
-Complexity of the architecture (therefore development effort, robustness, maintainability, etc):
-  * On the SG side
-  * On the workflow engine side (what is required to wrap it).
-Could be measured by counting the number of required functions,
-e.g. submit job, call a workflow service, etc. Ponderated by the
-"complexity" of the features between 1 (easy) and 3 (tricky). Look at
-other types of measures.
+We use four main criteria to evaluate the architectures: integration
+effort, robustness, modularity and scalability. Criteria break down to
+specific metrics for which low values indicate good performance.
 
-Provided features:
-  * Fine-grained workflow monitoring (debugging, progress report)
-  * Scalability, load-balancing between engines
-  * Support for multiple engines in the same gateway. Allows to reuse several different workflows in a single gateway but less complex than meta workflows. But do we really need meta workflows?
-  * Support for meta workflows
+\todo{Explain these metrics better, including why the related features
+  are useful.}
 
-"A Formal Approach to Support Interoperability in Scientific
-Meta-workflows" (reviewed for IWSG and JoGC) has a formal model to
-evaluate CGI and FGI.
+\emph{Integration effort} measures the development effort required to build
+the system, assuming that none of the components could initially
+communicate with each other. It is measured by counting the number of
+interactions to develop and components to modify on the architectural
+diagram. It breaks down to 2 metrics that relate to the direction of
+interaction arrows on the architectural diagrams:
+\begin{itemize}[itemsep=0cm]
+\item Interactions from the science gateway. 
+\item Iteractions from the workflow engine. 
+\end{itemize}
 
-"four major approaches for workflow interoperability include a,b,c,d" (see Terstyanzky et al, "Enabling scientific workflow sharing through CGI...", FGCS 2014) -> read this ref.
+\emph{Robustness} relates to the complexity of the architecture and
+its ability to provide relevant debugging information about the
+workflow executions. It breaks down to 3 metrics:
+\begin{itemize}[itemsep=0cm]
+\item Components: number of unique components involved in workflow executions.
+\item Interactions: number of unique interactions involved in workflow executions. 
+\item Debugging: availability of debugging information about workflow activities
+  (1: difficult to obtain; 0: easy to access).
+\end{itemize}
 
-http://www.oreilly.com/programming/free/files/software-architecture-patterns.pdf
+\emph{Modularity} measures the ability to replace elements in the
+architecture, or to integrate new elements. Elements include
+workflow engines, workflows and infrastructure. It breaks down in 4
+metrics:
+\begin{itemize}[itemsep=0cm]
+\item New engine type: number of interactions or components to modify
+  to integrate a new workflow engine type in the architecture.
+\item New engine version: number of interactions or components to
+  modify to integrate a new version of a workflow engine in the
+  architecture, assuming that another version of the same engine type
+  is already available.
+\item New infrastructure: number of interactions or components to
+  modify to integrate a new type of infrastructure in the
+  architecture.
+\item New workflow: number of interactions required to integrate a new
+  workflow in the architecture.
+\item Meta-workflow: ability to describe meta-workflows from existing
+  workflows (1: not available; 0: available).
+\end{itemize}
+
+\emph{Scalability} measures the ability of the architecture to support
+high workloads. It breaks down into 4 metrics:
+\begin{itemize}[itemsep=0cm]
+\item New engine instance (0: automated; 1: manual intervention needed).
+\item Elastic engines (0: easy to implement; 1: difficult to implement).
+\item Distributed engines (0: possible; 1: not possible).
+\item Task scheduling (0: difficult; 1: more difficult).
+\end{itemize}
+
+%"A Formal Approach to Support Interoperability in Scientific
+%Meta-workflows" (reviewed for IWSG and JoGC) has a formal model to
+%evaluate CGI and FGI.
+%"four major approaches for workflow interoperability include a,b,c,d" (see Terstyanzky et al, "Enabli%ng scientific workflow sharing through CGI...", FGCS 2014) -> read this ref.
+% http://www.oreilly.com/programming/free/files/software-architecture-patterns.pdf
 
@@
 
-\paragraph{Service}
+See Table~\ref{table:evaluation}.
+
+We focus on the integration of workflow engines in science gateways,
+i.e. we assume that the science gateway is already interfaced with the
+infrastructure.
+
+\paragraph{Service invocation} Integrating a workflow engine as a
+service requires 3 interactions. First, the science gateway has to be
+extended to call the engine interface (\texttt{d}), which might
+require some adaptation on the workflow engine too, for instance
+wrapping in a web service. In addition, the workflow engine has to be
+interfaced with the infrastructure, which requires the two
+interactions \texttt{b} and \texttt{$c_2$}.
+
 + Easy to implement in the SG (just call the service)
@@
 
+\begin{table*}
+\centering
+\begin{tabular}{rcccccc}
+\textbf{Metric}                      & \textbf{Tight}
+                                     & \textbf{Service}
+                                     & \textbf{Sub-task}
+                                     & \textbf{Pool}
+                                     & \textbf{Nested}
+                                     & \textbf{Import} \\
+\multicolumn{7}{c}{\cellcolor[HTML]{EEEEEE}\textbf{Integration effort}}\\
+From science gateway                 & \multirow{2}{*}{3}
+                                     & \cellcolor[HTML]{99FF99}1
+                                     & \cellcolor[HTML]{99FF99}2  
+                                     & \cellcolor[HTML]{99FF99}2
+                                     & \cellcolor[HTML]{99FF99}2
+                                     & \cellcolor[HTML]{99AA99}3 \\
+From workflow engine                 &
+                                     & \cellcolor[HTML]{99FF99}1 
+                                     & \cellcolor[HTML]{99FF99}2
+                                     & \cellcolor[HTML]{99AA99}2
+                                     & \cellcolor[HTML]{99AA99}2
+                                     & \cellcolor[HTML]{99FF99}1 \\
+\textbf{Total}                       & \cellcolor[HTML]{99FF99}\textbf{3}
+                                     & \cellcolor[HTML]{99FF99}\textbf{3}
+                                     & \cellcolor[HTML]{99FF99}\textbf{3}
+                                     & \cellcolor[HTML]{99AA99}\textbf{4}
+                                     & \cellcolor[HTML]{99AA99}\textbf{4}
+                                     & \cellcolor[HTML]{99AA99}\textbf{4} \\
+
+\multicolumn{7}{c}{\cellcolor[HTML]{EEEEEE}\textbf{Robustness}}\\
+Components                           & \cellcolor[HTML]{99FF99}2
+                                     & \cellcolor[HTML]{99DD99}3
+                                     & \cellcolor[HTML]{99FF99}2
+                                     & \cellcolor[HTML]{99AA99}4  
+                                     & \cellcolor[HTML]{99AA99}4 
+                                     & \cellcolor[HTML]{99AA99}4 \\
+Interactions                         & \cellcolor[HTML]{99FF99}2    
+                                     & \cellcolor[HTML]{99DD99}3 
+                                     & \cellcolor[HTML]{99DD99}3      
+                                     & \cellcolor[HTML]{99AA99}4  
+                                     & \cellcolor[HTML]{99AA99}4      
+                                     & \cellcolor[HTML]{99AA99}4 \\
+Debugging                            & \cellcolor[HTML]{99AA99}0          
+                                     & \cellcolor[HTML]{99FF99}0
+                                     & \cellcolor[HTML]{99AA99}1
+                                     & \cellcolor[HTML]{99FF99}0
+                                     & \cellcolor[HTML]{99AA99}1
+                                     & \cellcolor[HTML]{99FF99}0 \\
+\textbf{Total}                       & \cellcolor[HTML]{99FF99}\textbf{4}
+                                     & \cellcolor[HTML]{99CC99}\textbf{6}
+                                     & \cellcolor[HTML]{99CC99}\textbf{6}
+                                     & \cellcolor[HTML]{99BB99}\textbf{8}
+                                     & \cellcolor[HTML]{99AA99}\textbf{9}
+                                     & \cellcolor[HTML]{99BB99}\textbf{8} \\
+\multicolumn{7}{c}{\cellcolor[HTML]{EEEEEE}\textbf{Modularity}}\\
+New engine type                      & \cellcolor[HTML]{99AA99}3
+                                     & \cellcolor[HTML]{99DD99}2    
+                                     & \cellcolor[HTML]{99FF99}1
+                                     & \cellcolor[HTML]{99DD99}2
+                                     & \cellcolor[HTML]{99DD99}2
+                                     & \cellcolor[HTML]{99AA99}3 \\
+New engine version                   & \cellcolor[HTML]{99AA99}1
+                                     & \cellcolor[HTML]{99FF99}0
+                                     & \cellcolor[HTML]{99FF99}0
+                                     & \cellcolor[HTML]{99FF99}0
+                                     & \cellcolor[HTML]{99FF99}0
+                                     & \cellcolor[HTML]{99FF99}0 \\
+New infrastructure                   & \cellcolor[HTML]{99FF99}2
+                                     & \cellcolor[HTML]{99FF99}2
+                                     & \cellcolor[HTML]{99FF99}2
+                                     & \cellcolor[HTML]{99FF99}2
+                                     & \cellcolor[HTML]{99FF99}2
+                                     & \cellcolor[HTML]{99FF99}2 \\
+New workflow                         & \cellcolor[HTML]{99FF99}1
+                                     & \cellcolor[HTML]{99FF99}1
+                                     & \cellcolor[HTML]{99FF99}1
+                                     & \cellcolor[HTML]{99FF99}1
+                                     & \cellcolor[HTML]{99FF99}1
+                                     & \cellcolor[HTML]{99AA99}2 \\
+Meta-workflow                        & \cellcolor[HTML]{99AA99}1
+                                     & \cellcolor[HTML]{99AA99}1  
+                                     & \cellcolor[HTML]{99AA99}1
+                                     & \cellcolor[HTML]{99AA99}1
+                                     & \cellcolor[HTML]{99FF99}0
+                                     & \cellcolor[HTML]{99AA99}1 \\
+\textbf{Total}                       & \cellcolor[HTML]{99AA99}\textbf{8}
+                                     & \cellcolor[HTML]{99EE99}\textbf{6}
+                                     & \cellcolor[HTML]{99FF99}\textbf{5}
+                                     & \cellcolor[HTML]{99EE99}\textbf{6}
+                                     & \cellcolor[HTML]{99FF99}\textbf{5}
+                                     & \cellcolor[HTML]{99AA99}\textbf{8} \\
+\multicolumn{7}{c}{\cellcolor[HTML]{EEEEEE}\textbf{Scalability}}\\
+New engine instance                  & \cellcolor[HTML]{99AA99}1
+                                     & \cellcolor[HTML]{99AA99}1
+                                     & \cellcolor[HTML]{99FF99}0
+                                     & \cellcolor[HTML]{99FF99}0
+                                     & \cellcolor[HTML]{99AA99}1
+                                     & \cellcolor[HTML]{99AA99}1 \\
+Elastic engines                      & \cellcolor[HTML]{99AA99}2
+                                     & \cellcolor[HTML]{99DD99}1
+                                     & \cellcolor[HTML]{99FF99}0
+                                     & \cellcolor[HTML]{99FF99}0
+                                     & \cellcolor[HTML]{99DD99}1
+                                     & \cellcolor[HTML]{99DD99}1 \\
+Distributed engines                  & \cellcolor[HTML]{99AA99}1
+                                     & \cellcolor[HTML]{99AA99}1
+                                     & \cellcolor[HTML]{99AA99}1
+                                     & \cellcolor[HTML]{99AA99}1
+                                     & \cellcolor[HTML]{99FF99}0
+                                     & \cellcolor[HTML]{99AA99}1 \\
+Task scheduling                      & \cellcolor[HTML]{99FF99}0
+                                     & \cellcolor[HTML]{99FF99}0
+                                     & \cellcolor[HTML]{99AA99}1
+                                     & \cellcolor[HTML]{99FF99}0
+                                     & \cellcolor[HTML]{99FF99}0
+                                     & \cellcolor[HTML]{99FF99}0 \\
+\textbf{Total}                       & \cellcolor[HTML]{99AA99}\textbf{4}
+                                     & \cellcolor[HTML]{99BB99}\textbf{3}
+                                     & \cellcolor[HTML]{99DD99}\textbf{2}
+                                     & \cellcolor[HTML]{99FF99}\textbf{1}
+                                     & \cellcolor[HTML]{99DD99}\textbf{2}
+                                     & \cellcolor[HTML]{99BB99}\textbf{3} \\
+\multicolumn{7}{c}{\cellcolor[HTML]{EEEEEE}\textbf{Grand total}}\\
+                                     & \cellcolor[HTML]{99CC99}\textbf{19}
+                                     & \cellcolor[HTML]{99BB99}\textbf{18}
+                                     & \cellcolor[HTML]{99FF99}\textbf{16}
+                                     & \cellcolor[HTML]{99CC99}\textbf{19}
+                                     & \cellcolor[HTML]{99DD99}\textbf{20}
+                                     & \cellcolor[HTML]{99AA99}\textbf{23}
+\end{tabular}
+\caption{Architecture evaluation. Lower values (brighter colors) indicate better performance.}
+\label{table:evaluation}
+\end{table*}
+
 \section{Discussion}
@@
 
+\section*{References}
+
 \bibliographystyle{elsarticle-num} 

### 2016-04-12 figures/export.sh
Renamed images folder into figures

@@ -3,7 +3,7 @@
 for name in tight service sub-task agent nested-1 nested-2 import-1 import-2
 do
   echo "Exporting ${name}.svg"
-  inkscape -D -z --file=images/${name}.svg --export-pdf=images/${name}.pdf --export-latex
-  sed s,${name}\.pdf,images/${name}\.pdf,g images/${name}.pdf_tex > images/${name}.pdf_tex.new
-  \mv images/${name}.pdf_tex.new images/${name}.pdf_tex
+  inkscape -D -z --file=figures/${name}.svg --export-pdf=figures/${name}.pdf --export-latex
+  sed s,${name}\.pdf,figures/${name}\.pdf,g figures/${name}.pdf_tex > figures/${name}.pdf_tex.new
+  \mv figures/${name}.pdf_tex.new figures/${name}.pdf_tex
 done

### 2016-04-12 paper.tex
Renamed images folder into figures
Wrote section 4.3

@@
 \section{Introduction}
+
+\todo{Say that the review is done based on our experience with VIP,
+  CBRAIN, and to some extent SHIWA.}
 
@@
 \def\svgwidth{0.5\columnwidth}
-\input{images/tight.pdf_tex}
+\input{figures/tight.pdf_tex}
 \caption{Tight integration}
@@
 See Figure~\ref{archi:service}. The workflow engine is available
-externally to the science gateway. The science gateway controls it
-through a specific interaction that might be implemented as a
-web-service call (e.g. RESTful or SOAP), as a command-line or as any
-other method that separates the workflow engine from the science
-gateway. The workflow engine might be invoked either as a black box
-that completely masks the infrastructure and workflow activities, or
-as a grey box that allows for some interaction with them.  The
-workflow engine is responsible for controlling the tasks on the
-infrastructure, and performing the required data transfers to execute
-them. User data is usually managed through the science gateway,
-although it might as well be delivered by the workflow engine directly
-to the user.
+externally to the science gateway, in a service. The science gateway
+controls the service through a specific interaction that might be
+implemented as a web-service call (e.g. RESTful or SOAP), as a
+command-line or as any other method that offers a well-defined
+interface to the workflow engine. The workflow engine might be invoked
+either as a black box that completely masks the infrastructure and
+workflow activities, or as a grey box that allows for some interaction
+with them.  The workflow engine is responsible for controlling the
+tasks on the infrastructure, and performing the required data
+transfers to execute them. User data is usually managed through the
+science gateway, although it might as well be delivered by the
+workflow engine directly to the user.
 
@@
 \def\svgwidth{0.5\columnwidth}
-\input{images/service.pdf_tex}
+\input{figures/service.pdf_tex}
 \caption{Service integration}
@@
 \def\svgwidth{0.5\columnwidth}
-\input{images/sub-task.pdf_tex}
+\input{figures/sub-task.pdf_tex}
 \caption{Sub-tasking}
@@
 \def\svgwidth{0.8\columnwidth}
-\input{images/agent.pdf_tex}
+\input{figures/agent.pdf_tex}
 \caption{Pool model}
@@
 \def\svgwidth{0.48\columnwidth}
-\input{images/nested-1.pdf_tex}
+\input{figures/nested-1.pdf_tex}
 \def\svgwidth{0.48\columnwidth}
-\input{images/nested-2.pdf_tex}
+\input{figures/nested-2.pdf_tex}
 \caption{Nested workflows. Left: abstract model. Right: instantiation with service invocation.}
@@
 \def\svgwidth{0.48\columnwidth}
-\input{images/import-1.pdf_tex}
+\input{figures/import-1.pdf_tex}
 \def\svgwidth{0.48\columnwidth}
-\input{images/import-2.pdf_tex}
+\input{figures/import-2.pdf_tex}
 \caption{Workflow import. Left: abstract model. Right: instantiation with service invocation.}
@@
 
-\subsection{Criteria}
+See Table~\ref{table:evaluation}.
+
+We focus on the integration of workflow engines in science gateways,
+i.e. we assume that the science gateway is already interfaced with the
+infrastructure.
+
+\subsection{Evaluation metrics}
 
@@
 \item Interactions: number of unique interactions involved in workflow executions. 
-\item Debugging: availability of debugging information about workflow activities
-  (1: difficult to obtain; 0: easy to access).
+\item Debugging: availability of debugging information about workflow
+  activities (1: difficult to obtain; 0: easy to access). Fine-grained
+  information about workflow activity is required to properly
+  troubleshoot workflow executions.
 \end{itemize}
@@
 \item New engine type: number of interactions or components to modify
-  to integrate a new workflow engine type in the architecture.
+  to integrate a new workflow engine type in the
+  architecture. Modification of an interaction is deemed necessary
+  when the parameters involved in this interaction are
+  modified. Modification of a component is required when its code
+  needs to be modified or recompiled.
 \item New engine version: number of interactions or components to
@@
   architecture, assuming that another version of the same engine type
-  is already available.
+  is already available. We assume that different versions of a
+  workflow engine share the same interface, i.e. they can be invoked
+  using the same software. When this is not the case, the different
+  versions have to be considered as different engine types.
 \item New infrastructure: number of interactions or components to
@@
 \item New engine instance (0: automated; 1: manual intervention needed).
-\item Elastic engines (0: easy to implement; 1: difficult to implement).
+\item Elastic engines (0: easy to implement; 1: difficult to
+  implement; 2: not available).
 \item Distributed engines (0: possible; 1: not possible).
@@
 
-\subsection{Evaluation}
+\subsection{Tight integration}
 
-See Table~\ref{table:evaluation}.
+\subsection{Service invocation}
 
-We focus on the integration of workflow engines in science gateways,
-i.e. we assume that the science gateway is already interfaced with the
-infrastructure.
+\paragraph{Integration} Integrating a workflow engine as a service
+requires to develop 3 interactions and 1 component. First, a workflow
+service has to be developed and interfaced with the infrastructure
+through interactions \texttt{b} and \texttt{c$_2$}
+(\texttt{I$_2$=3}). In addtion, the science gateway has to be extended
+to call the engine interface \texttt{d} (\texttt{I$_1$=1}).
 
-\paragraph{Service invocation} Integrating a workflow engine as a
-service requires 3 interactions. First, the science gateway has to be
-extended to call the engine interface (\texttt{d}), which might
-require some adaptation on the workflow engine too, for instance
-wrapping in a web service. In addition, the workflow engine has to be
-interfaced with the infrastructure, which requires the two
-interactions \texttt{b} and \texttt{$c_2$}.
+\paragraph{Robustness} Workflow execution requires 3 components: the
+science gateway, the workflow service, and the infrastructure
+(\texttt{R$_1$=3}). It involves the 4 interactions \texttt{b},
+\texttt{$c_1$}, \texttt{$c_2$} and \texttt{d}
+(\texttt{R$_2$=4}). Fine-grained debugging information is usually easy
+to obtain since the workflow service provides direct access to the
+engine (\texttt{R$_3$=0}).
 
-+ Easy to implement in the SG (just call the service)
-- Load-balancing between services is difficult
-(refer to VIP's experience).
-- Need to wrap workflow engines in
-services.
-- Need to know where to submit each workflow, e.g. when
-multiple workflows are involved.
-- The science gateway has no
-knowledge of the detailed progress of the workflow, in particular
-job-level information (statuses, stdout, stderr). If we wanted to do
-that, specific feedback channels have to be developed, which requires more work (see VIP).
+\paragraph{Modularity} Adding a new type of workflow engine requires
+to implement the corresponding workflow service, to modify interaction
+\texttt{d}, and to implement interactions \texttt{b} and
+\texttt{$c_2$} (\texttt{M$_1$=4}). New engine versions can be added by
+updating the workflow service without modifying any interaction
+(\texttt{M$_2$=0}). Adding a new type of infrastructure requires
+updates in interactions \texttt{b}, \texttt{$c_1$} and \texttt{$c_2$}
+(\texttt{M$_3$=3}). New workflows are added in the science gateway or
+in the workflow engine through interaction \texttt{a} only
+(\texttt{M$_4$=1}). Meta-workflow are not supported by default
+(\texttt{M$_5$=1}).
 
-\paragraph{Sub-task}
+\paragraph{Scalability} The service architecture supports multiple
+engine instances through multiple workflow services. In VIP for
+instance, this feature has been available from release 1.17 (April
+2016). A basic load-balancing mechanism is available that sends new
+workflow executions to the engine instance that has the least active
+executions. To avoid ``black-hole'' syndromes created by failing
+engine instances, engine instances are automatically disabled when
+workflows cannot be submitted to them. Adding a new engine instance,
+however, requires manual intervention to declare the new instance in
+the science gateway (\texttt{S$_1$=1}). Consequently, elastic engines
+are difficult to implement because they require a mechanism to update
+the science gateway configuration when a new engine instance is
+available (\texttt{S$_2$=1}). Distributing the execution of a single
+workflow in multiple engines is usually not possible unless the
+workflow engine has specific abilities (\texttt{S$_3$=1}). The
+scheduling of tasks on the infrastructure is as complex as in any
+other architecture since the workflow engine might implement any kind
+of scheduling policy (\texttt{S$_4$=0}).
+
+\subsection{Sub-task}
 
@@
 
-\paragraph{Pool}
+\subsection{Pool}
 
@@
 - Needs a complete framework (the pool). Most likely third-party software as it's a lot of work to implement.
+
+\subsection{Nested workflows}
+
+\subsection{Workflow import}
 
@@
 \multicolumn{7}{c}{\cellcolor[HTML]{EEEEEE}\textbf{Integration effort}}\\
-From science gateway                 & \multirow{2}{*}{3}
+Science gateway -- \texttt{I$_1$}    & \multirow{2}{*}{3}
                                      & \cellcolor[HTML]{99FF99}1
@@
                                      & \cellcolor[HTML]{99AA99}3 \\
-From workflow engine                 &
-                                     & \cellcolor[HTML]{99FF99}1 
+Workflow engine -- \texttt{I$_2$}    &
+                                     & \cellcolor[HTML]{99FF99}3 
                                      & \cellcolor[HTML]{99FF99}2
@@
 \textbf{Total}                       & \cellcolor[HTML]{99FF99}\textbf{3}
-                                     & \cellcolor[HTML]{99FF99}\textbf{3}
+                                     & \cellcolor[HTML]{99FF99}\textbf{4}
                                      & \cellcolor[HTML]{99FF99}\textbf{3}
@@
 \multicolumn{7}{c}{\cellcolor[HTML]{EEEEEE}\textbf{Robustness}}\\
-Components                           & \cellcolor[HTML]{99FF99}2
+Components  --   \texttt{R$_1$}      & \cellcolor[HTML]{99FF99}2
                                      & \cellcolor[HTML]{99DD99}3
@@
                                      & \cellcolor[HTML]{99AA99}4 \\
-Interactions                         & \cellcolor[HTML]{99FF99}2    
-                                     & \cellcolor[HTML]{99DD99}3 
+Interactions -- \texttt{R$_2$}       & \cellcolor[HTML]{99FF99}3    
+                                     & \cellcolor[HTML]{99DD99}4
                                      & \cellcolor[HTML]{99DD99}3      
@@
                                      & \cellcolor[HTML]{99AA99}4 \\
-Debugging                            & \cellcolor[HTML]{99AA99}0          
+Debugging --    \texttt{R$_3$}       & \cellcolor[HTML]{99AA99}0          
                                      & \cellcolor[HTML]{99FF99}0
@@
 \textbf{Total}                       & \cellcolor[HTML]{99FF99}\textbf{4}
-                                     & \cellcolor[HTML]{99CC99}\textbf{6}
+                                     & \cellcolor[HTML]{99CC99}\textbf{7}
                                      & \cellcolor[HTML]{99CC99}\textbf{6}
@@
 \multicolumn{7}{c}{\cellcolor[HTML]{EEEEEE}\textbf{Modularity}}\\
-New engine type                      & \cellcolor[HTML]{99AA99}3
-                                     & \cellcolor[HTML]{99DD99}2    
+New engine type -- \texttt{M$_1$}    & \cellcolor[HTML]{99AA99}3
+                                     & \cellcolor[HTML]{99DD99}4
                                      & \cellcolor[HTML]{99FF99}1
@@
                                      & \cellcolor[HTML]{99AA99}3 \\
-New engine version                   & \cellcolor[HTML]{99AA99}1
+New engine version -- \texttt{M$_2$} & \cellcolor[HTML]{99AA99}1
                                      & \cellcolor[HTML]{99FF99}0
@@
                                      & \cellcolor[HTML]{99FF99}0 \\
-New infrastructure                   & \cellcolor[HTML]{99FF99}2
-                                     & \cellcolor[HTML]{99FF99}2
+New infrastructure -- \texttt{M$_3$} & \cellcolor[HTML]{99FF99}2
+                                     & \cellcolor[HTML]{99FF99}3
                                      & \cellcolor[HTML]{99FF99}2
@@
                                      & \cellcolor[HTML]{99FF99}2 \\
-New workflow                         & \cellcolor[HTML]{99FF99}1
+New workflow -- \texttt{M$_4$}       & \cellcolor[HTML]{99FF99}1
                                      & \cellcolor[HTML]{99FF99}1
@@
                                      & \cellcolor[HTML]{99AA99}2 \\
-Meta-workflow                        & \cellcolor[HTML]{99AA99}1
+Meta-workflow  -- \texttt{M$_5$}     & \cellcolor[HTML]{99AA99}1
                                      & \cellcolor[HTML]{99AA99}1  
@@
 \textbf{Total}                       & \cellcolor[HTML]{99AA99}\textbf{8}
-                                     & \cellcolor[HTML]{99EE99}\textbf{6}
+                                     & \cellcolor[HTML]{99EE99}\textbf{9}
                                      & \cellcolor[HTML]{99FF99}\textbf{5}
@@
 \multicolumn{7}{c}{\cellcolor[HTML]{EEEEEE}\textbf{Scalability}}\\
-New engine instance                  & \cellcolor[HTML]{99AA99}1
+New engine instance -- \texttt{S$_1$}& \cellcolor[HTML]{99AA99}1
                                      & \cellcolor[HTML]{99AA99}1
@@
                                      & \cellcolor[HTML]{99AA99}1 \\
-Elastic engines                      & \cellcolor[HTML]{99AA99}2
+Elastic engines -- \texttt{S$_2$}    & \cellcolor[HTML]{99AA99}2
                                      & \cellcolor[HTML]{99DD99}1
@@
                                      & \cellcolor[HTML]{99DD99}1 \\
-Distributed engines                  & \cellcolor[HTML]{99AA99}1
+Distributed engines -- \texttt{S$_3$}& \cellcolor[HTML]{99AA99}1
                                      & \cellcolor[HTML]{99AA99}1
@@
                                      & \cellcolor[HTML]{99AA99}1 \\
-Task scheduling                      & \cellcolor[HTML]{99FF99}0
+Task scheduling -- \texttt{S$_4$}    & \cellcolor[HTML]{99FF99}0
                                      & \cellcolor[HTML]{99FF99}0
@@
                                      & \cellcolor[HTML]{99CC99}\textbf{19}
-                                     & \cellcolor[HTML]{99BB99}\textbf{18}
+                                     & \cellcolor[HTML]{99BB99}\textbf{23}
                                      & \cellcolor[HTML]{99FF99}\textbf{16}
@@
 \section{Discussion}
+
+Comparison between architectures, per criterion (robustness, etc).
+
+This ignores pre-existing work (e.g. a workflow engine is already available as a service). Migration across architectures too.
 

### 2016-04-12 paper.tex
Wrote section 4.4

@@
 
+\todo{Interactions c1 and c2 are merged in c since this is done
+  completely by the science gateway}
+
 This model is implemented in CBRAIN~\cite{SHER-14} where it is used to
@@
 
-\emph{Integration effort} measures the development effort required to build
-the system, assuming that none of the components could initially
-communicate with each other. It is measured by counting the number of
-interactions to develop and components to modify on the architectural
-diagram. It breaks down to 2 metrics that relate to the direction of
-interaction arrows on the architectural diagrams:
+\todo{Refer to the metric codes.}
+
+\emph{Integration} measures the development effort required to build
+the architecture, assuming that the science gateway and infrastructure
+are already integrated. It is measured by counting on the
+architectural diagram the number of interactions and components to
+develop or modify. It breaks down to 4 metrics that quantify the
+developments related to:
 \begin{itemize}[itemsep=0cm]
-\item Interactions from the science gateway. 
-\item Iteractions from the workflow engine. 
+\item Science gateway (\texttt{I$_1$}),
+\item Workflow engine (\texttt{I$_2$}),
+\item Infrastructure (\texttt{I$_3$}),
+\item Other components (\texttt{I$_4$}).
 \end{itemize}
+Interactions are counted only once, in the component from which they
+originate.
 
@@
   modified. Modification of a component is required when its code
-  needs to be modified or recompiled.
+  needs to be modified or recompiled (science gateway or workflow
+  service), or when a new piece of software has to be installed
+  (infrastructure only).
 \item New engine version: number of interactions or components to
@@
 through interactions \texttt{b} and \texttt{c$_2$}
-(\texttt{I$_2$=3}). In addtion, the science gateway has to be extended
-to call the engine interface \texttt{d} (\texttt{I$_1$=1}).
+(\texttt{I$_2$}=3). In addtion, the science gateway has to be extended
+to call the engine interface \texttt{d} (\texttt{I$_1$}=1).
 
@@
 science gateway, the workflow service, and the infrastructure
-(\texttt{R$_1$=3}). It involves the 4 interactions \texttt{b},
+(\texttt{R$_1$}=3). It involves the 4 interactions \texttt{b},
 \texttt{$c_1$}, \texttt{$c_2$} and \texttt{d}
-(\texttt{R$_2$=4}). Fine-grained debugging information is usually easy
+(\texttt{R$_2$}=4). Fine-grained debugging information is usually easy
 to obtain since the workflow service provides direct access to the
-engine (\texttt{R$_3$=0}).
+engine (\texttt{R$_3$}=0).
 
@@
 \texttt{d}, and to implement interactions \texttt{b} and
-\texttt{$c_2$} (\texttt{M$_1$=4}). New engine versions can be added by
+\texttt{$c_2$} (\texttt{M$_1$}=4). New engine versions can be added by
 updating the workflow service without modifying any interaction
-(\texttt{M$_2$=0}). Adding a new type of infrastructure requires
+(\texttt{M$_2$}=0). Adding a new type of infrastructure requires
 updates in interactions \texttt{b}, \texttt{$c_1$} and \texttt{$c_2$}
-(\texttt{M$_3$=3}). New workflows are added in the science gateway or
+(\texttt{M$_3$}=3). New workflows are added in the science gateway or
 in the workflow engine through interaction \texttt{a} only
-(\texttt{M$_4$=1}). Meta-workflow are not supported by default
-(\texttt{M$_5$=1}).
+(\texttt{M$_4$}=1). Meta-workflow are not supported by default
+(\texttt{M$_5$}=1).
 
@@
 however, requires manual intervention to declare the new instance in
-the science gateway (\texttt{S$_1$=1}). Consequently, elastic engines
+the science gateway (\texttt{S$_1$}=1). Consequently, elastic engines
 are difficult to implement because they require a mechanism to update
 the science gateway configuration when a new engine instance is
-available (\texttt{S$_2$=1}). Distributing the execution of a single
+available (\texttt{S$_2$}=1). Distributing the execution of a single
 workflow in multiple engines is usually not possible unless the
-workflow engine has specific abilities (\texttt{S$_3$=1}). The
+workflow engine has specific abilities (\texttt{S$_3$}=1). The
 scheduling of tasks on the infrastructure is as complex as in any
 other architecture since the workflow engine might implement any kind
-of scheduling policy (\texttt{S$_4$=0}).
+of scheduling policy (\texttt{S$_4$}=0).
 
@@
 
-Science gateway has to make sure that the machine on which a workflow
-task executes is reliable as several sub-tasks will depend on the
-success of this task. Or workflow engine task has to be recoverable.
+\paragraph{Integration} Integrating a workflow engine as a sub-task
+only requires to implement interaction \texttt{e} (\texttt{I$_2$}=1)
+and to install the workflow engine on the infrastructure
+(\texttt{I$_3$}=1). A new task type is created in the science gateway
+only when a new workflow has to be integrated (see M$_4$), but this is
+not required to integrate the engine itself (\texttt{I$_1$}=0).
 
-Performance: there could be a deadlock when there is contention. Also,
-the workflow task has to be reliable (for instance, proper walltime
-estimation), which is difficult because these tasks are
-long. Restarting a workflow task may mean that all the tasks in the
-workflow have to be restarted, depending on the capability of the
-workflow engine. This complexifies task scheduling (a system could
-ignore these issues but they may backfire).
+\paragraph{Robustness} Only 2 components and 3 interactions are
+involved in workflow execution (\texttt{R$_1$}=2,
+\texttt{R$_2$}=3). Obtaining fine-grained information about workflow
+activities is not straightforward since the science gateway has no
+knowledge about the workflow topology, and the workflow engine is
+integrated as a task (\texttt{R$_3$}=1).
+
+\paragraph{Modularity} Integrating a new type of workflow engine
+requires to develop interaction \texttt{e} and to install the engine
+on the infrastructure (\texttt{M$_1$}=2). Updating an engine version
+requires updates only on the infrastructure (\texttt{M$_2$}=1). Adding
+a new infrastructure requires to update interactions \texttt{b} and
+\texttt{c} in the science gateway (\texttt{M$_3$}=2). New workflows are integrated by
+creating a new task in the science gateway through interaction
+\texttt{a} (\texttt{M$_4$}=1), and meta-workflows are not available
+(\texttt{M$_5$}=1).
+
+\paragraph{Scalability}
+New engine instances are spawned and executed on the infrastructure as
+any other task upon user submission (\texttt{S$_1$}=0,
+\texttt{S$_2$}=0). Such scalability properties are one of the major
+interests of the sub-task architecture. Distributed engines are not
+supported by default (\texttt{S$_3$}=1). Task scheduling is slightly
+more complex than in the other approaches due to the special role of
+the task that executes the workflow engine (\texttt{S$_4$}=1). Indeed,
+the reliability of this task is critical since all the sub-tasks in
+the workflow depend on it and, depending on the recovery capabilities
+of the workflow engine, may need to be resubmitted if the workflow
+task fails. The workflow task is also longer than all its sub-tasks,
+which increases its chances of failure. In addition, task parameters,
+for instance estimated walltime, are more difficult to estimate for
+the workflow task than for sub-tasks which may generate issues such as
+selection of wrong batch queues on clusters. Finally, the
+interdependencies between the workflow task and its sub-tasks may
+create deadlocks when there is contention. Say for instance that only
+1 computing resource is available for the science gateway and that the
+workflow task is running on it and submits sub-tasks, then the
+sub-tasks could only execute when the resource is available, which
+will never happen because the workflow task will not complete until
+the sub-tasks complete. This configuration can be generalized to an
+infrastructure with $n$ resources where $n$ workflows are
+submitted. In practice, however, the number of submitted workflows
+usually remains lower than the number of computing resources available
+on this infrastructure, which makes such deadlocks unlikely to happen.
 
@@
                                      & \textbf{Import} \\
-\multicolumn{7}{c}{\cellcolor[HTML]{EEEEEE}\textbf{Integration effort}}\\
+\multicolumn{7}{c}{\cellcolor[HTML]{EEEEEE}\textbf{Integration}}\\
 Science gateway -- \texttt{I$_1$}    & \multirow{2}{*}{3}
                                      & \cellcolor[HTML]{99FF99}1
-                                     & \cellcolor[HTML]{99FF99}2  
+                                     & \cellcolor[HTML]{99FF99}0  
                                      & \cellcolor[HTML]{99FF99}2
@@
                                      & \cellcolor[HTML]{99FF99}3 
-                                     & \cellcolor[HTML]{99FF99}2
+                                     & \cellcolor[HTML]{99FF99}1
                                      & \cellcolor[HTML]{99AA99}2
@@
                                      & \cellcolor[HTML]{99FF99}1 \\
+Infrastructure -- \texttt{I$_3$}   & \cellcolor[HTML]{99FF99}0
+                                     & \cellcolor[HTML]{99FF99}0
+                                     & \cellcolor[HTML]{99FF99}1
+                                     & \cellcolor[HTML]{99AA99}0
+                                     & \cellcolor[HTML]{99AA99}0
+                                     & \cellcolor[HTML]{99FF99}0 \\
+Other -- \texttt{I$_4$}              & \cellcolor[HTML]{99FF99}0
+                                     & \cellcolor[HTML]{99FF99}0
+                                     & \cellcolor[HTML]{99FF99}0
+                                     & \cellcolor[HTML]{99AA99}2
+                                     & \cellcolor[HTML]{99AA99}0
+                                     & \cellcolor[HTML]{99FF99}0 \\
 \textbf{Total}                       & \cellcolor[HTML]{99FF99}\textbf{3}
                                      & \cellcolor[HTML]{99FF99}\textbf{4}
-                                     & \cellcolor[HTML]{99FF99}\textbf{3}
+                                     & \cellcolor[HTML]{99FF99}\textbf{2}
                                      & \cellcolor[HTML]{99AA99}\textbf{4}
@@
                                      & \cellcolor[HTML]{99DD99}4
-                                     & \cellcolor[HTML]{99FF99}1
+                                     & \cellcolor[HTML]{99FF99}2
                                      & \cellcolor[HTML]{99DD99}2
@@
                                      & \cellcolor[HTML]{99FF99}0
-                                     & \cellcolor[HTML]{99FF99}0
+                                     & \cellcolor[HTML]{99FF99}1
                                      & \cellcolor[HTML]{99FF99}0
@@
                                      & \cellcolor[HTML]{99EE99}\textbf{9}
-                                     & \cellcolor[HTML]{99FF99}\textbf{5}
+                                     & \cellcolor[HTML]{99FF99}\textbf{7}
                                      & \cellcolor[HTML]{99EE99}\textbf{6}
@@
                                      & \cellcolor[HTML]{99BB99}\textbf{23}
-                                     & \cellcolor[HTML]{99FF99}\textbf{16}
+                                     & \cellcolor[HTML]{99FF99}\textbf{17}
                                      & \cellcolor[HTML]{99CC99}\textbf{19}

### 2016-04-12 paper.tex
Wrote section 4.5

@@
 
-\todo{Introduce the notion of architectural diagram with an abstract figure.}
+\todo{Introduce the notion of architectural diagram with an abstract figure. Mention arrows, boxes, box with dotted lines, red color.}
 
@@
 
-= Evaluation
-+ Scale-up and load balancing
-+ Multi-language
-- Needs a complete framework (the pool). Most likely third-party software as it's a lot of work to implement.
+\paragraph{Integration} Integrating workflow engines through the pool
+model requires interaction \texttt{d} in the science gateway
+(\texttt{I$_1$=1}), interactions \texttt{b} and \texttt{c$_2$} in the
+workflow engine (\texttt{I$_2$=2}), the development of the workflow
+pool, agent and interaction \texttt{f} between them
+(\texttt{I$_4$=3}).
+
+\paragraph{Robustness} The science gateway, workflow pool, agent and
+infrastructure are involved in a workflow execution
+(\texttt{R$_1$=4}), and these components are connected through 4
+interactions (\texttt{R$_2$=4}). Accessing debugging information is
+not likely to be an issue since the workflow pool could implement
+specific functions for that (\texttt{R$_3$=0}).
+
+\paragraph{Modularity} Adding a new engine type requires to wrap the
+engine in the agent and to update interactions \texttt{b} and
+\texttt{c$_2$} (\texttt{M$_1$=3}). Updating the version of an engine
+is transparent (\texttt{M$_2$=0}), and integrating a new
+infrastructure requires updates in interactions \texttt{b},
+\texttt{c$_1$} and \texttt{c$_2$} (\texttt{M$_3$=3}). Integrating a
+new workflow is done through interaction \texttt{a} only
+(\texttt{M$_4$=1}) and meta-workflows are not available by default
+(\texttt{M$_5$=1}).
+
+\paragraph{Scalability} New engine instances only require new agents,
+which is easily automated (\texttt{S$_1$=0}) and by design very
+suitable for elastic computing (\texttt{S$_2$=0}). For instance,
+auto-scaling rules can be implemented to start new agents when the
+workload in the science gateway exceeds a certain threshold \todo{ref
+  needed}. Distributed engines are not available by default
+(\texttt{S$_3$=1}) and task scheduling is as complex as in any other
+architecture (\texttt{S$_4$=0}).
 
@@
                                      & \cellcolor[HTML]{99FF99}0  
-                                     & \cellcolor[HTML]{99FF99}2
+                                     & \cellcolor[HTML]{99FF99}1
                                      & \cellcolor[HTML]{99FF99}2
@@
                                      & \cellcolor[HTML]{99FF99}0
-                                     & \cellcolor[HTML]{99AA99}2
+                                     & \cellcolor[HTML]{99AA99}3
                                      & \cellcolor[HTML]{99AA99}0
@@
                                      & \cellcolor[HTML]{99FF99}\textbf{2}
-                                     & \cellcolor[HTML]{99AA99}\textbf{4}
+                                     & \cellcolor[HTML]{99AA99}\textbf{6}
                                      & \cellcolor[HTML]{99AA99}\textbf{4}
@@
                                      & \cellcolor[HTML]{99FF99}2
-                                     & \cellcolor[HTML]{99DD99}2
+                                     & \cellcolor[HTML]{99DD99}3
                                      & \cellcolor[HTML]{99DD99}2
@@
                                      & \cellcolor[HTML]{99FF99}2
-                                     & \cellcolor[HTML]{99FF99}2
+                                     & \cellcolor[HTML]{99FF99}3
                                      & \cellcolor[HTML]{99FF99}2
@@
                                      & \cellcolor[HTML]{99FF99}\textbf{7}
-                                     & \cellcolor[HTML]{99EE99}\textbf{6}
+                                     & \cellcolor[HTML]{99EE99}\textbf{8}
                                      & \cellcolor[HTML]{99FF99}\textbf{5}
@@
                                      & \cellcolor[HTML]{99FF99}\textbf{17}
-                                     & \cellcolor[HTML]{99CC99}\textbf{19}
+                                     & \cellcolor[HTML]{99CC99}\textbf{23}
                                      & \cellcolor[HTML]{99DD99}\textbf{20}
@@
 \end{tabular}
-\caption{Architecture evaluation. Lower values (brighter colors) indicate better performance.}
+\caption{Architecture evaluation.
+Lower values (brighter colors) indicate better performance.
+\todo{update colors and find a more relevant way to compute a grand total without favoring criteria that have more metrics than the others (e.g.
+normalize each criterion between 0 and 1).}}
 \label{table:evaluation}

### 2016-04-12 paper.tex
Wrote section 4.2

@@
 
-\todo{Introduce the notion of architectural diagram with an abstract figure. Mention arrows, boxes, box with dotted lines, red color.}
+\todo{Introduce the notion of architectural diagram with an abstract
+  figure. Mention arrows, boxes, box with dotted lines, red color,
+  tasks, data.}
 
@@
 
-See Table~\ref{table:evaluation}.
-
-We focus on the integration of workflow engines in science gateways,
-i.e. we assume that the science gateway is already interfaced with the
-infrastructure.
+See Table~\ref{table:evaluation}. We focus on the integration of
+workflow engines in science gateways, i.e. we assume that the science
+gateway is already interfaced with the infrastructure.
 
@@
 \begin{itemize}[itemsep=0cm]
-\item New engine instance (0: automated; 1: manual intervention needed).
+\item New engine instance (0: automated; 1: manual intervention needed; 2: not possible).
 \item Elastic engines (0: easy to implement; 1: difficult to
@@
 \subsection{Tight integration}
+
+\paragraph{Integration} Tightly integrating a workflow engine in a
+science gateway requires to modify the science gateway
+(\texttt{I$_1$=1}) and to develop interactions \texttt{b}, and
+\texttt{c$_2$} in the workflow engine (\texttt{I$_2$=2}). No other
+modification is required (\texttt{I$_3$}=\texttt{I$_4$}=0).
+
+\paragraph{Robustness} The science gateway and the infrastructure are
+involved in a workflow execution (\texttt{R$_1$}=2), with interactions
+\texttt{b}, \texttt{c$_1$} and \texttt{c$_2$}
+(\texttt{R$_2$}=3). Debugging is not an issue since the science
+gateway can retrieve any information from the workflow engine
+directly (\texttt{R$_3$}=0).
+
+\paragraph{Modularity} Integrating a new type of workflow engine
+requires to modify the science gateway as well as interactions
+\texttt{b} and \texttt{c$_2$} (\texttt{M$_1$}=3). Updating a workflow
+engine version requires modifications in the science gateway
+(\texttt{M$_2$}=1). Adding a new infrastructure generates updates in
+interactions \texttt{b}, \texttt{c$_1$} and \texttt{c$_2$}
+(\texttt{M$_3$}=3). Inserting a new workflow is done through
+interaction \texttt{a} (\texttt{M$_4$}=1). Meta-workflows are not
+supported by default (\texttt{M$_4$}=1).
+
+\paragraph{Scalability} Adding a new engine instance requires a new
+instance of the science gateway, which is in general not possible
+since load-balancing mechanisms available in web servers are usually
+meant to deal with short web requests rather than long workflow
+executions (\texttt{S$_1$}=2). Coherently, elastic engines are usually
+not possible either (\texttt{S$_2$}=2).  Distributed engines are not
+available by default (\texttt{S$_3$}=1). The scheduling of tasks on
+the infrastructure is as complex as in any other architecture since
+the workflow engine might implement any kind of scheduling policy
+(\texttt{S$_4$}=0).
 
@@
 \multicolumn{7}{c}{\cellcolor[HTML]{EEEEEE}\textbf{Integration}}\\
-Science gateway -- \texttt{I$_1$}    & \multirow{2}{*}{3}
+Science gateway -- \texttt{I$_1$}    & \cellcolor[HTML]{99FF99}1
                                      & \cellcolor[HTML]{99FF99}1
@@
                                      & \cellcolor[HTML]{99AA99}3 \\
-Workflow engine -- \texttt{I$_2$}    &
+Workflow engine -- \texttt{I$_2$}    & \cellcolor[HTML]{99AA99}2
                                      & \cellcolor[HTML]{99FF99}3 
@@
                                      & \cellcolor[HTML]{99FF99}0 \\
-\textbf{Total}                       & \cellcolor[HTML]{99FF99}\textbf{4}
+\textbf{Total}                       & \cellcolor[HTML]{99FF99}\textbf{5}
                                      & \cellcolor[HTML]{99CC99}\textbf{7}
@@
                                      & \cellcolor[HTML]{99FF99}0 \\
-New infrastructure -- \texttt{M$_3$} & \cellcolor[HTML]{99FF99}2
+New infrastructure -- \texttt{M$_3$} & \cellcolor[HTML]{99FF99}3
                                      & \cellcolor[HTML]{99FF99}3
@@
                                      & \cellcolor[HTML]{99AA99}1 \\
-\textbf{Total}                       & \cellcolor[HTML]{99AA99}\textbf{8}
+\textbf{Total}                       & \cellcolor[HTML]{99AA99}\textbf{9}
                                      & \cellcolor[HTML]{99EE99}\textbf{9}
@@
 \multicolumn{7}{c}{\cellcolor[HTML]{EEEEEE}\textbf{Scalability}}\\
-New engine instance -- \texttt{S$_1$}& \cellcolor[HTML]{99AA99}1
+New engine instance -- \texttt{S$_1$}& \cellcolor[HTML]{99AA99}2
                                      & \cellcolor[HTML]{99AA99}1
@@
                                      & \cellcolor[HTML]{99FF99}0 \\
-\textbf{Total}                       & \cellcolor[HTML]{99AA99}\textbf{4}
+\textbf{Total}                       & \cellcolor[HTML]{99AA99}\textbf{5}
                                      & \cellcolor[HTML]{99BB99}\textbf{3}
@@
 \multicolumn{7}{c}{\cellcolor[HTML]{EEEEEE}\textbf{Grand total}}\\
-                                     & \cellcolor[HTML]{99CC99}\textbf{19}
+                                     & \cellcolor[HTML]{99CC99}\textbf{22}
                                      & \cellcolor[HTML]{99BB99}\textbf{23}

### 2016-04-12 paper.tex
Wrote section 4.6

@@
 
+\todo{instantiation is for foreign workflows, i.e. 2 services. We could have another one with native workflows too.}
+
+\todo{both workflow engines have to access the infrastructure. If they
+  run on 2 infrastructures, c1 has to be duplicated.}
+
+\todo{define parent and child engine}
+
 \begin{figure}
@@
 
-\subsection{Nested workflows}
+\subsection{Nested workflows and service invocation}
+
+\paragraph{Integration} Setting up a nested workflow architecture with
+service invocation requires interaction \texttt{d} in the science
+gateway (\texttt{I$_1$=1}), and the development of 2 workflow services
+with 2 instances of interactions \texttt{b} and \texttt{c$_2$}
+(\texttt{I$_2$=5}).
+
+\paragraph{Robustness} Workflow execution involves the science
+gateway, 2 workflow services and the infrastructure
+(\texttt{R$_1$=4}), which includes 7 interactions
+(\texttt{R$_2$=7}). Debugging is difficult because the science gateway
+cannot directly access fine-grained information in the sub-workflow
+(\texttt{R$_3$=1}).
+
+\paragraph{Modularity} Adding a new type of \emph{parent} engine
+requires to implement the corresponding service, to implement
+interactions \texttt{b} and \texttt{c$_2$} in the parent engine, and
+to implement interaction \texttt{d} in the science gateway and in the
+parent service (\texttt{M$_1$=5}). Adding a new type of \emph{child}
+engine only requires to implement the corresponding service, to
+develop interactions \texttt{b} and \texttt{c$_2$} in the child
+engine, and to implement interaction \texttt{d} in the parent service
+(\texttt{M$_1$=4}). We use \texttt{M$_1$=4.5} in
+Table~\ref{table:evaluation} to reflect both conditions. Adding a new
+version in the parent or child engine only requires modifying this
+engine (\texttt{M$_2$=0}). Adding a new infrastructure requires to
+re-implement interactions \texttt{b} and \texttt{c$_2$} twice, and
+interaction \texttt{c$_1$} once (\texttt{M$_3$=5}). Adding a new
+workflow is done through interaction \texttt{a}
+(\texttt{M$_4$=1}). Meta-workflows are possible, which is one of the
+main interest of this architecture (\texttt{M$_5$=0}).
+
+\paragraph{Scalability} As in the service architecture, adding a new
+workflow instance requires manual configuration in the science gateway
+(instance of a parent engine), or in the parent engine (instance of a
+child engine) (\texttt{S$_1$=1}). Similarly, elastic engines are
+difficult to achieve (\texttt{S$_2$=1}). Distributed engines are
+possible, through meta-workflows (\texttt{S$_3$=0}). Task scheduling
+is more complex than in other architectures though, due to the fact
+that workflow execution is split in different engines (\texttt{S$_4$=1}).
 
@@
                                      & \cellcolor[HTML]{99FF99}1
-                                     & \cellcolor[HTML]{99FF99}2
+                                     & \cellcolor[HTML]{99FF99}1
                                      & \cellcolor[HTML]{99AA99}3 \\
@@
                                      & \cellcolor[HTML]{99AA99}2
-                                     & \cellcolor[HTML]{99AA99}2
+                                     & \cellcolor[HTML]{99AA99}5
                                      & \cellcolor[HTML]{99FF99}1 \\
@@
                                      & \cellcolor[HTML]{99AA99}\textbf{6}
-                                     & \cellcolor[HTML]{99AA99}\textbf{4}
+                                     & \cellcolor[HTML]{99AA99}\textbf{6}
                                      & \cellcolor[HTML]{99AA99}\textbf{4} \\
@@
                                      & \cellcolor[HTML]{99AA99}4  
-                                     & \cellcolor[HTML]{99AA99}4      
+                                     & \cellcolor[HTML]{99AA99}7      
                                      & \cellcolor[HTML]{99AA99}4 \\
@@
                                      & \cellcolor[HTML]{99BB99}\textbf{8}
-                                     & \cellcolor[HTML]{99AA99}\textbf{9}
+                                     & \cellcolor[HTML]{99AA99}\textbf{12}
                                      & \cellcolor[HTML]{99BB99}\textbf{8} \\
@@
                                      & \cellcolor[HTML]{99DD99}3
-                                     & \cellcolor[HTML]{99DD99}2
+                                     & \cellcolor[HTML]{99DD99}4.5
                                      & \cellcolor[HTML]{99AA99}3 \\
@@
                                      & \cellcolor[HTML]{99FF99}3
-                                     & \cellcolor[HTML]{99FF99}2
+                                     & \cellcolor[HTML]{99FF99}5
                                      & \cellcolor[HTML]{99FF99}2 \\
@@
                                      & \cellcolor[HTML]{99EE99}\textbf{8}
-                                     & \cellcolor[HTML]{99FF99}\textbf{5}
+                                     & \cellcolor[HTML]{99FF99}\textbf{10.5}
                                      & \cellcolor[HTML]{99AA99}\textbf{8} \\
@@
                                      & \cellcolor[HTML]{99FF99}0
-                                     & \cellcolor[HTML]{99FF99}0
+                                     & \cellcolor[HTML]{99FF99}1
                                      & \cellcolor[HTML]{99FF99}0 \\
@@
                                      & \cellcolor[HTML]{99FF99}\textbf{1}
-                                     & \cellcolor[HTML]{99DD99}\textbf{2}
+                                     & \cellcolor[HTML]{99DD99}\textbf{3}
                                      & \cellcolor[HTML]{99BB99}\textbf{3} \\
@@
                                      & \cellcolor[HTML]{99CC99}\textbf{23}
-                                     & \cellcolor[HTML]{99DD99}\textbf{20}
+                                     & \cellcolor[HTML]{99DD99}\textbf{31.5}
                                      & \cellcolor[HTML]{99AA99}\textbf{23}
@@
 \todo{update colors and find a more relevant way to compute a grand total without favoring criteria that have more metrics than the others (e.g.
-normalize each criterion between 0 and 1).}}
+normalize each criterion between 0 and 1).
+Same comment for each sub-total.
+M1 might be the same as integration.}}
 \label{table:evaluation}

### 2016-04-12 paper.tex
Wrote section 4.7.

@@
 
+\todo{add interaction a with the infrastructure for all architectures (application porting)}
+
 %http://ieeexplore.ieee.org/stamp/stamp.jsp?tp=&arnumber=4782949
@@
 
+\todo{explain that such an abstract architecture has to be concretized to allow for comparison with other architectures.
+Say also that we could envisage other concretizations but we don't include them as we don't have any experience with these systems.}
+
 \begin{figure}
@@
   service), or when a new piece of software has to be installed
-  (infrastructure only).
+  (infrastructure only). Adding a new type of workflow engine allows
+  to execute the workflows that have been described in the language(s)
+  used by the new engine.
 \item New engine version: number of interactions or components to
@@
 
-\subsection{Nested workflows and service invocation}
+\subsection{Nested workflows with service invocation}
 
@@
 
-\subsection{Workflow import}
+\subsection{Workflow import with service invocation}
+
+\paragraph{Integration} Compared to the service architecture, workflow
+import requires an additional interaction \texttt{g} in the science
+gateway (\texttt{I$_1$}=2, \texttt{I$_2$}=3).
+
+\paragraph{Robustness} Since workflow conversion is not involved in
+the execution (it is an offline process), metrics are as in the
+service architecture (\texttt{R$_1$=3}, \texttt{R$_2$=4},
+\texttt{R$_3$=0}).
+
+\paragraph{Modularity} Since adding a new type of workflow engine aims
+at supporting more workflows, we consider that it only requires to
+re-implement interaction \texttt{g} in this architecture. Note,
+however, that implementing interaction \texttt{g} can require very
+substantial work depending on the complexity of the language used by
+the new engine (\texttt{M$_1$=1}). Based on the same logic, adding a
+new engine version only requires modifying interaction \texttt{g}
+(\texttt{M$_2$=1}). As in the service architecture, interfacing with a
+new infrastructure requires modifications in the workflow service and
+in interactions \texttt{b} and \texttt{c$_2$}
+(\texttt{M$_3$=3}). Adding a new workflow is done through interactions
+\texttt{a} and \texttt{g} (\texttt{M$_4$=2}). Meta-workflows are
+available after import, by connecting workflows in the language used
+in the science gateway (\texttt{M$_5$=0}).
+
+\paragraph{Scalability}  Since workflow conversion is not involved in
+the execution (it is an offline process), metrics are as in the
+service architecture (\texttt{S$_1$=1}, \texttt{S$_2$=1},
+\texttt{S$_3$=1}, \texttt{S$_4$=0}).
 
@@
                                      & \cellcolor[HTML]{99FF99}1
-                                     & \cellcolor[HTML]{99AA99}3 \\
+                                     & \cellcolor[HTML]{99AA99}2 \\
 Workflow engine -- \texttt{I$_2$}    & \cellcolor[HTML]{99AA99}2
@@
                                      & \cellcolor[HTML]{99AA99}5
-                                     & \cellcolor[HTML]{99FF99}1 \\
+                                     & \cellcolor[HTML]{99FF99}3 \\
 Infrastructure -- \texttt{I$_3$}   & \cellcolor[HTML]{99FF99}0
@@
                                      & \cellcolor[HTML]{99AA99}\textbf{6}
-                                     & \cellcolor[HTML]{99AA99}\textbf{4} \\
+                                     & \cellcolor[HTML]{99AA99}\textbf{5} \\
 
@@
                                      & \cellcolor[HTML]{99AA99}4 
-                                     & \cellcolor[HTML]{99AA99}4 \\
+                                     & \cellcolor[HTML]{99AA99}3 \\
 Interactions -- \texttt{R$_2$}       & \cellcolor[HTML]{99FF99}3    
@@
                                      & \cellcolor[HTML]{99AA99}\textbf{12}
-                                     & \cellcolor[HTML]{99BB99}\textbf{8} \\
+                                     & \cellcolor[HTML]{99BB99}\textbf{7} \\
 \multicolumn{7}{c}{\cellcolor[HTML]{EEEEEE}\textbf{Modularity}}\\
@@
                                      & \cellcolor[HTML]{99DD99}4.5
-                                     & \cellcolor[HTML]{99AA99}3 \\
+                                     & \cellcolor[HTML]{99AA99}1 \\
 New engine version -- \texttt{M$_2$} & \cellcolor[HTML]{99AA99}1
@@
                                      & \cellcolor[HTML]{99FF99}0
-                                     & \cellcolor[HTML]{99FF99}0 \\
+                                     & \cellcolor[HTML]{99FF99}1 \\
 New infrastructure -- \texttt{M$_3$} & \cellcolor[HTML]{99FF99}3
@@
                                      & \cellcolor[HTML]{99FF99}5
-                                     & \cellcolor[HTML]{99FF99}2 \\
+                                     & \cellcolor[HTML]{99FF99}3 \\
 New workflow -- \texttt{M$_4$}       & \cellcolor[HTML]{99FF99}1
@@
                                      & \cellcolor[HTML]{99FF99}0
-                                     & \cellcolor[HTML]{99AA99}1 \\
+                                     & \cellcolor[HTML]{99AA99}0 \\
 \textbf{Total}                       & \cellcolor[HTML]{99AA99}\textbf{9}
@@
                                      & \cellcolor[HTML]{99FF99}\textbf{10.5}
-                                     & \cellcolor[HTML]{99AA99}\textbf{8} \\
+                                     & \cellcolor[HTML]{99AA99}\textbf{7} \\
 \multicolumn{7}{c}{\cellcolor[HTML]{EEEEEE}\textbf{Scalability}}\\
@@
                                      & \cellcolor[HTML]{99DD99}\textbf{31.5}
-                                     & \cellcolor[HTML]{99AA99}\textbf{23}
+                                     & \cellcolor[HTML]{99AA99}\textbf{22}
 \end{tabular}
@@
 
+The difficulty of implementing interaction g is not properly
+reflected. For instance, fsl uses bash, which would be very difficult to convert.
+
 \section{Conclusion}

### 2016-04-13 paper.tex
Updates on Figures
Updates in paper.tex
Added figure on graphical notations

@@
 \usepackage[x11names,dvipsnames,table]{xcolor} %for use in color links
+\usepackage{subfig}
 \usepackage{svg}
 \usepackage{url}
+
 \usepackage{xspace}
@@
 
-\newcommand{\todo}[1]{\color{blue}\xspace\emph{#1}\xspace\color{black}}
-\newcommand{\note}[1]{\color{blue}\textbf{Note: }\textit{#1} \color{black}}
+\newcommand{\todo}[1]{\color{blue}\xspace[\emph{#1}]\xspace\color{black}}
+\newcommand{\note}[1]{\color{green}\textbf{Note: }\textit{#1} \color{black}}
 
@@
 \end{frontmatter}
-
-%\authors{Tristan Glatard, Marc-Etienne Rousseau, Pierre Rioux, Samir
-%Das, Natacha Beck, Reza Adalat, Pierre Bellec, Pierre-Olivier Quirion,
-%S\'ilvia D. Olabarriaga, Sorina Camarasu-Pop, Alan C. Evans}
-% Rafael? Ewa? Peter Kacsuk? Johan? Satra? Najma?
 
@@
 
-Architectures are defined below from their main components and the
-interactions between these components.
-
-\todo{Introduce the notion of architectural diagram with an abstract
-  figure. Mention arrows, boxes, box with dotted lines, red color,
-  tasks, data.}
+Figure~\ref{fig:notations} shows the graphical notations used to
+describe the architectures in the reminder of this
+Section. Architectures are described from their main software
+components and interactions.
 
@@
 
+\begin{figure}
+\centering
+\def\svgwidth{\columnwidth}
+\input{figures/notations.pdf_tex}
+\caption{Graphical notations}
+\label{fig:notations}
+\end{figure}
+
 %http://ieeexplore.ieee.org/stamp/stamp.jsp?tp=&arnumber=4782949
+
+\begin{figure*}
+\centering
+\hspace*{0.2\columnwidth}
+\subfloat[Tight integration]{
+\def\svgwidth{0.6\columnwidth}
+\input{figures/tight.pdf_tex}
+\label{archi:tight}
+}
+\hfill
+\subfloat[Service invocation]{
+\def\svgwidth{0.6\columnwidth}
+\input{figures/service.pdf_tex}
+\label{archi:service}
+\hspace*{0.2\columnwidth}
+}\\
+\hspace*{0.2\columnwidth}
+\subfloat[Sub-tasking]{
+\def\svgwidth{0.6\columnwidth}
+\input{figures/sub-task.pdf_tex}
+\label{archi:sub-task}
+}
+\hfill
+\subfloat[Pool]{
+\def\svgwidth{0.6\columnwidth}
+\input{figures/agent.pdf_tex}
+\label{archi:agent}
+\hspace*{0.2\columnwidth}
+}\\
+\subfloat[Nested workflows. Left: abstract model. Right: instantiation with service invocation.]{
+\def\svgwidth{0.9\columnwidth}
+\input{figures/nested-1.pdf_tex}
+\def\svgwidth{0.9\columnwidth}
+\input{figures/nested-2.pdf_tex}
+\label{archi:nested}
+}\\
+\subfloat[Workflow import. Left: abstract model. Right: instantiation with service invocation.]{
+\def\svgwidth{0.5\columnwidth}
+\input{figures/import-1.pdf_tex}
+\hfill
+\def\svgwidth{0.5\columnwidth}
+\input{figures/import-2.pdf_tex}
+\label{archi:import}
+}
+\caption{Architectures}
+\label{fig:architectures}
+\end{figure*}
 
@@
 
-\begin{figure}
-\centering
-\def\svgwidth{0.5\columnwidth}
-\input{figures/tight.pdf_tex}
-\caption{Tight integration}
-\label{archi:tight}
-\end{figure}
-
 \subsection{Service invocation}
@@
 that use it~\cite{kacsuk2014science}.
-\begin{figure}
-\centering
-\def\svgwidth{0.5\columnwidth}
-\input{figures/service.pdf_tex}
-\caption{Service integration}
-\label{archi:service}
-\end{figure}
 
@@
 jobs).
-\begin{figure}
-\centering
-\def\svgwidth{0.5\columnwidth}
-\input{figures/sub-task.pdf_tex}
-\caption{Sub-tasking}
-\label{archi:sub-task}
-\end{figure}
 
@@
 SHIWA pool~\cite{ROGE-13}.
-\begin{figure}
-\centering
-\def\svgwidth{0.8\columnwidth}
-\input{figures/agent.pdf_tex}
-\caption{Pool model}
-\label{archi:agent}
-\end{figure}
 
@@
 
-\begin{figure}
-\centering
-\def\svgwidth{0.48\columnwidth}
-\input{figures/nested-1.pdf_tex}
-\def\svgwidth{0.48\columnwidth}
-\input{figures/nested-2.pdf_tex}
-\caption{Nested workflows. Left: abstract model. Right: instantiation with service invocation.}
-\label{archi:nested}
-\end{figure}
-
 \subsection{Workflow import}
@@
 
-\begin{figure}
-\centering
-\def\svgwidth{0.48\columnwidth}
-\input{figures/import-1.pdf_tex}
-\def\svgwidth{0.48\columnwidth}
-\input{figures/import-2.pdf_tex}
-\caption{Workflow import. Left: abstract model. Right: instantiation with service invocation.}
-\label{archi:import}
-\end{figure}
+\todo{workflow import is an offline process.}
 

### 2016-04-13 paper.tex
Updates in section 3.1

@@
 describe the architectures in the reminder of this
-Section. Architectures are described from their main software
-components and interactions.
+Section. Architectures will be described from their main software
+components and interactions. Software components include science
+gateway and infrastructure in all architectures, and workflow service,
+workflow pool and agent are involved in some architectures. The
+workflow engine itself is represented by a specific symbol. Software
+interactions among these components are described in the next
+Section. Abstract interactions are specific types of interactions that
+may be implemented by various different software interactions. Red
+color is used to represent the components and interactions that need
+to be specifically developed to integrate workflow engines. Tasks
+refer to computing tasks that are created by the workflow engine and
+executed on the infrastructure. Data represents any type of file, or
+database that is involved in the workflow execution and stored on the
+infrastructure while the workflow executes. It does not cover workflow
+parameters such as strings, numbers, etc.
 
-\todo{add interaction a with the infrastructure for all architectures (application porting)}
+%http://ieeexplore.ieee.org/stamp/stamp.jsp?tp=&arnumber=4782949
 
@@
 \end{figure}
-
-%http://ieeexplore.ieee.org/stamp/stamp.jsp?tp=&arnumber=4782949
 
@@
 
-The interactions described below are labeled consistently with the
-notations used on Figures~\ref{archi:tight}-\ref{archi:import}.
+The interactions involved in the architectures are described below and
+labeled consistently with the notations used on
+Figure~\ref{fig:architectures}.
 
@@
 \item Workflow integration: consists in adding a new workflow to the
-  science gateway so that users can execute it. It is conducted by
-  science gateway administrators, developers or users, and it results
-  in an interface, for instance a web form, where users can enter the
-  parameters of the workflow to be executed. Integrating a workflow is
-  not the same process as integrating a workflow engine.
-\item Task control: operations to manage tasks on the infrastructure,
-  including: submission, monitoring, termination, deletion, etc.
-\item Data control: operations to manage data on
-  the infrastructure, such as: upload, download, deletion, browsing,
-  replication, caching, etc. Data movements can be triggered by the
-  user, to upload input data or download processed data -- interaction
-  \texttt{$c_1$}, or by the workflow engine, to transfer data across the
-  infrastructure -- interaction \texttt{$c_2$}. \todo{add c1 and c2 to the
-    Figures}
-\item Workflow control: operations to execute a workflow, including:
-  submission, monitoring, termination, etc. Workflow control can be
-  coarse-grained (a.k.a. black box) or fine-grained (white box). In a
-  coarse-grained model, workflow activities are masked. 
+  system so that users can execute it. It is triggered by science
+  administrator and it results in an interface, for instance a web
+  form, where users can enter the parameters of the workflow to be
+  executed. The interaction has two aspects: (\texttt{a$_1$}) the
+  programs used in the workflow are installed on the infrastructure,
+  which may or may not require administrative privileges on the
+  infrastructure; (\texttt{a$_2$}) the workflow has to be configured
+  in the science gateway so that it becomes available to users. Note
+  that integrating a workflow is not the same process as integrating a
+  workflow \emph{engine}.
+\item Task control: operations to manage tasks on the
+  infrastructure, including: authentication, submission, monitoring,
+  termination, deletion, etc. When the infrastructure is a grid or a
+  cloud, it is for instance implemented using libraries such as SAGA,
+  DRMAA, OCCI and similar initiatives \todo{refs needed}.
+\item Data control: operations to manage data on the infrastructure,
+  such as: upload, download, deletion, browsing, replication, caching,
+  etc. Data movements can be triggered by the user in the science
+  gateway (\texttt{$c_1$}), to upload input data or download processed
+  data. They can also be performed by the workflow engine
+  (\texttt{$c_2$}), to transfer data across the
+  infrastructure. 
+\item Workflow control: operations to execute a workflow with an
+  engine, including: workflow submission, monitoring, termination,
+  etc. Workflow control can be coarse-grained (a.k.a. black box) or
+  fine-grained (white box). In a coarse-grained model, workflow
+  activities are masked and the user only has a global view of the
+  workflow execution. In a fine-grained model, user is exposed to the
+  workflow topology, i.e. to the outputs of the individual tasks,
+  their statuses and so on. 
 \item Sub-task control: operations used by tasks to submit sub-tasks
   on the infrastructure, including: submission, monitoring,
-  termination, deletion, etc.
-\item Pool-agent: specific to the architecture described in~\ref{sec:pool}.
+  termination, deletion, etc. Sub-task control is similar to
+  interaction \texttt{b}, except that information about the parent
+  task which submitted a sub-task is usually available and used for
+  additional control. For instance, the parent task may wait for all
+  its sub-tasks to finish before finishing, and conversely all the
+  sub-tasks may be killed if the parent task is killed.
+\item Pool-agent: specific to the architecture described
+  in Section~\ref{sec:pool}. This is an interaction used when agents retrieve
+  work from a central pool. It covers agent registration and
+  de-registration to the pool, protocols to send work from the pool to
+  the agent, mechanisms to update work status, and so on. 
 \item Workflow conversion: translation from one workflow language to
-  another one.
+  another one. This interaction may not be available or implementable
+  for every workflow language. It has been developed mostly for
+  well-structured and relatively simple workflow language such as between GWorkflowDL
+  and Scufl~\cite{OLAB-09} and between the 4 languages that were involved in the
+  SHIWA initiative. In SHIWA, workflow conversion is done through an
+  intermediate representation, the Interoperable Workflow Intermediate
+  Representation~\cite{plankensteiner-montagnat-etal:2011}, which
+  allows to convert among $n$ workflow languages using $2n$
+  interactions instead of $n^2$.
 \end{enumerate}

### 2016-04-13 paper.tex
Updates in Section 3

@@
 
-Figure~\ref{fig:notations} shows the graphical notations used to
-describe the architectures in the reminder of this
-Section. Architectures will be described from their main software
-components and interactions. Software components include science
-gateway and infrastructure in all architectures, and workflow service,
-workflow pool and agent are involved in some architectures. The
-workflow engine itself is represented by a specific symbol. Software
-interactions among these components are described in the next
-Section. Abstract interactions are specific types of interactions that
-may be implemented by various different software interactions. Red
-color is used to represent the components and interactions that need
-to be specifically developed to integrate workflow engines. Tasks
-refer to computing tasks that are created by the workflow engine and
-executed on the infrastructure. Data represents any type of file, or
-database that is involved in the workflow execution and stored on the
-infrastructure while the workflow executes. It does not cover workflow
-parameters such as strings, numbers, etc.
+The architectures are diagramed in Figure~\ref{fig:architectures}
+using the graphical notations shown in
+Figure~\ref{fig:notations}. Architectures are described from their
+main software components and interactions. Software components include
+science gateway and infrastructure in all architectures, and workflow
+service, workflow pool and agent are involved in some
+architectures. The workflow engine itself is represented by a specific
+symbol. Software interactions among these components are described in
+the next Section. Abstract interactions are specific types of
+interactions that may be implemented by various different software
+interactions. Red color is used to represent the components and
+interactions that need to be specifically developed to integrate
+workflow engines. Tasks refer to computing tasks that are created by
+the workflow engine and executed on the infrastructure. Data
+represents any type of file, or database that is involved in the
+workflow execution and stored on the infrastructure while the workflow
+executes. It does not cover workflow parameters such as strings,
+numbers, etc. The reminder of this section describes the different
+types of interactions involved in the architectures, and details each
+architecture.
 
@@
 \item Workflow integration: consists in adding a new workflow to the
-  system so that users can execute it. It is triggered by science
-  administrator and it results in an interface, for instance a web
-  form, where users can enter the parameters of the workflow to be
-  executed. The interaction has two aspects: (\texttt{a$_1$}) the
-  programs used in the workflow are installed on the infrastructure,
-  which may or may not require administrative privileges on the
-  infrastructure; (\texttt{a$_2$}) the workflow has to be configured
-  in the science gateway so that it becomes available to users. Note
-  that integrating a workflow is not the same process as integrating a
-  workflow \emph{engine}.
-\item Task control: operations to manage tasks on the
-  infrastructure, including: authentication, submission, monitoring,
-  termination, deletion, etc. When the infrastructure is a grid or a
-  cloud, it is for instance implemented using libraries such as SAGA,
-  DRMAA, OCCI and similar initiatives \todo{refs needed}.
+  system so that users can execute it. It is triggered by an
+  administrator of the science gateway and it results in an interface,
+  for instance a web form, where users can enter the parameters of the
+  workflow to be executed. The interaction has two aspects:
+  (\texttt{a$_1$}) the programs used in the workflow are installed on
+  the infrastructure, which may or may not require administrative
+  privileges on the infrastructure; (\texttt{a$_2$}) the workflow has
+  to be configured in the science gateway so that it becomes available
+  to users. Note that integrating a workflow is not the same process
+  as integrating a workflow \emph{engine}.
+\item Task control: operations to manage tasks on the infrastructure,
+  including: authentication, submission, monitoring, termination,
+  deletion, etc. Controlling tasks requires to deal with the
+  heterogeneous batch managers and meta-schedulers that might be
+  available on the infrastructure. When the infrastructure is a grid
+  or a cloud, it is for instance implemented using libraries such as
+  SAGA, DRMAA, OCCI and similar initiatives \todo{refs needed}.
 \item Data control: operations to manage data on the infrastructure,
@@
   data. They can also be performed by the workflow engine
-  (\texttt{$c_2$}), to transfer data across the
-  infrastructure. 
+  (\texttt{$c_2$}), to transfer data across the infrastructure. The
+  infrastructure might offer various data storage backends with
+  heterogeneous interfaces. Tools and services such as
+  JSAGA~\cite{reynaud2010uniform} or Data Avenue~\cite{hajnal2014data}
+  can be used to homogeneize these interfaces.
 \item Workflow control: operations to execute a workflow with an
@@
 
+\todo{In the following sections, for each architecture, add a
+  description of a real system that implements it: integrated: ?,
+  service: VIP, sub-task: CBRAIN-PSOM, pool: SHIWA pool, nested
+  workflow: SHIWA CGI, import: SHIWA FGI}
+
 \subsection{Tight integration}
@@
 software components with the science gateway. For instance, the
-workflow engine might be a portlet in a Liferay portal, a controller
-in a Ruby on Rails application, and so on. The workflow engine and the
-science gateway usually share a database where application, users and
-other resources are stored. In this model, task control and data
-control are initiated from the science gateway. This is the model
-adopted in the Catania Science Gateway Framework~\cite{Ardizzone2012}
-(see specific documentation on
+workflow engine might be a portlet in a Liferay portal or a controller
+in a Ruby on Rails application. The workflow engine and the science
+gateway usually share a database where application, users and other
+resources are stored. In this model, task and data control are both
+initiated from the science gateway. Interactions \texttt{b} and
+\texttt{c$_2$} are initiated from the workflow engine while
+\texttt{c$_1$} comes from other parts of the science gateway, for
+instance a data management interface. As in any other model, the
+installation of new workflows in the science gateway (\texttt{a$_2$})
+and infrastructure (\texttt{a$_1$}) is done by an administrator, for
+security reasons. This is the model adopted in the Catania Science
+Gateway Framework~\cite{Ardizzone2012} (see specific documentation on
 workflows\footnote{\url{http://bit.ly/1oQrzvQ}}),
 CIPRES~\cite{miller2010creating} and LONI Pipeline
-Environment~\cite{dinov2009efficient}.\todo{Add better justification}
+Environment~\cite{dinov2009efficient}.
 
@@
 externally to the science gateway, in a service. The science gateway
-controls the service through a specific interaction that might be
-implemented as a web-service call (e.g. RESTful or SOAP), as a
-command-line or as any other method that offers a well-defined
+controls the service through a specific interaction (\texttt{d}) that
+might be implemented as a web-service call (e.g. RESTful or SOAP), as
+a command-line or as any other method that offers a well-defined
 interface to the workflow engine. The workflow engine might be invoked
 either as a black box that completely masks the infrastructure and
-workflow activities, or as a grey box that allows for some interaction
-with them.  The workflow engine is responsible for controlling the
-tasks on the infrastructure, and performing the required data
-transfers to execute them. User data is usually managed through the
-science gateway, although it might as well be delivered by the
-workflow engine directly to the user.
+workflow activities, or as a white box that allows for some
+interaction with them. The workflow engine is responsible for
+controlling the tasks on the infrastructure (\texttt{b}) and for
+performing the required data transfers to execute them
+(\texttt{c$_2$}). User data is usually managed through the science
+gateway (\texttt{c$_1$}), although it might as well be delivered by
+the workflow engine directly to the user.
 
@@
 framework~\cite{Kacsuk2012} and the numerous science gateway instances
-that use it~\cite{kacsuk2014science}.
+that rely on
+WS-PGRADE/gUSE~\cite{kacsuk2014science}. Figure~\ref{fig:vip-architecture}
+shows the architecture of the Virtual Imaging Platform, where the
+MOTEUR workflow engine~\cite{GLAT-08i} is invoked as a service at step
+\texttt{3} (interaction \texttt{d} on
+Figure~\ref{fig:architectures}). Step \texttt{2} on
+Figure~\ref{fig:vip-architecture} corresponds to interaction
+\texttt{a$_1$}, and step \texttt{4} maps to interactions \texttt{b}
+and \texttt{c$_2$} (in VIP, workflow data transfers are performed by
+the tasks and embedded in their descriptions).
+
+\begin{figure*}
+\def\svgwidth{2\columnwidth}
+\input{figures/VIP.pdf_tex}
+\caption{Architecture of the Virtual Imaging Platform that integrates
+  the MOTEUR workflow engine through service invocation.}
+\label{fig:vip-architecture}
+\end{figure*}
 
@@
 
-See Figure~\ref{archi:sub-task}. The science gateway is responsible to
-execute the tasks on the infrastructure, dealing with heterogeneous
-batch managers and meta-schedulers. The workflow engine is a
-particular task that can submit sub-tasks to the science gateway. The
-workflow engine keeps track of the dependencies between the sub-tasks
-but their execution is delegated to the science gateway. The science
-gateway may implement mechanisms to deal with task dependencies such
-as basic dependency lists as available in most batch
-managers. Although it has no global vision of the workflow, it can
-keep track of the sub-tasks submitted by a given task, for instance to
-be able to cancel them when the task is canceled. The fact that tasks
-are submitted through the science gateway does not restrict the range
-of possible execution models: pilot jobs, for instance, can still be
-used.
+See Figure~\ref{archi:sub-task}. The workflow engine is a particular
+task that can submit sub-tasks to the science gateway through
+interaction \texttt{e}. The workflow engine keeps track of the
+dependencies between the sub-tasks but their execution is delegated to
+the science gateway that executes them on the infrastructure through
+interaction \texttt{b}. Although the science gateway has no global
+vision of the workflow, it can keep track of the sub-tasks submitted
+by a given task, for instance to be able to cancel them when the task
+is canceled. The science gateway may also implement mechanisms to
+facilitate the handling of task dependencies, for instance basic
+dependency lists as available in most batch managers (see for instance
+attribute \texttt{depend} of option \texttt{-W} in
+Torque\footnote{\url{http://docs.adaptivecomputing.com}}).
 
-\todo{Interactions c1 and c2 are merged in c since this is done
-  completely by the science gateway}
+The science gateway also transfers both user and workflow data across
+the infrastructure, so that interactions \texttt{c$_1$} and
+\texttt{c$_2$} are both covered by \texttt{c}. In practice, both
+\texttt{c$_1$} and \texttt{c$_2$} can be implemented using the same
+pieces of code.
 
-This model is implemented in CBRAIN~\cite{SHER-14} where it is used to
-integrate the PSOM workflow engine and the FSL toolkit through its
-\texttt{fsl\_sub} tool. In particular, the CBRAIN-PSOM integration is
-described in~\cite{GLAT-16} and uses an agent computing model (a.k.a pilot
-jobs).
+The sub-tasking architecture is implemented in CBRAIN~\cite{SHER-14}
+where it is used to integrate the PSOM workflow
+engine~\cite{bellec2012pipeline} and the FSL toolkit. The CBRAIN-PSOM
+integration~\cite{GLAT-16} is illustrated on
+Figure~\ref{fig:cbrain-psom-architecture}. On this Figure, the science
+gateway is represented by components \texttt{CBRAIN portal} and
+\texttt{CBRAIN execution server} and the infrastructure consists of
+\texttt{Storage servers} and \texttt{Computing cluster with shared
+  file system}. Interaction \texttt{b} on
+Figure~\ref{fig:architectures} is implemented by steps \texttt{4},
+\texttt{5}, \texttt{7} and \texttt{8}. Interaction \texttt{c} is
+implemented by steps \texttt{3} and \texttt{10}. Interaction
+\texttt{e} is implemented by step \texttt{6}. The PSOM workflow engine
+adopts a pilot-job architecture~\cite{turilli2015comprehensive} where
+the a master coordinates workflow execution by submitting workers
+and establishing direct communication channels with them. Note how
+this peculiar execution model is totally supported by the sub-tasking
+architecture.
+
+The CBRAIN-FSL integration is also remarkable as it allows to leverage
+FSL pipelines that are written in a very low-level workflow language
+(Linux executables and script) that submits tasks uniformly through a
+specific tool called \texttt{fsl\_sub}. To integrate FSL in CBRAIN,
+\texttt{fsl\_sub} was modified to implement interaction
+\texttt{e}\footnote{See \texttt{fsl\_sub} modified script available at
+  \url{https://github.com/aces/cbrain-plugins-neuro}} \todo{Add a
+  diagram and maybe a code excerpt to illustrate that.}.
+
+\begin{figure}
+\def\svgwidth{\columnwidth}
+\input{figures/CBRAIN-PSOM.pdf_tex}
+\caption{Architecture of the CBRAIN-PSOM integration that illustrates
+  the sub-tasking model (Figure extracted from~\cite{GLAT-16}). 1:
+  User sends data and pipeline execution request to storage server(s)
+  and CBRAIN portal. 2: CBRAIN portal sends execution request to
+  execution server on cluster. 3: Execution server transfers data from
+  storage server(s). 4-5: Execution server starts PSOM master via task
+  scheduler. 6: PSOM master submits PSOM workers to execution
+  server. 7-8: Execution server starts PSOM workers through task
+  scheduler. 9: PSOM master and workers execute pipeline. 10.
+  Execution server transfers results to storage server(s). \todo{Improve graphics}}
+\label{fig:cbrain-psom-architecture}
+\end{figure}
 
@@
 gateway to a pool to which agents connect asynchronously to retrieve
-and execute workflows (interaction \texttt{f}). Agents may be started according to various
-policies, for instance to ensure load balancing. Agents may wrap
-different types of workflow engines. This model was implemented in the
-SHIWA pool~\cite{ROGE-13}.
+and execute workflows (interaction \texttt{f}). Agents may be started
+according to various policies, for instance to ensure load
+balancing. Agents may wrap different types of workflow engines. This
+model was implemented in the SHIWA pool~\cite{ROGE-13}.
 

### 2016-04-13 figures/export.sh

@@ -1,6 +1,6 @@
 #!/bin/bash
 
-for name in tight service sub-task agent nested-1 nested-2 import-1 import-2
+for name in tight service sub-task agent nested-1 nested-2 import-1 import-2 notations VIP CBRAIN-PSOM
 do
   echo "Exporting ${name}.svg"
   inkscape -D -z --file=figures/${name}.svg --export-pdf=figures/${name}.pdf --export-latex

### 2016-04-13 paper.tex
Minor updates

@@
 \usepackage{url}
-
+\usepackage{flushend}
 \usepackage{xspace}
@@
 
-FLI-IAM, Labex PRIMES, Ludmer Centre
+\todo{FLI-IAM, Labex PRIMES, Ludmer Centre, whatever grant is funding
+  POQ for integrating CBRAIN with PSOM.}
+
+We warmly thank Rafael Ferreira da Silva for implementing the VIP
+platform and creating Figure~\ref{fig:vip-architecture}.
 
@@
 
+\todo{Review use of capitals in references}
 \bibliographystyle{elsarticle-num} 

### 2016-04-15 paper.tex
Added figure on nested workflows
Updates in section 3.6

@@
 \begin{figure*}
-\def\svgwidth{2\columnwidth}
+\centering
+\def\svgwidth{1.5\columnwidth}
 \input{figures/VIP.pdf_tex}
@@
 Nested workflows have long been available in workflow engines, for
-instance in the Taverna workbench~\cite{oinn2004taverna}. This model
-was used by the SHIWA Simulation Platform to implement Coarse-Grained
-workflow interoperability~\cite{terstyanszky2014enabling}, i.e. to
-integrate various workflow engines in a consistent platform. Nested
-workflows are also often used implicitly to wrap scripts as
-black-boxes while these scripts actually correspond to complete workflows.
+instance in the Taverna workbench~\cite{oinn2004taverna}. They were
+also used by the SHIWA Science Gateway to implement so-called
+Coarse-Grained workflow
+interoperability~\cite{terstyanszky2014enabling}, i.e. to integrate
+various workflow engines in a consistent
+platform. Figure~\ref{fig:shiwa-architecture} shows the architecture
+used in the SHIWA Science Gateway for nested workflow execution. It
+corresponds to the nested worfklow with service invocation in
+Figure~\ref{archi:nested}. The parent workflow engine is WS-PGRADE,
+invoked as a service in the Science Gateway (step \texttt{1} on
+Figure~\ref{fig:shiwa-architecture}, interaction \texttt{d} on
+Figure~\ref{archi:nested}). Ten different child engines can be used by
+nested workflows, invoked through a ``submission service'' (step
+\texttt{2}, interaction \texttt{d}). Each of these engines can submit
+jobs and transfer data to a distributed computing infrastructure (DCI,
+step \texttt{3}, interactions \texttt{b} and \texttt{c$_2$}).
+
+Nested workflows
+are also often used implicitly to wrap scripts as black-boxes while
+these scripts actually correspond to complete workflows.
 
@@
 Say also that we could envisage other concretizations but we don't include them as we don't have any experience with these systems.}
+
+\begin{figure*}
+\centering
+\includegraphics[width=1.5\columnwidth]{figures/shiwa-science-gateway.pdf}
+\caption{Nested workflow execution through SHIWA Science Gateway
+  (Figure reproduced from~\cite{terstyanszky2014enabling} with
+  permission of the first author).}
+\label{fig:shiwa-architecture}
+\end{figure*}
 
@@
 format conversion from a native format to the science gateway
-format. This was implemented in the SHIWA Simulation Platform through
+format. This was implemented in the SHIWA Science Gateway through
 the IWIR language that provided a common language for portability
@@
 that workflow execution is split in different engines (\texttt{S$_4$=1}).
-
 \subsection{Workflow import with service invocation}

### 2016-04-15 paper.tex
Updates in section 3.6

@@
 Figure~\ref{fig:notations}. Architectures are described from their
-main software components and interactions. Software components include
-science gateway and infrastructure in all architectures, and workflow
-service, workflow pool and agent are involved in some
-architectures. The workflow engine itself is represented by a specific
-symbol. Software interactions among these components are described in
-the next Section. Abstract interactions are specific types of
-interactions that may be implemented by various different software
-interactions. Red color is used to represent the components and
+main software \emph{components} and \emph{interactions}. Software
+components include science gateway and infrastructure in all
+architectures, and workflow service, workflow pool and agent are
+involved in some architectures. The workflow engine itself is
+represented by a specific symbol. Software interactions among these
+components are described in the next Section. \emph{Abstract}
+interactions are specific types of interactions that may be
+implemented by various different software interactions. An
+architecture that has an abstract interaction is an abstract
+architecture. Red color is used to represent the components and
 interactions that need to be specifically developed to integrate
@@
 
-See Figure~\ref{archi:nested}. A nested workflow is an abstract
-architectural pattern where an activity of a master workflow is itself
-a workflow that is executed by another engine. Nested workflows can be
+See Figure~\ref{archi:nested}. In nested workflows (Figure~\ref{archi:nested}-Left), an activity of a
+\emph{parent} workflow executed by the \emph{parent} engine is itself
+a \emph{child} workflow that is executed by a \emph{child} engine. The
+parent and child engines might use different workflow description
+languages. They might also execute workflows on different
+infrastructures. The parent workflow is also called meta-workflow. The
+science gateway communicates with the parent engine through
+interaction \texttt{*$_1$}. The science gateway also communicate with
+the infrastructure to transfer user data through abstract interactions
+\texttt{*$_2$} and \texttt{*$_3$}. Both workflow engines communicate
+with the infrastructure through abstract interactions as well,
+\texttt{*$_4$} and \texttt{*$_5$}. The parent engine communicates with
+the child engine through abstract interaction
+\texttt{*$_6$}. Administrator installs workflows in the science
+gateway through interaction \texttt{a$_2$}, and installs software on
+infrastructures through interactions \texttt{a$_1$}.
+
+Nested workflows are abstract architectural patterns that can be
 instantiated in the various architectures described previously. We
-focus on instantiation with the service invocation model as this is
-the most used architecture. The master workflow is also called
-meta-workflow. The nested workflow might be described with a different
-language than the one used to describe the master workflow.
+focus on instantiation with the service invocation model(see
+Figure~\ref{archi:nested}-Right) as this is the most used
+architecture. In such an instantiation, we assume that the parent and
+child workflow engines are distinct pieces of software that require
+different workflow services invoked by distinct \texttt{d}
+interactions. If this is not the case, then workflow services can be
+collapsed in a single one with a \texttt{d} interaction with
+itself. Workflow engines communicate with infrastructures using
+\texttt{b} and \texttt{c$_2$}. Science gateway transfer user data to
+infrastructures using \texttt{c$_1$} interactions.
 
 Nested workflows have long been available in workflow engines, for
-instance in the Taverna workbench~\cite{oinn2004taverna}. They were
-also used by the SHIWA Science Gateway to implement so-called
-Coarse-Grained workflow
+instance in the Taverna workbench~\cite{oinn2004taverna}. They are
+also used implicitly in several platforms where workflow engines are
+wrapped in workflow activities as any other command-line tool. Nested
+workflows were also notably used by the SHIWA Science Gateway to
+implement so-called Coarse-Grained workflow
 interoperability~\cite{terstyanszky2014enabling}, i.e. to integrate
@@
 platform. Figure~\ref{fig:shiwa-architecture} shows the architecture
-used in the SHIWA Science Gateway for nested workflow execution. It
-corresponds to the nested worfklow with service invocation in
-Figure~\ref{archi:nested}. The parent workflow engine is WS-PGRADE,
-invoked as a service in the Science Gateway (step \texttt{1} on
+used in the SHIWA Science Gateway for nested workflow execution with
+service invocation as represented in Figure~\ref{archi:nested}. The
+parent workflow engine is WS-PGRADE, invoked as a service in the
+Science Gateway (step \texttt{1} on
 Figure~\ref{fig:shiwa-architecture}, interaction \texttt{d} on
 Figure~\ref{archi:nested}). Ten different child engines can be used by
-nested workflows, invoked through a ``submission service'' (step
-\texttt{2}, interaction \texttt{d}). Each of these engines can submit
-jobs and transfer data to a distributed computing infrastructure (DCI,
-step \texttt{3}, interactions \texttt{b} and \texttt{c$_2$}).
-
-Nested workflows
-are also often used implicitly to wrap scripts as black-boxes while
-these scripts actually correspond to complete workflows.
-
-\todo{instantiation is for foreign workflows, i.e. 2 services. We could have another one with native workflows too.}
-
-\todo{both workflow engines have to access the infrastructure. If they
-  run on 2 infrastructures, c1 has to be duplicated.}
-
-\todo{define parent and child engine}
-
-\todo{explain that such an abstract architecture has to be concretized to allow for comparison with other architectures.
-Say also that we could envisage other concretizations but we don't include them as we don't have any experience with these systems.}
+nested workflows, invoked through the \texttt{Submission service}
+(step \texttt{2}, interaction \texttt{d}). Each of these engines can
+submit jobs and transfer data to a distributed computing
+infrastructure (DCI, step \texttt{3}, interactions \texttt{b} and
+\texttt{c$_2$}). Data interactions (\texttt{c}) and application
+porting ones (\texttt{a}) are not represented on
+Figure~\ref{fig:shiwa-architecture}.
 
@@
 Same comment for each sub-total.
-M1 might be the same as integration.}}
+M1 might be the same as integration.
+Update nested workflow as infrastructures were separated.}}
 \label{table:evaluation}

### 2016-04-15 paper.tex
Updates in section 3.5

@@
 See Figure~\ref{archi:agent}. Workflows are submitted by the science
-gateway to a pool to which agents connect asynchronously to retrieve
-and execute workflows (interaction \texttt{f}). Agents may be started
-according to various policies, for instance to ensure load
-balancing. Agents may wrap different types of workflow engines. This
-model was implemented in the SHIWA pool~\cite{ROGE-13}.
+gateway to a pool through interaction \texttt{d}. Agents connect to
+the pool asynchronously to retrieve and execute workflows through
+interaction \texttt{f}. Agents may be started according to various
+policies, for instance to ensure load balancing. Agents may wrap
+different types of workflow engines. Workflow engine controls tasks
+and data on the infrastructure through interactions \texttt{b} and
+\texttt{c$_2$}, science gateway transfers user data through
+interaction \texttt{c$_1$}, and administrator installs workflow
+dependencies through interaction \texttt{a$_1$}.
+
+The pool model was implemented in the SHIWA pool~\cite{ROGE-13}, see
+Figure~\ref{fig:shiwa-pool-architecture}. In Figure
+~\ref{fig:shiwa-pool-architecture}, interaction \texttt{d} of
+Figure~\ref{archi:agent} is implemented in 3 different types of calls
+to the pool: workflow submission (\texttt{1} \& \texttt{2}), workflow
+status retrieval (\texttt{a} and \texttt{b}), and workflow retrieval
+(\texttt{13} \& \texttt{14}). In Figure
+~\ref{fig:shiwa-pool-architecture}, interaction \texttt{f} of
+Figure~\ref{archi:agent} is implemented through 2 types of calls:
+workflow instance retrieval (\texttt{4}, \texttt{5} and \texttt{6}),
+and workflow instance update (\texttt{11} and \texttt{12}). Workflow
+instance retrieval is used by the agents to fetch work from the
+pool. Workflow instance update is used by the agents to update
+workflow statuses. In the SHIWA pool, agents can wrap different types
+of workflow engines to execute workflows expressed with different
+languages. Calls \texttt{0}, \texttt{7}, \texttt{8} and \texttt{9} are
+used by workflow engine plugins to declare their supported language
+and launch engines, and by workflow engines to report their status to
+the agent. These calls are specific to the SHIWA Pool implementation
+of the \texttt{agent} component and therefore have no corresponding
+representation in Figure~\ref{archi:agent}.
+
+\begin{figure*}
+\centering
+\includegraphics[width=1.5\columnwidth]{figures/pool-interactions.pdf}
+\caption{Architecture of the SHIWA pool (Figure reproduced
+  from~\cite{ROGE-13}). Circle-terminated arrows indicate messages
+  that are broadcast to all pool clients.}
+\label{fig:shiwa-pool-architecture}
+\end{figure*}
 

### 2016-04-18 paper.tex
Major update in section 4 (metrics have changed).

@@
 \section{Architectures}
+\label{sec:architectures}
 
@@
 architecture. Red color is used to represent the components and
-interactions that need to be specifically developed to integrate
-workflow engines. Tasks refer to computing tasks that are created by
-the workflow engine and executed on the infrastructure. Data
-represents any type of file, or database that is involved in the
-workflow execution and stored on the infrastructure while the workflow
-executes. It does not cover workflow parameters such as strings,
-numbers, etc. The reminder of this section describes the different
-types of interactions involved in the architectures, and details each
-architecture.
+interactions that are specific to workflow execution, i.e. that are
+not involved in any other process. Tasks refer to computing tasks that
+are created by the workflow engine and executed on the
+infrastructure. Data represents any type of file, or database that is
+involved in the workflow execution and stored on the infrastructure
+while the workflow executes. It does not cover workflow parameters
+such as strings, numbers, etc. The remainder of this section describes
+the different types of interactions involved in the architectures, and
+details each architecture.
 
@@
 
-\todo{In the following sections, for each architecture, add a
-  description of a real system that implements it: integrated: ?,
-  service: VIP, sub-task: CBRAIN-PSOM, pool: SHIWA pool, nested
-  workflow: SHIWA CGI, import: SHIWA FGI}
-
 \subsection{Tight integration}
@@
 Environment~\cite{dinov2009efficient}.
+\todo{Add description of a real system.}
 
@@
 \todo{workflow import is an offline process.}
+\todo{Add description of a real system}
 
@@
 
-See Table~\ref{table:evaluation}. We focus on the integration of
-workflow engines in science gateways, i.e. we assume that the science
-gateway is already interfaced with the infrastructure.
+The architectures described in Section~\ref{sec:architectures} are
+evaluated on Table~\ref{table:evaluation}. We use 5 main criteria to
+evaluate the architectures: integration effort, robustness of workflow
+execution, extensibility, scalability and other features. Criteria
+break down to specific metrics where \emph{low value indicates good
+  performance}. For each criterion, a total score is computed by
+summing up the individual metrics. We ensure that the different
+metrics in a criterion measure similar entities so that summation
+makes sense. Metric values obtained by each architecture are
+normalized between 0 (best value) and 1 (worst value) to determine the
+color of the corresponding table cell (brighter color indicate better
+performance\footnote{Normalized metric value $m'$ is
+  $\frac{m-m_{\mathrm{min}}}{m_{\mathrm{min}}-m_{\mathrm{max}}}$ where
+  $m$ is the initial metric value,
+  $m_{\mathrm{min}}$/$m_{\mathrm{max}}$ are the minimal/maximal values
+  among all architectures. Cell color is \todo{\#99XX99, where
+    X=F-4m'})}. The criteria and metrics are explained hereafter.
 
 \subsection{Evaluation metrics}
-
-We use four main criteria to evaluate the architectures: integration
-effort, robustness, modularity and scalability. Criteria break down to
-specific metrics for which low values indicate good performance.
 
@@
 
-\todo{Refer to the metric codes.}
-
 \emph{Integration} measures the development effort required to build
-the architecture, assuming that the science gateway and infrastructure
-are already integrated. It is measured by counting on the
-architectural diagram the number of interactions and components to
-develop or modify. It breaks down to 4 metrics that quantify the
-developments related to:
+the architecture. It is obtained by counting the number of
+interactions and components on the architectural diagram. It breaks
+down to the following 2 metrics:
 \begin{itemize}[itemsep=0cm]
-\item Science gateway (\texttt{I$_1$}),
-\item Workflow engine (\texttt{I$_2$}),
-\item Infrastructure (\texttt{I$_3$}),
-\item Other components (\texttt{I$_4$}).
+\item Components \texttt{I$_1$}),
+\item Interactions (\texttt{I$_2$}).
 \end{itemize}
-Interactions are counted only once, in the component from which they
-originate.
-
-\emph{Robustness} relates to the complexity of the architecture and
-its ability to provide relevant debugging information about the
-workflow executions. It breaks down to 3 metrics:
-\begin{itemize}[itemsep=0cm]
-\item Components: number of unique components involved in workflow executions.
-\item Interactions: number of unique interactions involved in workflow executions. 
-\item Debugging: availability of debugging information about workflow
-  activities (1: difficult to obtain; 0: easy to access). Fine-grained
+One may wonder whether infrastructure should be part of
+\texttt{I$_1$}) since it is usually not developed by the groups who
+integrate workflow engines in science gateway. However, integrating an
+infrastructure in the architecture usually requires some technical
+effort (e.g. account creation, software installation, etc) which is
+why we keep it in the metric. The two metrics are summed to obtain
+the total score for this criterion, which measures the total number of
+software pieces to develop.
+
+\emph{Robustness of workflow execution} measures the likelihood that
+workflow execution fails due to errors coming from the software
+architecture. It is determined as the number of software components and
+interactions that are specific to workflow execution. It assume that
+complex architectures tend to be prone to failure.  These components
+and interactions are represented in red on the architecture diagram in
+Figure~\ref{fig:architectures}. Robustness breaks down to 2 metrics:
+\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
+\item Components (\texttt{R$_1$}): number of specific components
+  involved in workflow executions.
+\item Interactions (\texttt{R$_2$}): number of specific interactions
+  involved in workflow executions.
+\end{itemize}
+The two metrics are summed to obtain the total score for this
+criterion, which measures the total number of software pieces that are
+specifically involved in workflow execution. Robustness of workflow
+execution is critical since it directly impacts the end users of the
+science gateway.
+
+\emph{Extensibility} measures the difficulty to replace or add
+elements in the architecture. It is obtained as the number of
+interaction and component to modify when a new element is
+added. Modification of an interaction is deemed necessary when the
+parameters involved in this interaction are modified. Modification of
+a component is required when its code needs to be modified or
+recompiled (science gateway or workflow service), or when a new piece
+of software has to be installed (infrastructure only). Extensibility
+breaks down in 4 metrics depending on the type of element that has to
+be added or replaced:
+\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
+\item New engine type (\texttt{E$_1$}): number of interactions or
+  components to modify to integrate a new workflow engine type in the
+  architecture. Workflow engine belong to different types when they
+  cannot be invoked using the same interface. Adding a new type of
+  workflow engine allows to execute more workflows in the science gateway.
+\item New engine version (\texttt{E$_2$}): number of interactions or
+  components to modify to integrate a new version of a workflow engine
+  in the architecture, assuming that another version of the same
+  engine type is already available. Different versions of a workflow
+  engine share the same interface, i.e. they can be invoked using the
+  same software. When this is not the case, the different versions
+  have to be considered as different engine types. 
+\item New workflow (\texttt{E$_3$}): number of interactions required
+  to integrate a new workflow in the architecture, assuming that the
+  engine type and version required to execute this workflow are
+  already available.
+\item New infrastructure (\texttt{E$_4$}): number of interactions or components to
+  modify to integrate a new type of infrastructure in the
+  architecture. Adding a new infrastructure allows to provide more
+  computing or storage power, to access specific types of resources
+  (e.g. GPUs, clouds), or to enforce execution policies
+  (e.g. constrain data to remain in a particular network domain).
+\item New science gateway: \todo{Not sure if this makes sense}.
+\end{itemize}
+The 4 metrics are summed to obtain a global index that measures the
+difficulty to extend the architecture.
+
+\emph{Scalability} corresponds to the ability of the architecture to
+cope with high workloads. It is measured by reviewing the features
+that enable particular aspects of scalability. Features are evaluated
+using a 3-level metric: 0 means that the feature is very easy to
+enable, 1 means that it can be implemented but with some difficulty,
+and 2 means that it is not possible to implement the feature without a
+nonsensical amount of effort. Four different features are identified:
+\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
+\item Adding engine instance (\texttt{S$_1$}): this feature
+  corresponds to the possibility to add more than 1 engine instance in
+  the architecture. Workflow engines may require important amounts of
+  resources when a lot of workflows, or large workflows, are
+  executed. At some point, it may be required for the science gateway
+  to distribute the load to several engines. \texttt{S$_1$}=0 when
+  adding a new engine instance is a fully automated process, i.e. the
+  workflow engine only has to be started. In this case, elastic
+  engines are possible, i.e. some kind of auto-scaling mechanism can
+  be implemented to control the number of engine instances in the
+  architecture. \texttt{S$_1$}=1 when adding a new engine instance
+  requires some form of manual intervention, which prevents easy
+  implementations of elastic engines. \texttt{S$_1$}=2 when new engine
+  instances cannot be added.
+\item Distributed engines (\texttt{S$_2$}): possibility to distribute
+  the execution \emph{of a single workflow} among different engine
+  instances. In our scope, this feature focuses on the capabilities of
+  the architecture rather than these of the workflow
+  engine. \texttt{S$_2$}=0 when distributed engines are possible in
+  the architecture and \texttt{S$_2$}=1 when they require specific
+  developments in the workflow engine.
+\item Task scheduling (\texttt{S$_3$}): task scheduling is a difficult
+  issue that depends more on the implementation of specific algorithms
+  in the science gateway, workflow engine and infrastructure than on
+  the architecture used to integrate the workflow engine in the
+  science gateway. Some architectures, however, complexify the task
+  scheduling problem by introducing additional software layers or
+  create tasks with specific characteristics. \texttt{S$_3$}=0 when
+  the architecture does not introduce any additional complexity to the
+  scheduling problem and \texttt{S$_3$}=1 otherwise.
+\end{itemize}
+The 3 metrics are summed to obtain a global measure of the scalability
+issues available in the architecture.
+
+\emph{Other features} include:
+\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
+\item Meta-workflow (\texttt{O$_1$}): ability to describe meta-workflows from existing
+  workflows.  \todo{More details about why this is important?} \texttt{O$_1$}=0 when the feature is not available, \texttt{O$_1$}=1 otherwise.
+\item Fine-grained debugging (\texttt{O$_2$}): availability of
+  debugging information about workflow activities. Fine-grained
   information about workflow activity is required to properly
-  troubleshoot workflow executions.
+  troubleshoot workflow executions.  \texttt{O$_2$}=0 when such
+  information is easy to access, \texttt{O$_2$}=1
+  otherwise. \todo{More details about why this is important?}
 \end{itemize}
-
-\emph{Modularity} measures the ability to replace elements in the
-architecture, or to integrate new elements. Elements include
-workflow engines, workflows and infrastructure. It breaks down in 4
-metrics:
-\begin{itemize}[itemsep=0cm]
-\item New engine type: number of interactions or components to modify
-  to integrate a new workflow engine type in the
-  architecture. Modification of an interaction is deemed necessary
-  when the parameters involved in this interaction are
-  modified. Modification of a component is required when its code
-  needs to be modified or recompiled (science gateway or workflow
-  service), or when a new piece of software has to be installed
-  (infrastructure only). Adding a new type of workflow engine allows
-  to execute the workflows that have been described in the language(s)
-  used by the new engine.
-\item New engine version: number of interactions or components to
-  modify to integrate a new version of a workflow engine in the
-  architecture, assuming that another version of the same engine type
-  is already available. We assume that different versions of a
-  workflow engine share the same interface, i.e. they can be invoked
-  using the same software. When this is not the case, the different
-  versions have to be considered as different engine types. 
-\item New infrastructure: number of interactions or components to
-  modify to integrate a new type of infrastructure in the
-  architecture.
-\item New workflow: number of interactions required to integrate a new
-  workflow in the architecture.
-\item Meta-workflow: ability to describe meta-workflows from existing
-  workflows (1: not available; 0: available).
-\end{itemize}
-
-\emph{Scalability} measures the ability of the architecture to support
-high workloads. It breaks down into 4 metrics:
-\begin{itemize}[itemsep=0cm]
-\item New engine instance (0: automated; 1: manual intervention needed; 2: not possible).
-\item Elastic engines (0: easy to implement; 1: difficult to
-  implement; 2: not available).
-\item Distributed engines (0: possible; 1: not possible).
-\item Task scheduling (0: difficult; 1: more difficult).
-\end{itemize}
+These 2 metrics are summed to obtain a total number of missing other
+features in the architecture.  \todo{Maybe we could revise the
+  categories to avoid such a vague ``other features''.}
 
@@
 
+The different architectures described in
+Figure~\ref{fig:architectures} are evaluated along these metrics in
+the remainder of this Section.
+
 \subsection{Tight integration}
 
-\paragraph{Integration} Tightly integrating a workflow engine in a
-science gateway requires to modify the science gateway
-(\texttt{I$_1$=1}) and to develop interactions \texttt{b}, and
-\texttt{c$_2$} in the workflow engine (\texttt{I$_2$=2}). No other
-modification is required (\texttt{I$_3$}=\texttt{I$_4$}=0).
-
-\paragraph{Robustness} The science gateway and the infrastructure are
-involved in a workflow execution (\texttt{R$_1$}=2), with interactions
-\texttt{b}, \texttt{c$_1$} and \texttt{c$_2$}
-(\texttt{R$_2$}=3). Debugging is not an issue since the science
-gateway can retrieve any information from the workflow engine
-directly (\texttt{R$_3$}=0).
-
-\paragraph{Modularity} Integrating a new type of workflow engine
+\paragraph{Integration} This architecture does not require any
+component in addition to the science gateway and infrastructure
+(\texttt{I$_1$=2}). It involves 5 interactions: \texttt{a$_1$},
+\texttt{a$_2$}, \texttt{b}, \texttt{c$_1$} and \texttt{c$_2$}
+(\texttt{I$_2$=5}).
+
+\paragraph{Robustness} No component is specific to workflow execution
+(\texttt{R$_1$}=0), but interactions \texttt{b} and \texttt{c$_2$}
+(\texttt{R$_2$}=2) are.
+
+\paragraph{Extensibility} Integrating a new type of workflow engine
 requires to modify the science gateway as well as interactions
-\texttt{b} and \texttt{c$_2$} (\texttt{M$_1$}=3). Updating a workflow
+\texttt{b} and \texttt{c$_2$} (\texttt{E$_1$}=3). Updating a workflow
 engine version requires modifications in the science gateway
-(\texttt{M$_2$}=1). Adding a new infrastructure generates updates in
+(\texttt{E$_2$}=1).  Inserting a new workflow is done through
+interaction \texttt{a} (\texttt{E$_3$}=1). Adding a new infrastructure generates updates in
 interactions \texttt{b}, \texttt{c$_1$} and \texttt{c$_2$}
-(\texttt{M$_3$}=3). Inserting a new workflow is done through
-interaction \texttt{a} (\texttt{M$_4$}=1). Meta-workflows are not
-supported by default (\texttt{M$_4$}=1).
+(\texttt{E$_4$}=3).
 
@@
 meant to deal with short web requests rather than long workflow
-executions (\texttt{S$_1$}=2). Coherently, elastic engines are usually
-not possible either (\texttt{S$_2$}=2).  Distributed engines are not
-available by default (\texttt{S$_3$}=1). The scheduling of tasks on
+executions (\texttt{S$_1$}=2). Distributed engines are not
+available by default (\texttt{S$_2$}=1). The scheduling of tasks on
 the infrastructure is as complex as in any other architecture since
 the workflow engine might implement any kind of scheduling policy
-(\texttt{S$_4$}=0).
+(\texttt{S$_3$}=0).
+
+\paragraph{Other features} Meta-workflows are not supported by default
+(\texttt{O$_1$}=1).  Debugging is not an issue since the science
+gateway can retrieve any information from the workflow engine directly
+(\texttt{O$_2$}=0).
 
@@
 
-\paragraph{Integration} Integrating a workflow engine as a service
-requires to develop 3 interactions and 1 component. First, a workflow
-service has to be developed and interfaced with the infrastructure
-through interactions \texttt{b} and \texttt{c$_2$}
-(\texttt{I$_2$}=3). In addtion, the science gateway has to be extended
-to call the engine interface \texttt{d} (\texttt{I$_1$}=1).
-
-\paragraph{Robustness} Workflow execution requires 3 components: the
-science gateway, the workflow service, and the infrastructure
-(\texttt{R$_1$}=3). It involves the 4 interactions \texttt{b},
-\texttt{$c_1$}, \texttt{$c_2$} and \texttt{d}
-(\texttt{R$_2$}=4). Fine-grained debugging information is usually easy
-to obtain since the workflow service provides direct access to the
-engine (\texttt{R$_3$}=0).
-
-\paragraph{Modularity} Adding a new type of workflow engine requires
+\paragraph{Integration} Service invocation requires a workflow service
+in addition to the science gateway and infrastructure
+(\texttt{I$_1$}=3). The architecture involves 6 interactions:
+\texttt{a$_1$}, \texttt{a$_2$}, \texttt{b}, \texttt{c$_1$},
+\texttt{c$_2$} and \texttt{d} (\texttt{I$_2$}=6).
+
+\paragraph{Robustness} The workflow service is a component specific to
+workflow execution (\texttt{R$_1$}=1). Workflow execution also
+involves 3 specific interactions: \texttt{b}, 
+\texttt{c$_2$} and \texttt{d} (\texttt{R$_2$}=3).
+
+\paragraph{Extensibility} Adding a new type of workflow engine requires
 to implement the corresponding workflow service, to modify interaction
 \texttt{d}, and to implement interactions \texttt{b} and
-\texttt{$c_2$} (\texttt{M$_1$}=4). New engine versions can be added by
+\texttt{$c_2$} (\texttt{E$_1$}=4). New engine versions can be added by
 updating the workflow service without modifying any interaction
-(\texttt{M$_2$}=0). Adding a new type of infrastructure requires
+(\texttt{E$_2$}=0).  New workflows are added in the science gateway or
+in the workflow engine through interaction \texttt{a} only
+(\texttt{E$_3$}=1). Adding a new type of infrastructure requires
 updates in interactions \texttt{b}, \texttt{$c_1$} and \texttt{$c_2$}
-(\texttt{M$_3$}=3). New workflows are added in the science gateway or
-in the workflow engine through interaction \texttt{a} only
-(\texttt{M$_4$}=1). Meta-workflow are not supported by default
-(\texttt{M$_5$}=1).
+(\texttt{E$_4$}=3).
 
@@
 the science gateway configuration when a new engine instance is
-available (\texttt{S$_2$}=1). Distributing the execution of a single
+available. Distributing the execution of a single
 workflow in multiple engines is usually not possible unless the
-workflow engine has specific abilities (\texttt{S$_3$}=1). The
+workflow engine has specific abilities (\texttt{S$_2$}=1). The
 scheduling of tasks on the infrastructure is as complex as in any
 other architecture since the workflow engine might implement any kind
-of scheduling policy (\texttt{S$_4$}=0).
+of scheduling policy (\texttt{S$_3$}=0).
+
+\paragraph{Other features} Meta-workflow are not supported by default
+(\texttt{O$_1$}=1). Fine-grained debugging information is usually easy
+to obtain since the workflow service provides direct access to the
+engine (\texttt{O$_2$}=0).
 
@@
 
-\paragraph{Integration} Integrating a workflow engine as a sub-task
-only requires to implement interaction \texttt{e} (\texttt{I$_2$}=1)
-and to install the workflow engine on the infrastructure
-(\texttt{I$_3$}=1). A new task type is created in the science gateway
-only when a new workflow has to be integrated (see M$_4$), but this is
-not required to integrate the engine itself (\texttt{I$_1$}=0).
-
-\paragraph{Robustness} Only 2 components and 3 interactions are
-involved in workflow execution (\texttt{R$_1$}=2,
-\texttt{R$_2$}=3). Obtaining fine-grained information about workflow
+\paragraph{Integration} The sub-task architecture requires only 2
+components (\texttt{I$_1$}=2).  It involves 5 interactions:
+\texttt{a$_1$}, \texttt{a$_2$}, \texttt{b}, \texttt{c} and
+\texttt{e} (\texttt{I$_2$}=5).
+
+\paragraph{Robustness} No component is specific to workflow execution
+(\texttt{R$_1$}=0), and only interaction \texttt{e} is
+(\texttt{R$_2$}=1).
+
+\paragraph{Extensibility} Integrating a new type of workflow engine
+requires to develop interaction \texttt{e} and to install the engine
+on the infrastructure (\texttt{E$_1$}=2). Updating an engine version
+requires updates only on the infrastructure (\texttt{E$_2$}=1).  New
+workflows are integrated by creating a new task in the science gateway
+through interaction \texttt{a} (\texttt{E$_3$}=1). Adding a new
+infrastructure requires to update interactions \texttt{b} and
+\texttt{c} in the science gateway (\texttt{E$_4$}=2).
+
+\paragraph{Other features} Meta-workflows are not available
+(\texttt{O$_1$}=1).  Obtaining fine-grained information about workflow
 activities is not straightforward since the science gateway has no
 knowledge about the workflow topology, and the workflow engine is
-integrated as a task (\texttt{R$_3$}=1).
-
-\paragraph{Modularity} Integrating a new type of workflow engine
-requires to develop interaction \texttt{e} and to install the engine
-on the infrastructure (\texttt{M$_1$}=2). Updating an engine version
-requires updates only on the infrastructure (\texttt{M$_2$}=1). Adding
-a new infrastructure requires to update interactions \texttt{b} and
-\texttt{c} in the science gateway (\texttt{M$_3$}=2). New workflows are integrated by
-creating a new task in the science gateway through interaction
-\texttt{a} (\texttt{M$_4$}=1), and meta-workflows are not available
-(\texttt{M$_5$}=1).
+integrated as a task (\texttt{O$_2$}=1).
+
+%\todo{Integrating a workflow engine as a sub-task
+%only requires to implement interaction \texttt{e} (\texttt{I$_2$}=1)
+%and to install the workflow engine on the infrastructure
+%(\texttt{I$_3$}=1). A new task type is created in the science gateway
+%only when a new workflow has to be integrated (see M$_4$), but this is
+%not required to integrate the engine itself (\texttt{I$_1$}=0).}
 
@@
 New engine instances are spawned and executed on the infrastructure as
-any other task upon user submission (\texttt{S$_1$}=0,
-\texttt{S$_2$}=0). Such scalability properties are one of the major
-interests of the sub-task architecture. Distributed engines are not
-supported by default (\texttt{S$_3$}=1). Task scheduling is slightly
-more complex than in the other approaches due to the special role of
-the task that executes the workflow engine (\texttt{S$_4$}=1). Indeed,
-the reliability of this task is critical since all the sub-tasks in
-the workflow depend on it and, depending on the recovery capabilities
-of the workflow engine, may need to be resubmitted if the workflow
-task fails. The workflow task is also longer than all its sub-tasks,
-which increases its chances of failure. In addition, task parameters,
-for instance estimated walltime, are more difficult to estimate for
-the workflow task than for sub-tasks which may generate issues such as
-selection of wrong batch queues on clusters. Finally, the
-interdependencies between the workflow task and its sub-tasks may
-create deadlocks when there is contention. Say for instance that only
-1 computing resource is available for the science gateway and that the
-workflow task is running on it and submits sub-tasks, then the
-sub-tasks could only execute when the resource is available, which
-will never happen because the workflow task will not complete until
-the sub-tasks complete. This configuration can be generalized to an
-infrastructure with $n$ resources where $n$ workflows are
-submitted. In practice, however, the number of submitted workflows
-usually remains lower than the number of computing resources available
-on this infrastructure, which makes such deadlocks unlikely to happen.
+any other task upon user submission (\texttt{S$_1$}=0). Such
+scalability properties are one of the major interests of the sub-task
+architecture. Distributed engines are not supported by default
+(\texttt{S$_2$}=1). Task scheduling is slightly more complex than in
+the other approaches due to the special role of the task that executes
+the workflow engine (\texttt{S$_3$}=1). Indeed, the reliability of
+this task is critical since all the sub-tasks in the workflow depend
+on it and, depending on the recovery capabilities of the workflow
+engine, may need to be resubmitted if the workflow task fails. The
+workflow task is also longer than all its sub-tasks, which increases
+its chances of failure. In addition, task parameters, for instance
+estimated walltime, are more difficult to estimate for the workflow
+task than for sub-tasks which may generate issues such as selection of
+wrong batch queues on clusters. Finally, the interdependencies between
+the workflow task and its sub-tasks may create deadlocks when there is
+contention. Say for instance that only 1 computing resource is
+available for the science gateway and that the workflow task is
+running on it and submits sub-tasks, then the sub-tasks could only
+execute when the resource is available, which will never happen
+because the workflow task will not complete until the sub-tasks
+complete. This configuration can be generalized to an infrastructure
+with $n$ resources where $n$ workflows are submitted. In practice,
+however, the number of submitted workflows usually remains lower than
+the number of computing resources available on this infrastructure,
+which makes such deadlocks unlikely to happen.
 
@@
 
-\paragraph{Integration} Integrating workflow engines through the pool
-model requires interaction \texttt{d} in the science gateway
-(\texttt{I$_1$=1}), interactions \texttt{b} and \texttt{c$_2$} in the
-workflow engine (\texttt{I$_2$=2}), the development of the workflow
-pool, agent and interaction \texttt{f} between them
-(\texttt{I$_4$=3}).
-
-\paragraph{Robustness} The science gateway, workflow pool, agent and
-infrastructure are involved in a workflow execution
-(\texttt{R$_1$=4}), and these components are connected through 4
-interactions (\texttt{R$_2$=4}). Accessing debugging information is
-not likely to be an issue since the workflow pool could implement
-specific functions for that (\texttt{R$_3$=0}).
-
-\paragraph{Modularity} Adding a new engine type requires to wrap the
-engine in the agent and to update interactions \texttt{b} and
-\texttt{c$_2$} (\texttt{M$_1$=3}). Updating the version of an engine
-is transparent (\texttt{M$_2$=0}), and integrating a new
-infrastructure requires updates in interactions \texttt{b},
-\texttt{c$_1$} and \texttt{c$_2$} (\texttt{M$_3$=3}). Integrating a
-new workflow is done through interaction \texttt{a} only
-(\texttt{M$_4$=1}) and meta-workflows are not available by default
-(\texttt{M$_5$=1}).
+\paragraph{Integration} The pool model requires a workflow pool and an
+agent in addition to the science gateway and infrastructure
+(\texttt{I$_1$=4}). It involves 7 interactions: \texttt{a$_1$},
+\texttt{a$_2$}, \texttt{b}, \texttt{c$_1$}, \texttt{c$_2$}, \texttt{d}
+and \texttt{f} (\texttt{I$_2$}=7).
+
+\paragraph{Robustness} The workflow pool and agent are specific to
+workflow execution (\texttt{R$_1$=2}). Interactions \texttt{b},
+\texttt{c$_2$}, \texttt{d} and \texttt{f} also are (\texttt{R$_2$=4}).
+
+\paragraph{Extensibility} Adding a new engine type requires to wrap
+the engine in the agent and to update interactions \texttt{b} and
+\texttt{c$_2$} (\texttt{E$_1$=3}). Updating the version of an engine
+is transparent (\texttt{E$_2$=0}), and integrating a new workflow is
+done through interaction \texttt{a} only
+(\texttt{E$_3$=1}). Integrating a new infrastructure requires updates
+in interactions \texttt{b}, \texttt{c$_1$} and \texttt{c$_2$}
+(\texttt{E$_4$=3}).
 
@@
 which is easily automated (\texttt{S$_1$=0}) and by design very
-suitable for elastic computing (\texttt{S$_2$=0}). For instance,
-auto-scaling rules can be implemented to start new agents when the
-workload in the science gateway exceeds a certain threshold \todo{ref
-  needed}. Distributed engines are not available by default
-(\texttt{S$_3$=1}) and task scheduling is as complex as in any other
-architecture (\texttt{S$_4$=0}).
+suitable for elastic computing. For instance, auto-scaling rules can
+be implemented to start new agents when the workload in the science
+gateway exceeds a certain threshold \todo{ref needed}. Distributed
+engines are not available by default (\texttt{S$_2$=1}) and task
+scheduling is as complex as in any other architecture
+(\texttt{S$_3$=0}).
+
+\paragraph{Other features} Meta-workflows are not available by default
+(\texttt{O$_1$=1}).  Accessing debugging information is not likely to
+be an issue since the workflow pool could implement specific functions
+for that (\texttt{O$_2$=0}).
 
@@
 \paragraph{Integration} Setting up a nested workflow architecture with
-service invocation requires interaction \texttt{d} in the science
-gateway (\texttt{I$_1$=1}), and the development of 2 workflow services
-with 2 instances of interactions \texttt{b} and \texttt{c$_2$}
-(\texttt{I$_2$=5}).
-
-\paragraph{Robustness} Workflow execution involves the science
-gateway, 2 workflow services and the infrastructure
-(\texttt{R$_1$=4}), which includes 7 interactions
-(\texttt{R$_2$=7}). Debugging is difficult because the science gateway
-cannot directly access fine-grained information in the sub-workflow
-(\texttt{R$_3$=1}).
-
-\paragraph{Modularity} Adding a new type of \emph{parent} engine
+service invocation requires a science gateway, 2 workflow services and
+2 infrastructures (\texttt{I$_1$=5}). The architecture involves 11
+interactions (\texttt{I$_2$=11}): \texttt{a$_1$} (twice: once for each
+infrastructure), \texttt{a$_2$}, \texttt{b} (twice: once for each
+infrastructure), \texttt{c$_1$} (twice: once for each infrastructure),
+\texttt{c$_2$} (twice: once for each workflow engine and
+infrastructure) and \texttt{d} (twice: once for each workflow engine).
+
+\paragraph{Robustness} The two workflow services are specific to
+workflow execution (\texttt{R$_1$=2}). Interactions \texttt{b}
+(counted twice), \texttt{c$_2$} (twice) and \texttt{d} (twice) also
+are (\texttt{R$_2$=6}).
+
+\paragraph{Extensibility} Adding a new type of \emph{parent} engine
 requires to implement the corresponding service, to implement
@@
 to implement interaction \texttt{d} in the science gateway and in the
-parent service (\texttt{M$_1$=5}). Adding a new type of \emph{child}
+parent service (\texttt{E$_1$=5}). Adding a new type of \emph{child}
 engine only requires to implement the corresponding service, to
@@
 engine, and to implement interaction \texttt{d} in the parent service
-(\texttt{M$_1$=4}). We use \texttt{M$_1$=4.5} in
+(\texttt{E$_1$=4}). We use \texttt{E$_1$=4.5} in
 Table~\ref{table:evaluation} to reflect both conditions. Adding a new
 version in the parent or child engine only requires modifying this
-engine (\texttt{M$_2$=0}). Adding a new infrastructure requires to
+engine (\texttt{E$_2$=0}). Adding a new
+workflow is done through interaction \texttt{a}
+(\texttt{E$_3$=1}). Adding a new infrastructure requires to
 re-implement interactions \texttt{b} and \texttt{c$_2$} twice, and
-interaction \texttt{c$_1$} once (\texttt{M$_3$=5}). Adding a new
-workflow is done through interaction \texttt{a}
-(\texttt{M$_4$=1}). Meta-workflows are possible, which is one of the
-main interest of this architecture (\texttt{M$_5$=0}).
+interaction \texttt{c$_1$} once (\texttt{E$_4$=5}).
 
@@
 child engine) (\texttt{S$_1$=1}). Similarly, elastic engines are
-difficult to achieve (\texttt{S$_2$=1}). Distributed engines are
-possible, through meta-workflows (\texttt{S$_3$=0}). Task scheduling
+difficult to achieve. Distributed engines are
+possible, through meta-workflows (\texttt{S$_2$=0}). Task scheduling
 is more complex than in other architectures though, due to the fact
-that workflow execution is split in different engines (\texttt{S$_4$=1}).
+that workflow execution is split in different engines (\texttt{S$_3$=1}).
+
+\paragraph{Other features} 
+Meta-workflows are possible, which is one of the main interest of this
+architecture (\texttt{O$_1$=0}).  Debugging is difficult because the
+science gateway cannot directly access fine-grained information in the
+sub-workflow (\texttt{O$_2$=1}).
+
 \subsection{Workflow import with service invocation}
 
-\paragraph{Integration} Compared to the service architecture, workflow
-import requires an additional interaction \texttt{g} in the science
-gateway (\texttt{I$_1$}=2, \texttt{I$_2$}=3).
+\paragraph{Integration} Workflow import with service invocation
+requires the same components as for service invocation
+(\texttt{I$_1$}=3). It requires an additional \texttt{g} interaction
+for workflow import (\texttt{I$_1$}=7)
 
@@
 the execution (it is an offline process), metrics are as in the
-service architecture (\texttt{R$_1$=3}, \texttt{R$_2$=4},
-\texttt{R$_3$=0}).
-
-\paragraph{Modularity} Since adding a new type of workflow engine aims
-at supporting more workflows, we consider that it only requires to
-re-implement interaction \texttt{g} in this architecture. Note,
+service architecture (\texttt{R$_1$=1}, \texttt{R$_2$=3}).
+
+\paragraph{Extensibility} Since adding a new type of workflow engine
+aims at supporting more workflows, we consider that it only requires
+to re-implement interaction \texttt{g} in this architecture. Note,
 however, that implementing interaction \texttt{g} can require very
 substantial work depending on the complexity of the language used by
-the new engine (\texttt{M$_1$=1}). Based on the same logic, adding a
+the new engine (\texttt{E$_1$=1}). Based on the same logic, adding a
 new engine version only requires modifying interaction \texttt{g}
-(\texttt{M$_2$=1}). As in the service architecture, interfacing with a
-new infrastructure requires modifications in the workflow service and
-in interactions \texttt{b} and \texttt{c$_2$}
-(\texttt{M$_3$=3}). Adding a new workflow is done through interactions
-\texttt{a} and \texttt{g} (\texttt{M$_4$=2}). Meta-workflows are
-available after import, by connecting workflows in the language used
-in the science gateway (\texttt{M$_5$=0}).
+(\texttt{E$_2$=1}).  Adding a new workflow is done through
+interactions \texttt{a} and \texttt{g} (\texttt{E$_3$=2}). As in the
+service architecture, interfacing with a new infrastructure requires
+modifications in the workflow service and in interactions \texttt{b}
+and \texttt{c$_2$} (\texttt{E$_4$=3}).
 
@@
 the execution (it is an offline process), metrics are as in the
-service architecture (\texttt{S$_1$=1}, \texttt{S$_2$=1},
-\texttt{S$_3$=1}, \texttt{S$_4$=0}).
+service architecture (\texttt{S$_1$=1},
+\texttt{S$_2$=1}, \texttt{S$_3$=0}).
+
+\paragraph{Other features} Meta-workflows are available after import,
+by connecting workflows in the language used in the science gateway
+(\texttt{O$_1$=0}). Debugging information is accessible as in the
+service invocation architecture (\texttt{O$_2$=0}).
 
@@
 \begin{tabular}{rcccccc}
-\textbf{Metric}                      & \textbf{Tight}
+                                     & \textbf{Tight}
                                      & \textbf{Service}
@@
                                      & \textbf{Import} \\
-\multicolumn{7}{c}{\cellcolor[HTML]{EEEEEE}\textbf{Integration}}\\
-Science gateway -- \texttt{I$_1$}    & \cellcolor[HTML]{99FF99}1
+\multicolumn{7}{c}{\cellcolor[HTML]{EEEEEE}\textbf{Integration effort}}\\
+  Total components -- \texttt{I$_1$} & \cellcolor[HTML]{99FF99}2
+                                     & \cellcolor[HTML]{99E899}3
+                                     & \cellcolor[HTML]{99FF99}2
+                                     & \cellcolor[HTML]{99D199}4
+                                     & \cellcolor[HTML]{99BB99}5
+                                     & \cellcolor[HTML]{99E899}3\\
+Total interactions -- \texttt{I$_2$} & \cellcolor[HTML]{99FF99}5
+                                     & \cellcolor[HTML]{99F399}6
+                                     & \cellcolor[HTML]{99FF99}5
+                                     & \cellcolor[HTML]{99E899}7
+                                     & \cellcolor[HTML]{99BB99}11
+                                     & \cellcolor[HTML]{99E899}7\\
+\textbf{Total} (total software pieces) & \cellcolor[HTML]{99FF99}7
+                                     & \cellcolor[HTML]{99EF99}9
+                                     & \cellcolor[HTML]{99FF99}7
+                                     & \cellcolor[HTML]{99E099}11
+                                     & \cellcolor[HTML]{99BB99}16
+                                     & \cellcolor[HTML]{99E899}10\\
+  \multicolumn{7}{c}{\cellcolor[HTML]{EEEEEE}\textbf{Robustness of workflow execution}}\\
+Specific components -- \texttt{R$_1$} & \cellcolor[HTML]{99FF99}0
+                                     & \cellcolor[HTML]{99DD99}1
+                                     & \cellcolor[HTML]{99FF99}0
+                                     & \cellcolor[HTML]{99BB99}2
+                                     & \cellcolor[HTML]{99BB99}2
+                                     & \cellcolor[HTML]{99DD99}1\\
+  Specific interactions -- \texttt{R$_2$} & \cellcolor[HTML]{99F199}2
+                                     & \cellcolor[HTML]{99E399}3
                                      & \cellcolor[HTML]{99FF99}1
+                                     & \cellcolor[HTML]{99D699}4
+                                     & \cellcolor[HTML]{99BB99}6
+                                     & \cellcolor[HTML]{99E399}3\\
+  \textbf{Total} (specific software pieces)& \cellcolor[HTML]{99F599}2
+                                     & \cellcolor[HTML]{99E199}4
+                                     & \cellcolor[HTML]{99FF99}1
+                                     & \cellcolor[HTML]{99CE99}6
+                                     & \cellcolor[HTML]{99BB99}8
+                                     & \cellcolor[HTML]{99E199}4\\
+\multicolumn{7}{c}{\cellcolor[HTML]{EEEEEE}\textbf{Extensibility}}\\
+  New engine type -- \texttt{E$_1$}  & \cellcolor[HTML]{99D899}3
+                                     & \cellcolor[HTML]{99C499}4
+                                     & \cellcolor[HTML]{99EB99}2
+                                     & \cellcolor[HTML]{99D899}3
+                                     & \cellcolor[HTML]{99BB99}4.5
+                                     & \cellcolor[HTML]{99FF99}1\\
+New engine version -- \texttt{E$_2$} & \cellcolor[HTML]{99BB99}1
                                      & \cellcolor[HTML]{99FF99}0
-                                     & \cellcolor[HTML]{99FF99}1
-                                     & \cellcolor[HTML]{99FF99}1
-                                     & \cellcolor[HTML]{99AA99}2 \\
-Workflow engine -- \texttt{I$_2$}    & \cellcolor[HTML]{99AA99}2
-                                     & \cellcolor[HTML]{99FF99}3 
-                                     & \cellcolor[HTML]{99FF99}1
-                                     & \cellcolor[HTML]{99AA99}2
-                                     & \cellcolor[HTML]{99AA99}5
-                                     & \cellcolor[HTML]{99FF99}3 \\
-Infrastructure -- \texttt{I$_3$}   & \cellcolor[HTML]{99FF99}0
-                                     & \cellcolor[HTML]{99FF99}0
-                                     & \cellcolor[HTML]{99FF99}1
-                                     & \cellcolor[HTML]{99AA99}0
-                                     & \cellcolor[HTML]{99AA99}0
-                                     & \cellcolor[HTML]{99FF99}0 \\
-Other -- \texttt{I$_4$}              & \cellcolor[HTML]{99FF99}0
+                                     & \cellcolor[HTML]{99BB99}1
                                      & \cellcolor[HTML]{99FF99}0
                                      & \cellcolor[HTML]{99FF99}0
-                                     & \cellcolor[HTML]{99AA99}3
-                                     & \cellcolor[HTML]{99AA99}0
-                                     & \cellcolor[HTML]{99FF99}0 \\
-\textbf{Total}                       & \cellcolor[HTML]{99FF99}\textbf{3}
-                                     & \cellcolor[HTML]{99FF99}\textbf{4}
-                                     & \cellcolor[HTML]{99FF99}\textbf{2}
-                                     & \cellcolor[HTML]{99AA99}\textbf{6}
-                                     & \cellcolor[HTML]{99AA99}\textbf{6}
-                                     & \cellcolor[HTML]{99AA99}\textbf{5} \\
-
-\multicolumn{7}{c}{\cellcolor[HTML]{EEEEEE}\textbf{Robustness}}\\
-Components  --   \texttt{R$_1$}      & \cellcolor[HTML]{99FF99}2
-                                     & \cellcolor[HTML]{99DD99}3
-                                     & \cellcolor[HTML]{99FF99}2
-                                     & \cellcolor[HTML]{99AA99}4  
-                                     & \cellcolor[HTML]{99AA99}4 
-                                     & \cellcolor[HTML]{99AA99}3 \\
-Interactions -- \texttt{R$_2$}       & \cellcolor[HTML]{99FF99}3    
-                                     & \cellcolor[HTML]{99DD99}4
-                                     & \cellcolor[HTML]{99DD99}3      
-                                     & \cellcolor[HTML]{99AA99}4  
-                                     & \cellcolor[HTML]{99AA99}7      
-                                     & \cellcolor[HTML]{99AA99}4 \\
-Debugging --    \texttt{R$_3$}       & \cellcolor[HTML]{99AA99}0          
-                                     & \cellcolor[HTML]{99FF99}0
-                                     & \cellcolor[HTML]{99AA99}1
-                                     & \cellcolor[HTML]{99FF99}0
-                                     & \cellcolor[HTML]{99AA99}1
-                                     & \cellcolor[HTML]{99FF99}0 \\
-\textbf{Total}                       & \cellcolor[HTML]{99FF99}\textbf{5}
-                                     & \cellcolor[HTML]{99CC99}\textbf{7}
-                                     & \cellcolor[HTML]{99CC99}\textbf{6}
-                                     & \cellcolor[HTML]{99BB99}\textbf{8}
-                                     & \cellcolor[HTML]{99AA99}\textbf{12}
-                                     & \cellcolor[HTML]{99BB99}\textbf{7} \\
-\multicolumn{7}{c}{\cellcolor[HTML]{EEEEEE}\textbf{Modularity}}\\
-New engine type -- \texttt{M$_1$}    & \cellcolor[HTML]{99AA99}3
-                                     & \cellcolor[HTML]{99DD99}4
-                                     & \cellcolor[HTML]{99FF99}2
-                                     & \cellcolor[HTML]{99DD99}3
-                                     & \cellcolor[HTML]{99DD99}4.5
-                                     & \cellcolor[HTML]{99AA99}1 \\
-New engine version -- \texttt{M$_2$} & \cellcolor[HTML]{99AA99}1
-                                     & \cellcolor[HTML]{99FF99}0
-                                     & \cellcolor[HTML]{99FF99}1
-                                     & \cellcolor[HTML]{99FF99}0
-                                     & \cellcolor[HTML]{99FF99}0
-                                     & \cellcolor[HTML]{99FF99}1 \\
-New infrastructure -- \texttt{M$_3$} & \cellcolor[HTML]{99FF99}3
-                                     & \cellcolor[HTML]{99FF99}3
-                                     & \cellcolor[HTML]{99FF99}2
-                                     & \cellcolor[HTML]{99FF99}3
-                                     & \cellcolor[HTML]{99FF99}5
-                                     & \cellcolor[HTML]{99FF99}3 \\
-New workflow -- \texttt{M$_4$}       & \cellcolor[HTML]{99FF99}1
+                                     & \cellcolor[HTML]{99BB99}1\\
+  New workflow -- \texttt{E$_3$} & \cellcolor[HTML]{99FF99}1
                                      & \cellcolor[HTML]{99FF99}1
@@
                                      & \cellcolor[HTML]{99FF99}1
-                                     & \cellcolor[HTML]{99AA99}2 \\
-Meta-workflow  -- \texttt{M$_5$}     & \cellcolor[HTML]{99AA99}1
-                                     & \cellcolor[HTML]{99AA99}1  
-                                     & \cellcolor[HTML]{99AA99}1
-                                     & \cellcolor[HTML]{99AA99}1
-                                     & \cellcolor[HTML]{99FF99}0
-                                     & \cellcolor[HTML]{99AA99}0 \\
-\textbf{Total}                       & \cellcolor[HTML]{99AA99}\textbf{9}
-                                     & \cellcolor[HTML]{99EE99}\textbf{9}
-                                     & \cellcolor[HTML]{99FF99}\textbf{7}
-                                     & \cellcolor[HTML]{99EE99}\textbf{8}
-                                     & \cellcolor[HTML]{99FF99}\textbf{10.5}
-                                     & \cellcolor[HTML]{99AA99}\textbf{7} \\
+                                     & \cellcolor[HTML]{99BB99}2\\
+New infrastructure -- \texttt{E$_4$} & \cellcolor[HTML]{99BB99}4
+                                     & \cellcolor[HTML]{99BB99}4
+                                     & \cellcolor[HTML]{99FF99}3
+                                     & \cellcolor[HTML]{99BB99}4
+                                     & \cellcolor[HTML]{99BB99}4
+                                     & \cellcolor[HTML]{99BB99}4\\
+  \textbf{Total} (difficulty to extend) & \cellcolor[HTML]{99C899}9
+                                     & \cellcolor[HTML]{99C899}9
+                                     & \cellcolor[HTML]{99FF99}7
+                                     & \cellcolor[HTML]{99E399}8
+                                     & \cellcolor[HTML]{99BB99}9.5
+                                     & \cellcolor[HTML]{99E399}8\\
 \multicolumn{7}{c}{\cellcolor[HTML]{EEEEEE}\textbf{Scalability}}\\
-New engine instance -- \texttt{S$_1$}& \cellcolor[HTML]{99AA99}2
-                                     & \cellcolor[HTML]{99AA99}1
-                                     & \cellcolor[HTML]{99FF99}0
-                                     & \cellcolor[HTML]{99FF99}0
-                                     & \cellcolor[HTML]{99AA99}1
-                                     & \cellcolor[HTML]{99AA99}1 \\
-Elastic engines -- \texttt{S$_2$}    & \cellcolor[HTML]{99AA99}2
+New engine instance -- \texttt{S$_1$}& \cellcolor[HTML]{99BB99}2
                                      & \cellcolor[HTML]{99DD99}1
@@
                                      & \cellcolor[HTML]{99DD99}1
-                                     & \cellcolor[HTML]{99DD99}1 \\
-Distributed engines -- \texttt{S$_3$}& \cellcolor[HTML]{99AA99}1
-                                     & \cellcolor[HTML]{99AA99}1
-                                     & \cellcolor[HTML]{99AA99}1
-                                     & \cellcolor[HTML]{99AA99}1
+                                     & \cellcolor[HTML]{99DD99}1\\
+Distributed engines -- \texttt{S$_2$}& \cellcolor[HTML]{99BB99}1
+                                     & \cellcolor[HTML]{99BB99}1
+                                     & \cellcolor[HTML]{99BB99}1
+                                     & \cellcolor[HTML]{99BB99}1
                                      & \cellcolor[HTML]{99FF99}0
-                                     & \cellcolor[HTML]{99AA99}1 \\
-Task scheduling -- \texttt{S$_4$}    & \cellcolor[HTML]{99FF99}0
+                                     & \cellcolor[HTML]{99BB99}1\\
+Task scheduling -- \texttt{S$_3$}    & \cellcolor[HTML]{99FF99}0
                                      & \cellcolor[HTML]{99FF99}0
-                                     & \cellcolor[HTML]{99AA99}1
+                                     & \cellcolor[HTML]{99BB99}1
                                      & \cellcolor[HTML]{99FF99}0
+                                     & \cellcolor[HTML]{99BB99}1
+                                     & \cellcolor[HTML]{99FF99}0\\
+\textbf{Total} (scalability issues)  & \cellcolor[HTML]{99BB99}3
+                                     & \cellcolor[HTML]{99DD99}2
+                                     & \cellcolor[HTML]{99DD99}2
                                      & \cellcolor[HTML]{99FF99}1
-                                     & \cellcolor[HTML]{99FF99}0 \\
-\textbf{Total}                       & \cellcolor[HTML]{99AA99}\textbf{5}
-                                     & \cellcolor[HTML]{99BB99}\textbf{3}
-                                     & \cellcolor[HTML]{99DD99}\textbf{2}
-                                     & \cellcolor[HTML]{99FF99}\textbf{1}
-                                     & \cellcolor[HTML]{99DD99}\textbf{3}
-                                     & \cellcolor[HTML]{99BB99}\textbf{3} \\
-\multicolumn{7}{c}{\cellcolor[HTML]{EEEEEE}\textbf{Grand total}}\\
-                                     & \cellcolor[HTML]{99CC99}\textbf{22}
-                                     & \cellcolor[HTML]{99BB99}\textbf{23}
-                                     & \cellcolor[HTML]{99FF99}\textbf{17}
-                                     & \cellcolor[HTML]{99CC99}\textbf{23}
-                                     & \cellcolor[HTML]{99DD99}\textbf{31.5}
-                                     & \cellcolor[HTML]{99AA99}\textbf{22}
+                                     & \cellcolor[HTML]{99DD99}2
+                                     & \cellcolor[HTML]{99DD99}2\\
+\multicolumn{7}{c}{\cellcolor[HTML]{EEEEEE}\textbf{Other features}}\\
+  Meta-workflow  -- \texttt{O$_1$}    & \cellcolor[HTML]{99BB99}1
+                                     & \cellcolor[HTML]{99BB99}1
+                                     & \cellcolor[HTML]{99BB99}1
+                                     & \cellcolor[HTML]{99BB99}1
+                                     & \cellcolor[HTML]{99FF99}0
+                                     & \cellcolor[HTML]{99FF99}0\\
+  Fine-grained debugging -- \texttt{O$_2$}   & \cellcolor[HTML]{99FF99}0
+                                     & \cellcolor[HTML]{99FF99}0
+                                     & \cellcolor[HTML]{99BB99}1
+                                     & \cellcolor[HTML]{99FF99}0
+                                     & \cellcolor[HTML]{99BB99}1
+                                     & \cellcolor[HTML]{99FF99}0\\
+  \textbf{Total} (missing other features) & \cellcolor[HTML]{99DD99}1
+                                     & \cellcolor[HTML]{99DD99}1
+                                     & \cellcolor[HTML]{99BB99}2
+                                     & \cellcolor[HTML]{99DD99}1
+                                     & \cellcolor[HTML]{99DD99}1
+                                     & \cellcolor[HTML]{99FF99}0\\
 \end{tabular}
-\caption{Architecture evaluation.
-Lower values (brighter colors) indicate better performance.
-\todo{update colors and find a more relevant way to compute a grand total without favoring criteria that have more metrics than the others (e.g.
-normalize each criterion between 0 and 1).
-Same comment for each sub-total.
-M1 might be the same as integration.
-Update nested workflow as infrastructures were separated.}}
+
+\caption{Architecture evaluation. Lower values (brighter colors) indicate better performance.}
 \label{table:evaluation}
@@
 
-Comparison between architectures, per criterion (robustness, etc).
-
-This ignores pre-existing work (e.g. a workflow engine is already available as a service). Migration across architectures too.
+Comparison between architectures, per criterion (robustness,
+etc). Which architecture is best in each category?
+
+Limitations: 
+\begin{itemize}
+\item This ignores pre-existing work (e.g. a workflow engine is already available as a service). Migration across architectures too.
+\item The difficulty of implementing interaction g is not properly
+reflected. For instance, fsl uses bash, which would be very difficult to convert.
+\item We are well aware that all the
+components and interactions may not require the same level of
+development effort depending on the particular targeted platform.
+\end{itemize}
 
@@
 
-The difficulty of implementing interaction g is not properly
-reflected. For instance, fsl uses bash, which would be very difficult to convert.
-
 \section{Conclusion}
+
+We abstracted architectures so that we can compare them using
+objective metrics. We illustrated these architectural patterns on
+examples of real systems.
+
+What this work highlights and recommendations for science gateway
+developers.
 
@@
 
-\todo{Review use of capitals in references}
 \bibliographystyle{elsarticle-num} 

### 2016-04-18 paper.tex
Edits in section 5 and 4

@@
 The architectures described in Section~\ref{sec:architectures} are
-evaluated on Table~\ref{table:evaluation}. We use 5 main criteria to
+evaluated in Table~\ref{table:evaluation}. We use 5 main criteria to
 evaluate the architectures: integration effort, robustness of workflow
@@
 metrics in a criterion measure similar entities so that summation
-makes sense. Metric values obtained by each architecture are
-normalized between 0 (best value) and 1 (worst value) to determine the
-color of the corresponding table cell (brighter color indicate better
-performance\footnote{Normalized metric value $m'$ is
-  $\frac{m-m_{\mathrm{min}}}{m_{\mathrm{min}}-m_{\mathrm{max}}}$ where
-  $m$ is the initial metric value,
-  $m_{\mathrm{min}}$/$m_{\mathrm{max}}$ are the minimal/maximal values
-  among all architectures. Cell color is \todo{\#99XX99, where
-    X=F-4m'})}. The criteria and metrics are explained hereafter.
+makes sense. The
+criteria and metrics are explained hereafter.
 
@@
 
-\todo{Explain these metrics better, including why the related features
-  are useful.}
-
 \emph{Integration} measures the development effort required to build
-the architecture. It is obtained by counting the number of
+the architecture. It is obtained by counting the total number of
 interactions and components on the architectural diagram. It breaks
-down to the following 2 metrics:
-\begin{itemize}[itemsep=0cm]
-\item Components \texttt{I$_1$}),
-\item Interactions (\texttt{I$_2$}).
+down to the folloowing 2 metrics:
+\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
+\item Total number of components (\texttt{I$_1$}),
+\item Total number of interactions (\texttt{I$_2$}).
 \end{itemize}
-One may wonder whether infrastructure should be part of
-\texttt{I$_1$}) since it is usually not developed by the groups who
+One may wonder whether infrastructure should be counted in
+\texttt{I$_1$} since it is usually not developed by the groups who
 integrate workflow engines in science gateway. However, integrating an
@@
 effort (e.g. account creation, software installation, etc) which is
-why we keep it in the metric. The two metrics are summed to obtain
-the total score for this criterion, which measures the total number of
+why we keep it in the metric. The two metrics are summed to obtain the
+total score for this criterion, which measures the total number of
 software pieces to develop.
@@
 \emph{Robustness of workflow execution} measures the likelihood that
-workflow execution fails due to errors coming from the software
-architecture. It is determined as the number of software components and
-interactions that are specific to workflow execution. It assume that
-complex architectures tend to be prone to failure.  These components
-and interactions are represented in red on the architecture diagram in
-Figure~\ref{fig:architectures}. Robustness breaks down to 2 metrics:
+workflow execution fails due to errors in the components or
+interactions of the software architecture. It is a critical criterion
+since it directly impacts the end users of the science gateway.
+Robustness is determined as the number of software components and
+interactions that are specific to workflow execution, i.e. not reused
+by other functions:
 \begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
 \item Components (\texttt{R$_1$}): number of specific components
-  involved in workflow executions.
+  involved in workflow executions. Science
+gateway, for instance, is \emph{not} specific to workflow execution
+since it is used to authenticate users, add new workflows, transfer
+user data, etc. 
 \item Interactions (\texttt{R$_2$}): number of specific interactions
@@
 \end{itemize}
-The two metrics are summed to obtain the total score for this
+These specific components and interactions are represented in red on
+the architecture diagrams in Figure~\ref{fig:architectures}. The two
+metrics assume that complex architectures tend to be prone to
+failure. The two metrics are summed to obtain the total score for this
 criterion, which measures the total number of software pieces that are
-specifically involved in workflow execution. Robustness of workflow
-execution is critical since it directly impacts the end users of the
-science gateway.
+specifically involved in workflow execution.
 
 \emph{Extensibility} measures the difficulty to replace or add
-elements in the architecture. It is obtained as the number of
+elements in the architecture. It is determined as the number of
 interaction and component to modify when a new element is
@@
 
-\caption{Architecture evaluation. Lower values (brighter colors) indicate better performance.}
+\caption{Architecture evaluation. Lower values (brighter colors) indicate better performance. On each row, metric values are
+  normalized between 0 (best value) and 1 (worst value) to determine the
+  color of the corresponding table cell. More precisely, the normalized metric value $m'$ is
+  defined as
+  $\frac{m-m_{\mathrm{min}}}{m_{\mathrm{min}}-m_{\mathrm{max}}}$ where
+  $m$ is the initial metric value, $m_{\mathrm{min}}$/$m_{\mathrm{max}}$
+  are the minimal/maximal values among all architectures. The RGB hexadecimal color code of the cell
+  is \texttt{\#99XX99}, where X=F-4m'.}
 \label{table:evaluation}
@@
 etc). Which architecture is best in each category?
+
+Overall comparison, see Table~\ref{table:overall}.
+
+\begin{table*}
+\centering
+\begin{tabular}{cccccc}
+  \textbf{Tight}
+  & \textbf{Service}
+  & \textbf{Sub-task}
+  & \textbf{Pool}
+  & \textbf{Nested}
+  & \textbf{Import} \\
+\cellcolor[HTML]{99E599}2.4
+  & \cellcolor[HTML]{99E599}2.5
+  & \cellcolor[HTML]{99FF99}1.5
+  & \cellcolor[HTML]{99EF99}2.1
+  & \cellcolor[HTML]{99BB99}4.0
+  & \cellcolor[HTML]{99FA99}1.7\\
+\end{tabular}
+\caption{Overall evaluation. Brighter colors indicate better performance.}
+\label{table:overall}
+\end{table*}
 

### 2016-04-19 paper.tex
Updates in section 4

@@
 
-\emph{Integration} measures the development effort required to build
+\emph{Integration effort} measures the development required to build
 the architecture. It is obtained by counting the total number of
-interactions and components on the architectural diagram. It breaks
-down to the folloowing 2 metrics:
+interactions and components on the architecture diagram. It breaks
+down to the following 2 metrics:
 \begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
@@
 \emph{Robustness of workflow execution} measures the likelihood that
-workflow execution fails due to errors in the components or
-interactions of the software architecture. It is a critical criterion
-since it directly impacts the end users of the science gateway.
+workflow execution fails due to errors in the components or with the
+interactions in the software architecture. 
 Robustness is determined as the number of software components and
@@
 \end{itemize}
-These specific components and interactions are represented in red on
-the architecture diagrams in Figure~\ref{fig:architectures}. The two
-metrics assume that complex architectures tend to be prone to
-failure. The two metrics are summed to obtain the total score for this
+These specific components and interactions are represented in red in
+the architecture diagrams in
+Figure~\ref{fig:architectures}. \texttt{R$_1$} and \texttt{R$_2$}
+assume that complex architectures tend to be prone to
+failure. \texttt{R$_1$} and \texttt{R$_2$}  are summed to obtain the total score for this
 criterion, which measures the total number of software pieces that are
@@
 elements in the architecture. It is determined as the number of
-interaction and component to modify when a new element is
-added. Modification of an interaction is deemed necessary when the
-parameters involved in this interaction are modified. Modification of
-a component is required when its code needs to be modified or
-recompiled (science gateway or workflow service), or when a new piece
-of software has to be installed (infrastructure only). Extensibility
+interactions and components to modify when a new element is
+added. Modification of a component is required when its code needs to
+be modified or recompiled (science gateway or workflow service), or
+when a new piece of software has to be installed (infrastructure
+only). Modification of an interaction is deemed necessary when the
+parameters involved in this interaction are modified.  Extensibility
 breaks down in 4 metrics depending on the type of element that has to
@@
 \item New engine type (\texttt{E$_1$}): number of interactions or
-  components to modify to integrate a new workflow engine type in the
-  architecture. Workflow engine belong to different types when they
-  cannot be invoked using the same interface. Adding a new type of
-  workflow engine allows to execute more workflows in the science gateway.
+  components to modify to integrate a new type of workflow engine in
+  the architecture. Workflow engines belong to different types when
+  they cannot be invoked using the same interface. Adding a new type
+  of workflow engine allows to execute more workflows in the science
+  gateway.
 \item New engine version (\texttt{E$_2$}): number of interactions or
@@
   have to be considered as different engine types. 
-\item New workflow (\texttt{E$_3$}): number of interactions required
-  to integrate a new workflow in the architecture, assuming that the
-  engine type and version required to execute this workflow are
-  already available.
+\item New workflow (\texttt{E$_3$}): adding a new workflow is a very
+  common operation that does not require modifying software components
+  or interactions. We measure the difficulty to integrate a new
+  workflow by counting the number of interactions that need to be
+  invoked to integrate a new workflow in the architecture, assuming
+  that the engine type and version required to execute this workflow
+  are already available.
 \item New infrastructure (\texttt{E$_4$}): number of interactions or components to
@@
   (e.g. constrain data to remain in a particular network domain).
-\item New science gateway: \todo{Not sure if this makes sense}.
+%\item New science gateway: \todo{Not sure if this makes sense}.
 \end{itemize}
@@
 \emph{Scalability} corresponds to the ability of the architecture to
-cope with high workloads. It is measured by reviewing the features
-that enable particular aspects of scalability. Features are evaluated
-using a 3-level metric: 0 means that the feature is very easy to
-enable, 1 means that it can be implemented but with some difficulty,
-and 2 means that it is not possible to implement the feature without a
-nonsensical amount of effort. Four different features are identified:
+cope with high workloads. It is measured by counting the potential
+scalability issues in the architectures, i.e. the lack of scalability
+features. Features are evaluated using a 3-level metric: 0 means that
+the feature is very easy to enable, 1 means that it can be implemented
+but with some difficulty, and 2 means that the feature could only be
+implemented with a nonsensical amount of effort. Four different
+features are identified:
 \begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
-\item Adding engine instance (\texttt{S$_1$}): this feature
-  corresponds to the possibility to add more than 1 engine instance in
-  the architecture. Workflow engines may require important amounts of
-  resources when a lot of workflows, or large workflows, are
-  executed. At some point, it may be required for the science gateway
-  to distribute the load to several engines. \texttt{S$_1$}=0 when
-  adding a new engine instance is a fully automated process, i.e. the
-  workflow engine only has to be started. In this case, elastic
-  engines are possible, i.e. some kind of auto-scaling mechanism can
-  be implemented to control the number of engine instances in the
-  architecture. \texttt{S$_1$}=1 when adding a new engine instance
-  requires some form of manual intervention, which prevents easy
-  implementations of elastic engines. \texttt{S$_1$}=2 when new engine
-  instances cannot be added.
+\item Multiple engine instances (\texttt{S$_1$}): possibility to have
+  more than 1 engine instance in the architecture. Workflow engines
+  may require important amounts of resources when a lot of workflows,
+  or large workflows, are executed. At some point, it may be required
+  for the science gateway to distribute the load to several
+  engines. \texttt{S$_1$}=0 when adding a new engine instance is a
+  fully automated process, i.e. the workflow engine only has to be
+  started. In this case, elastic engines are possible, i.e. some kind
+  of auto-scaling mechanism can be implemented to control the number
+  of engine instances in the architecture. \texttt{S$_1$}=1 when
+  adding a new engine instance requires some form of manual
+  intervention, which prevents easy implementations of elastic
+  engines. \texttt{S$_1$}=2 when new engine instances cannot be added.
 \item Distributed engines (\texttt{S$_2$}): possibility to distribute
@@
 The 3 metrics are summed to obtain a global measure of the scalability
-issues available in the architecture.
+issues that are present in the architecture.
 
@@
 \begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
-\item Meta-workflow (\texttt{O$_1$}): ability to describe meta-workflows from existing
-  workflows.  \todo{More details about why this is important?} \texttt{O$_1$}=0 when the feature is not available, \texttt{O$_1$}=1 otherwise.
+\item Meta-workflow (\texttt{O$_1$}): ability to describe
+  meta-workflows from existing workflows.  \todo{More details about
+    why this is important?} \texttt{O$_1$}=0 when the feature is
+  available, \texttt{O$_1$}=1 otherwise.
 \item Fine-grained debugging (\texttt{O$_2$}): availability of
-  debugging information about workflow activities. Fine-grained
-  information about workflow activity is required to properly
-  troubleshoot workflow executions.  \texttt{O$_2$}=0 when such
-  information is easy to access, \texttt{O$_2$}=1
+  fine-grained debugging information about workflow activities. When
+  such information is available, workflow is seen as a white
+  box. Fine-grained information about workflow activity is required to
+  properly troubleshoot workflow executions.  \texttt{O$_2$}=0 when
+  such information is easy to access, \texttt{O$_2$}=1
   otherwise. \todo{More details about why this is important?}
 \end{itemize}
-These 2 metrics are summed to obtain a total number of missing other
+\texttt{O$_1$} and \texttt{O$_2$} are summed to obtain a total number of missing other
 features in the architecture.  \todo{Maybe we could revise the
@@
 \paragraph{Robustness} No component is specific to workflow execution
-(\texttt{R$_1$}=0), but interactions \texttt{b} and \texttt{c$_2$}
-(\texttt{R$_2$}=2) are.
+(\texttt{R$_1$}=0), but interactions \texttt{b} and \texttt{c$_2$} are
+(\texttt{R$_2$}=2).
 
@@
 (\texttt{E$_2$}=1).  Inserting a new workflow is done through
-interaction \texttt{a} (\texttt{E$_3$}=1). Adding a new infrastructure generates updates in
+interactions \texttt{a$_1$} and \texttt{a$_2$} (\texttt{E$_3$}=2). Adding a new infrastructure generates updates in
 interactions \texttt{b}, \texttt{c$_1$} and \texttt{c$_2$}
@@
 instance of the science gateway, which is in general not possible
-since load-balancing mechanisms available in web servers are usually
-meant to deal with short web requests rather than long workflow
-executions (\texttt{S$_1$}=2). Distributed engines are not
-available by default (\texttt{S$_2$}=1). The scheduling of tasks on
-the infrastructure is as complex as in any other architecture since
-the workflow engine might implement any kind of scheduling policy
-(\texttt{S$_3$}=0).
+(\texttt{S$_1$}=2). Distributed engines are not available by default
+(\texttt{S$_2$}=1). The scheduling of tasks on the infrastructure is
+as complex as in any other architecture since the workflow engine
+might implement any kind of scheduling policy (\texttt{S$_3$}=0).
 
@@
 (\texttt{E$_2$}=0).  New workflows are added in the science gateway or
-in the workflow engine through interaction \texttt{a} only
-(\texttt{E$_3$}=1). Adding a new type of infrastructure requires
+in the workflow engine through interactions \texttt{a$_1$} and \texttt{a$_2$}
+(\texttt{E$_3$}=2). Adding a new type of infrastructure requires
 updates in interactions \texttt{b}, \texttt{$c_1$} and \texttt{$c_2$}
@@
 workflows are integrated by creating a new task in the science gateway
-through interaction \texttt{a} (\texttt{E$_3$}=1). Adding a new
+through interactions \texttt{a$_1$} and \texttt{a$_2$}  (\texttt{E$_3$}=2). Adding a new
 infrastructure requires to update interactions \texttt{b} and
@@
 New engine instances are spawned and executed on the infrastructure as
-any other task upon user submission (\texttt{S$_1$}=0). Such
-scalability properties are one of the major interests of the sub-task
-architecture. Distributed engines are not supported by default
-(\texttt{S$_2$}=1). Task scheduling is slightly more complex than in
-the other approaches due to the special role of the task that executes
-the workflow engine (\texttt{S$_3$}=1). Indeed, the reliability of
-this task is critical since all the sub-tasks in the workflow depend
-on it and, depending on the recovery capabilities of the workflow
-engine, may need to be resubmitted if the workflow task fails. The
-workflow task is also longer than all its sub-tasks, which increases
-its chances of failure. In addition, task parameters, for instance
-estimated walltime, are more difficult to estimate for the workflow
-task than for sub-tasks which may generate issues such as selection of
-wrong batch queues on clusters. Finally, the interdependencies between
-the workflow task and its sub-tasks may create deadlocks when there is
+any other task upon user submission (\texttt{S$_1$}=0). This is a
+major interest of the sub-task architecture. Distributed engines are
+not supported by default (\texttt{S$_2$}=1). Task scheduling is
+slightly more complex than in the other approaches due to the special
+role of the task that executes the workflow engine
+(\texttt{S$_3$}=1). Indeed, the reliability of this task is critical
+since all the sub-tasks in the workflow depend on it and, depending on
+the recovery capabilities of the workflow engine, may need to be
+resubmitted if the workflow task fails. The workflow task is also
+longer than all its sub-tasks, which increases its chances of
+failure. In addition, task parameters, for instance estimated
+walltime, are more difficult to estimate for the workflow task than
+for sub-tasks which may generate issues such as selection of wrong
+batch queues on clusters. Finally, the interdependencies between the
+workflow task and its sub-tasks may create deadlocks when there is
 contention. Say for instance that only 1 computing resource is
@@
 \texttt{c$_2$} (\texttt{E$_1$=3}). Updating the version of an engine
-is transparent (\texttt{E$_2$=0}), and integrating a new workflow is
-done through interaction \texttt{a} only
-(\texttt{E$_3$=1}). Integrating a new infrastructure requires updates
+is transparent (\texttt{E$_2$=0}) \todo{explain why, also for service}, and integrating a new workflow is
+done through interactions \texttt{a$_1$} and \texttt{a$_2$}
+(\texttt{E$_3$=2}). Integrating a new infrastructure requires updates
 in interactions \texttt{b}, \texttt{c$_1$} and \texttt{c$_2$}
@@
 engine (\texttt{E$_2$=0}). Adding a new
-workflow is done through interaction \texttt{a}
-(\texttt{E$_3$=1}). Adding a new infrastructure requires to
+workflow is done through interaction \texttt{a$_1$} and \texttt{a$_2$}
+(\texttt{E$_3$=2}). Adding a new infrastructure requires to
 re-implement interactions \texttt{b} and \texttt{c$_2$} twice, and
@@
 aims at supporting more workflows, we consider that it only requires
-to re-implement interaction \texttt{g} in this architecture. Note,
+to re-implement interaction \texttt{g} in this architecture  (\texttt{E$_1$=1}). Note,
 however, that implementing interaction \texttt{g} can require very
 substantial work depending on the complexity of the language used by
-the new engine (\texttt{E$_1$=1}). Based on the same logic, adding a
+the new engine. Based on the same logic, adding a
 new engine version only requires modifying interaction \texttt{g}
 (\texttt{E$_2$=1}).  Adding a new workflow is done through
-interactions \texttt{a} and \texttt{g} (\texttt{E$_3$=2}). As in the
+interactions \texttt{a$_1$}, \texttt{a$_2$} and \texttt{g} (\texttt{E$_3$=3}). As in the
 service architecture, interfacing with a new infrastructure requires
@@
                                      & \cellcolor[HTML]{99E899}7\\
-\textbf{Total} (total software pieces) & \cellcolor[HTML]{99FF99}7
-                                     & \cellcolor[HTML]{99EF99}9
-                                     & \cellcolor[HTML]{99FF99}7
-                                     & \cellcolor[HTML]{99E099}11
-                                     & \cellcolor[HTML]{99BB99}16
-                                     & \cellcolor[HTML]{99E899}10\\
+\textbf{Total} (total software pieces) & \cellcolor[HTML]{99FF99}\textbf{7}
+                                     & \cellcolor[HTML]{99EF99}\textbf{9}
+                                     & \cellcolor[HTML]{99FF99}\textbf{7}
+                                     & \cellcolor[HTML]{99E099}\textbf{11}
+                                     & \cellcolor[HTML]{99BB99}\textbf{16}
+                                     & \cellcolor[HTML]{99E899}\textbf{10}\\
   \multicolumn{7}{c}{\cellcolor[HTML]{EEEEEE}\textbf{Robustness of workflow execution}}\\
@@
                                      & \cellcolor[HTML]{99E399}3\\
-  \textbf{Total} (specific software pieces)& \cellcolor[HTML]{99F599}2
-                                     & \cellcolor[HTML]{99E199}4
-                                     & \cellcolor[HTML]{99FF99}1
-                                     & \cellcolor[HTML]{99CE99}6
-                                     & \cellcolor[HTML]{99BB99}8
-                                     & \cellcolor[HTML]{99E199}4\\
+  \textbf{Total} (specific software pieces)& \cellcolor[HTML]{99F599}\textbf{2}
+                                     & \cellcolor[HTML]{99E199}\textbf{4}
+                                     & \cellcolor[HTML]{99FF99}\textbf{1}
+                                     & \cellcolor[HTML]{99CE99}\textbf{6}
+                                     & \cellcolor[HTML]{99BB99}\textbf{8}
+                                     & \cellcolor[HTML]{99E199}\textbf{4}\\
 \multicolumn{7}{c}{\cellcolor[HTML]{EEEEEE}\textbf{Extensibility}}\\
@@
                                      & \cellcolor[HTML]{99BB99}1\\
-  New workflow -- \texttt{E$_3$} & \cellcolor[HTML]{99FF99}1
-                                     & \cellcolor[HTML]{99FF99}1
-                                     & \cellcolor[HTML]{99FF99}1
-                                     & \cellcolor[HTML]{99FF99}1
-                                     & \cellcolor[HTML]{99FF99}1
-                                     & \cellcolor[HTML]{99BB99}2\\
-New infrastructure -- \texttt{E$_4$} & \cellcolor[HTML]{99BB99}4
-                                     & \cellcolor[HTML]{99BB99}4
-                                     & \cellcolor[HTML]{99FF99}3
-                                     & \cellcolor[HTML]{99BB99}4
-                                     & \cellcolor[HTML]{99BB99}4
-                                     & \cellcolor[HTML]{99BB99}4\\
-  \textbf{Total} (difficulty to extend) & \cellcolor[HTML]{99C899}9
-                                     & \cellcolor[HTML]{99C899}9
-                                     & \cellcolor[HTML]{99FF99}7
-                                     & \cellcolor[HTML]{99E399}8
-                                     & \cellcolor[HTML]{99BB99}9.5
-                                     & \cellcolor[HTML]{99E399}8\\
+  New workflow -- \texttt{E$_3$} & \cellcolor[HTML]{99FF99}2
+                                     & \cellcolor[HTML]{99FF99}2
+                                     & \cellcolor[HTML]{99FF99}2
+                                     & \cellcolor[HTML]{99FF99}2
+                                     & \cellcolor[HTML]{99FF99}2
+                                     & \cellcolor[HTML]{99BB99}3\\
+New infrastructure -- \texttt{E$_4$} & \cellcolor[HTML]{99E899}3
+                                     & \cellcolor[HTML]{99E899}3
+                                     & \cellcolor[HTML]{99FF99}2
+                                     & \cellcolor[HTML]{99E899}3
+                                     & \cellcolor[HTML]{99BB99}5
+                                     & \cellcolor[HTML]{99E899}3\\
+  \textbf{Total} (difficulty to extend) & \cellcolor[HTML]{99E099}\textbf{9}
+                                     & \cellcolor[HTML]{99E099}\textbf{9}
+                                     & \cellcolor[HTML]{99FF99}\textbf{7}
+                                     & \cellcolor[HTML]{99EF99}\textbf{8}
+                                     & \cellcolor[HTML]{99BB99}\textbf{11.5}
+                                     & \cellcolor[HTML]{99EF99}\textbf{8}\\
 \multicolumn{7}{c}{\cellcolor[HTML]{EEEEEE}\textbf{Scalability}}\\
-New engine instance -- \texttt{S$_1$}& \cellcolor[HTML]{99BB99}2
+Multiple engine instances -- \texttt{S$_1$}& \cellcolor[HTML]{99BB99}2
                                      & \cellcolor[HTML]{99DD99}1
@@
                                      & \cellcolor[HTML]{99FF99}0\\
-\textbf{Total} (scalability issues)  & \cellcolor[HTML]{99BB99}3
-                                     & \cellcolor[HTML]{99DD99}2
-                                     & \cellcolor[HTML]{99DD99}2
-                                     & \cellcolor[HTML]{99FF99}1
-                                     & \cellcolor[HTML]{99DD99}2
-                                     & \cellcolor[HTML]{99DD99}2\\
+\textbf{Total} (scalability issues)  & \cellcolor[HTML]{99BB99}\textbf{3}
+                                     & \cellcolor[HTML]{99DD99}\textbf{2}
+                                     & \cellcolor[HTML]{99DD99}\textbf{2}
+                                     & \cellcolor[HTML]{99FF99}\textbf{1}
+                                     & \cellcolor[HTML]{99DD99}\textbf{2}
+                                     & \cellcolor[HTML]{99DD99}\textbf{2}\\
 \multicolumn{7}{c}{\cellcolor[HTML]{EEEEEE}\textbf{Other features}}\\
@@
                                      & \cellcolor[HTML]{99FF99}0\\
-  \textbf{Total} (missing other features) & \cellcolor[HTML]{99DD99}1
-                                     & \cellcolor[HTML]{99DD99}1
-                                     & \cellcolor[HTML]{99BB99}2
-                                     & \cellcolor[HTML]{99DD99}1
-                                     & \cellcolor[HTML]{99DD99}1
-                                     & \cellcolor[HTML]{99FF99}0\\
+  \textbf{Total} (missing other features) & \cellcolor[HTML]{99DD99}\textbf{1}
+                                     & \cellcolor[HTML]{99DD99}\textbf{1}
+                                     & \cellcolor[HTML]{99BB99}\textbf{2}
+                                     & \cellcolor[HTML]{99DD99}\textbf{1}
+                                     & \cellcolor[HTML]{99DD99}\textbf{1}
+                                     & \cellcolor[HTML]{99FF99}\textbf{0}\\
 \end{tabular}
@@
   & \textbf{Import} \\
-\cellcolor[HTML]{99E599}2.4
-  & \cellcolor[HTML]{99E599}2.5
-  & \cellcolor[HTML]{99FF99}1.5
-  & \cellcolor[HTML]{99EF99}2.1
+\cellcolor[HTML]{99EE99}2.1
+  & \cellcolor[HTML]{99EE99}2.1
+  & \cellcolor[HTML]{99FE99}1.5
+  & \cellcolor[HTML]{99F499}1.9
   & \cellcolor[HTML]{99BB99}4.0
-  & \cellcolor[HTML]{99FA99}1.7\\
+  & \cellcolor[HTML]{99FF99}1.5\\
 \end{tabular}

### 2016-04-19 paper.tex
Updates in Section 3

@@
 components include science gateway and infrastructure in all
-architectures, and workflow service, workflow pool and agent are
+architectures while workflow service, workflow pool and agent are
 involved in some architectures. The workflow engine itself is
-represented by a specific symbol. Software interactions among these
-components are described in the next Section. \emph{Abstract}
-interactions are specific types of interactions that may be
-implemented by various different software interactions. An
-architecture that has an abstract interaction is an abstract
-architecture. Red color is used to represent the components and
-interactions that are specific to workflow execution, i.e. that are
-not involved in any other process. Tasks refer to computing tasks that
-are created by the workflow engine and executed on the
+represented by a specific symbol with dotted lines. Software
+interactions among these components are represented with
+arrows. \emph{Abstract} interactions are specific types of
+interactions that may be implemented by various different software
+interactions. They are represented with dotted arrows and labeled with
+ \texttt{*}. An architecture that has an abstract interaction is an
+abstract architecture. Red color is used to represent the components
+and interactions that are specific to workflow execution, i.e. that
+are not involved in any other process. Tasks refer to computing tasks
+that are created by the workflow engine and executed on the
 infrastructure. Data represents any type of file, or database that is
@@
   the infrastructure, which may or may not require administrative
-  privileges on the infrastructure; (\texttt{a$_2$}) the workflow has
-  to be configured in the science gateway so that it becomes available
-  to users. Note that integrating a workflow is not the same process
-  as integrating a workflow \emph{engine}.
+  privileges on the infrastructure; (\texttt{a$_2$}) the workflow is
+  configured in the science gateway so that it becomes available to
+  users. Note that integrating a workflow is not the same process as
+  integrating a workflow \emph{engine}.
 \item Task control: operations to manage tasks on the infrastructure,
@@
 
+\begin{figure*}
+\centering
+\def\svgwidth{1.5\columnwidth}
+\input{figures/VIP.pdf_tex}
+\caption{Architecture of the Virtual Imaging Platform that integrates
+  the MOTEUR engine through service invocation.}
+\label{fig:vip-architecture}
+\end{figure*}
+
 See Figure~\ref{archi:service}. The workflow engine is available
@@
 Figure~\ref{fig:vip-architecture} corresponds to interaction
-\texttt{a$_1$}, and step \texttt{4} maps to interactions \texttt{b}
+\texttt{c$_1$}, and step \texttt{4} maps to interactions \texttt{b}
 and \texttt{c$_2$} (in VIP, workflow data transfers are performed by
 the tasks and embedded in their descriptions).
-
-\begin{figure*}
-\centering
-\def\svgwidth{1.5\columnwidth}
-\input{figures/VIP.pdf_tex}
-\caption{Architecture of the Virtual Imaging Platform that integrates
-  the MOTEUR workflow engine through service invocation.}
-\label{fig:vip-architecture}
-\end{figure*}
 
@@
 Figure~\ref{fig:architectures} is implemented by steps \texttt{4},
-\texttt{5}, \texttt{7} and \texttt{8}. Interaction \texttt{c} is
-implemented by steps \texttt{3} and \texttt{10}. Interaction
-\texttt{e} is implemented by step \texttt{6}. The PSOM workflow engine
-adopts a pilot-job architecture~\cite{turilli2015comprehensive} where
-the a master coordinates workflow execution by submitting workers
-and establishing direct communication channels with them. Note how
-this peculiar execution model is totally supported by the sub-tasking
-architecture.
+\texttt{5}, \texttt{7} and \texttt{8} (regular interactions with batch
+manager). Interaction \texttt{c} is implemented by steps \texttt{3}
+and \texttt{10}. Interaction \texttt{e} is implemented by step
+\texttt{6}. The PSOM workflow engine adopts a pilot-job
+architecture~\cite{turilli2015comprehensive} where the a master
+coordinates workflow execution by submitting workers and establishing
+direct communication channels with them. Note how this peculiar
+execution model is totally supported by the sub-tasking architecture.
 
@@
   the sub-tasking model (Figure extracted from~\cite{GLAT-16}). 1:
-  User sends data and pipeline execution request to storage server(s)
+  User sends data and workflow execution request to storage server(s)
   and CBRAIN portal. 2: CBRAIN portal sends execution request to
@@
   server. 7-8: Execution server starts PSOM workers through task
-  scheduler. 9: PSOM master and workers execute pipeline. 10.
+  scheduler. 9: PSOM master and workers execute workflow. 10.
   Execution server transfers results to storage server(s). \todo{Improve graphics}}
@@
 \texttt{c$_2$}, science gateway transfers user data through
-interaction \texttt{c$_1$}, and administrator installs workflow
-dependencies through interaction \texttt{a$_1$}.
+interaction \texttt{c$_1$}, and administrator installs workflows
+through interaction \texttt{a$_1$} and \texttt{a$_2$}.
 
@@
 of workflow engines to execute workflows expressed with different
-languages. Calls \texttt{0}, \texttt{7}, \texttt{8} and \texttt{9} are
-used by workflow engine plugins to declare their supported language
-and launch engines, and by workflow engines to report their status to
-the agent. These calls are specific to the SHIWA Pool implementation
-of the \texttt{agent} component and therefore have no corresponding
-representation in Figure~\ref{archi:agent}.
+languages. Calls \texttt{0}, \texttt{7}, \texttt{8} and \texttt{9} on
+Figure~\ref{fig:shiwa-pool-architecture} are used by workflow engine
+plugins to declare their supported language and launch engines, and by
+workflow engines to report their status to the agent. These calls are
+specific to the SHIWA Pool implementation of the \texttt{agent}
+component and therefore have no corresponding representation in
+Figure~\ref{archi:agent}.
 
@@
 \emph{parent} workflow executed by the \emph{parent} engine is itself
-a \emph{child} workflow that is executed by a \emph{child} engine. The
+a workflow, called \emph{child} workflow, that is executed by a \emph{child} engine. The
 parent and child engines might use different workflow description
@@
 instantiated in the various architectures described previously. We
-focus on instantiation with the service invocation model(see
+focus on instantiation with the service invocation model (see
 Figure~\ref{archi:nested}-Right) as this is the most used
@@
 wrapped in workflow activities as any other command-line tool. Nested
-workflows were also notably used by the SHIWA Science Gateway to
-implement so-called Coarse-Grained workflow
+workflows were notably used by the SHIWA Science Gateway to implement
+so-called Coarse-Grained workflow
 interoperability~\cite{terstyanszky2014enabling}, i.e. to integrate
@@
 service invocation as represented in Figure~\ref{archi:nested}. The
-parent workflow engine is WS-PGRADE, invoked as a service in the
-Science Gateway (step \texttt{1} on
+parent workflow engine is \texttt{WS-PGRADE}, invoked as a service in
+the Science Gateway (step \texttt{1} on
 Figure~\ref{fig:shiwa-architecture}, interaction \texttt{d} on
@@
 submit jobs and transfer data to a distributed computing
-infrastructure (DCI, step \texttt{3}, interactions \texttt{b} and
-\texttt{c$_2$}). Data interactions (\texttt{c}) and application
+infrastructure (\texttt{DCI}, step \texttt{3}, interactions \texttt{b}
+and \texttt{c$_2$}). Data interactions (\texttt{c}) and application
 porting ones (\texttt{a}) are not represented on

### 2016-04-19 paper.tex
Wrote section 5

@@
 
-Comparison between architectures, per criterion (robustness,
-etc). Which architecture is best in each category?
-
-Overall comparison, see Table~\ref{table:overall}.
-
+\subsection{Comparison between architectures}
+
+Tight integration and sub-tasking require the least amount of
+integration effort, closely followed by service integration and
+workflow import (with service invocation). Pool is a bit behind but
+still before nested workflows (with service invocation) which requires
+more integration than the other architectures. Robustness leads to the
+same ordering of architectures, with tight integration and sub-tasking
+in the top group, service integration and workflow import (with
+service invocation) close behind, pool in a third group, and nested
+workflows (with service invocation) at the end. This ordering is
+consistent across metrics, and it reflects the global complexity of
+the architectures.
+
+Regarding extensibility, most architectures are overall comparable,
+except nested workflows (with service invocation) which is
+significantly behind. This is explained by the complexity of the
+nested workflow architecture, with 2 infrastructures and 2 workflow
+services. Sub-tasking and workflow import (with service invocation)
+perform a bit better to integrate new engines (\texttt{E$_1$}), which
+is interesting given the current profusion of engines with different
+characteristics.  However, sub-tasking, workflow import (with service
+invocation) and tight integration perform worse than the others to add
+engine versions (\texttt{E$_2$}) due to the need to update
+infrastructure (sub-tasking), science gateway (tight integration), or
+workflow converter (workflow import). Workflow import performs worse
+than the others to integrate new workflows (\texttt{E$_3$}) because of
+the language conversion step. All architectures except nested
+workflows perform the same to integrate new infrastructures (\texttt{E$_4$}).
+
+The pool architecture is overall the most scalable, which is not a
+surprise since it was designed precisely for scalability. Tight
+integration is the least scalable and all the other architectures
+perform the same. The overall scalability score, however, should not
+mask the unique characteristics of architectures regarding this
+criterion. For instance, nested workflows are the only architecture
+that can easily accommodate distributed workflow execution, which can
+be critical in some cases. At the same time, the scheduling
+constraints created by the sub-tasking and nested workflow
+architectures could well become showstoppers depending on the type of
+targeted infrastructure. Non-reliable infrastructures, for example,
+could hardly cope with workflow engines being wrapped in computing
+tasks as done in sub-tasking. The availability of multiple engine
+instances could also become a critical feature for science gateways
+with important workloads, which would favor sub-tasking and pool.
+
+Differences in other features should not be neglected. Nested
+workflows (with service invocation) and workflow import (with service
+invocation) are the only architectures that support meta-workflows,
+which may be interesting in some cases. The availability of
+fine-grained debugging information may also be critical to efficient
+user support.
+
+Table~\ref{table:overall} provides an overall comparison between
+architectures, based on the metrics in
+Table~\ref{table:evaluation}. While Table~\ref{table:overall} provides
+a global comparison between architectures, it should not be used
+without the detailed metrics reported on Table~\ref{table:evaluation}
+and discussed above. Overall, sub-tasking, and workflow import (with
+service invocation) perform a bit better than tight integration,
+service invocation and pool. Nested workflows stands a bit behind for
+the reasons mentioned previously.
 \begin{table*}
 \centering
-\begin{tabular}{cccccc}
-  \textbf{Tight}
+\begin{tabular}{rcccccc}
+                                    & \textbf{Tight}
                                     & \textbf{Service}
@@
                                     & \textbf{Import} \\
-\cellcolor[HTML]{99EE99}2.1
-  & \cellcolor[HTML]{99EE99}2.1
-  & \cellcolor[HTML]{99FE99}1.5
-  & \cellcolor[HTML]{99F499}1.9
-  & \cellcolor[HTML]{99BB99}4.0
-  & \cellcolor[HTML]{99FF99}1.5\\
+  Integration effort (normalized total) & 
+                                    \cellcolor[HTML]{99FF99}{0.00}
+                                    & \cellcolor[HTML]{99EF99}{0.22}
+                                    & \cellcolor[HTML]{99FF99}{0.00}
+                                    & \cellcolor[HTML]{99E099}{0.44}
+                                    & \cellcolor[HTML]{99BB99}{1.00}
+                                      & \cellcolor[HTML]{99E899}{0.33}\\                           
+Robustness (normalized total) & 
+                                \cellcolor[HTML]{99F599}{0.14}
+                                    & \cellcolor[HTML]{99E199}{0.43}
+                                    & \cellcolor[HTML]{99FF99}{0.00}
+                                    & \cellcolor[HTML]{99CE99}{0.71}
+                                    & \cellcolor[HTML]{99BB99}{1.00}
+                                    & \cellcolor[HTML]{99E199}{0.43}\\
+  Extensibility (normalized total)  & \cellcolor[HTML]{99E099}{0.44}
+                                     & \cellcolor[HTML]{99E099}{0.44}
+                                     & \cellcolor[HTML]{99FF99}{0.00}
+                                     & \cellcolor[HTML]{99EF99}{0.22}
+                                     & \cellcolor[HTML]{99BB99}{1.00}
+                                     & \cellcolor[HTML]{99EF99}{0.22}\\
+Scalability (normalized total)  & \cellcolor[HTML]{99BB99}{1.00}
+                                     & \cellcolor[HTML]{99DD99}{0.50}
+                                     & \cellcolor[HTML]{99DD99}{0.50}
+                                     & \cellcolor[HTML]{99FF99}{0.00}
+                                     & \cellcolor[HTML]{99DD99}{0.50}
+                                     & \cellcolor[HTML]{99DD99}{0.50}\\
+Other features (normalized total) & \cellcolor[HTML]{99DD99}{0.50}
+                                     & \cellcolor[HTML]{99DD99}{0.50}
+                                     & \cellcolor[HTML]{99BB99}{1.00}
+                                     & \cellcolor[HTML]{99DD99}{0.50}
+                                     & \cellcolor[HTML]{99DD99}{0.50}
+                                     & \cellcolor[HTML]{99FF99}{0.00}\\
+                                    & \cellcolor[HTML]{99EE99}\textbf{2.1}
+                                    & \cellcolor[HTML]{99EE99}\textbf{2.1}
+                                    & \cellcolor[HTML]{99FE99}\textbf{1.5}
+                                    & \cellcolor[HTML]{99F499}\textbf{1.9}
+                                    & \cellcolor[HTML]{99BB99}\textbf{4.0}
+                                    & \cellcolor[HTML]{99FF99}\textbf{1.5}\\
 \end{tabular}
-\caption{Overall evaluation. Brighter colors indicate better performance.}
+\caption{Overall evaluation. Brighter colors and lower scores indicate better performance. Scores
+  are obtained by summing the normalized total scores ($m'$ values) of
+  each criterion in Table~\ref{table:evaluation} and the colors are obtained by normalizing these scores
+  as done in Table~\ref{table:evaluation}. }
 \label{table:overall}
@@
 
-Limitations: 
-\begin{itemize}
-\item This ignores pre-existing work (e.g. a workflow engine is already available as a service). Migration across architectures too.
-\item The difficulty of implementing interaction g is not properly
-reflected. For instance, fsl uses bash, which would be very difficult to convert.
-\item We are well aware that all the
-components and interactions may not require the same level of
-development effort depending on the particular targeted platform.
-\end{itemize}
-
-These architectures can be combined (give examples).
+\subsection{Limitations}
+
+A few limitations remain with our evaluation. First of all, the
+evaluation is done at a quite high-level of abstraction that may be a
+bit artificial when applied to real systems. Abstracting architectures
+from their implementation in specific systems is fundamental to the
+progress of software engineering of science gateways, but we also
+realize that this comes at the cost of realism.
+
+In particular, the presented architectures are pure patterns that may
+be blended in actual systems. The distinction between tight
+integration and service invocation, for instance, may not be that
+clear in practice. Service invocation may also be combined with
+sub-tasking in some cases.
+
+Moreover, all the interactions involved in the architectures were
+treated equally, as if they all required similar amounts of
+development effort, which is realistically not the case. For example,
+interaction \texttt{g} (workflow language conversion) clearly requires
+more effort than interaction \texttt{d} (service invocation), and may
+create much more reliability issues as well. However, quantifying the
+robustness and amount of development associated with each interaction
+and software component cannot realistically be done without entering
+into the details of a particular system.
+
+The particular case of interaction \texttt{g} (workflow language
+conversion) should be further discussed since it is very likely that
+this interaction could not be easily generalized to any workflow
+language. For instance, converting FSL, PSOM or Nipype pipelines to
+any other workflow language is very tricky because these engines rely
+on general-purpose programming languages (bash, Octave, Python, etc)
+which are much richer than scientific workflow languages.
+
+The abstract nested workflows and workflow import patterns were
+instantiated with service invocation so that they can be analyzed in
+the same framework as the other architectures. Other types of
+instantiation, for instance nested workflows with sub-tasking, could
+also be envisaged. We chose to limit ourselves to instantiations with
+service invocation because the resulting architectures are implemented in 
+real systems, and because service invocation is a largely
+used architecture. Nested workflows with sub-tasking, in particular,
+could be an interesting architecture to explore.
+
+Finally, there is no need to say that the particular technical or
+historical context of a science gateway project may influence the
+choice of an architecture to integrate workflow engines. For instance,
+some workflow engines may be already available as web services, which
+would tend to favor service invocation, and other science gateways may
+have strongly tested task and data control features (interactions
+\texttt{b} and \texttt{c}), which would favor sub-tasking. The
+migration cost between architectures is totally ignored as well.
 

### 2016-04-19 paper.tex
Wrote section 2.1

@@
 high-level workflow systems to help developers access distributed
-infrastructures such as grids and web services, resulting in tools
-such as Askalon, Hyperflow, MOTEUR, Pegasus, Swift, Taverna, Triana,
-VizTrails, WS-PGRADE/gUSE, etc. Such workflow engines usually describe
-applications in a specific language that offers operators for parallel
-computing, visual description and edition, links with domain-specific
-tool repositories, etc. Descriptions and experiments conducted with
-e-Science workflow engines were published in journals such as Future
-Generation Computer Systems and conference venues such as
-WORKS.
+infrastructures such as clusters, grids, clouds and web services,
+resulting in tools such as Askalon~\cite{fahringer2005askalon},
+Hyperflow~\cite{balis2016hyperflow}, MOTEUR~\cite{GLAT-08i},
+Pegasus~\cite{deelman2005pegasus,Deelman201517},
+Swift~\cite{zhao2007swift}, Taverna~\cite{oinn2004taverna},
+Triana~\cite{taylor2007triana}, VisTrails~\cite{callahan2006managing},
+WS-PGRADE/gUSE~\cite{Kacsuk2012}, etc. Such workflow engines usually
+describe applications in a specific language that offers parallel data
+and control flow constructs, visual edition features, links with
+domain-specific tool repositories, provenance recording, etc.
+% Descriptions and experiments
+%conducted with e-Science workflow engines were published in journals
+%such as Future Generation Computer Systems and conference venues such
+%as WORKS.
 
 At the same time, specific toolboxes were emerging in various
-scientific domains to facilitate interactions between different
-software components. In neuroimaging for instance, tools such as
-Nipype, PSOM, SPM and FSL provide abstractions and functions to define
-processes that handle the data flow between other processes. Soon,
-such tools were interfaced to computing systems, in particular
-clusters: they were extended to create cluster tasks, handle their
-dependencies, and execute them on clusters. For instance, FSL can
-launch tasks on SGE through its \texttt{fsl\_sub} tool, Nipype
-\todo{...}, PSOM \todo{...}, and SPM \todo{...}. Although distributed
-infrastructures are hardly handled, such domain-specific tools have to
-be called workflow engines. They represent a tremendous opportunity
-for science gateways to leverage existing tools and
-applications. Whether these should be combined with e-Science workflow
-engines is an open question.
-
-A workflow engine is defined as a software that submits jobs to a
-cluster, grid or cloud. Workflows may be expressed in any language,
-including scripts. Several workflow engines may be combined in the
-same science gateway. Workflow: a process that submits tasks to the
-infrastructure. FSL tools, say Feat, can be considered as workflows
-when they are configured to submit tasks to clusters using fslsub. A
-workflow consists of activities. We are talking only about concrete
-workflows with a configuration (see terstyansky et al 2014 "enabling
-scientific workflow sharing...")
-
-  Workflow engine: executes workflows, i.e. process their dependencies
-  and create computing tasks. Might transfer data too. A workflow
-  engine is not supposed to be aware of the science gateway.
+scientific domains to facilitate the assembly of software components
+in consistent pipelines. For instance, in neuroimaging, our primary
+domain of interest, tools such as Nipype~\cite{gorgolewski2011nipype},
+PSOM~\cite{bellec2012pipeline}, SPM~\cite{ashburner2011spm} and
+FSL~\cite{Jenkinson2012782} provide abstractions and functions to
+handle the data and computing flow between processes implemented in a
+variety of programming languages. Such tools were interfaced to
+computing systems, in particular clusters, to create tasks, handle
+their dependencies, and execute them. For instance, FSL can launch
+tasks on SGE through its \texttt{fsl\_sub} tool, Nipype has execution
+plugins for S/OGE, Torque, IPython, ssh and multi-processor servers,
+and PSOM \todo{...}. Some of these tools also support advanced
+features such as provenance tracking (Nipype, PSOM), redundancy
+detection across analyses to avoid re-computation (PSOM), and so on. A
+wide-array of workflows have been implemented using these pipelines
+and are now widely shared across neuroimaging groups
+world-wide. \todo{Provide more context on that?} These domain-specific
+engines represent a tremendous opportunity for science gateways to
+leverage existing tools and applications. They nicely complement
+e-Science engines that are more oriented towards the exploitation of
+distributed computing infrastructures, in particular grids and clouds.
+
+We are looking for software architectures that are able to integrate
+e-Science and domain-specific workflow engines in science
+gateways. For the sake of the analysis conducted in this paper, we
+define a \emph{workflow engine} (also abbreviated \emph{engine}) as a
+software that submits interdependent computing \emph{tasks} to an
+infrastructure (local server, cluster, grid or cloud) based on a
+workflow description (a.k.a \emph{workflow}) and input data that may
+be files, data base entries, or simple parameter values. Although
+simplistic, this definition covers both e-Science and domain-specific
+engines. Some workflow engines, usually e-Science ones, may transfer
+data across the infrastructure, and others, usually domain-specific
+ones, may leave it to external processes. Workflows may be expressed
+in any language, including high-level XML or JSON dialects such as
+\texttt{Scufl} or \texttt{Hyperflow}, and low-level scripts such as
+\texttt{bash}.  \todo{Mention other workflow definitions,
+  e.g. abstract/concrete, etc?}
 
@@
 
+Several workflow engines may be combined in the same
+science gateway.
+
+A workflow
+  engine is not supposed to be aware of the science gateway.
+
 \subsection{Infrastructures}
-
-\todo{Maybe this is not relevant}
 
@@
   etc. Workflow control can be coarse-grained (a.k.a. black box) or
-  fine-grained (white box). In a coarse-grained model, workflow
-  activities are masked and the user only has a global view of the
-  workflow execution. In a fine-grained model, user is exposed to the
-  workflow topology, i.e. to the outputs of the individual tasks,
-  their statuses and so on. 
+  fine-grained (white box). In a coarse-grained model, the various
+  tasks created by a workflow execution are masked and the user only
+  has a global view of the workflow execution. In a fine-grained
+  model, user is exposed to the workflow topology, i.e. to the outputs
+  of the individual tasks, their statuses and so on.
 \item Sub-task control: operations used by tasks to submit sub-tasks
@@
 either as a black box that completely masks the infrastructure and
-workflow activities, or as a white box that allows for some
-interaction with them. The workflow engine is responsible for
-controlling the tasks on the infrastructure (\texttt{b}) and for
-performing the required data transfers to execute them
-(\texttt{c$_2$}). User data is usually managed through the science
-gateway (\texttt{c$_1$}), although it might as well be delivered by
-the workflow engine directly to the user.
+tasks, or as a white box that allows for some interaction with
+them. The workflow engine is responsible for controlling the tasks on
+the infrastructure (\texttt{b}) and for performing the required data
+transfers to execute them (\texttt{c$_2$}). User data is usually
+managed through the science gateway (\texttt{c$_1$}), although it
+might as well be delivered by the workflow engine directly to the
+user.
 
@@
 
-See Figure~\ref{archi:nested}. In nested workflows (Figure~\ref{archi:nested}-Left), an activity of a
-\emph{parent} workflow executed by the \emph{parent} engine is itself
-a workflow, called \emph{child} workflow, that is executed by a \emph{child} engine. The
+See Figure~\ref{archi:nested}. In nested workflows
+(Figure~\ref{archi:nested}-Left), a task of a \emph{parent} workflow
+executed by the \emph{parent} engine is itself a workflow, called
+\emph{child} workflow, that is executed by a \emph{child} engine. The
 parent and child engines might use different workflow description
@@
 also used implicitly in several platforms where workflow engines are
-wrapped in workflow activities as any other command-line tool. Nested
+wrapped in workflow tasks as any other command-line tool. Nested
 workflows were notably used by the SHIWA Science Gateway to implement
@@
 \item Fine-grained debugging (\texttt{O$_2$}): availability of
-  fine-grained debugging information about workflow activities. When
+  fine-grained debugging information about workflow tasks. When
   such information is available, workflow is seen as a white
-  box. Fine-grained information about workflow activity is required to
+  box. Fine-grained information about workflow tasks is required to
   properly troubleshoot workflow executions.  \texttt{O$_2$}=0 when
@@
 (\texttt{O$_1$}=1).  Obtaining fine-grained information about workflow
-activities is not straightforward since the science gateway has no
+tasks is not straightforward since the science gateway has no
 knowledge about the workflow topology, and the workflow engine is

### 2016-04-20 paper.tex
updates in section 2.1 and 2.2

@@
 plugins for S/OGE, Torque, IPython, ssh and multi-processor servers,
-and PSOM \todo{...}. Some of these tools also support advanced
-features such as provenance tracking (Nipype, PSOM), redundancy
-detection across analyses to avoid re-computation (PSOM), and so on. A
-wide-array of workflows have been implemented using these pipelines
-and are now widely shared across neuroimaging groups
-world-wide. \todo{Provide more context on that?} These domain-specific
-engines represent a tremendous opportunity for science gateways to
-leverage existing tools and applications. They nicely complement
-e-Science engines that are more oriented towards the exploitation of
-distributed computing infrastructures, in particular grids and clouds.
+and PSOM is interfaced to Torque, SGE, MOAB, LSF and Condor. Some of
+these tools also support advanced features such as provenance tracking
+(Nipype, PSOM), redundancy detection across analyses to avoid
+re-computation (PSOM), and so on. A wide-array of workflows have been
+implemented using these pipelines and are now widely shared across
+neuroimaging groups world-wide. \todo{Provide more context on that?}
+These domain-specific engines represent a tremendous opportunity for
+science gateways to leverage existing tools and applications. They
+nicely complement e-Science engines that are more oriented towards the
+exploitation of distributed computing infrastructures, in particular
+grids and clouds.
 
@@
 
-Science Gateway: An ecosystem, usually accessible through a web
-  portal, that provides tools to access distributed
-  infrastructures. Tools usually help users manage data transfers,
-  task execution and authentication on multiple computing and storage
-  locations. Examples: CBRAIN, VIP, NSG, etc
+Science gateways are ecosystems of software tools and services to
+access distributed infrastructures, usually through a web interface.
+Tools usually help users manage data transfers, task execution and
+authentication on multiple computing and storage locations. Workflow
+engines are of course part of this ecosystem since they are useful
+systems to implement and execute applications.
+
+Several models
+
+Examples: CBRAIN, VIP, NSG, etc
 

### 2016-04-20 paper.tex
Added some text in section 2.2. Sections 1 and 2 might need to be merged.

@@
 
-\todo{Say that the review is done based on our experience with VIP,
-  CBRAIN, and to some extent SHIWA.}
+\todo{Section still to be written}
 
@@
 
+\todo{This section still needs to be finished}
+
 \subsection{Workflow engines}
@@
 In the last decade, the e-Science workflow community has developed
-high-level workflow systems to help developers access distributed
-infrastructures such as clusters, grids, clouds and web services,
-resulting in tools such as Askalon~\cite{fahringer2005askalon},
+workflow systems to help developers access distributed infrastructures
+such as clusters, grids, clouds and web services, resulting in tools
+such as Askalon~\cite{fahringer2005askalon},
 Hyperflow~\cite{balis2016hyperflow}, MOTEUR~\cite{GLAT-08i},
@@
 Triana~\cite{taylor2007triana}, VisTrails~\cite{callahan2006managing},
-WS-PGRADE/gUSE~\cite{Kacsuk2012}, etc. Such workflow engines usually
-describe applications in a specific language that offers parallel data
-and control flow constructs, visual edition features, links with
-domain-specific tool repositories, provenance recording, etc.
+WS-PGRADE/gUSE~\cite{Kacsuk2012}\footnote{\todo{WS-PGRADE/gUSE is the
+    science gateway, check what the engine is called now}}, etc. Such
+workflow engines usually describe applications in a  language
+that offers parallel data and control flow constructs, visual edition
+features, links with domain-specific tool repositories, provenance
+recording, etc.
 % Descriptions and experiments
@@
 scientific domains to facilitate the assembly of software components
-in consistent pipelines. For instance, in neuroimaging, our primary
-domain of interest, tools such as Nipype~\cite{gorgolewski2011nipype},
+in consistent pipelines. In neuroimaging, our primary domain of
+interest, tools such as Nipype~\cite{gorgolewski2011nipype},
 PSOM~\cite{bellec2012pipeline}, SPM~\cite{ashburner2011spm} and
@@
 infrastructure (local server, cluster, grid or cloud) based on a
-workflow description (a.k.a \emph{workflow}) and input data that may
-be files, data base entries, or simple parameter values. Although
+workflow description (a.k.a \emph{workflow}) and using input data that
+may be files, data base entries, or simple parameter values. Although
 simplistic, this definition covers both e-Science and domain-specific
@@
 \texttt{bash}.  \todo{Mention other workflow definitions,
-  e.g. abstract/concrete, etc?}
+  e.g. abstract/concrete, etc?} \todo{Use ttt everywhere for technical
+  terms (ssh, etc)?}
 
@@
 Science gateways are ecosystems of software tools and services to
-access distributed infrastructures, usually through a web interface.
-Tools usually help users manage data transfers, task execution and
-authentication on multiple computing and storage locations. Workflow
-engines are of course part of this ecosystem since they are useful
-systems to implement and execute applications.
-
-Several models
-
-Examples: CBRAIN, VIP, NSG, etc
-
-Multi-engine, multi-language.
-
-Several workflow engines may be combined in the same
-science gateway.
-
-A workflow
-  engine is not supposed to be aware of the science gateway.
+access data and tools deployed on distributed infrastructures. They
+are used to share resources in a community and to provide performance
+through facilitated access to storage and computing power. They are
+usually accessible through a web interface that helps users manage
+access rights, data transfers, task execution and authentication on
+multiple computing and storage locations. Workflow engines are of
+course part of this ecosystem as core components to implement and
+execute applications.
+
+% Software properties: integration effort, extensibility, robustness,
+% scalability, specific features
+
+Integration between workflow engines and science gateways varies
+across systems. Some science gateways may be tailored to a particular
+workflow engine that manages critical functions such as task
+execution, data transfers and authentication. Other gateways may be
+more general and host applications executed by different types of
+engines. Extensibility is an important property of such integration:
+new workflows are added frequently, different types and versions of
+workflow engines may be integrated over time, and different kinds of
+infrastructure may be covered.
+
+Scalability is also an important concern, since science gateways are
+multi-user systems, manage substantial amounts of data and execute
+computationally-intensive analyses. For this purpose, science gateways
+may include different instances of the same engine type and implement
+some load-balancing policy among them. They may also enforce
+scheduling policies on the infrastructure, e.g. to ensure fairness
+among users~\cite{FERR-14} or to improve
+fault-tolerance~\cite{FERR-13}. \todo{Remove these self-citations or
+  cite more works on this topic.}
+
+Robustness is a key part of science gateway user experience. In this
+respect, the multiplicity of software components and functions is a
+challenge as it often leads to complex architectures that are
+therefore prone to failure. To offer good user experience, science
+gateways should remain as transparent as possible, which requires
+robustness.
+
+Specific features may also be available in science
+gateways, such as data vizualisation and quality control, workflow
+edition, debugging instruments, social tools among users, etc.
+
+Various science gateways have been developed in the last years,
+including science gateway frameworks used as building blocks to
+assemble science gateways. The main frameworks are
+WS-PGRADE/gUSE~\cite{Kacsuk2012}, the Catania Science Gateway
+Framework~\cite{Ardizzone2012} and Apache
+Airavata~\cite{marru2011apache} \todo{According to
+  https://airavata.apache.org, Genapp is based on Airavata, we could
+  refer to it somewhere}. Numerous science gateway instances were
+built using such frameworks, in various disciplines. Other gateway
+instances were also built independently from any framework, using
+regular web development tools and various libraries to manage
+distributed systems.
+
+This paper builds on experience from the development and sustained
+operations of CBRAIN (2008-now), the Virtual Imaging Platform
+(2009-now), and on our involvement with the
+SHIWA\footnote{\url{http://www.shiwa-workflow.eu}} European project
+(2010-2012) that built the SHIWA Simulation Platform.
+
+CBRAIN\footnote{\url{http://cbrain.mcgill.ca}}~\cite{SHER-14} is an
+open-source, web-based collaborative platform addressing the
+challenges of data-heavy, computation-intensive neuroimaging
+research. It offers transparent web-based access to remote data
+sources, distributed computing sites, and an array of processing and
+visualization tools, within a secure environment. CBRAIN is deployed
+at a wide range of computing sites, including nine national HPC
+centers in Canada, one in Korea, one in Germany, and on numerous local
+servers. As September 2015, CBRAIN has served over 370 users across 50
+cities in 19 countries who produced several high-impact results in the
+field of neurosciences.
+
+The Virtual Imaging Platform
+(VIP\footnote{\url{http://www.creatis.insa-lyon.fr/vip}})~\cite{GLAT-13}
+is a web portal for medical simulation and image data analysis. It
+leverages computing and storage resources available in the biomed
+Virtual Organization of the European Grid Infrastructure (EGI) to
+offer an open service to academic researchers worldwide. VIP
+completely masks the infrastructure to enable a transparent user
+experience. It uses the MOTEUR workflow engine~\cite{GLAT-13} and the
+DIRAC pilot-job framework to execute applications on EGI. Almost 900
+users were registered in VIP in April 2016 and the average CPU
+consumption between January 2013 and March 2016 was 35 CPU years per
+month.
+
+\todo{SHIWA Simulation Platform.}
 
@@
 workflow execution fails due to errors in the components or with the
-interactions in the software architecture. 
-Robustness is determined as the number of software components and
-interactions that are specific to workflow execution, i.e. not reused
-by other functions:
+interactions in the software architecture. Note that errors coming
+from the infrastructure (e.g. unavailable data or terminated tasks) or
+workflows (e.g., wrong user input or application crashes) are not
+covered since they do not stem directly from the software
+architecture. Robustness is determined as the number of software
+components and interactions that are specific to workflow execution,
+i.e. not reused by other functions:
 \begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
@@
 
+\todo{Section still to be written}
+
 We abstracted architectures so that we can compare them using
@@
 
+\todo{Section still to be written}
+
 \todo{FLI-IAM, Labex PRIMES, Ludmer Centre, whatever grant is funding

### 2016-04-20 paper.tex
Changed labels to Roman numbers on subfigures of Figure 2

@@
 \usepackage[x11names,dvipsnames,table]{xcolor} %for use in color links
-\usepackage{subfig}
+\usepackage[subrefformat=parens,labelformat=parens]{subfig}
 \usepackage{svg}
@@
 \newcommand{\note}[1]{\color{green}\textbf{Note: }\textit{#1} \color{black}}
+
+\renewcommand\thesubfigure{\Roman{subfigure}}
 
@@
 %% Text of abstract
-
+  We investigate software architectures to integrate workflow engines
+  in science gateways. Based on our experience with the development
+  and operations of the CBRAIN and VIP science gateways, and our past
+  involvement in the SHIWA project, we abstract the software
+  components and interactions in 6 architectures that we compare using
+  metrics measuring integration effort, robustness, extensibility,
+  scalability and other specific features. \todo{a word on results}
 \end{abstract}
@@
 
-See Figure~\ref{archi:tight}. The workflow engine is tightly
+See Figure~\subref*{archi:tight}. The workflow engine is tightly
 integrated with the science gateway, which means that it is deployed
@@
 
-See Figure~\ref{archi:service}. The workflow engine is available
+See Figure~\subref*{archi:service}. The workflow engine is available
 externally to the science gateway, in a service. The science gateway
@@
 
-See Figure~\ref{archi:sub-task}. The workflow engine is a particular
+See Figure~\subref*{archi:sub-task}. The workflow engine is a particular
 task that can submit sub-tasks to the science gateway through
@@
 \label{sec:pool}
-See Figure~\ref{archi:agent}. Workflows are submitted by the science
+See Figure~\subref*{archi:agent}. Workflows are submitted by the science
 gateway to a pool through interaction \texttt{d}. Agents connect to
@@
 ~\ref{fig:shiwa-pool-architecture}, interaction \texttt{d} of
-Figure~\ref{archi:agent} is implemented in 3 different types of calls
+Figure~\subref*{archi:agent} is implemented in 3 different types of calls
 to the pool: workflow submission (\texttt{1} \& \texttt{2}), workflow
@@
 ~\ref{fig:shiwa-pool-architecture}, interaction \texttt{f} of
-Figure~\ref{archi:agent} is implemented through 2 types of calls:
+Figure~\subref*{archi:agent} is implemented through 2 types of calls:
 workflow instance retrieval (\texttt{4}, \texttt{5} and \texttt{6}),
@@
 component and therefore have no corresponding representation in
-Figure~\ref{archi:agent}.
+Figure~\subref*{archi:agent}.
 
@@
 
-See Figure~\ref{archi:nested}. In nested workflows
-(Figure~\ref{archi:nested}-Left), a task of a \emph{parent} workflow
+See Figure~\subref*{archi:nested}. In nested workflows
+(Figure~\subref*{archi:nested}-Left), a task of a \emph{parent} workflow
 executed by the \emph{parent} engine is itself a workflow, called
@@
 focus on instantiation with the service invocation model (see
-Figure~\ref{archi:nested}-Right) as this is the most used
+Figure~\subref*{archi:nested}-Right) as this is the most used
 architecture. In such an instantiation, we assume that the parent and
@@
 used in the SHIWA Science Gateway for nested workflow execution with
-service invocation as represented in Figure~\ref{archi:nested}. The
+service invocation as represented in Figure~\subref*{archi:nested}. The
 parent workflow engine is \texttt{WS-PGRADE}, invoked as a service in
@@
 Figure~\ref{fig:shiwa-architecture}, interaction \texttt{d} on
-Figure~\ref{archi:nested}). Ten different child engines can be used by
+Figure~\subref*{archi:nested}). Ten different child engines can be used by
 nested workflows, invoked through the \texttt{Submission service}
@@
 
-See Figure~\ref{archi:import}. This is an abstract model that we
+See Figure~\subref*{archi:import}. This is an abstract model that we
 instantiate with the service invocation architecture for

### 2016-04-20 paper.tex
Section 1 is now readable, although not final. Section 2 was merged into Section 1.

@@
 
-\title{Architectures for workflow integration in science gateways}
+\title{Software architectures for workflow integration in science gateways}
 
@@
 \begin{keyword}
-%% keywords here, in the form: keyword \sep keyword
-
-%% PACS codes here, in the form: \PACS code \sep code
-
-%% MSC codes here, in the form: \MSC code \sep code
-%% or \MSC[2008] code \sep code (2000 is the default)
-
+Workflow engines \sep science gateways \sep software architecture.
 \end{keyword}
@@
 
-\todo{Section still to be written}
-
-\paragraph{Context} The question of integrating workflow engines in
-science gateways can be seen at various levels, corresponding to
-various definitions of workflows. One level is the SHIWA level, where
-it was considered that workflow engines are aware of the DCI (and 'D'
-is important because of data transfers, proxies, etc). Another level
-is to remove 'D' from the definition: a workflow engine becomes a
-program that submits jobs, potentially only to local clusters. It
-opens a whole new class of workflow engines that we used to consider
-as "applications". For instance, in neuroinformatics: Nipype, PSOM,
-but also FSL through the fslsub tool. A workflow engine is not
-supposed to be aware of the science gateway.  A wide-array of workflow
-engines are available with specificies coming from the application
-domain, available tools, etc Their integration in science gateways
-becomes critical. Several of the examples presented in the
-paper will be taken from medical image analysis, in particular
-neuroimaging.
-
-\paragraph{Goal} this paper reviews and compares the architectures to
-integrate workflow engines in science gateways.
-
-\paragraph{Contributions}
-\begin{itemize}
-\item We evaluate architectures based on our experience with existing systems
-\item We propose a new architecture
-\end{itemize}
-
-\section{Workflow engines and science gateways}
-
-\todo{This section still needs to be finished}
+Workflow engines are critical components of the ecosystem of tools and
+services offered by science gateways to leverage distributed
+infrastructures for the efficient execution of applications through
+the web. Several software architectures can be adopted to integrate
+workflow engines in science gateways, with important consequences on
+the required development effort and on the characteristics of the
+resulting system. This paper describes and compares such architectures
+based on system-independent representations of their main components
+and interactions.
+
+Our analysis is fed by experience gathered from the development and
+sustained operation of the CBRAIN~\cite{SHER-14} and
+VIP~\cite{GLAT-13} science gateways for medical image analysis during
+the past 7 years. It also builds on the lessons learned from several
+science gateway and workflow-related projects, including the SHIWA
+Simulation Platform~\cite{terstyanszky2014enabling}.
+
+The remainder of this Section provides background information and
+definitions about workflow engines and science gateways. In
+Section~\ref{sec:architectures}, we describe 6 architectures in a
+consistent framework based on the functional interactions between the
+main software components involved. In Section~\ref{sec:evaluation}, we
+introduce metrics to measure integration effort, robustness,
+extensibility, scalability and other specific features related to the
+architectures. We use these metrics to evaluate and compare the 6
+architectures. The paper closes on a discussion and conclusion that
+highlight architectural recommendations for science gateway and
+workflow engine developers.
 
@@
 workflow systems to help developers access distributed infrastructures
-such as clusters, grids, clouds and web services, resulting in tools
-such as Askalon~\cite{fahringer2005askalon},
+such as clusters, grids, clouds and web services. These efforts
+resulted in tools such as Askalon~\cite{fahringer2005askalon},
 Hyperflow~\cite{balis2016hyperflow}, MOTEUR~\cite{GLAT-08i},
@@
     science gateway, check what the engine is called now}}, etc. Such
-workflow engines usually describe applications in a  language
+workflow engines usually describe applications in a specific language
 that offers parallel data and control flow constructs, visual edition
 features, links with domain-specific tool repositories, provenance
-recording, etc.
+recording, among other features.
 % Descriptions and experiments
@@
 
-We are looking for software architectures that are able to integrate
-e-Science and domain-specific workflow engines in science
-gateways. For the sake of the analysis conducted in this paper, we
+For the sake of the analysis conducted in this paper, we
 define a \emph{workflow engine} (also abbreviated \emph{engine}) as a
@@
 
+\todo{The infrastructure consists of the computing and storage resources
+involved in the workflow execution, and the software services used to
+access these resources. Infrastructures can be servers, clusters,
+grids or clouds. Some workflow engines may assume specific
+characteristics about the infrastructure, such as the presence of a
+shared file system between the computing nodes.
+}
+
 \subsection{Science gateways}
 
-Science gateways are ecosystems of software tools and services to
-access data and tools deployed on distributed infrastructures. They
+Science gateways 
 are used to share resources in a community and to provide performance
@@
 
-This paper builds on experience from the development and sustained
-operations of CBRAIN (2008-now), the Virtual Imaging Platform
-(2009-now), and on our involvement with the
-SHIWA\footnote{\url{http://www.shiwa-workflow.eu}} European project
-(2010-2012) that built the SHIWA Simulation Platform.
-
+\todo{Transition needed}
 CBRAIN\footnote{\url{http://cbrain.mcgill.ca}}~\cite{SHER-14} is an
@@
 \todo{SHIWA Simulation Platform.}
-
-\subsection{Infrastructures}
-
-The infrastructure consists of the computing and storage resources
-involved in the workflow execution, and the software services used to
-access these resources. Infrastructures can be servers, clusters,
-grids or clouds. Some workflow engines may assume specific
-characteristics about the infrastructure, such as the presence of a
-shared file system between the computing nodes.
 
@@
 \section{Evaluation}
+
+\label{sec:evaluation}
 

### 2016-04-20 clean-pdf-history.sh
A script to clean the repo history from pdf versions.

@@ -1,0 +1,6 @@
+#!/bin/bash
+
+# Clean the repo history from pdf versions to reduce its size (former pdf versions can still be built from source)
+
+java -jar $BFG_HOME/bfg-1.12.12.jar --delete-files paper.pdf
+git reflog expire --expire=now --all && git gc --prune=now --aggressive

### 2016-04-20 paper.tex
First note commit -test-

@@
   infrastructure might offer various data storage backends with
-  heterogeneous interfaces. Tools and services such as
-  JSAGA~\cite{reynaud2010uniform} or Data Avenue~\cite{hajnal2014data}
-  can be used to homogeneize these interfaces.
+  heterogeneous interfaces. \note{this -c2- can be also trigerred by the 
+  user} Tools and services such as JSAGA~\cite{reynaud2010uniform} or
+  Data Avenue~\cite{hajnal2014data} can be used to homogeneize these 
+  interfaces.
 \item Workflow control: operations to execute a workflow with an

### 2016-04-20 paper.tex
Added a conclusion

--- 2016-04-20 paper.tex
@@
                                     & \cellcolor[HTML]{99EE99}\textbf{2.1}
-                                    & \cellcolor[HTML]{99FE99}\textbf{1.5}
+                                    & \cellcolor[HTML]{99FF99}\textbf{1.5}
                                     & \cellcolor[HTML]{99F499}\textbf{1.9}
@@
 
-\todo{Section still to be written}
-
-We abstracted architectures so that we can compare them using
-objective metrics. We illustrated these architectural patterns on
-examples of real systems.
-
-What this work highlights and recommendations for science gateway
-developers.
+% Summary of contributions
+We reviewed architectures used to integration workflow engines in
+science gateways, we described them in a system-independent framework
+suitable for comparison, we illustrated them on real systems, we
+developed quantitative metrics for evaluation, we evaluated these
+metrics for all the architectures, and we discussed the pros and cons
+of all the architectures based on these metrics.
+
+% Impact of the work
+To the best of our knowledge, our work is the first of its kind. Our
+results can be used as recommendation for science gateway and workflow
+engine developers to build new platforms, or to upgrade existing ones.
 
@@
 
-\todo{Section still to be written}
-
-\todo{FLI-IAM, Labex PRIMES, Ludmer Centre, whatever grant is funding
-  POQ for integrating CBRAIN with PSOM.}
-
-We warmly thank Rafael Ferreira da Silva for implementing the VIP
-platform and creating Figure~\ref{fig:vip-architecture}.
+We warmly thank Rafael Ferreira da Silva for its contribution to the
+VIP platform and for creating Figure~\ref{fig:vip-architecture}. This
+work is in the scope of the LABEX PRIMES (ANR-11- LABX-0063) of
+Universit\'e de Lyon, within the program ``Investissements d’Avenir''
+(ANR-11-IDEX-0007) operated by the French National Research Agency
+(ANR). This work has been made possible with the support of the Irving
+Ludmer Family Foundation and the Ludmer Centre for Neuroinformatics
+and Mental Health.
 

### 2016-04-21 paper.tex
Changed color of notes
Updates in Section 1

@@
 
+\definecolor{ao(english)}{rgb}{0.0, 0.5, 0.0}
 \newcommand{\todo}[1]{\color{blue}\xspace[\emph{#1}]\xspace\color{black}}
-\newcommand{\note}[1]{\color{green}\textbf{Note: }\textit{#1} \color{black}}
+\newcommand{\note}[1]{\color{ao(english)}\textbf{Note: }\textit{#1} \color{black}}
 
@@
   metrics measuring integration effort, robustness, extensibility,
-  scalability and other specific features. \todo{a word on results}
+  scalability and other specific features. Results show the pros and
+  cons of the different architectures. Tight integration and
+  sub-tasking require the least amount of integration effort and are
+  also the most robust. Most architectures perform the same in terms
+  of extensibility. The pool model is the most
+  scalable. Meta-workflows are only available in nested workflows and
+  workflow import.
 \end{abstract}
@@
 The remainder of this Section provides background information and
-definitions about workflow engines and science gateways. In
-Section~\ref{sec:architectures}, we describe 6 architectures in a
-consistent framework based on the functional interactions between the
-main software components involved. In Section~\ref{sec:evaluation}, we
-introduce metrics to measure integration effort, robustness,
-extensibility, scalability and other specific features related to the
-architectures. We use these metrics to evaluate and compare the 6
-architectures. The paper closes on a discussion and conclusion that
-highlight architectural recommendations for science gateway and
-workflow engine developers.
+definitions about workflow engines, science gateways and
+infrastructures. In Section~\ref{sec:architectures}, we describe 6
+architectures in a consistent framework based on the functional
+interactions between the main software components involved. In
+Section~\ref{sec:evaluation}, we introduce metrics to measure
+integration effort, robustness, extensibility, scalability and other
+specific features related to the architectures. We use these metrics
+to evaluate and compare the 6 architectures. The paper closes on a
+discussion and conclusion that highlight architectural recommendations
+for science gateway and workflow engine developers.
 
@@
 Triana~\cite{taylor2007triana}, VisTrails~\cite{callahan2006managing},
-WS-PGRADE/gUSE~\cite{Kacsuk2012}\footnote{\todo{WS-PGRADE/gUSE is the
-    science gateway, check what the engine is called now}}, etc. Such
+WS-PGRADE~\cite{Kacsuk2012}, etc. Such
 workflow engines usually describe applications in a specific language
@@
 their dependencies, and execute them. For instance, FSL can launch
-tasks on SGE through its \texttt{fsl\_sub} tool, Nipype has execution
+tasks on SGE through its fsl\_sub tool, Nipype has execution
 plugins for S/OGE, Torque, IPython, ssh and multi-processor servers,
@@
 in any language, including high-level XML or JSON dialects such as
-\texttt{Scufl} or \texttt{Hyperflow}, and low-level scripts such as
-\texttt{bash}.  \todo{Mention other workflow definitions,
-  e.g. abstract/concrete, etc?} \todo{Use ttt everywhere for technical
-  terms (ssh, etc)?}
-
-\todo{The infrastructure consists of the computing and storage resources
-involved in the workflow execution, and the software services used to
-access these resources. Infrastructures can be servers, clusters,
-grids or clouds. Some workflow engines may assume specific
-characteristics about the infrastructure, such as the presence of a
-shared file system between the computing nodes.
-}
+Scufl or Hyperflow, and low-level scripts such as
+Bash.
 
@@
 scheduling policies on the infrastructure, e.g. to ensure fairness
-among users~\cite{FERR-14} or to improve
-fault-tolerance~\cite{FERR-13}. \todo{Remove these self-citations or
-  cite more works on this topic.}
+among users~\cite{FERR-14,CPE:CPE3708} or to improve
+fault-tolerance~\cite{wilkins2008teragrid,FERR-13}.
 
@@
 Framework~\cite{Ardizzone2012} and Apache
-Airavata~\cite{marru2011apache} \todo{According to
-  https://airavata.apache.org, Genapp is based on Airavata, we could
-  refer to it somewhere}. Numerous science gateway instances were
-built using such frameworks, in various disciplines. Other gateway
-instances were also built independently from any framework, using
-regular web development tools and various libraries to manage
-distributed systems.
-
-\todo{Transition needed}
+Airavata~\cite{marru2011apache}. Numerous science gateway instances
+were built using such frameworks, in various disciplines. Other
+gateway instances were also built independently from any framework,
+using regular web development tools and various libraries to manage
+distributed systems. CBRAIN and VIP are two of such science gateways
+that were initially developed for medical imaging analysis and from
+which we gathered experience about the architectures described in
+this paper.
+
 CBRAIN\footnote{\url{http://cbrain.mcgill.ca}}~\cite{SHER-14} is an
@@
 
-\todo{SHIWA Simulation Platform.}
+The architectures described hereafter also build on experience with
+the SHIWA Simulation Platform~\cite{terstyanszky2014enabling}. The
+SHIWA Simulation Platform is a WS-PGRADE/gUSE science gateway that was
+developed in European projects
+SHIWA\footnote{\url{http://www.shiwa-workflow.eu}} and
+ER-flow\footnote{\url{http://www.erflow.eu}}. It focuses on workflow
+interoperability for various disciplines, in particular through the
+building of meta-workflows, i.e. workflows that are built out of other
+workflows. It implements the so-called Coarse-Grained Interoperability
+framework where workflow engines are integrated in the science gateway
+as other applications and invoked through a dedicated service. The
+Fine-Grained Interoperability framework was also developed in SHIWA,
+where conversions between workflow languages are achieved through a
+common intermediate
+representation~\cite{plankensteiner-prodan-etal:2013}.
+
+\subsection{Infrastructures}
+
+In this paper, infrastructure consists of the computing and storage
+resources involved in the workflow execution, and the software
+services used to access these resources. Infrastructures can be
+servers, clusters, grids or clouds. Some workflow engines and science
+gateways may assume specific characteristics about the infrastructure,
+such as the presence of a shared file system between the computing
+nodes, the availability of a global task meta-scheduler, the presence
+of a file catalog, etc. In the analysis presented below, such specific
+functions are not discussed. Instead, we consider the infrastructure
+as an abstract system that can execute tasks and store files
+regardless of the enabling mechanisms.
 
@@
 (Linux executables and script) that submits tasks uniformly through a
-specific tool called \texttt{fsl\_sub}. To integrate FSL in CBRAIN,
-\texttt{fsl\_sub} was modified to implement interaction
-\texttt{e}\footnote{See \texttt{fsl\_sub} modified script available at
+specific tool called fsl\_sub. To integrate FSL in CBRAIN,
+fsl\_sub was modified to implement interaction
+\texttt{e}\footnote{See modified fsl\_sub script available at
   \url{https://github.com/aces/cbrain-plugins-neuro}} \todo{Add a
@@
 any other workflow language is very tricky because these engines rely
-on general-purpose programming languages (bash, Octave, Python, etc)
+on general-purpose programming languages (Bash, Octave, Python, etc)
 which are much richer than scientific workflow languages.

### 2016-04-21 paper.tex
Updates in Section 2. Modified command to add notes.

@@
 \newcommand{\todo}[1]{\color{blue}\xspace[\emph{#1}]\xspace\color{black}}
-\newcommand{\note}[1]{\color{ao(english)}\textbf{Note: }\textit{#1} \color{black}}
+\newcommand{\note}[2]{\textbf{\Large{\color{blue}\footnote{{\color{blue}\textbf{#1:} \textit{#2}\color{black}}}}}}
 
@@
 %% use the tnoteref command within \title for footnotes;
-%% use the tnotetext command for theassociated footnote;
+%% use the tnotetext command for theassociated footnotes;
 %% use the fnref command within \author or \address for footnotes;
@@
   available on the infrastructure. When the infrastructure is a grid
-  or a cloud, it is for instance implemented using libraries such as
-  SAGA, DRMAA, OCCI and similar initiatives \todo{refs needed}.
+  or a cloud, it is for instance implemented using libraries
+  implementing standards such as SAGA~\cite{goodale2006saga},
+  DRMAA~\cite{troger2012distributed} or OCCI~\cite{edmonds2012toward}.
 \item Data control: operations to manage data on the infrastructure,
@@
   data. They can also be performed by the workflow engine
-  (\texttt{$c_2$}), to transfer data across the infrastructure. The
-  infrastructure might offer various data storage backends with
-  heterogeneous interfaces. \note{this -c2- can be also trigerred by the 
-  user} Tools and services such as JSAGA~\cite{reynaud2010uniform} or
-  Data Avenue~\cite{hajnal2014data} can be used to homogeneize these 
-  interfaces.
+  (\texttt{$c_2$}), to transfer workflow data (inputs, outputs, or
+  intermediate results) across the infrastructure so that tasks can
+  use it. The infrastructure might offer various data storage backends
+  with heterogeneous interfaces. \note{Marc-Etienne}{this -c2- can be also trigerred
+    by the user}\note{Tristan}{No, it can't: this is by definition an interaction between the workflow engine and the infrastructure.
+It is used to transfer data to/from tasks.
+I reworded a bit, is it clearer now?} Tools and services such as
+  JSAGA~\cite{reynaud2010uniform} or Data Avenue~\cite{hajnal2014data}
+  can be used to homogeneize these interfaces.
 \item Workflow control: operations to execute a workflow with an
@@
   scheduler. 9: PSOM master and workers execute workflow. 10.
-  Execution server transfers results to storage server(s). \todo{Improve graphics}}
+  Execution server transfers results to storage server(s).}
 \label{fig:cbrain-psom-architecture}

### 2016-04-21 paper.tex
Added CBRAIN-FSL architecture

@@
 common intermediate
-representation~\cite{plankensteiner-prodan-etal:2013}.
+representation~\cite{plankensteiner-prodan-etal:2013}. \note{Tristan}{Silvia,
+  do you think it is legitimate to refer to SHIWA like this in the
+  paper, or should we involve Gabor in the writing?}
 
@@
 The sub-tasking architecture is implemented in CBRAIN~\cite{SHER-14}
-where it is used to integrate the PSOM workflow
-engine~\cite{bellec2012pipeline} and the FSL toolkit. The CBRAIN-PSOM
-integration~\cite{GLAT-16} is illustrated on
-Figure~\ref{fig:cbrain-psom-architecture}. On this Figure, the science
+where it is used to integrate the FSL toolkit~\cite{Jenkinson2012782}
+and the PSOM workflow engine~\cite{bellec2012pipeline}.
+
+\begin{figure}
+\def\svgwidth{\columnwidth}
+\input{figures/CBRAIN-FSL.pdf_tex}
+\caption{Architecture of the CBRAIN-FSL integration (sub-tasking). 1:
+  User sends data and workflow execution request to storage server(s)
+  and CBRAIN portal. 2: CBRAIN portal sends execution request to
+  execution server on cluster. 3: Execution server transfers data from
+  storage server(s). 4-5: Execution server starts FSL tool (pipeline)
+  via task scheduler. 6: FSL pipeline submits sub-tasks execution
+  server through fsl\_sub. 7-8: Execution server starts sub-tasks
+  through task scheduler. 9.  Execution server transfers results to
+  storage server(s).}
+\label{fig:cbrain-fsl-architecture}
+\end{figure}
+The CBRAIN-FSL integration allows to leverage FSL pipelines that are
+written in a very low-level workflow language (Linux executables and
+script) that submits tasks uniformly through a specific tool called
+fsl\_sub. It is represented in
+Figure~\ref{fig:cbrain-fsl-architecture}.  On this Figure, the science
 gateway is represented by components \texttt{CBRAIN portal} and
@@
 manager). Interaction \texttt{c} is implemented by steps \texttt{3}
-and \texttt{10}. Interaction \texttt{e} is implemented by step
-\texttt{6}. The PSOM workflow engine adopts a pilot-job
-architecture~\cite{turilli2015comprehensive} where the a master
-coordinates workflow execution by submitting workers and establishing
-direct communication channels with them. Note how this peculiar
-execution model is totally supported by the sub-tasking architecture.
-
-The CBRAIN-FSL integration is also remarkable as it allows to leverage
-FSL pipelines that are written in a very low-level workflow language
-(Linux executables and script) that submits tasks uniformly through a
-specific tool called fsl\_sub. To integrate FSL in CBRAIN,
-fsl\_sub was modified to implement interaction
-\texttt{e}\footnote{See modified fsl\_sub script available at
-  \url{https://github.com/aces/cbrain-plugins-neuro}} \todo{Add a
-  diagram and maybe a code excerpt to illustrate that.}.
+and \texttt{9}. Interaction \texttt{e} is implemented by step
+\texttt{6} through a modified version of the fsl\_sub
+script\footnote{See modified fsl\_sub script available at
+  \url{https://github.com/aces/cbrain-plugins-neuro}}.
 
@@
 \input{figures/CBRAIN-PSOM.pdf_tex}
-\caption{Architecture of the CBRAIN-PSOM integration that illustrates
-  the sub-tasking model (Figure extracted from~\cite{GLAT-16}). 1:
-  User sends data and workflow execution request to storage server(s)
-  and CBRAIN portal. 2: CBRAIN portal sends execution request to
-  execution server on cluster. 3: Execution server transfers data from
-  storage server(s). 4-5: Execution server starts PSOM master via task
+\caption{Architecture of the CBRAIN-PSOM integration (sub-tasking,
+  Figure extracted from~\cite{GLAT-16}). 1: User sends data and
+  workflow execution request to storage server(s) and CBRAIN
+  portal. 2: CBRAIN portal sends execution request to execution server
+  on cluster. 3: Execution server transfers data from storage
+  server(s). 4-5: Execution server starts PSOM master via task
   scheduler. 6: PSOM master submits PSOM workers to execution
@@
 \end{figure}
+The CBRAIN-PSOM integration~\cite{GLAT-16} is illustrated on
+Figure~\ref{fig:cbrain-psom-architecture}. The PSOM workflow engine
+adopts a pilot-job architecture~\cite{turilli2015comprehensive} where
+the a master coordinates workflow execution by submitting workers and
+establishing direct communication channels with them. Note how this
+peculiar execution model is totally supported by the sub-tasking
+architecture.
 

### 2016-04-21 paper.tex
Added a summary table in section 2

@@
 \usepackage{multirow}
+\usepackage{pbox}
 
@@
 Triana~\cite{taylor2007triana}, VisTrails~\cite{callahan2006managing},
-WS-PGRADE~\cite{Kacsuk2012}, etc. Such
-workflow engines usually describe applications in a specific language
-that offers parallel data and control flow constructs, visual edition
-features, links with domain-specific tool repositories, provenance
-recording, among other features.
+WS-PGRADE~\cite{Kacsuk2012}, etc. Such workflow engines usually
+describe applications in a specific language that offers parallel data
+and control flow constructs, visual edition features, links with
+domain-specific tool repositories, provenance recording, among other
+features (see review in~\cite{deelman2009workflows}).
 % Descriptions and experiments
@@
 
-The architectures are diagramed in Figure~\ref{fig:architectures}
+The architectures are diagrammed in Figure~\ref{fig:architectures}
 using the graphical notations shown in
@@
 represented by a specific symbol with dotted lines. Software
-interactions among these components are represented with
-arrows. \emph{Abstract} interactions are specific types of
-interactions that may be implemented by various different software
-interactions. They are represented with dotted arrows and labeled with
- \texttt{*}. An architecture that has an abstract interaction is an
-abstract architecture. Red color is used to represent the components
-and interactions that are specific to workflow execution, i.e. that
-are not involved in any other process. Tasks refer to computing tasks
-that are created by the workflow engine and executed on the
+interactions among these components are represented with arrows.
+\emph{Abstract} interactions are specific types of interactions that
+may be implemented by various different software interactions. They
+are represented with dotted arrows and labeled with \texttt{*}. An
+architecture that has an abstract interaction is an abstract
+architecture.  Red color is used to represent the components and
+interactions that are specific to workflow execution, i.e. that are
+not involved in any other process. Tasks refer to computing tasks that
+are created by the workflow engine and executed on the
 infrastructure. Data represents any type of file, or database that is
@@
 while the workflow executes. It does not cover workflow parameters
-such as strings, numbers, etc. The remainder of this section describes
-the different types of interactions involved in the architectures, and
-details each architecture.
+such as strings, numbers, etc.
+
+The remainder of this section describes the different types of
+interactions involved in the architectures, details architectures and
+illustrates them using examples of real
+systems. Table~\ref{table:system-classification} summarizes the
+classification of a few systems by architecture.
 
@@
 and infrastructure (\texttt{a$_1$}) is done by an administrator, for
-security reasons. This is the model adopted in the Catania Science
-Gateway Framework~\cite{Ardizzone2012} (see specific documentation on
-workflows\footnote{\url{http://bit.ly/1oQrzvQ}}),
-CIPRES~\cite{miller2010creating} and LONI Pipeline
-Environment~\cite{dinov2009efficient}.
-\todo{Add description of a real system.}
+security reasons. This is the model adopted in LONI Pipeline
+Environment~\cite{dinov2009efficient} and in the Catania Science
+Gateway Framework~\cite{Ardizzone2012} which is used for instance in
+the DECIDE gateway instance for
+neuroimaging~\cite{ardizzone2012decide} (see specific documentation on
+workflows\footnote{\url{http://bit.ly/1oQrzvQ}}). \todo{Add
+  description of a real system.\note{Tristan}{The Catania Science
+    Gateway Framework would be a good illustration here (workflow
+    engine is a Java portlet), but I can't find a suitable figure
+    about it.}}
+
+%CIPRES~\cite{miller2010creating} (on which the Neuroscience gateway
+%described in~\cite{sivagnanam2015early} is based) % \note{Tristan}{There is no workflow engine in there.}
 
@@
 Toolkit~\cite{DBLP:journals/scpe/SzejnfeldDKKKKLPTWDNW10}, Virtual
-Imaging Platform~\cite{GLAT-13}, the WS-PGRADE/gUSE
+Imaging Platform~\cite{GLAT-13}, the system
+in~\cite{wu2010accelerating}, the WS-PGRADE/gUSE
 framework~\cite{Kacsuk2012} and the numerous science gateway instances
-that rely on
-WS-PGRADE/gUSE~\cite{kacsuk2014science}. Figure~\ref{fig:vip-architecture}
-shows the architecture of the Virtual Imaging Platform, where the
-MOTEUR workflow engine~\cite{GLAT-08i} is invoked as a service at step
+that rely on WS-PGRADE/gUSE, for instance
+e-BioInfra~\cite{shahand2015data}, MoSGrid~\cite{krüger2014mosgrid},
+and many others~\cite{kacsuk2014science}. The
+integration between the HubZero science gateway and Pegasus workflow
+engine performed in~\cite{CPE:CPE3257} also use a service architecture
+where the \texttt{d} interaction is implemented as a set of calls to
+the Pegasus command-line tools (e.g. pegasus-status,
+pegasus-statistics, etc). Figure~\ref{fig:vip-architecture} shows the
+architecture of the Virtual Imaging Platform, where the MOTEUR
+workflow engine~\cite{GLAT-08i} is invoked as a service at step
 \texttt{3} (interaction \texttt{d} on
@@
 platform. Figure~\ref{fig:shiwa-architecture} shows the architecture
-used in the SHIWA Science Gateway for nested workflow execution with
+used in the SHIWA Simulation Platform for nested workflow execution with
 service invocation as represented in Figure~\subref*{archi:nested}. The
@@
 format conversion from a native format to the science gateway
-format. This was implemented in the SHIWA Science Gateway through
-the IWIR language that provided a common language for portability
-across grid workflow
+format. Workflow import is usually an offline process which is not
+involved in workflow execution. Workflow import was implemented in the SHIWA
+Simulation Platform through the IWIR language that provided a common
+language for portability across grid workflow
 systems~\cite{plankensteiner-prodan-etal:2013}.
 
-\todo{workflow import is an offline process.}
 \todo{Add description of a real system}
+
+\todo{Refer to ``From the desktop to the grid: scalable bioinformatics via workflow conversion Luis de la GarzaEmail author, Johannes Veit, Andras Szolek, Marc Röttig, Stephan Aiche, Sandra Gesing, Knut Reinert and Oliver Kohlbacher''}
+
+\begin{table*}
+\centering
+\rowcolors{1}{white}{gray!25}
+\begin{tabular}{ll}
+  \textbf{Architecture} & \textbf{System} \\
+  \hline
+  Tight & \pbox{1.5\columnwidth}{
+          Catania Science Gateway Framework~\cite{Ardizzone2012}, 
+          DECIDE~\cite{ardizzone2012decide}$^{c,n}$, LONI Pipeline Environment$^n$~\cite{dinov2009efficient}
+          }
+  \\
+  Service & \pbox{1.5\columnwidth}{
+            Apache Airvata~\cite{marru2011apache}, e-BioInfra~\cite{shahand2015data}$^{w,n}$, HubZero with Pegasus~\cite{CPE:CPE3257}, MoSGrid~\cite{krüger2014mosgrid}$^w$, System in~\cite{wu2010accelerating}, Vine Toolkit~\cite{DBLP:journals/scpe/SzejnfeldDKKKKLPTWDNW10}, Virtual Imaging Platform~\cite{GLAT-13}$^n$, WS-PGRADE/gUSE framework~\cite{Kacsuk2012},
+            Science gateways in ~\cite{kacsuk2014science}$^w$
+            } \\
+  Sub-task & CBRAIN and PSOM~\cite{GLAT-16}$^n$, CBRAIN and FSL$^n$\\
+  Pool & SHIWA pool~\cite{ROGE-13}\\
+  Nested & SHIWA Simulation Platform (Coarse-Grained Interoperability~\cite{terstyanszky2014enabling})$^w$\todo{others}\\
+  Import & SHIWA Simulation Platform (Fine-Grained Interoperability)~\cite{plankensteiner-prodan-etal:2013}$^w$
+\end{tabular}
+\caption{Classification of science gateway based on the architecture used to integrate workflow engines.
+$^c$: based on the Catania Science Gateway Framework.
+$^w$: based on WS-PGRADE/gUSE.
+$^n$: used for neuroimaging.}
+\label{table:system-classification}
+\end{table*}
 
@@
 evaluate the architectures: integration effort, robustness of workflow
-execution, extensibility, scalability and other features. Criteria
+execution, extensibility, scalability and other specific features. Criteria
 break down to specific metrics where \emph{low value indicates good
@@
 
-\emph{Other features} include:
+\emph{Specific features} include:
 \begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
@@
 \end{itemize}
-\texttt{O$_1$} and \texttt{O$_2$} are summed to obtain a total number of missing other
-features in the architecture.  \todo{Maybe we could revise the
-  categories to avoid such a vague ``other features''.}
+\texttt{O$_1$} and \texttt{O$_2$} are summed to obtain a total number of missing specific
+features in the architecture.
 
@@
 
-\paragraph{Other features} Meta-workflows are not supported by default
+\paragraph{Specific features} Meta-workflows are not supported by default
 (\texttt{O$_1$}=1).  Debugging is not an issue since the science
@@
 
-\paragraph{Other features} Meta-workflow are not supported by default
+\paragraph{Specific features} Meta-workflow are not supported by default
 (\texttt{O$_1$}=1). Fine-grained debugging information is usually easy
@@
 
-\paragraph{Other features} Meta-workflows are not available
+\paragraph{Specific features} Meta-workflows are not available
 (\texttt{O$_1$}=1).  Obtaining fine-grained information about workflow
@@
 be implemented to start new agents when the workload in the science
-gateway exceeds a certain threshold \todo{ref needed}. Distributed
+gateway exceeds a certain threshold~\cite{lorido2012auto}. Distributed
 engines are not available by default (\texttt{S$_2$=1}) and task
@@
 
-\paragraph{Other features} Meta-workflows are not available by default
+\paragraph{Specific features} Meta-workflows are not available by default
 (\texttt{O$_1$=1}).  Accessing debugging information is not likely to
@@
 
-\paragraph{Other features} 
+\paragraph{Specific features} 
 Meta-workflows are possible, which is one of the main interest of this
@@
 
-\paragraph{Other features} Meta-workflows are available after import,
+\paragraph{Specific features} Meta-workflows are available after import,
 by connecting workflows in the language used in the science gateway
@@
                                      & \cellcolor[HTML]{99DD99}\textbf{2}\\
-\multicolumn{7}{c}{\cellcolor[HTML]{EEEEEE}\textbf{Other features}}\\
+\multicolumn{7}{c}{\cellcolor[HTML]{EEEEEE}\textbf{Specific features}}\\
   Meta-workflow  -- \texttt{O$_1$}    & \cellcolor[HTML]{99BB99}1
@@
                                      & \cellcolor[HTML]{99FF99}0\\
-  \textbf{Total} (missing other features) & \cellcolor[HTML]{99DD99}\textbf{1}
+  \textbf{Total} (missing features) & \cellcolor[HTML]{99DD99}\textbf{1}
                                      & \cellcolor[HTML]{99DD99}\textbf{1}
@@
 
-Differences in other features should not be neglected. Nested
+Differences in other specific features should not be neglected. Nested
 workflows (with service invocation) and workflow import (with service
@@
                                      & \cellcolor[HTML]{99DD99}{0.50}\\
-Other features (normalized total) & \cellcolor[HTML]{99DD99}{0.50}
+Specific features (normalized total) & \cellcolor[HTML]{99DD99}{0.50}
                                      & \cellcolor[HTML]{99DD99}{0.50}

### 2016-04-21 paper.tex
Fixed a few todos

@@
 security reasons. This is the model adopted in LONI Pipeline
-Environment~\cite{dinov2009efficient} and in the Catania Science
-Gateway Framework~\cite{Ardizzone2012} which is used for instance in
-the DECIDE gateway instance for
+Environment~\cite{dinov2009efficient}, in the Distributed application
+runtime environment (DARE~\cite{maddineni2012distributed}), and in the
+Catania Science Gateway Framework~\cite{Ardizzone2012} which is used
+for instance in the DECIDE gateway for 
 neuroimaging~\cite{ardizzone2012decide} (see specific documentation on
-workflows\footnote{\url{http://bit.ly/1oQrzvQ}}). \todo{Add
-  description of a real system.\note{Tristan}{The Catania Science
-    Gateway Framework would be a good illustration here (workflow
-    engine is a Java portlet), but I can't find a suitable figure
-    about it.}}
-
+workflows\footnote{\url{http://bit.ly/1oQrzvQ}}). 
+%The Catania Science Gateway Framework would be a good illustration here (workflow engine is a Java portlet), but I can't find a suitable figure about it.
 %CIPRES~\cite{miller2010creating} (on which the Neuroscience gateway
@@
           Catania Science Gateway Framework~\cite{Ardizzone2012}, 
+          Distributed application runtime environment (DARE~\cite{maddineni2012distributed}),
           DECIDE~\cite{ardizzone2012decide}$^{c,n}$, LONI Pipeline Environment$^n$~\cite{dinov2009efficient}
@@
 \item Meta-workflow (\texttt{O$_1$}): ability to describe
-  meta-workflows from existing workflows.  \todo{More details about
-    why this is important?} \texttt{O$_1$}=0 when the feature is
-  available, \texttt{O$_1$}=1 otherwise.
+  meta-workflows from existing workflows. Meta-workflows offer an
+  additional level of flexibility to build worfklows from reusable
+  components. \texttt{O$_1$}=0 when the feature is available,
+  \texttt{O$_1$}=1 otherwise.
 \item Fine-grained debugging (\texttt{O$_2$}): availability of
-  fine-grained debugging information about workflow tasks. When
-  such information is available, workflow is seen as a white
+  fine-grained debugging information about workflow tasks. When such
+  information is available, workflow is seen as a white
   box. Fine-grained information about workflow tasks is required to
   properly troubleshoot workflow executions.  \texttt{O$_2$}=0 when
-  such information is easy to access, \texttt{O$_2$}=1
-  otherwise. \todo{More details about why this is important?}
+  such information is easy to access, \texttt{O$_2$}=1 otherwise.
 \end{itemize}
@@
 
-\paragraph{Extensibility} Adding a new type of workflow engine requires
-to implement the corresponding workflow service, to modify interaction
-\texttt{d}, and to implement interactions \texttt{b} and
+\paragraph{Extensibility} Adding a new type of workflow engine
+requires to implement the corresponding workflow service, to modify
+interaction \texttt{d}, and to implement interactions \texttt{b} and
 \texttt{$c_2$} (\texttt{E$_1$}=4). New engine versions can be added by
-updating the workflow service without modifying any interaction
-(\texttt{E$_2$}=0).  New workflows are added in the science gateway or
-in the workflow engine through interactions \texttt{a$_1$} and \texttt{a$_2$}
-(\texttt{E$_3$}=2). Adding a new type of infrastructure requires
-updates in interactions \texttt{b}, \texttt{$c_1$} and \texttt{$c_2$}
-(\texttt{E$_4$}=3).
+updating the workflow service without modifying any interaction or
+component (\texttt{E$_2$}=0). We consider that updating the engine
+version in a workflow service is transparent since the only goal of
+this component is to wrap the engine.  New workflows are added in the
+science gateway or in the workflow engine through interactions
+\texttt{a$_1$} and \texttt{a$_2$} (\texttt{E$_3$}=2). Adding a new
+type of infrastructure requires updates in interactions \texttt{b},
+\texttt{$c_1$} and \texttt{$c_2$} (\texttt{E$_4$}=3).
 
@@
 \texttt{c$_2$} (\texttt{E$_1$=3}). Updating the version of an engine
-is transparent (\texttt{E$_2$=0}) \todo{explain why, also for service}, and integrating a new workflow is
+is transparent (\texttt{E$_2$=0}) since it only requires updating the
+agent that is a component dedicated to the engine. Integrating a new workflow is
 done through interactions \texttt{a$_1$} and \texttt{a$_2$}

### 2016-04-21 paper.tex
Fixed typo and updated abstract

@@
   involvement in the SHIWA project, we abstract the software
-  components and interactions in 6 architectures that we compare using
-  metrics measuring integration effort, robustness, extensibility,
-  scalability and other specific features. Results show the pros and
-  cons of the different architectures. Tight integration and
-  sub-tasking require the least amount of integration effort and are
-  also the most robust. Most architectures perform the same in terms
-  of extensibility. The pool model is the most
-  scalable. Meta-workflows are only available in nested workflows and
-  workflow import.
+  components and interactions in 6 architectures. We use these
+  architectures to classify 18 existing systems and we compare these
+  architectures using metrics measuring integration effort,
+  robustness, extensibility, scalability and other specific
+  features. Results show the pros and cons of the different
+  architectures. Tight integration and sub-tasking require the least
+  amount of integration effort and are also the most robust. Most
+  architectures perform the same in terms of extensibility. The pool
+  model is the most scalable. Meta-workflows are only available in
+  nested workflows and workflow import.
 \end{abstract}
@@
 \begin{tabular}{ll}
-  \textbf{Architecture} & \textbf{System} \\
+  \textbf{Architecture} & \textbf{Systems} \\
   \hline

### 2016-04-22 paper.tex
Global read-through, several fixes, and tentative author list.

@@
 
-\title{Software architectures for workflow integration in science gateways}
+\title{Software architectures to integrate workflow engines in science gateways}
 
@@
 
-\author{}
-
-\address{}
+\author[mcgill,creatis]{Tristan Glatard}
+\author[mcgill]{Marc-Etienne Rousseau}
+\author[creatis]{Sorina Camarasu-Pop}
+
+\author[mcgill]{Reza Adalat}
+\author[mcgill]{Natacha Beck}
+\author[mcgill]{Najmeh Khalili-Mahani}
+\author[criugm]{Pierre-Olivier Quirion}
+\author[mcgill]{Pierre Rioux}
+
+\author[criugm]{Pierre Bellec}
+\author[mcgill]{Alan C. Evans}
+
+\address[mcgill]{McConnell Brain Imaging Centre, Montreal Neurological Institute, McGill University, Canada.}
+\address[creatis]{University of Lyon, CNRS, INSERM, CREATIS, Villeurbanne, France.}
+\address[criugm]{Centre de Recherche de l'Institut de G\'eriatrie de Montr\'eal CRIUGM, Montreal, QC, Canada.}
 
 \begin{abstract}
-%% Text of abstract
   We investigate software architectures to integrate workflow engines
@@
   and operations of the CBRAIN and VIP science gateways, and our past
-  involvement in the SHIWA project, we abstract the software
-  components and interactions in 6 architectures. We use these
-  architectures to classify 18 existing systems and we compare these
-  architectures using metrics measuring integration effort,
-  robustness, extensibility, scalability and other specific
-  features. Results show the pros and cons of the different
-  architectures. Tight integration and sub-tasking require the least
-  amount of integration effort and are also the most robust. Most
-  architectures perform the same in terms of extensibility. The pool
-  model is the most scalable. Meta-workflows are only available in
-  nested workflows and workflow import.
+  involvement in the SHIWA project, we abstract software components
+  and interactions in 6 architectures that we use to classify 18
+  systems. We compare these architectures using metrics measuring
+  integration effort, robustness, extensibility, scalability and other
+  specific features. Results show the pros and cons of the different
+  architectures: tight integration and sub-tasking require the least
+  amount of integration effort and are also the most robust, most
+  architectures perform the same in terms of extensibility, the pool
+  model is the most scalable and meta-workflows are only available in
+  nested workflows and workflow import. These results provide insights
+  to software architects on how to integrate science gateways and
+  workflow engines.
 \end{abstract}
@@
 workflow engines in science gateways, with important consequences on
-the required development effort and on the characteristics of the
-resulting system. This paper describes and compares such architectures
-based on system-independent representations of their main components
-and interactions.
+the required development effort and on the resulting system. This
+paper describes, exemplifies and compares such architectures, based on
+system-independent representations of their main components and
+interactions.
 
@@
 the past 7 years. It also builds on the lessons learned from several
-science gateway and workflow-related projects, including the SHIWA
-Simulation Platform~\cite{terstyanszky2014enabling}.
+science gateway and workflow-related projects, including
+SHIWA\footnote{\url{http://www.shiwa-workflow.eu}} and related
+initiatives.
 
@@
 integration effort, robustness, extensibility, scalability and other
-specific features related to the architectures. We use these metrics
-to evaluate and compare the 6 architectures. The paper closes on a
+specific features of the architectures. We use these metrics to
+evaluate and compare the 6 architectures. The paper closes on a
 discussion and conclusion that highlight architectural recommendations
-for science gateway and workflow engine developers.
+for science gateway and workflow engine architects.
 
@@
 
-In the last decade, the e-Science workflow community has developed
-workflow systems to help developers access distributed infrastructures
-such as clusters, grids, clouds and web services. These efforts
-resulted in tools such as Askalon~\cite{fahringer2005askalon},
+In the last decade, the e-Science community has developed workflow
+systems to help developers access distributed infrastructures such as
+clusters, grids, clouds and web services. These efforts resulted in
+tools such as Askalon~\cite{fahringer2005askalon},
 Hyperflow~\cite{balis2016hyperflow}, MOTEUR~\cite{GLAT-08i},
@@
 Triana~\cite{taylor2007triana}, VisTrails~\cite{callahan2006managing},
-WS-PGRADE~\cite{Kacsuk2012}, etc. Such workflow engines usually
-describe applications in a specific language that offers parallel data
-and control flow constructs, visual edition features, links with
-domain-specific tool repositories, provenance recording, among other
-features (see review in~\cite{deelman2009workflows}).
+WS-PGRADE~\cite{Kacsuk2012} and others. Such workflow engines usually
+describe applications in a high-level language that offers specific
+data and control flow constructs, parallelization operators, visual
+edition features, links with domain-specific tool repositories,
+provenance recording, among other features. An overview of workflow
+systems features and capability is available
+in~\cite{deelman2009workflows}.
 % Descriptions and experiments
@@
 
-At the same time, specific toolboxes were emerging in various
-scientific domains to facilitate the assembly of software components
-in consistent pipelines. In neuroimaging, our primary domain of
-interest, tools such as Nipype~\cite{gorgolewski2011nipype},
-PSOM~\cite{bellec2012pipeline}, SPM~\cite{ashburner2011spm} and
-FSL~\cite{Jenkinson2012782} provide abstractions and functions to
+At the same time, toolboxes were emerging in various scientific
+domains to facilitate the assembly of software components in
+consistent ``pipelines''. In neuroimaging, our primary domain of
+interest, tools such as Nipype (Neuroimaging in Python, Pipelines and
+Interfaces~\cite{gorgolewski2011nipype}), PSOM (Pipeline System for
+Octave and Matlab~\cite{bellec2012pipeline}), SPM (Statistical
+Parametric Mapping~\cite{ashburner2011spm}) and FSL (FMRIB Software
+Library~\cite{Jenkinson2012782}) provide abstractions and functions to
 handle the data and computing flow between processes implemented in a
 variety of programming languages. Such tools were interfaced to
-computing systems, in particular clusters, to create tasks, handle
-their dependencies, and execute them. For instance, FSL can launch
-tasks on SGE through its fsl\_sub tool, Nipype has execution
-plugins for S/OGE, Torque, IPython, ssh and multi-processor servers,
-and PSOM is interfaced to Torque, SGE, MOAB, LSF and Condor. Some of
-these tools also support advanced features such as provenance tracking
-(Nipype, PSOM), redundancy detection across analyses to avoid
-re-computation (PSOM), and so on. A wide-array of workflows have been
-implemented using these pipelines and are now widely shared across
-neuroimaging groups world-wide. \todo{Provide more context on that?}
-These domain-specific engines represent a tremendous opportunity for
-science gateways to leverage existing tools and applications. They
-nicely complement e-Science engines that are more oriented towards the
+computing systems, in particular clusters, to handle high-throughput
+task executions. For instance, FSL can launch tasks on SGE through its
+fsl\_sub tool, Nipype has execution plugins for S/OGE, Torque,
+IPython, ssh and multi-processor servers, and PSOM is interfaced to
+Torque, SGE, MOAB, LSF and Condor. Some of these tools also support
+advanced features such as provenance tracking (Nipype, PSOM),
+redundancy detection across analyses to avoid re-computation (PSOM),
+and so on. A wide-array of workflows have been implemented using these
+pipelines and are now widely shared across neuroimaging groups
+world-wide. These
+domain-specific engines represent a tremendous opportunity for science
+gateways to leverage existing tools and applications. They nicely
+complement e-Science engines that are more oriented towards the
 exploitation of distributed computing infrastructures, in particular
@@
 
-For the sake of the analysis conducted in this paper, we
-define a \emph{workflow engine} (also abbreviated \emph{engine}) as a
+For the sake of the analysis conducted in this paper, we define a
+\emph{workflow engine} (also abbreviated \emph{engine}) as a piece of
 software that submits interdependent computing \emph{tasks} to an
-infrastructure (local server, cluster, grid or cloud) based on a
-workflow description (a.k.a \emph{workflow}) and using input data that
-may be files, data base entries, or simple parameter values. Although
-simplistic, this definition covers both e-Science and domain-specific
-engines. Some workflow engines, usually e-Science ones, may transfer
-data across the infrastructure, and others, usually domain-specific
-ones, may leave it to external processes. Workflows may be expressed
-in any language, including high-level XML or JSON dialects such as
-Scufl or Hyperflow, and low-level scripts such as
-Bash.
+\emph{infrastructure} (local server, cluster, grid or cloud) based on
+a workflow description (a.k.a \emph{workflow}) and using input data
+that may consist of files, database entries or simple parameter
+values. Although simplistic, this definition covers both e-Science and
+domain-specific engines. Some workflow engines, usually e-Science
+ones, may transfer data across the infrastructure, and others, usually
+domain-specific ones, may leave it to external processes. Workflows
+may be expressed in any language, including high-level XML or JSON
+dialects such as Scufl or Hyperflow, and low-level scripting languages
+such as Bash.
 
@@
 
-Science gateways 
-are used to share resources in a community and to provide performance
-through facilitated access to storage and computing power. They are
-usually accessible through a web interface that helps users manage
-access rights, data transfers, task execution and authentication on
-multiple computing and storage locations. Workflow engines are of
-course part of this ecosystem as core components to implement and
-execute applications.
+Science gateways are used to share resources in a community and to
+provide performance through facilitated access to storage and
+computing power. They are usually accessible through a web interface
+that helps users manage access rights, data transfers, task execution
+and authentication on multiple computing and storage
+locations. Workflow engines are part of this ecosystem as core
+components to implement and execute applications.
 
@@
 across systems. Some science gateways may be tailored to a particular
-workflow engine that manages critical functions such as task
-execution, data transfers and authentication. Other gateways may be
-more general and host applications executed by different types of
-engines. Extensibility is an important property of such integration:
+engine that manages critical functions such as task execution, data
+transfers and authentication. Other gateways may be more general and
+host applications executed by different types of
+engines. Extensibility is an important property of the integration, as
 new workflows are added frequently, different types and versions of
 workflow engines may be integrated over time, and different kinds of
-infrastructure may be covered.
+infrastructure can be targeted.
 
 Scalability is also an important concern, since science gateways are
-multi-user systems, manage substantial amounts of data and execute
-computationally-intensive analyses. For this purpose, science gateways
+multi-user systems that manage substantial amounts of data and execute
+computationally-intensive analyses. To this end, science gateways
 may include different instances of the same engine type and implement
-some load-balancing policy among them. They may also enforce
-scheduling policies on the infrastructure, e.g. to ensure fairness
+some load-balancing policy among them. They may also use advanced
+task scheduling policies on the infrastructure, e.g. to ensure fairness
 among users~\cite{FERR-14,CPE:CPE3708} or to improve
@@
 Robustness is a key part of science gateway user experience. In this
-respect, the multiplicity of software components and functions is a
-challenge as it often leads to complex architectures that are
-therefore prone to failure. To offer good user experience, science
-gateways should remain as transparent as possible, which requires
-robustness.
-
-Specific features may also be available in science
-gateways, such as data vizualisation and quality control, workflow
-edition, debugging instruments, social tools among users, etc.
-
-Various science gateways have been developed in the last years,
-including science gateway frameworks used as building blocks to
-assemble science gateways. The main frameworks are
-WS-PGRADE/gUSE~\cite{Kacsuk2012}, the Catania Science Gateway
-Framework~\cite{Ardizzone2012} and Apache
-Airavata~\cite{marru2011apache}. Numerous science gateway instances
-were built using such frameworks, in various disciplines. Other
+respect, the multiplicity of software components and interactions
+involved in architectures is a challenge as it often leads to complex
+systems that are therefore prone to failure. To offer good user
+experience, science gateways should remain as transparent as possible,
+which requires robustness.
+
+Specific features may also be available in science gateways, such as
+data vizualisation and quality control, workflow edition, debugging
+instruments, social tools among users, etc.
+
+Various science gateways have been developed, including frameworks
+used as building blocks to assemble science gateways. The main
+frameworks are Apache Airavata~\cite{marru2011apache}, the Catania
+Science Gateway Framework~\cite{Ardizzone2012} and
+WS-PGRADE/gUSE~\cite{Kacsuk2012}. Numerous science gateway instances
+were built using such frameworks\note{Tristan}{Add references}. Other
 gateway instances were also built independently from any framework,
-using regular web development tools and various libraries to manage
-distributed systems. CBRAIN and VIP are two of such science gateways
-that were initially developed for medical imaging analysis and from
-which we gathered experience about the architectures described in
+using regular web development tools and libraries to access
+distributed systems. CBRAIN and VIP are two of such science gateways,
+from which we gathered experience about the architectures described in
 this paper.
@@
 experience. It uses the MOTEUR workflow engine~\cite{GLAT-13} and the
-DIRAC pilot-job framework to execute applications on EGI. Almost 900
-users were registered in VIP in April 2016 and the average CPU
-consumption between January 2013 and March 2016 was 35 CPU years per
-month.
+DIRAC pilot-job framework~\cite{tsaregorodtsev2008dirac} to execute
+applications on EGI. Almost 900 users were registered in VIP as of
+April 2016 and the average CPU consumption between January 2013 and
+March 2016 was 35 CPU years per month.
 
@@
 as other applications and invoked through a dedicated service. The
-Fine-Grained Interoperability framework was also developed in SHIWA,
-where conversions between workflow languages are achieved through a
-common intermediate
+Fine-Grained Interoperability framework was also developed in the
+SHIWA project, where conversions between workflow languages are
+achieved through a common intermediate
 representation~\cite{plankensteiner-prodan-etal:2013}. \note{Tristan}{Silvia,
@@
 services used to access these resources. Infrastructures can be
-servers, clusters, grids or clouds. Some workflow engines and science
-gateways may assume specific characteristics about the infrastructure,
-such as the presence of a shared file system between the computing
-nodes, the availability of a global task meta-scheduler, the presence
-of a file catalog, etc. In the analysis presented below, such specific
-functions are not discussed. Instead, we consider the infrastructure
-as an abstract system that can execute tasks and store files
-regardless of the enabling mechanisms.
+computing or file servers, databases, clusters, grids or clouds. Some
+workflow engines and science gateways may assume specific
+characteristics about the infrastructure, such as the presence of a
+shared file system between the computing nodes, the availability of a
+global task meta-scheduler, the presence of a file catalog, etc. In
+the analysis presented below, such specific functions are not
+discussed. Instead, we consider the infrastructure as an abstract
+system that can execute tasks and store data regardless of the
+enabling mechanisms.
 
@@
 
-The architectures are diagrammed in Figure~\ref{fig:architectures}
-using the graphical notations shown in
-Figure~\ref{fig:notations}. Architectures are described from their
-main software \emph{components} and \emph{interactions}. Software
-components include science gateway and infrastructure in all
-architectures while workflow service, workflow pool and agent are
-involved in some architectures. The workflow engine itself is
-represented by a specific symbol with dotted lines. Software
-interactions among these components are represented with arrows.
-\emph{Abstract} interactions are specific types of interactions that
-may be implemented by various different software interactions. They
-are represented with dotted arrows and labeled with \texttt{*}. An
-architecture that has an abstract interaction is an abstract
-architecture.  Red color is used to represent the components and
-interactions that are specific to workflow execution, i.e. that are
-not involved in any other process. Tasks refer to computing tasks that
-are created by the workflow engine and executed on the
-infrastructure. Data represents any type of file, or database that is
-involved in the workflow execution and stored on the infrastructure
-while the workflow executes. It does not cover workflow parameters
-such as strings, numbers, etc.
+Architectures to integrate workflow engines in science gateways are
+diagrammed in Figure~\ref{fig:architectures} using the graphical
+notations shown in Figure~\ref{fig:notations}.
+
+Architectures are described from their main software \emph{components}
+and \emph{interactions}. Software components include science gateway
+and infrastructure in all architectures while workflow service,
+workflow pool and agent are involved in some architectures only. The
+workflow engine itself is represented by a specific symbol with dotted
+lines.
+
+Software interactions among these components are represented with
+arrows.  \emph{Abstract} interactions are specific types of
+interactions that may be implemented using different components and
+interactions. They are represented with dotted arrows and labeled with
+\texttt{*}. An architecture that has an abstract interaction is an
+abstract architecture that can be instantiated in various ways.
+
+Red color is used to represent the components and interactions that
+are specific to workflow execution, i.e. that are not involved in any
+other process. Tasks refer to computing tasks created by the workflow
+engine and executed on the infrastructure. Data represents any type of
+file or database involved in workflow execution and stored on the
+infrastructure. It does not cover workflow parameters such as strings,
+numbers, etc.
 
 The remainder of this section describes the different types of
-interactions involved in the architectures, details architectures and
-illustrates them using examples of real
+interactions involved in the architectures. Architectures are then
+detailed and illustrated using examples of real
 systems. Table~\ref{table:system-classification} summarizes the
@@
 \end{figure*}
-
-\subsection{Interactions}
-
-The interactions involved in the architectures are described below and
-labeled consistently with the notations used on
-Figure~\ref{fig:architectures}.
-
-\begin{enumerate}[leftmargin=0cm,itemindent=0.6cm,label=\texttt{(\alph*)}]
-
-\item Workflow integration: consists in adding a new workflow to the
-  system so that users can execute it. It is triggered by an
-  administrator of the science gateway and it results in an interface,
-  for instance a web form, where users can enter the parameters of the
-  workflow to be executed. The interaction has two aspects:
-  (\texttt{a$_1$}) the programs used in the workflow are installed on
-  the infrastructure, which may or may not require administrative
-  privileges on the infrastructure; (\texttt{a$_2$}) the workflow is
-  configured in the science gateway so that it becomes available to
-  users. Note that integrating a workflow is not the same process as
-  integrating a workflow \emph{engine}.
-\item Task control: operations to manage tasks on the infrastructure,
-  including: authentication, submission, monitoring, termination,
-  deletion, etc. Controlling tasks requires to deal with the
-  heterogeneous batch managers and meta-schedulers that might be
-  available on the infrastructure. When the infrastructure is a grid
-  or a cloud, it is for instance implemented using libraries
-  implementing standards such as SAGA~\cite{goodale2006saga},
-  DRMAA~\cite{troger2012distributed} or OCCI~\cite{edmonds2012toward}.
-\item Data control: operations to manage data on the infrastructure,
-  such as: upload, download, deletion, browsing, replication, caching,
-  etc. Data movements can be triggered by the user in the science
-  gateway (\texttt{$c_1$}), to upload input data or download processed
-  data. They can also be performed by the workflow engine
-  (\texttt{$c_2$}), to transfer workflow data (inputs, outputs, or
-  intermediate results) across the infrastructure so that tasks can
-  use it. The infrastructure might offer various data storage backends
-  with heterogeneous interfaces. \note{Marc-Etienne}{this -c2- can be also trigerred
-    by the user}\note{Tristan}{No, it can't: this is by definition an interaction between the workflow engine and the infrastructure.
-It is used to transfer data to/from tasks.
-I reworded a bit, is it clearer now?} Tools and services such as
-  JSAGA~\cite{reynaud2010uniform} or Data Avenue~\cite{hajnal2014data}
-  can be used to homogeneize these interfaces.
-\item Workflow control: operations to execute a workflow with an
-  engine, including: workflow submission, monitoring, termination,
-  etc. Workflow control can be coarse-grained (a.k.a. black box) or
-  fine-grained (white box). In a coarse-grained model, the various
-  tasks created by a workflow execution are masked and the user only
-  has a global view of the workflow execution. In a fine-grained
-  model, user is exposed to the workflow topology, i.e. to the outputs
-  of the individual tasks, their statuses and so on.
-\item Sub-task control: operations used by tasks to submit sub-tasks
-  on the infrastructure, including: submission, monitoring,
-  termination, deletion, etc. Sub-task control is similar to
-  interaction \texttt{b}, except that information about the parent
-  task which submitted a sub-task is usually available and used for
-  additional control. For instance, the parent task may wait for all
-  its sub-tasks to finish before finishing, and conversely all the
-  sub-tasks may be killed if the parent task is killed.
-\item Pool-agent: specific to the architecture described
-  in Section~\ref{sec:pool}. This is an interaction used when agents retrieve
-  work from a central pool. It covers agent registration and
-  de-registration to the pool, protocols to send work from the pool to
-  the agent, mechanisms to update work status, and so on. 
-\item Workflow conversion: translation from one workflow language to
-  another one. This interaction may not be available or implementable
-  for every workflow language. It has been developed mostly for
-  well-structured and relatively simple workflow language such as between GWorkflowDL
-  and Scufl~\cite{OLAB-09} and between the 4 languages that were involved in the
-  SHIWA initiative. In SHIWA, workflow conversion is done through an
-  intermediate representation, the Interoperable Workflow Intermediate
-  Representation~\cite{plankensteiner-montagnat-etal:2011}, which
-  allows to convert among $n$ workflow languages using $2n$
-  interactions instead of $n^2$.
-\end{enumerate}
-
-\subsection{Tight integration}
-
-See Figure~\subref*{archi:tight}. The workflow engine is tightly
-integrated with the science gateway, which means that it is deployed
-on the same machine and potentially shares code, libraries and other
-software components with the science gateway. For instance, the
-workflow engine might be a portlet in a Liferay portal or a controller
-in a Ruby on Rails application. The workflow engine and the science
-gateway usually share a database where application, users and other
-resources are stored. In this model, task and data control are both
-initiated from the science gateway. Interactions \texttt{b} and
-\texttt{c$_2$} are initiated from the workflow engine while
-\texttt{c$_1$} comes from other parts of the science gateway, for
-instance a data management interface. As in any other model, the
-installation of new workflows in the science gateway (\texttt{a$_2$})
-and infrastructure (\texttt{a$_1$}) is done by an administrator, for
-security reasons. This is the model adopted in LONI Pipeline
-Environment~\cite{dinov2009efficient}, in the Distributed application
-runtime environment (DARE~\cite{maddineni2012distributed}), and in the
-Catania Science Gateway Framework~\cite{Ardizzone2012} which is used
-for instance in the DECIDE gateway for 
-neuroimaging~\cite{ardizzone2012decide} (see specific documentation on
-workflows\footnote{\url{http://bit.ly/1oQrzvQ}}). 
-%The Catania Science Gateway Framework would be a good illustration here (workflow engine is a Java portlet), but I can't find a suitable figure about it.
-%CIPRES~\cite{miller2010creating} (on which the Neuroscience gateway
-%described in~\cite{sivagnanam2015early} is based) % \note{Tristan}{There is no workflow engine in there.}
-
-\subsection{Service invocation}
-
-\begin{figure*}
-\centering
-\def\svgwidth{1.5\columnwidth}
-\input{figures/VIP.pdf_tex}
-\caption{Architecture of the Virtual Imaging Platform that integrates
-  the MOTEUR engine through service invocation.}
-\label{fig:vip-architecture}
-\end{figure*}
-
-See Figure~\subref*{archi:service}. The workflow engine is available
-externally to the science gateway, in a service. The science gateway
-controls the service through a specific interaction (\texttt{d}) that
-might be implemented as a web-service call (e.g. RESTful or SOAP), as
-a command-line or as any other method that offers a well-defined
-interface to the workflow engine. The workflow engine might be invoked
-either as a black box that completely masks the infrastructure and
-tasks, or as a white box that allows for some interaction with
-them. The workflow engine is responsible for controlling the tasks on
-the infrastructure (\texttt{b}) and for performing the required data
-transfers to execute them (\texttt{c$_2$}). User data is usually
-managed through the science gateway (\texttt{c$_1$}), although it
-might as well be delivered by the workflow engine directly to the
-user.
-
-This architecture is largely adopted, in systems such as Apache
-Airvata~\cite{marru2011apache}, Vine
-Toolkit~\cite{DBLP:journals/scpe/SzejnfeldDKKKKLPTWDNW10}, Virtual
-Imaging Platform~\cite{GLAT-13}, the system
-in~\cite{wu2010accelerating}, the WS-PGRADE/gUSE
-framework~\cite{Kacsuk2012} and the numerous science gateway instances
-that rely on WS-PGRADE/gUSE, for instance
-e-BioInfra~\cite{shahand2015data}, MoSGrid~\cite{krüger2014mosgrid},
-and many others~\cite{kacsuk2014science}. The
-integration between the HubZero science gateway and Pegasus workflow
-engine performed in~\cite{CPE:CPE3257} also use a service architecture
-where the \texttt{d} interaction is implemented as a set of calls to
-the Pegasus command-line tools (e.g. pegasus-status,
-pegasus-statistics, etc). Figure~\ref{fig:vip-architecture} shows the
-architecture of the Virtual Imaging Platform, where the MOTEUR
-workflow engine~\cite{GLAT-08i} is invoked as a service at step
-\texttt{3} (interaction \texttt{d} on
-Figure~\ref{fig:architectures}). Step \texttt{2} on
-Figure~\ref{fig:vip-architecture} corresponds to interaction
-\texttt{c$_1$}, and step \texttt{4} maps to interactions \texttt{b}
-and \texttt{c$_2$} (in VIP, workflow data transfers are performed by
-the tasks and embedded in their descriptions).
-
-\subsection{Sub-tasking}
-
-See Figure~\subref*{archi:sub-task}. The workflow engine is a particular
-task that can submit sub-tasks to the science gateway through
-interaction \texttt{e}. The workflow engine keeps track of the
-dependencies between the sub-tasks but their execution is delegated to
-the science gateway that executes them on the infrastructure through
-interaction \texttt{b}. Although the science gateway has no global
-vision of the workflow, it can keep track of the sub-tasks submitted
-by a given task, for instance to be able to cancel them when the task
-is canceled. The science gateway may also implement mechanisms to
-facilitate the handling of task dependencies, for instance basic
-dependency lists as available in most batch managers (see for instance
-attribute \texttt{depend} of option \texttt{-W} in
-Torque\footnote{\url{http://docs.adaptivecomputing.com}}).
-
-The science gateway also transfers both user and workflow data across
-the infrastructure, so that interactions \texttt{c$_1$} and
-\texttt{c$_2$} are both covered by \texttt{c}. In practice, both
-\texttt{c$_1$} and \texttt{c$_2$} can be implemented using the same
-pieces of code.
-
-The sub-tasking architecture is implemented in CBRAIN~\cite{SHER-14}
-where it is used to integrate the FSL toolkit~\cite{Jenkinson2012782}
-and the PSOM workflow engine~\cite{bellec2012pipeline}.
-
-\begin{figure}
-\def\svgwidth{\columnwidth}
-\input{figures/CBRAIN-FSL.pdf_tex}
-\caption{Architecture of the CBRAIN-FSL integration (sub-tasking). 1:
-  User sends data and workflow execution request to storage server(s)
-  and CBRAIN portal. 2: CBRAIN portal sends execution request to
-  execution server on cluster. 3: Execution server transfers data from
-  storage server(s). 4-5: Execution server starts FSL tool (pipeline)
-  via task scheduler. 6: FSL pipeline submits sub-tasks execution
-  server through fsl\_sub. 7-8: Execution server starts sub-tasks
-  through task scheduler. 9.  Execution server transfers results to
-  storage server(s).}
-\label{fig:cbrain-fsl-architecture}
-\end{figure}
-The CBRAIN-FSL integration allows to leverage FSL pipelines that are
-written in a very low-level workflow language (Linux executables and
-script) that submits tasks uniformly through a specific tool called
-fsl\_sub. It is represented in
-Figure~\ref{fig:cbrain-fsl-architecture}.  On this Figure, the science
-gateway is represented by components \texttt{CBRAIN portal} and
-\texttt{CBRAIN execution server} and the infrastructure consists of
-\texttt{Storage servers} and \texttt{Computing cluster with shared
-  file system}. Interaction \texttt{b} on
-Figure~\ref{fig:architectures} is implemented by steps \texttt{4},
-\texttt{5}, \texttt{7} and \texttt{8} (regular interactions with batch
-manager). Interaction \texttt{c} is implemented by steps \texttt{3}
-and \texttt{9}. Interaction \texttt{e} is implemented by step
-\texttt{6} through a modified version of the fsl\_sub
-script\footnote{See modified fsl\_sub script available at
-  \url{https://github.com/aces/cbrain-plugins-neuro}}.
-
-\begin{figure}
-\def\svgwidth{\columnwidth}
-\input{figures/CBRAIN-PSOM.pdf_tex}
-\caption{Architecture of the CBRAIN-PSOM integration (sub-tasking,
-  Figure extracted from~\cite{GLAT-16}). 1: User sends data and
-  workflow execution request to storage server(s) and CBRAIN
-  portal. 2: CBRAIN portal sends execution request to execution server
-  on cluster. 3: Execution server transfers data from storage
-  server(s). 4-5: Execution server starts PSOM master via task
-  scheduler. 6: PSOM master submits PSOM workers to execution
-  server. 7-8: Execution server starts PSOM workers through task
-  scheduler. 9: PSOM master and workers execute workflow. 10.
-  Execution server transfers results to storage server(s).}
-\label{fig:cbrain-psom-architecture}
-\end{figure}
-The CBRAIN-PSOM integration~\cite{GLAT-16} is illustrated on
-Figure~\ref{fig:cbrain-psom-architecture}. The PSOM workflow engine
-adopts a pilot-job architecture~\cite{turilli2015comprehensive} where
-the a master coordinates workflow execution by submitting workers and
-establishing direct communication channels with them. Note how this
-peculiar execution model is totally supported by the sub-tasking
-architecture.
-
-\subsection{Pool model}
-\label{sec:pool}
-See Figure~\subref*{archi:agent}. Workflows are submitted by the science
-gateway to a pool through interaction \texttt{d}. Agents connect to
-the pool asynchronously to retrieve and execute workflows through
-interaction \texttt{f}. Agents may be started according to various
-policies, for instance to ensure load balancing. Agents may wrap
-different types of workflow engines. Workflow engine controls tasks
-and data on the infrastructure through interactions \texttt{b} and
-\texttt{c$_2$}, science gateway transfers user data through
-interaction \texttt{c$_1$}, and administrator installs workflows
-through interaction \texttt{a$_1$} and \texttt{a$_2$}.
-
-The pool model was implemented in the SHIWA pool~\cite{ROGE-13}, see
-Figure~\ref{fig:shiwa-pool-architecture}. In Figure
-~\ref{fig:shiwa-pool-architecture}, interaction \texttt{d} of
-Figure~\subref*{archi:agent} is implemented in 3 different types of calls
-to the pool: workflow submission (\texttt{1} \& \texttt{2}), workflow
-status retrieval (\texttt{a} and \texttt{b}), and workflow retrieval
-(\texttt{13} \& \texttt{14}). In Figure
-~\ref{fig:shiwa-pool-architecture}, interaction \texttt{f} of
-Figure~\subref*{archi:agent} is implemented through 2 types of calls:
-workflow instance retrieval (\texttt{4}, \texttt{5} and \texttt{6}),
-and workflow instance update (\texttt{11} and \texttt{12}). Workflow
-instance retrieval is used by the agents to fetch work from the
-pool. Workflow instance update is used by the agents to update
-workflow statuses. In the SHIWA pool, agents can wrap different types
-of workflow engines to execute workflows expressed with different
-languages. Calls \texttt{0}, \texttt{7}, \texttt{8} and \texttt{9} on
-Figure~\ref{fig:shiwa-pool-architecture} are used by workflow engine
-plugins to declare their supported language and launch engines, and by
-workflow engines to report their status to the agent. These calls are
-specific to the SHIWA Pool implementation of the \texttt{agent}
-component and therefore have no corresponding representation in
-Figure~\subref*{archi:agent}.
-
-\begin{figure*}
-\centering
-\includegraphics[width=1.5\columnwidth]{figures/pool-interactions.pdf}
-\caption{Architecture of the SHIWA pool (Figure reproduced
-  from~\cite{ROGE-13}). Circle-terminated arrows indicate messages
-  that are broadcast to all pool clients.}
-\label{fig:shiwa-pool-architecture}
-\end{figure*}
-
-\subsection{Nested workflows}
-
-See Figure~\subref*{archi:nested}. In nested workflows
-(Figure~\subref*{archi:nested}-Left), a task of a \emph{parent} workflow
-executed by the \emph{parent} engine is itself a workflow, called
-\emph{child} workflow, that is executed by a \emph{child} engine. The
-parent and child engines might use different workflow description
-languages. They might also execute workflows on different
-infrastructures. The parent workflow is also called meta-workflow. The
-science gateway communicates with the parent engine through
-interaction \texttt{*$_1$}. The science gateway also communicate with
-the infrastructure to transfer user data through abstract interactions
-\texttt{*$_2$} and \texttt{*$_3$}. Both workflow engines communicate
-with the infrastructure through abstract interactions as well,
-\texttt{*$_4$} and \texttt{*$_5$}. The parent engine communicates with
-the child engine through abstract interaction
-\texttt{*$_6$}. Administrator installs workflows in the science
-gateway through interaction \texttt{a$_2$}, and installs software on
-infrastructures through interactions \texttt{a$_1$}.
-
-Nested workflows are abstract architectural patterns that can be
-instantiated in the various architectures described previously. We
-focus on instantiation with the service invocation model (see
-Figure~\subref*{archi:nested}-Right) as this is the most used
-architecture. In such an instantiation, we assume that the parent and
-child workflow engines are distinct pieces of software that require
-different workflow services invoked by distinct \texttt{d}
-interactions. If this is not the case, then workflow services can be
-collapsed in a single one with a \texttt{d} interaction with
-itself. Workflow engines communicate with infrastructures using
-\texttt{b} and \texttt{c$_2$}. Science gateway transfer user data to
-infrastructures using \texttt{c$_1$} interactions.
-
-Nested workflows have long been available in workflow engines, for
-instance in the Taverna workbench~\cite{oinn2004taverna}. They are
-also used implicitly in several platforms where workflow engines are
-wrapped in workflow tasks as any other command-line tool. Nested
-workflows were notably used by the SHIWA Science Gateway to implement
-so-called Coarse-Grained workflow
-interoperability~\cite{terstyanszky2014enabling}, i.e. to integrate
-various workflow engines in a consistent
-platform. Figure~\ref{fig:shiwa-architecture} shows the architecture
-used in the SHIWA Simulation Platform for nested workflow execution with
-service invocation as represented in Figure~\subref*{archi:nested}. The
-parent workflow engine is \texttt{WS-PGRADE}, invoked as a service in
-the Science Gateway (step \texttt{1} on
-Figure~\ref{fig:shiwa-architecture}, interaction \texttt{d} on
-Figure~\subref*{archi:nested}). Ten different child engines can be used by
-nested workflows, invoked through the \texttt{Submission service}
-(step \texttt{2}, interaction \texttt{d}). Each of these engines can
-submit jobs and transfer data to a distributed computing
-infrastructure (\texttt{DCI}, step \texttt{3}, interactions \texttt{b}
-and \texttt{c$_2$}). Data interactions (\texttt{c}) and application
-porting ones (\texttt{a}) are not represented on
-Figure~\ref{fig:shiwa-architecture}.
-
-\begin{figure*}
-\centering
-\includegraphics[width=1.5\columnwidth]{figures/shiwa-science-gateway.pdf}
-\caption{Nested workflow execution through SHIWA Science Gateway
-  (Figure reproduced from~\cite{terstyanszky2014enabling} with
-  permission of the first author).}
-\label{fig:shiwa-architecture}
-\end{figure*}
-
-\subsection{Workflow import}
-
-See Figure~\subref*{archi:import}. This is an abstract model that we
-instantiate with the service invocation architecture for
-consistency. Workflows are integrated in the science gateway through
-format conversion from a native format to the science gateway
-format. Workflow import is usually an offline process which is not
-involved in workflow execution. Workflow import was implemented in the SHIWA
-Simulation Platform through the IWIR language that provided a common
-language for portability across grid workflow
-systems~\cite{plankensteiner-prodan-etal:2013}.
-
-\todo{Add description of a real system}
-
-\todo{Refer to ``From the desktop to the grid: scalable bioinformatics via workflow conversion Luis de la GarzaEmail author, Johannes Veit, Andras Szolek, Marc Röttig, Stephan Aiche, Sandra Gesing, Knut Reinert and Oliver Kohlbacher''}
-
 \begin{table*}
@@
 
+\subsection{Interactions}
+
+The interactions involved in the architectures are described below and
+labeled consistently with the notations used on
+Figure~\ref{fig:architectures}.
+
+\begin{enumerate}[leftmargin=0cm,itemindent=0.65cm,label=\texttt{(\alph*)}]
+
+\item Workflow integration: consists in adding a new workflow to the
+  system so that users can execute it. It is triggered by an
+  administrator of the science gateway and it results in an interface,
+  for instance a web form, where users can enter the parameters of the
+  workflow to be executed. The interaction has two aspects:
+  (\texttt{a$_1$}) the programs used in the workflow are installed on
+  the infrastructure, which may require administrative privileges on
+  the infrastructure; (\texttt{a$_2$}) the workflow is configured in
+  the science gateway so that it becomes available to users. Note that
+  integrating a workflow is not the same process as integrating a
+  workflow \emph{engine}.
+\item Task control: operations to manage tasks on the infrastructure,
+  including: authentication, submission, monitoring, termination,
+  deletion, etc. Controlling tasks requires to deal with the
+  heterogeneous batch managers and meta-schedulers that might be
+  available on the infrastructure. When the infrastructure is a grid
+  or a cloud, it may for instance be achieved using libraries that
+  implement standards such as SAGA (Simple API for Grid
+  Applications~\cite{goodale2006saga}), DRMAA (Distributed Resource
+  Management Application API~\cite{troger2012distributed}) or OCCI
+  (Open Cloud Computing Interface~\cite{edmonds2012toward}).
+\item Data control: operations to manage data on the infrastructure,
+  such as: upload, download, deletion, browsing, replication, caching,
+  etc. Data movements can be triggered by the user in the science
+  gateway (\texttt{$c_1$}), to upload input data or download processed
+  data. They can also be performed by the workflow engine
+  (\texttt{$c_2$}), to transfer workflow data (inputs, outputs, or
+  intermediate results) across the infrastructure so that tasks can
+  use it. The infrastructure might offer various data storage backends
+  with heterogeneous interfaces. \note{Marc-Etienne}{this -c2- can be also trigerred
+    by the user}\note{Tristan}{No, it can't: this is by definition an interaction between the workflow engine and the infrastructure.
+It is used to transfer data to/from tasks.
+I reworded a bit, is it clearer now?} Tools and services such as
+  JSAGA~\cite{reynaud2010uniform} or Data Avenue~\cite{hajnal2014data}
+  can be used to homogeneize these interfaces.
+\item Workflow control: operations to control workflow execution in an
+  engine, including: workflow submission, monitoring, termination,
+  etc. Workflow control can be coarse-grained (black box) or
+  fine-grained (white box). In a coarse-grained model, the various
+  tasks created by a workflow execution are masked and the user only
+  has a global view of the workflow execution. In a fine-grained
+  model, user is exposed to the workflow topology, i.e. to the outputs
+  of the individual tasks, their statuses and so on.
+\item Sub-task control: operations used by tasks to submit sub-tasks
+  on the infrastructure, including: submission, monitoring,
+  termination, deletion, etc. Sub-task control is similar to
+  interaction \texttt{b}, except that information about the parent
+  task which submitted a sub-task is usually available and used for
+  additional control. For instance, the parent task may wait for all
+  its sub-tasks to finish before finishing, and conversely all the
+  sub-tasks may be killed if the parent task is killed.
+\item Pool-agent: specific to the pool architecture described in
+  Section~\ref{sec:pool}. This is an interaction used when agents
+  retrieve work from a central pool. It covers agent registration and
+  de-registration to the pool, protocols to send work from the pool to
+  the agent, mechanisms to update work status, and so on. A similar
+  type of interaction is used in pilot-job
+  systems~\cite{turilli2015comprehensive} and other agent-based
+  computing models.
+\item Workflow conversion: translation from one workflow language to
+  another one. This interaction may not be available or implementable
+  for every workflow language. It has been developed mostly for
+  well-structured and relatively simple workflow language such as between GWorkflowDL
+  and Scufl~\cite{OLAB-09} and between the 4 languages that were involved in the
+  SHIWA initiative. In SHIWA, workflow conversion is done through an
+  intermediate representation, the Interoperable Workflow Intermediate
+  Representation~\cite{plankensteiner-montagnat-etal:2011}, which
+  allows to convert among $n$ workflow languages using $2n$
+  interactions instead of $n^2$.
+\end{enumerate}
+
+\subsection{Tight integration}
+
+See Figure~\subref*{archi:tight}. The workflow engine is tightly
+integrated with the science gateway, which means that it is deployed
+on the same machine and potentially shares code, libraries and other
+software components with the science gateway. For instance, the
+workflow engine might be a portlet in a Liferay portal or a controller
+in a Ruby on Rails application. The workflow engine and the science
+gateway usually share a database where application, users and other
+resources are stored. In this model, task and data control are both
+initiated from the science gateway. Interactions \texttt{b} and
+\texttt{c$_2$} are initiated from the workflow engine while
+\texttt{c$_1$} comes from other parts of the science gateway, for
+instance a data management interface. As in any other model, the
+installation of new workflows in the science gateway (\texttt{a$_2$})
+and infrastructure (\texttt{a$_1$}) is done by an administrator, for
+security reasons. This is the model adopted in the Catania Science
+Gateway Framework~\cite{Ardizzone2012} (see specific documentation on
+workflows\footnote{\url{http://bit.ly/1oQrzvQ}}), in the Distributed
+application runtime environment (DARE~\cite{maddineni2012distributed})
+and in LONI Pipeline Environment~\cite{dinov2009efficient}.
+%The Catania Science Gateway Framework would be a good illustration here (workflow engine is a Java portlet), but I can't find a suitable figure about it.
+%CIPRES~\cite{miller2010creating} (on which the Neuroscience gateway
+%described in~\cite{sivagnanam2015early} is based) % \note{Tristan}{There is no workflow engine in there.}
+
+\subsection{Service invocation}
+
+\begin{figure*}
+\centering
+\def\svgwidth{1.5\columnwidth}
+\input{figures/VIP.pdf_tex}
+\caption{Architecture of the Virtual Imaging Platform (service
+  invocation).  Step \texttt{3} corresponds to interaction \texttt{d}
+  on Figure~\ref{fig:architectures}, step \texttt{2} corresponds to
+  interaction \texttt{c$_1$}, and step \texttt{4} maps to interactions
+  \texttt{b} and \texttt{c$_2$} (in VIP, workflow data transfers are
+  performed by the tasks and embedded in their descriptions).}
+\label{fig:vip-architecture}
+\end{figure*}
+
+See Figure~\subref*{archi:service}. The workflow engine is available
+externally to the science gateway, in a service. The science gateway
+controls the service through a specific interaction (\texttt{d}) that
+might be implemented as a web-service call (e.g. RESTful or SOAP), as
+a command-line or as any other method that offers a well-defined
+interface to the workflow engine. The workflow engine might be invoked
+either as a black box that completely masks the infrastructure and
+tasks, or as a white box that allows for some interaction with
+them. The workflow engine is responsible for controlling the tasks on
+the infrastructure (\texttt{b}) and for performing the required data
+transfers to execute them (\texttt{c$_2$}). User data is usually
+managed through the science gateway (\texttt{c$_1$}), although it
+might as well be delivered by the workflow engine directly to the
+user.
+
+This architecture is largely adopted, in systems such as Apache
+Airvata~\cite{marru2011apache}, Vine
+Toolkit~\cite{DBLP:journals/scpe/SzejnfeldDKKKKLPTWDNW10}, Virtual
+Imaging Platform~\cite{GLAT-13}, the system
+in~\cite{wu2010accelerating} and the WS-PGRADE/gUSE
+framework~\cite{Kacsuk2012}. The integration between the
+HubZero science gateway and Pegasus workflow engine performed
+in~\cite{CPE:CPE3257} also uses a service architecture where the
+\texttt{d} interaction is implemented as a set of calls to Pegasus
+command-line tools (e.g. pegasus-status, pegasus-statistics,
+etc). Figure~\ref{fig:vip-architecture} shows the architecture of the
+Virtual Imaging Platform, where the MOTEUR workflow
+engine~\cite{GLAT-08i} is invoked through a service.
+
+\subsection{Sub-tasking}
+
+See Figure~\subref*{archi:sub-task}. The workflow engine is a particular
+task that can submit sub-tasks to the science gateway through
+interaction \texttt{e}. The workflow engine keeps track of the
+dependencies between the sub-tasks, but their execution is delegated to
+the science gateway that executes them on the infrastructure through
+interaction \texttt{b}. Although the science gateway has no global
+vision of the workflow, it can keep track of the sub-tasks submitted
+by a given task, for instance to be able to cancel them when the task
+is canceled. The science gateway may also implement mechanisms to
+facilitate the handling of task dependencies, for instance basic
+dependency lists as available in most batch managers (see for instance
+attribute \texttt{depend} of option \texttt{-W} in
+Torque\footnote{\url{http://docs.adaptivecomputing.com}}).
+
+The science gateway also transfers both user and workflow data across
+the infrastructure, so that interactions \texttt{c$_1$} and
+\texttt{c$_2$} are both covered by \texttt{c}. In practice, both
+\texttt{c$_1$} and \texttt{c$_2$} can be implemented using the same
+pieces of code.
+
+The sub-tasking architecture is implemented in CBRAIN~\cite{SHER-14}
+where it is used to integrate the FSL toolkit~\cite{Jenkinson2012782}
+and the PSOM workflow engine~\cite{bellec2012pipeline}.
+
+The CBRAIN-FSL integration allows to leverage FSL pipelines written in
+low-level workflow languages (Linux executables and scripts) that
+submit tasks uniformly through a specific tool called fsl\_sub. It is
+diagrammed in Figure~\ref{fig:cbrain-fsl-architecture}, where the
+science gateway is represented by components \texttt{CBRAIN portal}
+and \texttt{CBRAIN execution server} and the infrastructure consists
+of \texttt{Storage servers} and \texttt{Computing cluster with shared
+  file system}.
+
+The CBRAIN-PSOM integration~\cite{GLAT-16} is shown on
+Figure~\ref{fig:cbrain-psom-architecture}. The PSOM workflow engine
+adopts a pilot-job architecture~\cite{turilli2015comprehensive} where
+a master coordinates workflow execution by submitting workers and
+establishing direct communication channels with them. Note how this
+peculiar execution model is totally supported by the sub-tasking
+architecture.
+
+\begin{figure*}
+\centering
+\subfloat[CBRAIN-FSL integration.]{
+\def\svgwidth{\columnwidth}
+\input{figures/CBRAIN-FSL.pdf_tex}
+\label{fig:cbrain-fsl-architecture}
+} \hfill \subfloat[CBRAIN-PSOM integration. Figure
+reproduced from~\cite{GLAT-16}]{ \def\svgwidth{\columnwidth}
+  \input{figures/CBRAIN-PSOM.pdf_tex}
+\label{fig:cbrain-psom-architecture}
+}
+\caption{CBRAIN architecture for workflow engine integration
+  (sub-tasking).  1: User sends data and workflow execution request to
+  storage server(s) and CBRAIN portal. 2: CBRAIN portal sends
+  execution request to execution server on cluster. 3: Execution
+  server transfers data from storage server(s). 4-5: Execution server
+  starts workflow engine (FSL tool or PSOM master) via task
+  scheduler. 6: Workflow engine submits sub-tasks to execution server
+  (FSL tasks or PSOM agents). 7-8: Execution server starts sub-tasks
+  through task scheduler. 8': (PSOM only) PSOM master and workers
+  execute workflow. 9. Execution server transfers results to storage
+  server(s). Interaction \texttt{b} in Figure~\ref{fig:architectures}
+  is implemented by steps \texttt{4}, \texttt{5}, \texttt{7} and
+  \texttt{8} (regular interactions with batch manager). Interaction
+  \texttt{c} is implemented by steps \texttt{3} and
+  \texttt{9}. Interaction \texttt{e} is implemented by step \texttt{6}
+  (for FSL: through a modified version of the fsl\_sub script
+  available in \url{https://github.com/aces/cbrain-plugins-neuro}; for
+  PSOM: through a specific development in PSOM).}
+\label{fig:cbrain-sub-tasking}
+\end{figure*}
+
+\subsection{Pool model}
+\label{sec:pool}
+See Figure~\subref*{archi:agent}. Workflows are submitted by the science
+gateway to a pool through interaction \texttt{d}. Agents connect to
+the pool asynchronously to retrieve and execute workflows through
+interaction \texttt{f}. Agents may be started according to various
+policies, for instance to ensure load balancing. Agents may wrap
+different types of workflow engines. Workflow engine controls tasks
+and data on the infrastructure through interactions \texttt{b} and
+\texttt{c$_2$}, science gateway transfers user data through
+interaction \texttt{c$_1$}, and administrator installs workflows
+through interaction \texttt{a$_1$} and \texttt{a$_2$}.
+
+The pool model was implemented in the SHIWA pool~\cite{ROGE-13}
+diagrammed in Figure~\ref{fig:shiwa-pool-architecture}. In this
+Figure, interaction \texttt{d} of Figure~\subref*{archi:agent} is
+implemented in 3 different types of calls to the pool: workflow
+submission (\texttt{1} \& \texttt{2}), workflow status retrieval
+(\texttt{a} and \texttt{b}), and workflow retrieval (\texttt{13} \&
+\texttt{14}). In Figure ~\ref{fig:shiwa-pool-architecture},
+interaction \texttt{f} of Figure~\subref*{archi:agent} is implemented
+through 2 types of calls: workflow instance retrieval (\texttt{4},
+\texttt{5} and \texttt{6}), and workflow instance update (\texttt{11}
+and \texttt{12}). Workflow instance retrieval is used by the agents to
+fetch work from the pool. Workflow instance update is used by the
+agents to update workflow statuses. In the SHIWA pool, agents can wrap
+different types of workflow engines to execute workflows expressed
+with different languages. Calls \texttt{0}, \texttt{7}, \texttt{8} and
+\texttt{9} on Figure~\ref{fig:shiwa-pool-architecture} are used by
+workflow engine plugins to declare their supported language and launch
+engines, and by workflow engines to report their status to the
+agent. These calls are specific to the SHIWA Pool implementation of
+the \texttt{agent} component and therefore have no corresponding
+representation in Figure~\subref*{archi:agent}.
+
+\begin{figure*}
+\centering
+\includegraphics[width=1.5\columnwidth]{figures/pool-interactions.pdf}
+\caption{Architecture of the SHIWA pool. Circle-terminated arrows indicate messages
+  that are broadcast to all pool clients. Figure reproduced
+  from~\cite{ROGE-13}.}
+\label{fig:shiwa-pool-architecture}
+\end{figure*}
+
+\subsection{Nested workflows}
+
+See Figure~\subref*{archi:nested}. In nested workflows
+(Figure~\subref*{archi:nested}-Left), a task of a \emph{parent}
+workflow executed by the \emph{parent} engine is itself a workflow,
+called \emph{child} workflow, that is executed by a \emph{child}
+engine. The parent and child engines might use different workflow
+description languages. They might also execute workflows on different
+infrastructures. The parent workflow is also called meta-workflow. The
+science gateway communicates with the parent engine through abstract
+interaction \texttt{*$_1$}. The science gateway also communicates with
+the infrastructure to transfer user data through abstract interactions
+\texttt{*$_2$} and \texttt{*$_3$}. Both workflow engines communicate
+with the infrastructure through abstract interactions \texttt{*$_4$}
+and \texttt{*$_5$}. The parent engine communicates with the child
+engine through abstract interaction \texttt{*$_6$}. Administrator
+installs workflows through interactions \texttt{a$_1$} and
+\texttt{a$_2$}.
+
+Nested workflows are abstract architectural patterns that can be
+instantiated in the various architectures described previously. We
+focus on instantiation with the service invocation model
+(Figure~\subref*{archi:nested}-Right) as this is the most used
+architecture. In such an instantiation, we assume that the parent and
+child workflow engines are distinct pieces of software that require
+different workflow services invoked by distinct \texttt{d}
+interactions. If this is not the case, then workflow services can be
+collapsed in a single one with a \texttt{d} interaction with
+itself. Workflow engines communicate with infrastructures using
+\texttt{b} and \texttt{c$_2$}. Science gateway transfers user data to
+infrastructures using \texttt{c$_1$} interactions.
+
+Nested workflows have long been available in workflow engines, for
+instance in the Taverna workbench~\cite{oinn2004taverna}. They are
+also used implicitly in several platforms where workflow engines are
+wrapped in workflow tasks as any other command-line tool. Nested
+workflows were notably used by the SHIWA Science Gateway to implement
+so-called Coarse-Grained workflow
+interoperability~\cite{terstyanszky2014enabling}, i.e. to integrate
+various workflow engines in a consistent
+platform. Figure~\ref{fig:shiwa-architecture} shows the architecture
+used in the SHIWA Simulation Platform for nested workflow execution
+with service invocation. 
+\begin{figure*}
+\centering
+\includegraphics[width=1.5\columnwidth]{figures/shiwa-science-gateway.pdf}
+\caption{Nested workflow execution through SHIWA Science Gateway. The
+  parent workflow engine is \texttt{WS-PGRADE}, invoked as a service
+  in the Science Gateway (step \texttt{1}, interaction \texttt{d} on
+  Figure 2(V)-Right. Ten different child engines can be used by nested
+  workflows, invoked through the \texttt{Submission service} (step
+  \texttt{2}, interaction \texttt{d}). Each of these engines can
+  submit tasks and transfer data to a distributed computing
+  infrastructure (\texttt{DCI}, step \texttt{3}, interactions
+  \texttt{b} and \texttt{c$_2$}). Data interactions (\texttt{c}) and
+  application porting ones (\texttt{a}) are not represented. Figure
+  reproduced from~\cite{terstyanszky2014enabling} with permission of
+  the first author.}
+\label{fig:shiwa-architecture}
+\end{figure*}
+
+\subsection{Workflow import}
+
+See Figure~\subref*{archi:import}. This is an abstract model that we
+instantiate with the service invocation architecture for
+consistency. Workflows are integrated in the science gateway through
+format conversion from a native format to the science gateway
+format. Workflow import is usually an offline process which is not
+involved in workflow execution. Workflow import was implemented in the SHIWA
+Simulation Platform through the IWIR language that provided a common
+language for portability across grid workflow
+systems~\cite{plankensteiner-prodan-etal:2013}.
+
+\todo{Add description of a real system}
+
+\todo{Refer to ``From the desktop to the grid: scalable bioinformatics via workflow conversion Luis de la GarzaEmail author, Johannes Veit, Andras Szolek, Marc Röttig, Stephan Aiche, Sandra Gesing, Knut Reinert and Oliver Kohlbacher''}
+
 \section{Evaluation}
@@
 criteria and metrics are explained hereafter.
+
+\begin{table*}
+\centering
+\begin{tabular}{rcccccc}
+                                     & \textbf{Tight}
+                                     & \textbf{Service}
+                                     & \textbf{Sub-task}
+                                     & \textbf{Pool}
+                                     & \textbf{Nested}
+                                     & \textbf{Import} \\
+\cellcolor[HTML]{EEEEEE}\textbf{Integration effort}& \multicolumn{6}{l}{\cellcolor[HTML]{EEEEEE}}\\
+  Total components -- \texttt{I$_1$} & \cellcolor[HTML]{99FF99}2
+                                     & \cellcolor[HTML]{99E899}3
+                                     & \cellcolor[HTML]{99FF99}2
+                                     & \cellcolor[HTML]{99D199}4
+                                     & \cellcolor[HTML]{99BB99}5
+                                     & \cellcolor[HTML]{99E899}3\\
+Total interactions -- \texttt{I$_2$} & \cellcolor[HTML]{99FF99}5
+                                     & \cellcolor[HTML]{99F399}6
+                                     & \cellcolor[HTML]{99FF99}5
+                                     & \cellcolor[HTML]{99E899}7
+                                     & \cellcolor[HTML]{99BB99}11
+                                     & \cellcolor[HTML]{99E899}7\\
+\textbf{Total} (total software pieces) & \cellcolor[HTML]{99FF99}\textbf{7}
+                                     & \cellcolor[HTML]{99EF99}\textbf{9}
+                                     & \cellcolor[HTML]{99FF99}\textbf{7}
+                                     & \cellcolor[HTML]{99E099}\textbf{11}
+                                     & \cellcolor[HTML]{99BB99}\textbf{16}
+                                     & \cellcolor[HTML]{99E899}\textbf{10}\\
+\cellcolor[HTML]{EEEEEE}\textbf{Robustness}& \multicolumn{6}{l}{\cellcolor[HTML]{EEEEEE}}\\
+Specific components -- \texttt{R$_1$} & \cellcolor[HTML]{99FF99}0
+                                     & \cellcolor[HTML]{99DD99}1
+                                     & \cellcolor[HTML]{99FF99}0
+                                     & \cellcolor[HTML]{99BB99}2
+                                     & \cellcolor[HTML]{99BB99}2
+                                     & \cellcolor[HTML]{99DD99}1\\
+  Specific interactions -- \texttt{R$_2$} & \cellcolor[HTML]{99F199}2
+                                     & \cellcolor[HTML]{99E399}3
+                                     & \cellcolor[HTML]{99FF99}1
+                                     & \cellcolor[HTML]{99D699}4
+                                     & \cellcolor[HTML]{99BB99}6
+                                     & \cellcolor[HTML]{99E399}3\\
+  \textbf{Total} (execution software pieces)& \cellcolor[HTML]{99F599}\textbf{2}
+                                     & \cellcolor[HTML]{99E199}\textbf{4}
+                                     & \cellcolor[HTML]{99FF99}\textbf{1}
+                                     & \cellcolor[HTML]{99CE99}\textbf{6}
+                                     & \cellcolor[HTML]{99BB99}\textbf{8}
+                                     & \cellcolor[HTML]{99E199}\textbf{4}\\
+\cellcolor[HTML]{EEEEEE}\textbf{Extensibility}& \multicolumn{6}{l}{\cellcolor[HTML]{EEEEEE}}\\
+  New engine type -- \texttt{E$_1$}  & \cellcolor[HTML]{99D899}3
+                                     & \cellcolor[HTML]{99C499}4
+                                     & \cellcolor[HTML]{99EB99}2
+                                     & \cellcolor[HTML]{99D899}3
+                                     & \cellcolor[HTML]{99BB99}4.5
+                                     & \cellcolor[HTML]{99FF99}1\\
+New engine version -- \texttt{E$_2$} & \cellcolor[HTML]{99BB99}1
+                                     & \cellcolor[HTML]{99FF99}0
+                                     & \cellcolor[HTML]{99BB99}1
+                                     & \cellcolor[HTML]{99FF99}0
+                                     & \cellcolor[HTML]{99FF99}0
+                                     & \cellcolor[HTML]{99BB99}1\\
+  New workflow -- \texttt{E$_3$} & \cellcolor[HTML]{99FF99}2
+                                     & \cellcolor[HTML]{99FF99}2
+                                     & \cellcolor[HTML]{99FF99}2
+                                     & \cellcolor[HTML]{99FF99}2
+                                     & \cellcolor[HTML]{99FF99}2
+                                     & \cellcolor[HTML]{99BB99}3\\
+New infrastructure -- \texttt{E$_4$} & \cellcolor[HTML]{99E899}3
+                                     & \cellcolor[HTML]{99E899}3
+                                     & \cellcolor[HTML]{99FF99}2
+                                     & \cellcolor[HTML]{99E899}3
+                                     & \cellcolor[HTML]{99BB99}5
+                                     & \cellcolor[HTML]{99E899}3\\
+  \textbf{Total} (difficulty to extend) & \cellcolor[HTML]{99E099}\textbf{9}
+                                     & \cellcolor[HTML]{99E099}\textbf{9}
+                                     & \cellcolor[HTML]{99FF99}\textbf{7}
+                                     & \cellcolor[HTML]{99EF99}\textbf{8}
+                                     & \cellcolor[HTML]{99BB99}\textbf{11.5}
+                                     & \cellcolor[HTML]{99EF99}\textbf{8}\\
+\cellcolor[HTML]{EEEEEE}\textbf{Scalability}& \multicolumn{6}{l}{\cellcolor[HTML]{EEEEEE}}\\
+Multiple engine instances -- \texttt{S$_1$}& \cellcolor[HTML]{99BB99}2
+                                     & \cellcolor[HTML]{99DD99}1
+                                     & \cellcolor[HTML]{99FF99}0
+                                     & \cellcolor[HTML]{99FF99}0
+                                     & \cellcolor[HTML]{99DD99}1
+                                     & \cellcolor[HTML]{99DD99}1\\
+Distributed engines -- \texttt{S$_2$}& \cellcolor[HTML]{99BB99}1
+                                     & \cellcolor[HTML]{99BB99}1
+                                     & \cellcolor[HTML]{99BB99}1
+                                     & \cellcolor[HTML]{99BB99}1
+                                     & \cellcolor[HTML]{99FF99}0
+                                     & \cellcolor[HTML]{99BB99}1\\
+Task scheduling -- \texttt{S$_3$}    & \cellcolor[HTML]{99FF99}0
+                                     & \cellcolor[HTML]{99FF99}0
+                                     & \cellcolor[HTML]{99BB99}1
+                                     & \cellcolor[HTML]{99FF99}0
+                                     & \cellcolor[HTML]{99BB99}1
+                                     & \cellcolor[HTML]{99FF99}0\\
+\textbf{Total} (scalability issues)  & \cellcolor[HTML]{99BB99}\textbf{3}
+                                     & \cellcolor[HTML]{99DD99}\textbf{2}
+                                     & \cellcolor[HTML]{99DD99}\textbf{2}
+                                     & \cellcolor[HTML]{99FF99}\textbf{1}
+                                     & \cellcolor[HTML]{99DD99}\textbf{2}
+                                     & \cellcolor[HTML]{99DD99}\textbf{2}\\
+\cellcolor[HTML]{EEEEEE}\textbf{Specific features}& \multicolumn{6}{l}{\cellcolor[HTML]{EEEEEE}}\\
+  Meta-workflow  -- \texttt{O$_1$}    & \cellcolor[HTML]{99BB99}1
+                                     & \cellcolor[HTML]{99BB99}1
+                                     & \cellcolor[HTML]{99BB99}1
+                                     & \cellcolor[HTML]{99BB99}1
+                                     & \cellcolor[HTML]{99FF99}0
+                                     & \cellcolor[HTML]{99FF99}0\\
+  Fine-grained debugging -- \texttt{O$_2$}   & \cellcolor[HTML]{99FF99}0
+                                     & \cellcolor[HTML]{99FF99}0
+                                     & \cellcolor[HTML]{99BB99}1
+                                     & \cellcolor[HTML]{99FF99}0
+                                     & \cellcolor[HTML]{99BB99}1
+                                     & \cellcolor[HTML]{99FF99}0\\
+  \textbf{Total} (missing features) & \cellcolor[HTML]{99DD99}\textbf{1}
+                                     & \cellcolor[HTML]{99DD99}\textbf{1}
+                                     & \cellcolor[HTML]{99BB99}\textbf{2}
+                                     & \cellcolor[HTML]{99DD99}\textbf{1}
+                                     & \cellcolor[HTML]{99DD99}\textbf{1}
+                                     & \cellcolor[HTML]{99FF99}\textbf{0}\\
+\end{tabular}
+
+\caption{Architecture evaluation. Lower values (brighter colors) indicate better performance. On each row, metric values are
+  normalized between 0 (best value) and 1 (worst value) to determine the
+  color of the corresponding table cell. The normalized metric value $m'$ is
+  defined as
+  $\frac{m-m_{\mathrm{min}}}{m_{\mathrm{min}}-m_{\mathrm{max}}}$ where
+  $m$ is the initial metric value, $m_{\mathrm{min}}$ and $m_{\mathrm{max}}$
+  are the minimal and maximal values among all architectures. The RGB hexadecimal color code of the cell
+  is \texttt{\#99XX99}, where X=F-4m'. \todo{Use 6. Round to the nearest integer.}}
+\label{table:evaluation}
+\end{table*}
 
@@
 why we keep it in the metric. The two metrics are summed to obtain the
-total score for this criterion, which measures the total number of
-software pieces to develop.
+total score for integration effort, which measures the total number of
+software pieces to develop (pieces include interactions and
+components).
 
@@
 from the infrastructure (e.g. unavailable data or terminated tasks) or
-workflows (e.g., wrong user input or application crashes) are not
+workflows (e.g., wrong user input or application errors) are not
 covered since they do not stem directly from the software
@@
 components and interactions that are specific to workflow execution,
-i.e. not reused by other functions:
+which assumes that complex architectures tend to be prone to failure:
 \begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
@@
 the architecture diagrams in
-Figure~\ref{fig:architectures}. \texttt{R$_1$} and \texttt{R$_2$}
-assume that complex architectures tend to be prone to
-failure. \texttt{R$_1$} and \texttt{R$_2$}  are summed to obtain the total score for this
+Figure~\ref{fig:architectures}. \texttt{R$_1$} and \texttt{R$_2$}  are summed to obtain the total score for this
 criterion, which measures the total number of software pieces that are
@@
 elements in the architecture. It is determined as the number of
-interactions and components to modify when a new element is
-added. Modification of a component is required when its code needs to
-be modified or recompiled (science gateway or workflow service), or
-when a new piece of software has to be installed (infrastructure
-only). Modification of an interaction is deemed necessary when the
-parameters involved in this interaction are modified.  Extensibility
-breaks down in 4 metrics depending on the type of element that has to
-be added or replaced:
+interactions and components that need to be modified when a new
+element is added. Modification of a component is required when its
+code needs to be updated or recompiled (science gateway or workflow
+service), or when a new piece of software has to be installed
+(infrastructure only). Modification of an interaction is deemed
+necessary when the parameters involved in this interaction are
+modified.  Extensibility breaks down in 4 metrics depending on the
+type of element that has to be added or replaced:
 \begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
@@
   engine share the same interface, i.e. they can be invoked using the
-  same software. When this is not the case, the different versions
-  have to be considered as different engine types. 
+  same software. When this is not the case, the different versions are
+  considered as different engine types.
 \item New workflow (\texttt{E$_3$}): adding a new workflow is a very
@@
 \item Fine-grained debugging (\texttt{O$_2$}): availability of
-  fine-grained debugging information about workflow tasks. When such
-  information is available, workflow is seen as a white
-  box. Fine-grained information about workflow tasks is required to
-  properly troubleshoot workflow executions.  \texttt{O$_2$}=0 when
-  such information is easy to access, \texttt{O$_2$}=1 otherwise.
+  fine-grained debugging information about workflow tasks (white-box
+  workflow).  Fine-grained information about workflow tasks is
+  required to properly troubleshoot workflow executions.
+  \texttt{O$_2$}=0 when such information is easy to access,
+  \texttt{O$_2$}=1 otherwise.
 \end{itemize}
@@
 
-The different architectures described in
-Figure~\ref{fig:architectures} are evaluated along these metrics in
-the remainder of this Section.
+The architectures described in Figure~\ref{fig:architectures} are
+evaluated along these metrics in the remainder of this Section.
 
@@
 
-\paragraph{Integration} This architecture does not require any
+\paragraph{Integration effort} This architecture does not require any
 component in addition to the science gateway and infrastructure
@@
 instance of the science gateway, which is in general not possible
-(\texttt{S$_1$}=2). Distributed engines are not available by default
+(\texttt{S$_1$}=2)\note{Tristan}{What-about load-balancing between web
+  server instances?}. Distributed engines are not available by default
 (\texttt{S$_2$}=1). The scheduling of tasks on the infrastructure is
@@
 
-\paragraph{Integration} Service invocation requires a workflow service
+\paragraph{Integration effort} Service invocation requires a workflow service
 in addition to the science gateway and infrastructure
@@
 component (\texttt{E$_2$}=0). We consider that updating the engine
-version in a workflow service is transparent since the only goal of
-this component is to wrap the engine.  New workflows are added in the
-science gateway or in the workflow engine through interactions
-\texttt{a$_1$} and \texttt{a$_2$} (\texttt{E$_3$}=2). Adding a new
-type of infrastructure requires updates in interactions \texttt{b},
-\texttt{$c_1$} and \texttt{$c_2$} (\texttt{E$_4$}=3).
+version in a workflow service does not require code updates in the
+service since the only goal of this component is to wrap the engine.
+New workflows are added in the science gateway or in the workflow
+engine through interactions \texttt{a$_1$} and \texttt{a$_2$}
+(\texttt{E$_3$}=2). Adding a new type of infrastructure requires
+updates in interactions \texttt{b}, \texttt{$c_1$} and \texttt{$c_2$}
+(\texttt{E$_4$}=3).
 
@@
 
-\paragraph{Integration} The sub-task architecture requires only 2
+\paragraph{Integration effort} The sub-task architecture requires only 2
 components (\texttt{I$_1$}=2).  It involves 5 interactions:
@@
 
-\paragraph{Integration} The pool model requires a workflow pool and an
+\paragraph{Integration effort} The pool model requires a workflow pool and an
 agent in addition to the science gateway and infrastructure
@@
 
-\paragraph{Integration} Setting up a nested workflow architecture with
+\paragraph{Integration effort} Setting up a nested workflow architecture with
 service invocation requires a science gateway, 2 workflow services and
 2 infrastructures (\texttt{I$_1$=5}). The architecture involves 11
-interactions (\texttt{I$_2$=11}): \texttt{a$_1$} (twice: once for each
+interactions (\texttt{I$_2$=11}): \texttt{a$_1$} (appears twice: once for each
 infrastructure), \texttt{a$_2$}, \texttt{b} (twice: once for each
@@
 
-\paragraph{Integration} Workflow import with service invocation
+\paragraph{Integration effort} Workflow import with service invocation
 requires the same components as for service invocation
 (\texttt{I$_1$}=3). It requires an additional \texttt{g} interaction
-for workflow import (\texttt{I$_1$}=7)
+for workflow import (\texttt{I$_1$}=7).
 
@@
 
-\begin{table*}
-\centering
-\begin{tabular}{rcccccc}
-                                     & \textbf{Tight}
-                                     & \textbf{Service}
-                                     & \textbf{Sub-task}
-                                     & \textbf{Pool}
-                                     & \textbf{Nested}
-                                     & \textbf{Import} \\
-\multicolumn{7}{c}{\cellcolor[HTML]{EEEEEE}\textbf{Integration effort}}\\
-  Total components -- \texttt{I$_1$} & \cellcolor[HTML]{99FF99}2
-                                     & \cellcolor[HTML]{99E899}3
-                                     & \cellcolor[HTML]{99FF99}2
-                                     & \cellcolor[HTML]{99D199}4
-                                     & \cellcolor[HTML]{99BB99}5
-                                     & \cellcolor[HTML]{99E899}3\\
-Total interactions -- \texttt{I$_2$} & \cellcolor[HTML]{99FF99}5
-                                     & \cellcolor[HTML]{99F399}6
-                                     & \cellcolor[HTML]{99FF99}5
-                                     & \cellcolor[HTML]{99E899}7
-                                     & \cellcolor[HTML]{99BB99}11
-                                     & \cellcolor[HTML]{99E899}7\\
-\textbf{Total} (total software pieces) & \cellcolor[HTML]{99FF99}\textbf{7}
-                                     & \cellcolor[HTML]{99EF99}\textbf{9}
-                                     & \cellcolor[HTML]{99FF99}\textbf{7}
-                                     & \cellcolor[HTML]{99E099}\textbf{11}
-                                     & \cellcolor[HTML]{99BB99}\textbf{16}
-                                     & \cellcolor[HTML]{99E899}\textbf{10}\\
-  \multicolumn{7}{c}{\cellcolor[HTML]{EEEEEE}\textbf{Robustness of workflow execution}}\\
-Specific components -- \texttt{R$_1$} & \cellcolor[HTML]{99FF99}0
-                                     & \cellcolor[HTML]{99DD99}1
-                                     & \cellcolor[HTML]{99FF99}0
-                                     & \cellcolor[HTML]{99BB99}2
-                                     & \cellcolor[HTML]{99BB99}2
-                                     & \cellcolor[HTML]{99DD99}1\\
-  Specific interactions -- \texttt{R$_2$} & \cellcolor[HTML]{99F199}2
-                                     & \cellcolor[HTML]{99E399}3
-                                     & \cellcolor[HTML]{99FF99}1
-                                     & \cellcolor[HTML]{99D699}4
-                                     & \cellcolor[HTML]{99BB99}6
-                                     & \cellcolor[HTML]{99E399}3\\
-  \textbf{Total} (specific software pieces)& \cellcolor[HTML]{99F599}\textbf{2}
-                                     & \cellcolor[HTML]{99E199}\textbf{4}
-                                     & \cellcolor[HTML]{99FF99}\textbf{1}
-                                     & \cellcolor[HTML]{99CE99}\textbf{6}
-                                     & \cellcolor[HTML]{99BB99}\textbf{8}
-                                     & \cellcolor[HTML]{99E199}\textbf{4}\\
-\multicolumn{7}{c}{\cellcolor[HTML]{EEEEEE}\textbf{Extensibility}}\\
-  New engine type -- \texttt{E$_1$}  & \cellcolor[HTML]{99D899}3
-                                     & \cellcolor[HTML]{99C499}4
-                                     & \cellcolor[HTML]{99EB99}2
-                                     & \cellcolor[HTML]{99D899}3
-                                     & \cellcolor[HTML]{99BB99}4.5
-                                     & \cellcolor[HTML]{99FF99}1\\
-New engine version -- \texttt{E$_2$} & \cellcolor[HTML]{99BB99}1
-                                     & \cellcolor[HTML]{99FF99}0
-                                     & \cellcolor[HTML]{99BB99}1
-                                     & \cellcolor[HTML]{99FF99}0
-                                     & \cellcolor[HTML]{99FF99}0
-                                     & \cellcolor[HTML]{99BB99}1\\
-  New workflow -- \texttt{E$_3$} & \cellcolor[HTML]{99FF99}2
-                                     & \cellcolor[HTML]{99FF99}2
-                                     & \cellcolor[HTML]{99FF99}2
-                                     & \cellcolor[HTML]{99FF99}2
-                                     & \cellcolor[HTML]{99FF99}2
-                                     & \cellcolor[HTML]{99BB99}3\\
-New infrastructure -- \texttt{E$_4$} & \cellcolor[HTML]{99E899}3
-                                     & \cellcolor[HTML]{99E899}3
-                                     & \cellcolor[HTML]{99FF99}2
-                                     & \cellcolor[HTML]{99E899}3
-                                     & \cellcolor[HTML]{99BB99}5
-                                     & \cellcolor[HTML]{99E899}3\\
-  \textbf{Total} (difficulty to extend) & \cellcolor[HTML]{99E099}\textbf{9}
-                                     & \cellcolor[HTML]{99E099}\textbf{9}
-                                     & \cellcolor[HTML]{99FF99}\textbf{7}
-                                     & \cellcolor[HTML]{99EF99}\textbf{8}
-                                     & \cellcolor[HTML]{99BB99}\textbf{11.5}
-                                     & \cellcolor[HTML]{99EF99}\textbf{8}\\
-\multicolumn{7}{c}{\cellcolor[HTML]{EEEEEE}\textbf{Scalability}}\\
-Multiple engine instances -- \texttt{S$_1$}& \cellcolor[HTML]{99BB99}2
-                                     & \cellcolor[HTML]{99DD99}1
-                                     & \cellcolor[HTML]{99FF99}0
-                                     & \cellcolor[HTML]{99FF99}0
-                                     & \cellcolor[HTML]{99DD99}1
-                                     & \cellcolor[HTML]{99DD99}1\\
-Distributed engines -- \texttt{S$_2$}& \cellcolor[HTML]{99BB99}1
-                                     & \cellcolor[HTML]{99BB99}1
-                                     & \cellcolor[HTML]{99BB99}1
-                                     & \cellcolor[HTML]{99BB99}1
-                                     & \cellcolor[HTML]{99FF99}0
-                                     & \cellcolor[HTML]{99BB99}1\\
-Task scheduling -- \texttt{S$_3$}    & \cellcolor[HTML]{99FF99}0
-                                     & \cellcolor[HTML]{99FF99}0
-                                     & \cellcolor[HTML]{99BB99}1
-                                     & \cellcolor[HTML]{99FF99}0
-                                     & \cellcolor[HTML]{99BB99}1
-                                     & \cellcolor[HTML]{99FF99}0\\
-\textbf{Total} (scalability issues)  & \cellcolor[HTML]{99BB99}\textbf{3}
-                                     & \cellcolor[HTML]{99DD99}\textbf{2}
-                                     & \cellcolor[HTML]{99DD99}\textbf{2}
-                                     & \cellcolor[HTML]{99FF99}\textbf{1}
-                                     & \cellcolor[HTML]{99DD99}\textbf{2}
-                                     & \cellcolor[HTML]{99DD99}\textbf{2}\\
-\multicolumn{7}{c}{\cellcolor[HTML]{EEEEEE}\textbf{Specific features}}\\
-  Meta-workflow  -- \texttt{O$_1$}    & \cellcolor[HTML]{99BB99}1
-                                     & \cellcolor[HTML]{99BB99}1
-                                     & \cellcolor[HTML]{99BB99}1
-                                     & \cellcolor[HTML]{99BB99}1
-                                     & \cellcolor[HTML]{99FF99}0
-                                     & \cellcolor[HTML]{99FF99}0\\
-  Fine-grained debugging -- \texttt{O$_2$}   & \cellcolor[HTML]{99FF99}0
-                                     & \cellcolor[HTML]{99FF99}0
-                                     & \cellcolor[HTML]{99BB99}1
-                                     & \cellcolor[HTML]{99FF99}0
-                                     & \cellcolor[HTML]{99BB99}1
-                                     & \cellcolor[HTML]{99FF99}0\\
-  \textbf{Total} (missing features) & \cellcolor[HTML]{99DD99}\textbf{1}
-                                     & \cellcolor[HTML]{99DD99}\textbf{1}
-                                     & \cellcolor[HTML]{99BB99}\textbf{2}
-                                     & \cellcolor[HTML]{99DD99}\textbf{1}
-                                     & \cellcolor[HTML]{99DD99}\textbf{1}
-                                     & \cellcolor[HTML]{99FF99}\textbf{0}\\
-\end{tabular}
-
-\caption{Architecture evaluation. Lower values (brighter colors) indicate better performance. On each row, metric values are
-  normalized between 0 (best value) and 1 (worst value) to determine the
-  color of the corresponding table cell. More precisely, the normalized metric value $m'$ is
-  defined as
-  $\frac{m-m_{\mathrm{min}}}{m_{\mathrm{min}}-m_{\mathrm{max}}}$ where
-  $m$ is the initial metric value, $m_{\mathrm{min}}$/$m_{\mathrm{max}}$
-  are the minimal/maximal values among all architectures. The RGB hexadecimal color code of the cell
-  is \texttt{\#99XX99}, where X=F-4m'.}
-\label{table:evaluation}
-\end{table*}
-
 \section{Discussion}
@@
 Tight integration and sub-tasking require the least amount of
-integration effort, closely followed by service integration and
-workflow import (with service invocation). Pool is a bit behind but
-still before nested workflows (with service invocation) which requires
-more integration than the other architectures. Robustness leads to the
-same ordering of architectures, with tight integration and sub-tasking
-in the top group, service integration and workflow import (with
-service invocation) close behind, pool in a third group, and nested
-workflows (with service invocation) at the end. This ordering is
-consistent across metrics, and it reflects the global complexity of
-the architectures.
+integration effort, followed by service integration, workflow import
+(with service invocation) and pool. Nested workflows (with service
+invocation) requires more integration than the other
+architectures. Robustness roughly leads to the same ordering of
+architectures, with tight integration and sub-tasking in the top
+group, service integration and workflow import (with service
+invocation) close behind, pool in a third group, and nested workflows
+(with service invocation) at the end. This ordering is consistent
+across metrics, it reflects the global complexity of the
+architectures.
 
@@
 mask the unique characteristics of architectures regarding this
-criterion. For instance, nested workflows are the only architecture
-that can easily accommodate distributed workflow execution, which can
-be critical in some cases. At the same time, the scheduling
-constraints created by the sub-tasking and nested workflow
-architectures could well become showstoppers depending on the type of
-targeted infrastructure. Non-reliable infrastructures, for example,
-could hardly cope with workflow engines being wrapped in computing
-tasks as done in sub-tasking. The availability of multiple engine
-instances could also become a critical feature for science gateways
-with important workloads, which would favor sub-tasking and pool.
+criterion. Nested workflows are the only architecture that can easily
+accommodate distributed workflow execution, which can be critical in
+some cases. At the same time, the scheduling constraints created by
+the sub-tasking and nested workflow architectures could well become
+showstoppers depending on the type of targeted
+infrastructure. Non-reliable infrastructures, for example, could
+hardly cope with workflow engines being wrapped in computing tasks as
+done in sub-tasking. The availability of multiple engine instances
+could also become a critical feature for science gateways with
+important workloads, which would favor sub-tasking and pool.
 
@@
 Table~\ref{table:overall} provides an overall comparison between
-architectures, based on the metrics in
-Table~\ref{table:evaluation}. While Table~\ref{table:overall} provides
-a global comparison between architectures, it should not be used
-without the detailed metrics reported on Table~\ref{table:evaluation}
-and discussed above. Overall, sub-tasking, and workflow import (with
-service invocation) perform a bit better than tight integration,
-service invocation and pool. Nested workflows stands a bit behind for
-the reasons mentioned previously.
+architectures, based on the metrics in Table~\ref{table:evaluation}.
+Overall, sub-tasking, and workflow import (with service invocation)
+perform a bit better than tight integration, service invocation and
+pool. Nested workflows stands a bit behind for the reasons mentioned
+previously.
 \begin{table*}
@@
 
-A few limitations remain with our evaluation. First of all, the
-evaluation is done at a quite high-level of abstraction that may be a
-bit artificial when applied to real systems. Abstracting architectures
-from their implementation in specific systems is fundamental to the
-progress of software engineering of science gateways, but we also
-realize that this comes at the cost of realism.
-
-In particular, the presented architectures are pure patterns that may
-be blended in actual systems. The distinction between tight
-integration and service invocation, for instance, may not be that
-clear in practice. Service invocation may also be combined with
-sub-tasking in some cases.
+A few limitations remain that should be considered when using the
+results of our evaluation. First of all, the evaluation is done at a
+quite high-level of abstraction that may be a bit artificial when
+applied to real systems. Abstracting architectures from their
+implementation in specific systems is useful to compare them, but we
+also realize that this comes at the cost of realism.  In particular,
+the presented architectures are pure patterns that may be blended in
+actual systems. The distinction between tight integration and service
+invocation, for instance, may not be that clear in practice. Service
+invocation may also be combined with sub-tasking in some cases.
 
@@
 any other workflow language is very tricky because these engines rely
-on general-purpose programming languages (Bash, Octave, Python, etc)
-which are much richer than scientific workflow languages.
+on general-purpose programming languages such as Bash, Octave and
+Python, which are much richer than scientific workflow languages.
 
@@
 also be envisaged. We chose to limit ourselves to instantiations with
-service invocation because the resulting architectures are implemented in 
-real systems, and because service invocation is a largely
-used architecture. Nested workflows with sub-tasking, in particular,
-could be an interesting architecture to explore.
-
-Finally, there is no need to say that the particular technical or
-historical context of a science gateway project may influence the
-choice of an architecture to integrate workflow engines. For instance,
-some workflow engines may be already available as web services, which
-would tend to favor service invocation, and other science gateways may
-have strongly tested task and data control features (interactions
-\texttt{b} and \texttt{c}), which would favor sub-tasking. The
-migration cost between architectures is totally ignored as well.
+service invocation because the resulting architectures are implemented
+in real systems, and because service invocation is largely
+used. Nested workflows with sub-tasking, in particular, could be an
+interesting architecture to explore.
+
+Finally, the particular technical or historical context of a science
+gateway project may obviously influence the choice of an architecture
+to integrate workflow engines. For instance, some workflow engines may
+be already available as web services, which would tend to favor
+service invocation, and other science gateways may have strongly
+tested task and data control features (interactions \texttt{b} and
+\texttt{c}), which would favor sub-tasking. The migration cost between
+architectures is totally ignored as well.
 
@@
 % Summary of contributions
-We reviewed architectures used to integration workflow engines in
+We reviewed architectures used to integrate workflow engines in
 science gateways, we described them in a system-independent framework
@@
 % Impact of the work
-To the best of our knowledge, our work is the first of its kind. Our
-results can be used as recommendation for science gateway and workflow
-engine developers to build new platforms, or to upgrade existing ones.
+To the best of our knowledge, our work is the first of its kind
+\note{Tristan}{Be more accurate}. Our results can be used as
+recommendation for science gateway and workflow engine architects to
+build new platforms, or to upgrade existing ones.
 

### 2016-04-25 paper.tex
fixed structure of the paper
Intro - 2.1 minor edits
Minor edits and comments abstract to up to 2.1 (excluded).

@@
 \author[mcgill,creatis]{Tristan Glatard}
-\author[mcgill]{Marc-Etienne Rousseau}
+\author[mcgill]{Marc-\'Etienne Rousseau}
 \author[creatis]{Sorina Camarasu-Pop}
@@
 
+\note{Marc-Etienne}{Do we want to use MCIN instead of BIC?}
 \address[mcgill]{McConnell Brain Imaging Centre, Montreal Neurological Institute, McGill University, Canada.}
@@
 \begin{abstract}
-  We investigate software architectures to integrate workflow engines
+  We investigate software architectures commonly used to integrate workflow engines
   in science gateways. Based on our experience with the development
@@
   nested workflows and workflow import. These results provide insights
-  to software architects on how to integrate science gateways and
+  for software architects on how to integrate science gateways and
   workflow engines.
@@
 and so on. A wide-array of workflows have been implemented using these
-pipelines and are now widely shared across neuroimaging groups
+pipelines and are now shared across neuroimaging groups
 world-wide. These
@@
 ones, may transfer data across the infrastructure, and others, usually
-domain-specific ones, may leave it to external processes. Workflows
+domain-specific ones, may leave this role to external processes. Workflows
 may be expressed in any language, including high-level XML or JSON
@@
 Science gateways are used to share resources in a community and to
-provide performance through facilitated access to storage and
-computing power. They are usually accessible through a web interface
+provide increased performance and capacity through facilitated access to storage and
+computing power. They are often accessible through a web interface
 that helps users manage access rights, data transfers, task execution
@@
 which requires robustness.
+\note{Marc-Etienne}{I find that something is weak in this last
+paragraph. Robustness will help with transparency, sure, but it seems only a small facet.
+Also, the text seem to timidly favor architectures with less components/interactions
+without much explanation... Complex systems can be robust, but at greater costs...
+Here is one:
+"Architectures showing reduced system complexity will greatly facilitate the implementation
+effort towards robust interactions which, in turn, have a positive impact on characteristics
+such as gateway predictability, transparence, general reliability and results traceability and
+reproducibility; characteristics that are also key to a good user experience."}
 
@@
 CBRAIN\footnote{\url{http://cbrain.mcgill.ca}}~\cite{SHER-14} is an
-open-source, web-based collaborative platform addressing the
+open-source\footnote{\url{https://github.com/aces/cbrain}}, web-based collaborative platform addressing the
 challenges of data-heavy, computation-intensive neuroimaging
@@
 visualization tools, within a secure environment. CBRAIN is deployed
-at a wide range of computing sites, including nine national HPC
+at a wide range of computing sites, including a network of national HPC
 centers in Canada, one in Korea, one in Germany, and on numerous local
-servers. As September 2015, CBRAIN has served over 370 users across 50
-cities in 19 countries who produced several high-impact results in the
-field of neurosciences.
+servers. As February 2016, CBRAIN has served over 420 users across 50
+cities in 22 countries and has a national computing allocation of 660 CPU years
+per annum.
 
@@
 In this paper, infrastructure consists of the computing and storage
-resources involved in the workflow execution, and the software
-services used to access these resources. Infrastructures can be
+resources involved in the workflow execution, as well as the software
+services used to access these resources. Infrastructures can be composed of
 computing or file servers, databases, clusters, grids or clouds. Some
@@
 notations shown in Figure~\ref{fig:notations}.
+\note{Marc-e}{A few things bother me a bit with the notation fig., making it harder to assimilate
+1- The interactions are padded in the middle of other stuff (Administrator, Data)
+2- The Software component example says "Science gateway", should it not be empty?
+3- The red arrow to explain the use of red; it looks like a very specific type 
+of interactions instead of the desired "any interactions or components can be red meaning...",
+especially with 'a' and 'd' which the reader does not know the meaning yet.
+Maybe a little red arrow and square with "Red used when specific to workflow execution"?
+I don't have a great solution...
+4- Order in text jumps all over the fig notations. Explanations are given for, in order:
+Components, workflow engine, interactions, abstract interactions, red color, tasks, data (no Administrator?)...
+Could re-ordering the fig fix this? A figure figure legend could go a long way too (are there some in this journal?)
+}
 
@@
 
-Software interactions among these components are represented with
+Software interactions among these components are represented with labeled
 arrows.  \emph{Abstract} interactions are specific types of
@@
 since it is used to authenticate users, add new workflows, transfer
-user data, etc. 
 \item Interactions (\texttt{R$_2$}): number of specific interactions

### 2016-04-25 paper.tex
Minor comments and suggestions

--- 2016-04-22 paper.tex
@@
 
-\title{Software architectures to integrate workflow engines in science gateways}
+\title{Software Architectures to Integrate Workflow Engines in Science Gateways}
 
@@
 using regular web development tools and libraries to access
-distributed systems. CBRAIN and VIP are two of such science gateways,
+distributed systems. CBRAIN and VIP are two such science gateways,
 from which we gathered experience about the architectures described in
@@
 centers in Canada, one in Korea, one in Germany, and on numerous local
-servers. As September 2015, CBRAIN has served over 370 users across 50
+servers. By September 2015, CBRAIN has served over 370 users across 50
 cities in 19 countries who produced several high-impact results in the
@@
 DIRAC pilot-job framework~\cite{tsaregorodtsev2008dirac} to execute
-applications on EGI. Almost 900 users were registered in VIP as of
+applications on EGI. Almost 900 users were registered in VIP by
 April 2016 and the average CPU consumption between January 2013 and
@@
 SHIWA Simulation Platform is a WS-PGRADE/gUSE science gateway that was
-developed in European projects
+developed in the European projects
 SHIWA\footnote{\url{http://www.shiwa-workflow.eu}} and
@@
 workflow pool and agent are involved in some architectures only. The
-workflow engine itself is represented by a specific symbol with dotted
-lines.
-
-Software interactions among these components are represented with
+workflow engine itself is represented by a specific symbol enclosed in a dotted rectangle.
+
+Software interactions among these components are represented with plain black
 arrows.  \emph{Abstract} interactions are specific types of
@@
 gateway usually share a database where application, users and other
-resources are stored. In this model, task and data control are both
+resources are stored\note{sorina}{VIP and Moteur also share the WorkflowsDB}. 
+In this model, task and data control are both
 initiated from the science gateway. Interactions \texttt{b} and
@@
 submission (\texttt{1} \& \texttt{2}), workflow status retrieval
-(\texttt{a} and \texttt{b}), and workflow retrieval (\texttt{13} \&
+(\texttt{a} and \texttt{b})\note{sorina}{I suppose letters come here 
+from the initial pool-model figure; it's a bit puzzling when mixed with 
+letters from Fig 2}, and workflow retrieval (\texttt{13} \&
 \texttt{14}). In Figure ~\ref{fig:shiwa-pool-architecture},
@@
   meta-workflows from existing workflows. Meta-workflows offer an
-  additional level of flexibility to build worfklows from reusable
+  additional level of flexibility to build workflows from reusable
   components. \texttt{O$_1$}=0 when the feature is available,
@@
 (\texttt{R$_1$}=0), and only interaction \texttt{e} is
-(\texttt{R$_2$}=1).
+(\texttt{R$_2$}=1)\note{sorina}{I understand why this is 1 (and why interactions 
+c and b are not counted here), but I wonder whether we can simply ignore these interactions
+because they are not specific to workflow execution. As far as robustness is concerned,
+the transfers have to be successful anyway...}.
 
@@
 workflow task and its sub-tasks may create deadlocks when there is
-contention. Say for instance that only 1 computing resource is
+contention. For instance, if only 1 computing resource is
 available for the science gateway and that the workflow task is
@@
 services. Sub-tasking and workflow import (with service invocation)
-perform a bit better to integrate new engines (\texttt{E$_1$}), which
+perform a bit better when integrating new engines (\texttt{E$_1$}), which
 is interesting given the current profusion of engines with different
 characteristics.  However, sub-tasking, workflow import (with service
-invocation) and tight integration perform worse than the others to add
+invocation) and tight integration perform worse than the others when adding
 engine versions (\texttt{E$_2$}) due to the need to update
@@
 workflow converter (workflow import). Workflow import performs worse
-than the others to integrate new workflows (\texttt{E$_3$}) because of
+than the others when integrating new workflows (\texttt{E$_3$}) because of
 the language conversion step. All architectures except nested
-workflows perform the same to integrate new infrastructures (\texttt{E$_4$}).
+workflows perform the same when integrating new infrastructures (\texttt{E$_4$}).
 
@@
 the sub-tasking and nested workflow architectures could well become
-showstoppers depending on the type of targeted
+showstoppers depending on the type of the targeted
 infrastructure. Non-reliable infrastructures, for example, could

### 2016-04-25 paper.tex

--- 2016-04-25 paper.tex
@@
 I don't have a great solution...
+Red has special meaning, maybe remove it from the other pics (workflow eng, tasks)
+as it creates lots of visual interference when looking for red?
 4- Order in text jumps all over the fig notations. Explanations are given for, in order:
@@
 since it is used to authenticate users, add new workflows, transfer
+user data, etc.
 \item Interactions (\texttt{R$_2$}): number of specific interactions

### 2016-04-25 paper.tex

--- 2016-04-25 paper.tex
@@
 visualization tools, within a secure environment. CBRAIN is deployed
-at a wide range of computing sites, including nine national HPC
+at a wide range of computing sites, including a network of national HPC
 centers in Canada, one in Korea, one in Germany, and on numerous local
-servers. By September 2015, CBRAIN has served over 370 users across 50
-cities in 19 countries who produced several high-impact results in the
-field of neurosciences.
+servers. As February 2016, CBRAIN has served over 420 users across 50
+cities in 22 countries and has a national computing allocation of 660 CPU years
+per annum.
 
@@
 notations shown in Figure~\ref{fig:notations}.
+\note{Marc-e}{A few things bother me a bit with the notation fig., making it harder to assimilate
+1- The interactions are padded in the middle of other stuff (Administrator, Data)
+2- The Software component example says "Science gateway", should it not be empty?
+3- The red arrow to explain the use of red; it looks like a very specific type
+of interactions instead of the desired "any interactions or components can be red meaning...",
+especially with 'a' and 'd' which the reader does not know the meaning yet.
+Maybe a little red arrow and square with "Red used when specific to workflow execution"?
+I don't have a great solution...
+Red has special meaning, maybe remove it from the other pics (workflow eng, tasks)
+as it creates lots of visual interference when looking for red?
+4- Order in text jumps all over the fig notations. Explanations are given for, in order:
+Components, workflow engine, interactions, abstract interactions, red color, tasks, data (no Administrator?)...
+Could re-ordering the fig fix this? A figure figure legend could go a long way too (are there some in this journal?)
+}
 
@@
 
-Software interactions among these components are represented with plain black
-arrows.  \emph{Abstract} interactions are specific types of
-interactions that may be implemented using different components and
-interactions. They are represented with dotted arrows and labeled with
-\texttt{*}. An architecture that has an abstract interaction is an
-abstract architecture that can be instantiated in various ways.
+Software interactions among these components are represented with
+labeled, plain black arrows.  \emph{Abstract} interactions are
+specific types of interactions that may be implemented using different
+components and interactions. They are represented with dotted arrows
+and labeled with \texttt{*}. An architecture that has an abstract
+interaction is an abstract architecture that can be instantiated in
+various ways.
 

### 2016-04-25 paper.tex
minor edits up to section 2.1

--- 2016-04-25 paper.tex
@@
 \renewcommand\thesubfigure{\Roman{subfigure}}
+
+\journal{Future Generation Computer Systems}
 
@@
 
-\note{Marc-Etienne}{Do we want to use MCIN instead of BIC?}
 \address[mcgill]{McConnell Brain Imaging Centre, Montreal Neurological Institute, McGill University, Canada.}
@@
 \begin{abstract}
-  We investigate software architectures commonly used to integrate workflow engines
+  We investigate software architectures to integrate workflow engines
   in science gateways. Based on our experience with the development
@@
   nested workflows and workflow import. These results provide insights
-  for software architects on how to integrate science gateways and
+  to software architects on how to integrate science gateways and
   workflow engines.
@@
 
-\journal{Future Generation Computer Systems}
-
-\maketitle
+%\maketitle
 
@@
 IPython, ssh and multi-processor servers, and PSOM is interfaced to
-Torque, SGE, MOAB, LSF and Condor. Some of these tools also support
+Torque, SGE, MOAB, LSF, and HTCondor. Some of these tools also support
 advanced features such as provenance tracking (Nipype, PSOM),
@@
 and so on. A wide-array of workflows have been implemented using these
-pipelines and are now shared across neuroimaging groups
+pipelines and are now widely shared across neuroimaging groups
 world-wide. These
@@
 ones, may transfer data across the infrastructure, and others, usually
-domain-specific ones, may leave this role to external processes. Workflows
+domain-specific ones, may leave it to external processes. Workflows
 may be expressed in any language, including high-level XML or JSON
@@
 Science gateways are used to share resources in a community and to
-provide increased performance and capacity through facilitated access to storage and
-computing power. They are often accessible through a web interface
+provide performance through facilitated access to storage and
+computing power. They are usually accessible through a web interface
 that helps users manage access rights, data transfers, task execution
@@
 which requires robustness.
-\note{Marc-Etienne}{I find that something is weak in this last
-paragraph. Robustness will help with transparency, sure, but it seems only a small facet.
-Also, the text seem to timidly favor architectures with less components/interactions
-without much explanation... Complex systems can be robust, but at greater costs...
-Here is one:
-"Architectures showing reduced system complexity will greatly facilitate the implementation
-effort towards robust interactions which, in turn, have a positive impact on characteristics
-such as gateway predictability, transparence, general reliability and results traceability and
-reproducibility; characteristics that are also key to a good user experience."}
 
@@
 CBRAIN\footnote{\url{http://cbrain.mcgill.ca}}~\cite{SHER-14} is an
-open-source\footnote{\url{https://github.com/aces/cbrain}}, web-based collaborative platform addressing the
+open-source, web-based collaborative platform addressing the
 challenges of data-heavy, computation-intensive neuroimaging
@@
 visualization tools, within a secure environment. CBRAIN is deployed
-at a wide range of computing sites, including a network of national HPC
+at a wide range of computing sites, including nine national HPC
 centers in Canada, one in Korea, one in Germany, and on numerous local
-servers. As February 2016, CBRAIN has served over 420 users across 50
-cities in 22 countries and has a national computing allocation of 660 CPU years
-per annum.
+servers. As September 2015, CBRAIN has served over 370 users across 50
+cities in 19 countries who produced several high-impact results in the
+field of neurosciences.
 
@@
 In this paper, infrastructure consists of the computing and storage
-resources involved in the workflow execution, as well as the software
-services used to access these resources. Infrastructures can be composed of
+resources involved in the workflow execution, and the software
+services used to access these resources. Infrastructures can be
 computing or file servers, databases, clusters, grids or clouds. Some
@@
 notations shown in Figure~\ref{fig:notations}.
-\note{Marc-e}{A few things bother me a bit with the notation fig., making it harder to assimilate
-1- The interactions are padded in the middle of other stuff (Administrator, Data)
-2- The Software component example says "Science gateway", should it not be empty?
-3- The red arrow to explain the use of red; it looks like a very specific type
-of interactions instead of the desired "any interactions or components can be red meaning...",
-especially with 'a' and 'd' which the reader does not know the meaning yet.
-Maybe a little red arrow and square with "Red used when specific to workflow execution"?
-I don't have a great solution...
-Red has special meaning, maybe remove it from the other pics (workflow eng, tasks)
-as it creates lots of visual interference when looking for red?
-4- Order in text jumps all over the fig notations. Explanations are given for, in order:
-Components, workflow engine, interactions, abstract interactions, red color, tasks, data (no Administrator?)...
-Could re-ordering the fig fix this? A figure figure legend could go a long way too (are there some in this journal?)
-}
 
@@
 
-Software interactions among these components are represented with labeled
+Software interactions among these components are represented with
 arrows.  \emph{Abstract} interactions are specific types of
@@
 \centering
+\footnotesize
 \rowcolors{1}{white}{gray!25}
@@
 \end{tabular}
-\caption{Classification of science gateway based on the architecture used to integrate workflow engines.
+\caption{Classification of science gateways based on the architecture used to integrate workflow engines.
 $^c$: based on the Catania Science Gateway Framework.
@@
   workflow \emph{engine}.
+
 \item Task control: operations to manage tasks on the infrastructure,
@@
   Applications~\cite{goodale2006saga}), DRMAA (Distributed Resource
-  Management Application API~\cite{troger2012distributed}) or OCCI
+  Management Application API~\cite{troger2012distributed}), or OCCI
   (Open Cloud Computing Interface~\cite{edmonds2012toward}).
@@
   interaction \texttt{b}, except that information about the parent
-  task which submitted a sub-task is usually available and used for
+  task, which submitted a sub-task, is usually available and used for
   additional control. For instance, the parent task may wait for all
-  its sub-tasks to finish before finishing, and conversely all the
-  sub-tasks may be killed if the parent task is killed.
+  its sub-tasks to complete before finishing, and conversely all the
+  sub-tasks may be terminated if the parent task is killed.
 \item Pool-agent: specific to the pool architecture described in
@@
   well-structured and relatively simple workflow language such as between GWorkflowDL
-  and Scufl~\cite{OLAB-09} and between the 4 languages that were involved in the
+  and Scufl~\cite{OLAB-09}, and between the 4 languages that were involved in the
   SHIWA initiative. In SHIWA, workflow conversion is done through an

### 2016-04-25 paper.tex
edits to section 2

@@
   Pool & SHIWA pool~\cite{ROGE-13}\\
-  Nested & SHIWA Simulation Platform (Coarse-Grained Interoperability~\cite{terstyanszky2014enabling})$^w$\todo{others}\\
+  Nested & \pbox{1.5\columnwidth}{SHIWA Simulation Platform (Coarse-Grained Interoperability~\cite{terstyanszky2014enabling})$^w$, HubZero with Pegasus (via hierarchical workflows)~\cite{Deelman201517} \todo{others}
+  }\\
   Import & SHIWA Simulation Platform (Fine-Grained Interoperability)~\cite{plankensteiner-prodan-etal:2013}$^w$
@@
 framework~\cite{Kacsuk2012}. The integration between the
-HubZero science gateway and Pegasus workflow engine performed
+HubZero science gateway and the Pegasus workflow engine performed
 in~\cite{CPE:CPE3257} also uses a service architecture where the
-\texttt{d} interaction is implemented as a set of calls to Pegasus
-command-line tools (e.g. pegasus-status, pegasus-statistics,
-etc). Figure~\ref{fig:vip-architecture} shows the architecture of the
+\texttt{d} interaction is implemented as a set of calls to Pegasus'
+command-line tools (e.g. pegasus-status, pegasus-plan,
+etc.). Figure~\ref{fig:vip-architecture} shows the architecture of the
 Virtual Imaging Platform, where the MOTEUR workflow
@@
 (\texttt{a} and \texttt{b}), and workflow retrieval (\texttt{13} \&
-\texttt{14}). In Figure ~\ref{fig:shiwa-pool-architecture},
+\texttt{14}). In Figure~\ref{fig:shiwa-pool-architecture},
 interaction \texttt{f} of Figure~\subref*{archi:agent} is implemented
@@
 collapsed in a single one with a \texttt{d} interaction with
-itself. Workflow engines communicate with infrastructures using
+itself. An example of such interaction is the use of hierarchical
+workflows in Pegasus~\cite{Deelman201517}.
+Workflow engines communicate with infrastructures using
 \texttt{b} and \texttt{c$_2$}. Science gateway transfers user data to
@@
 
-See Figure~\subref*{archi:import}. This is an abstract model that we
-instantiate with the service invocation architecture for
+See Figure~\subref*{archi:import}. This is an abstract model 
+instantiated with the service invocation architecture for
 consistency. Workflows are integrated in the science gateway through

### 2016-04-25 paper.tex

@@
 workflow pool and agent are involved in some architectures only. The
-workflow engine itself is represented by a specific symbol with dotted
-lines.
+workflow engine itself is represented by a specific symbol enclosed in a dotted rectangle.
 
 Software interactions among these components are represented with
-arrows.  \emph{Abstract} interactions are specific types of
-interactions that may be implemented using different components and
-interactions. They are represented with dotted arrows and labeled with
-\texttt{*}. An architecture that has an abstract interaction is an
-abstract architecture that can be instantiated in various ways.
+labeled, plain black arrows.  \emph{Abstract} interactions are
+specific types of interactions that may be implemented using different
+components and interactions. They are represented with dotted arrows
+and labeled with \texttt{*}. An architecture that has an abstract
+interaction is an abstract architecture that can be instantiated in
+various ways.
 
@@
 submission (\texttt{1} \& \texttt{2}), workflow status retrieval
-(\texttt{a} and \texttt{b}), and workflow retrieval (\texttt{13} \&
+(\texttt{a} and \texttt{b})\note{sorina}{I suppose letters come here 
+from the initial pool-model figure; it's a bit puzzling when mixed with 
+letters from Fig 2}, and workflow retrieval (\texttt{13} \&
 \texttt{14}). In Figure~\ref{fig:shiwa-pool-architecture},

### 2016-04-25 paper.tex
Updated McGill's affiliation
edits to section 3.1

--- 2016-04-25 paper.tex
@@
 execution, extensibility, scalability and other specific features. Criteria
-break down to specific metrics where \emph{low value indicates good
+break down to specific metrics where \emph{low value indicates better
 performance}. For each criterion, a total score is computed by
@@
 metrics in a criterion measure similar entities so that summation
-makes sense. The
+does not incur in loss of accuracy. The
 criteria and metrics are explained hereafter.
@@
 \begin{table*}
+\footnotesize
 \centering
@@
 \begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
-\item Total number of components (\texttt{I$_1$}),
+\item Total number of components (\texttt{I$_1$}), and
 \item Total number of interactions (\texttt{I$_2$}).
@@
 infrastructure in the architecture usually requires some technical
-effort (e.g. account creation, software installation, etc) which is
+effort (e.g., account creation, software installation, etc.), which is
 why we keep it in the metric. The two metrics are summed to obtain the
@@
 interactions in the software architecture. Note that errors coming
-from the infrastructure (e.g. unavailable data or terminated tasks) or
+from the infrastructure (e.g., unavailable data or terminated tasks) or
 workflows (e.g., wrong user input or application errors) are not
@@
 criterion, which measures the total number of software pieces that are
-specifically involved in workflow execution.
+specifically involved in the workflow execution.
 
@@
   or interactions. We measure the difficulty to integrate a new
-  workflow by counting the number of interactions that need to be
-  invoked to integrate a new workflow in the architecture, assuming
+  workflow by counting the number of interactions required
+  to integrate a new workflow in the architecture, assuming
   that the engine type and version required to execute this workflow
@@
   (e.g. GPUs, clouds), or to enforce execution policies 
-  (e.g. constrain data to remain in a particular network domain).
+  (e.g.
+constrain data to remain in a particular network domain).\note{Rafael}{Adding a new engine type may also allow run in different infrastructures.
+Maybe it is worth to put a sentence to clarify that the metric account for it separately.}
 %\item New science gateway: \todo{Not sure if this makes sense}.
@@
   more than 1 engine instance in the architecture. Workflow engines
-  may require important amounts of resources when a lot of workflows,
+  may require important amounts of resources when several workflows,
   or large workflows, are executed. At some point, it may be required
-  for the science gateway to distribute the load to several
+  for the science gateway to distribute the load among several
   engines. \texttt{S$_1$}=0 when adding a new engine instance is a
@@
   adding a new engine instance requires some form of manual
-  intervention, which prevents easy implementations of elastic
-  engines. \texttt{S$_1$}=2 when new engine instances cannot be added.
+  intervention (e.g., two-factor authentication), which prevents easy 
+  implementations of elastic engines. \texttt{S$_1$}=2 when new engine 
+  instances cannot be added.
 \item Distributed engines (\texttt{S$_2$}): possibility to distribute
@@
   engine. \texttt{S$_2$}=0 when distributed engines are possible in
-  the architecture and \texttt{S$_2$}=1 when they require specific
-  developments in the workflow engine.
+  the architecture, and \texttt{S$_2$}=1 when they require specific
+  developments in the workflow engine.\note{Rafael}{$S_2$ would be 2 if the paradigm does not allow to distribute the orchestration?
+e.g., workflow executions implemented within MPI jobs, in Pegasus PMC and dispel4Py.}
 \item Task scheduling (\texttt{S$_3$}): task scheduling is a difficult
@@
   the architecture does not introduce any additional complexity to the
-  scheduling problem and \texttt{S$_3$}=1 otherwise.
+  scheduling problem, and \texttt{S$_3$}=1 otherwise.
 \end{itemize}

### 2016-04-25 paper.tex
Fixed a few notes

--- 2016-04-25 paper.tex
@@
 \newcommand{\note}[2]{\textbf{\Large{\color{blue}\footnote{{\color{blue}\textbf{#1:} \textit{#2}\color{black}}}}}}
+\newcommand{\closednote}[4]{}
 
@@
 
-\note{Marc-Etienne}{Do we want to use MCIN instead of BIC?}
-\address[mcgill]{McConnell Brain Imaging Centre, Montreal Neurological Institute, McGill University, Canada.}
+\address[mcgill]{McGill Centre for Integrative Neuroscience, Montreal Neurological Institute, McGill University, Canada.}
 \address[creatis]{University of Lyon, CNRS, INSERM, CREATIS, Villeurbanne, France.}
@@
 
-Robustness is a key part of science gateway user experience. In this
-respect, the multiplicity of software components and interactions
-involved in architectures is a challenge as it often leads to complex
-systems that are therefore prone to failure. To offer good user
-experience, science gateways should remain as transparent as possible,
-which requires robustness.
-\note{Marc-Etienne}{I find that something is weak in this last
-paragraph. Robustness will help with transparency, sure, but it seems only a small facet.
-Also, the text seem to timidly favor architectures with less components/interactions
-without much explanation... Complex systems can be robust, but at greater costs...
-Here is one:
-"Architectures showing reduced system complexity will greatly facilitate the implementation
-effort towards robust interactions which, in turn, have a positive impact on characteristics
-such as gateway predictability, transparence, general reliability and results traceability and
-reproducibility; characteristics that are also key to a good user experience."}
+Architectures showing reduced system complexity will greatly
+facilitate the implementation effort towards robust interactions
+which, in turn, have a positive impact on characteristics such as
+gateway predictability, transparence, general reliability and results
+traceability and reproducibility; characteristics that are also key to
+a good user experience. \closednote{Marc-Etienne}{I find that
+  something is weak in this last paragraph. Robustness will help with
+  transparency, sure, but it seems only a small facet.  Also, the text
+  seem to timidly favor architectures with less
+  components/interactions without much explanation... Complex systems
+  can be robust, but at greater costs...  Here is one: "Architectures
+  showing reduced system complexity will greatly facilitate the
+  implementation effort towards robust interactions which, in turn,
+  have a positive impact on characteristics such as gateway
+  predictability, transparence, general reliability and results
+  traceability and reproducibility; characteristics that are also key
+  to a good user experience."}{Tristan}{Agreed.}
 
@@
 diagrammed in Figure~\ref{fig:architectures} using the graphical
-notations shown in Figure~\ref{fig:notations}.
-\note{Marc-e}{A few things bother me a bit with the notation fig., making it harder to assimilate
-1- The interactions are padded in the middle of other stuff (Administrator, Data)
-2- The Software component example says "Science gateway", should it not be empty?
-3- The red arrow to explain the use of red; it looks like a very specific type
-of interactions instead of the desired "any interactions or components can be red meaning...",
-especially with 'a' and 'd' which the reader does not know the meaning yet.
-Maybe a little red arrow and square with "Red used when specific to workflow execution"?
-I don't have a great solution...
-Red has special meaning, maybe remove it from the other pics (workflow eng, tasks)
-as it creates lots of visual interference when looking for red?
-4- Order in text jumps all over the fig notations. Explanations are given for, in order:
-Components, workflow engine, interactions, abstract interactions, red color, tasks, data (no Administrator?)...
-Could re-ordering the fig fix this? A figure figure legend could go a long way too (are there some in this journal?)
-}
+notations shown in Figure~\ref{fig:notations}.  \closednote{Marc-e}{A
+  few things bother me a bit with the notation fig., making it harder
+  to assimilate 1- The interactions are padded in the middle of other
+  stuff (Administrator, Data) 2- The Software component example says
+  "Science gateway", should it not be empty?  3- The red arrow to
+  explain the use of red; it looks like a very specific type of
+  interactions instead of the desired "any interactions or components
+  can be red meaning...", especially with 'a' and 'd' which the reader
+  does not know the meaning yet.  Maybe a little red arrow and square
+  with "Red used when specific to workflow execution"?  I don't have a
+  great solution...  Red has special meaning, maybe remove it from the
+  other pics (workflow eng, tasks) as it creates lots of visual
+  interference when looking for red?  4- Order in text jumps all over
+  the fig notations. Explanations are given for, in order: Components,
+  workflow engine, interactions, abstract interactions, red color,
+  tasks, data (no Administrator?)...  Could re-ordering the fig fix
+  this? A figure figure legend could go a long way too (are there some
+  in this journal?)  }{Tristan}{Marc-E, is Figure 1 better now?}
 
@@
 infrastructure. It does not cover workflow parameters such as strings,
-numbers, etc.
+numbers, etc. Administrator refers to a science gateway user with
+extended privileges to install applications.
 
@@
 \centering
-\def\svgwidth{\columnwidth}
+\def\svgwidth{0.7\columnwidth}
 \input{figures/notations.pdf_tex}
@@
   use it. The infrastructure might offer various data storage backends
-  with heterogeneous interfaces. \note{Marc-Etienne}{this -c2- can be also trigerred
-    by the user}\note{Tristan}{No, it can't: this is by definition an interaction between the workflow engine and the infrastructure.
+  with heterogeneous interfaces. \closednote{Marc-Etienne}{this -c2- can be also trigerred
+    by the user}{Tristan}{No, it can't: this is by definition an interaction between the workflow engine and the infrastructure.
 It is used to transfer data to/from tasks.
@@
 gateway usually share a database where application, users and other
-resources are stored\note{sorina}{VIP and Moteur also share the WorkflowsDB}. 
-In this model, task and data control are both
-initiated from the science gateway. Interactions \texttt{b} and
-\texttt{c$_2$} are initiated from the workflow engine while
-\texttt{c$_1$} comes from other parts of the science gateway, for
-instance a data management interface. As in any other model, the
+resources are stored\closednote{sorina}{VIP and Moteur also share the
+  WorkflowsDB}{Tristan}{Yes, you are right. This is an example where
+  such abstract architectures may not completely apply in real
+  life. See comment about ``blended'' architectures in the
+  discussion. In case of VIP, the sharing of the DB is more an
+  implementation detail rather than a strong architectural pattern: we
+  could avoid sharing the DB if we wanted.}.  In this model, task and
+data control are both initiated from the science gateway. Interactions
+\texttt{b} and \texttt{c$_2$} are initiated from the workflow engine
+while \texttt{c$_1$} comes from other parts of the science gateway,
+for instance a data management interface. As in any other model, the
 installation of new workflows in the science gateway (\texttt{a$_2$})
@@
 submission (\texttt{1} \& \texttt{2}), workflow status retrieval
-(\texttt{a} and \texttt{b})\note{sorina}{I suppose letters come here 
+(\texttt{0} \& \texttt{3})\closednote{sorina}{I suppose letters come here 
 from the initial pool-model figure; it's a bit puzzling when mixed with 
-letters from Fig 2}, and workflow retrieval (\texttt{13} \&
+letters from Fig 2}{Tristan}{Fixed.}, and workflow retrieval (\texttt{13} \&
 \texttt{14}). In Figure~\ref{fig:shiwa-pool-architecture},

### 2016-04-25 paper.tex
Minor updates

@@
 controls the service through a specific interaction (\texttt{d}) that
-might be implemented as a web-service call (e.g. RESTful or SOAP), as
+might be implemented as a web-service call (e.g., RESTful or SOAP), as
 a command-line or as any other method that offers a well-defined
@@
 \texttt{d} interaction is implemented as a set of calls to Pegasus'
-command-line tools (e.g. pegasus-status, pegasus-plan,
+command-line tools (e.g., pegasus-status, pegasus-plan,
 etc.). Figure~\ref{fig:vip-architecture} shows the architecture of the
@@
 evaluate the architectures: integration effort, robustness of workflow
-execution, extensibility, scalability and other specific features. Criteria
-break down to specific metrics where \emph{low value indicates better
-performance}. For each criterion, a total score is computed by
-summing up the individual metrics. We ensure that the different
-metrics in a criterion measure similar entities so that summation
-does not incur in loss of accuracy. The
-criteria and metrics are explained hereafter.
+execution, extensibility, scalability and other specific
+features. Criteria break down to specific metrics where \emph{lower
+  value indicates better performance}. For each criterion, a total
+score is computed by summing up the individual metrics. We ensure that
+the different metrics in a criterion measure similar entities so that
+they can be summed. The criteria and metrics are explained hereafter.
 
@@
   computing or storage power, to access specific types of resources
-  (e.g. GPUs, clouds), or to enforce execution policies 
-  (e.g.
-constrain data to remain in a particular network domain).\note{Rafael}{Adding a new engine type may also allow run in different infrastructures.
+  (e.g., GPUs, clouds), or to enforce execution policies 
+  (e.g., constrain data to remain in a particular network domain).\note{Rafael}{Adding a new engine type may also allow run in different infrastructures.
 Maybe it is worth to put a sentence to clarify that the metric account for it separately.}

### 2016-04-25 paper.tex
Improved display of notes and closed notes.

@@
 \usepackage{pbox}
+\usepackage{pdfcomment}
 
@@
 \newcommand{\todo}[1]{\color{blue}\xspace[\emph{#1}]\xspace\color{black}}
-\newcommand{\note}[2]{\textbf{\Large{\color{blue}\footnote{{\color{blue}\textbf{#1:} \textit{#2}\color{black}}}}}}
-\newcommand{\closednote}[4]{}
+\newcommand{\note}[2]{\pdfmargincomment[color=yellow,author=#1,open=true]{#2}}
+\newcommand{\closednote}[4]{\pdfmargincomment[color=gray,author=#1-#3,open=true]{#1: #2 \textCR \textCR #3: #4}}
 
@@
   traceability and reproducibility; characteristics that are also key
-  to a good user experience."}{Tristan}{Agreed.}
+  to a good user experience."}{Tristan}{Agreed, fixed as suggested.}
 

### 2016-04-25 paper.tex
Updates in table colors.

@@
 \definecolor{ao(english)}{rgb}{0.0, 0.5, 0.0}
-\newcommand{\todo}[1]{\color{blue}\xspace[\emph{#1}]\xspace\color{black}}
+\newcommand{\todo}[2]{\pdfmargincomment[color=red,author=#1,open=true]{#2}}
 \newcommand{\note}[2]{\pdfmargincomment[color=yellow,author=#1,open=true]{#2}}
@@
   Pool & SHIWA pool~\cite{ROGE-13}\\
-  Nested & \pbox{1.5\columnwidth}{SHIWA Simulation Platform (Coarse-Grained Interoperability~\cite{terstyanszky2014enabling})$^w$, HubZero with Pegasus (via hierarchical workflows)~\cite{Deelman201517} \todo{others}
+  Nested & \pbox{1.5\columnwidth}{SHIWA Simulation Platform (Coarse-Grained Interoperability~\cite{terstyanszky2014enabling})$^w$, HubZero with Pegasus (via hierarchical workflows)~\cite{Deelman201517} \todo{Tristan}{Add others}
   }\\
@@
 
-\todo{Add description of a real system}
-
-\todo{Refer to ``From the desktop to the grid: scalable bioinformatics via workflow conversion Luis de la GarzaEmail author, Johannes Veit, Andras Szolek, Marc Röttig, Stephan Aiche, Sandra Gesing, Knut Reinert and Oliver Kohlbacher''}
+\todo{Tristan}{Add description of a real system}
+
+\todo{Tristan}{Refer to ``From the desktop to the grid: scalable bioinformatics via workflow conversion Luis de la GarzaEmail author, Johannes Veit, Andras Szolek, Marc Röttig, Stephan Aiche, Sandra Gesing, Knut Reinert and Oliver Kohlbacher''}
 
@@
   Total components -- \texttt{I$_1$} & \cellcolor[HTML]{99FF99}2
-                                     & \cellcolor[HTML]{99E899}3
+                                     & \cellcolor[HTML]{99DD99}3
                                      & \cellcolor[HTML]{99FF99}2
-                                     & \cellcolor[HTML]{99D199}4
-                                     & \cellcolor[HTML]{99BB99}5
-                                     & \cellcolor[HTML]{99E899}3\\
+                                     & \cellcolor[HTML]{99BB99}4
+                                     & \cellcolor[HTML]{999999}5
+                                     & \cellcolor[HTML]{99DD99}3\\
 Total interactions -- \texttt{I$_2$} & \cellcolor[HTML]{99FF99}5
-                                     & \cellcolor[HTML]{99F399}6
+                                     & \cellcolor[HTML]{99EE99}6
                                      & \cellcolor[HTML]{99FF99}5
-                                     & \cellcolor[HTML]{99E899}7
-                                     & \cellcolor[HTML]{99BB99}11
-                                     & \cellcolor[HTML]{99E899}7\\
+                                     & \cellcolor[HTML]{99DD99}7
+                                     & \cellcolor[HTML]{999999}11
+                                     & \cellcolor[HTML]{99DD99}7\\
 \textbf{Total} (total software pieces) & \cellcolor[HTML]{99FF99}\textbf{7}
-                                     & \cellcolor[HTML]{99EF99}\textbf{9}
+                                     & \cellcolor[HTML]{99E899}\textbf{9}
                                      & \cellcolor[HTML]{99FF99}\textbf{7}
-                                     & \cellcolor[HTML]{99E099}\textbf{11}
-                                     & \cellcolor[HTML]{99BB99}\textbf{16}
-                                     & \cellcolor[HTML]{99E899}\textbf{10}\\
+                                     & \cellcolor[HTML]{99D199}\textbf{11}
+                                     & \cellcolor[HTML]{999999}\textbf{16}
+                                     & \cellcolor[HTML]{99DD99}\textbf{10}\\
 \cellcolor[HTML]{EEEEEE}\textbf{Robustness}& \multicolumn{6}{l}{\cellcolor[HTML]{EEEEEE}}\\
 Specific components -- \texttt{R$_1$} & \cellcolor[HTML]{99FF99}0
-                                     & \cellcolor[HTML]{99DD99}1
+                                     & \cellcolor[HTML]{99CC99}1
                                      & \cellcolor[HTML]{99FF99}0
-                                     & \cellcolor[HTML]{99BB99}2
-                                     & \cellcolor[HTML]{99BB99}2
-                                     & \cellcolor[HTML]{99DD99}1\\
-  Specific interactions -- \texttt{R$_2$} & \cellcolor[HTML]{99F199}2
-                                     & \cellcolor[HTML]{99E399}3
+                                     & \cellcolor[HTML]{999999}2
+                                     & \cellcolor[HTML]{999999}2
+                                     & \cellcolor[HTML]{99CC99}1\\
+  Specific interactions -- \texttt{R$_2$} & \cellcolor[HTML]{99EA99}2
+                                     & \cellcolor[HTML]{99D699}3
                                      & \cellcolor[HTML]{99FF99}1
-                                     & \cellcolor[HTML]{99D699}4
-                                     & \cellcolor[HTML]{99BB99}6
-                                     & \cellcolor[HTML]{99E399}3\\
-  \textbf{Total} (execution software pieces)& \cellcolor[HTML]{99F599}\textbf{2}
-                                     & \cellcolor[HTML]{99E199}\textbf{4}
+                                     & \cellcolor[HTML]{99C199}4
+                                     & \cellcolor[HTML]{999999}6
+                                     & \cellcolor[HTML]{99D699}3\\
+  \textbf{Total} (execution software pieces)& \cellcolor[HTML]{99F099}\textbf{2}
+                                     & \cellcolor[HTML]{99D399}\textbf{4}
                                      & \cellcolor[HTML]{99FF99}\textbf{1}
-                                     & \cellcolor[HTML]{99CE99}\textbf{6}
-                                     & \cellcolor[HTML]{99BB99}\textbf{8}
-                                     & \cellcolor[HTML]{99E199}\textbf{4}\\
+                                     & \cellcolor[HTML]{99B699}\textbf{6}
+                                     & \cellcolor[HTML]{999999}\textbf{8}
+                                     & \cellcolor[HTML]{99D399}\textbf{4}\\
 \cellcolor[HTML]{EEEEEE}\textbf{Extensibility}& \multicolumn{6}{l}{\cellcolor[HTML]{EEEEEE}}\\
-  New engine type -- \texttt{E$_1$}  & \cellcolor[HTML]{99D899}3
-                                     & \cellcolor[HTML]{99C499}4
-                                     & \cellcolor[HTML]{99EB99}2
-                                     & \cellcolor[HTML]{99D899}3
-                                     & \cellcolor[HTML]{99BB99}4.5
+  New engine type -- \texttt{E$_1$}  & \cellcolor[HTML]{99C499}3
+                                     & \cellcolor[HTML]{99A799}4
+                                     & \cellcolor[HTML]{99E199}2
+                                     & \cellcolor[HTML]{99C499}3
+                                     & \cellcolor[HTML]{999999}4.5
                                      & \cellcolor[HTML]{99FF99}1\\
-New engine version -- \texttt{E$_2$} & \cellcolor[HTML]{99BB99}1
+New engine version -- \texttt{E$_2$} & \cellcolor[HTML]{999999}1
                                      & \cellcolor[HTML]{99FF99}0
-                                     & \cellcolor[HTML]{99BB99}1
+                                     & \cellcolor[HTML]{999999}1
                                      & \cellcolor[HTML]{99FF99}0
                                      & \cellcolor[HTML]{99FF99}0
-                                     & \cellcolor[HTML]{99BB99}1\\
+                                     & \cellcolor[HTML]{999999}1\\
   New workflow -- \texttt{E$_3$} & \cellcolor[HTML]{99FF99}2
@@
                                      & \cellcolor[HTML]{99FF99}2
-                                     & \cellcolor[HTML]{99BB99}3\\
-New infrastructure -- \texttt{E$_4$} & \cellcolor[HTML]{99E899}3
-                                     & \cellcolor[HTML]{99E899}3
+                                     & \cellcolor[HTML]{999999}3\\
+New infrastructure -- \texttt{E$_4$} & \cellcolor[HTML]{99DD99}3
+                                     & \cellcolor[HTML]{99DD99}3
                                      & \cellcolor[HTML]{99FF99}2
-                                     & \cellcolor[HTML]{99E899}3
-                                     & \cellcolor[HTML]{99BB99}5
-                                     & \cellcolor[HTML]{99E899}3\\
-  \textbf{Total} (difficulty to extend) & \cellcolor[HTML]{99E099}\textbf{9}
-                                     & \cellcolor[HTML]{99E099}\textbf{9}
+                                     & \cellcolor[HTML]{99DD99}3
+                                     & \cellcolor[HTML]{999999}5
+                                     & \cellcolor[HTML]{99DD99}3\\
+  \textbf{Total} (difficulty to extend) & \cellcolor[HTML]{99D199}\textbf{9}
+                                     & \cellcolor[HTML]{99D199}\textbf{9}
                                      & \cellcolor[HTML]{99FF99}\textbf{7}
-                                     & \cellcolor[HTML]{99EF99}\textbf{8}
-                                     & \cellcolor[HTML]{99BB99}\textbf{11.5}
-                                     & \cellcolor[HTML]{99EF99}\textbf{8}\\
+                                     & \cellcolor[HTML]{99E899}\textbf{8}
+                                     & \cellcolor[HTML]{999999}\textbf{11.5}
+                                     & \cellcolor[HTML]{99E899}\textbf{8}\\
 \cellcolor[HTML]{EEEEEE}\textbf{Scalability}& \multicolumn{6}{l}{\cellcolor[HTML]{EEEEEE}}\\
-Multiple engine instances -- \texttt{S$_1$}& \cellcolor[HTML]{99BB99}2
-                                     & \cellcolor[HTML]{99DD99}1
+Multiple engine instances -- \texttt{S$_1$}& \cellcolor[HTML]{999999}2
+                                     & \cellcolor[HTML]{99CC99}1
                                      & \cellcolor[HTML]{99FF99}0
                                      & \cellcolor[HTML]{99FF99}0
-                                     & \cellcolor[HTML]{99DD99}1
-                                     & \cellcolor[HTML]{99DD99}1\\
-Distributed engines -- \texttt{S$_2$}& \cellcolor[HTML]{99BB99}1
-                                     & \cellcolor[HTML]{99BB99}1
-                                     & \cellcolor[HTML]{99BB99}1
-                                     & \cellcolor[HTML]{99BB99}1
+                                     & \cellcolor[HTML]{99CC99}1
+                                     & \cellcolor[HTML]{99CC99}1\\
+Distributed engines -- \texttt{S$_2$}& \cellcolor[HTML]{999999}1
+                                     & \cellcolor[HTML]{999999}1
+                                     & \cellcolor[HTML]{999999}1
+                                     & \cellcolor[HTML]{999999}1
                                      & \cellcolor[HTML]{99FF99}0
-                                     & \cellcolor[HTML]{99BB99}1\\
+                                     & \cellcolor[HTML]{999999}1\\
 Task scheduling -- \texttt{S$_3$}    & \cellcolor[HTML]{99FF99}0
                                      & \cellcolor[HTML]{99FF99}0
-                                     & \cellcolor[HTML]{99BB99}1
+                                     & \cellcolor[HTML]{999999}1
                                      & \cellcolor[HTML]{99FF99}0
-                                     & \cellcolor[HTML]{99BB99}1
+                                     & \cellcolor[HTML]{999999}1
                                      & \cellcolor[HTML]{99FF99}0\\
-\textbf{Total} (scalability issues)  & \cellcolor[HTML]{99BB99}\textbf{3}
-                                     & \cellcolor[HTML]{99DD99}\textbf{2}
-                                     & \cellcolor[HTML]{99DD99}\textbf{2}
+\textbf{Total} (scalability issues)  & \cellcolor[HTML]{999999}\textbf{3}
+                                     & \cellcolor[HTML]{99CC99}\textbf{2}
+                                     & \cellcolor[HTML]{99CC99}\textbf{2}
                                      & \cellcolor[HTML]{99FF99}\textbf{1}
-                                     & \cellcolor[HTML]{99DD99}\textbf{2}
-                                     & \cellcolor[HTML]{99DD99}\textbf{2}\\
+                                     & \cellcolor[HTML]{99CC99}\textbf{2}
+                                     & \cellcolor[HTML]{99CC99}\textbf{2}\\
 \cellcolor[HTML]{EEEEEE}\textbf{Specific features}& \multicolumn{6}{l}{\cellcolor[HTML]{EEEEEE}}\\
-  Meta-workflow  -- \texttt{O$_1$}    & \cellcolor[HTML]{99BB99}1
-                                     & \cellcolor[HTML]{99BB99}1
-                                     & \cellcolor[HTML]{99BB99}1
-                                     & \cellcolor[HTML]{99BB99}1
+  Meta-workflow  -- \texttt{O$_1$}    & \cellcolor[HTML]{999999}1
+                                     & \cellcolor[HTML]{999999}1
+                                     & \cellcolor[HTML]{999999}1
+                                     & \cellcolor[HTML]{999999}1
                                      & \cellcolor[HTML]{99FF99}0
@@
                                      & \cellcolor[HTML]{99FF99}0
-                                     & \cellcolor[HTML]{99BB99}1
+                                     & \cellcolor[HTML]{999999}1
                                      & \cellcolor[HTML]{99FF99}0
-                                     & \cellcolor[HTML]{99BB99}1
+                                     & \cellcolor[HTML]{999999}1
                                      & \cellcolor[HTML]{99FF99}0\\
-  \textbf{Total} (missing features) & \cellcolor[HTML]{99DD99}\textbf{1}
-                                     & \cellcolor[HTML]{99DD99}\textbf{1}
-                                     & \cellcolor[HTML]{99BB99}\textbf{2}
-                                     & \cellcolor[HTML]{99DD99}\textbf{1}
-                                     & \cellcolor[HTML]{99DD99}\textbf{1}
+  \textbf{Total} (missing features) & \cellcolor[HTML]{99CC99}\textbf{1}
+                                     & \cellcolor[HTML]{99CC99}\textbf{1}
+                                     & \cellcolor[HTML]{999999}\textbf{2}
+                                     & \cellcolor[HTML]{99CC99}\textbf{1}
+                                     & \cellcolor[HTML]{99CC99}\textbf{1}
                                      & \cellcolor[HTML]{99FF99}\textbf{0}\\
@@
   are the minimal and maximal values among all architectures. The RGB hexadecimal color code of the cell
-  is \texttt{\#99XX99}, where X=F-4m'. \todo{Use 6. Round to the nearest integer.}}
+  is \texttt{\#99XX99}, where X=F-4m'. \closednote{Tristan}{Use 6. Round to the nearest integer.}{Tristan}{Fixed.}}
 \label{table:evaluation}
@@
                                     \cellcolor[HTML]{99FF99}{0.00}
-                                    & \cellcolor[HTML]{99EF99}{0.22}
+                                    & \cellcolor[HTML]{99E899}{0.22}
                                     & \cellcolor[HTML]{99FF99}{0.00}
-                                    & \cellcolor[HTML]{99E099}{0.44}
-                                    & \cellcolor[HTML]{99BB99}{1.00}
-                                      & \cellcolor[HTML]{99E899}{0.33}\\
+                                    & \cellcolor[HTML]{99D199}{0.44}
+                                    & \cellcolor[HTML]{999999}{1.00}
+                                      & \cellcolor[HTML]{99DD99}{0.33}\\
 Robustness (normalized total) &
-                                \cellcolor[HTML]{99F599}{0.14}
-                                    & \cellcolor[HTML]{99E199}{0.43}
+                                \cellcolor[HTML]{99F099}{0.14}
+                                    & \cellcolor[HTML]{99D399}{0.43}
                                     & \cellcolor[HTML]{99FF99}{0.00}
-                                    & \cellcolor[HTML]{99CE99}{0.71}
-                                    & \cellcolor[HTML]{99BB99}{1.00}
-                                    & \cellcolor[HTML]{99E199}{0.43}\\
-  Extensibility (normalized total)  & \cellcolor[HTML]{99E099}{0.44}
-                                     & \cellcolor[HTML]{99E099}{0.44}
+                                    & \cellcolor[HTML]{99B699}{0.71}
+                                    & \cellcolor[HTML]{999999}{1.00}
+                                    & \cellcolor[HTML]{99D399}{0.43}\\
+  Extensibility (normalized total)  & \cellcolor[HTML]{99D199}{0.44}
+                                     & \cellcolor[HTML]{99D199}{0.44}
                                      & \cellcolor[HTML]{99FF99}{0.00}
-                                     & \cellcolor[HTML]{99EF99}{0.22}
-                                     & \cellcolor[HTML]{99BB99}{1.00}
-                                     & \cellcolor[HTML]{99EF99}{0.22}\\
-Scalability (normalized total)  & \cellcolor[HTML]{99BB99}{1.00}
-                                     & \cellcolor[HTML]{99DD99}{0.50}
-                                     & \cellcolor[HTML]{99DD99}{0.50}
+                                     & \cellcolor[HTML]{99E899}{0.22}
+                                     & \cellcolor[HTML]{999999}{1.00}
+                                     & \cellcolor[HTML]{99E899}{0.22}\\
+Scalability (normalized total)  & \cellcolor[HTML]{999999}{1.00}
+                                     & \cellcolor[HTML]{99CC99}{0.50}
+                                     & \cellcolor[HTML]{99CC99}{0.50}
                                      & \cellcolor[HTML]{99FF99}{0.00}
-                                     & \cellcolor[HTML]{99DD99}{0.50}
-                                     & \cellcolor[HTML]{99DD99}{0.50}\\
-Specific features (normalized total) & \cellcolor[HTML]{99DD99}{0.50}
-                                     & \cellcolor[HTML]{99DD99}{0.50}
-                                     & \cellcolor[HTML]{99BB99}{1.00}
-                                     & \cellcolor[HTML]{99DD99}{0.50}
-                                     & \cellcolor[HTML]{99DD99}{0.50}
-                                     & \cellcolor[HTML]{99FF99}{0.00}\\
-                                    & \cellcolor[HTML]{99EE99}\textbf{2.1}
-                                    & \cellcolor[HTML]{99EE99}\textbf{2.1}
-                                    & \cellcolor[HTML]{99FF99}\textbf{1.5}
-                                    & \cellcolor[HTML]{99F499}\textbf{1.9}
-                                    & \cellcolor[HTML]{99BB99}\textbf{4.0}
+                                     & \cellcolor[HTML]{99CC99}{0.50}
+                                     & \cellcolor[HTML]{99CC99}{0.50}\\
+Specific features (normalized total) & \cellcolor[HTML]{99CC99}{0.50}
+                                     & \cellcolor[HTML]{99CC99}{0.50}
+                                     & \cellcolor[HTML]{999999}{1.00}
+                                     & \cellcolor[HTML]{99CC99}{0.50}
+                                     & \cellcolor[HTML]{99CC99}{0.50}
+                                     & \cellcolor[HTML]{99DD99}{0.00}\\
+                                    & \cellcolor[HTML]{99E699}\textbf{2.1}
+                                    & \cellcolor[HTML]{99E699}\textbf{2.1}
+                                    & \cellcolor[HTML]{99FE99}\textbf{1.5}
+                                    & \cellcolor[HTML]{99EE99}\textbf{1.9}
+                                    & \cellcolor[HTML]{999999}\textbf{4.0}
                                     & \cellcolor[HTML]{99FF99}\textbf{1.5}\\

### 2016-04-26 paper.tex
overall minor edits and comments

@@
 
-\paragraph{Specific features} Meta-workflow are not supported by default
+\paragraph{Specific features} Meta-workflows are not supported by default
 (\texttt{O$_1$}=1). Fine-grained debugging information is usually easy
@@
 infrastructure) and \texttt{d} (twice: once for each workflow engine).
+\note{Rafael}{I like the idea of what has been discussed in the `integration effort', however I think the name may lead to misinterpretations.
+For example, according to Table 2, it requires more effort to integrate in a Tight architecture than in a service-based architecture.
+Considering the principles of software engineer, it is simpler to integrate/maintain a system using decoupled services.
+Maybe it should be renamed to something related to 'dependencies'.}
 
@@
 invocation) are the only architectures that support meta-workflows,
-which may be interesting in some cases. The availability of
-fine-grained debugging information may also be critical to efficient
+which may be interesting in some cases.\note{Rafael}{I would suggest to provide some
+more examples when this could be interesting.}
+The availability of fine-grained debugging information may also be critical to efficient
 user support.
@@
 pool. Nested workflows stands a bit behind for the reasons mentioned
-previously.
+previously.\note{Rafael}{I would emphasize that this analysis is only for illustration purposes, since it does not account for different weights in metrics.}
 \begin{table*}
@@
 of all the architectures based on these metrics.
+\note{Rafael}{One thing that is missing in the conclusion or the limitations section is to state that this work does not intend to quantify the amount of work required to develop science gateways using a specific architecture, but the goal is to highlight what are the main components and interactions for each architecture, and the main advantages and drawbacks.
+This should also be stated in the introduction.}
 

### 2016-04-26 paper.tex
Full read with minors notes

--- 2016-04-25 paper.tex
@@
 \usepackage{pbox}
-\usepackage{pdfcomment}
 
@@
 \newcommand{\todo}[1]{\color{blue}\xspace[\emph{#1}]\xspace\color{black}}
-\newcommand{\note}[2]{\pdfmargincomment[color=yellow,author=#1,open=true]{#2}}
-\newcommand{\closednote}[4]{\pdfmargincomment[color=gray,author=#1-#3,open=true]{#1: #2 \textCR \textCR #3: #4}}
+\newcommand{\note}[2]{\textbf{\Large{\color{blue}\footnote{{\color{blue}\textbf{#1:} \textit{#2}\color{black}}}}}}
 
 \renewcommand\thesubfigure{\Roman{subfigure}}
-
-\journal{Future Generation Computer Systems}
 
@@
 
-\title{Software Architectures to Integrate Workflow Engines in Science Gateways}
+\title{Software architectures to integrate workflow engines in science gateways}
 
@@
 
-\address[mcgill]{McGill Centre for Integrative Neuroscience, Montreal Neurological Institute, McGill University, Canada.}
+\note{Marc-Etienne}{Do we want to use MCIN instead of BIC?}
+\address[mcgill]{McConnell Brain Imaging Centre, Montreal Neurological Institute, McGill University, Canada.}
 \address[creatis]{University of Lyon, CNRS, INSERM, CREATIS, Villeurbanne, France.}
@@
 
-%\maketitle
+\journal{Future Generation Computer Systems}
+
+\maketitle
 
@@
 IPython, ssh and multi-processor servers, and PSOM is interfaced to
-Torque, SGE, MOAB, LSF, and HTCondor. Some of these tools also support
+Torque, SGE, MOAB, LSF and Condor. Some of these tools also support
 advanced features such as provenance tracking (Nipype, PSOM),
@@
 
-Architectures showing reduced system complexity will greatly
-facilitate the implementation effort towards robust interactions
-which, in turn, have a positive impact on characteristics such as
-gateway predictability, transparence, general reliability and results
-traceability and reproducibility; characteristics that are also key to
-a good user experience. \closednote{Marc-Etienne}{I find that
-  something is weak in this last paragraph. Robustness will help with
-  transparency, sure, but it seems only a small facet.  Also, the text
-  seem to timidly favor architectures with less
-  components/interactions without much explanation... Complex systems
-  can be robust, but at greater costs...  Here is one: "Architectures
-  showing reduced system complexity will greatly facilitate the
-  implementation effort towards robust interactions which, in turn,
-  have a positive impact on characteristics such as gateway
-  predictability, transparence, general reliability and results
-  traceability and reproducibility; characteristics that are also key
-  to a good user experience."}{Tristan}{Agreed, fixed as suggested.}
+Robustness is a key part of science gateway user experience. In this
+respect, the multiplicity of software components and interactions
+involved in architectures is a challenge as it often leads to complex
+systems that are therefore prone to failure. To offer good user
+experience, science gateways should remain as transparent as possible,
+which requires robustness.
+\note{Marc-Etienne}{I find that something is weak in this last
+paragraph. Robustness will help with transparency, sure, but it seems only a small facet.
+Also, the text seem to timidly favor architectures with less components/interactions
+without much explanation... Complex systems can be robust, but at greater costs...
+Here is one:
+"Architectures showing reduced system complexity will greatly facilitate the implementation
+effort towards robust interactions which, in turn, have a positive impact on characteristics
+such as gateway predictability, transparence, general reliability and results traceability and
+reproducibility; characteristics that are also key to a good user experience."}
 
@@
 using regular web development tools and libraries to access
-distributed systems. CBRAIN and VIP are two such science gateways,
+distributed systems. CBRAIN and VIP are two of such science gateways,
 from which we gathered experience about the architectures described in
@@
 DIRAC pilot-job framework~\cite{tsaregorodtsev2008dirac} to execute
-applications on EGI. Almost 900 users were registered in VIP by
+applications on EGI. Almost 900 users were registered in VIP as of
 April 2016 and the average CPU consumption between January 2013 and
@@
 SHIWA Simulation Platform is a WS-PGRADE/gUSE science gateway that was
-developed in the European projects
+developed in European projects
 SHIWA\footnote{\url{http://www.shiwa-workflow.eu}} and
@@
 diagrammed in Figure~\ref{fig:architectures} using the graphical
-notations shown in Figure~\ref{fig:notations}.  \closednote{Marc-e}{A
-  few things bother me a bit with the notation fig., making it harder
-  to assimilate 1- The interactions are padded in the middle of other
-  stuff (Administrator, Data) 2- The Software component example says
-  "Science gateway", should it not be empty?  3- The red arrow to
-  explain the use of red; it looks like a very specific type of
-  interactions instead of the desired "any interactions or components
-  can be red meaning...", especially with 'a' and 'd' which the reader
-  does not know the meaning yet.  Maybe a little red arrow and square
-  with "Red used when specific to workflow execution"?  I don't have a
-  great solution...  Red has special meaning, maybe remove it from the
-  other pics (workflow eng, tasks) as it creates lots of visual
-  interference when looking for red?  4- Order in text jumps all over
-  the fig notations. Explanations are given for, in order: Components,
-  workflow engine, interactions, abstract interactions, red color,
-  tasks, data (no Administrator?)...  Could re-ordering the fig fix
-  this? A figure figure legend could go a long way too (are there some
-  in this journal?)  }{Tristan}{Marc-E, is Figure 1 better now?}
+notations shown in Figure~\ref{fig:notations}.
+\note{Marc-e}{A few things bother me a bit with the notation fig., making it harder to assimilate
+1- The interactions are padded in the middle of other stuff (Administrator, Data)
+2- The Software component example says "Science gateway", should it not be empty?
+3- The red arrow to explain the use of red; it looks like a very specific type
+of interactions instead of the desired "any interactions or components can be red meaning...",
+especially with 'a' and 'd' which the reader does not know the meaning yet.
+Maybe a little red arrow and square with "Red used when specific to workflow execution"?
+I don't have a great solution...
+Red has special meaning, maybe remove it from the other pics (workflow eng, tasks)
+as it creates lots of visual interference when looking for red?
+4- Order in text jumps all over the fig notations. Explanations are given for, in order:
+Components, workflow engine, interactions, abstract interactions, red color, tasks, data (no Administrator?)...
+Could re-ordering the fig fix this? A figure figure legend could go a long way too (are there some in this journal?)
+}
 
@@
 workflow pool and agent are involved in some architectures only. The
-workflow engine itself is represented by a specific symbol enclosed in a dotted rectangle.
-
-Software interactions among these components are represented with
-labeled, plain black arrows.  \emph{Abstract} interactions are
-specific types of interactions that may be implemented using different
-components and interactions. They are represented with dotted arrows
-and labeled with \texttt{*}. An architecture that has an abstract
-interaction is an abstract architecture that can be instantiated in
-various ways.
+workflow engine itself is represented by a specific symbol with dotted
+lines.
+
+Software interactions among these components are represented with labeled
+arrows.  \emph{Abstract} interactions are specific types of
+interactions that may be implemented using different components and
+interactions. They are represented with dotted arrows and labeled with
+\texttt{*}. An architecture that has an abstract interaction is an
+abstract architecture that can be instantiated in various ways.
 
@@
 infrastructure. It does not cover workflow parameters such as strings,
-numbers, etc. Administrator refers to a science gateway user with
-extended privileges to install applications.
+numbers, etc.
 
@@
 \centering
-\def\svgwidth{0.7\columnwidth}
+\def\svgwidth{\columnwidth}
 \input{figures/notations.pdf_tex}
@@
 \centering
-\footnotesize
 \rowcolors{1}{white}{gray!25}
@@
   Pool & SHIWA pool~\cite{ROGE-13}\\
-  Nested & \pbox{1.5\columnwidth}{SHIWA Simulation Platform (Coarse-Grained Interoperability~\cite{terstyanszky2014enabling})$^w$, HubZero with Pegasus (via hierarchical workflows)~\cite{Deelman201517} \todo{others}
-  }\\
+  Nested & SHIWA Simulation Platform (Coarse-Grained Interoperability~\cite{terstyanszky2014enabling})$^w$\todo{others}\\
   Import & SHIWA Simulation Platform (Fine-Grained Interoperability)~\cite{plankensteiner-prodan-etal:2013}$^w$
 \end{tabular}
-\caption{Classification of science gateways based on the architecture used to integrate workflow engines.
+\caption{Classification of science gateway based on the architecture used to integrate workflow engines.
 $^c$: based on the Catania Science Gateway Framework.
@@
   integrating a workflow is not the same process as integrating a
-  workflow \emph{engine}.
-
+  workflow \emph{engine}. \note{Marc-e}{automation?}
 \item Task control: operations to manage tasks on the infrastructure,
@@
   Applications~\cite{goodale2006saga}), DRMAA (Distributed Resource
-  Management Application API~\cite{troger2012distributed}), or OCCI
+  Management Application API~\cite{troger2012distributed}) or OCCI
   (Open Cloud Computing Interface~\cite{edmonds2012toward}).
@@
   use it. The infrastructure might offer various data storage backends
-  with heterogeneous interfaces. \closednote{Marc-Etienne}{this -c2- can be also trigerred
-    by the user}{Tristan}{No, it can't: this is by definition an interaction between the workflow engine and the infrastructure.
+  with heterogeneous interfaces. \note{Marc-Etienne}{this -c2- can be also trigerred
+    by the user}\note{Tristan}{No, it can't: this is by definition an interaction between the workflow engine and the infrastructure.
 It is used to transfer data to/from tasks.
-I reworded a bit, is it clearer now?} Tools and services such as
-  JSAGA~\cite{reynaud2010uniform} or Data Avenue~\cite{hajnal2014data}
+I reworded a bit, is it clearer now?}
+    \note{Marc-e}{I meant, "using c1", a user in some platforms (CBRAIN), can trigger
+    some operations across the infrastructure, as in c2... not just up/download data.
+    And I see now that this is in fact the case in III, c is not differentiated.
+    You can delete this thread.}
+  Tools and services such as JSAGA~\cite{reynaud2010uniform} or Data Avenue~\cite{hajnal2014data}
   can be used to homogeneize these interfaces.
@@
   interaction \texttt{b}, except that information about the parent
-  task, which submitted a sub-task, is usually available and used for
+  task which submitted a sub-task is usually available and used for
   additional control. For instance, the parent task may wait for all
-  its sub-tasks to complete before finishing, and conversely all the
-  sub-tasks may be terminated if the parent task is killed.
+  its sub-tasks to finish before finishing, and conversely all the
+  sub-tasks may be killed if the parent task is killed.
 \item Pool-agent: specific to the pool architecture described in
@@
   well-structured and relatively simple workflow language such as between GWorkflowDL
-  and Scufl~\cite{OLAB-09}, and between the 4 languages that were involved in the
+  and Scufl~\cite{OLAB-09} and between the 4 languages that were involved in the
   SHIWA initiative. In SHIWA, workflow conversion is done through an
@@
 gateway usually share a database where application, users and other
-resources are stored\closednote{sorina}{VIP and Moteur also share the
-  WorkflowsDB}{Tristan}{Yes, you are right. This is an example where
-  such abstract architectures may not completely apply in real
-  life. See comment about ``blended'' architectures in the
-  discussion. In case of VIP, the sharing of the DB is more an
-  implementation detail rather than a strong architectural pattern: we
-  could avoid sharing the DB if we wanted.}.  In this model, task and
-data control are both initiated from the science gateway. Interactions
-\texttt{b} and \texttt{c$_2$} are initiated from the workflow engine
-while \texttt{c$_1$} comes from other parts of the science gateway,
-for instance a data management interface. As in any other model, the
+resources are stored. In this model, task and data control are both
+initiated from the science gateway. Interactions \texttt{b} and
+\texttt{c$_2$} are initiated from the workflow engine while
+\texttt{c$_1$} comes from other parts of the science gateway, for
+instance a data management interface. As in any other model, the
 installation of new workflows in the science gateway (\texttt{a$_2$})
@@
 controls the service through a specific interaction (\texttt{d}) that
-might be implemented as a web-service call (e.g., RESTful or SOAP), as
+might be implemented as a web-service call (e.g. RESTful or SOAP), as
 a command-line or as any other method that offers a well-defined
@@
 framework~\cite{Kacsuk2012}. The integration between the
-HubZero science gateway and the Pegasus workflow engine performed
+HubZero science gateway and Pegasus workflow engine performed
 in~\cite{CPE:CPE3257} also uses a service architecture where the
-\texttt{d} interaction is implemented as a set of calls to Pegasus'
-command-line tools (e.g., pegasus-status, pegasus-plan,
-etc.). Figure~\ref{fig:vip-architecture} shows the architecture of the
+\texttt{d} interaction is implemented as a set of calls to Pegasus
+command-line tools (e.g. pegasus-status, pegasus-statistics,
+etc). Figure~\ref{fig:vip-architecture} shows the architecture of the
 Virtual Imaging Platform, where the MOTEUR workflow
@@
 submission (\texttt{1} \& \texttt{2}), workflow status retrieval
-(\texttt{0} \& \texttt{3})\closednote{sorina}{I suppose letters come here 
-from the initial pool-model figure; it's a bit puzzling when mixed with 
-letters from Fig 2}{Tristan}{Fixed.}, and workflow retrieval (\texttt{13} \&
-\texttt{14}). In Figure~\ref{fig:shiwa-pool-architecture},
+(\texttt{a} and \texttt{b}), and workflow retrieval (\texttt{13} \&
+\texttt{14}). In Figure ~\ref{fig:shiwa-pool-architecture},
 interaction \texttt{f} of Figure~\subref*{archi:agent} is implemented
@@
 collapsed in a single one with a \texttt{d} interaction with
-itself. An example of such interaction is the use of hierarchical
-workflows in Pegasus~\cite{Deelman201517}.
-Workflow engines communicate with infrastructures using
+itself. Workflow engines communicate with infrastructures using
 \texttt{b} and \texttt{c$_2$}. Science gateway transfers user data to
@@
 
-See Figure~\subref*{archi:import}. This is an abstract model 
-instantiated with the service invocation architecture for
+See Figure~\subref*{archi:import}. This is an abstract model that we
+instantiate with the service invocation architecture for
 consistency. Workflows are integrated in the science gateway through
@@
 evaluate the architectures: integration effort, robustness of workflow
-execution, extensibility, scalability and other specific
-features. Criteria break down to specific metrics where \emph{lower
-  value indicates better performance}. For each criterion, a total
-score is computed by summing up the individual metrics. We ensure that
-the different metrics in a criterion measure similar entities so that
-they can be summed. The criteria and metrics are explained hereafter.
+execution, extensibility, scalability and other specific features. Criteria
+break down to specific metrics where \emph{low value indicates good
+  performance}. For each criterion, a total score is computed by
+summing up the individual metrics. We ensure that the different
+metrics in a criterion measure similar entities so that summation
+makes sense. The
+criteria and metrics are explained hereafter.
 
 \begin{table*}
-\footnotesize
 \centering
@@
 \begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
-\item Total number of components (\texttt{I$_1$}), and
+\item Total number of components (\texttt{I$_1$}),
 \item Total number of interactions (\texttt{I$_2$}).
@@
 infrastructure in the architecture usually requires some technical
-effort (e.g., account creation, software installation, etc.), which is
+effort (e.g. account creation, software installation, etc) which is
 why we keep it in the metric. The two metrics are summed to obtain the
@@
 interactions in the software architecture. Note that errors coming
-from the infrastructure (e.g., unavailable data or terminated tasks) or
+from the infrastructure (e.g. unavailable data or terminated tasks) or
 workflows (e.g., wrong user input or application errors) are not
@@
 components and interactions that are specific to workflow execution,
-which assumes that complex architectures tend to be prone to failure:
+which assumes that complex architectures tend to be more prone to failure:
 \begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
@@
 criterion, which measures the total number of software pieces that are
-specifically involved in the workflow execution.
+specifically involved in workflow execution.
 
@@
   or interactions. We measure the difficulty to integrate a new
-  workflow by counting the number of interactions required
-  to integrate a new workflow in the architecture, assuming
+  workflow by counting the number of interactions that need to be
+  invoked to integrate a new workflow in the architecture, assuming
   that the engine type and version required to execute this workflow
@@
   computing or storage power, to access specific types of resources
-  (e.g., GPUs, clouds), or to enforce execution policies 
-  (e.g., constrain data to remain in a particular network domain).\note{Rafael}{Adding a new engine type may also allow run in different infrastructures.
-Maybe it is worth to put a sentence to clarify that the metric account for it separately.}
+  (e.g. GPUs, clouds), or to enforce execution policies
+  (e.g. constrain data to remain in a particular network domain).
 %\item New science gateway: \todo{Not sure if this makes sense}.
@@
   more than 1 engine instance in the architecture. Workflow engines
-  may require important amounts of resources when several workflows,
+  may require important amounts of resources when a lot of workflows,
   or large workflows, are executed. At some point, it may be required
-  for the science gateway to distribute the load among several
+  for the science gateway to distribute the load to several
   engines. \texttt{S$_1$}=0 when adding a new engine instance is a
@@
   adding a new engine instance requires some form of manual
-  intervention (e.g., two-factor authentication), which prevents easy 
-  implementations of elastic engines. \texttt{S$_1$}=2 when new engine 
-  instances cannot be added.
+  intervention, which prevents easy implementations of elastic
+  engines. \texttt{S$_1$}=2 when new engine instances cannot be added.
 \item Distributed engines (\texttt{S$_2$}): possibility to distribute
@@
   engine. \texttt{S$_2$}=0 when distributed engines are possible in
-  the architecture, and \texttt{S$_2$}=1 when they require specific
-  developments in the workflow engine.\note{Rafael}{$S_2$ would be 2 if the paradigm does not allow to distribute the orchestration?
-e.g., workflow executions implemented within MPI jobs, in Pegasus PMC and dispel4Py.}
+  the architecture and \texttt{S$_2$}=1 when they require specific
+  developments in the workflow engine.
 \item Task scheduling (\texttt{S$_3$}): task scheduling is a difficult
@@
   the architecture does not introduce any additional complexity to the
-  scheduling problem, and \texttt{S$_3$}=1 otherwise.
+  scheduling problem and \texttt{S$_3$}=1 otherwise.
 \end{itemize}
@@
   meta-workflows from existing workflows. Meta-workflows offer an
-  additional level of flexibility to build workflows from reusable
+  additional level of flexibility to build worfklows from reusable
   components. \texttt{O$_1$}=0 when the feature is available,
@@
 (\texttt{S$_1$}=2)\note{Tristan}{What-about load-balancing between web
-  server instances?}. Distributed engines are not available by default
+  server instances?} \note{Marc-e}{Would that really be considered a real architectural feature or an IT hack?}.
+  Distributed engines are not available by default
 (\texttt{S$_2$}=1). The scheduling of tasks on the infrastructure is
@@
 (\texttt{R$_1$}=0), and only interaction \texttt{e} is
-(\texttt{R$_2$}=1)\note{sorina}{I understand why this is 1 (and why interactions 
-c and b are not counted here), but I wonder whether we can simply ignore these interactions
-because they are not specific to workflow execution. As far as robustness is concerned,
-the transfers have to be successful anyway...}.
+(\texttt{R$_2$}=1).
 
@@
 workflow task and its sub-tasks may create deadlocks when there is
-contention. For instance, if only 1 computing resource is
+contention. Say for instance that only 1 computing resource is
 available for the science gateway and that the workflow task is
@@
 services. Sub-tasking and workflow import (with service invocation)
-perform a bit better when integrating new engines (\texttt{E$_1$}), which
+perform a bit better to integrate new engines (\texttt{E$_1$}), which
 is interesting given the current profusion of engines with different
 characteristics.  However, sub-tasking, workflow import (with service
-invocation) and tight integration perform worse than the others when adding
+invocation) and tight integration perform worse than the others to add
 engine versions (\texttt{E$_2$}) due to the need to update
-infrastructure (sub-tasking), science gateway (tight integration), or
+infrastructure \note{Marc-e}{confusion with physical?}(sub-tasking), science gateway (tight integration), or
 workflow converter (workflow import). Workflow import performs worse
-than the others when integrating new workflows (\texttt{E$_3$}) because of
+than the others to integrate new workflows (\texttt{E$_3$}) because of
 the language conversion step. All architectures except nested
-workflows perform the same when integrating new infrastructures (\texttt{E$_4$}).
+workflows perform the same to integrate new infrastructures (\texttt{E$_4$}).
 
@@
 the sub-tasking and nested workflow architectures could well become
-showstoppers depending on the type of the targeted
+showstoppers depending on the type of targeted
 infrastructure. Non-reliable infrastructures, for example, could
 hardly cope with workflow engines being wrapped in computing tasks as
-done in sub-tasking. The availability of multiple engine instances
+done in sub-tasking \note{Marc-e}{free for all? WE dev involved?}. The availability of multiple engine instances
 could also become a critical feature for science gateways with
@@
 also realize that this comes at the cost of realism. In particular,
-the presented architectures are pure patterns that may be blended in
+the presented architectures are theoritical patterns that may be blended in
 actual systems. The distinction between tight integration and service
-invocation, for instance, may not be that clear in practice. Service
-invocation may also be combined with sub-tasking in some cases.
+invocation, for instance, may not be always that clear in practice. Service
+invocation may also be combined with sub-tasking in some cases. However, the
+criteria discussed in this work would still apply to broadly compare and categorize
+such cases of hydrid architectures.
 
@@
 language. For instance, converting FSL, PSOM or Nipype pipelines to
-any other workflow language is very tricky because these engines rely
+any other workflow language is problematic because these engines rely
 on general-purpose programming languages such as Bash, Octave and
@@
 \texttt{c}), which would favor sub-tasking. The migration cost between
-architectures is totally ignored as well.
+architectures has been ignored as well.
 
@@
 Ludmer Family Foundation and the Ludmer Centre for Neuroinformatics
-and Mental Health.
+and Mental Health. \note{Marc-e}{Would it be proper to thank EGI? Compute Canada? CANARIE?}
 

### 2016-04-26 paper.tex

--- 2016-04-26 paper.tex
@@
   use it. The infrastructure might offer various data storage backends
-  with heterogeneous interfaces. \closednote{Marc-Etienne}{this -c2- can be also trigerred
+  with heterogeneous interfaces. \closenote{Marc-Etienne}{this -c2- can be also trigerred
     by the user}{Tristan}{No, it can't: this is by definition an interaction between the workflow engine and the infrastructure.
 It is used to transfer data to/from tasks.
-I reworded a bit, is it clearer now?} Tools and services such as
-  JSAGA~\cite{reynaud2010uniform} or Data Avenue~\cite{hajnal2014data}
+I reworded a bit, is it clearer now?}
+    {Marc-e}{I meant, "using c1", a user in some platforms (CBRAIN), can trigger
+    some operations across the infrastructure, as in c2... not just up/download data.
+    And I clearly see now that this is in fact the case in III, c is not differentiated. Closing this thread for good.}
+  Tools and services such as JSAGA~\cite{reynaud2010uniform} or Data Avenue~\cite{hajnal2014data}
   can be used to homogeneize these interfaces.
@@
 engine versions (\texttt{E$_2$}) due to the need to update
-infrastructure (sub-tasking), science gateway (tight integration), or
+infrastructure \note{Marc-e}{confusion with physical?}(sub-tasking), science gateway (tight integration), or
 workflow converter (workflow import). Workflow import performs worse
@@
 hardly cope with workflow engines being wrapped in computing tasks as
-done in sub-tasking. The availability of multiple engine instances
+done in sub-tasking \note{Marc-e}{free for all? WE dev involved?}. The availability of multiple engine instances
 could also become a critical feature for science gateways with

### 2016-04-26 paper.tex
1.2 Robustness
In 1.2, added back a 2 sentence intro about robustness.
A few updates

--- 2016-04-26 paper.tex
@@
 WS-PGRADE/gUSE~\cite{Kacsuk2012}. Numerous science gateway instances
-were built using such frameworks\note{Tristan}{Add references}. Other
-gateway instances were also built independently from any framework,
-using regular web development tools and libraries to access
-distributed systems. CBRAIN and VIP are two such science gateways,
-from which we gathered experience about the architectures described in
-this paper.
+were built using such
+frameworks~\cite{kacsuk2014science,ardizzone2012decide}. Other gateway
+instances were also built independently from any framework, using
+regular web development tools and libraries to access distributed
+systems. CBRAIN and VIP are two such science gateways, from which we
+gathered experience about the architectures described in this paper.
 
@@
             Apache Airvata~\cite{marru2011apache}, e-BioInfra~\cite{shahand2015data}$^{w,n}$, HubZero with Pegasus~\cite{CPE:CPE3257}, MoSGrid~\cite{krüger2014mosgrid}$^w$, System in~\cite{wu2010accelerating}, Vine Toolkit~\cite{DBLP:journals/scpe/SzejnfeldDKKKKLPTWDNW10}, Virtual Imaging Platform~\cite{GLAT-13}$^n$, WS-PGRADE/gUSE framework~\cite{Kacsuk2012},
-            Science gateways in ~\cite{kacsuk2014science}$^w$
+            Science gateways in~\cite{kacsuk2014science}$^w$
             } \\
@@
   Pool & SHIWA pool~\cite{ROGE-13}\\
-  Nested & \pbox{1.5\columnwidth}{SHIWA Simulation Platform (Coarse-Grained Interoperability~\cite{terstyanszky2014enabling})$^w$, HubZero with Pegasus (via hierarchical workflows)~\cite{Deelman201517} \todo{Tristan}{Add others}
+  Nested & \pbox{1.5\columnwidth}{SHIWA Simulation Platform (Coarse-Grained Interoperability~\cite{terstyanszky2014enabling})$^w$, HubZero with Pegasus (via hierarchical workflows)~\cite{Deelman201517}, Tavaxy~\cite{Abouelhoda2012}.
            }\\
-  Import & SHIWA Simulation Platform (Fine-Grained Interoperability)~\cite{plankensteiner-prodan-etal:2013}$^w$
+  Import & SHIWA Simulation Platform (Fine-Grained Interoperability)~\cite{plankensteiner-prodan-etal:2013}$^w$, Tavaxy~\cite{Abouelhoda2012}, system in~\cite{delaGarza2016}.
 \end{tabular}
@@
 workflows\footnote{\url{http://bit.ly/1oQrzvQ}}), in the Distributed
-application runtime environment (DARE~\cite{maddineni2012distributed})
-and in LONI Pipeline Environment~\cite{dinov2009efficient}.
+application runtime environment
+(DARE~\cite{maddineni2012distributed}), in
+Galaxy~\cite{goecks2010galaxy}, and in LONI Pipeline
+Environment~\cite{dinov2009efficient}. Note that tightly integrated
+architectures may provide advanced workflow edition features which are
+not covered by our analysis.
 %The Catania Science Gateway Framework would be a good illustration here (workflow engine is a Java portlet), but I can't find a suitable figure about it.
@@
 format. Workflow import is usually an offline process which is not
-involved in workflow execution. Workflow import was implemented in the SHIWA
-Simulation Platform through the IWIR language that provided a common
-language for portability across grid workflow
-systems~\cite{plankensteiner-prodan-etal:2013}.
-
-\todo{Tristan}{Add description of a real system}
-
-\todo{Tristan}{Refer to ``From the desktop to the grid: scalable bioinformatics via workflow conversion Luis de la GarzaEmail author, Johannes Veit, Andras Szolek, Marc Röttig, Stephan Aiche, Sandra Gesing, Knut Reinert and Oliver Kohlbacher''}
+involved in workflow execution. Workflow import was implemented in the
+SHIWA Simulation Platform through the IWIR language that provided a
+common language for portability across grid workflow
+systems~\cite{plankensteiner-prodan-etal:2013}.  Other systems rely on
+workflow import. Tavaxy~\cite{Abouelhoda2012} allows to import, merge
+and run Taverna~\cite{oinn2004taverna} and
+Galaxy~\cite{goecks2010galaxy} workflows. When some workflow parts
+cannot be imported, workflows are run by Tavaxy as nested workflows
+using their native engine. The work in~\cite{delaGarza2016} describes
+workflow import from KNIME~\cite{Berthold2008} to WS-PGRADE/gUSE and
+from Galaxy~\cite{goecks2010galaxy} to WS-PGRADE/gUSE.
+\closednote{Tristan}{Add description of a real system}{Tristan}{Would
+  require to reproduce a figure from another paper. The concept seems
+  simple enough to not require a figure.}
+
+%Furthermore, both systems provide web-portals allowing users to share and publish their workflows: These are the myExperiment portal for Taverna [23, 24] and the Public Pages for Galaxy [13].
 
@@
   are already available.
-\item New infrastructure (\texttt{E$_4$}): number of interactions or components to
-  modify to integrate a new type of infrastructure in the
-  architecture. Adding a new infrastructure allows to provide more
+\item New infrastructure (\texttt{E$_4$}): number of interactions or
+  components to modify to integrate a new type of infrastructure in
+  the architecture. Adding a new infrastructure allows to provide more
   computing or storage power, to access specific types of resources
-  (e.g., GPUs, clouds), or to enforce execution policies 
-  (e.g., constrain data to remain in a particular network domain).\note{Rafael}{Adding a new engine type may also allow run in different infrastructures.
-Maybe it is worth to put a sentence to clarify that the metric account for it separately.}
+  (e.g., GPUs, clouds), or to enforce execution policies (e.g.,
+  constrain data to remain in a particular network domain). 
 %\item New science gateway: \todo{Not sure if this makes sense}.
@@
 The 4 metrics are summed to obtain a global index that measures the
-difficulty to extend the architecture.
+difficulty to extend the architecture. Note that some extensions may
+involve several metrics in practice. For instance, adding a new type
+of engine may help integrate new infrastructures when interactions
+\texttt{b} and \texttt{c$_2$} are already present for the new engine
+and infrastructure.  \closednote{Rafael}{Adding a new engine type may
+  also allow run in different infrastructures. Maybe it is worth to
+  put a sentence to clarify that the metric account for it
+  separately.}{Tristan}{It will not *allow* because interaction c1
+  would still need to be developed. But it will definitely facilitate
+  it. I added a note on this after the description of the metrics.}
 
@@
 tested task and data control features (interactions \texttt{b} and
-\texttt{c}), which would favor sub-tasking. The migration cost between
-architectures is totally ignored as well.
+\texttt{c}), which would favor sub-tasking. Similarly, adding a new
+type of engine may facilitate integration of a new infrastructure when
+interactions \texttt{b} and \texttt{c$_2$} are already available. The
+migration cost between architectures is totally ignored as well.
 

### 2016-04-26 paper.tex

--- 2016-04-26 paper.tex
@@
 
+Robustness is a highly desirable characteristic in production gateways. In this respect, the 
+multiplicity of software components and interactions involved in architectures is
+a challenge as it often leads to complex systems.
 Architectures showing reduced system complexity will greatly
@@
   are already available.
-\item New infrastructure (\texttt{E$_4$}): number of interactions or components to
-  modify to integrate a new type of infrastructure in the
-  architecture. Adding a new infrastructure allows to provide more
+\item New infrastructure (\texttt{E$_4$}): number of interactions or
+  components to modify to integrate a new type of infrastructure in
+  the architecture. Adding a new infrastructure allows to provide more
   computing or storage power, to access specific types of resources
-  (e.g., GPUs, clouds), or to enforce execution policies
-  (e.g., constrain data to remain in a particular network domain).\note{Rafael}{Adding a new engine type may also allow run in different infrastructures.
-Maybe it is worth to put a sentence to clarify that the metric account for it separately.}
+  (e.g., GPUs, clouds), or to enforce execution policies (e.g.,
+  constrain data to remain in a particular network domain). 
 %\item New science gateway: \todo{Not sure if this makes sense}.
@@
 tested task and data control features (interactions \texttt{b} and
-\texttt{c}), which would favor sub-tasking. The migration cost between
-architectures has been ignored as well.
+\texttt{c}), which would favor sub-tasking. Similarly, adding a new
+type of engine may facilitate integration of a new infrastructure when
+interactions \texttt{b} and \texttt{c$_2$} are already available. The
+migration cost between architectures has been ignored as well.
 

### 2016-04-26 paper.tex
Typo
Fixed a few notes

@@
 \newcommand{\closednote}[4]{\pdfmargincomment[color=gray,author=#1-#3,open=true]{#1: #2 \textCR \textCR #3: #4}}
+\newcommand{\answerednote}[4]{\pdfmargincomment[color=green,author=#1-#3,open=true]{#1: #2 \textCR \textCR #3: #4}}
 
@@
   use it. The infrastructure might offer various data storage backends
-  with heterogeneous interfaces. \closenote{Marc-Etienne}{this -c2- can be also trigerred
+  with heterogeneous interfaces. \closednote{Marc-Etienne}{this -c2- can be also trigerred
     by the user}{Tristan}{No, it can't: this is by definition an interaction between the workflow engine and the infrastructure.
@@
   engine. \texttt{S$_2$}=0 when distributed engines are possible in
-  the architecture, and \texttt{S$_2$}=1 when they require specific
-  developments in the workflow engine.\note{Rafael}{$S_2$ would be 2 if the paradigm does not allow to distribute the orchestration?
-e.g., workflow executions implemented within MPI jobs, in Pegasus PMC and dispel4Py.}
+  the architecture, \texttt{S$_2$}=1 when they require specific
+  developments in the workflow engine..  \closednote{Rafael}{$S_2$
+    would be 2 if the paradigm does not allow to distribute the
+    orchestration? e.g., workflow executions implemented within MPI
+    jobs, in Pegasus PMC and dispel4Py.}{Tristan}{The problem is that
+    it is very difficult to define ``does not allow to''. For
+    instance, why couldn't you hack Pegasus PMC so much that it
+    becomes possible to distribute workflow executions in several MPI
+    jobs coordinated by an MPI job? Moreover, S2=2 is not used so I'd
+    prefer to leave it like this, unless you have a suggestion to
+    reflect this in the text.}
 \item Task scheduling (\texttt{S$_3$}): task scheduling is a difficult
@@
 instance of the science gateway, which is in general not possible
-(\texttt{S$_1$}=2)\note{Tristan}{What-about load-balancing between web
-  server instances?} \note{Marc-e}{Would that really be considered a real architectural feature or an IT hack?}.
-  Distributed engines are not available by default
-(\texttt{S$_2$}=1). The scheduling of tasks on the infrastructure is
-as complex as in any other architecture since the workflow engine
-might implement any kind of scheduling policy (\texttt{S$_3$}=0).
+(\texttt{S$_1$}=2)\answerednote{Tristan}{What-about load-balancing between web
+  server instances?}{Marc-e}{Would that really be considered a
+  real architectural feature or an IT hack?}.  Distributed engines are
+not available by default (\texttt{S$_2$}=1). The scheduling of tasks
+on the infrastructure is as complex as in any other architecture since
+the workflow engine might implement any kind of scheduling policy
+(\texttt{S$_3$}=0).
 
@@
 (\texttt{R$_1$}=0), and only interaction \texttt{e} is
-(\texttt{R$_2$}=1)\note{sorina}{I understand why this is 1 (and why interactions
-c and b are not counted here), but I wonder whether we can simply ignore these interactions
-because they are not specific to workflow execution. As far as robustness is concerned,
-the transfers have to be successful anyway...}.
+(\texttt{R$_2$}=1)\answerednote{sorina}{I understand why this is 1 (and why
+  interactions c and b are not counted here), but I wonder whether we
+  can simply ignore these interactions because they are not specific
+  to workflow execution. As far as robustness is concerned, the
+  transfers have to be successful anyway...}{Tristan}{We cannot just
+  ignore these interactions in the architecture because they are
+  required for the system to work. The transfers have to be successful
+  but data movements in the workflow will use the same code as data
+  movements done in the science gateway, which will make this
+  architecture more robust than service for instance. In VIP, for
+  instance, the code used in the GASW wrapper to transfer data (c2) is
+  completely independent from the ``file transfer'' module in the
+  platform (c1), which is detrimental to robustness}.
 
@@
 
-\paragraph{Integration effort} Setting up a nested workflow architecture with
-service invocation requires a science gateway, 2 workflow services and
-2 infrastructures (\texttt{I$_1$=5}). The architecture involves 11
-interactions (\texttt{I$_2$=11}): \texttt{a$_1$} (appears twice: once for each
-infrastructure), \texttt{a$_2$}, \texttt{b} (twice: once for each
-infrastructure), \texttt{c$_1$} (twice: once for each infrastructure),
-\texttt{c$_2$} (twice: once for each workflow engine and
-infrastructure) and \texttt{d} (twice: once for each workflow engine).
-\note{Rafael}{I like the idea of what has been discussed in the `integration effort', however I think the name may lead to misinterpretations.
-For example, according to Table 2, it requires more effort to integrate in a Tight architecture than in a service-based architecture.
-Considering the principles of software engineer, it is simpler to integrate/maintain a system using decoupled services.
-Maybe it should be renamed to something related to 'dependencies'.}
+\paragraph{Integration effort} Setting up a nested workflow
+architecture with service invocation requires a science gateway, 2
+workflow services and 2 infrastructures (\texttt{I$_1$=5}). The
+architecture involves 11 interactions (\texttt{I$_2$=11}):
+\texttt{a$_1$} (appears twice: once for each infrastructure),
+\texttt{a$_2$}, \texttt{b} (twice: once for each infrastructure),
+\texttt{c$_1$} (twice: once for each infrastructure), \texttt{c$_2$}
+(twice: once for each workflow engine and infrastructure) and
+\texttt{d} (twice: once for each workflow engine).
+\answerednote{Rafael}{I like the idea of what has been discussed in
+  the `integration effort', however I think the name may lead to
+  misinterpretations. For example, according to Table 2, it requires
+  more effort to integrate in a Tight architecture than in a
+  service-based architecture. Considering the principles of software
+  engineer, it is simpler to integrate/maintain a system using
+  decoupled services. Maybe it should be renamed to something related
+  to 'dependencies'.}{Tristan}{In our scope I'm not sure that it's
+  simpler to integrate a service architecture than a tightly
+  integrated one. But I take your point. 'Dependencies' is not very
+  suitable either. What about 'development effort'?}
 
@@
 services. Sub-tasking and workflow import (with service invocation)
-perform a bit better when integrating new engines (\texttt{E$_1$}), which
-is interesting given the current profusion of engines with different
-characteristics.  However, sub-tasking, workflow import (with service
-invocation) and tight integration perform worse than the others when adding
-engine versions (\texttt{E$_2$}) due to the need to update
-infrastructure \note{Marc-e}{confusion with physical?}(sub-tasking), science gateway (tight integration), or
+perform a bit better when integrating new engines (\texttt{E$_1$}),
+which is interesting given the current profusion of engines with
+different characteristics.  However, sub-tasking, workflow import
+(with service invocation) and tight integration perform worse than the
+others when adding engine versions (\texttt{E$_2$}) due to the need to
+update infrastructure \note{Marc-e}{confusion with
+  physical?}(sub-tasking), science gateway (tight integration), or
 workflow converter (workflow import). Workflow import performs worse
-than the others when integrating new workflows (\texttt{E$_3$}) because of
-the language conversion step. All architectures except nested
-workflows perform the same when integrating new infrastructures (\texttt{E$_4$}).
+than the others when integrating new workflows (\texttt{E$_3$})
+because of the language conversion step. All architectures except
+nested workflows perform the same when integrating new infrastructures
+(\texttt{E$_4$}).
 
@@
 invocation) are the only architectures that support meta-workflows,
-which may be interesting in some cases.\note{Rafael}{I would suggest to provide some
-more examples when this could be interesting.}
-The availability of fine-grained debugging information may also be critical to efficient
-user support.
+which may be interesting in some cases. Also, sub-tasking and nested
+workflows may complexify fine-grained debugging. \closednote{Rafael}{I would
+  suggest to provide some more examples when this could be
+  interesting.}{Tristan}{Done.}  The availability of fine-grained debugging
+information may also be critical to efficient user support.
 
@@
 architectures, based on the metrics in Table~\ref{table:evaluation}.
-Overall, sub-tasking, and workflow import (with service invocation)
-perform a bit better than tight integration, service invocation and
-pool. Nested workflows stands a bit behind for the reasons mentioned
-previously.\note{Rafael}{I would emphasize that this analysis is only for illustration purposes, since it does not account for different weights in metrics.}
+Note that this analysis is only for illustration purposes since it
+assumes that all criteria have equal weights while real systems would
+favor some of them. Overall, sub-tasking, and workflow import (with
+service invocation) perform a bit better than tight integration,
+service invocation and pool. Nested workflows stands a bit behind for
+the reasons mentioned previously.\closednote{Rafael}{I would emphasize
+  that this analysis is only for illustration purposes, since it does
+  not account for different weights in metrics.}{Tristan}{Done.}
 \begin{table*}
@@
 metrics for all the architectures, and we discussed the pros and cons
-of all the architectures based on these metrics.
-\note{Rafael}{One thing that is missing in the conclusion or the limitations section is to state that this work does not intend to quantify the amount of work required to develop science gateways using a specific architecture, but the goal is to highlight what are the main components and interactions for each architecture, and the main advantages and drawbacks.
-This should also be stated in the introduction.}
+of all the architectures based on these metrics.  \note{Rafael}{One
+  thing that is missing in the conclusion or the limitations section
+  is to state that this work does not intend to quantify the amount of
+  work required to develop science gateways using a specific
+  architecture, but the goal is to highlight what are the main
+  components and interactions for each architecture, and the main
+  advantages and drawbacks. This should also be stated in the
+  introduction.}
 

### 2016-04-26 paper.tex
Answered a note

@@
 metrics for all the architectures, and we discussed the pros and cons
-of all the architectures based on these metrics.  \note{Rafael}{One
+of all the architectures based on these metrics.  \answerednote{Rafael}{One
   thing that is missing in the conclusion or the limitations section
@@
   advantages and drawbacks. This should also be stated in the
-  introduction.}
+  introduction.}{Tristan}{This is in contradiction with the metric on
+  'integration effort'. What do you think?}
 

### 2016-04-26 paper.tex
Fixed \closednote command
Added Rafael as a co-author. Updates in acknowledgement.

@@
 \newcommand{\note}[2]{\pdfmargincomment[color=yellow,author=#1,open=true]{#2}}
-\newcommand{\closednote}[4]{\pdfmargincomment[color=gray,author=#1-#3,open=true]{#1: #2 \textCR \textCR #3: #4}}
+\newcommand{\closednote}[4]{\pdfmargincomment[color=gray,author=#1-#3,open=false]{#1: #2 \textCR \textCR #3: #4}}
 \newcommand{\answerednote}[4]{\pdfmargincomment[color=green,author=#1-#3,open=true]{#1: #2 \textCR \textCR #3: #4}}
@@
 \author[mcgill]{Natacha Beck}
+\author[isi]{Rafael Ferreira da Silva}
 \author[mcgill]{Najmeh Khalili-Mahani}
@@
 \address[criugm]{Centre de Recherche de l'Institut de G\'eriatrie de Montr\'eal CRIUGM, Montreal, QC, Canada.}
+\address[isi]{University of Southern California, Information Sciences Institute, Marina del Rey, CA, USA.}
 
@@
 
-We warmly thank Rafael Ferreira da Silva for its contribution to the
-VIP platform and for creating Figure~\ref{fig:vip-architecture}. This
-work is in the scope of the LABEX PRIMES (ANR-11- LABX-0063) of
+This work is in the scope of the LABEX PRIMES (ANR-11- LABX-0063) of
 Universit\'e de Lyon, within the program ``Investissements d’Avenir''
 (ANR-11-IDEX-0007) operated by the French National Research Agency
-(ANR). This work has been made possible with the support of the Irving
-Ludmer Family Foundation and the Ludmer Centre for Neuroinformatics
-and Mental Health. \note{Marc-e}{Would it be proper to thank EGI? Compute Canada? CANARIE?}
+(ANR). This work also falls into the scope of the scientific topics of
+the French National Grid Institute (IdG). We thank the site
+administrators of the European Grid Initiative and the GGUS support
+for their help related to the VIP platform. We are grateful for the
+computing cycles, storage, and support obtained from Compute Canada
+(\url{https://computecanada.ca}) for CBRAIN. This work has been made
+possible with the support of the Irving Ludmer Family Foundation and
+the Ludmer Centre for Neuroinformatics and Mental
+Health. \answerednote{Marc-e}{Would it be proper to thank EGI?
+  Compute Canada? CANARIE?}{Tristan}{Added EGI and Compute Canada. Not sure if
+  Canarie should be added. What about Calcul QC?}
 

### 2016-04-26 paper.tex
Fixed issue #8.

@@
 \newcommand{\answerednote}[4]{\pdfmargincomment[color=green,author=#1-#3,open=true]{#1: #2 \textCR \textCR #3: #4}}
+\newcommand{\closedanswerednote}[6]{\pdfmargincomment[color=gray,author=#1-#3-#5,open=true]{#1: #2 \textCR \textCR #3: #4 \textCR \textCR #5: #6}}
 
@@
 \begin{abstract}
-  We investigate software architectures commonly used to integrate workflow engines
-  in science gateways. Based on our experience with the development
-  and operations of the CBRAIN and VIP science gateways, and our past
-  involvement in the SHIWA project, we abstract software components
-  and interactions in 6 architectures that we use to classify 18
-  systems. We compare these architectures using metrics measuring
-  integration effort, robustness, extensibility, scalability and other
-  specific features. Results show the pros and cons of the different
-  architectures: tight integration and sub-tasking require the least
-  amount of integration effort and are also the most robust, most
-  architectures perform the same in terms of extensibility, the pool
-  model is the most scalable and meta-workflows are only available in
-  nested workflows and workflow import. These results provide insights
-  for software architects on how to integrate science gateways and
-  workflow engines.
+  We investigate software architectures commonly used to integrate
+  workflow engines in science gateways. Based on our experience with
+  the development and operations of the CBRAIN and VIP science
+  gateways, and our past involvement in the SHIWA project, we abstract
+  software components and interactions in 6 architectures that we use
+  to classify 18 systems. We compare these architectures using metrics
+  measuring global complexity, robustness, extensibility, scalability
+  and other specific features. Results show the pros and cons of the
+  different architectures: tight integration and sub-tasking are the
+  simplest architectures and should therefore be the easiest to
+  implement and the most robust ones, most architectures perform the
+  same in terms of extensibility, the pool model is the most scalable
+  and meta-workflows are only available in nested workflows and
+  workflow import. These results provide insights for software
+  architects on how to integrate science gateways and workflow
+  engines.
 \end{abstract}
@@
 Section~\ref{sec:evaluation}, we introduce metrics to measure
-integration effort, robustness, extensibility, scalability and other
+global complexity, robustness, extensibility, scalability and other
 specific features of the architectures. We use these metrics to
@@
 evaluated in Table~\ref{table:evaluation}. We use 5 main criteria to
-evaluate the architectures: integration effort, robustness of workflow
-execution, extensibility, scalability and other specific
-features. Criteria break down to specific metrics where \emph{lower
-  value indicates better performance}. For each criterion, a total
-score is computed by summing up the individual metrics. We ensure that
-the different metrics in a criterion measure similar entities so that
-they can be summed. The criteria and metrics are explained hereafter.
+evaluate the architectures: global complexity, execution complexity,
+extensibility, scalability and other specific features. Criteria break
+down to specific metrics where \emph{lower value indicates better
+  performance}. For each criterion, a total score is computed by
+summing up the individual metrics. We ensure that the different
+metrics in a criterion measure similar entities so that they can be
+summed. The criteria and metrics are explained hereafter.
 
@@
                                      & \textbf{Import} \\
-\cellcolor[HTML]{EEEEEE}\textbf{Integration effort}& \multicolumn{6}{l}{\cellcolor[HTML]{EEEEEE}}\\
+\cellcolor[HTML]{EEEEEE}\textbf{Global complexity}& \multicolumn{6}{l}{\cellcolor[HTML]{EEEEEE}}\\
   Total components -- \texttt{I$_1$} & \cellcolor[HTML]{99FF99}2
@@
 
-\emph{Integration effort} measures the development required to build
-the architecture. It is obtained by counting the total number of
+\emph{Global complexity} is obtained by counting the total number of
 interactions and components on the architecture diagram. It breaks
@@
 why we keep it in the metric. The two metrics are summed to obtain the
-total score for integration effort, which measures the total number of
-software pieces to develop (pieces include interactions and
-components).
+total score for global complexity. Global complexity measures the
+total number of software pieces to develop, including interactions and
+components. It may be used to measure the development effort required
+to build the architecture.
 
@@
 interactions in the software architecture. Note that errors coming
-from the infrastructure (e.g., unavailable data or terminated tasks) or
-workflows (e.g., wrong user input or application errors) are not
+from the infrastructure (e.g., unavailable data or terminated tasks)
+or workflows (e.g., wrong user input or application errors) are not
 covered since they do not stem directly from the software
-architecture. Robustness is determined as the number of software
-components and interactions that are specific to workflow execution,
-which assumes that complex architectures tend to be more prone to failure:
+architecture. Robustness is measured here as a consequence of global
+complexity since complex architectures tend to be more prone to
+failure. More precisely, robustness is determined as the number of
+software components and interactions that are specific to workflow
+execution:
 \begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
@@
 \end{itemize}
-These specific components and interactions are represented in red in
-the architecture diagrams in
-Figure~\ref{fig:architectures}. \texttt{R$_1$} and \texttt{R$_2$}  are summed to obtain the total score for this
-criterion, which measures the total number of software pieces that are
-specifically involved in the workflow execution.
+These specific components and interactions are the ones represented in
+red in the architecture diagrams in
+Figure~\ref{fig:architectures}. \texttt{R$_1$} and \texttt{R$_2$} are
+summed to obtain the total score for this criterion, which measures
+the total number of software pieces that are specifically involved in
+the workflow execution.
 
 \emph{Extensibility} measures the difficulty to replace or add
-elements in the architecture. It is determined as the number of
-interactions and components that need to be modified when a new
-element is added. Modification of a component is required when its
-code needs to be updated or recompiled (science gateway or workflow
-service), or when a new piece of software has to be installed
-(infrastructure only). Modification of an interaction is deemed
-necessary when the parameters involved in this interaction are
-modified.  Extensibility breaks down in 4 metrics depending on the
-type of element that has to be added or replaced:
+elements in the architecture. It is also a consequence of global
+complexity and it is determined as the number of interactions and
+components that need to be modified when a new element is
+added. Modification of a component is required when its code needs to
+be updated or recompiled (science gateway or workflow service), or
+when a new piece of software has to be installed (infrastructure
+only). Modification of an interaction is deemed necessary when the
+parameters involved in this interaction are modified.  Extensibility
+breaks down in 4 metrics depending on the type of element that has to
+be added or replaced:
 \begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
@@
 
-\paragraph{Integration effort} This architecture does not require any
+\paragraph{Global complexity} This architecture does not require any
 component in addition to the science gateway and infrastructure
@@
 
-\paragraph{Integration effort} Service invocation requires a workflow service
+\paragraph{Global complexity} Service invocation requires a workflow service
 in addition to the science gateway and infrastructure
@@
 
-\paragraph{Integration effort} The sub-task architecture requires only 2
+\paragraph{Global complexity} The sub-task architecture requires only 2
 components (\texttt{I$_1$}=2).  It involves 5 interactions:
@@
 
-\paragraph{Integration effort} The pool model requires a workflow pool and an
+\paragraph{Global complexity} The pool model requires a workflow pool and an
 agent in addition to the science gateway and infrastructure
@@
 
-\paragraph{Integration effort} Setting up a nested workflow
+\paragraph{Global complexity} Setting up a nested workflow
 architecture with service invocation requires a science gateway, 2
@@
 \texttt{d} (twice: once for each workflow engine).
-\answerednote{Rafael}{I like the idea of what has been discussed in
+\closedanswerednote{Rafael}{I like the idea of what has been discussed in
   the `integration effort', however I think the name may lead to
@@
   integrated one. But I take your point. 'Dependencies' is not very
-  suitable either. What about 'development effort'?}
+  suitable either. What about 'development effort'?}{Tristan}{Replaced
+  'Integration effort' with 'global complexity', as discussed in
+  github issue \#8.}
 
@@
 
-\paragraph{Integration effort} Workflow import with service invocation
+\paragraph{Global complexity} Workflow import with service invocation
 requires the same components as for service invocation
@@
 
-Tight integration and sub-tasking require the least amount of
-integration effort, followed by service integration, workflow import
-(with service invocation) and pool. Nested workflows (with service
+Tight integration and sub-tasking are globally the simplest
+architectures, followed by service integration, workflow import (with
+service invocation) and pool. Nested workflows (with service
 invocation) requires more integration than the other
@@
                                     & \textbf{Import} \\
-  Integration effort (normalized total) &
+  Global complexity (normalized total) &
                                     \cellcolor[HTML]{99FF99}{0.00}
@@
 Moreover, all the interactions involved in the architectures were
-treated equally, as if they all required similar amounts of
-development effort, which is realistically not the case. For example,
-interaction \texttt{g} (workflow language conversion) clearly requires
-more effort than interaction \texttt{d} (service invocation), and may
-create much more reliability issues as well. However, quantifying the
-robustness and amount of development associated with each interaction
-and software component cannot realistically be done without entering
-into the details of a particular system.
+treated equally while some of them are obviously more complex
+than others. For example, interaction \texttt{g} (workflow language
+conversion) is clearly more complex than interaction \texttt{d}
+(service invocation), and may create much more reliability issues as
+well. However, quantifying the robustness and amount of development
+associated with each interaction and software component cannot
+realistically be done without entering into the details of a
+particular system.
 
@@
 metrics for all the architectures, and we discussed the pros and cons
-of all the architectures based on these metrics.  \answerednote{Rafael}{One
-  thing that is missing in the conclusion or the limitations section
-  is to state that this work does not intend to quantify the amount of
-  work required to develop science gateways using a specific
-  architecture, but the goal is to highlight what are the main
-  components and interactions for each architecture, and the main
-  advantages and drawbacks. This should also be stated in the
-  introduction.}{Tristan}{This is in contradiction with the metric on
-  'integration effort'. What do you think?}
+of all the architectures based on these metrics.
+\closedanswerednote{Rafael}{One thing that is missing in the
+  conclusion or the limitations section is to state that this work
+  does not intend to quantify the amount of work required to develop
+  science gateways using a specific architecture, but the goal is to
+  highlight what are the main components and interactions for each
+  architecture, and the main advantages and drawbacks. This should
+  also be stated in the introduction.}{Tristan}{This is in
+  contradiction with the metric on 'integration effort'. What do you
+  think?}{Tristan}{Fixed, as discussed in github \#8.}
 

### 2016-04-26 paper.tex
Author addition
Samir should be added?

--- 2016-04-26 paper.tex
@@
 \newcommand{\closednote}[4]{\pdfmargincomment[color=gray,author=#1-#3,open=true]{#1: #2 \textCR \textCR #3: #4}}
-\newcommand{\answerednote}[4]{\pdfmargincomment[color=green,author=#1-#3,open=true]{#1: #2 \textCR \textCR #3: #4}}
 
@@
 \author[mcgill]{Alan C. Evans}
+
+\note{Marc-e}{Add Samir Das?}
 
@@
 
-Robustness is a highly desirable characteristic in production gateways. In this respect, the 
-multiplicity of software components and interactions involved in architectures is
-a challenge as it often leads to complex systems.
 Architectures showing reduced system complexity will greatly
@@
 WS-PGRADE/gUSE~\cite{Kacsuk2012}. Numerous science gateway instances
-were built using such
-frameworks~\cite{kacsuk2014science,ardizzone2012decide}. Other gateway
-instances were also built independently from any framework, using
-regular web development tools and libraries to access distributed
-systems. CBRAIN and VIP are two such science gateways, from which we
-gathered experience about the architectures described in this paper.
+were built using such frameworks\note{Tristan}{Add references}. Other
+gateway instances were also built independently from any framework,
+using regular web development tools and libraries to access
+distributed systems. CBRAIN and VIP are two such science gateways,
+from which we gathered experience about the architectures described in
+this paper.
 
@@
             Apache Airvata~\cite{marru2011apache}, e-BioInfra~\cite{shahand2015data}$^{w,n}$, HubZero with Pegasus~\cite{CPE:CPE3257}, MoSGrid~\cite{krüger2014mosgrid}$^w$, System in~\cite{wu2010accelerating}, Vine Toolkit~\cite{DBLP:journals/scpe/SzejnfeldDKKKKLPTWDNW10}, Virtual Imaging Platform~\cite{GLAT-13}$^n$, WS-PGRADE/gUSE framework~\cite{Kacsuk2012},
-            Science gateways in~\cite{kacsuk2014science}$^w$
+            Science gateways in ~\cite{kacsuk2014science}$^w$
             } \\
@@
   Pool & SHIWA pool~\cite{ROGE-13}\\
-  Nested & \pbox{1.5\columnwidth}{SHIWA Simulation Platform (Coarse-Grained Interoperability~\cite{terstyanszky2014enabling})$^w$, HubZero with Pegasus (via hierarchical workflows)~\cite{Deelman201517}, Tavaxy~\cite{Abouelhoda2012}.
+  Nested & \pbox{1.5\columnwidth}{SHIWA Simulation Platform (Coarse-Grained Interoperability~\cite{terstyanszky2014enabling})$^w$, HubZero with Pegasus (via hierarchical workflows)~\cite{Deelman201517} \todo{Tristan}{Add others}
   }\\
-  Import & SHIWA Simulation Platform (Fine-Grained Interoperability)~\cite{plankensteiner-prodan-etal:2013}$^w$, Tavaxy~\cite{Abouelhoda2012}, system in~\cite{delaGarza2016}.
+  Import & SHIWA Simulation Platform (Fine-Grained Interoperability)~\cite{plankensteiner-prodan-etal:2013}$^w$
 \end{tabular}
@@
   integrating a workflow is not the same process as integrating a
-  workflow \emph{engine}. \note{Marc-e}{automation?}
+  workflow \emph{engine}.
+
 \item Task control: operations to manage tasks on the infrastructure,
@@
 It is used to transfer data to/from tasks.
-I reworded a bit, is it clearer now?}
-    {Marc-e}{I meant, "using c1", a user in some platforms (CBRAIN), can trigger
-    some operations across the infrastructure, as in c2... not just up/download data.
-    And I clearly see now that this is in fact the case in III, c is not differentiated. Closing this thread for good.}
-  Tools and services such as JSAGA~\cite{reynaud2010uniform} or Data Avenue~\cite{hajnal2014data}
+I reworded a bit, is it clearer now?} Tools and services such as
+  JSAGA~\cite{reynaud2010uniform} or Data Avenue~\cite{hajnal2014data}
   can be used to homogeneize these interfaces.
@@
 workflows\footnote{\url{http://bit.ly/1oQrzvQ}}), in the Distributed
-application runtime environment
-(DARE~\cite{maddineni2012distributed}), in
-Galaxy~\cite{goecks2010galaxy}, and in LONI Pipeline
-Environment~\cite{dinov2009efficient}. Note that tightly integrated
-architectures may provide advanced workflow edition features which are
-not covered by our analysis.
+application runtime environment (DARE~\cite{maddineni2012distributed})
+and in LONI Pipeline Environment~\cite{dinov2009efficient}.
 %The Catania Science Gateway Framework would be a good illustration here (workflow engine is a Java portlet), but I can't find a suitable figure about it.
@@
 format. Workflow import is usually an offline process which is not
-involved in workflow execution. Workflow import was implemented in the
-SHIWA Simulation Platform through the IWIR language that provided a
-common language for portability across grid workflow
-systems~\cite{plankensteiner-prodan-etal:2013}.  Other systems rely on
-workflow import. Tavaxy~\cite{Abouelhoda2012} allows to import, merge
-and run Taverna~\cite{oinn2004taverna} and
-Galaxy~\cite{goecks2010galaxy} workflows. When some workflow parts
-cannot be imported, workflows are run by Tavaxy as nested workflows
-using their native engine. The work in~\cite{delaGarza2016} describes
-workflow import from KNIME~\cite{Berthold2008} to WS-PGRADE/gUSE and
-from Galaxy~\cite{goecks2010galaxy} to WS-PGRADE/gUSE.
-\closednote{Tristan}{Add description of a real system}{Tristan}{Would
-  require to reproduce a figure from another paper. The concept seems
-  simple enough to not require a figure.}
-
-%Furthermore, both systems provide web-portals allowing users to share and publish their workflows: These are the myExperiment portal for Taverna [23, 24] and the Public Pages for Galaxy [13].
+involved in workflow execution. Workflow import was implemented in the SHIWA
+Simulation Platform through the IWIR language that provided a common
+language for portability across grid workflow
+systems~\cite{plankensteiner-prodan-etal:2013}.
+
+\todo{Tristan}{Add description of a real system}
+
+\todo{Tristan}{Refer to ``From the desktop to the grid: scalable bioinformatics via workflow conversion Luis de la GarzaEmail author, Johannes Veit, Andras Szolek, Marc Röttig, Stephan Aiche, Sandra Gesing, Knut Reinert and Oliver Kohlbacher''}
 
@@
 components and interactions that are specific to workflow execution,
-which assumes that complex architectures tend to be more prone to failure:
+which assumes that complex architectures tend to be prone to failure:
 \begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
@@
   are already available.
-\item New infrastructure (\texttt{E$_4$}): number of interactions or
-  components to modify to integrate a new type of infrastructure in
-  the architecture. Adding a new infrastructure allows to provide more
+\item New infrastructure (\texttt{E$_4$}): number of interactions or components to
+  modify to integrate a new type of infrastructure in the
+  architecture. Adding a new infrastructure allows to provide more
   computing or storage power, to access specific types of resources
-  (e.g., GPUs, clouds), or to enforce execution policies (e.g.,
-  constrain data to remain in a particular network domain). 
+  (e.g., GPUs, clouds), or to enforce execution policies
+  (e.g., constrain data to remain in a particular network domain).\note{Rafael}{Adding a new engine type may also allow run in different infrastructures.
+Maybe it is worth to put a sentence to clarify that the metric account for it separately.}
 %\item New science gateway: \todo{Not sure if this makes sense}.
@@
 The 4 metrics are summed to obtain a global index that measures the
-difficulty to extend the architecture. Note that some extensions may
-involve several metrics in practice. For instance, adding a new type
-of engine may help integrate new infrastructures when interactions
-\texttt{b} and \texttt{c$_2$} are already present for the new engine
-and infrastructure.  \closednote{Rafael}{Adding a new engine type may
-  also allow run in different infrastructures. Maybe it is worth to
-  put a sentence to clarify that the metric account for it
-  separately.}{Tristan}{It will not *allow* because interaction c1
-  would still need to be developed. But it will definitely facilitate
-  it. I added a note on this after the description of the metrics.}
+difficulty to extend the architecture.
 
@@
   engine. \texttt{S$_2$}=0 when distributed engines are possible in
-  the architecture, \texttt{S$_2$}=1 when they require specific
-  developments in the workflow engine..  \closednote{Rafael}{$S_2$
-    would be 2 if the paradigm does not allow to distribute the
-    orchestration? e.g., workflow executions implemented within MPI
-    jobs, in Pegasus PMC and dispel4Py.}{Tristan}{The problem is that
-    it is very difficult to define ``does not allow to''. For
-    instance, why couldn't you hack Pegasus PMC so much that it
-    becomes possible to distribute workflow executions in several MPI
-    jobs coordinated by an MPI job? Moreover, S2=2 is not used so I'd
-    prefer to leave it like this, unless you have a suggestion to
-    reflect this in the text.}
+  the architecture, and \texttt{S$_2$}=1 when they require specific
+  developments in the workflow engine.\note{Rafael}{$S_2$ would be 2 if the paradigm does not allow to distribute the orchestration?
+e.g., workflow executions implemented within MPI jobs, in Pegasus PMC and dispel4Py.}
 \item Task scheduling (\texttt{S$_3$}): task scheduling is a difficult
@@
 instance of the science gateway, which is in general not possible
-(\texttt{S$_1$}=2)\answerednote{Tristan}{What-about load-balancing between web
-  server instances?}{Marc-e}{Would that really be considered a
-  real architectural feature or an IT hack?}.  Distributed engines are
-not available by default (\texttt{S$_2$}=1). The scheduling of tasks
-on the infrastructure is as complex as in any other architecture since
-the workflow engine might implement any kind of scheduling policy
-(\texttt{S$_3$}=0).
+(\texttt{S$_1$}=2)\note{Tristan}{What-about load-balancing between web
+  server instances?}. Distributed engines are not available by default
+(\texttt{S$_2$}=1). The scheduling of tasks on the infrastructure is
+as complex as in any other architecture since the workflow engine
+might implement any kind of scheduling policy (\texttt{S$_3$}=0).
 
@@
 (\texttt{R$_1$}=0), and only interaction \texttt{e} is
-(\texttt{R$_2$}=1)\answerednote{sorina}{I understand why this is 1 (and why
-  interactions c and b are not counted here), but I wonder whether we
-  can simply ignore these interactions because they are not specific
-  to workflow execution. As far as robustness is concerned, the
-  transfers have to be successful anyway...}{Tristan}{We cannot just
-  ignore these interactions in the architecture because they are
-  required for the system to work. The transfers have to be successful
-  but data movements in the workflow will use the same code as data
-  movements done in the science gateway, which will make this
-  architecture more robust than service for instance. In VIP, for
-  instance, the code used in the GASW wrapper to transfer data (c2) is
-  completely independent from the ``file transfer'' module in the
-  platform (c1), which is detrimental to robustness}.
+(\texttt{R$_2$}=1)\note{sorina}{I understand why this is 1 (and why interactions
+c and b are not counted here), but I wonder whether we can simply ignore these interactions
+because they are not specific to workflow execution. As far as robustness is concerned,
+the transfers have to be successful anyway...}.
 
@@
 
-\paragraph{Integration effort} Setting up a nested workflow
-architecture with service invocation requires a science gateway, 2
-workflow services and 2 infrastructures (\texttt{I$_1$=5}). The
-architecture involves 11 interactions (\texttt{I$_2$=11}):
-\texttt{a$_1$} (appears twice: once for each infrastructure),
-\texttt{a$_2$}, \texttt{b} (twice: once for each infrastructure),
-\texttt{c$_1$} (twice: once for each infrastructure), \texttt{c$_2$}
-(twice: once for each workflow engine and infrastructure) and
-\texttt{d} (twice: once for each workflow engine).
-\answerednote{Rafael}{I like the idea of what has been discussed in
-  the `integration effort', however I think the name may lead to
-  misinterpretations. For example, according to Table 2, it requires
-  more effort to integrate in a Tight architecture than in a
-  service-based architecture. Considering the principles of software
-  engineer, it is simpler to integrate/maintain a system using
-  decoupled services. Maybe it should be renamed to something related
-  to 'dependencies'.}{Tristan}{In our scope I'm not sure that it's
-  simpler to integrate a service architecture than a tightly
-  integrated one. But I take your point. 'Dependencies' is not very
-  suitable either. What about 'development effort'?}
+\paragraph{Integration effort} Setting up a nested workflow architecture with
+service invocation requires a science gateway, 2 workflow services and
+2 infrastructures (\texttt{I$_1$=5}). The architecture involves 11
+interactions (\texttt{I$_2$=11}): \texttt{a$_1$} (appears twice: once for each
+infrastructure), \texttt{a$_2$}, \texttt{b} (twice: once for each
+infrastructure), \texttt{c$_1$} (twice: once for each infrastructure),
+\texttt{c$_2$} (twice: once for each workflow engine and
+infrastructure) and \texttt{d} (twice: once for each workflow engine).
+\note{Rafael}{I like the idea of what has been discussed in the `integration effort', however I think the name may lead to misinterpretations.
+For example, according to Table 2, it requires more effort to integrate in a Tight architecture than in a service-based architecture.
+Considering the principles of software engineer, it is simpler to integrate/maintain a system using decoupled services.
+Maybe it should be renamed to something related to 'dependencies'.}
 
@@
 services. Sub-tasking and workflow import (with service invocation)
-perform a bit better when integrating new engines (\texttt{E$_1$}),
-which is interesting given the current profusion of engines with
-different characteristics.  However, sub-tasking, workflow import
-(with service invocation) and tight integration perform worse than the
-others when adding engine versions (\texttt{E$_2$}) due to the need to
-update infrastructure \note{Marc-e}{confusion with
-  physical?}(sub-tasking), science gateway (tight integration), or
+perform a bit better when integrating new engines (\texttt{E$_1$}), which
+is interesting given the current profusion of engines with different
+characteristics.  However, sub-tasking, workflow import (with service
+invocation) and tight integration perform worse than the others when adding
+engine versions (\texttt{E$_2$}) due to the need to update
+infrastructure (sub-tasking), science gateway (tight integration), or
 workflow converter (workflow import). Workflow import performs worse
-than the others when integrating new workflows (\texttt{E$_3$})
-because of the language conversion step. All architectures except
-nested workflows perform the same when integrating new infrastructures
-(\texttt{E$_4$}).
+than the others when integrating new workflows (\texttt{E$_3$}) because of
+the language conversion step. All architectures except nested
+workflows perform the same when integrating new infrastructures (\texttt{E$_4$}).
 
@@
 hardly cope with workflow engines being wrapped in computing tasks as
-done in sub-tasking \note{Marc-e}{free for all? WE dev involved?}. The availability of multiple engine instances
+done in sub-tasking. The availability of multiple engine instances
 could also become a critical feature for science gateways with
@@
 invocation) are the only architectures that support meta-workflows,
-which may be interesting in some cases. Also, sub-tasking and nested
-workflows may complexify fine-grained debugging. \closednote{Rafael}{I would
-  suggest to provide some more examples when this could be
-  interesting.}{Tristan}{Done.}  The availability of fine-grained debugging
-information may also be critical to efficient user support.
+which may be interesting in some cases.\note{Rafael}{I would suggest to provide some
+more examples when this could be interesting.}
+The availability of fine-grained debugging information may also be critical to efficient
+user support.
 
@@
 architectures, based on the metrics in Table~\ref{table:evaluation}.
-Note that this analysis is only for illustration purposes since it
-assumes that all criteria have equal weights while real systems would
-favor some of them. Overall, sub-tasking, and workflow import (with
-service invocation) perform a bit better than tight integration,
-service invocation and pool. Nested workflows stands a bit behind for
-the reasons mentioned previously.\closednote{Rafael}{I would emphasize
-  that this analysis is only for illustration purposes, since it does
-  not account for different weights in metrics.}{Tristan}{Done.}
+Overall, sub-tasking, and workflow import (with service invocation)
+perform a bit better than tight integration, service invocation and
+pool. Nested workflows stands a bit behind for the reasons mentioned
+previously.\note{Rafael}{I would emphasize that this analysis is only for illustration purposes, since it does not account for different weights in metrics.}
 \begin{table*}
@@
 also realize that this comes at the cost of realism.  In particular,
-the presented architectures are theoritical patterns that may be blended in
+the presented architectures are pure patterns that may be blended in
 actual systems. The distinction between tight integration and service
-invocation, for instance, may not be always that clear in practice. Service
-invocation may also be combined with sub-tasking in some cases. However, the
-criteria discussed in this work would still apply to broadly compare and categorize
-such cases of hydrid architectures.
+invocation, for instance, may not be that clear in practice. Service
+invocation may also be combined with sub-tasking in some cases.
 
@@
 language. For instance, converting FSL, PSOM or Nipype pipelines to
-any other workflow language is problematic because these engines rely
+any other workflow language is very tricky because these engines rely
 on general-purpose programming languages such as Bash, Octave and
@@
 tested task and data control features (interactions \texttt{b} and
-\texttt{c}), which would favor sub-tasking. Similarly, adding a new
-type of engine may facilitate integration of a new infrastructure when
-interactions \texttt{b} and \texttt{c$_2$} are already available. The
-migration cost between architectures has been ignored as well.
+\texttt{c}), which would favor sub-tasking. The migration cost between
+architectures is totally ignored as well.
 
@@
 metrics for all the architectures, and we discussed the pros and cons
-of all the architectures based on these metrics.  \answerednote{Rafael}{One
-  thing that is missing in the conclusion or the limitations section
-  is to state that this work does not intend to quantify the amount of
-  work required to develop science gateways using a specific
-  architecture, but the goal is to highlight what are the main
-  components and interactions for each architecture, and the main
-  advantages and drawbacks. This should also be stated in the
-  introduction.}{Tristan}{This is in contradiction with the metric on
-  'integration effort'. What do you think?}
+of all the architectures based on these metrics.
+\note{Rafael}{One thing that is missing in the conclusion or the limitations section is to state that this work does not intend to quantify the amount of work required to develop science gateways using a specific architecture, but the goal is to highlight what are the main components and interactions for each architecture, and the main advantages and drawbacks.
+This should also be stated in the introduction.}
 
@@
 Ludmer Family Foundation and the Ludmer Centre for Neuroinformatics
-and Mental Health. \note{Marc-e}{Would it be proper to thank EGI? Compute Canada? CANARIE?}
+and Mental Health.
 

### 2016-04-26 paper.tex
Fixed note on multiple engines in tight integrations.

--- 2016-04-26 paper.tex
@@
 instance of the science gateway, which is in general not possible
-(\texttt{S$_1$}=2)\answerednote{Tristan}{What-about load-balancing between web
-  server instances?}{Marc-e}{Would that really be considered a
-  real architectural feature or an IT hack?}.  Distributed engines are
-not available by default (\texttt{S$_2$}=1). The scheduling of tasks
-on the infrastructure is as complex as in any other architecture since
-the workflow engine might implement any kind of scheduling policy
-(\texttt{S$_3$}=0).
+(\texttt{S$_1$}=2). Specific IT setups such as load-balancing between
+web server instances might be used to deal with this
+issue. \closedanswerednote{Tristan}{What-about load-balancing between web
+  server instances?}{Marc-e}{Would that really be considered a real
+  architectural feature or an IT hack?}{Tristan}{I agree, we should
+  keep IT setups out of the picture. Depending on the internals of the
+  science gateway, such hacks may or may not be feasible. I added a
+  sentence to mention this possibility, without changing the score.}
+Distributed engines are not available by default
+(\texttt{S$_2$}=1). The scheduling of tasks on the infrastructure is
+as complex as in any other architecture since the workflow engine
+might implement any kind of scheduling policy (\texttt{S$_3$}=0).
 

### 2016-04-26 paper.tex
Merge remote-tracking branch 'glatard/master'
# Conflicts:
#	paper.tex fixed
Closed note based on discussion in #7.

@@
 (\texttt{R$_1$}=0), and only interaction \texttt{e} is
-(\texttt{R$_2$}=1)\answerednote{sorina}{I understand why this is 1 (and why
-  interactions c and b are not counted here), but I wonder whether we
-  can simply ignore these interactions because they are not specific
-  to workflow execution. As far as robustness is concerned, the
-  transfers have to be successful anyway...}{Tristan}{We cannot just
-  ignore these interactions in the architecture because they are
+(\texttt{R$_2$}=1)\closedanswerednote{sorina}{I understand why this is 1
+  (and why interactions c and b are not counted here), but I wonder
+  whether we can simply ignore these interactions because they are not
+  specific to workflow execution. As far as robustness is concerned,
+  the transfers have to be successful anyway...}{Tristan}{We cannot
+  just ignore these interactions in the architecture because they are
   required for the system to work. The transfers have to be successful
@@
   completely independent from the ``file transfer'' module in the
-  platform (c1), which is detrimental to robustness}.
+  platform (c1), which is detrimental to robustness}{Tristan}{Closed,
+  see github issue \#7.}.
 

### 2016-04-26 paper.tex
Minor updates

@@
   integrating a workflow is not the same process as integrating a
-  workflow \emph{engine}. \note{Marc-e}{automation?}
+  workflow \emph{engine}. 
 \item Task control: operations to manage tasks on the infrastructure,
@@
 others when adding engine versions (\texttt{E$_2$}) due to the need to
-update infrastructure \note{Marc-e}{confusion with
-  physical?}(sub-tasking), science gateway (tight integration), or
-workflow converter (workflow import). Workflow import performs worse
-than the others when integrating new workflows (\texttt{E$_3$})
-because of the language conversion step. All architectures except
-nested workflows perform the same when integrating new infrastructures
-(\texttt{E$_4$}).
+update infrastructure (sub-tasking), science gateway (tight
+integration), or workflow converter (workflow import). Workflow import
+performs worse than the others when integrating new workflows
+(\texttt{E$_3$}) because of the language conversion step. All
+architectures except nested workflows perform the same when
+integrating new infrastructures (\texttt{E$_4$}).
 
@@
 
+% Literature on SG architecture Very few works go beyond the
+% description of architectures of particular systems. The work in
+% http://link.springer.com/chapter/10.1007/11596141_3 discusses
+% service-oriented architectures for science gateways, with a workflow
+% engine. The work in
+% http://docs.lib.purdue.edu/cgi/viewcontent.cgi?article=1414&context=nanopub
+% highlights 4 TeraGrid science gateways (GridChem, LEAD, nanoHUB,
+% CABIG). Architectures are described but there is no real focus on
+% workflow engines. GridChem doesn't include any workflow engine,
+% which btw make it relevant for integration via sub-tasking. LEAD has
+% a workflow engine.
+
 \section{Acknowledgments}

### 2016-04-26 paper.tex
Acknowledgements fixed CBRAIN

--- 2016-04-26 paper.tex
@@
   integrating a workflow is not the same process as integrating a
-  workflow \emph{engine}. \note{Marc-e}{automation?}
+  workflow \emph{engine}.
 \item Task control: operations to manage tasks on the infrastructure,
@@
 instance of the science gateway, which is in general not possible
-(\texttt{S$_1$}=2). Specific IT setups such as load-balancing between
-web server instances might be used to deal with this
-issue. \closedanswerednote{Tristan}{What-about load-balancing between web
-  server instances?}{Marc-e}{Would that really be considered a real
-  architectural feature or an IT hack?}{Tristan}{I agree, we should
-  keep IT setups out of the picture. Depending on the internals of the
-  science gateway, such hacks may or may not be feasible. I added a
-  sentence to mention this possibility, without changing the score.}
-Distributed engines are not available by default
-(\texttt{S$_2$}=1). The scheduling of tasks on the infrastructure is
-as complex as in any other architecture since the workflow engine
-might implement any kind of scheduling policy (\texttt{S$_3$}=0).
+(\texttt{S$_1$}=2)\answerednote{Tristan}{What-about load-balancing between web
+  server instances?}{Marc-e}{Would that really be considered a
+  real architectural feature or an IT hack?}.  Distributed engines are
+not available by default (\texttt{S$_2$}=1). The scheduling of tasks
+on the infrastructure is as complex as in any other architecture since
+the workflow engine might implement any kind of scheduling policy
+(\texttt{S$_3$}=0).
 
@@
 administrators of the European Grid Initiative and the GGUS support
-for their help related to the VIP platform. We are grateful for the
+for their help related to the VIP platform. The CBRAIN team is grateful for the
 computing cycles, storage, and support obtained from Compute Canada
-(\url{https://computecanada.ca}) for CBRAIN. This work has been made
-possible with the support of the Irving Ludmer Family Foundation and
-the Ludmer Centre for Neuroinformatics and Mental
-Health. \answerednote{Marc-e}{Would it be proper to thank EGI?
+(\url{https://computecanada.ca}) and platform development program from CANARIE
+ (\url{http://www.canarie.ca/}). This work has been made possible with the support of the Irving Ludmer Family Foundation and
+the Ludmer Centre for Neuroinformatics and Mental Health. \answerednote{Marc-e}{Would it be proper to thank EGI?
   Compute Canada? CANARIE?}{Tristan}{Added EGI and Compute Canada. Not sure if
-  Canarie should be added. What about Calcul QC?}
+  Canarie should be added. What about Calcul QC?}{Marc-e}{CQ, no. CANARIE, I'd
+  say yes since they funded a major chunk of CBRAIN's development.}
 

### 2016-04-27 paper.tex
Following @merous' comment, added 1 to all E4 values (to add a1).

@@
 \author[mcgill]{Natacha Beck}
+\author[mcgill]{Samir Das}
 \author[isi]{Rafael Ferreira da Silva}
@@
 \author[mcgill]{Alan C. Evans}
-
-\note{Marc-e}{Add Samir Das?}
 
@@
                                      & \cellcolor[HTML]{999999}3\\
-New infrastructure -- \texttt{E$_4$} & \cellcolor[HTML]{99DD99}3
-                                     & \cellcolor[HTML]{99DD99}3
-                                     & \cellcolor[HTML]{99FF99}2
-                                     & \cellcolor[HTML]{99DD99}3
-                                     & \cellcolor[HTML]{999999}5
-                                     & \cellcolor[HTML]{99DD99}3\\
-  \textbf{Total} (difficulty to extend) & \cellcolor[HTML]{99D199}\textbf{9}
-                                     & \cellcolor[HTML]{99D199}\textbf{9}
-                                     & \cellcolor[HTML]{99FF99}\textbf{7}
-                                     & \cellcolor[HTML]{99E899}\textbf{8}
-                                     & \cellcolor[HTML]{999999}\textbf{11.5}
-                                     & \cellcolor[HTML]{99E899}\textbf{8}\\
+New infrastructure -- \texttt{E$_4$} & \cellcolor[HTML]{99DD99}4
+                                     & \cellcolor[HTML]{99DD99}4
+                                     & \cellcolor[HTML]{99FF99}3
+                                     & \cellcolor[HTML]{99DD99}4
+                                     & \cellcolor[HTML]{999999}6
+                                     & \cellcolor[HTML]{99DD99}4\\
+  \textbf{Total} (difficulty to extend) & \cellcolor[HTML]{99D199}\textbf{10}
+                                     & \cellcolor[HTML]{99D199}\textbf{10}
+                                     & \cellcolor[HTML]{99FF99}\textbf{8}
+                                     & \cellcolor[HTML]{99E899}\textbf{9}
+                                     & \cellcolor[HTML]{999999}\textbf{12.5}
+                                     & \cellcolor[HTML]{99E899}\textbf{9}\\
 \cellcolor[HTML]{EEEEEE}\textbf{Scalability}& \multicolumn{6}{l}{\cellcolor[HTML]{EEEEEE}}\\
@@
 (\texttt{E$_2$}=1).  Inserting a new workflow is done through
-interactions \texttt{a$_1$} and \texttt{a$_2$} (\texttt{E$_3$}=2). Adding a new infrastructure generates updates in
-interactions \texttt{b}, \texttt{c$_1$} and \texttt{c$_2$}
-(\texttt{E$_4$}=3).
+interactions \texttt{a$_1$} and \texttt{a$_2$}
+(\texttt{E$_3$}=2). Adding a new infrastructure generates updates in
+interactions \texttt{a$_1$}, \texttt{b}, \texttt{c$_1$} and
+\texttt{c$_2$} (\texttt{E$_4$}=4).
 
@@
 (\texttt{E$_3$}=2). Adding a new type of infrastructure requires
-updates in interactions \texttt{b}, \texttt{$c_1$} and \texttt{$c_2$}
-(\texttt{E$_4$}=3).
+updates in interactions \texttt{a$_1$}, \texttt{b}, \texttt{$c_1$} and \texttt{$c_2$}
+(\texttt{E$_4$}=4).
 
@@
 through interactions \texttt{a$_1$} and \texttt{a$_2$}  (\texttt{E$_3$}=2). Adding a new
-infrastructure requires to update interactions \texttt{b} and
-\texttt{c} in the science gateway (\texttt{E$_4$}=2).
+infrastructure requires to update interactions \texttt{a$_1$}, \texttt{b} and
+\texttt{c} in the science gateway (\texttt{E$_4$}=3).
 
@@
 (\texttt{E$_3$=2}). Integrating a new infrastructure requires updates
-in interactions \texttt{b}, \texttt{c$_1$} and \texttt{c$_2$}
-(\texttt{E$_4$=3}).
+in interactions \texttt{a$_1$}, \texttt{b}, \texttt{c$_1$} and \texttt{c$_2$}
+(\texttt{E$_4$=4}).
 
@@
 (\texttt{E$_3$=2}). Adding a new infrastructure requires to
-re-implement interactions \texttt{b} and \texttt{c$_2$} twice, and
-interaction \texttt{c$_1$} once (\texttt{E$_4$=5}).
+re-implement interactions \texttt{a$_1$}, \texttt{b} and \texttt{c$_2$} twice, and
+interaction \texttt{c$_1$} once (\texttt{E$_4$=6}).
 
@@
 service architecture, interfacing with a new infrastructure requires
-modifications in the workflow service and in interactions \texttt{b}
-and \texttt{c$_2$} (\texttt{E$_4$=3}).
+modifications in the workflow service and in interactions \texttt{a$_1$}, \texttt{b}
+and \texttt{c$_2$} (\texttt{E$_4$=4}).
 

### 2016-04-27 paper.tex
Precised description of E2 following @merous' comment.

@@
   same software. When this is not the case, the different versions are
-  considered as different engine types.
+  considered as different engine types. Components are not treated
+  equally regarding engine version updates. Components whose only
+  function is to host the workflow engine, i.e. workflow services and
+  agents, are not counted in E$_2$ because updates in such components
+  are assumed to be straightforward. On the contrary, modifications of
+  components that have other functions than wrapping the engine,
+  i.e. science gateway and infrastructure, are counted in E$_2$
+  because updates in these components require more effort, e.g. new
+  release of the science gateway, gaining administrative privileges on
+  the infrastructure, etc.
 \item New workflow (\texttt{E$_3$}): adding a new workflow is a very
@@
 updating the workflow service without modifying any interaction or
-component (\texttt{E$_2$}=0). We consider that updating the engine
-version in a workflow service does not require code updates in the
-service since the only goal of this component is to wrap the engine.
-New workflows are added in the science gateway or in the workflow
-engine through interactions \texttt{a$_1$} and \texttt{a$_2$}
-(\texttt{E$_3$}=2). Adding a new type of infrastructure requires
-updates in interactions \texttt{a$_1$}, \texttt{b}, \texttt{$c_1$} and \texttt{$c_2$}
-(\texttt{E$_4$}=4).
+component (\texttt{E$_2$}=0). Updating the engine version in a
+workflow service does not count since the only goal of this component
+is to wrap the engine.  New workflows are added in the science gateway
+or in the workflow engine through interactions \texttt{a$_1$} and
+\texttt{a$_2$} (\texttt{E$_3$}=2). Adding a new type of infrastructure
+requires updates in interactions \texttt{a$_1$}, \texttt{b},
+\texttt{$c_1$} and \texttt{$c_2$} (\texttt{E$_4$}=4).
 

### 2016-04-27 paper.tex
Added a sentence on science gateways that do not include any workflow engine (following @merous' comment)

@@
 
-Science gateways are used to share resources in a community and to
-provide increased performance and capacity through facilitated access to storage and
-computing power. They are often accessible through a web interface
-that helps users manage access rights, data transfers, task execution
-and authentication on multiple computing and storage
+Science gateways are used to share resources within a community and to
+provide increased performance and capacity through facilitated access
+to storage and computing power. They are often accessible through a
+web interface that helps users manage access rights, data transfers,
+task execution and authentication on multiple computing and storage
 locations. Workflow engines are part of this ecosystem as core
@@
 transfers and authentication. Other gateways may be more general and
-host applications executed by different types of
-engines. Extensibility is an important property of the integration, as
-new workflows are added frequently, different types and versions of
+host applications executed by different types of engines. Finally,
+some science gateways may not use any workflow engine and focus on the
+execution of simple tasks -- this class of systems is not covered by
+our analysis.
+
+Extensibility is an important property of the integration, as new
+workflows are added frequently, different types and versions of
 workflow engines may be integrated over time, and different kinds of
@@
 
-Scalability is also an important concern, since science gateways are
+Scalability is also a crucial concern, since science gateways are
 multi-user systems that manage substantial amounts of data and execute

### 2016-04-27 paper.tex
Added a paragraph in discussion (@merous comment).

@@
 
-Finally, the particular technical or historical context of a science
-gateway project may obviously influence the choice of an architecture
-to integrate workflow engines. For instance, some workflow engines may
-be already available as web services, which would tend to favor
-service invocation, and other science gateways may have strongly
-tested task and data control features (interactions \texttt{b} and
-\texttt{c}), which would favor sub-tasking. Similarly, adding a new
-type of engine may facilitate integration of a new infrastructure when
-interactions \texttt{b} and \texttt{c$_2$} are already available. The
-migration cost between architectures has been ignored as well.
+The particular technical or historical context of a science gateway
+project may obviously influence the choice of an architecture to
+integrate workflow engines. For instance, some workflow engines may be
+already available as web services, which would tend to favor service
+invocation, and other science gateways may have strongly tested task
+and data control features (interactions \texttt{b} and \texttt{c}),
+which would favor sub-tasking. Similarly, adding a new type of engine
+may facilitate integration of a new infrastructure when interactions
+\texttt{b} and \texttt{c$_2$} are already available. The migration
+cost between architectures has been ignored as well.
+
+Finally, workflow engines are only one of the many aspects involved in
+the design of a science gateway. Other properties definitely influence
+the design of a software architecture, for instance collaborative
+features, data visualization, search, authentication, accounting and
+so on.
 

### 2016-04-27 paper.tex
Fixed note in conclusion

@@
 suitable for comparison, we illustrated them on real systems, we
-developed quantitative metrics for evaluation, we evaluated these
-metrics for all the architectures, and we discussed the pros and cons
-of all the architectures based on these metrics.
+developed quantitative metrics for evaluation, and we discussed the
+pros and cons of all the presented architectures based on these metrics.
 \closedanswerednote{Rafael}{One thing that is missing in the
@@
 % Impact of the work
-To the best of our knowledge, our work is the first of its kind
-\note{Tristan}{Be more accurate}. Our results can be used as
-recommendation for science gateway and workflow engine architects to
-build new platforms, or to upgrade existing ones.
+To the best of our knowledge, our work is the first of its kind. So
+far, the literature on science gateways and workflow engines has
+focused on the description of particular systems, or on the
+presentation of a particular architecture. For instance, the work
+in~\cite{Gannon2005} discusses service-oriented architectures for
+science gateways with workflow engines.  Instead, our analysis
+abstracts and evaluates architectural patterns independently from any
+particular system. It provides general insight to integrate science
+gateways and workflow engines.
 
@@
  (\url{http://www.canarie.ca/}). This work has been made possible with the support of the Irving Ludmer Family Foundation and
-the Ludmer Centre for Neuroinformatics and Mental Health. \answerednote{Marc-e}{Would it be proper to thank EGI?
+the Ludmer Centre for Neuroinformatics and Mental Health. \closedanswerednote{Marc-e}{Would it be proper to thank EGI?
   Compute Canada? CANARIE?}{Tristan}{Added EGI and Compute Canada. Not sure if

### 2016-04-27 paper.tex
Various small edits
Intro parts

--- 2016-04-27 paper.tex
@@
 \author[mcgill]{Natacha Beck}
-\author[mcgill]{Samir Das}
 \author[isi]{Rafael Ferreira da Silva}
@@
 variety of programming languages. Such tools were interfaced to
-computing systems, in particular clusters, to handle high-throughput
+computing systems, in particular cluster schedulers, to handle high-throughput
 task executions. For instance, FSL can launch tasks on SGE through its
@@
 that may consist of files, database entries or simple parameter
-values. Although simplistic, this definition covers both e-Science and
-domain-specific engines. Some workflow engines, usually e-Science
+values. Although simplistic, this definition covers both e-Science workflow systems and
+domain-specific pipelines.\note{Marc-e}{To remain consistent with above descriptions} Some workflow engines, usually e-Science
 ones, may transfer data across the infrastructure, and others, usually
@@
 
-Science gateways are used to share resources within a community and to
-provide increased performance and capacity through facilitated access
-to storage and computing power. They are often accessible through a
-web interface that helps users manage access rights, data transfers,
-task execution and authentication on multiple computing and storage
+Science gateways are used to share resources in a community and to
+provide increased performance and capacity through facilitated access to storage and
+computing power. They are often accessible through a web interface
+that helps users manage access rights, data transfers, task execution
+and authentication on multiple computing and storage
 locations. Workflow engines are part of this ecosystem as core
@@
 transfers and authentication. Other gateways may be more general and
-host applications executed by different types of engines. Finally,
-some science gateways may not use any workflow engine and focus on the
-execution of simple tasks -- this class of systems is not covered by
-our analysis.
-
-Extensibility is an important property of the integration, as new
-workflows are added frequently, different types and versions of
+host applications executed by different types of
+engines. Extensibility is an important property of the integration, as
+new workflows are added frequently, different types and versions of
 workflow engines may be integrated over time, and different kinds of
@@
 
-Scalability is also a crucial concern, since science gateways are
+Scalability is also an important concern, since science gateways are
 multi-user systems that manage substantial amounts of data and execute
@@
 frameworks~\cite{kacsuk2014science,ardizzone2012decide}. Other gateway
-instances were also built independently from any framework, using
+instances were also built independently from such specialized framework, using
 regular web development tools and libraries to access distributed
@@
 at a wide range of computing sites, including a network of national HPC
-centers in Canada, one in Korea, one in Germany, and on numerous local
-servers. As February 2016, CBRAIN has served over 420 users across 50
-cities in 22 countries and has a national computing allocation of 660 CPU years
-per annum.
+centers in Canada, a cluster in Germany, and on numerous local research
+servers. CBRAIN is a task-centric platform where various workflow engines
+can be deployed on the infrastructure as any other task would. As February 2016,
+CBRAIN has served over 420 users across 50 cities in 22 countries and has a
+national computing allocation of 660 CPU years per annum.
 
@@
 global task meta-scheduler, the presence of a file catalog, etc. In
-the analysis presented below, such specific functions are not
+the analysis presented below, such specific requirements are not
 discussed. Instead, we consider the infrastructure as an abstract
@@
                                      & \cellcolor[HTML]{999999}3\\
-New infrastructure -- \texttt{E$_4$} & \cellcolor[HTML]{99DD99}4
-                                     & \cellcolor[HTML]{99DD99}4
-                                     & \cellcolor[HTML]{99FF99}3
-                                     & \cellcolor[HTML]{99DD99}4
-                                     & \cellcolor[HTML]{999999}6
-                                     & \cellcolor[HTML]{99DD99}4\\
-  \textbf{Total} (difficulty to extend) & \cellcolor[HTML]{99D199}\textbf{10}
-                                     & \cellcolor[HTML]{99D199}\textbf{10}
-                                     & \cellcolor[HTML]{99FF99}\textbf{8}
-                                     & \cellcolor[HTML]{99E899}\textbf{9}
-                                     & \cellcolor[HTML]{999999}\textbf{12.5}
-                                     & \cellcolor[HTML]{99E899}\textbf{9}\\
+New infrastructure -- \texttt{E$_4$} & \cellcolor[HTML]{99DD99}3
+                                     & \cellcolor[HTML]{99DD99}3
+                                     & \cellcolor[HTML]{99FF99}2
+                                     & \cellcolor[HTML]{99DD99}3
+                                     & \cellcolor[HTML]{999999}5
+                                     & \cellcolor[HTML]{99DD99}3\\
+  \textbf{Total} (difficulty to extend) & \cellcolor[HTML]{99D199}\textbf{9}
+                                     & \cellcolor[HTML]{99D199}\textbf{9}
+                                     & \cellcolor[HTML]{99FF99}\textbf{7}
+                                     & \cellcolor[HTML]{99E899}\textbf{8}
+                                     & \cellcolor[HTML]{999999}\textbf{11.5}
+                                     & \cellcolor[HTML]{99E899}\textbf{8}\\
 \cellcolor[HTML]{EEEEEE}\textbf{Scalability}& \multicolumn{6}{l}{\cellcolor[HTML]{EEEEEE}}\\
@@
   same software. When this is not the case, the different versions are
-  considered as different engine types. Components are not treated
-  equally regarding engine version updates. Components whose only
-  function is to host the workflow engine, i.e. workflow services and
-  agents, are not counted in E$_2$ because updates in such components
-  are assumed to be straightforward. On the contrary, modifications of
-  components that have other functions than wrapping the engine,
-  i.e. science gateway and infrastructure, are counted in E$_2$
-  because updates in these components require more effort, e.g. new
-  release of the science gateway, gaining administrative privileges on
-  the infrastructure, etc.
+  considered as different engine types.
 \item New workflow (\texttt{E$_3$}): adding a new workflow is a very
@@
 (\texttt{E$_2$}=1).  Inserting a new workflow is done through
-interactions \texttt{a$_1$} and \texttt{a$_2$}
-(\texttt{E$_3$}=2). Adding a new infrastructure generates updates in
-interactions \texttt{a$_1$}, \texttt{b}, \texttt{c$_1$} and
-\texttt{c$_2$} (\texttt{E$_4$}=4).
+interactions \texttt{a$_1$} and \texttt{a$_2$} (\texttt{E$_3$}=2). Adding a new infrastructure generates updates in
+interactions \texttt{b}, \texttt{c$_1$} and \texttt{c$_2$}
+(\texttt{E$_4$}=3).
 
@@
 updating the workflow service without modifying any interaction or
-component (\texttt{E$_2$}=0). Updating the engine version in a
-workflow service does not count since the only goal of this component
-is to wrap the engine.  New workflows are added in the science gateway
-or in the workflow engine through interactions \texttt{a$_1$} and
-\texttt{a$_2$} (\texttt{E$_3$}=2). Adding a new type of infrastructure
-requires updates in interactions \texttt{a$_1$}, \texttt{b},
-\texttt{$c_1$} and \texttt{$c_2$} (\texttt{E$_4$}=4).
+component (\texttt{E$_2$}=0). We consider that updating the engine
+version in a workflow service does not require code updates in the
+service since the only goal of this component is to wrap the engine.
+New workflows are added in the science gateway or in the workflow
+engine through interactions \texttt{a$_1$} and \texttt{a$_2$}
+(\texttt{E$_3$}=2). Adding a new type of infrastructure requires
+updates in interactions \texttt{b}, \texttt{$c_1$} and \texttt{$c_2$}
+(\texttt{E$_4$}=3).
 
@@
 through interactions \texttt{a$_1$} and \texttt{a$_2$}  (\texttt{E$_3$}=2). Adding a new
-infrastructure requires to update interactions \texttt{a$_1$}, \texttt{b} and
-\texttt{c} in the science gateway (\texttt{E$_4$}=3).
+infrastructure requires to update interactions \texttt{b} and
+\texttt{c} in the science gateway (\texttt{E$_4$}=2).
 
@@
 (\texttt{E$_3$=2}). Integrating a new infrastructure requires updates
-in interactions \texttt{a$_1$}, \texttt{b}, \texttt{c$_1$} and \texttt{c$_2$}
-(\texttt{E$_4$=4}).
+in interactions \texttt{b}, \texttt{c$_1$} and \texttt{c$_2$}
+(\texttt{E$_4$=3}).
 
@@
 (\texttt{E$_3$=2}). Adding a new infrastructure requires to
-re-implement interactions \texttt{a$_1$}, \texttt{b} and \texttt{c$_2$} twice, and
-interaction \texttt{c$_1$} once (\texttt{E$_4$=6}).
+re-implement interactions \texttt{b} and \texttt{c$_2$} twice, and
+interaction \texttt{c$_1$} once (\texttt{E$_4$=5}).
 
@@
 service architecture, interfacing with a new infrastructure requires
-modifications in the workflow service and in interactions \texttt{a$_1$}, \texttt{b}
-and \texttt{c$_2$} (\texttt{E$_4$=4}).
+modifications in the workflow service and in interactions \texttt{b}
+and \texttt{c$_2$} (\texttt{E$_4$=3}).
 

### 2016-04-27 paper.tex
A few updates

--- 2016-04-27 paper.tex
@@
 \author[mcgill]{Natacha Beck}
-\author[mcgill]{Samir Das}
+\author[mcgill]{\\Samir Das}
 \author[isi]{Rafael Ferreira da Silva}
@@
 \address[creatis]{University of Lyon, CNRS, INSERM, CREATIS, Villeurbanne, France.}
-\address[criugm]{Centre de Recherche de l'Institut de G\'eriatrie de Montr\'eal CRIUGM, Montreal, QC, Canada.}
+\address[criugm]{Centre de Recherche de l'Institut de G\'eriatrie de Montr\'eal CRIUGM, Montr\'eal, QC, Canada.}
 \address[isi]{University of Southern California, Information Sciences Institute, Marina del Rey, CA, USA.}
@@
   We investigate software architectures commonly used to integrate
-  workflow engines in science gateways. Based on our experience with
-  the development and operations of the CBRAIN and VIP science
-  gateways, and our past involvement in the SHIWA project, we abstract
-  software components and interactions in 6 architectures that we use
-  to classify 18 systems. We compare these architectures using metrics
-  measuring global complexity, robustness, extensibility, scalability
-  and other specific features. Results show the pros and cons of the
-  different architectures: tight integration and sub-tasking are the
-  simplest architectures and should therefore be the easiest to
-  implement and the most robust ones, most architectures perform the
-  same in terms of extensibility, the pool model is the most scalable
-  and meta-workflows are only available in nested workflows and
-  workflow import. These results provide insights for software
-  architects on how to integrate science gateways and workflow
-  engines.
+  workflow engines in science gateways, based on our experience with
+  the CBRAIN and VIP platforms, and on our past involvement in the
+  SHIWA project. We abstract software components and interactions in 6
+  architectures that we use to classify 18 systems. We compare these
+  architectures using metrics measuring global complexity, robustness,
+  extensibility, scalability and other specific features. Results show
+  the pros and cons of the different architectures: tight integration
+  and sub-tasking are the simplest architectures and should therefore
+  be the easiest to implement and the most robust ones, most
+  architectures perform the same in terms of extensibility, the pool
+  model is the most scalable and meta-workflows are only available in
+  nested workflows and workflow import. Our analysis provides insights
+  for software architects on how to integrate science gateways and
+  workflow engines.
 \end{abstract}
@@
 \begin{keyword}
-Workflow engines \sep science gateways \sep software architecture.
+Workflow engines \sep science gateways \sep software architectures.
 \end{keyword}
@@
 
-Workflow engines are critical components of the ecosystem of tools and
-services offered by science gateways to leverage distributed
-infrastructures for the efficient execution of applications through
-the web. Several software architectures can be adopted to integrate
-workflow engines in science gateways, with important consequences on
-the required development effort and on the resulting system. This
-paper describes, exemplifies and compares such architectures, based on
+Workflow engines are critical components in the ecosystem of tools and
+services offered by science gateways for the efficient and transparent
+execution of applications on distributed infrastructures. Several
+software architectures can be adopted to integrate workflow engines in
+science gateways, with important consequences on the required
+development effort and on the resulting system. This paper describes,
+exemplifies and compares such architectures, based on
 system-independent representations of their main components and
@@
 In the last decade, the e-Science community has developed workflow
-systems to help developers access distributed infrastructures such as
-clusters, grids, clouds and web services. These efforts resulted in
-tools such as Askalon~\cite{fahringer2005askalon},
+systems to help application developers access distributed
+infrastructures such as clusters, grids, clouds and web
+services. These efforts resulted in tools such as
+Askalon~\cite{fahringer2005askalon},
 Hyperflow~\cite{balis2016hyperflow}, MOTEUR~\cite{GLAT-08i},
@@
 WS-PGRADE~\cite{Kacsuk2012} and others. Such workflow engines usually
-describe applications in a high-level language that offers specific
-data and control flow constructs, parallelization operators, visual
-edition features, links with domain-specific tool repositories,
-provenance recording, among other features. An overview of workflow
-systems features and capability is available
-in~\cite{deelman2009workflows}.
-% Descriptions and experiments
-%conducted with e-Science workflow engines were published in journals
-%such as Future Generation Computer Systems and conference venues such
-%as WORKS.
+describe applications in a high-level language with specific data and
+control flow constructs, parallelization operators, visual edition
+features, links with domain-specific tool repositories, provenance
+recording and other features. An overview of workflow system
+capabilities is available in~\cite{deelman2009workflows}.
 
@@
 variety of programming languages. Such tools were interfaced to
-computing systems, in particular clusters, to handle high-throughput
-task executions. For instance, FSL can launch tasks on SGE through its
-fsl\_sub tool, Nipype has execution plugins for S/OGE, Torque,
+computing infrastructures, in particular clusters, to execute tasks at
+a high throughput. For instance, FSL can launch tasks on SGE through
+its fsl\_sub tool, Nipype has execution plugins for S/OGE, Torque,
 IPython, ssh and multi-processor servers, and PSOM is interfaced to
@@
 pipelines and are now shared across neuroimaging groups
-world-wide. These
-domain-specific engines represent a tremendous opportunity for science
-gateways to leverage existing tools and applications. They nicely
-complement e-Science engines that are more oriented towards the
-exploitation of distributed computing infrastructures, in particular
-grids and clouds.
+world-wide. These domain-specific engines represent a tremendous
+opportunity for science gateways to leverage existing tools and
+applications. They nicely complement e-Science engines that are more
+oriented towards the exploitation of distributed computing
+infrastructures, in particular grids and clouds.
 
@@
 web interface that helps users manage access rights, data transfers,
-task execution and authentication on multiple computing and storage
+task execution, and authentication on multiple computing and storage
 locations. Workflow engines are part of this ecosystem as core
@@
 achieved through a common intermediate
-representation~\cite{plankensteiner-prodan-etal:2013}. \note{Tristan}{Silvia,
-  do you think it is legitimate to refer to SHIWA like this in the
-  paper, or should we involve Gabor in the writing?}
+representation~\cite{plankensteiner-prodan-etal:2013}.
 
@@
 The interactions involved in the architectures are described below and
-labeled consistently with the notations used on
-Figure~\ref{fig:architectures}.
+labeled consistently with Figure~\ref{fig:architectures}.
 
@@
   use it. The infrastructure might offer various data storage backends
-  with heterogeneous interfaces. \closednote{Marc-Etienne}{this -c2- can be also trigerred
+  with heterogeneous interfaces. \closedanswerednote{Marc-Etienne}{this -c2- can be also trigerred
     by the user}{Tristan}{No, it can't: this is by definition an interaction between the workflow engine and the infrastructure.
@@
   (FSL tasks or PSOM agents). 7-8: Execution server starts sub-tasks
-  through task scheduler. 8': (PSOM only) PSOM master and workers
-  execute workflow. 9. Execution server transfers results to storage
-  server(s). Interaction \texttt{b} in Figure~\ref{fig:architectures}
-  is implemented by steps \texttt{4}, \texttt{5}, \texttt{7} and
-  \texttt{8} (regular interactions with batch manager). Interaction
-  \texttt{c} is implemented by steps \texttt{3} and
-  \texttt{9}. Interaction \texttt{e} is implemented by step \texttt{6}
-  (for FSL: through a modified version of the fsl\_sub script
-  available in \url{https://github.com/aces/cbrain-plugins-neuro}; for
-  PSOM: through a specific development in PSOM).}
+  through task scheduler. FSL sub-tasks will run locally
+  instead of being submitted to CBRAIN through interaction 6. 8':
+  (PSOM only) PSOM master and workers execute workflow. 9. Execution
+  server transfers results to storage server(s). Interaction
+  \texttt{b} in Figure~\ref{fig:architectures} is implemented by steps
+  \texttt{4}, \texttt{5}, \texttt{7} and \texttt{8} (regular
+  interactions with batch manager). Interaction \texttt{c} is
+  implemented by steps \texttt{3} and \texttt{9}. Interaction
+  \texttt{e} is implemented by step \texttt{6} (for FSL: through a
+  modified version of the fsl\_sub script available in
+  \url{https://github.com/aces/cbrain-plugins-neuro}; for PSOM:
+  through a specific development in PSOM).}
 \label{fig:cbrain-sub-tasking}

### 2016-04-27 paper.tex
Bold other names
For sub-tasking, Global Complexity and other small changes.

--- 2016-04-27 paper.tex
@@
 See Figure~\subref*{archi:service}. The workflow engine is available
-externally to the science gateway, in a service. The science gateway
+externally to the science gateway, as a service. The science gateway
 controls the service through a specific interaction (\texttt{d}) that
@@
 
-\subsection{Sub-tasking}
+\subsection{Sub-tasking} \note{Marc-e}{May I propose another name here? Either
+"Task-centric" or "Task Oriented". My reasoning is that the WE is just another
+task, as well as the sub-tasks themselves.}
 
@@
 establishing direct communication channels with them. Note how this
-peculiar execution model is totally supported by the sub-tasking
+peculiar execution model is well supported by the sub-tasking
 architecture.
@@
 letters from Fig 2}{Tristan}{Fixed.}, and workflow retrieval (\texttt{13} \&
-\texttt{14}). In Figure~\ref{fig:shiwa-pool-architecture},
+\texttt{14}). \note{Marc-e}{Maybe easier to follow if the following would be in fig
+ legend instead of here, as for 2 previous figures} In Figure~\ref{fig:shiwa-pool-architecture},
 interaction \texttt{f} of Figure~\subref*{archi:agent} is implemented
@@
 infrastructure in the architecture usually requires some technical
-effort (e.g., account creation, software installation, etc.), which is
+effort (e.g., account creation, software installation, APIs, etc.), which is
 why we keep it in the metric. The two metrics are summed to obtain the
@@
 on the infrastructure (\texttt{E$_1$}=2). Updating an engine version
-requires updates only on the infrastructure (\texttt{E$_2$}=1).  New
+in this architechture shares the same mechanism as version updates of existing
+tasks or tools, which requires an update on the infrastructure (\texttt{E$_2$}=1).  New
 workflows are integrated by creating a new task in the science gateway
@@
 architecture involves 11 interactions (\texttt{I$_2$=11}):
-\texttt{a$_1$} (appears twice: once for each infrastructure),
-\texttt{a$_2$}, \texttt{b} (twice: once for each infrastructure),
-\texttt{c$_1$} (twice: once for each infrastructure), \texttt{c$_2$}
-(twice: once for each workflow engine and infrastructure) and
+\texttt{a$_1$}, \texttt{a$_2$}, \texttt{b}, \texttt{c$_1$} (twice: once for each infrastructure),
+\texttt{c$_2$} (twice: once for each workflow engine and infrastructure) and
 \texttt{d} (twice: once for each workflow engine).
@@
   'Integration effort' with 'global complexity', as discussed in
-  github issue \#8.}
+  github issue \#8.} \note{Marc-e}{I understand both points. However, complexity is
+  addressed in Robustness as well... I liked the idea of development effort. It's a tough call.}
 

### 2016-04-27 paper.tex

--- 2016-04-27 paper.tex
@@
 variety of programming languages. Such tools were interfaced to
-computing infrastructures, in particular clusters, to execute tasks at
+computing infrastructures, in particular cluster schedulers, to execute tasks at
 a high throughput. For instance, FSL can launch tasks on SGE through

### 2016-04-28 paper.tex
Fixed #11

--- 2016-04-27 paper.tex
@@
   We investigate software architectures commonly used to integrate
-  workflow engines in science gateways, based on our experience with
-  the CBRAIN and VIP platforms, and on our past involvement in the
-  SHIWA project. We abstract software components and interactions in 6
-  architectures that we use to classify 18 systems. We compare these
-  architectures using metrics measuring global complexity, robustness,
-  extensibility, scalability and other specific features. Results show
-  the pros and cons of the different architectures: tight integration
-  and sub-tasking are the simplest architectures and should therefore
-  be the easiest to implement and the most robust ones, most
-  architectures perform the same in terms of extensibility, the pool
-  model is the most scalable and meta-workflows are only available in
-  nested workflows and workflow import. Our analysis provides insights
-  for software architects on how to integrate science gateways and
-  workflow engines.
+  workflow engines in science gateways. We abstract software
+  components and interactions in 6 architectures that we use to
+  classify 18 systems. We compare these architectures using metrics
+  measuring global complexity, robustness, extensibility, scalability
+  and other specific features. Results show the pros and cons of the
+  different architectures: tight integration and sub-tasking are the
+  simplest architectures and should therefore be the easiest to
+  implement and the most robust ones, most architectures perform the
+  same in terms of extensibility, the pool model is the most scalable
+  and meta-workflows are only available in nested workflows and
+  workflow import. Our analysis provides insights for software
+  architects on how to integrate science gateways and workflow
+  engines.
 \end{abstract}
@@
 regular web development tools and libraries to access distributed
-systems. CBRAIN and VIP are two such science gateways, from which we
-gathered experience about the architectures described in this paper.
-
-CBRAIN\footnote{\url{http://cbrain.mcgill.ca}}~\cite{SHER-14} is an
-open-source\footnote{\url{https://github.com/aces/cbrain}}, web-based collaborative platform addressing the
-challenges of data-heavy, computation-intensive neuroimaging
-research. It offers transparent web-based access to remote data
-sources, distributed computing sites, and an array of processing and
-visualization tools, within a secure environment. CBRAIN is deployed
-at a wide range of computing sites, including a network of national HPC
-centers in Canada, one in Korea, one in Germany, and on numerous local
-servers. As February 2016, CBRAIN has served over 420 users across 50
-cities in 22 countries and has a national computing allocation of 660 CPU years
-per annum.
-
-The Virtual Imaging Platform
-(VIP\footnote{\url{http://www.creatis.insa-lyon.fr/vip}})~\cite{GLAT-13}
-is a web portal for medical simulation and image data analysis. It
-leverages computing and storage resources available in the biomed
-Virtual Organization of the European Grid Infrastructure (EGI) to
-offer an open service to academic researchers worldwide. VIP
-completely masks the infrastructure to enable a transparent user
-experience. It uses the MOTEUR workflow engine~\cite{GLAT-13} and the
-DIRAC pilot-job framework~\cite{tsaregorodtsev2008dirac} to execute
-applications on EGI. Almost 900 users were registered in VIP by
-April 2016 and the average CPU consumption between January 2013 and
-March 2016 was 35 CPU years per month.
-
-The architectures described hereafter also build on experience with
-the SHIWA Simulation Platform~\cite{terstyanszky2014enabling}. The
-SHIWA Simulation Platform is a WS-PGRADE/gUSE science gateway that was
-developed in the European projects
-SHIWA\footnote{\url{http://www.shiwa-workflow.eu}} and
-ER-flow\footnote{\url{http://www.erflow.eu}}. It focuses on workflow
-interoperability for various disciplines, in particular through the
-building of meta-workflows, i.e. workflows that are built out of other
-workflows. It implements the so-called Coarse-Grained Interoperability
-framework where workflow engines are integrated in the science gateway
-as other applications and invoked through a dedicated service. The
-Fine-Grained Interoperability framework was also developed in the
-SHIWA project, where conversions between workflow languages are
-achieved through a common intermediate
-representation~\cite{plankensteiner-prodan-etal:2013}.
+systems, for instance CBRAIN~\cite{SHER-14} and VIP~\cite{GLAT-13}.
 

### 2016-04-28 paper.tex

--- 2016-04-27 paper.tex
@@
 regular web development tools and libraries to access distributed
-systems. CBRAIN and VIP are two such science gateways, from which we
-gathered experience about the architectures described in this paper.
-
-CBRAIN\footnote{\url{http://cbrain.mcgill.ca}}~\cite{SHER-14} is an
-open-source\footnote{\url{https://github.com/aces/cbrain}}, web-based collaborative platform addressing the
-challenges of data-heavy, computation-intensive neuroimaging
-research. It offers transparent web-based access to remote data
-sources, distributed computing sites, and an array of processing and
-visualization tools, within a secure environment. CBRAIN is deployed
-at a wide range of computing sites, including a network of national HPC
-centers in Canada, a cluster in Germany, and on numerous local research
-servers. CBRAIN is a task-centric platform where various workflow engines
-can be deployed on the infrastructure as any other task would. As February 2016,
-CBRAIN has served over 420 users across 50 cities in 22 countries and has a
-national computing allocation of 660 CPU years per annum.
-
-The Virtual Imaging Platform
-(VIP\footnote{\url{http://www.creatis.insa-lyon.fr/vip}})~\cite{GLAT-13}
-is a web portal for medical simulation and image data analysis. It
-leverages computing and storage resources available in the biomed
-Virtual Organization of the European Grid Infrastructure (EGI) to
-offer an open service to academic researchers worldwide. VIP
-completely masks the infrastructure to enable a transparent user
-experience. It uses the MOTEUR workflow engine~\cite{GLAT-13} and the
-DIRAC pilot-job framework~\cite{tsaregorodtsev2008dirac} to execute
-applications on EGI. Almost 900 users were registered in VIP by
-April 2016 and the average CPU consumption between January 2013 and
-March 2016 was 35 CPU years per month.
-
-The architectures described hereafter also build on experience with
-the SHIWA Simulation Platform~\cite{terstyanszky2014enabling}. The
-SHIWA Simulation Platform is a WS-PGRADE/gUSE science gateway that was
-developed in the European projects
-SHIWA\footnote{\url{http://www.shiwa-workflow.eu}} and
-ER-flow\footnote{\url{http://www.erflow.eu}}. It focuses on workflow
-interoperability for various disciplines, in particular through the
-building of meta-workflows, i.e. workflows that are built out of other
-workflows. It implements the so-called Coarse-Grained Interoperability
-framework where workflow engines are integrated in the science gateway
-as other applications and invoked through a dedicated service. The
-Fine-Grained Interoperability framework was also developed in the
-SHIWA project, where conversions between workflow languages are
-achieved through a common intermediate
-representation~\cite{plankensteiner-prodan-etal:2013}.
+systems, for instance CBRAIN~\cite{SHER-14} and VIP~\cite{GLAT-13}.
 

### 2016-04-28 paper.tex
Small edits following @merous' corrections.

@@
 variety of programming languages. Such tools were interfaced to
-computing infrastructures, in particular cluster schedulers, to execute tasks at
+computing infrastructures, in particular clusters\closednote{Tristan}{Reverted from Marc's 'cluster scheduler'}{}{}, to execute tasks at
 a high throughput. For instance, FSL can launch tasks on SGE through
@@
 values. Although simplistic, this definition covers both e-Science workflow systems and
-domain-specific pipelines.\note{Marc-e}{To remain consistent with above descriptions} Some workflow engines, usually e-Science
+domain-specific pipelines.\closednote{Marc-e}{To remain consistent with above descriptions}{Tristan}{yes!} Some workflow engines, usually e-Science
 ones, may transfer data across the infrastructure, and others, usually
@@
 frameworks~\cite{kacsuk2014science,ardizzone2012decide}. Other gateway
-instances were also built independently from such specialized framework, using
-regular web development tools and libraries to access distributed
-systems, for instance CBRAIN~\cite{SHER-14} and VIP~\cite{GLAT-13}.
+instances were also built independently from such specialized
+framework, using regular web development tools and libraries to access
+distributed systems. CBRAIN~\cite{SHER-14} and VIP~\cite{GLAT-13} are
+two such gateways.
 
@@
 on the infrastructure (\texttt{E$_1$}=2). Updating an engine version
-in this architechture shares the same mechanism as version updates of existing
-tasks or tools, which requires an update on the infrastructure (\texttt{E$_2$}=1).  New
-workflows are integrated by creating a new task in the science gateway
-through interactions \texttt{a$_1$} and \texttt{a$_2$}  (\texttt{E$_3$}=2). Adding a new
-infrastructure requires to update interactions \texttt{a$_1$}, \texttt{b} and
-\texttt{c} in the science gateway (\texttt{E$_4$}=3).
+in this architechture shares the same mechanism as version updates of
+other tasks, which requires an update on the
+infrastructure (\texttt{E$_2$}=1).  New workflows are integrated by
+creating a new task in the science gateway through interactions
+\texttt{a$_1$} and \texttt{a$_2$} (\texttt{E$_3$}=2). Adding a new
+infrastructure requires to update interactions \texttt{a$_1$},
+\texttt{b} and \texttt{c} in the science gateway (\texttt{E$_4$}=3).
 
@@
 architecture involves 11 interactions (\texttt{I$_2$=11}):
-\texttt{a$_1$}, \texttt{a$_2$}, \texttt{b}, \texttt{c$_1$} (twice: once for each infrastructure),
-\texttt{c$_2$} (twice: once for each workflow engine and infrastructure) and
-\texttt{d} (twice: once for each workflow engine).
+\texttt{a$_1$} (twice), \texttt{a$_2$}, \texttt{b} (twice), \texttt{c$_1$} (twice),
+\texttt{c$_2$} (twice) and
+\texttt{d} (twice).
 \closedanswerednote{Rafael}{I like the idea of what has been discussed in

### 2016-04-28 paper.tex
Renamed 'sub-tasking' with 'task encapsulation', following @merous' comment.

@@
 \newcommand{\note}[2]{\pdfmargincomment[color=yellow,author=#1,open=true]{#2}}
-\newcommand{\closednote}[4]{\pdfmargincomment[color=gray,author=#1-#3,open=false]{#1: #2 \textCR \textCR #3: #4}}
+\newcommand{\closednote}[4]{} % closed notes are hidden. This should obviously be a parameter of \note rather than a separate command.
 \newcommand{\answerednote}[4]{\pdfmargincomment[color=green,author=#1-#3,open=true]{#1: #2 \textCR \textCR #3: #4}}
@@
   and other specific features. Results show the pros and cons of the
-  different architectures: tight integration and sub-tasking are the
+  different architectures: tight integration and task encapsulation are the
   simplest architectures and should therefore be the easiest to
@@
 \hspace*{0.2\columnwidth}
-\subfloat[Sub-tasking]{
+\subfloat[Task encapsulation]{
 \def\svgwidth{0.6\columnwidth}
@@
             } \\
-  Sub-task & CBRAIN and PSOM~\cite{GLAT-16}$^n$, CBRAIN and FSL$^n$\\
+  Task & CBRAIN and PSOM~\cite{GLAT-16}$^n$, CBRAIN and FSL$^n$\\
   Pool & SHIWA pool~\cite{ROGE-13}\\
@@
 
-\subsection{Sub-tasking} \note{Marc-e}{May I propose another name here? Either
-"Task-centric" or "Task Oriented". My reasoning is that the WE is just another
-task, as well as the sub-tasks themselves.}
-
-See Figure~\subref*{archi:sub-task}. The workflow engine is a particular
-task that can submit sub-tasks to the science gateway through
-interaction \texttt{e}. The workflow engine keeps track of the
-dependencies between the sub-tasks, but their execution is delegated to
-the science gateway that executes them on the infrastructure through
-interaction \texttt{b}. Although the science gateway has no global
-vision of the workflow, it can keep track of the sub-tasks submitted
-by a given task, for instance to be able to cancel them when the task
-is canceled. The science gateway may also implement mechanisms to
-facilitate the handling of task dependencies, for instance basic
-dependency lists as available in most batch managers (see for instance
-attribute \texttt{depend} of option \texttt{-W} in
+\subsection{Task encapsulation} \answerednote{Marc-e}{May I propose
+  another name here? Either "Task-centric" or "Task Oriented". My
+  reasoning is that the WE is just another task, as well as the
+  sub-tasks themselves.}{Tristan}{I agree: sub-tasking is not the
+  integration model, it is the process that engines use when they
+  are integrated as tasks. I think 'task encapsulation' is more
+  consistent  with 'service invocation' than 'task-oriented'}
+
+See Figure~\subref*{archi:sub-task}. The workflow engine is wrapped in
+a particular task that can submit sub-tasks to the science gateway
+through interaction \texttt{e}. The workflow engine keeps track of the
+dependencies between the sub-tasks, but their execution is delegated
+to the science gateway that executes them on the infrastructure
+through interaction \texttt{b}. Although the science gateway has no
+global vision of the workflow, it can keep track of the sub-tasks
+submitted by a given task, for instance to be able to cancel them when
+the task is canceled. The science gateway may also implement
+mechanisms to facilitate the handling of task dependencies, for
+instance basic dependency lists as available in most batch managers
+(see for instance attribute \texttt{depend} of option \texttt{-W} in
 Torque\footnote{\url{http://docs.adaptivecomputing.com}}).
@@
 
-The sub-tasking architecture is implemented in CBRAIN~\cite{SHER-14}
-where it is used to integrate the FSL toolkit~\cite{Jenkinson2012782}
-and the PSOM workflow engine~\cite{bellec2012pipeline}.
+Task encapsulation is implemented in CBRAIN~\cite{SHER-14} where it is
+used to integrate the FSL toolkit~\cite{Jenkinson2012782} and the PSOM
+workflow engine~\cite{bellec2012pipeline}.
 
@@
 establishing direct communication channels with them. Note how this
-peculiar execution model is well supported by the sub-tasking
-architecture.
+peculiar execution model is well supported by task encapsulation.
 
@@
 \caption{CBRAIN architecture for workflow engine integration
-  (sub-tasking).  1: User sends data and workflow execution request to
+  (task encapsulation).  1: User sends data and workflow execution request to
   storage server(s) and CBRAIN portal. 2: CBRAIN portal sends
@@
                                      & \textbf{Service}
-                                     & \textbf{Sub-task}
+                                     & \textbf{Task}
                                      & \textbf{Pool}
@@
 
-\subsection{Sub-task}
-
-\paragraph{Global complexity} The sub-task architecture requires only 2
+\subsection{Task encapsulation}
+
+\paragraph{Global complexity} Task encapsulation requires only 2
 components (\texttt{I$_1$}=2).  It involves 5 interactions:
@@
 any other task upon user submission (\texttt{S$_1$}=0). This is a
-major interest of the sub-task architecture. Distributed engines are
+major interest of task encapsulation. Distributed engines are
 not supported by default (\texttt{S$_2$}=1). Task scheduling is
@@
 
-\subsection{Pool}
+\subsection{Pool model}
 
@@
 
-Tight integration and sub-tasking are globally the simplest
+Tight integration and task encapsulation are globally the simplest
 architectures, followed by service integration, workflow import (with
@@
 architectures. Robustness roughly leads to the same ordering of
-architectures, with tight integration and sub-tasking in the top
+architectures, with tight integration and task encapsulation in the top
 group, service integration and workflow import (with service
@@
 nested workflow architecture, with 2 infrastructures and 2 workflow
-services. Sub-tasking and workflow import (with service invocation)
+services. Task encapsulation and workflow import (with service invocation)
 perform a bit better when integrating new engines (\texttt{E$_1$}),
 which is interesting given the current profusion of engines with
-different characteristics.  However, sub-tasking, workflow import
+different characteristics.  However, task encapsulation, workflow import
 (with service invocation) and tight integration perform worse than the
 others when adding engine versions (\texttt{E$_2$}) due to the need to
-update infrastructure (sub-tasking), science gateway (tight
+update infrastructure (task encapsulation), science gateway (tight
 integration), or workflow converter (workflow import). Workflow import
@@
 some cases. At the same time, the scheduling constraints created by
-the sub-tasking and nested workflow architectures could well become
+task encapsulation and nested workflows could well become
 showstoppers depending on the type of the targeted
@@
 hardly cope with workflow engines being wrapped in computing tasks as
-done in sub-tasking. The availability of multiple engine instances
+done in task encapsulation. The availability of multiple engine instances
 could also become a critical feature for science gateways with
-important workloads, which would favor sub-tasking and pool.
+important workloads, which would favor task encapsulation and pool.
 
@@
 invocation) are the only architectures that support meta-workflows,
-which may be interesting in some cases. Also, sub-tasking and nested
+which may be interesting in some cases. Also, task encapsulation and nested
 workflows may complexify fine-grained debugging. \closednote{Rafael}{I would
@@
 assumes that all criteria have equal weights while real systems would
-favor some of them. Overall, sub-tasking, and workflow import (with
+favor some of them. Overall, task encapsulation, and workflow import (with
 service invocation) perform a bit better than tight integration,
@@
                                     & \textbf{Service}
-                                    & \textbf{Sub-task}
+                                    & \textbf{Task}
                                     & \textbf{Pool}
@@
 invocation, for instance, may not be always that clear in practice. Service
-invocation may also be combined with sub-tasking in some cases. However, the
+invocation may also be combined with task encapsulation in some cases. However, the
 criteria discussed in this work would still apply to broadly compare and categorize
@@
 the same framework as the other architectures. Other types of
-instantiation, for instance nested workflows with sub-tasking, could
+instantiation, for instance nested workflows with task encapsulation, could
 also be envisaged. We chose to limit ourselves to instantiations with
@@
 in real systems, and because service invocation is largely
-used. Nested workflows with sub-tasking, in particular, could be an
+used. Nested workflows with task encapsulation, in particular, could be an
 interesting architecture to explore.
@@
 and data control features (interactions \texttt{b} and \texttt{c}),
-which would favor sub-tasking. Similarly, adding a new type of engine
+which would favor task encapsulation. Similarly, adding a new type of engine
 may facilitate integration of a new infrastructure when interactions

### 2016-04-28 paper.tex
Fixed @merous note on description of pool architecture

@@
   invocation).  Step \texttt{3} corresponds to interaction \texttt{d}
-  on Figure~\ref{fig:architectures}, step \texttt{2} corresponds to
+  on Figure~\ref{fig:architectures}(II), step \texttt{2} corresponds to
   interaction \texttt{c$_1$}, and step \texttt{4} maps to interactions
@@
 
-\subsection{Task encapsulation} \answerednote{Marc-e}{May I propose
+\subsection{Task encapsulation}
+\closednote{Marc-e}{May I propose
   another name here? Either "Task-centric" or "Task Oriented". My
@@
   server transfers results to storage server(s). Interaction
-  \texttt{b} in Figure~\ref{fig:architectures} is implemented by steps
+  \texttt{b} in Figure~\ref{fig:architectures}(III) is implemented by steps
   \texttt{4}, \texttt{5}, \texttt{7} and \texttt{8} (regular
@@
 The pool model was implemented in the SHIWA pool~\cite{ROGE-13}
-diagrammed in Figure~\ref{fig:shiwa-pool-architecture}. In this
-Figure, interaction \texttt{d} of Figure~\subref*{archi:agent} is
-implemented in 3 different types of calls to the pool: workflow
-submission (\texttt{1} \& \texttt{2}), workflow status retrieval
-(\texttt{0} \& \texttt{3})\closednote{sorina}{I suppose letters come here
-from the initial pool-model figure; it's a bit puzzling when mixed with
-letters from Fig 2}{Tristan}{Fixed.}, and workflow retrieval (\texttt{13} \&
-\texttt{14}). \note{Marc-e}{Maybe easier to follow if the following would be in fig
- legend instead of here, as for 2 previous figures} In Figure~\ref{fig:shiwa-pool-architecture},
-interaction \texttt{f} of Figure~\subref*{archi:agent} is implemented
-through 2 types of calls: workflow instance retrieval (\texttt{4},
-\texttt{5} and \texttt{6}), and workflow instance update (\texttt{11}
-and \texttt{12}). Workflow instance retrieval is used by the agents to
-fetch work from the pool. Workflow instance update is used by the
-agents to update workflow statuses. In the SHIWA pool, agents can wrap
-different types of workflow engines to execute workflows expressed
-with different languages. Calls \texttt{0}, \texttt{7}, \texttt{8} and
-\texttt{9} on Figure~\ref{fig:shiwa-pool-architecture} are used by
-workflow engine plugins to declare their supported language and launch
-engines, and by workflow engines to report their status to the
-agent. These calls are specific to the SHIWA Pool implementation of
-the \texttt{agent} component and therefore have no corresponding
-representation in Figure~\subref*{archi:agent}.
+diagrammed in Figure~\ref{fig:shiwa-pool-architecture}. In the SHIWA
+pool, agents can wrap different types of workflow engines to execute
+workflows expressed with different languages.
 
@@
 \includegraphics[width=1.5\columnwidth]{figures/pool-interactions.pdf}
-\caption{Architecture of the SHIWA pool. Circle-terminated arrows indicate messages
-  that are broadcast to all pool clients. Figure reproduced
+\caption{Architecture of the SHIWA pool. Circle-terminated arrows
+  indicate messages that are broadcast to all pool
+  clients. Interaction \texttt{d} of Figure~\ref{fig:architectures}(IV) is
+  implemented in 3 different types of calls to the pool: workflow
+  submission (\texttt{1} \& \texttt{2}), workflow status retrieval
+  (\texttt{0} \& \texttt{3})\closednote{sorina}{I suppose letters come
+    here from the initial pool-model figure; it's a bit puzzling when
+    mixed with letters from Fig 2}{Tristan}{Fixed.}, and workflow
+  retrieval (\texttt{13} \& \texttt{14}). \closednote{Marc-e}{Maybe
+    easier to follow if the following would be in fig legend instead
+    of here, as for 2 previous figures}{Tristan}{fixed} Interaction
+  \texttt{f} is implemented through 2 types of calls: workflow
+  instance retrieval (\texttt{4}, \texttt{5} and \texttt{6}), and
+  workflow instance update (\texttt{11} and \texttt{12}). Workflow
+  instance retrieval is used by the agents to fetch work from the
+  pool. Workflow instance update is used by the agents to update
+  workflow statuses.  Calls \texttt{0}, \texttt{7}, \texttt{8} and
+  \texttt{9} are used by
+  workflow engine plugins to declare their supported language and to
+  launch engines, and by workflow engines to report their status to
+  the agent. These calls are specific to the SHIWA Pool implementation
+  of the \texttt{agent} component and therefore have no corresponding
+  representation in Figure~\ref{fig:architectures}(IV). Figure reproduced
   from~\cite{ROGE-13}.}
@@
   in the Science Gateway (step \texttt{1}, interaction \texttt{d} on
-  Figure 2(V)-Right. Ten different child engines can be used by nested
+  Figure~\ref{fig:architectures}(V)-Right. Ten different child engines can be used by nested
   workflows, invoked through the \texttt{Submission service} (step
@@
 
+\enlargethispage{5mm}
 \bibliographystyle{elsarticle-num}

### 2016-04-28 paper.tex
Replaced 'global complexity' with 'integration challenge'

@@
 \newcommand{\answerednote}[4]{\pdfmargincomment[color=green,author=#1-#3,open=true]{#1: #2 \textCR \textCR #3: #4}}
-\newcommand{\closedanswerednote}[6]{\pdfmargincomment[color=gray,author=#1-#3-#5,open=true]{#1: #2 \textCR \textCR #3: #4 \textCR \textCR #5: #6}}
+\newcommand{\closedanswerednote}[6]{}
 
@@
   classify 18 systems. We compare these architectures using metrics
-  measuring global complexity, robustness, extensibility, scalability
+  measuring integration challenge, robustness, extensibility, scalability
   and other specific features. Results show the pros and cons of the
@@
 Section~\ref{sec:evaluation}, we introduce metrics to measure
-global complexity, robustness, extensibility, scalability and other
+integration challenge, robustness, extensibility, scalability and other
 specific features of the architectures. We use these metrics to
@@
 evaluated in Table~\ref{table:evaluation}. We use 5 main criteria to
-evaluate the architectures: global complexity, execution complexity,
+evaluate the architectures: integration challenge, execution complexity,
 extensibility, scalability and other specific features. Criteria break
 down to specific metrics where \emph{lower value indicates better
-  performance}. For each criterion, a total score is computed by
+  performance}. For each criterion, a global score is computed by
 summing up the individual metrics. We ensure that the different
@@
                                      & \textbf{Import} \\
-\cellcolor[HTML]{EEEEEE}\textbf{Global complexity}& \multicolumn{6}{l}{\cellcolor[HTML]{EEEEEE}}\\
+\cellcolor[HTML]{EEEEEE}\textbf{Integration challenge}& \multicolumn{6}{l}{\cellcolor[HTML]{EEEEEE}}\\
   Total components -- \texttt{I$_1$} & \cellcolor[HTML]{99FF99}2
@@
 
-\emph{Global complexity} is obtained by counting the total number of
+\emph{Integration challenge} is obtained by counting the total number of
 interactions and components on the architecture diagram. It breaks
@@
 effort (e.g., account creation, software installation, APIs, etc.), which is
-why we keep it in the metric. The two metrics are summed to obtain the
-total score for global complexity. Global complexity measures the
+why we keep it in the metric. The two metrics are summed to obtain a
+global score that measures the
 total number of software pieces to develop, including interactions and
@@
 Figure~\ref{fig:architectures}. \texttt{R$_1$} and \texttt{R$_2$} are
-summed to obtain the total score for this criterion, which measures
+summed to obtain the global score for this criterion, which measures
 the total number of software pieces that are specifically involved in
@@
 
-\paragraph{Global complexity} This architecture does not require any
+\paragraph{Integration challenge} This architecture does not require any
 component in addition to the science gateway and infrastructure
@@
 
-\paragraph{Global complexity} Service invocation requires a workflow service
+\paragraph{Integration challenge} Service invocation requires a workflow service
 in addition to the science gateway and infrastructure
@@
 
-\paragraph{Global complexity} Task encapsulation requires only 2
+\paragraph{Integration challenge} Task encapsulation requires only 2
 components (\texttt{I$_1$}=2).  It involves 5 interactions:
@@
 
-\paragraph{Global complexity} The pool model requires a workflow pool and an
+\paragraph{Integration challenge} The pool model requires a workflow pool and an
 agent in addition to the science gateway and infrastructure
@@
 
-\paragraph{Global complexity} Setting up a nested workflow
+\paragraph{Integration challenge} Setting up a nested workflow
 architecture with service invocation requires a science gateway, 2
@@
   'Integration effort' with 'global complexity', as discussed in
-  github issue \#8.} \note{Marc-e}{I understand both points. However, complexity is
-  addressed in Robustness as well... I liked the idea of development effort. It's a tough call.}
+  github issue \#8.}
 
@@
 are (\texttt{R$_2$=6}).
+\note{Marc-e}{I understand both points. However, complexity is
+  addressed in Robustness as well... I liked the idea of development effort. It's a tough call.}
 
@@
 
-\paragraph{Global complexity} Workflow import with service invocation
+\paragraph{Integration challenge} Workflow import with service invocation
 requires the same components as for service invocation
@@
                                     & \textbf{Import} \\
-  Global complexity (normalized total) &
+  Integration (normalized total)    &
                                     \cellcolor[HTML]{99FF99}{0.00}
@@
 \caption{Overall evaluation. Brighter colors and lower scores indicate better performance. Scores
-  are obtained by summing the normalized total scores ($m'$ values) of
+  are obtained by summing the normalized global scores ($m'$ values) of
   each criterion in Table~\ref{table:evaluation} and the colors are obtained by normalizing these scores

### 2016-04-28 paper.tex
Closed note
Updates in cell colors in tables.

@@
                                      & \cellcolor[HTML]{99FF99}\textbf{7}
-                                     & \cellcolor[HTML]{99D199}\textbf{11}
+                                     & \cellcolor[HTML]{99D299}\textbf{11}
                                      & \cellcolor[HTML]{999999}\textbf{16}
@@
                                      & \cellcolor[HTML]{99CC99}1\\
-  Specific interactions -- \texttt{R$_2$} & \cellcolor[HTML]{99EA99}2
+  Specific interactions -- \texttt{R$_2$} & \cellcolor[HTML]{99EB99}2
                                      & \cellcolor[HTML]{99D699}3
                                      & \cellcolor[HTML]{99FF99}1
-                                     & \cellcolor[HTML]{99C199}4
+                                     & \cellcolor[HTML]{99C299}4
                                      & \cellcolor[HTML]{999999}6
@@
 \cellcolor[HTML]{EEEEEE}\textbf{Extensibility}& \multicolumn{6}{l}{\cellcolor[HTML]{EEEEEE}}\\
-  New engine type -- \texttt{E$_1$}  & \cellcolor[HTML]{99C499}3
-                                     & \cellcolor[HTML]{99A799}4
-                                     & \cellcolor[HTML]{99E199}2
-                                     & \cellcolor[HTML]{99C499}3
+  New engine type -- \texttt{E$_1$}  & \cellcolor[HTML]{99C599}3
+                                     & \cellcolor[HTML]{99A899}4
+                                     & \cellcolor[HTML]{99E299}2
+                                     & \cellcolor[HTML]{99C599}3
                                      & \cellcolor[HTML]{999999}4.5
@@
                                      & \cellcolor[HTML]{99DD99}4\\
-  \textbf{Total} (difficulty to extend) & \cellcolor[HTML]{99D199}\textbf{10}
-                                     & \cellcolor[HTML]{99D199}\textbf{10}
+  \textbf{Total} (difficulty to extend) & \cellcolor[HTML]{99D299}\textbf{10}
+                                     & \cellcolor[HTML]{99D299}\textbf{10}
                                      & \cellcolor[HTML]{99FF99}\textbf{8}
@@
   are the minimal and maximal values among all architectures. The RGB hexadecimal color code of the cell
-  is \texttt{\#99XX99}, where X=F-4m'. \closednote{Tristan}{Use 6. Round to the nearest integer.}{Tristan}{Fixed.}}
+  is \texttt{\#99XX99}, where X=\texttt{round}(F-6m') (\texttt{round} rounds a number to the nearest integer). \closednote{Tristan}{Use 6. Round to the nearest integer.}{Tristan}{Fixed.}}
 \label{table:evaluation}
@@
 are (\texttt{R$_2$=6}).
-\note{Marc-e}{I understand both points. However, complexity is
-  addressed in Robustness as well... I liked the idea of development effort. It's a tough call.}
+\closednote{Marc-e}{I understand both points. However, complexity is
+  addressed in Robustness as well... I liked the idea of development effort. It's a tough call.}{Tristan}{Maybe 'integration challenge' is a good compromise.}
 
@@
                                     & \cellcolor[HTML]{99FF99}{0.00}
-                                    & \cellcolor[HTML]{99D199}{0.44}
+                                    & \cellcolor[HTML]{99D299}{0.44}
                                     & \cellcolor[HTML]{999999}{1.00}
@@
                                     & \cellcolor[HTML]{99D399}{0.43}\\
-  Extensibility (normalized total)  & \cellcolor[HTML]{99D199}{0.44}
-                                     & \cellcolor[HTML]{99D199}{0.44}
+  Extensibility (normalized total)  & \cellcolor[HTML]{99D299}{0.44}
+                                     & \cellcolor[HTML]{99D299}{0.44}
                                      & \cellcolor[HTML]{99FF99}{0.00}
@@
                                      & \cellcolor[HTML]{99CC99}{0.50}
-                                     & \cellcolor[HTML]{99DD99}{0.00}\\
-                                    & \cellcolor[HTML]{99E699}\textbf{2.1}
+                                     & \cellcolor[HTML]{99FF99}{0.00}\\
+                                    & \cellcolor[HTML]{99E799}\textbf{2.1}
                                     & \cellcolor[HTML]{99E699}\textbf{2.1}
                                     & \cellcolor[HTML]{99FE99}\textbf{1.5}
-                                    & \cellcolor[HTML]{99EE99}\textbf{1.9}
+                                    & \cellcolor[HTML]{99EF99}\textbf{1.9}
                                     & \cellcolor[HTML]{999999}\textbf{4.0}

### 2016-04-28 paper.tex
Updates in introduction

@@
   We investigate software architectures commonly used to integrate
-  workflow engines in science gateways. We abstract software
-  components and interactions in 6 architectures that we use to
-  classify 18 systems. We compare these architectures using metrics
-  measuring integration challenge, robustness, extensibility, scalability
-  and other specific features. Results show the pros and cons of the
-  different architectures: tight integration and task encapsulation are the
-  simplest architectures and should therefore be the easiest to
-  implement and the most robust ones, most architectures perform the
-  same in terms of extensibility, the pool model is the most scalable
-  and meta-workflows are only available in nested workflows and
-  workflow import. Our analysis provides insights for software
-  architects on how to integrate science gateways and workflow
-  engines.
+  workflow engines in science gateways. We describe and evaluate 6
+  architectures with metrics measuring integration challenge,
+  robustness, extensibility, scalability and other specific
+  features. Results show the pros and cons of the different
+  architectures: tight integration and task encapsulation are the
+  simplest architectures and are therefore the easiest to integrate
+  and the most robust ones, most architectures perform the same in
+  terms of extensibility, the pool model is the most scalable and
+  meta-workflows are only available in nested workflows and workflow
+  import. Our analysis provides insights for software architects on
+  how to integrate science gateways and workflow engines.
 \end{abstract}
@@
 the past 7 years. It also builds on the lessons learned from several
-science gateway and workflow-related projects, including
+science gateway and workflow projects, including
 SHIWA\footnote{\url{http://www.shiwa-workflow.eu}} and related
@@
 infrastructures. In Section~\ref{sec:architectures}, we describe 6
-architectures in a consistent framework based on the functional
-interactions between the main software components involved. In
-Section~\ref{sec:evaluation}, we introduce metrics to measure
-integration challenge, robustness, extensibility, scalability and other
-specific features of the architectures. We use these metrics to
-evaluate and compare the 6 architectures. The paper closes on a
-discussion and conclusion that highlight architectural recommendations
-for science gateway and workflow engine architects.
+architectures in a consistent framework that highlight the functional
+interactions between their main software components. In
+Section~\ref{sec:evaluation}, we evaluate the architectures with
+metrics measuring integration challenge, robustness, extensibility,
+scalability and other specific features. The paper closes on a
+discussion and conclusion that compare architectures and highlight
+recommendations for science gateway architects.
 

### 2016-04-29 paper.tex
Updates in abstract and introduction

@@
   meta-workflows are only available in nested workflows and workflow
-  import. Our analysis provides insights for software architects on
-  how to integrate science gateways and workflow engines.
+  import. Our analysis provides insights for science gateway
+  architects and workflow engine developers.
 \end{abstract}
@@
 services offered by science gateways for the efficient and transparent
-execution of applications on distributed infrastructures. Several
+exploitation of distributed infrastructures. Several
 software architectures can be adopted to integrate workflow engines in
-science gateways, with important consequences on the required
-development effort and on the resulting system. This paper describes,
+science gateways, with important consequences on the 
+development effort required and on the resulting system. This paper describes,
 exemplifies and compares such architectures, based on
@@
 discussion and conclusion that compare architectures and highlight
-recommendations for science gateway architects.
+recommendations for science gateway architects and workflow engine developers.
 
@@
 computing infrastructures, in particular clusters\closednote{Tristan}{Reverted from Marc's 'cluster scheduler'}{}{}, to execute tasks at
-a high throughput. For instance, FSL can launch tasks on SGE through
-its fsl\_sub tool, Nipype has execution plugins for S/OGE, Torque,
-IPython, ssh and multi-processor servers, and PSOM is interfaced to
-Torque, SGE, MOAB, LSF, and HTCondor. Some of these tools also support
+a high throughput. Some of these tools also support
 advanced features such as provenance tracking (Nipype, PSOM),
@@
 Integration between workflow engines and science gateways varies
-across systems. Some science gateways may be tailored to a particular
-engine that manages critical functions such as task execution, data
-transfers and authentication. Other gateways may be more general and
-host applications executed by different types of engines. Finally,
-some science gateways may not use any workflow engine and focus on the
-execution of simple tasks -- this class of systems is not covered by
-our analysis.
+across systems. Some science gateways are tailored to a particular
+engine while other ones are more general and host applications
+executed by different types of engines. Some gateways may not even use any
+workflow engine and focus on the execution of simple tasks.
 
@@
 workflow engines may be integrated over time, and different kinds of
-infrastructure can be targeted.
-
-Scalability is also a crucial concern, since science gateways are
-multi-user systems that manage substantial amounts of data and execute
-computationally-intensive analyses. To this end, science gateways
-may include different instances of the same engine type and implement
-some load-balancing policy among them. They may also use advanced
-task scheduling policies on the infrastructure, e.g. to ensure fairness
-among users~\cite{FERR-14,CPE:CPE3708} or to improve
+infrastructure can be targeted.Scalability is also a crucial concern
+for such multi-user systems that manage substantial amounts of data
+and execute computationally-intensive analyses. To this end, science
+gateways may balance the load among different instances of the same
+engine. They may also use advanced task scheduling policies on the
+infrastructure, e.g. to ensure fairness among
+users~\cite{FERR-14,CPE:CPE3708} or to improve
 fault-tolerance~\cite{wilkins2008teragrid,FERR-13}.
 
-Robustness is a highly desirable characteristic in production gateways. In this respect, the
-multiplicity of software components and interactions involved in architectures is
-a challenge as it often leads to complex systems.
-Architectures showing reduced system complexity will greatly
-facilitate the implementation effort towards robust interactions
-which, in turn, have a positive impact on characteristics such as
-gateway predictability, transparence, general reliability and results
-traceability and reproducibility; characteristics that are also key to
-a good user experience. \closednote{Marc-Etienne}{I find that
-  something is weak in this last paragraph. Robustness will help with
-  transparency, sure, but it seems only a small facet.  Also, the text
-  seem to timidly favor architectures with less
-  components/interactions without much explanation... Complex systems
-  can be robust, but at greater costs...  Here is one: "Architectures
+Robustness is a highly desirable characteristic as well. Architectures
 showing reduced system complexity will greatly facilitate the
-  implementation effort towards robust interactions which, in turn,
-  have a positive impact on characteristics such as gateway
-  predictability, transparence, general reliability and results
-  traceability and reproducibility; characteristics that are also key
-  to a good user experience."}{Tristan}{Agreed, fixed as suggested.}
+implementation effort towards robust interactions which, in turn, have
+a positive impact on characteristics such as gateway predictability,
+transparence, general reliability and results traceability and
+reproducibility; characteristics that are also key to a good user
+experience. \closednote{Marc-Etienne}{I find that something is weak in
+  this last paragraph. Robustness will help with transparency, sure,
+  but it seems only a small facet.  Also, the text seem to timidly
+  favor architectures with less components/interactions without much
+  explanation... Complex systems can be robust, but at greater
+  costs...  Here is one: "Architectures showing reduced system
+  complexity will greatly facilitate the implementation effort towards
+  robust interactions which, in turn, have a positive impact on
+  characteristics such as gateway predictability, transparence,
+  general reliability and results traceability and reproducibility;
+  characteristics that are also key to a good user
+  experience."}{Tristan}{Agreed, fixed as
+  suggested.}\note{Tristan}{Work on the this sentence again.}
 
@@
 Various science gateways have been developed, including frameworks
-used as building blocks to assemble science gateways. The main
-frameworks are Apache Airavata~\cite{marru2011apache}, the Catania
-Science Gateway Framework~\cite{Ardizzone2012} and
-WS-PGRADE/gUSE~\cite{Kacsuk2012}. Numerous science gateway instances
-were built using such
-frameworks~\cite{kacsuk2014science,ardizzone2012decide}. Other gateway
-instances were also built independently from such specialized
-framework, using regular web development tools and libraries to access
-distributed systems. CBRAIN~\cite{SHER-14} and VIP~\cite{GLAT-13} are
-two such gateways.
-
-\subsection{Infrastructures}
+used as building blocks to assemble other systems, for instance Apache
+Airavata~\cite{marru2011apache}, the Catania Science Gateway
+Framework~\cite{Ardizzone2012} and
+WS-PGRADE/gUSE~\cite{Kacsuk2012}. Numerous science gateways were built
+using such frameworks~\cite{kacsuk2014science,ardizzone2012decide},
+and other ones, e.g. CBRAIN~\cite{SHER-14} and VIP~\cite{GLAT-13},
+were built without any framework.
+
+\subsection{Infrastructure}
 
 In this paper, infrastructure consists of the computing and storage
-resources involved in the workflow execution, as well as the software
-services used to access these resources. Infrastructures can be composed of
-computing or file servers, databases, clusters, grids or clouds. Some
-workflow engines and science gateways may assume specific
-characteristics about the infrastructure, such as the presence of a
-shared file system between the computing nodes, the availability of a
-global task meta-scheduler, the presence of a file catalog, etc. In
-the analysis presented below, such specific requirements are not
-discussed. Instead, we consider the infrastructure as an abstract
-system that can execute tasks and store data regardless of the
-enabling mechanisms.
+resources involved in workflow execution, as well as the software
+services used to access these resources. Infrastructure can be
+composed of computing or file servers, databases, clusters, grids or
+clouds. Some workflow engines and science gateways may assume specific
+characteristics, such as the presence of a shared file system between
+the computing nodes, the availability of a global task meta-scheduler,
+the presence of a file catalog, etc. In the analysis presented below,
+such specific requirements are not discussed. Instead, we consider the
+infrastructure as an abstract system that can execute tasks and store
+data regardless of the enabling mechanisms.
 

### 2016-04-29 paper.tex
Update paper.tex
read up to line 1070 or so

@@
 
-\title{Software Architectures to Integrate Workflow Engines in Science Gateways}
+\title{Software Architectures for Integrating Workflow Engines in Science Gateways} \note{Naj}{'what' software architectures? what is their qualifier, adjective?}
 
@@
 \begin{abstract}
-  We investigate software architectures commonly used to integrate
-  workflow engines in science gateways. We describe and evaluate 6
-  architectures with metrics measuring integration challenge,
+  We investigate six software architectures commonly used to integrate
+  workflow engines in science gateways. We describe and evaluate these software
+  architectures with metrics that measure integration complexity,
   robustness, extensibility, scalability and other specific
-  features. Results show the pros and cons of the different
-  architectures: tight integration and task encapsulation are the
-  simplest architectures and are therefore the easiest to integrate
-  and the most robust ones, most architectures perform the same in
-  terms of extensibility, the pool model is the most scalable and
-  meta-workflows are only available in nested workflows and workflow
+  features\note{Naj}{Please be precise, what are they?}. Results illustrate the pros and cons of different
+  architectures: Simple architectures with tight integration and task encapsulation are the
+  the easiest to integrate and are most robust.Extensibility is equivalent in most architectures.
+The pool model is the most scalable one and meta-workflows are only available in nested workflows and workflow
   import. Our analysis provides insights for science gateway
-  architects and workflow engine developers.
+  architects and workflow engine developers. \note{Naj}{I find this abstract to be vague, it can be made more concise and poignant}
 \end{abstract}
@@
 
-Workflow engines are critical components in the ecosystem of tools and
-services offered by science gateways for the efficient and transparent
-exploitation of distributed infrastructures. Several
+Workflow engines are critical for the efficient and transparent
+exploitation of distributed infrastructures in the ecosystem of tools and
+services offered by science gateways . Several
 software architectures can be adopted to integrate workflow engines in
-science gateways, with important consequences on the 
-development effort required and on the resulting system. This paper describes,
-exemplifies and compares such architectures, based on
-system-independent representations of their main components and
-interactions.
-
-Our analysis is fed by experience gathered from the development and
+science gateways, however 
+development effort required and the resulting system vary across different architectures.
+This paper and our analytical approach to comparison of software architectures is informed by our experince in the development and
 sustained operation of the CBRAIN~\cite{SHER-14} and
 VIP~\cite{GLAT-13} science gateways for medical image analysis during
-the past 7 years. It also builds on the lessons learned from several
+the past 7 years; as well as by lessons learned from several
 science gateway and workflow projects, including
 SHIWA\footnote{\url{http://www.shiwa-workflow.eu}} and related
-initiatives.
-
-The remainder of this Section provides background information and
-definitions about workflow engines, science gateways and
-infrastructures. In Section~\ref{sec:architectures}, we describe 6
-architectures in a consistent framework that highlight the functional
+initiatives. This paper describes,
+provides examples and compares such architectures, based on
+system-independent representations of their main components and
+interactions. In this section we provide background information and
+definitions of different workflow engines, science gateways and
+infrastructures. In Section~\ref{sec:architectures}, we describe six
+architectures within a consistent framework that underlines the functional
 interactions between their main software components. In
-Section~\ref{sec:evaluation}, we evaluate the architectures with
+Section~\ref{sec:evaluation}, different architectures are evaluated with
 metrics measuring integration challenge, robustness, extensibility,
-scalability and other specific features. The paper closes on a
-discussion and conclusion that compare architectures and highlight
+scalability and other specific features. We compare different architectures and highlight
 recommendations for science gateway architects and workflow engine developers.
@@
 
-At the same time, toolboxes were emerging in various scientific
+At the same time, toolboxes have been emerging in various scientific
 domains to facilitate the assembly of software components in
@@
 a high throughput. Some of these tools also support
-advanced features such as provenance tracking (Nipype, PSOM),
+advanced features such as provenance tracking (Nipype, PSOM), \note{Naj}{should we not mention PMP which is used for CIVET?}
 redundancy detection across analyses to avoid re-computation (PSOM),
@@
 
-For the sake of the analysis conducted in this paper, we define a
-\emph{workflow engine} (also abbreviated \emph{engine}) as a piece of
+In this paper, the working definition of a
+\emph{workflow engine} (also abbreviated \emph{engine}) is "a piece of
 software that submits interdependent computing \emph{tasks} to an
@@
 that may consist of files, database entries or simple parameter
-values. Although simplistic, this definition covers both e-Science workflow systems and
+values." Although simplistic, this definition covers both e-Science workflow systems and
 domain-specific pipelines.\closednote{Marc-e}{To remain consistent with above descriptions}{Tristan}{yes!} Some workflow engines, usually e-Science
@@
 locations. Workflow engines are part of this ecosystem as core
-components to implement and execute applications.
+components to implement and execute applications. \note{provide examples?}
 
@@
 workflow engines may be integrated over time, and different kinds of
-infrastructure can be targeted.Scalability is also a crucial concern
+infrastructure can be targeted. Scalability is also a crucial concern
 for such multi-user systems that manage substantial amounts of data
@@
 specific types of interactions that may be implemented using different
-components and interactions. They are represented with dotted arrows
+components and interactions. \note{Naj}{three 'interactions' in one sentence} They are represented with dotted arrows
 and labeled with \texttt{*}. An architecture that has an abstract
@@
 
-Red color is used to represent the components and interactions that
+Red color (WHERE?) is used to represent the components and interactions that
 are specific to workflow execution, i.e. that are not involved in any
@@
 numbers, etc. Administrator refers to a science gateway user with
-extended privileges to install applications.
+extended privileges to install applications. \note{Naj}{I propose a table of definitions.}
 
@@
   Service & \pbox{1.5\columnwidth}{
-            Apache Airvata~\cite{marru2011apache}, e-BioInfra~\cite{shahand2015data}$^{w,n}$, HubZero with Pegasus~\cite{CPE:CPE3257}, MoSGrid~\cite{krüger2014mosgrid}$^w$, System in~\cite{wu2010accelerating}, Vine Toolkit~\cite{DBLP:journals/scpe/SzejnfeldDKKKKLPTWDNW10}, Virtual Imaging Platform~\cite{GLAT-13}$^n$, WS-PGRADE/gUSE framework~\cite{Kacsuk2012},
+            Apache Airvata~\cite{marru2011apache}, e-BioInfra~\cite{shahand2015data}$^{w,n}$, HubZero with Pegasus~\cite{CPE:CPE3257}, MoSGrid~\cite{krüger2014mosgrid}$^w$, System in~\cite{wu2010accelerating}, Vine Toolkit~\cite{DBLP:journals/scpe/SzejnfeldDKKKKLPTWDNW10}, Virtual Imaging Platform~\cite{GLAT-13}$^n$, WS-PGRADE/gUSE framework~\cite{Kacsuk2012},
             Science gateways in~\cite{kacsuk2014science}$^w$
@@
 called \emph{child} workflow, that is executed by a \emph{child}
-engine. The parent and child engines might use different workflow
+engine. \note{Naj} {this reads funny} The parent and child engines might use different workflow
 description languages. They might also execute workflows on different
@@
 The 4 metrics are summed to obtain a global index that measures the
-difficulty to extend the architecture. Note that some extensions may
+difficulty to extend the architecture.
+\note{Naj}{I would rather compare each metric separately, this woudl also give a sense of how different architectures compare across metrics} Note that some extensions may
 involve several metrics in practice. For instance, adding a new type
@@
 but with some difficulty, and 2 means that the feature could only be
-implemented with a nonsensical amount of effort. Four different
+implemented with a nonsensical amount of effort\note{Naj}{how do you metricize "nonesensical amount fo effors?"}. Four different
 features are identified:

### 2016-04-29 paper.tex
Update paper.tex

@@
   the architecture used to integrate the workflow engine in the
-  science gateway. Some architectures, however, complexify the task
-  scheduling problem by introducing additional software layers or
-  create tasks with specific characteristics. \texttt{S$_3$}=0 when
+  science gateway. Some architectures, however, complicate the task
+  scheduling problem by introducing additional software layers, or by
+  creating tasks with specific characteristics. \texttt{S$_3$}=0 when
   the architecture does not introduce any additional complexity to the

### 2016-05-02 paper.tex
Updates in Sections 2 to 5

--- 2016-04-29 paper.tex
@@
 are specific to workflow execution, i.e. that are not involved in any
-other process. Tasks refer to computing tasks created by the workflow
-engine and executed on the infrastructure. Data represents any type of
-file or database involved in workflow execution and stored on the
-infrastructure. It does not cover workflow parameters such as strings,
-numbers, etc. Administrator refers to a science gateway user with
-extended privileges to install applications.
+other process. Tasks refer to computing tasks executed on the
+infrastructure. Data represents any type of file or database entry involved
+in workflow execution and stored on the infrastructure. It does not
+cover workflow parameters such as strings, numbers, etc. Administrator
+refers to a science gateway user with extended privileges to install
+applications.
 
@@
 The interactions involved in the architectures are described below and
-labeled consistently with Figure~\ref{fig:architectures}.
+labeled as in Figure~\ref{fig:architectures}.
 
@@
   (\texttt{a$_1$}) the programs used in the workflow are installed on
-  the infrastructure, which may require administrative privileges on
-  the infrastructure; (\texttt{a$_2$}) the workflow is configured in
-  the science gateway so that it becomes available to users. Note that
-  integrating a workflow is not the same process as integrating a
-  workflow \emph{engine}.
+  the infrastructure, which may require administrative privileges;
+  (\texttt{a$_2$}) the workflow is configured in the science gateway
+  so that it becomes available to users. Note that integrating a
+  workflow is not the same process as integrating a workflow
+  \emph{engine}.
 \item Task control: operations to manage tasks on the infrastructure,
@@
   model, user is exposed to the workflow topology, i.e. to the outputs
-  of the individual tasks, their statuses and so on.
+  of the individual tasks, their statuses, dependencies and so on.
 \item Sub-task control: operations used by tasks to submit sub-tasks
@@
   interaction \texttt{b}, except that information about the parent
-  task, which submitted a sub-task, is usually available and used for
+  of a sub-task is usually available and used for
   additional control. For instance, the parent task may wait for all
@@
   for every workflow language. It has been developed mostly for
-  well-structured and relatively simple workflow language such as between GWorkflowDL
+  well-structured and relatively simple workflow languages such as between GWorkflowDL
   and Scufl~\cite{OLAB-09}, and between the 4 languages that were involved in the
@@
 (DARE~\cite{maddineni2012distributed}), in
-Galaxy~\cite{goecks2010galaxy}, and in LONI Pipeline
+Galaxy~\cite{goecks2010galaxy}, and in the LONI Pipeline
 Environment~\cite{dinov2009efficient}. Note that tightly integrated
@@
   invocation).  Step \texttt{3} corresponds to interaction \texttt{d}
-  on Figure~\ref{fig:architectures}(II), step \texttt{2} corresponds to
-  interaction \texttt{c$_1$}, and step \texttt{4} maps to interactions
-  \texttt{b} and \texttt{c$_2$} (in VIP, workflow data transfers are
+  in Figure~\ref{fig:architectures}(II), step \texttt{2} corresponds to
+  interaction \texttt{c$_1$}, step \texttt{4} maps to interaction
+  \texttt{b}, and steps 7 and 9 map to \texttt{c$_2$} (in VIP, workflow data transfers are
   performed by the tasks and embedded in their descriptions).}
@@
 global vision of the workflow, it can keep track of the sub-tasks
-submitted by a given task, for instance to be able to cancel them when
+submitted by a given task, for instance to cancel them when
 the task is canceled. The science gateway may also implement
@@
 used to integrate the FSL toolkit~\cite{Jenkinson2012782} and the PSOM
-workflow engine~\cite{bellec2012pipeline}.
-
-The CBRAIN-FSL integration allows to leverage FSL pipelines written in
-low-level workflow languages (Linux executables and scripts) that
-submit tasks uniformly through a specific tool called fsl\_sub. It is
-diagrammed in Figure~\ref{fig:cbrain-fsl-architecture}, where the
-science gateway is represented by components \texttt{CBRAIN portal}
-and \texttt{CBRAIN execution server} and the infrastructure consists
-of \texttt{Storage servers} and \texttt{Computing cluster with shared
-  file system}.
-
-The CBRAIN-PSOM integration~\cite{GLAT-16} is shown on
-Figure~\ref{fig:cbrain-psom-architecture}. The PSOM workflow engine
+workflow engine~\cite{bellec2012pipeline}. The CBRAIN-FSL integration
+allows to leverage FSL pipelines written in low-level workflow
+languages (Linux executables and scripts) that submit tasks uniformly
+through a specific tool called fsl\_sub. It is diagrammed in
+Figure~\subref*{fig:cbrain-fsl-architecture}, where the science gateway is
+represented by components \texttt{CBRAIN portal} and \texttt{CBRAIN
+  execution server} and the infrastructure consists of \texttt{Storage
+  servers} and \texttt{Computing cluster with shared file system}.
+
+The CBRAIN-PSOM integration~\cite{GLAT-16} is shown in
+Figure~\subref*{fig:cbrain-psom-architecture}. The PSOM workflow engine
 adopts a pilot-job architecture~\cite{turilli2015comprehensive} where
@@
 } \hfill \subfloat[CBRAIN-PSOM integration. Figure
-reproduced from~\cite{GLAT-16}]{ \def\svgwidth{\columnwidth}
+reproduced from~\cite{GLAT-16}].{ \def\svgwidth{\columnwidth}
   \input{figures/CBRAIN-PSOM.pdf_tex}
@@
   through task scheduler. FSL sub-tasks will run locally
-  instead of being submitted to CBRAIN through interaction 6. 8':
+  instead of being submitted again to CBRAIN through interaction 6. 8':
   (PSOM only) PSOM master and workers execute workflow. 9. Execution
@@
 See Figure~\subref*{archi:agent}. Workflows are submitted by the science
-gateway to a pool through interaction \texttt{d}. Agents connect to
+gateway to a central pool through interaction \texttt{d}. Agents connect to
 the pool asynchronously to retrieve and execute workflows through
@@
   clients. Interaction \texttt{d} of Figure~\ref{fig:architectures}(IV) is
-  implemented in 3 different types of calls to the pool: workflow
+  implemented by 3 different calls to the pool: workflow
   submission (\texttt{1} \& \texttt{2}), workflow status retrieval
@@
 (Figure~\subref*{archi:nested}-Right) as this is the most used
-architecture. In such an instantiation, we assume that the parent and
+architecture. In the instantiation, we assume that the parent and
 child workflow engines are distinct pieces of software that require
 different workflow services invoked by distinct \texttt{d}
-interactions. If this is not the case, then workflow services can be
+interactions. If this is not the case then workflow services can be
 collapsed in a single one with a \texttt{d} interaction with
@@
   in the Science Gateway (step \texttt{1}, interaction \texttt{d} on
-  Figure~\ref{fig:architectures}(V)-Right. Ten different child engines can be used by nested
+  Figure~\ref{fig:architectures}(V)-Right). Ten different child engines can be used by nested
   workflows, invoked through the \texttt{Submission service} (step
@@
   infrastructure (\texttt{DCI}, step \texttt{3}, interactions
-  \texttt{b} and \texttt{c$_2$}). Data interactions (\texttt{c}) and
+  \texttt{b} and \texttt{c$_2$}). User data interactions (\texttt{c$_1$}) and
   application porting ones (\texttt{a}) are not represented. Figure
@@
 format conversion from a native format to the science gateway
-format. Workflow import is usually an offline process which is not
-involved in workflow execution. Workflow import was implemented in the
-SHIWA Simulation Platform through the IWIR language that provided a
-common language for portability across grid workflow
-systems~\cite{plankensteiner-prodan-etal:2013}.  Other systems rely on
-workflow import. Tavaxy~\cite{Abouelhoda2012} allows to import, merge
-and run Taverna~\cite{oinn2004taverna} and
-Galaxy~\cite{goecks2010galaxy} workflows. When some workflow parts
+format. Workflow import is usually an offline process that is not
+involved in workflow execution. A few systems have implemented
+workflow import. In the SHIWA Simulation Platform, it is implemented
+through the IWIR language that provides a common language for
+portability across grid workflow
+systems~\cite{plankensteiner-prodan-etal:2013}. Tavaxy~\cite{Abouelhoda2012}
+allows to import, merge and run Taverna~\cite{oinn2004taverna} and
+Galaxy~\cite{goecks2010galaxy} workflows; when some workflow parts
 cannot be imported, workflows are run by Tavaxy as nested workflows
@@
 evaluated in Table~\ref{table:evaluation}. We use 5 main criteria to
-evaluate the architectures: integration challenge, execution complexity,
+evaluate the architectures: integration challenge, robustness,
 extensibility, scalability and other specific features. Criteria break
@@
 
-\caption{Architecture evaluation. Lower values (brighter colors) indicate better performance. On each row, metric values are
-  normalized between 0 (best value) and 1 (worst value) to determine the
-  color of the corresponding table cell. The normalized metric value $m'$ is
-  defined as
-  $\frac{m-m_{\mathrm{min}}}{m_{\mathrm{min}}-m_{\mathrm{max}}}$ where
-  $m$ is the initial metric value, $m_{\mathrm{min}}$ and $m_{\mathrm{max}}$
-  are the minimal and maximal values among all architectures. The RGB hexadecimal color code of the cell
+\caption{Architecture evaluation. Lower values (brighter colors) indicate better performance. Cell color is set as follows: (1) on each row, metric values are
+  normalized between 0 (best value) and 1 (worst value):
+  $m'=\frac{m-m_{\mathrm{min}}}{m_{\mathrm{max}}-m_{\mathrm{min}}}$ where
+  $m$ is the metric value, $m_{\mathrm{min}}$ and $m_{\mathrm{max}}$
+  are the minimal and maximal values among all architectures; (2) the RGB hexadecimal color code of the cell
   is \texttt{\#99XX99}, where X=\texttt{round}(F-6m') (\texttt{round} rounds a number to the nearest integer). \closednote{Tristan}{Use 6. Round to the nearest integer.}{Tristan}{Fixed.}}
@@
 \emph{Extensibility} measures the difficulty to replace or add
-elements in the architecture. It is also a consequence of global
-complexity and it is determined as the number of interactions and
+elements in the architecture. It is determined as the number of interactions and
 components that need to be modified when a new element is
@@
 cope with high workloads. It is measured by counting the potential
-scalability issues in the architectures, i.e. the lack of scalability
+scalability issues in the architectures, i.e. the missing scalability
 features. Features are evaluated using a 3-level metric: 0 means that
@@
   the architecture, \texttt{S$_2$}=1 when they require specific
-  developments in the workflow engine..  \closednote{Rafael}{$S_2$
+  developments in the workflow engine.  \closednote{Rafael}{$S_2$
     would be 2 if the paradigm does not allow to distribute the
@@
 component (\texttt{E$_2$}=0). Updating the engine version in a
-workflow service does not count since the only goal of this component
+workflow service does not count in \texttt{E$_2$} since the only goal of this component
 is to wrap the engine.  New workflows are added in the science gateway
@@
 on the infrastructure (\texttt{E$_1$}=2). Updating an engine version
-in this architechture shares the same mechanism as version updates of
+in the architecture shares the same mechanism as version updates of
 other tasks, which requires an update on the
@@
 \texttt{b} and \texttt{c} in the science gateway (\texttt{E$_4$}=3).
-
-\paragraph{Specific features} Meta-workflows are not available
-(\texttt{O$_1$}=1).  Obtaining fine-grained information about workflow
-tasks is not straightforward since the science gateway has no
-knowledge about the workflow topology, and the workflow engine is
-integrated as a task (\texttt{O$_2$}=1).
 
@@
 walltime, are more difficult to estimate for the workflow task than
-for sub-tasks which may generate issues such as selection of wrong
+for the sub-tasks, which may generate issues such as selection of wrong
 batch queues on clusters. Finally, the interdependencies between the
@@
 contention. For instance, if only 1 computing resource is
-available for the science gateway and that the workflow task is
+available for the science gateway and if the workflow task is
 running on it and submits sub-tasks, then the sub-tasks could only
@@
 which makes such deadlocks unlikely to happen.
+
+\paragraph{Specific features} Meta-workflows are not available
+(\texttt{O$_1$}=1).  Obtaining fine-grained information about workflow
+tasks is not straightforward since the science gateway has no
+knowledge about the workflow topology, and the workflow engine is
+integrated as a task (\texttt{O$_2$}=1).
 
@@
 workflow execution (\texttt{R$_1$=2}). Interactions \texttt{b}
-(counted twice), \texttt{c$_2$} (twice) and \texttt{d} (twice) also
+(twice), \texttt{c$_2$} (twice) and \texttt{d} (twice) also
 are (\texttt{R$_2$=6}).
@@
 version in the parent or child engine only requires modifying this
-engine (\texttt{E$_2$=0}). Adding a new
-workflow is done through interaction \texttt{a$_1$} and \texttt{a$_2$}
+engine (\texttt{E$_2$=0}). Adding a new workflow is done through
+interaction \texttt{a$_1$} and \texttt{a$_2$}
 (\texttt{E$_3$=2}). Adding a new infrastructure requires to
-re-implement interactions \texttt{a$_1$}, \texttt{b} and \texttt{c$_2$} twice, and
-interaction \texttt{c$_1$} once (\texttt{E$_4$=6}).
+re-implement interactions \texttt{a$_1$}, \texttt{b} and
+\texttt{c$_2$} twice, and interaction \texttt{c$_1$} once so that it
+can be supported by both workflow engines (\texttt{E$_4$=6}).
 
@@
 substantial work depending on the complexity of the language used by
-the new engine. Based on the same logic, adding a
+the new engine. Similarly, adding a
 new engine version only requires modifying interaction \texttt{g}
@@
 service architecture, interfacing with a new infrastructure requires
-modifications in the workflow service and in interactions \texttt{a$_1$}, \texttt{b}
+modifications in interactions \texttt{a$_1$}, \texttt{b}, \texttt{c$_1$}
 and \texttt{c$_2$} (\texttt{E$_4$=4}).
@@
 
-Tight integration and task encapsulation are globally the simplest
-architectures, followed by service integration, workflow import (with
+Tight integration and task encapsulation are the simplest
+architectures to integrate, followed by service integration, workflow import (with
 service invocation) and pool. Nested workflows (with service
@@
 (with service invocation) at the end. This ordering is consistent
-across metrics, it reflects the global complexity of the
+across metrics; it reflects the global complexity of the
 architectures.
@@
 integration is the least scalable and all the other architectures
-perform the same. The overall scalability score, however, should not
+perform the same overall. The global scalability score, however, should not
 mask the unique characteristics of architectures regarding this
@@
 task encapsulation and nested workflows could well become
-showstoppers depending on the type of the targeted
+showstoppers depending on the type of targeted
 infrastructure. Non-reliable infrastructures, for example, could
@@
                                     & \textbf{Import} \\
-  Integration (normalized total)    &
+  Integration (global score)    &
                                     \cellcolor[HTML]{99FF99}{0.00}
@@
                                       & \cellcolor[HTML]{99DD99}{0.33}\\
-Robustness (normalized total) &
+Robustness (global score) &
                                 \cellcolor[HTML]{99F099}{0.14}
@@
                                     & \cellcolor[HTML]{99D399}{0.43}\\
-  Extensibility (normalized total)  & \cellcolor[HTML]{99D299}{0.44}
+  Extensibility (global score)  & \cellcolor[HTML]{99D299}{0.44}
                                      & \cellcolor[HTML]{99D299}{0.44}
@@
                                      & \cellcolor[HTML]{99E899}{0.22}\\
-Scalability (normalized total)  & \cellcolor[HTML]{999999}{1.00}
+Scalability (global score)  & \cellcolor[HTML]{999999}{1.00}
                                      & \cellcolor[HTML]{99CC99}{0.50}
@@
                                      & \cellcolor[HTML]{99CC99}{0.50}\\
-Specific features (normalized total) & \cellcolor[HTML]{99CC99}{0.50}
+Specific features (global score) & \cellcolor[HTML]{99CC99}{0.50}
                                      & \cellcolor[HTML]{99CC99}{0.50}
@@
   are obtained by summing the normalized global scores ($m'$ values) of
-  each criterion in Table~\ref{table:evaluation} and the colors are obtained by normalizing these scores
-  as done in Table~\ref{table:evaluation}. }
+  each criterion in Table~\ref{table:evaluation} and the colors are obtained as in Table~\ref{table:evaluation}. }
 \label{table:overall}
@@
 implementation in specific systems is useful to compare them, but we
-also realize that this comes at the cost of realism. In particular,
-the presented architectures are theoritical patterns that may be blended in
-actual systems. The distinction between tight integration and service
-invocation, for instance, may not be always that clear in practice. Service
-invocation may also be combined with task encapsulation in some cases. However, the
-criteria discussed in this work would still apply to broadly compare and categorize
-such cases of hydrid architectures.
+also realize that it comes at the cost of realism. In particular, the
+presented architectures are theoretical patterns that may be blended
+in actual systems. The distinction between tight integration and
+service invocation, for instance, may not always be that clear in
+practice. Service invocation may also be combined with task
+encapsulation in some cases. However, the criteria discussed in this
+work would still apply to broadly compare and categorize such cases of
+hydrid architectures.
 
@@
 the same framework as the other architectures. Other types of
-instantiation, for instance nested workflows with task encapsulation, could
-also be envisaged. We chose to limit ourselves to instantiations with
-service invocation because the resulting architectures are implemented
-in real systems, and because service invocation is largely
-used. Nested workflows with task encapsulation, in particular, could be an
-interesting architecture to explore.
+instantiation, for instance nested workflows with task encapsulation,
+could also be envisaged. We chose to limit ourselves to instantiations
+with service invocation because the resulting architectures are
+implemented in real systems, and because service invocation is largely
+used. Nevertheless, it could be interesting to explore other types of
+instantiations, in particular nested workflows with task
+encapsulation.
 
@@
 particular system. It provides general insight to integrate science
-gateways and workflow engines.
+gateways and workflow engines, which we hope will be useful for
+software architects.
 

### 2016-05-02 paper.tex

@@
 refers to a science gateway user with extended privileges to install
-applications.
+applications. \note{Naj}{I propose a table of definitions.}
 

### 2016-05-02 paper.tex
A few fixes following @najmahan's comments in https://github.com/glatard/FGCS-works-special-issue/pull/15/files

@@
 
-\title{Software Architectures for Integrating Workflow Engines in Science Gateways} \note{Naj}{'what' software architectures? what is their qualifier, adjective?}
+\title{An analysis of existing software architectures\\ to integrate workflow engines in science gateways\closednote{Naj}{'what' software architectures?
+what is their qualifier, adjective?}{Tristan}{There is no qualifier except 'for integrating workflow engines in science gatewats'.
+They are just software architectures :) The point of the paper is to describe them properly.
+It's a general formulation to cover any kind of software architecture that is used to integrated workflow engines in science gateways.
+Does it make sense?}}
 
@@
   We investigate six software architectures commonly used to integrate
-  workflow engines in science gateways. We describe and evaluate these software
+  workflow engines in science gateways. We describe and evaluate these
   architectures with metrics that measure integration complexity,
-  robustness, extensibility, scalability and other specific
-  features\note{Naj}{Please be precise, what are they?}. Results illustrate the pros and cons of different
-  architectures: Simple architectures with tight integration and task encapsulation are the
-  the easiest to integrate and are most robust.Extensibility is equivalent in most architectures.
-The pool model is the most scalable one and meta-workflows are only available in nested workflows and workflow
+  robustness, extensibility, scalability and support for
+  meta-workflows and fine-grained debugging\closednote{Naj}{Please be
+    precise, what are they?}{Tristan}{Done, thanks}. Results
+  illustrate the pros and cons of different architectures: Tight
+  integration and task encapsulation are the easiest to integrate and
+  are most robust. Extensibility is equivalent in most
+  architectures. The pool model is the most scalable one and
+  meta-workflows are only available in nested workflows and workflow
   import. Our analysis provides insights for science gateway
-  architects and workflow engine developers. \note{Naj}{I find this abstract to be vague, it can be made more concise and poignant}
+  architects and workflow engine developers. \note{Naj}{I find this
+    abstract to be vague, it can be made more concise and poignant}
 \end{abstract}
@@
 Workflow engines are critical for the efficient and transparent
-exploitation of distributed infrastructures in the ecosystem of tools and
-services offered by science gateways . Several
-software architectures can be adopted to integrate workflow engines in
-science gateways, however 
-development effort required and the resulting system vary across different architectures.
-This paper and our analytical approach to comparison of software architectures is informed by our experince in the development and
-sustained operation of the CBRAIN~\cite{SHER-14} and
-VIP~\cite{GLAT-13} science gateways for medical image analysis during
-the past 7 years; as well as by lessons learned from several
-science gateway and workflow projects, including
-SHIWA\footnote{\url{http://www.shiwa-workflow.eu}} and related
-initiatives. This paper describes,
-provides examples and compares such architectures, based on
-system-independent representations of their main components and
-interactions. In this section we provide background information and
-definitions of different workflow engines, science gateways and
-infrastructures. In Section~\ref{sec:architectures}, we describe six
-architectures within a consistent framework that underlines the functional
-interactions between their main software components. In
-Section~\ref{sec:evaluation}, different architectures are evaluated with
-metrics measuring integration challenge, robustness, extensibility,
-scalability and other specific features. We compare different architectures and highlight
-recommendations for science gateway architects and workflow engine developers.
+exploitation of distributed infrastructures in the ecosystem of tools
+and services offered by science gateways. Several software
+architectures can be adopted to integrate workflow engines in science
+gateways, however development effort required and the resulting system
+vary across different architectures.
+
+This paper describes, provides examples and compares such
+architectures, based on system-independent representations of their
+main components and interactions. It is informed by our experince in
+the development and sustained operation of the CBRAIN~\cite{SHER-14}
+and VIP~\cite{GLAT-13} science gateways for medical image analysis
+during the past 7 years; as well as by lessons learned from several
+science gateway and workflow projects such as
+SHIWA\footnote{\url{http://www.shiwa-workflow.eu}}.  In this section
+we provide background information and definitions of different
+workflow engines, science gateways and infrastructures. In
+Section~\ref{sec:architectures}, we describe six architectures within
+a consistent framework that underlines the functional interactions
+between their main software components. In
+Section~\ref{sec:evaluation}, the different architectures are
+evaluated with metrics measuring integration challenge, robustness,
+extensibility, scalability and other specific features. We finally
+compare the architectures and highlight recommendations for science
+gateway architects and workflow engine developers.
 
@@
 Interfaces~\cite{gorgolewski2011nipype}), PSOM (Pipeline System for
-Octave and Matlab~\cite{bellec2012pipeline}), SPM (Statistical
-Parametric Mapping~\cite{ashburner2011spm}) and FSL (FMRIB Software
+Octave and Matlab~\cite{bellec2012pipeline}), RPPL~\cite{1174106} and PMP
+(Poor Man's Pipeline~\cite{Ad-DabbaghY2006}), SPM (Statistical Parametric
+Mapping~\cite{ashburner2011spm}) and FSL (FMRIB Software
 Library~\cite{Jenkinson2012782}) provide abstractions and functions to
@@
 variety of programming languages. Such tools were interfaced to
-computing infrastructures, in particular clusters\closednote{Tristan}{Reverted from Marc's 'cluster scheduler'}{}{}, to execute tasks at
-a high throughput. Some of these tools also support
-advanced features such as provenance tracking (Nipype, PSOM), \note{Naj}{should we not mention PMP which is used for CIVET?}
+computing infrastructures, in particular
+clusters\closednote{Tristan}{Reverted from Marc's 'cluster
+  scheduler'}{}{}, to execute tasks at a high throughput. Some of
+these tools also support advanced features such as provenance tracking
+(Nipype, PSOM), \closednote{Naj}{should we not mention PMP which is
+  used for CIVET?}{Tristan}{Done, as discussed.}
 redundancy detection across analyses to avoid re-computation (PSOM),
@@
 
-In this paper, the working definition of a
-\emph{workflow engine} (also abbreviated \emph{engine}) is "a piece of
+In this paper, a
+\emph{workflow engine} (also abbreviated \emph{engine}) is a piece of
 software that submits interdependent computing \emph{tasks} to an
@@
 that may consist of files, database entries or simple parameter
-values." Although simplistic, this definition covers both e-Science workflow systems and
+values. Although simplistic, this definition covers both e-Science workflow systems and
 domain-specific pipelines.\closednote{Marc-e}{To remain consistent with above descriptions}{Tristan}{yes!} Some workflow engines, usually e-Science
@@
 locations. Workflow engines are part of this ecosystem as core
-components to implement and execute applications. \note{provide examples?}
+components to implement and execute
+applications. \closednote{Naj}{provide examples?}{Tristan}{Examples
+  are mentioned in the last paragraph of this section. Moved it next.}
+
+Various science gateways have been developed, including complete
+frameworks such as Apache Airavata~\cite{marru2011apache}, the Catania
+Science Gateway Framework~\cite{Ardizzone2012} and
+WS-PGRADE/gUSE~\cite{Kacsuk2012}. Numerous science gateways were built
+using such frameworks~\cite{kacsuk2014science,ardizzone2012decide} or
+as standalone systems~\cite{SHER-14,GLAT-13}. Most of these science
+gateway frameworks and instances include one or several workflow
+engines.
 
@@
 
-Various science gateways have been developed, including frameworks
-used as building blocks to assemble other systems, for instance Apache
-Airavata~\cite{marru2011apache}, the Catania Science Gateway
-Framework~\cite{Ardizzone2012} and
-WS-PGRADE/gUSE~\cite{Kacsuk2012}. Numerous science gateways were built
-using such frameworks~\cite{kacsuk2014science,ardizzone2012decide},
-and other ones, e.g. CBRAIN~\cite{SHER-14} and VIP~\cite{GLAT-13},
-were built without any framework.
-
 \subsection{Infrastructure}
@@
 diagrammed in Figure~\ref{fig:architectures} using the graphical
-notations shown in Figure~\ref{fig:notations}.  \closednote{Marc-e}{A
+notations shown in Figure~\ref{fig:notations}. This section describes the different types of
+interactions involved in the architectures. Architectures are then
+detailed and illustrated using examples of real
+systems. Table~\ref{table:system-classification} summarizes the
+classification of a few systems by architecture.
+
+\closednote{Marc-e}{A
   few things bother me a bit with the notation fig., making it harder
@@
   in this journal?)  }{Tristan}{Marc-E, is Figure 1 better now?}
-
-Architectures are described from their main software \emph{components}
-and \emph{interactions}. Software components include science gateway
-and infrastructure in all architectures while workflow service,
-workflow pool and agent are involved in some architectures only. The
-workflow engine itself is represented by a specific symbol enclosed in a dotted rectangle.
-
-Software interactions among these components are represented with
-labeled, plain black arrows.  \emph{Abstract} interactions are
-specific types of interactions that may be implemented using different
-components and interactions. \note{Naj}{three 'interactions' in one sentence} They are represented with dotted arrows
-and labeled with \texttt{*}. An architecture that has an abstract
-interaction is an abstract architecture that can be instantiated in
-various ways.
-
-Red color (WHERE?) is used to represent the components and interactions that
-are specific to workflow execution, i.e. that are not involved in any
-other process. Tasks refer to computing tasks executed on the
-infrastructure. Data represents any type of file or database entry involved
-in workflow execution and stored on the infrastructure. It does not
-cover workflow parameters such as strings, numbers, etc. Administrator
-refers to a science gateway user with extended privileges to install
-applications. \note{Naj}{I propose a table of definitions.}
-
-The remainder of this section describes the different types of
-interactions involved in the architectures. Architectures are then
-detailed and illustrated using examples of real
-systems. Table~\ref{table:system-classification} summarizes the
-classification of a few systems by architecture.
+\closednote{Naj}{I propose a table of definitions.}{Tristan}{Added
+  information in Figure 1 and removed the corresponding text.}
 
@@
 \centering
-\def\svgwidth{0.7\columnwidth}
+\def\svgwidth{0.8\columnwidth}
 \input{figures/notations.pdf_tex}
@@
   Service & \pbox{1.5\columnwidth}{
-            Apache Airvata~\cite{marru2011apache}, e-BioInfra~\cite{shahand2015data}$^{w,n}$, HubZero with Pegasus~\cite{CPE:CPE3257}, MoSGrid~\cite{krüger2014mosgrid}$^w$, System in~\cite{wu2010accelerating}, Vine Toolkit~\cite{DBLP:journals/scpe/SzejnfeldDKKKKLPTWDNW10}, Virtual Imaging Platform~\cite{GLAT-13}$^n$, WS-PGRADE/gUSE framework~\cite{Kacsuk2012},
+            Apache Airvata~\cite{marru2011apache}, e-BioInfra~\cite{shahand2015data}$^{w,n}$, HubZero with Pegasus~\cite{CPE:CPE3257}, MoSGrid~\cite{kruger2014mosgrid}$^w$, System in~\cite{wu2010accelerating}, Vine Toolkit~\cite{DBLP:journals/scpe/SzejnfeldDKKKKLPTWDNW10}, Virtual Imaging Platform~\cite{GLAT-13}$^n$, WS-PGRADE/gUSE framework~\cite{Kacsuk2012},
             Science gateways in~\cite{kacsuk2014science}$^w$
@@
 } \hfill \subfloat[CBRAIN-PSOM integration. Figure
-reproduced from~\cite{GLAT-16}].{ \def\svgwidth{\columnwidth}
-  \input{figures/CBRAIN-PSOM.pdf_tex}
+reproduced from~\cite{GLAT-16}.]{\def\svgwidth{\columnwidth}
+  \input{figures/CBRAIN-PSOM.pdf_tex
+}
 \label{fig:cbrain-psom-architecture}
@@
 See Figure~\subref*{archi:nested}. In nested workflows
-(Figure~\subref*{archi:nested}-Left), a task of a \emph{parent}
-workflow executed by the \emph{parent} engine is itself a workflow,
-called \emph{child} workflow, that is executed by a \emph{child}
-engine. \note{Naj} {this reads funny} The parent and child engines might use different workflow
-description languages. They might also execute workflows on different
-infrastructures. The parent workflow is also called meta-workflow. The
-science gateway communicates with the parent engine through abstract
-interaction \texttt{*$_1$}. The science gateway also communicates with
-the infrastructure to transfer user data through abstract interactions
-\texttt{*$_2$} and \texttt{*$_3$}. Both workflow engines communicate
-with the infrastructure through abstract interactions \texttt{*$_4$}
-and \texttt{*$_5$}. The parent engine communicates with the child
-engine through abstract interaction \texttt{*$_6$}. Administrator
-installs workflows through interactions \texttt{a$_1$} and
-\texttt{a$_2$}.
+(Figure~\subref*{archi:nested}-Left), workflows include other
+workflows that are executed by a \emph{child} engine. However, parent
+and child engines might use different languages and might run on
+different infrastructures. A parent workflow is
+also called meta-workflow. The science gateway communicates with the
+parent engine through abstract interaction \texttt{*$_1$}. The science
+gateway also communicates with the infrastructure to transfer user
+data through abstract interactions \texttt{*$_2$} and
+\texttt{*$_3$}. Both workflow engines communicate with the
+infrastructure through abstract interactions \texttt{*$_4$} and
+\texttt{*$_5$}. The parent engine communicates with the child engine
+through abstract interaction \texttt{*$_6$}. Administrator installs
+workflows through interactions \texttt{a$_1$} and \texttt{a$_2$}.
 
@@
 difficulty to extend the architecture.
-\note{Naj}{I would rather compare each metric separately, this woudl also give a sense of how different architectures compare across metrics} Note that some extensions may
+\closednote{Naj}{I would rather compare each metric separately, this woudl also give a sense of how different architectures compare across metrics}{Tristan}{Metrics are compared separately too, see Table 2 and discussion.} Note that some extensions may
 involve several metrics in practice. For instance, adding a new type
@@
 the feature is very easy to enable, 1 means that it can be implemented
-but with some difficulty, and 2 means that the feature could only be
-implemented with a nonsensical amount of effort\note{Naj}{how do you metricize "nonesensical amount fo effors?"}. Four different
-features are identified:
+but with some difficulty, and 2 means that the feature cannot
+realistically be implemented. \closednote{Naj}{how do you metricize
+  "nonesensical amount fo effors?"}{Tristan}{You don't :) Rephrased to
+  'cannot realistically be implemented'. At some point it's a matter
+  of common sense and experience to draw the line between what is
+  easily implemented and what is more tricky. I would like to avoid
+  'impossible' because nothing is impossible to implement if enough
+  effort is put.}. Four different features are identified:
 \begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
@@
   science gateway. Some architectures, however, complicate the task
-  scheduling problem by introducing additional software layers, or by
+  scheduling problem by introducing additional software layers or
   creating tasks with specific characteristics. \texttt{S$_3$}=0 when

### 2016-05-02 paper.tex
Fixed #13

@@
 Section~\ref{sec:evaluation}, the different architectures are
-evaluated with metrics measuring integration challenge, robustness,
+evaluated with metrics measuring integration complexity, robustness,
 extensibility, scalability and other specific features. We finally
@@
 evaluated in Table~\ref{table:evaluation}. We use 5 main criteria to
-evaluate the architectures: integration challenge, robustness,
+evaluate the architectures: integration complexity, robustness,
 extensibility, scalability and other specific features. Criteria break
@@
                                      & \textbf{Import} \\
-\cellcolor[HTML]{EEEEEE}\textbf{Integration challenge}& \multicolumn{6}{l}{\cellcolor[HTML]{EEEEEE}}\\
+\cellcolor[HTML]{EEEEEE}\textbf{Integration complexity}& \multicolumn{6}{l}{\cellcolor[HTML]{EEEEEE}}\\
   Total components -- \texttt{I$_1$} & \cellcolor[HTML]{99FF99}2
@@
 
-\emph{Integration challenge} is obtained by counting the total number of
+\emph{Integration complexity} is obtained by counting the total number of
 interactions and components on the architecture diagram. It breaks
@@
 
-\paragraph{Integration challenge} This architecture does not require any
+\paragraph{Integration complexity} This architecture does not require any
 component in addition to the science gateway and infrastructure
@@
 
-\paragraph{Integration challenge} Service invocation requires a workflow service
+\paragraph{Integration complexity} Service invocation requires a workflow service
 in addition to the science gateway and infrastructure
@@
 
-\paragraph{Integration challenge} Task encapsulation requires only 2
+\paragraph{Integration complexity} Task encapsulation requires only 2
 components (\texttt{I$_1$}=2).  It involves 5 interactions:
@@
 
-\paragraph{Integration challenge} The pool model requires a workflow pool and an
+\paragraph{Integration complexity} The pool model requires a workflow pool and an
 agent in addition to the science gateway and infrastructure
@@
 
-\paragraph{Integration challenge} Setting up a nested workflow
+\paragraph{Integration complexity} Setting up a nested workflow
 architecture with service invocation requires a science gateway, 2
@@
 
-\paragraph{Integration challenge} Workflow import with service invocation
+\paragraph{Integration complexity} Workflow import with service invocation
 requires the same components as for service invocation

### 2016-05-03 paper.tex
Added acknowledgments to Brain Canada PSG as well as CCNA.
Updates in abstract (#17)

@@
 
-\title{An analysis of existing software architectures\\ to integrate workflow engines in science gateways\closednote{Naj}{'what' software architectures?
+\title{A comprehensive analysis of existing software architectures\\ to integrate workflow engines in science gateways\closednote{Naj}{'what' software architectures?
 what is their qualifier, adjective?}{Tristan}{There is no qualifier except 'for integrating workflow engines in science gatewats'.
@@
   We investigate six software architectures commonly used to integrate
-  workflow engines in science gateways. We describe and evaluate these
-  architectures with metrics that measure integration complexity,
-  robustness, extensibility, scalability and support for
-  meta-workflows and fine-grained debugging\closednote{Naj}{Please be
-    precise, what are they?}{Tristan}{Done, thanks}. Results
-  illustrate the pros and cons of different architectures: Tight
+  workflow engines in science gateways. (1) In \emph{tight
+    integration}, the workflow engine shares software components with
+  the science gateway; (2) In \emph{service invocation}, the engine is
+  isolated and invoked through a specific software interface; (3) In
+  \emph{task encapsulation}, the engine is wrapped as a computing task
+  executed on the infrastructure; (4) In the \emph{pool model}, the
+  engine is bundled in an agent that connects to a central pool to
+  fetch and execute workflows; (5) In \emph{nested workflows}, the
+  engine is integrated as a child process of another engine; (6) In
+  \emph{workflow import}, the engine is integrated through workflow
+  language conversion. We describe and evaluate these architectures
+  with metrics that measure integration complexity, robustness,
+  extensibility, scalability and support for meta-workflows and
+  fine-grained debugging\closednote{Naj}{Please be precise, what are
+    they?}{Tristan}{Done, thanks}. Results provide insights for
+  science gateway architects and workflow engine developers. Tight
   integration and task encapsulation are the easiest to integrate and
@@
   meta-workflows are only available in nested workflows and workflow
-  import. Our analysis provides insights for science gateway
-  architects and workflow engine developers. \note{Naj}{I find this
-    abstract to be vague, it can be made more concise and poignant}
+  import. \note{Naj}{I find this abstract to be vague, it can be made
+    more concise and poignant}
 \end{abstract}
@@
 (\url{https://computecanada.ca}) and platform development program from CANARIE
- (\url{http://www.canarie.ca/}). This work has been made possible with the support of the Irving Ludmer Family Foundation and
+ (\url{http://www.canarie.ca/}). The integration between PSOM and CBRAIN was supported by a Brain 
+Canada Platform Support Grant, as well as the Canadian Consortium on
+Neurodegeneration in Aging (CCNA), through a grant from the Canadian Institute
+of Health Research and funding from several partners. This work has been made possible with the support of the Irving Ludmer Family Foundation and
 the Ludmer Centre for Neuroinformatics and Mental Health. \closedanswerednote{Marc-e}{Would it be proper to thank EGI?

### 2016-05-06 paper.tex
Reverted title to original one
A few fixes in the abstract and introduction

@@
 
-\title{A comprehensive analysis of existing software architectures\\ to integrate workflow engines in science gateways\closednote{Naj}{'what' software architectures?
+\title{Software architectures to integrate workflow engines in science gateways\closednote{Naj}{'what' software architectures?
 what is their qualifier, adjective?}{Tristan}{There is no qualifier except 'for integrating workflow engines in science gatewats'.
@@
   We investigate six software architectures commonly used to integrate
-  workflow engines in science gateways. (1) In \emph{tight
-    integration}, the workflow engine shares software components with
-  the science gateway; (2) In \emph{service invocation}, the engine is
-  isolated and invoked through a specific software interface; (3) In
-  \emph{task encapsulation}, the engine is wrapped as a computing task
-  executed on the infrastructure; (4) In the \emph{pool model}, the
-  engine is bundled in an agent that connects to a central pool to
-  fetch and execute workflows; (5) In \emph{nested workflows}, the
-  engine is integrated as a child process of another engine; (6) In
-  \emph{workflow import}, the engine is integrated through workflow
-  language conversion. We describe and evaluate these architectures
-  with metrics that measure integration complexity, robustness,
+  workflow engines in science gateways. In \emph{tight integration},
+  the workflow engine shares software components with the science
+  gateway. In \emph{service invocation}, the engine is isolated and
+  invoked through a specific software interface. In \emph{task
+    encapsulation}, the engine is wrapped as a computing task executed
+  on the infrastructure. In the \emph{pool model}, the engine is
+  bundled in an agent that connects to a central pool to fetch and
+  execute workflows. In \emph{nested workflows}, the engine is
+  integrated as a child process of another engine. In \emph{workflow
+    import}, the engine is integrated through workflow language
+  conversion. We describe and evaluate these architectures with
+  metrics that measure integration complexity, robustness,
   extensibility, scalability and support for meta-workflows and
@@
   meta-workflows are only available in nested workflows and workflow
-  import. \note{Naj}{I find this abstract to be vague, it can be made
-    more concise and poignant}
+  import. \closednote{Naj}{I find this abstract to be vague, it can be made
+    more concise and poignant}{Tristan}{Added a description of the
+    architectures, following @pbellec's suggestion.}
 \end{abstract}
@@
 architectures can be adopted to integrate workflow engines in science
-gateways, however development effort required and the resulting system
-vary across different architectures.
+gateways, with important consequences on the development effort
+required and resulting system.
 
@@
 SHIWA\footnote{\url{http://www.shiwa-workflow.eu}}.  In this section
-we provide background information and definitions of different
+we provide background information and definitions of 
 workflow engines, science gateways and infrastructures. In
@@
 Interfaces~\cite{gorgolewski2011nipype}), PSOM (Pipeline System for
-Octave and Matlab~\cite{bellec2012pipeline}), RPPL~\cite{1174106} and PMP
-(Poor Man's Pipeline~\cite{Ad-DabbaghY2006}), SPM (Statistical Parametric
-Mapping~\cite{ashburner2011spm}) and FSL (FMRIB Software
+Octave and Matlab~\cite{bellec2012pipeline}), RPPL~\cite{1174106} and
+PMP (Poor Man's Pipeline~\cite{Ad-DabbaghY2006}), SPM (Statistical
+Parametric Mapping~\cite{ashburner2011spm}) and FSL (FMRIB Software
 Library~\cite{Jenkinson2012782}) provide abstractions and functions to
@@
   scheduler'}{}{}, to execute tasks at a high throughput. Some of
-these tools also support advanced features such as provenance tracking
-(Nipype, PSOM), \closednote{Naj}{should we not mention PMP which is
-  used for CIVET?}{Tristan}{Done, as discussed.}
-redundancy detection across analyses to avoid re-computation (PSOM),
-and so on. A wide-array of workflows have been implemented using these
-pipelines and are now shared across neuroimaging groups
-world-wide. These domain-specific engines represent a tremendous
-opportunity for science gateways to leverage existing tools and
-applications. They nicely complement e-Science engines that are more
-oriented towards the exploitation of distributed computing
-infrastructures, in particular grids and clouds.
+these tools also support advanced features such as provenance
+tracking, \closednote{Naj}{should we not mention PMP which is used for
+  CIVET?}{Tristan}{Done, as discussed.}  redundancy detection across
+analyses to avoid re-computation, and so on. A wide-array of
+workflows have been implemented using these pipelines and are now
+shared across neuroimaging groups world-wide. These domain-specific
+engines represent a tremendous opportunity for science gateways to
+leverage existing tools and applications. They nicely complement
+e-Science engines that are more oriented towards the exploitation of
+distributed computing infrastructures, in particular grids and clouds.
 
@@
 using such frameworks~\cite{kacsuk2014science,ardizzone2012decide} or
-as standalone systems~\cite{SHER-14,GLAT-13}. Most of these science
-gateway frameworks and instances include one or several workflow
-engines.
+as standalone systems~\cite{SHER-14,GLAT-13}. Most of these frameworks
+and instances include one or several workflow engines.
 
@@
 executed by different types of engines. Some gateways may not even use any
-workflow engine and focus on the execution of simple tasks.
-
-Extensibility is an important property of the integration, as new
+workflow engine and instead focus on the execution of simple tasks.
+
+Extensibility is an important property of the integration. New
 workflows are added frequently, different types and versions of
 workflow engines may be integrated over time, and different kinds of
-infrastructure can be targeted. Scalability is also a crucial concern
-for such multi-user systems that manage substantial amounts of data
-and execute computationally-intensive analyses. To this end, science
-gateways may balance the load among different instances of the same
-engine. They may also use advanced task scheduling policies on the
-infrastructure, e.g. to ensure fairness among
-users~\cite{FERR-14,CPE:CPE3708} or to improve
-fault-tolerance~\cite{wilkins2008teragrid,FERR-13}.
-
-Robustness is a highly desirable characteristic as well. Architectures
-showing reduced system complexity will greatly facilitate the
+infrastructure can be targeted.
+
+Scalability is also a crucial concern for such multi-user,
+high-throughput systems. To this end, science gateways may balance the
+load among different instances of the same engine, and use advanced
+task scheduling policies on the infrastructure to improve performance,
+fault-tolerance and fairness among users.
+
+Robustness is highly desirable as well since it is key to a
+good user experience. Simple architectures facilitate the
 implementation effort towards robust interactions which, in turn, have
 a positive impact on characteristics such as gateway predictability,
-transparence, general reliability and results traceability and
-reproducibility; characteristics that are also key to a good user
-experience. \closednote{Marc-Etienne}{I find that something is weak in
-  this last paragraph. Robustness will help with transparency, sure,
-  but it seems only a small facet.  Also, the text seem to timidly
-  favor architectures with less components/interactions without much
-  explanation... Complex systems can be robust, but at greater
-  costs...  Here is one: "Architectures showing reduced system
+transparency, reliability, traceability and
+reproducibility. \closednote{Marc-Etienne}{I find that something is
+  weak in this last paragraph. Robustness will help with transparency,
+  sure, but it seems only a small facet.  Also, the text seem to
+  timidly favor architectures with less components/interactions
+  without much explanation... Complex systems can be robust, but at
+  greater costs...  Here is one: "Architectures showing reduced system
   complexity will greatly facilitate the implementation effort towards
@@
   experience."}{Tristan}{Agreed, fixed as
-  suggested.}\note{Tristan}{Work on the this sentence again.}
+  suggested.}\closednote{Tristan}{Work on the this sentence again.}{Tristan}{done.}
 

### 2016-05-09 paper.tex
Minor updates

@@
 control flow constructs, parallelization operators, visual edition
-features, links with domain-specific tool repositories, provenance
+tools, links with domain-specific application repositories, provenance
 recording and other features. An overview of workflow system
@@
   CIVET?}{Tristan}{Done, as discussed.}  redundancy detection across
-analyses to avoid re-computation, and so on. A wide-array of
-workflows have been implemented using these pipelines and are now
-shared across neuroimaging groups world-wide. These domain-specific
-engines represent a tremendous opportunity for science gateways to
-leverage existing tools and applications. They nicely complement
-e-Science engines that are more oriented towards the exploitation of
-distributed computing infrastructures, in particular grids and clouds.
+analyses to avoid re-computation, and so on. A wide array of workflows
+have been implemented using these pipelines and are now shared across
+neuroimaging groups world-wide, which represents a tremendous
+opportunity for science gateways to leverage. Domain-specific engines
+nicely complement e-Science systems that are more oriented towards the
+exploitation of distributed computing infrastructures, in particular
+grids and clouds.
 
@@
 values. Although simplistic, this definition covers both e-Science workflow systems and
-domain-specific pipelines.\closednote{Marc-e}{To remain consistent with above descriptions}{Tristan}{yes!} Some workflow engines, usually e-Science
+domain-specific pipeline systems.\closednote{Marc-e}{To remain consistent with above descriptions}{Tristan}{yes!} Some workflow engines, usually e-Science
 ones, may transfer data across the infrastructure, and others, usually
@@
 
-Various science gateways have been developed, including complete
-frameworks such as Apache Airavata~\cite{marru2011apache}, the Catania
-Science Gateway Framework~\cite{Ardizzone2012} and
+Various science gateways have been developed, including frameworks
+such as Apache Airavata~\cite{marru2011apache}, the Catania Science
+Gateway Framework~\cite{ardizzone2012decide} and
 WS-PGRADE/gUSE~\cite{Kacsuk2012}. Numerous science gateways were built
 using such frameworks~\cite{kacsuk2014science,ardizzone2012decide} or
-as standalone systems~\cite{SHER-14,GLAT-13}. Most of these frameworks
-and instances include one or several workflow engines.
+as standalone systems~\cite{SHER-14,GLAT-13}. Most of these systems
+include one or several workflow engines.
 
@@
 engine while other ones are more general and host applications
-executed by different types of engines. Some gateways may not even use any
-workflow engine and instead focus on the execution of simple tasks.
+executed by different types of engines. 
+%Some gateways may not even use any
+%workflow engine and instead focus on the execution of simple tasks.
 
@@
 
-Specific features may also be available in science gateways, such as
+Specific features may also be available, for instance
 data vizualisation and quality control, workflow edition, debugging
@@
 
-In this paper, infrastructure consists of the computing and storage
+Infrastructure consists of the computing and storage
 resources involved in workflow execution, as well as the software
@@
 composed of computing or file servers, databases, clusters, grids or
-clouds. Some workflow engines and science gateways may assume specific
+clouds. Some workflow engines and science gateways may require specific
 characteristics, such as the presence of a shared file system between
@@
   Tight & \pbox{1.5\columnwidth}{
-          Catania Science Gateway Framework~\cite{Ardizzone2012},
+          Catania Science Gateway Framework~\cite{ardizzone2012decide},
           Distributed application runtime environment (DARE~\cite{maddineni2012distributed}),
@@
   for instance a web form, where users can enter the parameters of the
-  workflow to be executed. The interaction has two aspects:
+  workflow to be executed. The interaction has two steps:
   (\texttt{a$_1$}) the programs used in the workflow are installed on
@@
   and Scufl~\cite{OLAB-09}, and between the 4 languages that were involved in the
-  SHIWA initiative. In SHIWA, workflow conversion is done through an
-  intermediate representation, the Interoperable Workflow Intermediate
-  Representation~\cite{plankensteiner-montagnat-etal:2011}, which
-  allows to convert among $n$ workflow languages using $2n$
-  interactions instead of $n^2$.
+  SHIWA initiative.
 \end{enumerate}
@@
 security reasons. This is the model adopted in the Catania Science
-Gateway Framework~\cite{Ardizzone2012} (see specific documentation on
+Gateway Framework~\cite{ardizzone2012decide} (see specific documentation on
 workflows\footnote{\url{http://bit.ly/1oQrzvQ}}), in the Distributed
@@
 The pool model was implemented in the SHIWA pool~\cite{ROGE-13}
-diagrammed in Figure~\ref{fig:shiwa-pool-architecture}. In the SHIWA
-pool, agents can wrap different types of workflow engines to execute
-workflows expressed with different languages.
+diagrammed in Figure~\ref{fig:shiwa-pool-architecture}, where agents
+can wrap different types of workflow engines to execute workflows
+expressed with different languages.
 
@@
 portability across grid workflow
-systems~\cite{plankensteiner-prodan-etal:2013}. Tavaxy~\cite{Abouelhoda2012}
+systems~\cite{plankensteiner-prodan-etal:2013} and allows to convert
+among $n$ workflow languages using $2n$ interactions instead of $n^2$.
+Tavaxy~\cite{Abouelhoda2012}
 allows to import, merge and run Taverna~\cite{oinn2004taverna} and
@@
 The architectures described in Section~\ref{sec:architectures} are
-evaluated in Table~\ref{table:evaluation}. We use 5 main criteria to
-evaluate the architectures: integration complexity, robustness,
-extensibility, scalability and other specific features. Criteria break
-down to specific metrics where \emph{lower value indicates better
-  performance}. For each criterion, a global score is computed by
-summing up the individual metrics. We ensure that the different
-metrics in a criterion measure similar entities so that they can be
-summed. The criteria and metrics are explained hereafter.
+evaluated in Table~\ref{table:evaluation} using 5 main criteria:
+integration complexity, robustness, extensibility, scalability and
+other specific features. Criteria break down to specific metrics where
+\emph{lower value indicates better performance}. For each criterion, a
+global score is computed by summing up the individual metrics. We
+ensure that the different metrics in a criterion measure similar
+entities so that they can be summed. The criteria and metrics are
+explained hereafter.
 
@@
   the architecture. Workflow engines belong to different types when
-  they cannot be invoked using the same interface. Adding a new type
-  of workflow engine allows to execute more workflows in the science
-  gateway.
+  they cannot be invoked using the same interface. 
 \item New engine version (\texttt{E$_2$}): number of interactions or
@@
   'impossible' because nothing is impossible to implement if enough
-  effort is put.}. Four different features are identified:
+  effort is put.} Four different features are identified:
 \begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
@@
 engine instances through multiple workflow services. In VIP for
-instance, this feature has been available from release 1.17 (April
-2016). A basic load-balancing mechanism is available that sends new
-workflow executions to the engine instance that has the least active
+instance, this feature has been available from release 1.17. A basic
+load-balancing mechanism is available that sends new workflow
+executions to the engine instance that has the least active
 executions. To avoid ``black-hole'' syndromes created by failing
@@
 the science gateway configuration when a new engine instance is
-available. Distributing the execution of a single
-workflow in multiple engines is usually not possible unless the
-workflow engine has specific abilities (\texttt{S$_2$}=1). The
-scheduling of tasks on the infrastructure is as complex as in any
-other architecture since the workflow engine might implement any kind
-of scheduling policy (\texttt{S$_3$}=0).
+available. Distributing the execution of a single workflow in multiple
+engines is usually not possible unless the workflow engine has
+specific abilities (\texttt{S$_2$}=1). The scheduling of tasks on the
+infrastructure is as complex as in any other architecture since the
+workflow engine might implement any kind of scheduling policy
+(\texttt{S$_3$}=0).
 
@@
 \enlargethispage{5mm}
-\bibliographystyle{elsarticle-num}
+\bibliographystyle{plain}
 \bibliography{biblio}

### 2016-05-10 paper.tex
A few minor fixes

@@
 infrastructures such as clusters, grids, clouds and web
-services. These efforts resulted in tools such as
+services. These efforts resulted in tools among which
 Askalon~\cite{fahringer2005askalon},
@@
 Swift~\cite{zhao2007swift}, Taverna~\cite{oinn2004taverna},
-Triana~\cite{taylor2007triana}, VisTrails~\cite{callahan2006managing},
-WS-PGRADE~\cite{Kacsuk2012} and others. Such workflow engines usually
+Triana~\cite{taylor2007triana}, VisTrails~\cite{callahan2006managing}
+and WS-PGRADE~\cite{Kacsuk2012}. Such workflow engines usually
 describe applications in a high-level language with specific data and
@@
 Interfaces~\cite{gorgolewski2011nipype}), PSOM (Pipeline System for
-Octave and Matlab~\cite{bellec2012pipeline}), RPPL~\cite{1174106} and
-PMP (Poor Man's Pipeline~\cite{Ad-DabbaghY2006}), SPM (Statistical
+Octave and Matlab~\cite{bellec2012pipeline}), 
+PMP (Poor Man's Pipeline~\cite{Ad-DabbaghY2006}), RPPL~\cite{1174106}, SPM (Statistical
 Parametric Mapping~\cite{ashburner2011spm}) and FSL (FMRIB Software
@@
 analyses to avoid re-computation, and so on. A wide array of workflows
-have been implemented using these pipelines and are now shared across
+have been implemented using these pipeline systems and are now shared across
 neuroimaging groups world-wide, which represents a tremendous
@@
 Scalability is also a crucial concern for such multi-user,
-high-throughput systems. To this end, science gateways may balance the
-load among different instances of the same engine, and use advanced
-task scheduling policies on the infrastructure to improve performance,
-fault-tolerance and fairness among users.
+high-throughput systems. For this purpose, science gateways may
+balance the load among different instances of the same engine, start
+new engines elastically, and use advanced task scheduling policies on
+the infrastructure to improve performance, fault-tolerance and
+fairness among users.
 
@@
 
-Specific features may also be available, for instance
+Other specific features may also be available, for instance
 data vizualisation and quality control, workflow edition, debugging
@@
 diagrammed in Figure~\ref{fig:architectures} using the graphical
-notations shown in Figure~\ref{fig:notations}. This section describes the different types of
-interactions involved in the architectures. Architectures are then
-detailed and illustrated using examples of real
-systems. Table~\ref{table:system-classification} summarizes the
-classification of a few systems by architecture.
+notations shown in
+Figure~\ref{fig:notations}. Table~\ref{table:system-classification}
+summarizes the classification of a few systems by architecture.
 
@@
 software components with the science gateway. For instance, the
-workflow engine might be a portlet in a Liferay portal or a controller
+workflow engine might be a portlet in a Liferay portal or a model
 in a Ruby on Rails application. The workflow engine and the science
-gateway usually share a database where application, users and other
+gateway usually share a database where applications, users and other
 resources are stored\closednote{sorina}{VIP and Moteur also share the
@@
 application runtime environment
-(DARE~\cite{maddineni2012distributed}), in
+(DARE)~\cite{maddineni2012distributed}, in
 Galaxy~\cite{goecks2010galaxy}, and in the LONI Pipeline
@@
 mechanisms to facilitate the handling of task dependencies, for
-instance basic dependency lists as available in most batch managers
+instance basic dependency lists as available in most cluster managers
 (see for instance attribute \texttt{depend} of option \texttt{-W} in
@@
   (FSL tasks or PSOM agents). 7-8: Execution server starts sub-tasks
-  through task scheduler. FSL sub-tasks will run locally
+  through task scheduler (FSL sub-tasks or PSOM workers). FSL sub-tasks will run locally
   instead of being submitted again to CBRAIN through interaction 6. 8':
@@
 \label{sec:pool}
-See Figure~\subref*{archi:agent}. Workflows are submitted by the science
-gateway to a central pool through interaction \texttt{d}. Agents connect to
-the pool asynchronously to retrieve and execute workflows through
-interaction \texttt{f}. Agents may be started according to various
-policies, for instance to ensure load balancing. Agents may wrap
-different types of workflow engines. Workflow engine controls tasks
-and data on the infrastructure through interactions \texttt{b} and
-\texttt{c$_2$}, science gateway transfers user data through
-interaction \texttt{c$_1$}, and administrator installs workflows
-through interaction \texttt{a$_1$} and \texttt{a$_2$}.
+See Figure~\subref*{archi:agent}. Workflows are submitted by the
+science gateway to a central pool through interaction
+\texttt{d}. Agents connect to the pool asynchronously to retrieve and
+execute workflows through interaction \texttt{f}. Agents may be
+started according to various policies, for instance to ensure load
+balancing. Workflow engine controls tasks and data on the
+infrastructure through interactions \texttt{b} and \texttt{c$_2$},
+science gateway transfers user data through interaction
+\texttt{c$_1$}, and administrator installs workflows through
+interaction \texttt{a$_1$} and \texttt{a$_2$}.
 
@@
 (Figure~\subref*{archi:nested}-Left), workflows include other
-workflows that are executed by a \emph{child} engine. However, parent
+workflows that are executed by a \emph{child} engine. Parent
 and child engines might use different languages and might run on
@@
 total number of software pieces to develop, including interactions and
-components. It may be used to measure the development effort required
-to build the architecture.
+components.
 
@@
 workflow execution fails due to errors in the components or with the
-interactions in the software architecture. Note that errors coming
+interactions in the software architecture. Errors coming
 from the infrastructure (e.g., unavailable data or terminated tasks)
@@
 parameters involved in this interaction are modified.  Extensibility
-breaks down in 4 metrics depending on the type of element that has to
+breaks down to 4 metrics depending on the type of element that has to
 be added or replaced:
@@
 service invocation) and pool. Nested workflows (with service
-invocation) requires more integration than the other
+invocation) require more integration than the other
 architectures. Robustness roughly leads to the same ordering of
@@
 (\texttt{E$_3$}) because of the language conversion step. All
-architectures except nested workflows perform the same when
+architectures except nested workflows are equivalent when
 integrating new infrastructures (\texttt{E$_4$}).
@@
 
-A few limitations remain that should be considered when using the
+A few limitations should be considered when using the
 results of our evaluation. First of all, the evaluation is done at a
@@
 focused on the description of particular systems, or on the
-presentation of a particular architecture. For instance, the work
-in~\cite{Gannon2005} discusses service-oriented architectures for
-science gateways with workflow engines.  Instead, our analysis
+presentation of a particular architecture.  Instead, our analysis
 abstracts and evaluates architectural patterns independently from any
@@
 % Literature on SG architecture Very few works go beyond the
-% description of architectures of particular systems. The work in
+% description of architectures of particular systems. For instance, the work
+%in~\cite{Gannon2005} discusses service-oriented architectures for
+% science gateways with workflow engines.  The work in
 % http://link.springer.com/chapter/10.1007/11596141_3 discusses

### 2016-05-13 paper.tex
Update paper.tex

--- 2016-05-06 paper.tex
@@
   We investigate six software architectures commonly used to integrate
-  workflow engines in science gateways. In \emph{tight integration},
-  the workflow engine shares software components with the science
-  gateway. In \emph{service invocation}, the engine is isolated and
-  invoked through a specific software interface. In \emph{task
-    encapsulation}, the engine is wrapped as a computing task executed
-  on the infrastructure. In the \emph{pool model}, the engine is
-  bundled in an agent that connects to a central pool to fetch and
-  execute workflows. In \emph{nested workflows}, the engine is
-  integrated as a child process of another engine. In \emph{workflow
-    import}, the engine is integrated through workflow language
-  conversion. We describe and evaluate these architectures with
-  metrics that measure integration complexity, robustness,
+  workflow engines in science gateways. (1) In \emph{tight
+    integration}, the workflow engine shares software components with
+  the science gateway; (2) In \emph{service invocation}, the engine is
+  isolated and invoked through a specific software interface; (3) In
+  \emph{task encapsulation}, the engine is wrapped as a computing task
+  executed on the infrastructure; (4) In the \emph{pool model}, the
+  engine is bundled in an agent that connects to a central pool to
+  fetch and execute workflows; (5) In \emph{nested workflows}, the
+  engine is integrated as a child process of another engine; (6) In
+  \emph{workflow import}, the engine is integrated through workflow
+  language conversion. We describe and evaluate these architectures
+  with metrics that measure integration complexity, robustness,
   extensibility, scalability and support for meta-workflows and
@@
   meta-workflows are only available in nested workflows and workflow
-  import. \closednote{Naj}{I find this abstract to be vague, it can be made
-    more concise and poignant}{Tristan}{Added a description of the
-    architectures, following @pbellec's suggestion.}
+  import. \note{Naj}{I find this abstract to be vague, it can be made
+    more concise and poignant}
 \end{abstract}
@@
 architectures can be adopted to integrate workflow engines in science
-gateways, with important consequences on the development effort
-required and resulting system.
+gateways, however development effort required and the resulting system
+vary across different architectures.
 
@@
 SHIWA\footnote{\url{http://www.shiwa-workflow.eu}}.  In this section
-we provide background information and definitions of 
+we provide background information and definitions of different
 workflow engines, science gateways and infrastructures. In
@@
 Interfaces~\cite{gorgolewski2011nipype}), PSOM (Pipeline System for
-Octave and Matlab~\cite{bellec2012pipeline}), RPPL~\cite{1174106} and
-PMP (Poor Man's Pipeline~\cite{Ad-DabbaghY2006}), SPM (Statistical
-Parametric Mapping~\cite{ashburner2011spm}) and FSL (FMRIB Software
+Octave and Matlab~\cite{bellec2012pipeline}), RPPL~\cite{1174106} and PMP
+(Poor Man's Pipeline~\cite{Ad-DabbaghY2006}), SPM (Statistical Parametric
+Mapping~\cite{ashburner2011spm}) and FSL (FMRIB Software
 Library~\cite{Jenkinson2012782}) provide abstractions and functions to
@@
   scheduler'}{}{}, to execute tasks at a high throughput. Some of
-these tools also support advanced features such as provenance
-tracking, \closednote{Naj}{should we not mention PMP which is used for
-  CIVET?}{Tristan}{Done, as discussed.}  redundancy detection across
-analyses to avoid re-computation, and so on. A wide-array of
-workflows have been implemented using these pipelines and are now
-shared across neuroimaging groups world-wide. These domain-specific
-engines represent a tremendous opportunity for science gateways to
-leverage existing tools and applications. They nicely complement
-e-Science engines that are more oriented towards the exploitation of
-distributed computing infrastructures, in particular grids and clouds.
+these tools also support advanced features such as provenance tracking
+(Nipype, PSOM), \closednote{Naj}{should we not mention PMP which is
+  used for CIVET?}{Tristan}{Done, as discussed.}
+redundancy detection across analyses to avoid re-computation (PSOM),
+and so on \note{naj}{please remove so on, and be more precise?}. A wide-array \note{Naj}{no need for hyphenation here} of workflows have been implemented using these
+pipelines and are now shared across neuroimaging groups
+world-wide. These domain-specific engines represent a tremendous
+opportunity for science gateways to leverage existing tools and
+applications. They nicely complement e-Science engines that are more
+oriented towards the exploitation of distributed computing
+infrastructures, in particular grids and clouds.
 
@@
 \emph{infrastructure} (local server, cluster, grid or cloud) based on
-a workflow description (a.k.a \emph{workflow}) and using input data
+a workflow description (a.k.a \emph{workflow}) and\note{Naj}{I think and is redundant here, you can replace is with a comma} using input data
 that may consist of files, database entries or simple parameter
@@
 using such frameworks~\cite{kacsuk2014science,ardizzone2012decide} or
-as standalone systems~\cite{SHER-14,GLAT-13}. Most of these frameworks
-and instances include one or several workflow engines.
+as standalone systems~\cite{SHER-14,GLAT-13}. Most of these science
+gateway frameworks and instances include one or several workflow
+engines.
 
@@
 across systems. Some science gateways are tailored to a particular
-engine while other ones are more general and host applications
+engine, while other ones are more general and host applications
 executed by different types of engines. Some gateways may not even use any
-workflow engine and instead focus on the execution of simple tasks.
-
-Extensibility is an important property of the integration. New
+workflow engine and focus on the execution of simple tasks.
+
+Extensibility is an important property of the integration, as new
 workflows are added frequently, different types and versions of
 workflow engines may be integrated over time, and different kinds of
-infrastructure can be targeted.
-
-Scalability is also a crucial concern for such multi-user,
-high-throughput systems. To this end, science gateways may balance the
-load among different instances of the same engine, and use advanced
-task scheduling policies on the infrastructure to improve performance,
-fault-tolerance and fairness among users.
-
-Robustness is highly desirable as well since it is key to a
-good user experience. Simple architectures facilitate the
-implementation effort towards robust interactions which, in turn, have
+infrastructure can be targeted. Scalability is also a crucial concern
+for such multi-user systems that manage substantial amounts of data
+and execute computationally-intensive analyses. To this end, science
+gateways may balance the load among different instances of the same
+engine. They may also use advanced task scheduling policies on the
+infrastructure, e.g. to ensure fairness among
+users~\cite{FERR-14,CPE:CPE3708} or to improve
+fault-tolerance~\cite{wilkins2008teragrid,FERR-13}.
+
+Robustness is a highly desirable characteristic as well. Architectures
+showing reduced system complexity will greatly facilitate the
+implementation effort towards robust interactions which, in turn, will have
 a positive impact on characteristics such as gateway predictability,
-transparency, reliability, traceability and
-reproducibility. \closednote{Marc-Etienne}{I find that something is
-  weak in this last paragraph. Robustness will help with transparency,
-  sure, but it seems only a small facet.  Also, the text seem to
-  timidly favor architectures with less components/interactions
-  without much explanation... Complex systems can be robust, but at
-  greater costs...  Here is one: "Architectures showing reduced system
+transparence, general reliability and results traceability and
+reproducibility; characteristics that are also key to a good user
+experience. \closednote{Marc-Etienne}{I find that something is weak in
+  this last paragraph. Robustness will help with transparency, sure,
+  but it seems only a small facet.  Also, the text seem to timidly
+  favor architectures with less components/interactions without much
+  explanation... Complex systems can be robust, but at greater
+  costs...  Here is one: "Architectures showing reduced system
   complexity will greatly facilitate the implementation effort towards
@@
   experience."}{Tristan}{Agreed, fixed as
-  suggested.}\closednote{Tristan}{Work on the this sentence again.}{Tristan}{done.}
+  suggested.}\note{Tristan}{Work on the this sentence again.}
 
@@
 data vizualisation and quality control, workflow edition, debugging
-instruments, social tools among users, etc.
+instruments, social tools among users, etc. \note{Naj}{I think using etc after such as, for example and e.g. is not correct. --sorry for being pedantic}
 
@@
 2016). A basic load-balancing mechanism is available that sends new
-workflow executions to the engine instance that has the least active
+workflow executions to the engine instance which has the least active
 executions. To avoid ``black-hole'' syndromes created by failing
@@
 for the sub-tasks, which may generate issues such as selection of wrong
-batch queues on clusters. Finally, the interdependencies between the
+batch queues on clusters \note{Naj}{The lay reader wonders why? an example like the one offered for interdependencies can help}. Finally, the interdependencies between the
 workflow task and its sub-tasks may create deadlocks when there is
@@
 the number of computing resources available on this infrastructure,
-which makes such deadlocks unlikely to happen.
+which makes such deadlocks unlikely to happen. \note{Naj}{so is this the solution to fixing the above mentioned limitations with scalability?}
 
@@
 which is easily automated (\texttt{S$_1$=0}) and by design very
-suitable for elastic computing. For instance, auto-scaling rules can
+suitable for elastic computing.\note{Naj}{I feel the concept of elastic computing should be mentioned in introductory sections} For instance, auto-scaling rules can
 be implemented to start new agents when the workload in the science
@@
 Regarding extensibility, most architectures are overall comparable,
-except nested workflows (with service invocation) which is
+except nested workflows (with service invocation) which are
 significantly behind. This is explained by the complexity of the
@@
 services. Task encapsulation and workflow import (with service invocation)
-perform a bit better when integrating new engines (\texttt{E$_1$}),
+perform slightly better when integrating new engines (\texttt{E$_1$}),
 which is interesting given the current profusion of engines with
-different characteristics.  However, task encapsulation, workflow import
+different characteristics.\note {Naj}{what is interesting?}  However, task encapsulation, workflow import
 (with service invocation) and tight integration perform worse than the
-others when adding engine versions (\texttt{E$_2$}) due to the need to
+others when adding engine versions (\texttt{E$_2$}), due to the need to
 update infrastructure (task encapsulation), science gateway (tight
@@
 
-The pool architecture is overall the most scalable, which is not a
-surprise since it was designed precisely for scalability. Tight
+The pool architecture is overall the most scalable, which is not surprising since it is designed precisely for scalability. Tight
 integration is the least scalable and all the other architectures
 perform the same overall. The global scalability score, however, should not
-mask the unique characteristics of architectures regarding this
+mask \note{Naj}{what do you mean by mask?}the unique characteristics of architectures regarding this
 criterion. Nested workflows are the only architecture that can easily
@@
 task encapsulation and nested workflows could well become
-showstoppers depending on the type of targeted
-infrastructure. Non-reliable infrastructures, for example, could
+showstoppers \note{Naj}{please don't use these from-film-marketing words?! :-) }depending on the type of targeted
+infrastructure.
+Non-reliable infrastructures \note{Naj}{you mean by the showstoppers, i.e.
+the nested workflows which can accommodate distributed executions but are limited by scheduling constraints?} , for example, could
 hardly cope with workflow engines being wrapped in computing tasks as
@@
 could also become a critical feature for science gateways with
-important workloads, which would favor task encapsulation and pool.
+important workloads, which would favor task encapsulation and pool. \note{Naj}{I feel this paragraph is a bit vague, and doesn't distill and contextualize the thorough work described before.}
 
@@
 invocation) are the only architectures that support meta-workflows,
-which may be interesting in some cases. Also, task encapsulation and nested
+which may be interesting in some cases \note{Naj}{Which?}. Also, task encapsulation and nested
 workflows may complexify fine-grained debugging. \closednote{Rafael}{I would
@@
 A few limitations remain that should be considered when using the
-results of our evaluation. First of all, the evaluation is done at a
-quite high-level of abstraction that may be a bit artificial when
-applied to real systems. Abstracting architectures from their
-implementation in specific systems is useful to compare them, but we
-also realize that it comes at the cost of realism. In particular, the
+results of our evaluation.
+First, our evaluation methods aimed to provide comparative metrics, however the high-level of abstraction used to derive these metrics hinders the complexity of real systems to some extent.
+applied to real systems.\note{Naj}{Here, you were bashing your article harshly, so I rewrote it}. In particular, the
 presented architectures are theoretical patterns that may be blended
-in actual systems. The distinction between tight integration and
+in actual systems.\note{Naj}{something about "theoretical pattern", and "blending" is jarring, aare these common words in the computer sci context?} The distinction between tight integration and
 service invocation, for instance, may not always be that clear in
@@
 (service invocation), and may create much more reliability issues as
-well. However, quantifying the robustness and amount of development
-associated with each interaction and software component cannot
-realistically be done without entering into the details of a
-particular system.
+well. However, without entering into the details of a
+particular system,  quantification of the robustness or the amount of development
+associated with each interaction and software component will not be precise.
 
 The particular case of interaction \texttt{g} (workflow language
-conversion) should be further discussed since it is very likely that
-this interaction could not be easily generalized to any workflow
+conversion) should be further discussed \note{Naj}{so are you discussing it here, or in the future?}since 
+this interaction may not be easily generalized to any workflow
 language. For instance, converting FSL, PSOM or Nipype pipelines to
@@
 instantiations, in particular nested workflows with task
-encapsulation.
+encapsulation.\note{Naj}{I think in the discussion, you may want to close the circle on a statement in introduction, which referred to your 7-year experience with CBRAIN and VIP}
 
@@
 integrate workflow engines. For instance, some workflow engines may be
-already available as web services, which would tend to favor service
+already available as web services\note{Naj}{This is a good place to offer some examples}, which would tend to favor service
 invocation, and other science gateways may have strongly tested task
@@
 % Summary of contributions
-We reviewed architectures used to integrate workflow engines in
-science gateways, we described them in a system-independent framework
-suitable for comparison, we illustrated them on real systems, we
-developed quantitative metrics for evaluation, and we discussed the
+We have systematically reviewed architectures used to integrate workflow engines in
+science gateways. These architectures are described in a system-independent framework
+suitable for comparison, illustrated on real systems, and evaluated using novel quantitative metrics that allow simple comparison across architectures. We have discussed the
 pros and cons of all the presented architectures based on these metrics.
@@
 % Impact of the work
-To the best of our knowledge, our work is the first of its kind. So
+To the best of our knowledge, our work is the first of its kind.\note{Naj}{instead of its kind, explain what kind it is, e.g. systematic review of all ...} So
 far, the literature on science gateways and workflow engines has

### 2016-05-13 paper.tex

--- 2016-05-10 paper.tex
@@
 tracking, \closednote{Naj}{should we not mention PMP which is used for
-  CIVET?}{Tristan}{Done, as discussed.}  redundancy detection across
-analyses to avoid re-computation, and so on. A wide array of workflows
+  CIVET?}{Tristan}{Done, as discussed.} or  redundancy detection across
+analyses to avoid re-computation. A wide array of workflows
 have been implemented using these pipeline systems and are now shared across
@@
 \emph{infrastructure} (local server, cluster, grid or cloud) based on
-a workflow description (a.k.a \emph{workflow}) and using input data
+a workflow description (a.k.a \emph{workflow}), \closednote{Naj}{I think and is redundant here, you can replace is with a comma}{Tristan}{yes!} using input data
 that may consist of files, database entries or simple parameter
@@
 across systems. Some science gateways are tailored to a particular
-engine while other ones are more general and host applications
-executed by different types of engines. 
-%Some gateways may not even use any
-%workflow engine and instead focus on the execution of simple tasks.
+engine, while other ones are more general and host applications
+executed by different types of engines. %Some gateways may not even use any
+%workflow engine and focus on the execution of simple tasks.
 
@@
 balance the load among different instances of the same engine, start
-new engines elastically, and use advanced task scheduling policies on
-the infrastructure to improve performance, fault-tolerance and
-fairness among users.
+new engines elastically using auto-scaling techniques such as the ones
+reviewed in~\cite{lorido2012auto}, and use advanced task scheduling
+policies on the infrastructure to improve performance, fault-tolerance
+and fairness among users.
 
@@
 data vizualisation and quality control, workflow edition, debugging
-instruments, social tools among users, etc.
+instruments, or social tools among users.
+\closednote{Naj}{I think using etc after such as, for example and e.g.
+is not correct.
+--sorry for being pedantic}{Tristan}{You are completely right, this is now fixed.}
 
@@
 load-balancing mechanism is available that sends new workflow
-executions to the engine instance that has the least active
+executions to the engine instance which has the least active
 executions. To avoid ``black-hole'' syndromes created by failing
@@
 any other task upon user submission (\texttt{S$_1$}=0). This is a
-major interest of task encapsulation. Distributed engines are
-not supported by default (\texttt{S$_2$}=1). Task scheduling is
-slightly more complex than in the other approaches due to the special
-role of the task that executes the workflow engine
-(\texttt{S$_3$}=1). Indeed, the reliability of this task is critical
-since all the sub-tasks in the workflow depend on it and, depending on
-the recovery capabilities of the workflow engine, may need to be
-resubmitted if the workflow task fails. The workflow task is also
-longer than all its sub-tasks, which increases its chances of
-failure. In addition, task parameters, for instance estimated
-walltime, are more difficult to estimate for the workflow task than
-for the sub-tasks, which may generate issues such as selection of wrong
-batch queues on clusters. Finally, the interdependencies between the
-workflow task and its sub-tasks may create deadlocks when there is
-contention. For instance, if only 1 computing resource is
-available for the science gateway and if the workflow task is
-running on it and submits sub-tasks, then the sub-tasks could only
-execute when the resource is available, which will never happen
-because the workflow task will not complete until the sub-tasks
-complete. This configuration can be generalized to an infrastructure
-with $n$ resources where $n$ workflows are submitted. In practice,
-however, the number of submitted workflows usually remains lower than
-the number of computing resources available on this infrastructure,
-which makes such deadlocks unlikely to happen.
+major interest of task encapsulation. Distributed engines are not
+supported by default (\texttt{S$_2$}=1). Task scheduling is slightly
+more complex than in the other approaches due to the special role of
+the task that executes the workflow engine (\texttt{S$_3$}=1). Indeed,
+the reliability of this task is critical since all the sub-tasks in
+the workflow depend on it and, depending on the recovery capabilities
+of the workflow engine, may need to be resubmitted if the workflow
+task fails. The workflow task is also longer than all its sub-tasks,
+which increases its chances of failure. In addition, task parameters,
+for instance estimated walltime, are more difficult to estimate for
+the workflow task than for the sub-tasks because workflows are by
+definition more complex than their sub-tasks: errors on sub-task
+parameter estimations accumulate in the workflow, and additional
+control constructs such as tests and loops may further increase the
+uncertainty. Such parameter estimation errors may generate issues such
+as selection of wrong batch queues on clusters or task termination due
+to exceeded quotas. \closednote{Naj}{The lay reader wonders why?  an
+  example like the one offered for interdependencies can
+  help}{Tristan}{Done, thanks.} Finally, the interdependencies between
+the workflow task and its sub-tasks may create deadlocks when there is
+contention. For instance, if only 1 computing resource is available
+for the science gateway and if the workflow task is running on it and
+submits sub-tasks, then the sub-tasks could only execute when the
+resource is available, which will never happen because the workflow
+task will not complete until the sub-tasks complete. This
+configuration can be generalized to an infrastructure with $n$
+resources where $n$ workflows are submitted. In practice, however, the
+number of submitted workflows usually remains lower than the number of
+computing resources available on this infrastructure, which makes such
+deadlocks unlikely to happen. \closednote{Naj}{so is this the solution to
+  fixing the above mentioned limitations with
+  scalability?}{Tristan}{Unfortunately this is not a solution. We just
+  say that this situation is unlikely to happen which is true.}
 
@@
 which is easily automated (\texttt{S$_1$=0}) and by design very
-suitable for elastic computing. For instance, auto-scaling rules can
+suitable for elastic computing.\closednote{Naj}{I feel the concept of elastic computing should be mentioned in introductory sections}{Tristan}{You are right, this is now fixed.} For instance, auto-scaling rules can
 be implemented to start new agents when the workload in the science
@@
 
-A few limitations should be considered when using the
-results of our evaluation. First of all, the evaluation is done at a
-quite high-level of abstraction that may be a bit artificial when
-applied to real systems. Abstracting architectures from their
-implementation in specific systems is useful to compare them, but we
-also realize that it comes at the cost of realism. In particular, the
-presented architectures are theoretical patterns that may be blended
-in actual systems. The distinction between tight integration and
+A few limitations should be considered when using the results of our
+evaluation. First, our evaluation methods aimed to provide comparative
+metrics, however the high-level of abstraction used to derive these
+metrics hinders the complexity of real systems to some
+extent.\closednote{Naj}{Here, you were bashing your article harshly,
+  so I rewrote it}{Tristan}{yup, thanks!}. In particular, the
+presented architectures are abstract patterns that may be mixed
+together in actual systems.\answerednote{Naj}{something about "theoretical
+  pattern", and "blending" is jarring, are these common words in the
+  computer sci context?}{Tristan}{Better now?} The distinction between tight integration and
 service invocation, for instance, may not always be that clear in
@@
 work would still apply to broadly compare and categorize such cases of
-hydrid architectures.
+hybrid architectures.
 
 Moreover, all the interactions involved in the architectures were
-treated equally while some of them are obviously more complex
-than others. For example, interaction \texttt{g} (workflow language
+treated equally while some of them are obviously more complex than
+others. For example, interaction \texttt{g} (workflow language
 conversion) is clearly more complex than interaction \texttt{d}
 (service invocation), and may create much more reliability issues as
-well. However, quantifying the robustness and amount of development
-associated with each interaction and software component cannot
-realistically be done without entering into the details of a
-particular system.
+well. However, without entering into the details of a particular
+system, quantification of the robustness or the amount of development
+associated with each interaction and software component will not be
+precise.
 
 The particular case of interaction \texttt{g} (workflow language
-conversion) should be further discussed since it is very likely that
-this interaction could not be easily generalized to any workflow
+conversion) should be further discussed \note{Naj}{so are you discussing it here, or in the future?}since 
+this interaction may not be easily generalized to any workflow
 language. For instance, converting FSL, PSOM or Nipype pipelines to

### 2016-05-13 paper.tex
Some additional fixes following @najmahan's comments

@@
 nested workflow architecture, with 2 infrastructures and 2 workflow
-services. Task encapsulation and workflow import (with service invocation)
-perform slightly better when integrating new engines (\texttt{E$_1$}),
-which is interesting given the current profusion of engines with
-different characteristics.\note {Naj}{what is interesting?}  However, task encapsulation, workflow import
-(with service invocation) and tight integration perform worse than the
-others when adding engine versions (\texttt{E$_2$}), due to the need to
-update infrastructure (task encapsulation), science gateway (tight
+services. Task encapsulation and workflow import (with service
+invocation) perform slightly better when integrating new engines
+(\texttt{E$_1$}), which makes them useful architectures for science
+gateways that do not focus on a particular engine.\closednote{Naj}{what is
+  interesting?}{Tristan}{Rephrased, thanks.}  However, task encapsulation, workflow import (with
+service invocation) and tight integration perform worse than the
+others when adding engine versions (\texttt{E$_2$}), due to the need
+to update infrastructure (task encapsulation), science gateway (tight
 integration), or workflow converter (workflow import). Workflow import
@@
 (\texttt{E$_3$}) because of the language conversion step. All
-architectures except nested workflows are equivalent when
-integrating new infrastructures (\texttt{E$_4$}).
-
-The pool architecture is overall the most scalable, which is not surprising since it is designed precisely for scalability. Tight
+architectures except nested workflows are equivalent when integrating
+new infrastructures (\texttt{E$_4$}).
+
+The pool architecture is overall the most scalable, which is not
+surprising since it is designed precisely for scalability. Tight
 integration is the least scalable and all the other architectures
-perform the same overall. The global scalability score, however, should not
-mask \note{Naj}{what do you mean by mask?}the unique characteristics of architectures regarding this
-criterion. Nested workflows are the only architecture that can easily
-accommodate distributed workflow execution, which can be critical in
-some cases. At the same time, the scheduling constraints created by
-task encapsulation and nested workflows could well become
-showstoppers \note{Naj}{please don't use these from-film-marketing words?! :-) }depending on the type of targeted
-infrastructure.
-Non-reliable infrastructures \note{Naj}{you mean by the showstoppers, i.e.
-the nested workflows which can accommodate distributed executions but are limited by scheduling constraints?} , for example, could
-hardly cope with workflow engines being wrapped in computing tasks as
-done in task encapsulation. The availability of multiple engine instances
-could also become a critical feature for science gateways with
-important workloads, which would favor task encapsulation and pool. \note{Naj}{I feel this paragraph is a bit vague, and doesn't distill and contextualize the thorough work described before.}
+perform the same overall. The global scalability score, however,
+should not conceal \closednote{Naj}{what do you mean by
+  mask?}{Tristan}{I guess 'hide' or 'conceal' are better
+  words. Fixed.} the unique characteristics of architectures regarding
+this criterion. Nested workflows are the only architecture that can
+easily accommodate distributed workflow execution, which can be
+critical in some cases. At the same time, the scheduling constraints
+created by task encapsulation and nested workflows may become problematic \closednote{Naj}{please don't use these from-film-marketing
+  words?! :-) }{Tristan}{OK!} depending on the type of targeted
+infrastructure. Non-reliable infrastructures \closednote{Naj}{you mean by
+  the showstoppers, i.e. the nested workflows which can accommodate
+  distributed executions but are limited by scheduling constraints?}{Tristan}{Yes. I rephrased a bit this part.} ,
+for example, could hardly cope with workflow engines being wrapped in
+computing tasks as done in task encapsulation. The availability of
+multiple engine instances could also become a critical feature for
+science gateways with important workloads, which would favor task
+encapsulation and pool. \note{Naj}{I feel this paragraph is a bit
+  vague, and doesn't distill and contextualize the thorough work
+  described before.}
 
@@
 workflows (with service invocation) and workflow import (with service
-invocation) are the only architectures that support meta-workflows,
-which may be interesting in some cases \note{Naj}{Which?}. Also, task encapsulation and nested
-workflows may complexify fine-grained debugging. \closednote{Rafael}{I would
-  suggest to provide some more examples when this could be
-  interesting.}{Tristan}{Done.}  The availability of fine-grained debugging
-information may also be critical to efficient user support.
+invocation) are the only architectures that support meta-workflows;
+task encapsulation and nested workflows may complexify
+fine-grained debugging. \closednote{Rafael}{I would suggest to provide
+  some more examples when this could be interesting.}{Tristan}{Done.}
+The availability of fine-grained debugging information may also be
+critical to efficient user support.
 
@@
 The particular case of interaction \texttt{g} (workflow language
-conversion) should be further discussed \note{Naj}{so are you discussing it here, or in the future?}since 
-this interaction may not be easily generalized to any workflow
-language. For instance, converting FSL, PSOM or Nipype pipelines to
-any other workflow language is problematic because these engines rely
-on general-purpose programming languages such as Bash, Octave and
-Python, which are much richer than scientific workflow languages.
+conversion) requires particular attention when implementing a real
+system \closednote{Naj}{so are you discussing it here, or in the
+  future?}{Tristan}{It's out of scope, a complete paper in itself. Rephrased this part.} since this interaction may not be easily generalized to any
+workflow language. For instance, converting FSL, PSOM or Nipype
+pipelines to any other workflow language is problematic because these
+engines rely on general-purpose programming languages such as Bash,
+Octave and Python, which are much richer than scientific workflow
+languages.
 
@@
 instantiations, in particular nested workflows with task
-encapsulation.\note{Naj}{I think in the discussion, you may want to close the circle on a statement in introduction, which referred to your 7-year experience with CBRAIN and VIP}
+encapsulation.\answerednote{Naj}{I think in the discussion, you may want to close the circle on a statement in introduction, which referred to your 7-year experience with CBRAIN and VIP}{Tristan}{Good idea, will think about it.}
 
@@
 % Impact of the work
-To the best of our knowledge, our work is the first of its kind.\note{Naj}{instead of its kind, explain what kind it is, e.g. systematic review of all ...} So
-far, the literature on science gateways and workflow engines has
-focused on the description of particular systems, or on the
-presentation of a particular architecture.  Instead, our analysis
-abstracts and evaluates architectural patterns independently from any
-particular system. It provides general insight to integrate science
-gateways and workflow engines, which we hope will be useful for
-software architects.
+To the best of our knowledge, our work is the first to systematically
+review and compare software architectures used to integrate workflow
+engines in science gateways.\closednote{Naj}{instead of its kind,
+  explain what kind it is, e.g. systematic review of all
+  ...}{Tristan}{Done.} So far, the literature on science gateways and
+workflow engines has focused on the description of particular systems,
+or on the presentation of a particular architecture.  Instead, our
+analysis abstracts and evaluates architectural patterns independently
+from any particular system. It provides general insight to integrate
+science gateways and workflow engines, which we hope will be useful
+for software architects.
 

### 2016-05-13 paper.tex
up to 2.1

@@
   conversion. We describe and evaluate these architectures with
-  metrics that measure integration complexity, robustness,
+  metrics for assessment of integration complexity, robustness,
   extensibility, scalability and support for meta-workflows and
@@
   integration and task encapsulation are the easiest to integrate and
-  are most robust. Extensibility is equivalent in most
+  the most robust. Extensibility is equivalent in most
   architectures. The pool model is the most scalable one and
@@
 and VIP~\cite{GLAT-13} science gateways for medical image analysis
-during the past 7 years; as well as by lessons learned from several
+during the past 7 years, as well as by lessons learned from several
 science gateway and workflow projects such as
@@
 values. Although simplistic, this definition covers both e-Science workflow systems and
-domain-specific pipeline systems.\closednote{Marc-e}{To remain consistent with above descriptions}{Tristan}{yes!} Some workflow engines, usually e-Science
-ones, may transfer data across the infrastructure, and others, usually
+domain-specific pipeline systems.\closednote{Marc-e}{To remain consistent with above descriptions}{Tristan}{yes!} 
+Some workflow engines, 
+usually from the e-Science community, 
+may transfer data across the infrastructure, and others, usually
 domain-specific ones, may leave this role to external processes. Workflows
@@
 across systems. Some science gateways are tailored to a particular
-engine, while other ones are more general and host applications
+engine, while others are more general and host applications
 executed by different types of engines. %Some gateways may not even use any
@@
 Architectures to integrate workflow engines in science gateways are
-diagrammed in Figure~\ref{fig:architectures} using the graphical
-notations shown in
+presented in Figure~\ref{fig:architectures} using the graphical
+notations defined in
 Figure~\ref{fig:notations}. Table~\ref{table:system-classification}
@@
 The interactions involved in the architectures are described below and
-labeled as in Figure~\ref{fig:architectures}.
+labeled (a,b,c, ...) as in Figure~\ref{fig:architectures}.
 
@@
   computing models.
-\item Workflow conversion: translation from one workflow language to
-  another one. This interaction may not be available or implementable
+\item Workflow conversion: translation from one workflow language into
+  another. This interaction may not be available or possible
   for every workflow language. It has been developed mostly for
-  well-structured and relatively simple workflow languages such as between GWorkflowDL
-  and Scufl~\cite{OLAB-09}, and between the 4 languages that were involved in the
+  translation between well-structured and relatively simple workflow languages such as GWorkflowDL
+  and Scufl~\cite{OLAB-09}, and the 4 languages that were involved in the
   SHIWA initiative.
+ \note{Silvia}{Should we add here the 4?
+ 	and for translation among the 5 workflow systems in the SHIWA project: Askalon, MOTEUR, Taverna, Triana and WS-PGRADE)} 
 \end{enumerate}

### 2016-05-13 paper.tex
up to 2.5

@@
 \label{sec:architectures}
+
+\note{Sil}{i changed above a few occurrences of "diagrammed". i know it is correct, but i am not used to seeing it. this is a matter of style, just forget my changes below then}
 
@@
   Service & \pbox{1.5\columnwidth}{
-            Apache Airvata~\cite{marru2011apache}, e-BioInfra~\cite{shahand2015data}$^{w,n}$, HubZero with Pegasus~\cite{CPE:CPE3257}, MoSGrid~\cite{kruger2014mosgrid}$^w$, System in~\cite{wu2010accelerating}, Vine Toolkit~\cite{DBLP:journals/scpe/SzejnfeldDKKKKLPTWDNW10}, Virtual Imaging Platform~\cite{GLAT-13}$^n$, WS-PGRADE/gUSE framework~\cite{Kacsuk2012},
+            Apache Airvata~\cite{marru2011apache}, Neurosciece Gateway~\cite{shahand2015data}$^{w,n}$, HubZero with Pegasus~\cite{CPE:CPE3257}, MoSGrid~\cite{kruger2014mosgrid}$^w$, System in~\cite{wu2010accelerating}, Vine Toolkit~\cite{DBLP:journals/scpe/SzejnfeldDKKKKLPTWDNW10}, Virtual Imaging Platform~\cite{GLAT-13}$^n$, WS-PGRADE/gUSE framework~\cite{Kacsuk2012},
             Science gateways in~\cite{kacsuk2014science}$^w$
@@
 user.
-
-This architecture is largely adopted, in systems such as Apache
+\note{Sil}{about c1: should it also show from engine to infra in the figure? this is what i understand from the above sentence}
+
+This architecture is largely adopted in systems such as Apache
 Airvata~\cite{marru2011apache}, Vine
 Toolkit~\cite{DBLP:journals/scpe/SzejnfeldDKKKKLPTWDNW10}, Virtual
-Imaging Platform~\cite{GLAT-13}, the system
-in~\cite{wu2010accelerating} and the WS-PGRADE/gUSE
+Imaging Platform~\cite{GLAT-13}, the systems
+in~\cite{wu2010accelerating,shahand:2012jgc} and the WS-PGRADE/gUSE
 framework~\cite{Kacsuk2012}. The integration between the
 HubZero science gateway and the Pegasus workflow engine performed
-in~\cite{CPE:CPE3257} also uses a service architecture where the
+in~\cite{CPE:CPE3257} also uses a service architecture, where the
 \texttt{d} interaction is implemented as a set of calls to Pegasus'
@@
 the infrastructure, so that interactions \texttt{c$_1$} and
-\texttt{c$_2$} are both covered by \texttt{c}. In practice, both
+\texttt{c$_2$} are both covered by \texttt{c}. 
+\note{Sil}{is this next sentence relevant?}In practice, both
 \texttt{c$_1$} and \texttt{c$_2$} can be implemented using the same
@@
 workflow engine~\cite{bellec2012pipeline}. The CBRAIN-FSL integration
-allows to leverage FSL pipelines written in low-level workflow
+allows leveraging FSL pipelines written in low-level workflow
 languages (Linux executables and scripts) that submit tasks uniformly
-through a specific tool called fsl\_sub. It is diagrammed in
+through a specific tool called fsl\_sub. This integration is shown in
 Figure~\subref*{fig:cbrain-fsl-architecture}, where the science gateway is

### 2016-05-13 paper.tex
One fix following @najmahan's comments

--- 2016-05-13 paper.tex
@@
   words?! :-) }{Tristan}{OK!} depending on the type of targeted
-infrastructure. Non-reliable infrastructures \closednote{Naj}{you mean by
+infrastructure. Non-reliable infrastructures\closednote{Naj}{you mean by
   the showstoppers, i.e. the nested workflows which can accommodate
-  distributed executions but are limited by scheduling constraints?}{Tristan}{Yes. I rephrased a bit this part.} ,
+  distributed executions but are limited by scheduling constraints?}{Tristan}{Yes. I rephrased a bit this part.},
 for example, could hardly cope with workflow engines being wrapped in

### 2016-05-13 paper.tex
Some fixes following @SilviaOlabarriaga's comments

@@
 
-\note{Sil}{i changed above a few occurrences of "diagrammed". i know it is correct, but i am not used to seeing it. this is a matter of style, just forget my changes below then}
+\closednote{Sil}{i changed above a few occurrences of "diagrammed".
+i know it is correct, but i am not used to seeing it.
+this is a matter of style, just forget my changes below then}{Tristan}{That's perfect, thanks!}
 
@@
 The interactions involved in the architectures are described below and
-labeled (a,b,c, ...) as in Figure~\ref{fig:architectures}.
+labeled (\texttt{a},\texttt{b},\texttt{c}, \ldots) as in Figure~\ref{fig:architectures}.
 
@@
 \item Workflow conversion: translation from one workflow language into
-  another. This interaction may not be available or possible
-  for every workflow language. It has been developed mostly for
-  translation between well-structured and relatively simple workflow languages such as GWorkflowDL
-  and Scufl~\cite{OLAB-09}, and the 4 languages that were involved in the
-  SHIWA initiative.
- \note{Silvia}{Should we add here the 4?
- 	and for translation among the 5 workflow systems in the SHIWA project: Askalon, MOTEUR, Taverna, Triana and WS-PGRADE)} 
+  another. This interaction may not be available or possible for every
+  workflow language. It has been developed mostly for translation
+  between well-structured and relatively simple workflow languages
+  such as GWorkflowDL and Scufl~\cite{OLAB-09}, and for translation
+  among the 5 workflow systems in the SHIWA project: Askalon, MOTEUR,
+  Taverna, Triana and WS-PGRADE.  \closednote{Silvia}{Should we add here the
+    4?  and for translation among the 5 workflow systems in the SHIWA
+    project: Askalon, MOTEUR, Taverna, Triana and WS-PGRADE)}{Tristan}{Yes, done!}
 \end{enumerate}
@@
 transfers to execute them (\texttt{c$_2$}). User data is usually
-managed through the science gateway (\texttt{c$_1$}), although it
-might as well be delivered by the workflow engine directly to the
-user.
-\note{Sil}{about c1: should it also show from engine to infra in the figure? this is what i understand from the above sentence}
+managed through the science gateway (\texttt{c$_1$}), although in some
+variations of the architecture (not represented in
+Figure~\ref{fig:architectures}) it might as well be delivered by the
+workflow engine directly to the user. \answerednote{Sil}{about c1: should it
+  also show from engine to infra in the figure? this is what i
+  understand from the above sentence}{Tristan}{Actually c1 is either
+  coming from the science gatewray or from the workflow engine but
+  then it merges with c2. I don't think we can cover both cases in a
+  single architecture, the latter case would be another variation of
+  this architecture. I modified a bit the text, is it clearer now?}
 
@@
 \texttt{c$_2$} are both covered by \texttt{c}. 
-\note{Sil}{is this next sentence relevant?}In practice, both
-\texttt{c$_1$} and \texttt{c$_2$} can be implemented using the same
-pieces of code.
+\closednote{Sil}{is this next sentence relevant?}{Tristan}{I just wanted to  justify why these two interactions can be grouped but I guess you're right (I removed the sentence.}%In practice, both
+%\texttt{c$_1$} and \texttt{c$_2$} can be implemented using the same
+%pieces of code.
 
@@
 workflows (with service invocation) and workflow import (with service
-invocation) are the only architectures that support meta-workflows;
-task encapsulation and nested workflows may complexify
-fine-grained debugging. \closednote{Rafael}{I would suggest to provide
-  some more examples when this could be interesting.}{Tristan}{Done.}
-The availability of fine-grained debugging information may also be
-critical to efficient user support.
+invocation) are the only architectures that support
+meta-workflows. Besides, task encapsulation and nested workflows may
+complexify fine-grained debugging. \closednote{Rafael}{I would suggest
+  to provide some more examples when this could be
+  interesting.}{Tristan}{Done.}  The availability of fine-grained
+debugging information may also be critical to efficient user support.
 
@@
 presented architectures are abstract patterns that may be mixed
-together in actual systems.\answerednote{Naj}{something about "theoretical
+together in actual systems.\closednote{Naj}{something about "theoretical
   pattern", and "blending" is jarring, are these common words in the

### 2016-05-13 paper.tex
done fornow with sec 2

--- 2016-05-13 paper.tex
@@
   Service & \pbox{1.5\columnwidth}{
-            Apache Airvata~\cite{marru2011apache}, Neurosciece Gateway~\cite{shahand2015data}$^{w,n}$, HubZero with Pegasus~\cite{CPE:CPE3257}, MoSGrid~\cite{kruger2014mosgrid}$^w$, System in~\cite{wu2010accelerating}, Vine Toolkit~\cite{DBLP:journals/scpe/SzejnfeldDKKKKLPTWDNW10}, Virtual Imaging Platform~\cite{GLAT-13}$^n$, WS-PGRADE/gUSE framework~\cite{Kacsuk2012},
+            Apache Airvata~\cite{marru2011apache}, Neuroscience Gateway~\cite{shahand2015data}$^{w,n}$, HubZero with Pegasus~\cite{CPE:CPE3257}, MoSGrid~\cite{kruger2014mosgrid}$^w$, System in~\cite{wu2010accelerating}, Vine Toolkit~\cite{DBLP:journals/scpe/SzejnfeldDKKKKLPTWDNW10}, Virtual Imaging Platform~\cite{GLAT-13}$^n$, WS-PGRADE/gUSE framework~\cite{Kacsuk2012},
             Science gateways in~\cite{kacsuk2014science}$^w$
@@
 different workflow services invoked by distinct \texttt{d}
-interactions. If this is not the case then workflow services can be
-collapsed in a single one with a \texttt{d} interaction with
+interactions. If this is not the case, then workflow services can be
+collapsed into a single one with a \texttt{d} interaction with
 itself. An example of such interaction is the use of hierarchical
@@
 \includegraphics[width=1.5\columnwidth]{figures/shiwa-science-gateway.pdf}
-\caption{Nested workflow execution through SHIWA Science Gateway. The
-  parent workflow engine is \texttt{WS-PGRADE}, invoked as a service
+\caption{Nested workflow execution through the SHIWA Science Gateway. The
+  parent workflow engine is \texttt{WS-PGRADE/gUSE}, invoked as a service
   in the Science Gateway (step \texttt{1}, interaction \texttt{d} on
@@
 workflow import. In the SHIWA Simulation Platform, it is implemented
-through the IWIR language that provides a common language for
+through the IWIR language, which provides a common language for
 portability across grid workflow
-systems~\cite{plankensteiner-prodan-etal:2013} and allows to convert
+systems~\cite{plankensteiner-prodan-etal:2013} and allows conversion
 among $n$ workflow languages using $2n$ interactions instead of $n^2$.
 Tavaxy~\cite{Abouelhoda2012}
-allows to import, merge and run Taverna~\cite{oinn2004taverna} and
+enables the import, merging and execution of Taverna~\cite{oinn2004taverna} and
 Galaxy~\cite{goecks2010galaxy} workflows; when some workflow parts
@@
   words?! :-) }{Tristan}{OK!} depending on the type of targeted
-infrastructure. Non-reliable infrastructures\closednote{Naj}{you mean by
+infrastructure. Non-reliable infrastructures \closednote{Naj}{you mean by
   the showstoppers, i.e. the nested workflows which can accommodate
-  distributed executions but are limited by scheduling constraints?}{Tristan}{Yes. I rephrased a bit this part.},
+  distributed executions but are limited by scheduling constraints?}{Tristan}{Yes. I rephrased a bit this part.} ,
 for example, could hardly cope with workflow engines being wrapped in

### 2016-05-13 paper.tex
up to integration complexity

@@
 \subsection{Science gateways}
+
+\label{sec:intro_sg}
 
@@
 The architectures described in Section~\ref{sec:architectures} are
-evaluated in Table~\ref{table:evaluation} using 5 main criteria:
+evaluated in Table~\ref{table:evaluation} using five main criteria:
 integration complexity, robustness, extensibility, scalability and
-other specific features. Criteria break down to specific metrics where
+other specific features. These are based on science gateway properties introduced in Section~\ref{sec:intro_sg}. Criteria break down to specific metrics where
 \emph{lower value indicates better performance}. For each criterion, a
-global score is computed by summing up the individual metrics. We
-ensure that the different metrics in a criterion measure similar
-entities so that they can be summed. The criteria and metrics are
+global score is computed by summing up the individual scores for each metric. We
+ensure that the different metrics used to calculate the global score for a criterion are comparable, so that they can be combined meaningfully. The criteria and metrics are
 explained hereafter.
+
+\note{Sil}{is any normalization used, for example, to the maximum values found in the analysis?}
 
@@
 
-\emph{Integration complexity} is obtained by counting the total number of
-interactions and components on the architecture diagram. It breaks
+Science gateway \emph{integration complexity} is measured through the number of software components and their interactions. It breaks
 down to the following 2 metrics:
@@
 One may wonder whether infrastructure should be counted in
-\texttt{I$_1$} since it is usually not developed by the groups who
-integrate workflow engines in science gateway. However, integrating an
-infrastructure in the architecture usually requires some technical
-effort (e.g., account creation, software installation, APIs, etc.), which is
-why we keep it in the metric. The two metrics are summed to obtain a
+\texttt{I$_1$}, since it is usually an external entity that is not developed nor maintained by the groups who
+integrate workflow engines in science gateway. We support its inclusion here because integrating an
+infrastructure into the system usually requires some technical
+effort (e.g., account creation, software installation, APIs, etc.), therefore increasing complexity. The two metrics are summed to obtain a
 global score that measures the
-total number of software pieces to develop, including interactions and
-components.
+total number of software pieces affected by integration.
 

### 2016-05-13 paper.tex
up to extensibility

@@
 
-\note{Sil}{is any normalization used, for example, to the maximum values found in the analysis?}
+\note{Sil}{normalization is explained in caption tab 2. i wonder whether something should be said here, as questions about the range of the score immediately pops to mind..}
 
@@
 
+\note{Sil}{i was wondering whether it would be more clear to present the metrics like this:
+\item \texttt{I$_1$}: total number of components, and
+
+now the parenthesis are next to text that does not really match the metric definition (exception in this case) could also use boldface
+	}
 Science gateway \emph{integration complexity} is measured through the number of software components and their interactions. It breaks
@@
 
-\emph{Robustness of workflow execution} measures the likelihood that
-workflow execution fails due to errors in the components or with the
-interactions in the software architecture. Errors coming
+\emph{Robustness of workflow execution} measures the likelihood that the
+workflow execution fails due to errors in the components or in the
+interactions of the software architecture. Errors coming
 from the infrastructure (e.g., unavailable data or terminated tasks)
 or workflows (e.g., wrong user input or application errors) are not
-covered since they do not stem directly from the software
+covered, since they do not stem directly from the software
 architecture. Robustness is measured here as a consequence of global
-complexity since complex architectures tend to be more prone to
-failure. More precisely, robustness is determined as the number of
-software components and interactions that are specific to workflow
-execution:
+complexity, since complex architectures tend to be more prone to
+failure. More precisely, robustness is determined by the number of
+software components and interactions represented in
+red in the architecture diagrams in
+Figure~\ref{fig:architectures}, which are specific to workflow
+execution. Two metrics are used:
 \begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
-\item Components (\texttt{R$_1$}): number of specific components
-  involved in workflow executions. Science
-gateway, for instance, is \emph{not} specific to workflow execution
+\item Number of specific components
+involved in workflow executions (\texttt{R$_1$}). The Science
+gateway component, for instance, is \emph{not} specific to workflow execution
 since it is used to authenticate users, add new workflows, transfer
-user data, etc.
-\item Interactions (\texttt{R$_2$}): number of specific interactions
-  involved in workflow executions.
+user data, etc., and
+\item Number of specific interactions
+involved in workflow executions (\texttt{R$_2$}).
+\note{Sil}{add here an example of an interaction that is not, for example, data transfers?}
 \end{itemize}
-These specific components and interactions are the ones represented in
-red in the architecture diagrams in
-Figure~\ref{fig:architectures}. \texttt{R$_1$} and \texttt{R$_2$} are
+\texttt{R$_1$} and \texttt{R$_2$} are
 summed to obtain the global score for this criterion, which measures
@@
 \emph{Extensibility} measures the difficulty to replace or add
-elements in the architecture. It is determined as the number of interactions and
+elements to the architecture. It is determined by the number of interactions and
 components that need to be modified when a new element is
@@
 parameters involved in this interaction are modified.  Extensibility
-breaks down to 4 metrics depending on the type of element that has to
+breaks down to 4 metrics depending on the type of element (engine, workflow or infrastructure) that has to
 be added or replaced:
 \begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
-\item New engine type (\texttt{E$_1$}): number of interactions or
-  components to modify to integrate a new type of workflow engine in
+\item New engine \texttt{E$_1$}: number of interactions or
+  components that need to be modified to integrate a new type of workflow engine in
   the architecture. Workflow engines belong to different types when
   they cannot be invoked using the same interface. 
-\item New engine version (\texttt{E$_2$}): number of interactions or
-  components to modify to integrate a new version of a workflow engine
+  \note{Si}{new interface = new interaction? maybe better to use one term only here. Maybe also refer to how this shows on figure 2}
+\item Engine version upgrade \texttt{E$_2$}: number of interactions or
+  components modified to integrate a new version of a workflow engine
   in the architecture, assuming that another version of the same
@@
   the infrastructure, etc.
-\item New workflow (\texttt{E$_3$}): adding a new workflow is a very
+  \note{Silv}{it seems to me that this metric cannot be calculated in a straighforward manner from the diagrams.. is that correct? if it can, give an example of how this is seen in the diagram}
+\item New workflow \texttt{E$_3$}: number of interactions required
+to integrate a new workflow in the architecture. Adding a new workflow is a very
   common operation that does not require modifying software components
-  or interactions. We measure the difficulty to integrate a new
-  workflow by counting the number of interactions required
-  to integrate a new workflow in the architecture, assuming
+  or interactions, assuming
   that the engine type and version required to execute this workflow
   are already available.
+  \note{sil}{again here it would be good to indicate how this is seen in the diagram}
 \item New infrastructure (\texttt{E$_4$}): number of interactions or
-  components to modify to integrate a new type of infrastructure in
-  the architecture. Adding a new infrastructure allows to provide more
-  computing or storage power, to access specific types of resources
-  (e.g., GPUs, clouds), or to enforce execution policies (e.g.,
+  components affected for the integration of a new type of infrastructure in
+  the architecture. Adding a new infrastructure typically aims at providing more
+  computing or storage power, enabling access to specific types of resources
+  (e.g., GPUs, clouds), or  enforcing execution policies (e.g.,
   constrain data to remain in a particular network domain).
@@
 \end{itemize}
-The 4 metrics are summed to obtain a global index that measures the
+The four metrics are summed to obtain a global score that measures the
 difficulty to extend the architecture.
@@
 involve several metrics in practice. For instance, adding a new type
-of engine may help integrate new infrastructures when interactions
+of engine may help integrating new infrastructures when interactions
 \texttt{b} and \texttt{c$_2$} are already present for the new engine
-and infrastructure.  \closednote{Rafael}{Adding a new engine type may
+and the infrastructure.  \closednote{Rafael}{Adding a new engine type may
   also allow run in different infrastructures. Maybe it is worth to

### 2016-05-13 paper.tex
done scalability

@@
 
-\emph{Scalability} corresponds to the ability of the architecture to
-cope with high workloads. It is measured by counting the potential
-scalability issues in the architectures, i.e. the missing scalability
-features. Features are evaluated using a 3-level metric: 0 means that
+\emph{Scalability} corresponds to the ability of an architecture to
+cope with high workloads through the following mechanisms: starting multiple engine instances, distributing engines and task scheduling. 
+This is measured by assessing the potential
+scalability difficulties due to (partial) absence of a given feature in the architecture.  
+Features are evaluated using a 3-level metric: 0 means that
 the feature is very easy to enable, 1 means that it can be implemented
 but with some difficulty, and 2 means that the feature cannot
-realistically be implemented. \closednote{Naj}{how do you metricize
+ be implemented in a realistic scenario. \closednote{Naj}{how do you metricize
   "nonesensical amount fo effors?"}{Tristan}{You don't :) Rephrased to
@@
   'impossible' because nothing is impossible to implement if enough
-  effort is put.} Four different features are identified:
+  effort is put.} Three different features are identified:
 \begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
-\item Multiple engine instances (\texttt{S$_1$}): possibility to have
-  more than 1 engine instance in the architecture. Workflow engines
+\item Multiple engine instances (\texttt{S$_1$}): measures the possibility \note{sil}{are we measuring possibility or potential?
+i think the second, because it is hypothetical.
+possibility brings me the idea of it is possible or not - which would be zero to almost all these features} of having
+  more than one engine instance in the architecture simultaneously. Workflow engines
   may require important amounts of resources when several workflows,
   or large workflows, are executed. At some point, it may be required
-  for the science gateway to distribute the load among several
-  engines. \texttt{S$_1$}=0 when adding a new engine instance is a
+  for the science gateway to means that adding a new engine instance is a
   fully automated process, i.e. the workflow engine only has to be
@@
   of auto-scaling mechanism can be implemented to control the number
-  of engine instances in the architecture. \texttt{S$_1$}=1 when
+  of engine instances in the architecture. \texttt{S$_1$}=1 means that
   adding a new engine instance requires some form of manual
-  intervention (e.g., two-factor authentication), which prevents easy
-  implementations of elastic engines. \texttt{S$_1$}=2 when new engine
+  intervention (e.g., two-factor authentication \note{sil}{what is this?}), which prevents easy
+  implementations of elastic engines. \texttt{S$_1$}=2 means that a new engine
   instances cannot be added.
-\item Distributed engines (\texttt{S$_2$}): possibility to distribute
+\item Distributed engines (\texttt{S$_2$}): measures the possibility of distributing
   the execution \emph{of a single workflow} among different engine
@@
   the architecture rather than these of the workflow
-  engine. \texttt{S$_2$}=0 when distributed engines are possible in
-  the architecture, \texttt{S$_2$}=1 when they require specific
+  engine. \texttt{S$_2$}=0 means that distributed engines are possible in
+  the architecture, and \texttt{S$_2$}=1 that they require specific
   developments in the workflow engine.  \closednote{Rafael}{$S_2$
@@
     reflect this in the text.}
-\item Task scheduling (\texttt{S$_3$}): task scheduling is a difficult
-  issue that depends more on the implementation of specific algorithms
+\item Task scheduling (\texttt{S$_3$}): measures ???
+\note{sil}{need to be more specific about what is being called task scheduling here.
+also explain why it is difficult, or dont say anything if it is common knowledge}.
+Task scheduling typically depends more on the implementation of specific algorithms
   in the science gateway, workflow engine and infrastructure than on
@@
   scheduling problem by introducing additional software layers or
-  creating tasks with specific characteristics. \texttt{S$_3$}=0 when
+  creating tasks with specific characteristics. \texttt{S$_3$}=0 means that
   the architecture does not introduce any additional complexity to the
@@
 \end{itemize}
-The 3 metrics are summed to obtain a global measure of the scalability
-issues that are present in the architecture.
+These three metrics are summed to obtain a global measure of the scalability
+potential of the architecture.
 

### 2016-05-13 paper.tex
done up to before 3.2

@@
 \begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
-\item Meta-workflow (\texttt{O$_1$}): ability to describe
+\item Meta-workflow (\texttt{O$_1$}): measures the ability to describe
   meta-workflows from existing workflows. Meta-workflows offer an
   additional level of flexibility to build workflows from reusable
-  components. \texttt{O$_1$}=0 when the feature is available,
+  components. \texttt{O$_1$}=0 means that the feature is available, and
   \texttt{O$_1$}=1 otherwise.
+  \note{si}{why here we do not have the "possible to implement" situation?}
+
 \item Fine-grained debugging (\texttt{O$_2$}): availability of
@@
   required to properly troubleshoot workflow executions.
-  \texttt{O$_2$}=0 when such information is easy to access,
+  \texttt{O$_2$}=0 means that such information is easy to access, and
   \texttt{O$_2$}=1 otherwise.

### 2016-05-13 paper.tex
done 3.2

@@
 
+\note{sil}{since we are dreaming, it seems to me that keeping provenance properly is also something to consider.
+the architecture can hide some details, and make it nearly impossible to keep provenance without hard coded hacks across the layers/components - i think}
+
 %"A Formal Approach to Support Interoperability in Scientific
@@
 component in addition to the science gateway and infrastructure
-(\texttt{I$_1$=2}). It involves 5 interactions: \texttt{a$_1$},
+(\texttt{I$_1$=2}) and it involves 5 interactions: \texttt{a$_1$},
 \texttt{a$_2$}, \texttt{b}, \texttt{c$_1$} and \texttt{c$_2$}
@@
 \paragraph{Extensibility} Integrating a new type of workflow engine
-requires to modify the science gateway as well as interactions
+requires modifying the science gateway as well as interactions
 \texttt{b} and \texttt{c$_2$} (\texttt{E$_1$}=3). Updating a workflow
@@
 \paragraph{Scalability} Adding a new engine instance requires a new
-instance of the science gateway, which is in general not possible
-(\texttt{S$_1$}=2). Specific IT setups such as load-balancing between
-web server instances might be used to deal with this
-issue. \closedanswerednote{Tristan}{What-about load-balancing between web
+instance of the complete science gateway
+%, which is in general not possible
+\note{sil}{why not? i agree that S=2, but i do not agree that it would be impossible. you say how below}
+(\texttt{S$_1$}=2). 
+Specific IT setups such as load-balancing between
+web server instances might be used to create new gateway instances, distributing the users, but would not solve the problem for the workload of a single workflow or user.
+\closedanswerednote{Tristan}{What-about load-balancing between web
   server instances?}{Marc-e}{Would that really be considered a real
@@
 Distributed engines are not available by default
-(\texttt{S$_2$}=1). The scheduling of tasks on the infrastructure is
+(\texttt{S$_2$}=1). 
+The scheduling of tasks on the infrastructure is
 as complex as in any other architecture since the workflow engine
 might implement any kind of scheduling policy (\texttt{S$_3$}=0).
+\note{sil}{so same as the others = "easy"? i think this requires more explanation. maybe we should introduce a score for "dont know"?}
 
 \paragraph{Specific features} Meta-workflows are not supported by default
-(\texttt{O$_1$}=1).  Debugging is not an issue since the science
+(\texttt{O$_1$}=1).  
+\note{sil}{why not? if the engine is ws-pgrade it can. i really think that meta-workflows is a property exclusively of the engine. you can see that in fig 2.V interaction *6}
+Debugging is not an issue since the science
 gateway can retrieve any information from the workflow engine directly
@@
 
+\note{sil}{should we then also thank SURF foundation (for the dutchgrid), FP7 (for SCI_BUS, SHIWA and ER-flow), and COMMIT/ and HUvA HPCN for the funding for the amc science gateway activities.
+My participation has been minimal, and the amc gateways are not discussed in detail (my fault), so maybe it would be an overkill}
+
 \section*{References}

### 2016-05-13 paper.tex
done 3.3

@@
 (\texttt{O$_1$}=1).  
-\note{sil}{why not? if the engine is ws-pgrade it can. i really think that meta-workflows is a property exclusively of the engine. you can see that in fig 2.V interaction *6}
+\note{sil}{why not? if the engine is ws-pgrade it does. i really think that meta-workflows (and nested workflows) is a property exclusively of the engine. you can see that in fig 2.V interaction *6}
 Debugging is not an issue since the science
@@
 \paragraph{Extensibility} Adding a new type of workflow engine
-requires to implement the corresponding workflow service, to modify
-interaction \texttt{d}, and to implement interactions \texttt{b} and
+requires implementing the corresponding workflow service,  modifying
+interaction \texttt{d}, and  implementing interactions \texttt{b} and
 \texttt{$c_2$} (\texttt{E$_1$}=4). New engine versions can be added by
@@
 
-\paragraph{Scalability} The service architecture supports multiple
+\paragraph{Scalability} The service architecture in principle supports multiple
 engine instances through multiple workflow services. In VIP for
@@
 load-balancing mechanism is available that sends new workflow
-executions to the engine instance which has the least active
+executions to the engine instance that has the least active
 executions. To avoid ``black-hole'' syndromes created by failing
-engine instances, engine instances are automatically disabled when
+ instances, engine instances are automatically disabled when
 workflows cannot be submitted to them. Adding a new engine instance,
@@
 engines is usually not possible unless the workflow engine has
-specific abilities (\texttt{S$_2$}=1). The scheduling of tasks on the
+specific abilities (\texttt{S$_2$}=1). 
+\note{sil}{is this score for VIP?
+i think that the score depends on the implementation of the science gateway, which could adopt some elastic mechanism within.
+so we should not say s_2=1 for the architecture}
+The scheduling of tasks on the
 infrastructure is as complex as in any other architecture since the
@@
 (\texttt{S$_3$}=0).
+\note{sil}{see comment in section 3.1}
 
 \paragraph{Specific features} Meta-workflows are not supported by default
-(\texttt{O$_1$}=1). Fine-grained debugging information is usually easy
+(\texttt{O$_1$}=1). 
+\note{sil}{see comment in section 3.1}
+Fine-grained debugging information is usually easy
 to obtain since the workflow service provides direct access to the
 engine (\texttt{O$_2$}=0).
+\note{sil}{this is really not the case. the fact that the wf service is a separate component that hides information is a big problem for debugging at the science gateway level. 
+we had a paper about this:
+	@inproceedings{olabarriaga2014scientific,
+		Author = {Olabarriaga, Silvia and Pierantoni, Gabrielle and Taffoni, Giuliano and Sciacca, Eva and Jaghoori, Mahdi and Korkhov, Vladimir and Castelli, Giuliano and Vuerli, Claudio and Becciani, Ugo and Carley, Eoin and others},
+		Booktitle = {e-Science (e-Science), 2014 IEEE 10th International Conference on},
+		Organization = {IEEE},
+		Pages = {298--305},
+		Title = {Scientific Workflow Management--For Whom?},
+		Year = {2014}}
+	}
 

### 2016-05-13 paper.tex
done 3.4

@@
 \paragraph{Robustness} No component is specific to workflow execution
-(\texttt{R$_1$}=0), and only interaction \texttt{e} is
+(\texttt{R$_1$}=0), and only interaction \texttt{e} is necessary
 (\texttt{R$_2$}=1)\closedanswerednote{sorina}{I understand why this is 1
@@
 \paragraph{Extensibility} Integrating a new type of workflow engine
-requires to develop interaction \texttt{e} and to install the engine
+requires developing interaction \texttt{e} and  installing the engine
 on the infrastructure (\texttt{E$_1$}=2). Updating an engine version
@@
 \texttt{a$_1$} and \texttt{a$_2$} (\texttt{E$_3$}=2). Adding a new
-infrastructure requires to update interactions \texttt{a$_1$},
+infrastructure requires updating interactions \texttt{a$_1$},
 \texttt{b} and \texttt{c} in the science gateway (\texttt{E$_4$}=3).
@@
 more complex than in the other approaches due to the special role of
-the task that executes the workflow engine (\texttt{S$_3$}=1). Indeed,
+the task that executes the workflow engine (\texttt{S$_3$}=1). 
+\note{sil}{i have the impression that the comments below are about robustness, and not so much about scalability.
+i also think that the level of detail here is much larger than in other 3.X sections.
+finally, i wonder whether what is explained here (which i must admit i could not follow completely in the scope of this evaluation or architectures) is valid only for this architecture in particular.
+i have the impression that this argument is more general }
+Indeed,
 the reliability of this task is critical since all the sub-tasks in
@@
 \paragraph{Specific features} Meta-workflows are not available
-(\texttt{O$_1$}=1).  Obtaining fine-grained information about workflow
+(\texttt{O$_1$}=1).  
+\note{sil}{see comment in section 3.1}
+Obtaining fine-grained information about workflow
 tasks is not straightforward since the science gateway has no
@@
 integrated as a task (\texttt{O$_2$}=1).
+\note{sil}{mm, i think the case here is the same as for service. or actually i would give this one a better score, because there are less components to cross}
 

### 2016-05-13 paper.tex
done 3.6

@@
 
-\paragraph{Extensibility} Adding a new engine type requires to wrap
-the engine in the agent and to update interactions \texttt{b} and
+\paragraph{Extensibility} Adding a new engine type requires wrapping
+the engine into the agent and updating interactions \texttt{b} and
 \texttt{c$_2$} (\texttt{E$_1$=3}). Updating the version of an engine
 is transparent (\texttt{E$_2$=0}) since it only requires updating the
-agent that is a component dedicated to the engine. Integrating a new workflow is
+agent, which is a component dedicated to the engine. Integrating a new workflow is
 done through interactions \texttt{a$_1$} and \texttt{a$_2$}
@@
 \paragraph{Scalability} New engine instances only require new agents,
-which is easily automated (\texttt{S$_1$=0}) and by design very
+which can be easily automated (\texttt{S$_1$=0}) and by design very
 suitable for elastic computing.\closednote{Naj}{I feel the concept of elastic computing should be mentioned in introductory sections}{Tristan}{You are right, this is now fixed.} For instance, auto-scaling rules can
@@
 \paragraph{Specific features} Meta-workflows are not available by default
-(\texttt{O$_1$=1}).  Accessing debugging information is not likely to
+(\texttt{O$_1$=1}).  
+\note{sil}{see comment in section 3.1}
+Accessing debugging information is not likely to
 be an issue since the workflow pool could implement specific functions
 for that (\texttt{O$_2$=0}).
+\note{sil}{i think the case here is the same as for service. or actually i would give this one a worse score, because there is one additional component - the pool - to cross}
 

### 2016-05-13 paper.tex
now it is true, done with 3.6

@@
 (twice), \texttt{c$_2$} (twice) and \texttt{d} (twice) also
-are (\texttt{R$_2$=6}).
+are necessary (\texttt{R$_2$=6}).
 \closednote{Marc-e}{I understand both points. However, complexity is
@@
 \paragraph{Extensibility} Adding a new type of \emph{parent} engine
-requires to implement the corresponding service, to implement
+requires implementing the corresponding service,  implementing
 interactions \texttt{b} and \texttt{c$_2$} in the parent engine, and
-to implement interaction \texttt{d} in the science gateway and in the
+implementing interaction \texttt{d} in the science gateway and in the
 parent service (\texttt{E$_1$=5}). Adding a new type of \emph{child}
-engine only requires to implement the corresponding service, to
-develop interactions \texttt{b} and \texttt{c$_2$} in the child
-engine, and to implement interaction \texttt{d} in the parent service
+engine only requires implementing the corresponding service, 
+developing interactions \texttt{b} and \texttt{c$_2$} in the child
+engine, and implementing interaction \texttt{d} in the parent service
 (\texttt{E$_1$=4}). We use \texttt{E$_1$=4.5} in
@@
 interaction \texttt{a$_1$} and \texttt{a$_2$}
-(\texttt{E$_3$=2}). Adding a new infrastructure requires to
-re-implement interactions \texttt{a$_1$}, \texttt{b} and
+(\texttt{E$_3$=2}). Adding a new infrastructure requires 
+re-implementing interactions \texttt{a$_1$}, \texttt{b} and
 \texttt{c$_2$} twice, and interaction \texttt{c$_1$} once so that it

### 2016-05-13 paper.tex
done section 3 for now

@@
 requires the same components as for service invocation
-(\texttt{I$_1$}=3). It requires an additional \texttt{g} interaction
-for workflow import (\texttt{I$_1$}=7).
+(\texttt{I$_1$}=3), and an additional \texttt{g} interaction (\texttt{I$_1$}=7).
 
 \paragraph{Robustness} Since workflow conversion is not involved in
-the execution (it is an offline process), metrics are as in the
+the execution (it is an offline process), the scores are the same as for the
 service architecture (\texttt{R$_1$=1}, \texttt{R$_2$=3}).
@@
 aims at supporting more workflows, we consider that it only requires
-to re-implement interaction \texttt{g} in this architecture  (\texttt{E$_1$=1}). Note,
-however, that implementing interaction \texttt{g} can require very
-substantial work depending on the complexity of the language used by
+re-implementing interaction \texttt{g} in this architecture  (\texttt{E$_1$=1}). Note,
+however, that implementing interaction \texttt{g} can require
+substantial work depending on the complexity of the workflow language used by
 the new engine. Similarly, adding a
@@
 \paragraph{Scalability}  Since workflow conversion is not involved in
-the execution (it is an offline process), metrics are as in the
+the execution (it is an offline process), the score is the same as for the
 service architecture (\texttt{S$_1$=1},
@@
 by connecting workflows in the language used in the science gateway
-(\texttt{O$_1$=0}). Debugging information is accessible as in the
+(\texttt{O$_1$=0}). 
+\note{sil}{i dont understand this about meta-workflows.
+translation does not imply composition, right?
+if the science gateway is allowed to implement workflow composition, then this is probably the case in other architectures as well?}
+Debugging information is accessible as in the
 service invocation architecture (\texttt{O$_2$=0}).
+\note{sil}{see notes 3.2}
 

### 2016-05-13 paper.tex
completed

@@
   $m$ is the metric value, $m_{\mathrm{min}}$ and $m_{\mathrm{max}}$
-  are the minimal and maximal values among all architectures; (2) the RGB hexadecimal color code of the cell
+  are the minimum and maximum values for all architectures; (2) the RGB hexadecimal color code of the cell
   is \texttt{\#99XX99}, where X=\texttt{round}(F-6m') (\texttt{round} rounds a number to the nearest integer). \closednote{Tristan}{Use 6. Round to the nearest integer.}{Tristan}{Fixed.}}
@@
   words?! :-) }{Tristan}{OK!} depending on the type of targeted
-infrastructure. Non-reliable infrastructures \closednote{Naj}{you mean by
+infrastructure. Non-reliable infrastructures\closednote{Naj}{you mean by
   the showstoppers, i.e. the nested workflows which can accommodate
@@
 invocation) are the only architectures that support meta-workflows;
-task encapsulation and nested workflows may complexify
+task encapsulation and nested workflows may hamper
 fine-grained debugging. \closednote{Rafael}{I would suggest to provide
@@
 architectures, based on the metrics in Table~\ref{table:evaluation}.
-Note that this analysis is only for illustration purposes since it
+Note that this analysis is only meant for illustration purposes, since it
 assumes that all criteria have equal weights while real systems would
@@
 service invocation) perform a bit better than tight integration,
-service invocation and pool. Nested workflows stands a bit behind for
+service invocation and pool. Nested workflows stand a bit behind for
 the reasons mentioned previously.\closednote{Rafael}{I would emphasize
@@
 extent.\closednote{Naj}{Here, you were bashing your article harshly,
-  so I rewrote it}{Tristan}{yup, thanks!}. In particular, the
+  so I rewrote it}{Tristan}{yup, thanks!} In particular, the
 presented architectures are abstract patterns that may be mixed
@@
 Moreover, all the interactions involved in the architectures were
-treated equally while some of them are obviously more complex than
+treated equally, whereas some of them are obviously more complex than
 others. For example, interaction \texttt{g} (workflow language
 conversion) is clearly more complex than interaction \texttt{d}
-(service invocation), and may create much more reliability issues as
-well. However, without entering into the details of a particular
+(service invocation), and may challenge reliability more strongly. However, without entering into the details of a particular
 system, quantification of the robustness or the amount of development
-associated with each interaction and software component will not be
+associated with each interaction and software component will hardly be
 precise.
@@
 system \closednote{Naj}{so are you discussing it here, or in the
-  future?}{Tristan}{It's out of scope, a complete paper in itself. Rephrased this part.} since this interaction may not be easily generalized to any
+  future?}{Tristan}{It's out of scope, a complete paper in itself. Rephrased this part.} since this interaction may not be easily generalizable to any
 workflow language. For instance, converting FSL, PSOM or Nipype
@@
 so on.
-
+\note{sil}{do you want to refer to the science gateway canvas paper here?
+	@inproceedings{shahand:2015ab,
+		Address = {Portland, OR, USA},
+		Author = {Shahand, Shayan and van Kampen, Antoine H. C. and Olabarriaga, S\'{\i}lvia D.},
+		Booktitle = {{Proceedings of the Science of Cyberinfrastructure: Research, Experience, Applications and Models}},
+		Series = {SCREAM'15},
+		Title = {{Science Gateway Canvas: A business reference model for Science Gateways}},
+		Year = {2015},
+	}
 \section{Conclusion}
@@
 review and compare software architectures used to integrate workflow
-engines in science gateways.\closednote{Naj}{instead of its kind,
+engines into science gateways.\closednote{Naj}{instead of its kind,
   explain what kind it is, e.g. systematic review of all

### 2016-05-13 paper.tex
done for now, really

@@
 \author[mcgill]{Pierre Rioux}
-
+\author[amc]{Si\'lvia D. Olabarriaga}
 \author[criugm]{Pierre Bellec}
@@
 \address[isi]{University of Southern California, Information Sciences Institute, Marina del Rey, CA, USA.}
+\address[amc]{Academic Medical Center of the University of Amsterdam, Department of Clinical Epidemiology, Biostatistics and Bioinformatics, Amsterdam, NL.}
 
 \begin{abstract}
+	\note{sil}{i find this a dry start. what about an intro sentence?
+		Workflow engines are critical for the efficient and transparent
+		exploitation of distributed infrastructures in the ecosystem of tools
+		and services offered by science gateways.}
   We investigate six software architectures commonly used to integrate
-  workflow engines in science gateways. In \emph{tight integration},
+  workflow engines into science gateways. In \emph{tight integration},
   the workflow engine shares software components with the science
@@
     they?}{Tristan}{Done, thanks}. Results provide insights for
-  science gateway architects and workflow engine developers. Tight
+  science gateway architects and workflow engine developers. 
+  \note{sil}{what are the insights for wf engine developers?. this is not clear in the discussion and conclusions. i think the focus lies more on the integration and not on one or the other} Tight
   integration and task encapsulation are the easiest to integrate and
@@
     architectures, following @pbellec's suggestion.}
+\note{sil{what about closing with this sentence from above:
+		These results provide insights for
+		science gateway architects and workflow engine developers. }
 \end{abstract}
@@
 
-\note{sil}{should we then also thank SURF foundation (for the dutchgrid), FP7 (for SCI_BUS, SHIWA and ER-flow), and COMMIT/ and HUvA HPCN for the funding for the amc science gateway activities.
+\note{sil}{should we then also thank the dutch crowd?
+	 We also acknowledge the Dutch national e-Infrastructure with the support of SURF Cooperative,  the Dutch national program COMMIT/ and the High Performance Computing and Networking (HPCN) Fund of the University of Amsterdam for their support to the science gateway actitivies at the AMC.
+	 We are also greatful to the financial support provided by FP7 E-INFRASTRUCTURE program for financial support to SCI_BUS, SHIWA and ER-flow projects.  
 	 My participation has been minimal, and the amc gateways are not discussed in detail (my fault), so maybe it would be an overkill}

### 2016-05-14 related.tex
some ideas for related work (in separate file)

\section{Related work}
\label{sec:related}

The early work \cite{yuTaxonomy} proposes a generic architecture for grid workflow systems that is based on the workflow reference model defined by the Workflow Management Coalition \cite{wfRefModel}. 
Deelman et al. \cite{deelmanFGCS} have further characterized the features of workflow systems and proposed a general architecture?? 
These two works, however, do not mention science gateways.

Shahand et al. \cite{shahand:2015ab} presents the Science Gateway Canvas, which is a business reference model where the most relevant functional components are organized into functional groups. Although workflow management is mentioned as a possible  component to coordinate distributed computations in science gateways, the paper does not comment on architecture to achieve this.

Olabarriaga et al. \cite{olabarriaga2014} presents a user-centered view of the ecosystem of science gateways with focus on workflow management. Their proposed layered system architecture includes gateway, workflow engine and distributed infrastructure components, similarly to the "service invocation" pattern. However other patterns are not identified.

@inproceedings{shahand:2015ab,
	Address = {Portland, OR, USA},
	Author = {Shahand, Shayan and van Kampen, Antoine H. C. and Olabarriaga, S\'{\i}lvia D.},
	Booktitle = {{Proceedings of the Science of Cyberinfrastructure: Research, Experience, Applications and Models}},
	Doi = {10.1145/2753524.2753527},
	Series = {SCREAM'15},
	Title = {{Science Gateway Canvas: A business reference model for Science Gateways}},
	Year = {2015},
	Bdsk-Url-1 = {http://dx.doi.org/10.1145/2753524.2753527}}

@inproceedings{olabarriaga2014,
	Author = {Olabarriaga, Silvia and Pierantoni, Gabrielle and Taffoni, Giuliano and Sciacca, Eva and Jaghoori, Mahdi and Korkhov, Vladimir and Castelli, Giuliano and Vuerli, Claudio and Becciani, Ugo and Carley, Eoin and others},
	Booktitle = {e-Science (e-Science), 2014 IEEE 10th International Conference on},
	Organization = {IEEE},
	Pages = {298--305},
	Title = {Scientific Workflow Management--For Whom?},
	Year = {2014}}

### 2016-05-14 related.tex
again related

@@
 However other patterns are not identified.
+
+@article{yuTaxonomy,
+	year={2005},
+	issn={1570-7873},
+	journal={Journal of Grid Computing},
+	volume={3},
+	number={3-4},
+	title={A Taxonomy of Workflow Management Systems for Grid Computing},
+	publisher={Springer Netherlands},
+	keywords={Grid computing; resource management; scheduling; taxonomy; workflow management},
+	author={J. Yu  and R. Buyya},
+	pages={171-200},
+	language={English}
+}
+
+@article{deelmanFGCS,
+	TITLE={Workflows and e-Science: An overview of workflow system features and capabilities}, 
+	author    = {E. Deelman and
+		D. Gannon and
+		M. S. Shields and
+		I. Taylor},
+	JOURNAL={Future Generation Computer Systems}, 
+	YEAR={2009},
+	VOLUME={25},
+	NUMBER={5},
+	PAGES={528-540},
+}
 

### 2016-05-14 paper.tex

@@
   the showstoppers, i.e. the nested workflows which can accommodate
-  distributed executions but are limited by scheduling constraints?}{Tristan}{Yes. I rephrased a bit this part.} ,
+  distributed executions but are limited by scheduling constraints?}{Tristan}{Yes. I rephrased a bit this part.},
 for example, could hardly cope with workflow engines being wrapped in
@@
 workflows (with service invocation) and workflow import (with service
-invocation) are the only architectures that support meta-workflows;
+invocation) are the only architectures that support meta-workflows. Besides,
 task encapsulation and nested workflows may hamper

### 2016-05-14 paper.tex
Some corrections following @SilviaOlabarriaga's comments.

@@
 \begin{abstract}
-	\note{sil}{i find this a dry start. what about an intro sentence?
+  \closednote{sil}{i find this a dry start. what about an intro sentence?
     Workflow engines are critical for the efficient and transparent
-		exploitation of distributed infrastructures in the ecosystem of tools
-		and services offered by science gateways.}
-  We investigate six software architectures commonly used to integrate
-  workflow engines into science gateways. In \emph{tight integration},
-  the workflow engine shares software components with the science
-  gateway. In \emph{service invocation}, the engine is isolated and
-  invoked through a specific software interface. In \emph{task
-    encapsulation}, the engine is wrapped as a computing task executed
-  on the infrastructure. In the \emph{pool model}, the engine is
-  bundled in an agent that connects to a central pool to fetch and
-  execute workflows. In \emph{nested workflows}, the engine is
-  integrated as a child process of another engine. In \emph{workflow
-    import}, the engine is integrated through workflow language
-  conversion. We describe and evaluate these architectures with
-  metrics for assessment of integration complexity, robustness,
-  extensibility, scalability and support for meta-workflows and
-  fine-grained debugging\closednote{Naj}{Please be precise, what are
-    they?}{Tristan}{Done, thanks}. Results provide insights for
-  science gateway architects and workflow engine developers. 
-  \note{sil}{what are the insights for wf engine developers?. this is not clear in the discussion and conclusions. i think the focus lies more on the integration and not on one or the other} Tight
+    exploitation of distributed infrastructures in the ecosystem of
+    tools and services offered by science gateways.}{Tristan}{Done, thanks.}  Workflow engines
+  are critical for the efficient and transparent exploitation of
+  distributed infrastructures in the ecosystem of tools and services
+  offered by science gateways.  We investigate six software
+  architectures commonly used to integrate workflow engines into
+  science gateways. In \emph{tight integration}, the workflow engine
+  shares software components with the science gateway. In
+  \emph{service invocation}, the engine is isolated and invoked
+  through a specific software interface. In \emph{task encapsulation},
+  the engine is wrapped as a computing task executed on the
+  infrastructure. In the \emph{pool model}, the engine is bundled in
+  an agent that connects to a central pool to fetch and execute
+  workflows. In \emph{nested workflows}, the engine is integrated as a
+  child process of another engine. In \emph{workflow import}, the
+  engine is integrated through workflow language conversion. We
+  describe and evaluate these architectures with metrics for
+  assessment of integration complexity, robustness, extensibility,
+  scalability and support for meta-workflows and fine-grained
+  debugging\closednote{Naj}{Please be precise, what are
+    they?}{Tristan}{Done, thanks}.\closednote{sil}{what are the insights for wf engine developers?. this is
+    not clear in the discussion and conclusions. i think the focus
+    lies more on the integration and not on one or the other}{Tristan}{Fixed by your next note.} Tight
   integration and task encapsulation are the easiest to integrate and
@@
   meta-workflows are only available in nested workflows and workflow
-  import. \closednote{Naj}{I find this abstract to be vague, it can be made
-    more concise and poignant}{Tristan}{Added a description of the
-    architectures, following @pbellec's suggestion.}
-\note{sil{what about closing with this sentence from above:
-		These results provide insights for
-		science gateway architects and workflow engine developers. }
+  import. \closednote{Naj}{I find this abstract to be vague, it can be
+    made more concise and poignant}{Tristan}{Added a description of
+    the architectures, following @pbellec's suggestion.}
+  \closednote{sil}{what about closing with this sentence from above: These
+      results provide insights for science gateway architects and
+      workflow engine developers. }{Tristan}{Done, thanks.}These results provide insights for
+		science gateway architects and developers.
 \end{abstract}
@@
 integration complexity, robustness, extensibility, scalability and
-other specific features. These are based on science gateway properties introduced in Section~\ref{sec:intro_sg}. Criteria break down to specific metrics where
-\emph{lower value indicates better performance}. For each criterion, a
-global score is computed by summing up the individual scores for each metric. We
-ensure that the different metrics used to calculate the global score for a criterion are comparable, so that they can be combined meaningfully. The criteria and metrics are
+other specific features. These are based on science gateway properties
+introduced in Section~\ref{sec:intro_sg}. Criteria break down to
+specific metrics where \emph{lower value indicates better
+  performance}. Colors in Table~\ref{table:evaluation} represent
+normalized scores, as explained in the Table caption. For each
+criterion, a global score is computed by summing up the individual
+scores for each metric. We ensure that the different metrics used to
+calculate the global score for a criterion are comparable, so that
+they can be combined meaningfully. The criteria and metrics are
 explained hereafter.
 
-\note{Sil}{normalization is explained in caption tab 2. i wonder whether something should be said here, as questions about the range of the score immediately pops to mind..}
+\closednote{Sil}{normalization is explained in caption tab 2.
+i wonder whether something should be said here, as questions about the range of the score immediately pops to mind..}{Tristan}{Added a sentence about it.}
 
@@
 
-\note{Sil}{i was wondering whether it would be more clear to present the metrics like this:
-\item \texttt{I$_1$}: total number of components, and
-
-now the parenthesis are next to text that does not really match the metric definition (exception in this case) could also use boldface
-	}
+\closednote{Sil}{i was wondering whether it would be more clear to present the metrics like this:
+\item \texttt{I$_1$}: total number of components, and now the parenthesis are next to text that does not really match the metric definition (exception in this case) could also use boldface}{Tristan}{Fixed, following discussion in https://github.com/glatard/FGCS-works-special-issue/pull/20}
 Science gateway \emph{integration complexity} is measured through the number of software components and their interactions. It breaks
@@
 \begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
-\item Total number of components (\texttt{I$_1$}), and
-\item Total number of interactions (\texttt{I$_2$}).
+\item Total components (\texttt{I$_1$}): total number of components in
+  the architecture.
+\item Total interactions (\texttt{I$_2$}): total number of
+  interactions in the architecture.
 \end{itemize}
@@
 \begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
-\item Number of specific components
-involved in workflow executions (\texttt{R$_1$}). The Science
+\item Specific components  (\texttt{R$_1$}): number of specific components
+involved in workflow executions. The Science
 gateway component, for instance, is \emph{not} specific to workflow execution
@@
 user data, etc., and
-\item Number of specific interactions
-involved in workflow executions (\texttt{R$_2$}).
-\note{Sil}{add here an example of an interaction that is not, for example, data transfers?}
+\item Specific interactions (\texttt{R$_2$}): number of specific
+  interactions involved in workflow executions. Data control
+  (interaction \texttt{c}), for instance, is \emph{not} specific to
+  workflow execution since it is also used by the science gateway to
+  transfer user data. \closednote{Sil}{add here an example of an interaction
+    that is not, for example, data transfers?}{Tristan}{Done.}
 \end{itemize}
@@
 \begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
-\item New engine \texttt{E$_1$}: number of interactions or
-  components that need to be modified to integrate a new type of workflow engine in
-  the architecture. Workflow engines belong to different types when
-  they cannot be invoked using the same interface. 
-  \note{Si}{new interface = new interaction? maybe better to use one term only here. Maybe also refer to how this shows on figure 2}
-\item Engine version upgrade \texttt{E$_2$}: number of interactions or
-  components modified to integrate a new version of a workflow engine
-  in the architecture, assuming that another version of the same
-  engine type is already available. Different versions of a workflow
-  engine share the same interface, i.e. they can be invoked using the
-  same software. When this is not the case, the different versions are
-  considered as different engine types. Components are not treated
+\item New engine (\texttt{E$_1$}): number of interactions or
+  components that need to be modified to integrate a new type of
+  workflow engine in the architecture. Workflow engines belong to
+  different types when they cannot be invoked using the same
+  implementation of the interactions.
+  \closednote{Si}{new interface = new interaction? maybe better to use
+    one term only here. Maybe also refer to how this shows on figure
+    2}{Tristan}{Fixed the interface vs. interaction issue.}
+\item Engine version upgrade (\texttt{E$_2$}): number of interactions
+  or components modified to integrate a new version of a workflow
+  engine in the architecture, assuming that another version of the
+  same engine type is already available. Different versions of a
+  workflow engine can be invoked using the same interaction
+  implementation(s). When this is not the case, the different versions
+  are considered as different engine types. Components are not treated
   equally regarding engine version updates. Components whose only
@@
   release of the science gateway, gaining administrative privileges on
-  the infrastructure, etc.
-  \note{Silv}{it seems to me that this metric cannot be calculated in a straighforward manner from the diagrams.. is that correct? if it can, give an example of how this is seen in the diagram}
-\item New workflow \texttt{E$_3$}: number of interactions required
-to integrate a new workflow in the architecture. Adding a new workflow is a very
-  common operation that does not require modifying software components
-  or interactions, assuming
-  that the engine type and version required to execute this workflow
-  are already available.
-  \note{sil}{again here it would be good to indicate how this is seen in the diagram}
+  the infrastructure, etc. In practice, E$_2$=0 when the workflow
+  engine is wrapped in a service or in an agent, and E$_2$=1 when it is
+  wrapped in another component.  \closednote{Silv}{it seems to me that this
+    metric cannot be calculated in a straighforward manner from the
+    diagrams.. is that correct? if it can, give an example of how this
+    is seen in the diagram}{Tristan}{Done}
+\item New workflow (\texttt{E$_3$}): number of interactions required
+  to integrate a new workflow in the architecture. Adding a new
+  workflow is a very common operation that does not require modifying
+  software components or interactions, assuming that the engine type
+  and version required to execute this workflow are already
+  available. In most cases, adding a new workflow requires only
+  interactions \texttt{a$_1$} and \texttt{a$_2$}, but \texttt{g}
+  is required as well when workflow conversion is used. \closednote{sil}{again
+    here it would be good to indicate how this is seen in the diagram}{Tristan}{done}
 \item New infrastructure (\texttt{E$_4$}): number of interactions or
@@
 \begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
-\item Multiple engine instances (\texttt{S$_1$}): measures the possibility \note{sil}{are we measuring possibility or potential?
+\item Multiple engine instances (\texttt{S$_1$}): measures the potential \closednote{sil}{are we measuring possibility or potential?
 i think the second, because it is hypothetical.
-possibility brings me the idea of it is possible or not - which would be zero to almost all these features} of having
+possibility brings me the idea of it is possible or not - which would be zero to almost all these features}{Tristan}{Fixed everywhere} of having
   more than one engine instance in the architecture simultaneously. Workflow engines
@@
   fully automated process, i.e. the workflow engine only has to be
-  started. In this case, elastic engines are possible, i.e. some kind
-  of auto-scaling mechanism can be implemented to control the number
+  started. In this case, elastic engines can be implemented through some kind
+  of auto-scaling mechanism to control the number
   of engine instances in the architecture. \texttt{S$_1$}=1 means that
   adding a new engine instance requires some form of manual
-  intervention (e.g., two-factor authentication \note{sil}{what is this?}), which prevents easy
-  implementations of elastic engines. \texttt{S$_1$}=2 means that a new engine
+  intervention, which prevents easy
+  implementations of elastic engines. \texttt{S$_1$}=2 means that new engine
   instances cannot be added.
-\item Distributed engines (\texttt{S$_2$}): measures the possibility of distributing
+\item Distributed engines (\texttt{S$_2$}): measures the potential of distributing
   the execution \emph{of a single workflow} among different engine
@@
   the architecture rather than these of the workflow
-  engine. \texttt{S$_2$}=0 means that distributed engines are possible in
+  engine. \texttt{S$_2$}=0 means that distributed engines are enabled by
   the architecture, and \texttt{S$_2$}=1 that they require specific
@@
     reflect this in the text.}
-\item Task scheduling (\texttt{S$_3$}): measures ???
-\note{sil}{need to be more specific about what is being called task scheduling here.
-also explain why it is difficult, or dont say anything if it is common knowledge}.
-Task scheduling typically depends more on the implementation of specific algorithms
-  in the science gateway, workflow engine and infrastructure than on
-  the architecture used to integrate the workflow engine in the
-  science gateway. Some architectures, however, complicate the task
-  scheduling problem by introducing additional software layers or
-  creating tasks with specific characteristics. \texttt{S$_3$}=0 means that
-  the architecture does not introduce any additional complexity to the
-  scheduling problem, and \texttt{S$_3$}=1 otherwise.
+\item Task scheduling (\texttt{S$_3$}): measures the complexity
+  introduced by the architecture regarding task scheduling on
+  the infrastructure. \closednote{sil}{need to be more specific about what
+    is being called task scheduling here. also explain why it is
+    difficult, or dont say anything if it is common knowledge}{Tristan}{OK, fixed.} Task
+  scheduling typically depends more on the implementation of specific
+  algorithms in the science gateway, workflow engine and
+  infrastructure than on the architecture used to integrate the
+  workflow engine in the science gateway. Some architectures, however,
+  complicate the task scheduling problem by introducing additional
+  software layers or creating tasks with specific
+  characteristics. \texttt{S$_3$}=0 means that the architecture does
+  not introduce any additional complexity to the scheduling problem,
+  and \texttt{S$_3$}=1 otherwise.
 \end{itemize}
@@
   additional level of flexibility to build workflows from reusable
-  components. \texttt{O$_1$}=0 means that the feature is available, and
+  components. \texttt{O$_1$}=0 means that meta-workflows are enabled by the architecture, and
   \texttt{O$_1$}=1 otherwise.
@@
 
-\paragraph{Scalability} The service architecture in principle supports multiple
-engine instances through multiple workflow services. In VIP for
-instance, this feature has been available from release 1.17. A basic
-load-balancing mechanism is available that sends new workflow
-executions to the engine instance that has the least active
-executions. To avoid ``black-hole'' syndromes created by failing
- instances, engine instances are automatically disabled when
-workflows cannot be submitted to them. Adding a new engine instance,
-however, requires manual intervention to declare the new instance in
-the science gateway (\texttt{S$_1$}=1). Consequently, elastic engines
-are difficult to implement because they require a mechanism to update
-the science gateway configuration when a new engine instance is
+\paragraph{Scalability} The service architecture in principle supports
+multiple engine instances through multiple workflow services. Adding a
+new engine instance, however, requires manual intervention to declare
+the new instance in the science gateway
+(\texttt{S$_1$}=1). Consequently, elastic engines are difficult to
+implement because they require a mechanism to update the science
+gateway configuration when a new engine instance is
 available. Distributing the execution of a single workflow in multiple
 engines is usually not possible unless the workflow engine has
-specific abilities (\texttt{S$_2$}=1). 
-\note{sil}{is this score for VIP?
-i think that the score depends on the implementation of the science gateway, which could adopt some elastic mechanism within.
-so we should not say s_2=1 for the architecture}
-The scheduling of tasks on the
-infrastructure is as complex as in any other architecture since the
-workflow engine might implement any kind of scheduling policy
-(\texttt{S$_3$}=0).
-\note{sil}{see comment in section 3.1}
+specific abilities (\texttt{S$_2$}=1).  \answerednote{sil}{is this score for
+  VIP? i think that the score depends on the implementation of the
+  science gateway, which could adopt some elastic mechanism within. so
+  we should not say s_2=1 for the architecture}{Tristan}{Removed VIP example.
+Score definition still has to be modified according to the discussion in @SilviaOlabarriaga's PR (\#20).} The scheduling of
+tasks on the infrastructure is as complex as in any other architecture
+since the workflow engine might implement any kind of scheduling
+policy (\texttt{S$_3$}=0).  \note{sil}{see comment in section 3.1}
 
@@
 the task that executes the workflow engine (\texttt{S$_3$}=1). 
-\note{sil}{i have the impression that the comments below are about robustness, and not so much about scalability.
+\closednote{sil}{i have the impression that the comments below are about robustness, and not so much about scalability.
 i also think that the level of detail here is much larger than in other 3.X sections.
 finally, i wonder whether what is explained here (which i must admit i could not follow completely in the scope of this evaluation or architectures) is valid only for this architecture in particular.
-i have the impression that this argument is more general }
+i have the impression that this argument is more general }{Tristan}{Rejected :) See discussion in PR #20.}
 Indeed,
@@
 child engine) (\texttt{S$_1$=1}). Similarly, elastic engines are
-difficult to achieve. Distributed engines are
-possible, through meta-workflows (\texttt{S$_2$=0}). Task scheduling
+difficult to achieve. Distributed engines can be implemented through meta-workflows (\texttt{S$_2$=0}). Task scheduling
 is more complex than in other architectures though, due to the fact
@@
 \paragraph{Specific features}
-Meta-workflows are possible, which is one of the main interest of this
+Meta-workflows can be implemented, which is one of the main interest of this
 architecture (\texttt{O$_1$=0}).  Debugging is difficult because the
@@
 so on.
-\note{sil}{do you want to refer to the science gateway canvas paper here?
-	@inproceedings{shahand:2015ab,
-		Address = {Portland, OR, USA},
-		Author = {Shahand, Shayan and van Kampen, Antoine H. C. and Olabarriaga, S\'{\i}lvia D.},
-		Booktitle = {{Proceedings of the Science of Cyberinfrastructure: Research, Experience, Applications and Models}},
-		Series = {SCREAM'15},
-		Title = {{Science Gateway Canvas: A business reference model for Science Gateways}},
-		Year = {2015},
-	}
+\closednote{sil}{do you want to refer to the science gateway canvas paper here?}{Tristan}{Done (by you) in the related work section.}
+
+\section{Related work}
+\label{sec:related}
+
+The early work~\cite{yuTaxonomy} proposes a generic architecture for
+grid workflow systems that is based on the workflow reference model
+defined by the Workflow Management Coalition~\cite{wfRefModel}.
+Deelman et al. \cite{deelman2009workflows} have further characterized
+the features of workflow systems and proposed a general architecture??
+These two works, however, do not mention science gateways.
+
+Shahand et al.~\cite{shahand:2015ab} presents the Science Gateway
+Canvas, which is a business reference model where the most relevant
+functional components are organized into functional groups. Although
+workflow management is mentioned as a possible component to coordinate
+distributed computations in science gateways, the paper does not
+comment on architecture to achieve this.
+
+Olabarriaga et al.~\cite{olabarriaga2014} presents a user-centered
+view of the ecosystem of science gateways with focus on workflow
+management. Their proposed layered system architecture includes
+gateway, workflow engine and distributed infrastructure components,
+similarly to the "service invocation" pattern. However other patterns
+are not identified.
+
 \section{Conclusion}
@@
 
-This work is in the scope of the LABEX PRIMES (ANR-11- LABX-0063) of
-Universit\'e de Lyon, within the program ``Investissements d’Avenir''
-(ANR-11-IDEX-0007) operated by the French National Research Agency
-(ANR). This work also falls into the scope of the scientific topics of
-the French National Grid Institute (IdG). We thank the site
-administrators of the European Grid Initiative and the GGUS support
-for their help related to the VIP platform. The CBRAIN team is grateful for the
-computing cycles, storage, and support obtained from Compute Canada
-(\url{https://computecanada.ca}) and platform development program from CANARIE
- (\url{http://www.canarie.ca/}). The integration between PSOM and CBRAIN was supported by a Brain 
-Canada Platform Support Grant, as well as the Canadian Consortium on
-Neurodegeneration in Aging (CCNA), through a grant from the Canadian Institute
-of Health Research and funding from several partners. This work has been made possible with the support of the Irving Ludmer Family Foundation and
-the Ludmer Centre for Neuroinformatics and Mental Health. \closedanswerednote{Marc-e}{Would it be proper to thank EGI?
-  Compute Canada? CANARIE?}{Tristan}{Added EGI and Compute Canada. Not sure if
-  Canarie should be added. What about Calcul QC?}{Marc-e}{CQ, no. CANARIE, I'd
-  say yes since they funded a major chunk of CBRAIN's development.}
-
-\note{sil}{should we then also thank the dutch crowd?
+This work has been made possible with the support of the Irving Ludmer
+Family Foundation and the Ludmer Centre for Neuroinformatics and
+Mental Health. The integration between PSOM and CBRAIN was supported
+by a Brain Canada Platform Support Grant, as well as the Canadian
+Consortium on Neurodegeneration in Aging (CCNA), through a grant from
+the Canadian Institute of Health Research and funding from several
+partners.  This work is in the scope of the LABEX PRIMES (ANR-11-
+LABX-0063) of Universit\'e de Lyon, within the program
+``Investissements d’Avenir'' (ANR-11-IDEX-0007) operated by the French
+National Research Agency (ANR). This work also falls into the scope of
+the scientific topics of the French National Grid Institute (IdG). The
+VIP team thanks the site administrators of the European Grid
+Initiative and the GGUS support for their help related to the VIP
+platform. The CBRAIN team is grateful for the computing cycles,
+storage, and support obtained from Compute Canada
+(\url{https://computecanada.ca}) and platform development program from
+CANARIE (\url{http://www.canarie.ca/}). We also acknowledge the Dutch
+national e-Infrastructure with the support of SURF Cooperative, the
+Dutch national program COMMIT/ and the High Performance Computing and
+Networking (HPCN) Fund of the University of Amsterdam for their
+support to the science gateway actitivies at the AMC.  We are also
+grateful to the financial support provided by FP7 E-INFRASTRUCTURE
+program for financial support to SCI-BUS, SHIWA and ER-flow
+projects. \closedanswerednote{Marc-e}{Would it be proper to thank EGI?
+  Compute Canada? CANARIE?}{Tristan}{Added EGI and Compute Canada. Not
+  sure if Canarie should be added. What about Calcul QC?}{Marc-e}{CQ,
+  no. CANARIE, I'd say yes since they funded a major chunk of CBRAIN's
+  development.} \closednote{sil}{should we then also thank the dutch crowd?
 	 We also acknowledge the Dutch national e-Infrastructure with the support of SURF Cooperative,  the Dutch national program COMMIT/ and the High Performance Computing and Networking (HPCN) Fund of the University of Amsterdam for their support to the science gateway actitivies at the AMC.
 	 We are also greatful to the financial support provided by FP7 E-INFRASTRUCTURE program for financial support to SCI_BUS, SHIWA and ER-flow projects.
-	 My participation has been minimal, and the amc gateways are not discussed in detail (my fault), so maybe it would be an overkill}
+My participation has been minimal, and the amc gateways are not discussed in detail (my fault), so maybe it would be an overkill}{Tristan}{done.}
 

### 2016-05-14 paper.tex
Removed all the closed notes (#21)

@@
 
-\title{Software architectures to integrate workflow engines in science gateways\closednote{Naj}{'what' software architectures?
-what is their qualifier, adjective?}{Tristan}{There is no qualifier except 'for integrating workflow engines in science gatewats'.
-They are just software architectures :) The point of the paper is to describe them properly.
-It's a general formulation to cover any kind of software architecture that is used to integrated workflow engines in science gateways.
-Does it make sense?}}
+\title{Software architectures to integrate workflow engines in science gateways}
 
@@
 \begin{abstract}
-  \closednote{sil}{i find this a dry start. what about an intro sentence?
   Workflow engines are critical for the efficient and transparent
   exploitation of distributed infrastructures in the ecosystem of
-    tools and services offered by science gateways.}{Tristan}{Done, thanks.}  Workflow engines
-  are critical for the efficient and transparent exploitation of
-  distributed infrastructures in the ecosystem of tools and services
-  offered by science gateways.  We investigate six software
-  architectures commonly used to integrate workflow engines into
-  science gateways. In \emph{tight integration}, the workflow engine
-  shares software components with the science gateway. In
+  tools and services offered by science gateways.  We investigate six
+  software architectures commonly used to integrate workflow engines
+  into science gateways. In \emph{tight integration}, the workflow
+  engine shares software components with the science gateway. In
   \emph{service invocation}, the engine is isolated and invoked
@@
   scalability and support for meta-workflows and fine-grained
-  debugging\closednote{Naj}{Please be precise, what are
-    they?}{Tristan}{Done, thanks}.\closednote{sil}{what are the insights for wf engine developers?. this is
-    not clear in the discussion and conclusions. i think the focus
-    lies more on the integration and not on one or the other}{Tristan}{Fixed by your next note.} Tight
-  integration and task encapsulation are the easiest to integrate and
-  the most robust. Extensibility is equivalent in most
-  architectures. The pool model is the most scalable one and
+  debugging. Tight integration and task encapsulation are the easiest
+  to integrate and the most robust. Extensibility is equivalent in
+  most architectures. The pool model is the most scalable one and
   meta-workflows are only available in nested workflows and workflow
-  import. \closednote{Naj}{I find this abstract to be vague, it can be
-    made more concise and poignant}{Tristan}{Added a description of
-    the architectures, following @pbellec's suggestion.}
-  \closednote{sil}{what about closing with this sentence from above: These
-      results provide insights for science gateway architects and
-      workflow engine developers. }{Tristan}{Done, thanks.}These results provide insights for
-		science gateway architects and developers.
+  import. These results provide insights for science gateway
+  architects and developers.
 \end{abstract}
@@
 computing infrastructures, in particular
-clusters\closednote{Tristan}{Reverted from Marc's 'cluster
-  scheduler'}{}{}, to execute tasks at a high throughput. Some of
+clusters, to execute tasks at a high throughput. Some of
 these tools also support advanced features such as provenance
-tracking, \closednote{Naj}{should we not mention PMP which is used for
-  CIVET?}{Tristan}{Done, as discussed.} or  redundancy detection across
+tracking, or  redundancy detection across
 analyses to avoid re-computation. A wide array of workflows
@@
 \emph{infrastructure} (local server, cluster, grid or cloud) based on
-a workflow description (a.k.a \emph{workflow}), \closednote{Naj}{I think and is redundant here, you can replace is with a comma}{Tristan}{yes!} using input data
+a workflow description (a.k.a \emph{workflow}), using input data
 that may consist of files, database entries or simple parameter
 values. Although simplistic, this definition covers both e-Science workflow systems and
-domain-specific pipeline systems.\closednote{Marc-e}{To remain consistent with above descriptions}{Tristan}{yes!} 
+domain-specific pipeline systems.
 Some workflow engines, 
@@
 components to implement and execute
-applications. \closednote{Naj}{provide examples?}{Tristan}{Examples
-  are mentioned in the last paragraph of this section. Moved it next.}
+applications.
 
@@
 transparency, reliability, traceability and
-reproducibility. \closednote{Marc-Etienne}{I find that something is
-  weak in this last paragraph. Robustness will help with transparency,
-  sure, but it seems only a small facet.  Also, the text seem to
-  timidly favor architectures with less components/interactions
-  without much explanation... Complex systems can be robust, but at
-  greater costs...  Here is one: "Architectures showing reduced system
-  complexity will greatly facilitate the implementation effort towards
-  robust interactions which, in turn, have a positive impact on
-  characteristics such as gateway predictability, transparence,
-  general reliability and results traceability and reproducibility;
-  characteristics that are also key to a good user
-  experience."}{Tristan}{Agreed, fixed as
-  suggested.}\closednote{Tristan}{Work on the this sentence again.}{Tristan}{done.}
+reproducibility.
 
@@
 instruments, or social tools among users.
-\closednote{Naj}{I think using etc after such as, for example and e.g.
-is not correct.
---sorry for being pedantic}{Tristan}{You are completely right, this is now fixed.}
 
@@
 
-\closednote{Sil}{i changed above a few occurrences of "diagrammed".
-i know it is correct, but i am not used to seeing it.
-this is a matter of style, just forget my changes below then}{Tristan}{That's perfect, thanks!}
-
 Architectures to integrate workflow engines in science gateways are
@@
 summarizes the classification of a few systems by architecture.
-
-\closednote{Marc-e}{A
-  few things bother me a bit with the notation fig., making it harder
-  to assimilate 1- The interactions are padded in the middle of other
-  stuff (Administrator, Data) 2- The Software component example says
-  "Science gateway", should it not be empty?  3- The red arrow to
-  explain the use of red; it looks like a very specific type of
-  interactions instead of the desired "any interactions or components
-  can be red meaning...", especially with 'a' and 'd' which the reader
-  does not know the meaning yet.  Maybe a little red arrow and square
-  with "Red used when specific to workflow execution"?  I don't have a
-  great solution...  Red has special meaning, maybe remove it from the
-  other pics (workflow eng, tasks) as it creates lots of visual
-  interference when looking for red?  4- Order in text jumps all over
-  the fig notations. Explanations are given for, in order: Components,
-  workflow engine, interactions, abstract interactions, red color,
-  tasks, data (no Administrator?)...  Could re-ordering the fig fix
-  this? A figure figure legend could go a long way too (are there some
-  in this journal?)  }{Tristan}{Marc-E, is Figure 1 better now?}
-\closednote{Naj}{I propose a table of definitions.}{Tristan}{Added
-  information in Figure 1 and removed the corresponding text.}
 
@@
   use it. The infrastructure might offer various data storage backends
-  with heterogeneous interfaces. \closedanswerednote{Marc-Etienne}{this -c2- can be also trigerred
-    by the user}{Tristan}{No, it can't: this is by definition an interaction between the workflow engine and the infrastructure.
-It is used to transfer data to/from tasks.
-I reworded a bit, is it clearer now?}
-    {Marc-e}{I meant, "using c1", a user in some platforms (CBRAIN), can trigger
-    some operations across the infrastructure, as in c2... not just up/download data.
-    And I clearly see now that this is in fact the case in III, c is not differentiated. Closing this thread for good.}
+  with heterogeneous interfaces. 
   Tools and services such as JSAGA~\cite{reynaud2010uniform} or Data Avenue~\cite{hajnal2014data}
@@
   among the 5 workflow systems in the SHIWA project: Askalon, MOTEUR,
-  Taverna, Triana and WS-PGRADE.  \closednote{Silvia}{Should we add here the
-    4?  and for translation among the 5 workflow systems in the SHIWA
-    project: Askalon, MOTEUR, Taverna, Triana and WS-PGRADE)}{Tristan}{Yes, done!}
+  Taverna, Triana and WS-PGRADE.  
 \end{enumerate}
@@
 gateway usually share a database where applications, users and other
-resources are stored\closednote{sorina}{VIP and Moteur also share the
-  WorkflowsDB}{Tristan}{Yes, you are right. This is an example where
-  such abstract architectures may not completely apply in real
-  life. See comment about ``blended'' architectures in the
-  discussion. In case of VIP, the sharing of the DB is more an
-  implementation detail rather than a strong architectural pattern: we
-  could avoid sharing the DB if we wanted.}.  In this model, task and
+resources are stored.  In this model, task and
 data control are both initiated from the science gateway. Interactions
@@
 \subsection{Task encapsulation}
-\closednote{Marc-e}{May I propose
-  another name here? Either "Task-centric" or "Task Oriented". My
-  reasoning is that the WE is just another task, as well as the
-  sub-tasks themselves.}{Tristan}{I agree: sub-tasking is not the
-  integration model, it is the process that engines use when they
-  are integrated as tasks. I think 'task encapsulation' is more
-  consistent  with 'service invocation' than 'task-oriented'}
 
@@
 \texttt{c$_2$} are both covered by \texttt{c}.
-\closednote{Sil}{is this next sentence relevant?}{Tristan}{I just wanted to  justify why these two interactions can be grouped but I guess you're right (I removed the sentence.}%In practice, both
-%\texttt{c$_1$} and \texttt{c$_2$} can be implemented using the same
-%pieces of code.
 
@@
   submission (\texttt{1} \& \texttt{2}), workflow status retrieval
-  (\texttt{0} \& \texttt{3})\closednote{sorina}{I suppose letters come
-    here from the initial pool-model figure; it's a bit puzzling when
-    mixed with letters from Fig 2}{Tristan}{Fixed.}, and workflow
-  retrieval (\texttt{13} \& \texttt{14}). \closednote{Marc-e}{Maybe
-    easier to follow if the following would be in fig legend instead
-    of here, as for 2 previous figures}{Tristan}{fixed} Interaction
+  (\texttt{0} \& \texttt{3}), and workflow
+  retrieval (\texttt{13} \& \texttt{14}). Interaction
   \texttt{f} is implemented through 2 types of calls: workflow
@@
 from Galaxy~\cite{goecks2010galaxy} to WS-PGRADE/gUSE.
-\closednote{Tristan}{Add description of a real system}{Tristan}{Would
-  require to reproduce a figure from another paper. The concept seems
-  simple enough to not require a figure.}
 
@@
 explained hereafter.
-
-\closednote{Sil}{normalization is explained in caption tab 2.
-i wonder whether something should be said here, as questions about the range of the score immediately pops to mind..}{Tristan}{Added a sentence about it.}
 
@@
   are the minimum and maximum values for all architectures; (2) the RGB hexadecimal color code of the cell
-  is \texttt{\#99XX99}, where X=\texttt{round}(F-6m') (\texttt{round} rounds a number to the nearest integer). \closednote{Tristan}{Use 6. Round to the nearest integer.}{Tristan}{Fixed.}}
+  is \texttt{\#99XX99}, where X=\texttt{round}(F-6m') (\texttt{round} rounds a number to the nearest integer). }
 \label{table:evaluation}
@@
 
-\closednote{Sil}{i was wondering whether it would be more clear to present the metrics like this:
-\item \texttt{I$_1$}: total number of components, and now the parenthesis are next to text that does not really match the metric definition (exception in this case) could also use boldface}{Tristan}{Fixed, following discussion in https://github.com/glatard/FGCS-works-special-issue/pull/20}
 Science gateway \emph{integration complexity} is measured through the number of software components and their interactions. It breaks
@@
   workflow execution since it is also used by the science gateway to
-  transfer user data. \closednote{Sil}{add here an example of an interaction
-    that is not, for example, data transfers?}{Tristan}{Done.}
+  transfer user data. 
 \end{itemize}
@@
   implementation of the interactions.
-  \closednote{Si}{new interface = new interaction? maybe better to use
-    one term only here. Maybe also refer to how this shows on figure
-    2}{Tristan}{Fixed the interface vs. interaction issue.}
  \item Engine version upgrade (\texttt{E$_2$}): number of interactions
@@
   engine is wrapped in a service or in an agent, and E$_2$=1 when it is
-  wrapped in another component.  \closednote{Silv}{it seems to me that this
-    metric cannot be calculated in a straighforward manner from the
-    diagrams.. is that correct? if it can, give an example of how this
-    is seen in the diagram}{Tristan}{Done}
+  wrapped in another component.  
 \item New workflow (\texttt{E$_3$}): number of interactions required
@@
   interactions \texttt{a$_1$} and \texttt{a$_2$}, but \texttt{g}
-  is required as well when workflow conversion is used. \closednote{sil}{again
-    here it would be good to indicate how this is seen in the diagram}{Tristan}{done}
+  is required as well when workflow conversion is used. 
 \item New infrastructure (\texttt{E$_4$}): number of interactions or
@@
 The four metrics are summed to obtain a global score that measures the
-difficulty to extend the architecture.
-\closednote{Naj}{I would rather compare each metric separately, this woudl also give a sense of how different architectures compare across metrics}{Tristan}{Metrics are compared separately too, see Table 2 and discussion.} Note that some extensions may
+difficulty to extend the architecture. Note that some extensions may
 involve several metrics in practice. For instance, adding a new type
@@
 \texttt{b} and \texttt{c$_2$} are already present for the new engine
-and the infrastructure.  \closednote{Rafael}{Adding a new engine type may
-  also allow run in different infrastructures. Maybe it is worth to
-  put a sentence to clarify that the metric account for it
-  separately.}{Tristan}{It will not *allow* because interaction c1
-  would still need to be developed. But it will definitely facilitate
-  it. I added a note on this after the description of the metrics.}
+and the infrastructure.
 
@@
 but with some difficulty, and 2 means that the feature cannot
- be implemented in a realistic scenario. \closednote{Naj}{how do you metricize
-  "nonesensical amount fo effors?"}{Tristan}{You don't :) Rephrased to
-  'cannot realistically be implemented'. At some point it's a matter
-  of common sense and experience to draw the line between what is
-  easily implemented and what is more tricky. I would like to avoid
-  'impossible' because nothing is impossible to implement if enough
-  effort is put.} Three different features are identified:
+ be implemented in a realistic scenario. Three different features are identified:
 \begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
-\item Multiple engine instances (\texttt{S$_1$}): measures the potential \closednote{sil}{are we measuring possibility or potential?
-i think the second, because it is hypothetical.
-possibility brings me the idea of it is possible or not - which would be zero to almost all these features}{Tristan}{Fixed everywhere} of having
+\item Multiple engine instances (\texttt{S$_1$}): measures the potential of having
   more than one engine instance in the architecture simultaneously. Workflow engines
@@
   the architecture, and \texttt{S$_2$}=1 that they require specific
-  developments in the workflow engine.  \closednote{Rafael}{$S_2$
-    would be 2 if the paradigm does not allow to distribute the
-    orchestration? e.g., workflow executions implemented within MPI
-    jobs, in Pegasus PMC and dispel4Py.}{Tristan}{The problem is that
-    it is very difficult to define ``does not allow to''. For
-    instance, why couldn't you hack Pegasus PMC so much that it
-    becomes possible to distribute workflow executions in several MPI
-    jobs coordinated by an MPI job? Moreover, S2=2 is not used so I'd
-    prefer to leave it like this, unless you have a suggestion to
-    reflect this in the text.}
+  developments in the workflow engine. 
 \item Task scheduling (\texttt{S$_3$}): measures the complexity
   introduced by the architecture regarding task scheduling on
-  the infrastructure. \closednote{sil}{need to be more specific about what
-    is being called task scheduling here. also explain why it is
-    difficult, or dont say anything if it is common knowledge}{Tristan}{OK, fixed.} Task
+  the infrastructure. Task
   scheduling typically depends more on the implementation of specific
@@
 web server instances might be used to create new gateway instances, distributing the users, but would not solve the problem for the workload of a single workflow or user.
-\closedanswerednote{Tristan}{What-about load-balancing between web
-  server instances?}{Marc-e}{Would that really be considered a real
-  architectural feature or an IT hack?}{Tristan}{I agree, we should
-  keep IT setups out of the picture. Depending on the internals of the
-  science gateway, such hacks may or may not be feasible. I added a
-  sentence to mention this possibility, without changing the score.}
 Distributed engines are not available by default
@@
 (\texttt{R$_1$}=0), and only interaction \texttt{e} is necessary
-(\texttt{R$_2$}=1)\closedanswerednote{sorina}{I understand why this is 1
-  (and why interactions c and b are not counted here), but I wonder
-  whether we can simply ignore these interactions because they are not
-  specific to workflow execution. As far as robustness is concerned,
-  the transfers have to be successful anyway...}{Tristan}{We cannot
-  just ignore these interactions in the architecture because they are
-  required for the system to work. The transfers have to be successful
-  but data movements in the workflow will use the same code as data
-  movements done in the science gateway, which will make this
-  architecture more robust than service for instance. In VIP, for
-  instance, the code used in the GASW wrapper to transfer data (c2) is
-  completely independent from the ``file transfer'' module in the
-  platform (c1), which is detrimental to robustness}{Tristan}{Closed,
-  see github issue \#7.}.
+(\texttt{R$_2$}=1).
 
@@
 the task that executes the workflow engine (\texttt{S$_3$}=1). 
-\closednote{sil}{i have the impression that the comments below are about robustness, and not so much about scalability.
-i also think that the level of detail here is much larger than in other 3.X sections.
-finally, i wonder whether what is explained here (which i must admit i could not follow completely in the scope of this evaluation or architectures) is valid only for this architecture in particular.
-i have the impression that this argument is more general }{Tristan}{Rejected :) See discussion in PR #20.}
 Indeed,
@@
 as selection of wrong batch queues on clusters or task termination due
-to exceeded quotas. \closednote{Naj}{The lay reader wonders why?  an
-  example like the one offered for interdependencies can
-  help}{Tristan}{Done, thanks.} Finally, the interdependencies between
+to exceeded quotas. Finally, the interdependencies between
 the workflow task and its sub-tasks may create deadlocks when there is
@@
 computing resources available on this infrastructure, which makes such
-deadlocks unlikely to happen. \closednote{Naj}{so is this the solution to
-  fixing the above mentioned limitations with
-  scalability?}{Tristan}{Unfortunately this is not a solution. We just
-  say that this situation is unlikely to happen which is true.}
+deadlocks unlikely to happen.
 
@@
 which can be easily automated (\texttt{S$_1$=0}) and by design very
-suitable for elastic computing.\closednote{Naj}{I feel the concept of elastic computing should be mentioned in introductory sections}{Tristan}{You are right, this is now fixed.} For instance, auto-scaling rules can
+suitable for elastic computing. For instance, auto-scaling rules can
 be implemented to start new agents when the workload in the science
@@
 \texttt{d} (twice).
-\closedanswerednote{Rafael}{I like the idea of what has been discussed in
-  the `integration effort', however I think the name may lead to
-  misinterpretations. For example, according to Table 2, it requires
-  more effort to integrate in a Tight architecture than in a
-  service-based architecture. Considering the principles of software
-  engineer, it is simpler to integrate/maintain a system using
-  decoupled services. Maybe it should be renamed to something related
-  to 'dependencies'.}{Tristan}{In our scope I'm not sure that it's
-  simpler to integrate a service architecture than a tightly
-  integrated one. But I take your point. 'Dependencies' is not very
-  suitable either. What about 'development effort'?}{Tristan}{Replaced
-  'Integration effort' with 'global complexity', as discussed in
-  github issue \#8.}
 
@@
 are necessary (\texttt{R$_2$=6}).
-\closednote{Marc-e}{I understand both points. However, complexity is
-  addressed in Robustness as well... I liked the idea of development effort. It's a tough call.}{Tristan}{Maybe 'integration challenge' is a good compromise.}
 
@@
 (\texttt{E$_1$}), which makes them useful architectures for science
-gateways that do not focus on a particular engine.\closednote{Naj}{what is
-  interesting?}{Tristan}{Rephrased, thanks.}  However, task encapsulation, workflow import (with
+gateways that do not focus on a particular engine.  However, task encapsulation, workflow import (with
 service invocation) and tight integration perform worse than the
@@
 perform the same overall. The global scalability score, however,
-should not conceal \closednote{Naj}{what do you mean by
-  mask?}{Tristan}{I guess 'hide' or 'conceal' are better
-  words. Fixed.} the unique characteristics of architectures regarding
+should not conceal the unique characteristics of architectures regarding
 this criterion. Nested workflows are the only architecture that can
@@
 critical in some cases. At the same time, the scheduling constraints
-created by task encapsulation and nested workflows may become problematic \closednote{Naj}{please don't use these from-film-marketing
-  words?! :-) }{Tristan}{OK!} depending on the type of targeted
-infrastructure. Non-reliable infrastructures\closednote{Naj}{you mean by
-  the showstoppers, i.e. the nested workflows which can accommodate
-  distributed executions but are limited by scheduling constraints?}{Tristan}{Yes. I rephrased a bit this part.},
+created by task encapsulation and nested workflows may become problematic  depending on the type of targeted
+infrastructure. Non-reliable infrastructures,
 for example, could hardly cope with workflow engines being wrapped in
@@
 task encapsulation and nested workflows may hamper
-fine-grained debugging. \closednote{Rafael}{I would suggest to provide
-  some more examples when this could be interesting.}{Tristan}{Done.}
+fine-grained debugging. 
 The availability of fine-grained debugging information may also be
@@
 service invocation and pool. Nested workflows stand a bit behind for
-the reasons mentioned previously.\closednote{Rafael}{I would emphasize
-  that this analysis is only for illustration purposes, since it does
-  not account for different weights in metrics.}{Tristan}{Done.}
+the reasons mentioned previously.
 \begin{table*}
@@
 metrics hinders the complexity of real systems to some
-extent.\closednote{Naj}{Here, you were bashing your article harshly,
-  so I rewrote it}{Tristan}{yup, thanks!} In particular, the
+extent. In particular, the
 presented architectures are abstract patterns that may be mixed
-together in actual systems.\closednote{Naj}{something about "theoretical
-  pattern", and "blending" is jarring, are these common words in the
-  computer sci context?}{Tristan}{Better now?} The distinction between tight integration and
+together in actual systems. The distinction between tight integration and
 service invocation, for instance, may not always be that clear in
@@
 conversion) requires particular attention when implementing a real
-system \closednote{Naj}{so are you discussing it here, or in the
-  future?}{Tristan}{It's out of scope, a complete paper in itself. Rephrased this part.} since this interaction may not be easily generalizable to any
+system since this interaction may not be easily generalizable to any
 workflow language. For instance, converting FSL, PSOM or Nipype
@@
 so on.
-\closednote{sil}{do you want to refer to the science gateway canvas paper here?}{Tristan}{Done (by you) in the related work section.}
 
@@
 pros and cons of all the presented architectures based on these metrics.
-\closedanswerednote{Rafael}{One thing that is missing in the
-  conclusion or the limitations section is to state that this work
-  does not intend to quantify the amount of work required to develop
-  science gateways using a specific architecture, but the goal is to
-  highlight what are the main components and interactions for each
-  architecture, and the main advantages and drawbacks. This should
-  also be stated in the introduction.}{Tristan}{This is in
-  contradiction with the metric on 'integration effort'. What do you
-  think?}{Tristan}{Fixed, as discussed in github \#8.}
 
@@
 review and compare software architectures used to integrate workflow
-engines into science gateways.\closednote{Naj}{instead of its kind,
-  explain what kind it is, e.g. systematic review of all
-  ...}{Tristan}{Done.} So far, the literature on science gateways and
+engines into science gateways. So far, the literature on science gateways and
 workflow engines has focused on the description of particular systems,
@@
 program for financial support to SCI-BUS, SHIWA and ER-flow
-projects. \closedanswerednote{Marc-e}{Would it be proper to thank EGI?
-  Compute Canada? CANARIE?}{Tristan}{Added EGI and Compute Canada. Not
-  sure if Canarie should be added. What about Calcul QC?}{Marc-e}{CQ,
-  no. CANARIE, I'd say yes since they funded a major chunk of CBRAIN's
-  development.} \closednote{sil}{should we then also thank the dutch crowd?
-	 We also acknowledge the Dutch national e-Infrastructure with the support of SURF Cooperative,  the Dutch national program COMMIT/ and the High Performance Computing and Networking (HPCN) Fund of the University of Amsterdam for their support to the science gateway actitivies at the AMC.
-	 We are also greatful to the financial support provided by FP7 E-INFRASTRUCTURE program for financial support to SCI_BUS, SHIWA and ER-flow projects.
-My participation has been minimal, and the amc gateways are not discussed in detail (my fault), so maybe it would be an overkill}{Tristan}{done.}
+projects.
 

### 2016-05-14 paper.tex
Up to Evaluation

@@
 architectures, based on system-independent representations of their
-main components and interactions. It is informed by our experince in
+main components and interactions. It is informed \note{vladimir}{is this a good word for this context, 'informed'? maybe 'based on'?} by our experience in
 the development and sustained operation of the CBRAIN~\cite{SHER-14}
@@
 workflow engines, science gateways and infrastructures. In
-Section~\ref{sec:architectures}, we describe six architectures within
+Section~\ref{sec:architectures}, we describe six architectures \note{vladimir}{why is it exactly six?
+are these architectures the ones that happened to be implemented during various projects, or is it a full set of all possibilities (I believe not)?
+I think, it must be stated clearly why the focus is given to these particular ones - we are reviewing and evaluating the approaches that have been used already in our projects} within
 a consistent framework that underlines the functional interactions
@@
 evaluated with metrics measuring integration complexity, robustness,
-extensibility, scalability and other specific features. We finally
+extensibility, scalability and other specific features \note{vladimir}{can we explain why exactly these metrics are used but not anything else?
+I think, we need to justify that these ones are especially important for the SG case; it would be nice to refer to some related research on software architecture metrics, Google gives about 2 million pages on this, where you can find metrics like cohesion, depth of scenario and so on}.
+We finally
 compare the architectures and highlight recommendations for science
@@
 Triana~\cite{taylor2007triana}, VisTrails~\cite{callahan2006managing}
-and WS-PGRADE~\cite{Kacsuk2012}. Such workflow engines usually
+and WS-PGRADE~\cite{Kacsuk2012} \note{vladimir}{As a past developer of it, I would suggest to refer to WS-VLAM as well, despite it is not very famous, if you do not mind :)}.
+Such workflow engines usually
 describe applications in a high-level language with specific data and
@@
   (\texttt{a$_1$}) the programs used in the workflow are installed on
-  the infrastructure, which may require administrative privileges;
+  the infrastructure, which may require administrative privileges \note{vladimir}{or bundled in packages considering all needed dependencies than can be deployed on the fly during the workflow execution};
   (\texttt{a$_2$}) the workflow is configured in the science gateway
@@
 format conversion from a native format to the science gateway
-format. Workflow import is usually an offline process that is not
+format.
+\note{vladimir}{Are we talking only about importing workflows from other wf languages?
+I would suggest to modify the title of the section to 'Workflow conversion and import' then, as some gateways use the term 'import' to import their own native workflows, like WS-PGRADE, and using import only in the context of converting workflows can be confusing} Workflow import is usually an offline process that is not
 involved in workflow execution. A few systems have implemented
@@
 covered, since they do not stem directly from the software
-architecture. Robustness is measured here as a consequence of global
+architecture. 
+\note{vladimir}{I believe, proper error handling through the different levels of wf engine/gateway must be especially mentioned even if not especially covered.
+One of serious issues we faced while working with gateways at the AMC, in particular with gUse/WS-PGRADE, was that when it failed sometimes it was completely not clear what was the problem exactly.
+So we might consider discussing not only the number of software components but how explicit and clear interactions are between them; how transparently we can monitor the execution through all the levels and thus detect and prevent possible malfunction}
+Robustness is measured here as a consequence of global
 complexity, since complex architectures tend to be more prone to

### 2016-05-14 paper.tex
including section 3

@@
   or large workflows, are executed. At some point, it may be required
-  for the science gateway to means that adding a new engine instance is a
+  for the science gateway to means \note{vladimir}{is this the right word here, 'to means'?} that adding a new engine instance is a
   fully automated process, i.e. the workflow engine only has to be
@@
   \note{si}{why here we do not have the "possible to implement" situation?}
+  \note{vladimir}{I agree with Silvia, we should have the intermediate values as well, since meta-workflows might be straightforwardly supported by the gateway, or they are still possible to implement but only with some workarounds.
+Like it was with SHIWA: first 6 months coarsed-grained interoperability looked more like a workaround till it was better integrated and supported by WS-PGRADE itself}
 
@@
   \texttt{O$_2$}=0 means that such information is easy to access, and
-  \texttt{O$_2$}=1 otherwise.
+  \texttt{O$_2$}=1 otherwise. \note{vladimir}{otherwise - does it mean the fine-grained info is available but hard to access or not available it all? the latter happens as well}
 \end{itemize}
@@
 (\texttt{R$_2$}=1).
+\note{vladimir}{there is a large piece of text in Scalability about reliability of the workflow engine task which on my opinion suits Robustness better}
 
@@
 task fails. The workflow task is also longer than all its sub-tasks,
-which increases its chances of failure. In addition, task parameters,
+which increases its chances of failure.\note{vladimir}{isn't this about robustness?} In addition, task parameters,
 for instance estimated walltime, are more difficult to estimate for
@@
 \texttt{d} (twice).
+\note{vladimir}{and if we have a meta-workflow with 3 and more sub-workflows the score will be even higher, is it a fair calculation for an architecture?
+we did not increase the scores for possible meta-workflows supported by wf engines in other cases}
 

### 2016-05-14 paper.tex

@@
 Triana~\cite{taylor2007triana}, VisTrails~\cite{callahan2006managing}
-and WS-PGRADE~\cite{Kacsuk2012} \note{vladimir}{As a past developer of it, I would suggest to refer to WS-VLAM as well, despite it is not very famous, if you do not mind :)}.
+and WS-PGRADE~\cite{Kacsuk2012} \note{vladimir}{As a past developer of it, I would suggest to refer to WS-VLAM~\cite{wsvlam} as well, despite it is not very famous, but it played an important role for Amsterdam e-science workflows}.
 Such workflow engines usually
@@
 \note{vladimir}{there is a large piece of text in Scalability about reliability of the workflow engine task which on my opinion suits Robustness better}
+\note{vladimir}{in addition, the claim of very good robustness of task encapsulation seems questionable to me as in this model you delegate the critical task of controlling workflow execution to, let's say, some 'random' node on the grid that you have no control of, not sure about its performance, connectivity and reliability and so on; however, considering the models and dependencies presented in fig.2 this evaluation of robustness metric may be correct}
 
@@
 workflows (with service invocation) and workflow import (with service
-invocation) are the only architectures that support meta-workflows. Besides,
+invocation) are the only architectures that support meta-workflows. \note{vladimir}{As mentioned earlier by Silvia, this might not be exactly the case, depending on the point of view} Besides,
 task encapsulation and nested workflows may hamper
@@
 are not identified.
+\note{vladimir}{@silvia, does it make sense to mention workflow classification efforts based on workflow motifs and so on?}
 

### 2016-05-14 paper.tex
Minor updates

--- 2016-05-14 paper.tex
@@
 \author[mcgill]{Pierre Rioux}
-\author[amc]{Si\'lvia D. Olabarriaga}
+\author[amc]{S\'ilvia D. Olabarriaga}
 \author[criugm]{Pierre Bellec}
@@
 architectures, based on system-independent representations of their
-main components and interactions. It is informed \note{vladimir}{is this a good word for this context, 'informed'? maybe 'based on'?} by our experience in
+main components and interactions. It is informed by our experince in
 the development and sustained operation of the CBRAIN~\cite{SHER-14}
@@
 workflow engines, science gateways and infrastructures. In
-Section~\ref{sec:architectures}, we describe six architectures \note{vladimir}{why is it exactly six?
-are these architectures the ones that happened to be implemented during various projects, or is it a full set of all possibilities (I believe not)?
-I think, it must be stated clearly why the focus is given to these particular ones - we are reviewing and evaluating the approaches that have been used already in our projects} within
+Section~\ref{sec:architectures}, we describe six architectures within
 a consistent framework that underlines the functional interactions
@@
 evaluated with metrics measuring integration complexity, robustness,
-extensibility, scalability and other specific features \note{vladimir}{can we explain why exactly these metrics are used but not anything else?
-I think, we need to justify that these ones are especially important for the SG case; it would be nice to refer to some related research on software architecture metrics, Google gives about 2 million pages on this, where you can find metrics like cohesion, depth of scenario and so on}.
-We finally
+extensibility, scalability and other specific features. We finally
 compare the architectures and highlight recommendations for science
@@
 Triana~\cite{taylor2007triana}, VisTrails~\cite{callahan2006managing}
-and WS-PGRADE~\cite{Kacsuk2012} \note{vladimir}{As a past developer of it, I would suggest to refer to WS-VLAM as well, despite it is not very famous, if you do not mind :)}.
-Such workflow engines usually
+and WS-PGRADE~\cite{Kacsuk2012}. Such workflow engines usually
 describe applications in a high-level language with specific data and
@@
   (\texttt{a$_1$}) the programs used in the workflow are installed on
-  the infrastructure, which may require administrative privileges \note{vladimir}{or bundled in packages considering all needed dependencies than can be deployed on the fly during the workflow execution};
+  the infrastructure, which may require administrative privileges;
   (\texttt{a$_2$}) the workflow is configured in the science gateway
@@
 format conversion from a native format to the science gateway
-format.
-\note{vladimir}{Are we talking only about importing workflows from other wf languages?
-I would suggest to modify the title of the section to 'Workflow conversion and import' then, as some gateways use the term 'import' to import their own native workflows, like WS-PGRADE, and using import only in the context of converting workflows can be confusing} Workflow import is usually an offline process that is not
+format. Workflow import is usually an offline process that is not
 involved in workflow execution. A few systems have implemented
@@
 covered, since they do not stem directly from the software
-architecture. 
-\note{vladimir}{I believe, proper error handling through the different levels of wf engine/gateway must be especially mentioned even if not especially covered.
-One of serious issues we faced while working with gateways at the AMC, in particular with gUse/WS-PGRADE, was that when it failed sometimes it was completely not clear what was the problem exactly.
-So we might consider discussing not only the number of software components but how explicit and clear interactions are between them; how transparently we can monitor the execution through all the levels and thus detect and prevent possible malfunction}
-Robustness is measured here as a consequence of global
+architecture. Robustness is measured here as a consequence of global
 complexity, since complex architectures tend to be more prone to

### 2016-05-15 paper.tex
Major updates following comments by @SilviaOlabarriaga and @vkorkhov

@@
 \author[mcgill]{Najmeh Khalili-Mahani}
+\author[stpetersburg]{Vladimir Korkhov}
 \author[criugm]{Pierre-Olivier Quirion}
@@
 \address[amc]{Academic Medical Center of the University of Amsterdam, Department of Clinical Epidemiology, Biostatistics and Bioinformatics, Amsterdam, NL.}
+\address[stpetersburg]{St. Petersburg State University, Russia.}
 
@@
   workflows. In \emph{nested workflows}, the engine is integrated as a
-  child process of another engine. In \emph{workflow import}, the
+  child process of another engine. In \emph{workflow conversion}, the
   engine is integrated through workflow language conversion. We
@@
   most architectures. The pool model is the most scalable one and
-  meta-workflows are only available in nested workflows and workflow
-  import. These results provide insights for science gateway
-  architects and developers.
+  meta-workflows are only available in task encapsulation, nested
+  workflows and workflow conversion. These results provide insights
+  for science gateway architects and developers.
 \end{abstract}
@@
 architectures, based on system-independent representations of their
-main components and interactions. It is informed by our experince in
-the development and sustained operation of the CBRAIN~\cite{SHER-14}
-and VIP~\cite{GLAT-13} science gateways for medical image analysis
-during the past 7 years, as well as by lessons learned from several
-science gateway and workflow projects such as
-SHIWA\footnote{\url{http://www.shiwa-workflow.eu}}.  In this section
-we provide background information and definitions of 
-workflow engines, science gateways and infrastructures. In
-Section~\ref{sec:architectures}, we describe six architectures within
-a consistent framework that underlines the functional interactions
-between their main software components. In
+main components and interactions. It is informed
+\closednote{vladimir}{is this a good word for this context,
+  'informed'? maybe 'based on'?}{Tristan}{I guess it is the right
+  word, keeping it as is.} by our experience in the development and
+sustained operation of the CBRAIN~\cite{SHER-14} and
+VIP~\cite{GLAT-13} science gateways for medical image analysis during
+the past 7 years, as well as by lessons learned from several science
+gateway and workflow projects such as
+SHIWA\footnote{\url{http://www.shiwa-workflow.eu}} and
+ER-flow\footnote{\url{http://www.erflow.eu}}.  In this section we
+provide background information and definitions of workflow engines,
+science gateways and infrastructures. In
+Section~\ref{sec:architectures}, we describe six architectures
+\closednote{vladimir}{why is it exactly six? are these architectures
+  the ones that happened to be implemented during various projects, or
+  is it a full set of all possibilities (I believe not)? I think, it
+  must be stated clearly why the focus is given to these particular
+  ones - we are reviewing and evaluating the approaches that have been
+  used already in our projects}{Tristan}{Fixed, thanks.} that cover
+the main patterns that we encountered in the various systems that we
+studied during the last few years. The architectures are described
+within a consistent framework that underlines the functional
+interactions between their main software components. In
 Section~\ref{sec:evaluation}, the different architectures are
 evaluated with metrics measuring integration complexity, robustness,
-extensibility, scalability and other specific features. We finally
-compare the architectures and highlight recommendations for science
-gateway architects and workflow engine developers.
+extensibility, scalability and other specific features. These metrics
+are specifically designed to measure the ability of the architectures
+to address workflow-related issues commonly encountered in science
+gateways. \closednote{vladimir}{can we explain why exactly these
+  metrics are used but not anything else? I think, we need to justify
+  that these ones are especially important for the SG case; it would
+  be nice to refer to some related research on software architecture
+  metrics, Google gives about 2 million pages on this, where you can
+  find metrics like cohesion, depth of scenario and so
+  on}{Tristan}{The metrics are specifically designed to cover our
+  particular workflow-SG context. Do you have any other
+  suggestion?} We finally compare the architectures and highlight
+recommendations for science gateway architects and 
+developers.
 
@@
 Swift~\cite{zhao2007swift}, Taverna~\cite{oinn2004taverna},
-Triana~\cite{taylor2007triana}, VisTrails~\cite{callahan2006managing}
-and WS-PGRADE~\cite{Kacsuk2012}. Such workflow engines usually
-describe applications in a high-level language with specific data and
-control flow constructs, parallelization operators, visual edition
-tools, links with domain-specific application repositories, provenance
+Triana~\cite{taylor2007triana}, VisTrails~\cite{callahan2006managing},
+WS-PGRADE~\cite{Kacsuk2012} and
+WS-VLAM~\cite{wsvlam}. \closednote{vladimir}{As a past developer of it, I
+  would suggest to refer to WS-VLAM~\cite{wsvlam} as well, despite it
+  is not very famous, but it played an important role for Amsterdam
+  e-science workflows}{Tristan}{Done, thanks.} Such workflow engines usually describe
+applications in a high-level language with specific data and control
+flow constructs, parallelization operators, visual edition tools,
+links with domain-specific application repositories, provenance
 recording and other features. An overview of workflow system
@@
 }\\
-\subfloat[Workflow import. Left: abstract model. Right: instantiation with service invocation.]{
+\subfloat[Workflow conversion. Left: abstract model. Right: instantiation with service invocation.]{
 \def\svgwidth{0.5\columnwidth}
@@
            }\\
-  Import & SHIWA Simulation Platform (Fine-Grained Interoperability)~\cite{plankensteiner-prodan-etal:2013}$^w$, Tavaxy~\cite{Abouelhoda2012}, system in~\cite{delaGarza2016}.
+  Conversion & SHIWA Simulation Platform (Fine-Grained Interoperability)~\cite{plankensteiner-prodan-etal:2013}$^w$, Tavaxy~\cite{Abouelhoda2012}, system in~\cite{delaGarza2016}.
 \end{tabular}
@@
   (\texttt{a$_1$}) the programs used in the workflow are installed on
-  the infrastructure, which may require administrative privileges;
-  (\texttt{a$_2$}) the workflow is configured in the science gateway
-  so that it becomes available to users. Note that integrating a
-  workflow is not the same process as integrating a workflow
-  \emph{engine}.
+  the infrastructure or prepared for on-the-fly installation, which
+  may require specific privileges; \closednote{vladimir}{or bundled in
+    packages considering all needed dependencies than can be deployed
+    on the fly during the workflow execution}{Tristan}{added, thanks!} (\texttt{a$_2$}) the
+  workflow is configured in the science gateway so that it becomes
+  available to users. Note that integrating a workflow is not the same
+  process as integrating a workflow \emph{engine}.
 \item Task control: operations to manage tasks on the infrastructure,
@@
 Figure~\ref{fig:architectures}) it might as well be delivered by the
-workflow engine directly to the user. \answerednote{Sil}{about c1: should it
+workflow engine directly to the user. \closednote{Sil}{about c1: should it
   also show from engine to infra in the figure? this is what i
@@
   single architecture, the latter case would be another variation of
-  this architecture. I modified a bit the text, is it clearer now?}
+  this architecture. I modified a bit the text to reflect this.}
 
@@
 See Figure~\subref*{archi:nested}. In nested workflows
-(Figure~\subref*{archi:nested}-Left), workflows include other
-workflows that are executed by a \emph{child} engine. Parent
+(Figure~\subref*{archi:nested}-Left), a workflow engine is integrated
+as a \emph{child} process of a \emph{parent} workflow engine. Parent
 and child engines might use different languages and might run on
-different infrastructures. A parent workflow is
-also called meta-workflow. The science gateway communicates with the
-parent engine through abstract interaction \texttt{*$_1$}. The science
-gateway also communicates with the infrastructure to transfer user
-data through abstract interactions \texttt{*$_2$} and
-\texttt{*$_3$}. Both workflow engines communicate with the
-infrastructure through abstract interactions \texttt{*$_4$} and
-\texttt{*$_5$}. The parent engine communicates with the child engine
-through abstract interaction \texttt{*$_6$}. Administrator installs
-workflows through interactions \texttt{a$_1$} and \texttt{a$_2$}.
+different infrastructures. A parent workflow is also a
+meta-workflow. The science gateway communicates with the parent engine
+through abstract interaction \texttt{*$_1$}. The science gateway also
+communicates with the infrastructure to transfer user data through
+abstract interactions \texttt{*$_2$} and \texttt{*$_3$}. Both workflow
+engines communicate with the infrastructure through abstract
+interactions \texttt{*$_4$} and \texttt{*$_5$}. The parent engine
+communicates with the child engine through abstract interaction
+\texttt{*$_6$}. Administrator installs workflows through interactions
+\texttt{a$_1$} and \texttt{a$_2$}.
 
@@
 
-\subsection{Workflow import}
+\subsection{Workflow conversion}
 
@@
 instantiated with the service invocation architecture for
-consistency. Workflows are integrated in the science gateway through
-format conversion from a native format to the science gateway
-format. Workflow import is usually an offline process that is not
-involved in workflow execution. A few systems have implemented
-workflow import. In the SHIWA Simulation Platform, it is implemented
-through the IWIR language, which provides a common language for
-portability across grid workflow
-systems~\cite{plankensteiner-prodan-etal:2013} and allows conversion
-among $n$ workflow languages using $2n$ interactions instead of $n^2$.
-Tavaxy~\cite{Abouelhoda2012}
-enables the import, merging and execution of Taverna~\cite{oinn2004taverna} and
+consistency. Workflow engines are integrated in the science gateway
+through workflow format conversion from a native format to the science
+gateway format. \closednote{vladimir}{Are we talking only about importing
+  workflows from other wf languages? I would suggest to modify the
+  title of the section to 'Workflow conversion and import' then, as
+  some gateways use the term 'import' to import their own native
+  workflows, like WS-PGRADE, and using import only in the context of
+  converting workflows can be confusing}{Tristan}{Good catch, the term
+  'import' is actually a bit vague. I guess 'conversion' is more
+  accurate. Replaced it everywhere in the text.} Workflow conversion
+is usually an offline process that is not involved in workflow
+execution. A few systems have implemented workflow conversion. In the
+SHIWA Simulation Platform, it is implemented through the IWIR
+language, which provides a common language for portability across grid
+workflow systems~\cite{plankensteiner-prodan-etal:2013} and allows
+conversion among $n$ workflow languages using $2n$ interactions
+instead of $n^2$.  Tavaxy~\cite{Abouelhoda2012} enables the import,
+merging and execution of Taverna~\cite{oinn2004taverna} and
 Galaxy~\cite{goecks2010galaxy} workflows; when some workflow parts
@@
 using their native engine. The work in~\cite{delaGarza2016} describes
-workflow import from KNIME~\cite{Berthold2008} to WS-PGRADE/gUSE and
-from Galaxy~\cite{goecks2010galaxy} to WS-PGRADE/gUSE.
+workflow conversion from KNIME~\cite{Berthold2008} to WS-PGRADE/gUSE
+and from Galaxy~\cite{goecks2010galaxy} to WS-PGRADE/gUSE.
 
@@
                                      & \textbf{Nested}
-                                     & \textbf{Import} \\
+                                     & \textbf{Conversion} \\
 \cellcolor[HTML]{EEEEEE}\textbf{Integration complexity}& \multicolumn{6}{l}{\cellcolor[HTML]{EEEEEE}}\\
@@
                                      & \cellcolor[HTML]{999999}1
-                                     & \cellcolor[HTML]{999999}1
+                                     & \cellcolor[HTML]{99FF99}0
                                      & \cellcolor[HTML]{999999}1
@@
   Fine-grained debugging -- \texttt{O$_2$}   & \cellcolor[HTML]{99FF99}0
-                                     & \cellcolor[HTML]{99FF99}0
                                      & \cellcolor[HTML]{999999}1
-                                     & \cellcolor[HTML]{99FF99}0
                                      & \cellcolor[HTML]{999999}1
-                                     & \cellcolor[HTML]{99FF99}0\\
-  \textbf{Total} (missing features) & \cellcolor[HTML]{99CC99}\textbf{1}
-                                     & \cellcolor[HTML]{99CC99}\textbf{1}
+                                     & \cellcolor[HTML]{999999}1
+                                     & \cellcolor[HTML]{999999}1
+                                     & \cellcolor[HTML]{999999}1\\
+  \textbf{Total} (missing features) & \cellcolor[HTML]{99FF99}\textbf{1}
                                      & \cellcolor[HTML]{999999}\textbf{2}
-                                     & \cellcolor[HTML]{99CC99}\textbf{1}
-                                     & \cellcolor[HTML]{99CC99}\textbf{1}
-                                     & \cellcolor[HTML]{99FF99}\textbf{0}\\
+                                     & \cellcolor[HTML]{99FF99}\textbf{1}
+                                     & \cellcolor[HTML]{999999}\textbf{2}
+                                     & \cellcolor[HTML]{99FF99}\textbf{1}
+                                     & \cellcolor[HTML]{99FF99}\textbf{1}\\
 \end{tabular}
@@
 covered, since they do not stem directly from the software
-architecture. Robustness is measured here as a consequence of global
+architecture. 
+\closednote{vladimir}{I believe, proper error handling through the different levels of wf engine/gateway must be especially mentioned even if not especially covered.
+One of serious issues we faced while working with gateways at the AMC, in particular with gUse/WS-PGRADE, was that when it failed sometimes it was completely not clear what was the problem exactly.
+So we might consider discussing not only the number of software components but how explicit and clear interactions are between them; how transparently we can monitor the execution through all the levels and thus detect and prevent possible malfunction}{Tristan}{In our current evaluation framework, this is supposed to be covered by O2.
+I still agree with you that O2 could be further refined.
+See Github issue #27}
+Robustness is measured here as a consequence of global
 complexity, since complex architectures tend to be more prone to
@@
 \begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
-\item Multiple engine instances (\texttt{S$_1$}): measures the potential of having
-  more than one engine instance in the architecture simultaneously. Workflow engines
-  may require important amounts of resources when several workflows,
-  or large workflows, are executed. At some point, it may be required
-  for the science gateway to means that adding a new engine instance is a
-  fully automated process, i.e. the workflow engine only has to be
-  started. In this case, elastic engines can be implemented through some kind
-  of auto-scaling mechanism to control the number
-  of engine instances in the architecture. \texttt{S$_1$}=1 means that
-  adding a new engine instance requires some form of manual
-  intervention, which prevents easy
-  implementations of elastic engines. \texttt{S$_1$}=2 means that new engine
-  instances cannot be added.
+\item Multiple engine instances (\texttt{S$_1$}): measures the
+  potential for a single gateway instance to use more than one engine
+  instance simultaneously. Workflow engines may require important
+  amounts of resources when several workflows, or large workflows, are
+  executed. At some point, it may be required for the science gateway
+  to distribute the load among several engines. \texttt{S$_1$}=0 means
+  \closednote{vladimir}{is this the right word here, 'to
+    means'?}{Tristan}{This sentence had been truncated, thanks.} that
+  adding a new engine instance is inherently supported in the
+  architecture. In this case, elastic engines can be implemented
+  through some kind of auto-scaling mechanism to control the number of
+  engine instances in the architecture. \texttt{S$_1$}=1 means that
+  multiple engine instances requires specific developments in the
+  science gateway or workflow engine. \texttt{S$_1$}=2 means that new
+  engine instances cannot be added.
 \item Distributed engines (\texttt{S$_2$}): measures the potential of distributing
@@
 \item Task scheduling (\texttt{S$_3$}): measures the complexity
-  introduced by the architecture regarding task scheduling on
+  added by the architecture to task scheduling on
   the infrastructure. Task
@@
   characteristics. \texttt{S$_3$}=0 means that the architecture does
-  not introduce any additional complexity to the scheduling problem,
+  not add any additional complexity to the scheduling problem,
   and \texttt{S$_3$}=1 otherwise.
@@
   additional level of flexibility to build workflows from reusable
-  components. \texttt{O$_1$}=0 means that meta-workflows are enabled by the architecture, and
-  \texttt{O$_1$}=1 otherwise.
-  \note{si}{why here we do not have the "possible to implement" situation?}
+  components. \texttt{O$_1$}=0 means that meta-workflows are
+  intrinsically enabled by the architecture (i.e. they can be
+  implemented using the components and interactions already in place),
+  amd \texttt{O$_1$}=1 means that they may be implemented with some
+  development effort. \closednote{si}{why here we do not have the
+    "possible to implement" situation?}{Tristan}{O$_1$=2 would never
+    be reached. Besides, that would make it difficult to sum O1 and O2
+    because they wouldn't measure the same type of thing. In the
+    current situation, we are only talking about the presence/absence
+    of a particular feature and this is why O1 and O2 can be added to
+    obtain a total number of 'missing features'.}
+  \closednote{vladimir}{I agree with Silvia, we should have the
+    intermediate values as well, since meta-workflows might be
+    straightforwardly supported by the gateway, or they are still
+    possible to implement but only with some workarounds. Like it was
+    with SHIWA: first 6 months coarsed-grained interoperability looked
+    more like a workaround till it was better integrated and supported
+    by WS-PGRADE itself}{Tristan}{Yes but in the situation you
+    describe, evolutions came from the wf engine implementation rather
+    than from the architecture to integrate it, which is our point
+    here.}
 
@@
   workflow).  Fine-grained information about workflow tasks is
-  required to properly troubleshoot workflow executions.
-  \texttt{O$_2$}=0 means that such information is easy to access, and
-  \texttt{O$_2$}=1 otherwise.
+  required to properly troubleshoot workflow executions. Its
+  availability and accuracy usually depends on the number of software
+  layers between the science gateway and the workflow
+  engine~\cite{olabarriaga2014}. \texttt{O$_2$}=0 means that
+  the information is directly accessible in the science gateway,
+  \texttt{O$_2$}=1 means that it is obtained through one or more software
+  interactions. \closednote{vladimir}{otherwise - does it mean the
+    fine-grained info is available but hard to access or not available
+    it all? the latter happens as well}{Tristan}{Precised the text.}
 \end{itemize}
@@
 
-\note{sil}{since we are dreaming, it seems to me that keeping provenance properly is also something to consider.
-the architecture can hide some details, and make it nearly impossible to keep provenance without hard coded hacks across the layers/components - i think}
+\closednote{sil}{since we are dreaming, it seems to me that keeping provenance properly is also something to consider.
+the architecture can hide some details, and make it nearly impossible to keep provenance without hard coded hacks across the layers/components - i think}{Tristan}{Fixed the definition according to the discussion in PR #20.}
 
@@
 %, which is in general not possible
-\note{sil}{why not? i agree that S=2, but i do not agree that it would be impossible. you say how below}
+\closednote{sil}{why not?
+i agree that S=2, but i do not agree that it would be impossible.
+you say how below}{Tristan}{Updated the definition of S$_1$ to explain that multiple engine instances have to be available in the same sg instance.}
 (\texttt{S$_1$}=2). 
 Specific IT setups such as load-balancing between
-web server instances might be used to create new gateway instances, distributing the users, but would not solve the problem for the workload of a single workflow or user.
+web server instances might be used to create new gateway instances, but they would not allow a single gateway instance to use multiple engine instances, which is the point here.
 Distributed engines are not available by default
 (\texttt{S$_2$}=1). 
-The scheduling of tasks on the infrastructure is
-as complex as in any other architecture since the workflow engine
-might implement any kind of scheduling policy (\texttt{S$_3$}=0).
-\note{sil}{so same as the others = "easy"? i think this requires more explanation. maybe we should introduce a score for "dont know"?}
-
-\paragraph{Specific features} Meta-workflows are not supported by default
-(\texttt{O$_1$}=1).  
-\note{sil}{why not? if the engine is ws-pgrade it does. i really think that meta-workflows (and nested workflows) is a property exclusively of the engine. you can see that in fig 2.V interaction *6}
-Debugging is not an issue since the science
-gateway can retrieve any information from the workflow engine directly
-(\texttt{O$_2$}=0).
+The architecture does not add any particular complexity to the task scheduling problem since the workflow engine
+may implement any kind of scheduling policy (\texttt{S$_3$}=0).
+\closednote{sil}{so same as the others = "easy"? i think this requires more explanation. maybe we should introduce a score for "dont know"?}{Tristan}{Fixed as discussed in PR #20}
+
+\paragraph{Specific features} Supporting meta-workflows requires
+specific developments in the workflow engine or science gateway to
+allow workflows to execute other workflows (\texttt{O$_1$}=1). For
+instance, specific Java objects may be implemented in the science
+gateway framework to execute different types of workflows and link
+them together.  \closednote{sil}{why not? if the engine is ws-pgrade
+  it does. i really think that meta-workflows (and nested workflows)
+  is a property exclusively of the engine. you can see that in fig 2.V
+  interaction *6}{Tristan}{If the engine is WS-PGRADE as used in SHIWA
+  CGI, then the architecture is nested workflows with service
+  invocation. In a tight integration, meta-workflows would be
+  expressed, e.g. as compositions of Liferay portlets that all
+  implement different types of engines.}  The science gateway can
+retrieve fine-grained debugging information from the workflow engine
+directly (\texttt{O$_2$}=0).
 
@@
 multiple engine instances through multiple workflow services. Adding a
-new engine instance, however, requires manual intervention to declare
-the new instance in the science gateway
-(\texttt{S$_1$}=1). Consequently, elastic engines are difficult to
-implement because they require a mechanism to update the science
-gateway configuration when a new engine instance is
-available. Distributing the execution of a single workflow in multiple
+new engine instance, however, requires specific developments in the science gateway
+(\texttt{S$_1$}=1). Distributing the execution of a single workflow in multiple
 engines is usually not possible unless the workflow engine has
-specific abilities (\texttt{S$_2$}=1).  \answerednote{sil}{is this score for
+specific abilities (\texttt{S$_2$}=1).  \closednote{sil}{is this score for
   VIP? i think that the score depends on the implementation of the
@@
   we should not say s_2=1 for the architecture}{Tristan}{Removed VIP example.
-Score definition still has to be modified according to the discussion in @SilviaOlabarriaga's PR (\#20).} The scheduling of
-tasks on the infrastructure is as complex as in any other architecture
-since the workflow engine might implement any kind of scheduling
-policy (\texttt{S$_3$}=0).  \note{sil}{see comment in section 3.1}
-
-\paragraph{Specific features} Meta-workflows are not supported by default
-(\texttt{O$_1$}=1). 
-\note{sil}{see comment in section 3.1}
-Fine-grained debugging information is usually easy
-to obtain since the workflow service provides direct access to the
-engine (\texttt{O$_2$}=0).
-\note{sil}{this is really not the case. the fact that the wf service is a separate component that hides information is a big problem for debugging at the science gateway level. 
-we had a paper about this:
-	@inproceedings{olabarriaga2014scientific,
-		Author = {Olabarriaga, Silvia and Pierantoni, Gabrielle and Taffoni, Giuliano and Sciacca, Eva and Jaghoori, Mahdi and Korkhov, Vladimir and Castelli, Giuliano and Vuerli, Claudio and Becciani, Ugo and Carley, Eoin and others},
-		Booktitle = {e-Science (e-Science), 2014 IEEE 10th International Conference on},
-		Organization = {IEEE},
-		Pages = {298--305},
-		Title = {Scientific Workflow Management--For Whom?},
-		Year = {2014}}
-	}
+Updated score definition according to the discussion in @SilviaOlabarriaga's PR (\#20).} The architecture does not add any particular complexity to the task scheduling problem
+since the workflow engine may implement any kind of scheduling
+policy (\texttt{S$_3$}=0).  \closednote{sil}{see comment in section 3.1}{Tristan}{Fixed the same way}
+
+\paragraph{Specific features} Meta-workflows require specific
+developments in the workflow engine or science gateway to allow
+workflows to execute other workflows.  (\texttt{O$_1$}=1).
+\closednote{sil}{see comment in section 3.1}{Tristan}{same answer} The
+science gateway needs to invoke interaction \texttt{d} to retrieve
+dine-grained debugging information from the workflow service
+(\texttt{O$_2$}=1).  \closednote{sil}{this is really not the case. the
+  fact that the wf service is a separate component that hides
+  information is a big problem for debugging at the science gateway
+  level.  we had a paper about this:
+  @inproceedings{olabarriaga2014scientific, Author = {Olabarriaga,
+      Silvia and Pierantoni, Gabrielle and Taffoni, Giuliano and
+      Sciacca, Eva and Jaghoori, Mahdi and Korkhov, Vladimir and
+      Castelli, Giuliano and Vuerli, Claudio and Becciani, Ugo and
+      Carley, Eoin and others}, Booktitle = {e-Science (e-Science),
+      2014 IEEE 10th International Conference on}, Organization =
+    {IEEE}, Pages = {298--305}, Title = {Scientific Workflow
+      Management--For Whom?}, Year = {2014}} }{Tristan}{I agree. Fixed
+  it and precised the definition of O2.}
 
@@
 (\texttt{R$_2$}=1).
+\closednote{vladimir}{there is a large piece of text in Scalability about reliability of the workflow engine task which on my opinion suits Robustness better}{Tristan}{I understand that that text may sound like robustness, but the robustness metrics are about the architecture, not the tasks, as explained in 3.1 (Errors coming from
+the infrastructure (e.g., unavailable data or termi-
+nated tasks) or workflows (e.g., wrong user input or
+application errors) are not covered).}
+\closednote{vladimir}{in addition, the claim of very good robustness of task encapsulation seems questionable to me as in this model you delegate the critical task of controlling workflow execution to, let's say, some 'random' node on the grid that you have no control of, not sure about its performance, connectivity and reliability and so on; however, considering the models and dependencies presented in fig.2 this evaluation of robustness metric may be correct}{Tristan}{You are right but the robustness issues that you are talking about are related to the infrastructure and not the architecture.
+This is really an issue related to task scheduling as the SG might decide to run the workflow tasks on reliable nodes of the infrastructure such as a dedicated cluster.
+So the issue that you mention is reflected in S3.}
 
@@
 task fails. The workflow task is also longer than all its sub-tasks,
-which increases its chances of failure. In addition, task parameters,
+which increases its chances of failure.
+\closednote{vladimir}{isn't this about robustness?}{Tristan}{Nope, this is task scheduling, therefore scalability in our evaluation framework.
+See answer to your previous note.} In addition, task parameters,
 for instance estimated walltime, are more difficult to estimate for
@@
 
-\paragraph{Specific features} Meta-workflows are not available
-(\texttt{O$_1$}=1).  
-\note{sil}{see comment in section 3.1}
-Obtaining fine-grained information about workflow
-tasks is not straightforward since the science gateway has no
-knowledge about the workflow topology, and the workflow engine is
-integrated as a task (\texttt{O$_2$}=1).
-\note{sil}{mm, i think the case here is the same as for service. or actually i would give this one a better score, because there are less components to cross}
+\paragraph{Specific features} Task encapsulation explicitly enables
+meta-workflows (\texttt{O$_1$}=0). Indeed, workflow engines may invoke
+other engines using the exact same interaction used to submit regular
+sub-tasks (interaction \texttt{e}), without any additional
+development. \closednote{sil}{see comment in section
+  3.1}{Tristan}{Fixed, actually updated the score of this architecture
+  now that the definition is more accurate.} Fine-grained debugging
+information is obtained through interaction \texttt{b}
+(\texttt{O$_2$}=1). \closednote{sil}{mm, i think the case here is the
+  same as for service. or actually i would give this one a better
+  score, because there are less components to cross}{Tristan}{Fixed the text to better match the updated definition of O2}
 
@@
 
-\paragraph{Scalability} New engine instances only require new agents,
-which can be easily automated (\texttt{S$_1$=0}) and by design very
-suitable for elastic computing. For instance, auto-scaling rules can
-be implemented to start new agents when the workload in the science
-gateway exceeds a certain threshold~\cite{lorido2012auto}. Distributed
-engines are not available by default (\texttt{S$_2$=1}) and task
-scheduling is as complex as in any other architecture
-(\texttt{S$_3$=0}).
-
-\paragraph{Specific features} Meta-workflows are not available by default
-(\texttt{O$_1$=1}).  
-\note{sil}{see comment in section 3.1}
-Accessing debugging information is not likely to
-be an issue since the workflow pool could implement specific functions
-for that (\texttt{O$_2$=0}).
-\note{sil}{i think the case here is the same as for service. or actually i would give this one a worse score, because there is one additional component - the pool - to cross}
+\paragraph{Scalability} By design, new engine instances only require
+new agents, which can be easily automated (\texttt{S$_1$=0}). For
+instance, auto-scaling rules can be implemented to start new agents
+when the workload in the science gateway exceeds a certain
+threshold~\cite{lorido2012auto}. Distributed engines are not available
+by default (\texttt{S$_2$=1}) and the architecture does not add any
+complexity to the task scheduling problem (\texttt{S$_3$=0}).
+
+\paragraph{Specific features} Meta-workflows require specific
+developments in the workflow engine to enable workflow submission
+(\texttt{O$_1$=1}).  \closednote{sil}{see comment in section 3.1}{Tristan}{Same answer.} Debugging information is accessed through interactions \texttt{d} and \texttt{f} (\texttt{O$_2$=1}).
+\closednote{sil}{i think the case here is the same as for service. or
+  actually i would give this one a worse score, because there is one
+  additional component - the pool - to cross}{Tristan}{Fixed using the updated definition of O2}
 
@@
 \texttt{d} (twice).
+\closednote{vladimir}{and if we have a meta-workflow with 3 and more sub-workflows the score will be even higher, is it a fair calculation for an architecture?
+we did not increase the scores for possible meta-workflows supported by wf engines in other cases}{Tristan}{Good point!
+You are totally right for meta-workflows, I added a sentence in the paragraph below.
+However, nested workflows are studied here as an architecture to integrate a workflow engine (and only one), not to describe meta-workflows.}
 
@@
 \paragraph{Scalability} As in the service architecture, adding a new
-workflow instance requires manual configuration in the science gateway
+engine instance requires specific developments in the science gateway
 (instance of a parent engine), or in the parent engine (instance of a
@@
 \paragraph{Specific features}
-Meta-workflows can be implemented, which is one of the main interest of this
-architecture (\texttt{O$_1$=0}).  Debugging is difficult because the
-science gateway cannot directly access fine-grained information in the
-sub-workflow (\texttt{O$_2$=1}).
-
-\subsection{Workflow import with service invocation}
-
-\paragraph{Integration complexity} Workflow import with service invocation
+Meta-workflows can be implemented, which is one of the main interest
+of this architecture (\texttt{O$_1$=0}). Note, however, that the
+complexity of the architecture increases with the number of engine
+types involved in meta-workflows: for instance, a meta-workflow with
+child workflows executed by 2 different engine types would require an
+additional workflow service and the corresponding interactions.
+Debugging information is accessed through interaction \texttt{d} (invoked twice) (\texttt{O$_2$=1}).
+
+\subsection{Workflow conversion with service invocation}
+
+\paragraph{Integration complexity} Workflow conversion with service invocation
 requires the same components as for service invocation
@@
 
-\paragraph{Specific features} Meta-workflows are available after import,
-by connecting workflows in the language used in the science gateway
-(\texttt{O$_1$=0}). 
-\note{sil}{i dont understand this about meta-workflows.
-translation does not imply composition, right?
-if the science gateway is allowed to implement workflow composition, then this is probably the case in other architectures as well?}
-Debugging information is accessible as in the
-service invocation architecture (\texttt{O$_2$=0}).
-\note{sil}{see notes 3.2}
+\paragraph{Specific features} Meta-workflows are available after
+conversion, by connecting workflows in the language used in the
+science gateway (\texttt{O$_1$=0}).  \closednote{sil}{i dont
+  understand this about meta-workflows. translation does not imply
+  composition, right?  if the science gateway is allowed to implement
+  workflow composition, then this is probably the case in other
+  architectures as well?}{Tristan}{Well at this level, connecting two
+  workflows expressed in the same language is straightforward,
+  although I agree that it might be tricky in some cases.} Debugging information is accessible
+as in the service invocation architecture (\texttt{O$_2$=1}).
+\closednote{sil}{see notes 3.2}{Tristan}{Fixed the same way}
 
@@
 Tight integration and task encapsulation are the simplest
-architectures to integrate, followed by service integration, workflow import (with
+architectures to integrate, followed by service integration, workflow conversion (with
 service invocation) and pool. Nested workflows (with service
@@
 architectures, with tight integration and task encapsulation in the top
-group, service integration and workflow import (with service
+group, service integration and workflow conversion (with service
 invocation) close behind, pool in a third group, and nested workflows
@@
 nested workflow architecture, with 2 infrastructures and 2 workflow
-services. Task encapsulation and workflow import (with service
+services. Task encapsulation and workflow conversion (with service
 invocation) perform slightly better when integrating new engines
 (\texttt{E$_1$}), which makes them useful architectures for science
-gateways that do not focus on a particular engine.  However, task encapsulation, workflow import (with
+gateways that do not focus on a particular engine.  However, task encapsulation, workflow conversion (with
 service invocation) and tight integration perform worse than the
@@
 to update infrastructure (task encapsulation), science gateway (tight
-integration), or workflow converter (workflow import). Workflow import
+integration), or workflow converter (workflow conversion). Workflow conversion
 performs worse than the others when integrating new workflows
@@
 
-Differences in other specific features should not be neglected. Nested
-workflows (with service invocation) and workflow import (with service
-invocation) are the only architectures that support meta-workflows. Besides,
+Differences in other specific features should not be neglected. Task encapsulation, nested
+workflows (with service invocation) and workflow conversion (with service
+invocation) are the only architectures that intrinsically support meta-workflows.
+\closednote{vladimir}{As mentioned earlier by Silvia, this might not be exactly the case, depending on the point of view}{Tristan}{The example given by Silvia (WS-PGRADE) actually comes down to the nested workflow architecture.
+I added 'intrinsically' to the sentence to better reflect that.} Besides,
 task encapsulation and nested workflows may hamper
@@
 architectures, based on the metrics in Table~\ref{table:evaluation}.
-Note that this analysis is only meant for illustration purposes, since it
-assumes that all criteria have equal weights while real systems would
-favor some of them. Overall, task encapsulation, and workflow import (with
-service invocation) perform a bit better than tight integration,
-service invocation and pool. Nested workflows stand a bit behind for
-the reasons mentioned previously.
+Note that this analysis is only meant for illustration purposes, since
+it assumes that all criteria have equal weights while real systems
+would favor some of them. Overall, task encapsulation performs a bit
+better than the other architectures and nested workflows stand a bit
+behind for the reasons mentioned previously.
 \begin{table*}
@@
                                     & \textbf{Nested}
-                                    & \textbf{Import} \\
+                                    & \textbf{Conversion} \\
   Integration (global score)    &
@@
                                      & \cellcolor[HTML]{99CC99}{0.50}\\
-Specific features (global score) & \cellcolor[HTML]{99CC99}{0.50}
-                                     & \cellcolor[HTML]{99CC99}{0.50}
+Specific features (global score) & \cellcolor[HTML]{99FF99}{0.00}
                                      & \cellcolor[HTML]{999999}{1.00}
-                                     & \cellcolor[HTML]{99CC99}{0.50}
-                                     & \cellcolor[HTML]{99CC99}{0.50}
+                                     & \cellcolor[HTML]{99FF99}{0.00}
+                                     & \cellcolor[HTML]{999999}{1.00}
+                                     & \cellcolor[HTML]{99FF99}{0.00}
                                      & \cellcolor[HTML]{99FF99}{0.00}\\
-                                    & \cellcolor[HTML]{99E799}\textbf{2.1}
-                                    & \cellcolor[HTML]{99E699}\textbf{2.1}
-                                    & \cellcolor[HTML]{99FE99}\textbf{1.5}
-                                    & \cellcolor[HTML]{99EF99}\textbf{1.9}
-                                    & \cellcolor[HTML]{999999}\textbf{4.0}
-                                    & \cellcolor[HTML]{99FF99}\textbf{1.5}\\
+                                    & \cellcolor[HTML]{99DA99}\textbf{1.6}
+                                    & \cellcolor[HTML]{99B899}\textbf{2.6}
+                                    & \cellcolor[HTML]{99FF99}\textbf{0.5}
+                                    & \cellcolor[HTML]{99BF99}\textbf{2.4}
+                                    & \cellcolor[HTML]{999999}\textbf{3.5}
+                                    & \cellcolor[HTML]{99DE99}\textbf{1.5}\\
 \end{tabular}
@@
 
-The abstract nested workflows and workflow import patterns were
+The abstract nested workflows and workflow conversion patterns were
 instantiated with service invocation so that they can be analyzed in
@@
 grid workflow systems that is based on the workflow reference model
-defined by the Workflow Management Coalition~\cite{wfRefModel}.
-Deelman et al. \cite{deelman2009workflows} have further characterized
-the features of workflow systems and proposed a general architecture??
+defined by the Workflow Management
+Coalition\footnote{\url{http://www.wfmc.org}}.  Deelman et
+al. \cite{deelman2009workflows} have further characterized the
+features of workflow systems and proposed a general architecture??
 These two works, however, do not mention science gateways.

### 2016-05-15 paper.tex
Minor updates

@@
 
+\label{sec:introduction-engines}
 In the last decade, the e-Science community has developed workflow
@@
 science gateways with important workloads, which would favor task
-encapsulation and pool. \note{Naj}{I feel this paragraph is a bit
+encapsulation and pool. \closednote{Naj}{I feel this paragraph is a bit
   vague, and doesn't distill and contextualize the thorough work
-  described before.}
+  described before.}{Tristan}{Rejected comment :)}
 
@@
 I added 'intrinsically' to the sentence to better reflect that.} Besides,
-task encapsulation and nested workflows may hamper
+tight integration is the only architecture that allows
 fine-grained debugging. 
-The availability of fine-grained debugging information may also be
+The availability of fine-grained debugging information may be
 critical to efficient user support.
@@
 used. Nevertheless, it could be interesting to explore other types of
-instantiations, in particular nested workflows with task
-encapsulation.\answerednote{Naj}{I think in the discussion, you may want to close the circle on a statement in introduction, which referred to your 7-year experience with CBRAIN and VIP}{Tristan}{Good idea, will think about it.}
+instantiations.\closednote{Naj}{I think in the discussion, you may want to close the circle on a statement in introduction, which referred to your 7-year experience with CBRAIN and VIP}{Tristan}{Not sure how to do this without making the conclusion sound too specific to these systems.}
 
@@
 project may obviously influence the choice of an architecture to
-integrate workflow engines. For instance, some workflow engines may be
-already available as web services\note{Naj}{This is a good place to offer some examples}, which would tend to favor service
+integrate workflow engines. For instance, many workflow engines are
+already available as web services\closednote{Naj}{This is a good place to offer some examples}{Tristan}{Many of them are, precised the text, not sure adding explicit examples makes sense at this stage.}, which tends to favor service
 invocation, and other science gateways may have strongly tested task
@@
 
-The early work~\cite{yuTaxonomy} proposes a generic architecture for
-grid workflow systems that is based on the workflow reference model
-defined by the Workflow Management
+There is abundant literature describing specific workflow engines and
+systems; some of the main references are cited in
+Section~\ref{sec:introduction-engines}. A few works described the
+architecture and features of workflow system at an abstract level. For
+instance, the early work~\cite{yuTaxonomy} proposes a generic
+architecture for grid workflow systems that is based on the workflow
+reference model defined by the Workflow Management
 Coalition\footnote{\url{http://www.wfmc.org}}.  Deelman et
 al. \cite{deelman2009workflows} have further characterized the
-features of workflow systems and proposed a general architecture??
-These two works, however, do not mention science gateways.
-
-Shahand et al.~\cite{shahand:2015ab} presents the Science Gateway
-Canvas, which is a business reference model where the most relevant
-functional components are organized into functional groups. Although
-workflow management is mentioned as a possible component to coordinate
-distributed computations in science gateways, the paper does not
-comment on architecture to achieve this.
+features of workflow systems. These two works, however, do not mention
+science gateways.
+
+Numerous science gateways have been described, as mentioned in
+Section~\ref{sec:intro_sg}. However, only a few works focused on
+science gateway architectures. Shahand et al.~\cite{shahand:2015ab}
+presents the Science Gateway Canvas, which is a business reference
+model where the most relevant functional components are organized into
+functional groups. Although workflow management is mentioned as a
+possible component to coordinate distributed computations in science
+gateways, the paper does not comment on architecture to achieve this.
 
@@
 gateway, workflow engine and distributed infrastructure components,
-similarly to the "service invocation" pattern. However other patterns
+similarly to the ``service invocation'' pattern. However other patterns
 are not identified.
-\note{vladimir}{@silvia, does it make sense to mention workflow classification efforts based on workflow motifs and so on?}
+
+\closednote{vladimir}{@silvia, does it make sense to mention workflow classification efforts based on workflow motifs and so on?}{Tristan}{I'm  not  sure. Which papers are you thinking of?}
 
@@
 % Summary of contributions
-We have systematically reviewed architectures used to integrate workflow engines in
-science gateways. These architectures are described in a system-independent framework
-suitable for comparison, illustrated on real systems, and evaluated using novel quantitative metrics that allow simple comparison across architectures. We have discussed the
-pros and cons of all the presented architectures based on these metrics.
+We have systematically reviewed architectures used to integrate
+workflow engines in science gateways. These architectures are
+described in a system-independent framework suitable for comparison,
+illustrated on real systems, and evaluated using novel quantitative
+metrics that allow simple comparison across architectures. We have
+discussed the pros and cons of all the presented architectures based
+on these metrics.
 

### 2016-05-15 paper.tex
Removed closed comments as they may introduce white spaces.

@@
 \newcommand{\todo}[2]{\pdfmargincomment[color=red,author=#1,open=true]{#2}}
-\newcommand{\note}[2]{\pdfmargincomment[color=yellow,author=#1,open=true]{#2}}
-\newcommand{\closednote}[4]{} % closed notes are hidden. This should obviously be a parameter of \note rather than a separate command.
-\newcommand{\answerednote}[4]{\pdfmargincomment[color=green,author=#1-#3,open=true]{#1: #2 \textCR \textCR #3: #4}}
-\newcommand{\closedanswerednote}[6]{}
+%\newcommand{\note}[2]{\pdfmargincomment[color=yellow,author=#1,open=true]{#2}}
+%\newcommand{\closednote}[4]{} % closed notes are hidden. This should obviously be a parameter of \note rather than a separate command.
+%\newcommand{\answerednote}[4]{\pdfmargincomment[color=green,author=#1-#3,open=true]{#1: #2 \textCR \textCR #3: #4}}
+%\newcommand{\closedanswerednote}[6]{}
 
@@
 main components and interactions. It is informed
-\closednote{vladimir}{is this a good word for this context,
-  'informed'? maybe 'based on'?}{Tristan}{I guess it is the right
-  word, keeping it as is.} by our experience in the development and
+ by our experience in the development and
 sustained operation of the CBRAIN~\cite{SHER-14} and
@@
 Section~\ref{sec:architectures}, we describe six architectures
-\closednote{vladimir}{why is it exactly six? are these architectures
-  the ones that happened to be implemented during various projects, or
-  is it a full set of all possibilities (I believe not)? I think, it
-  must be stated clearly why the focus is given to these particular
-  ones - we are reviewing and evaluating the approaches that have been
-  used already in our projects}{Tristan}{Fixed, thanks.} that cover
+ that cover
 the main patterns that we encountered in the various systems that we
@@
 to address workflow-related issues commonly encountered in science
-gateways. \closednote{vladimir}{can we explain why exactly these
-  metrics are used but not anything else? I think, we need to justify
-  that these ones are especially important for the SG case; it would
-  be nice to refer to some related research on software architecture
-  metrics, Google gives about 2 million pages on this, where you can
-  find metrics like cohesion, depth of scenario and so
-  on}{Tristan}{The metrics are specifically designed to cover our
-  particular workflow-SG context. Do you have any other
-  suggestion?} We finally compare the architectures and highlight
+gateways. We finally compare the architectures and highlight
 recommendations for science gateway architects and 
@@
 WS-PGRADE~\cite{Kacsuk2012} and
-WS-VLAM~\cite{wsvlam}. \closednote{vladimir}{As a past developer of it, I
-  would suggest to refer to WS-VLAM~\cite{wsvlam} as well, despite it
-  is not very famous, but it played an important role for Amsterdam
-  e-science workflows}{Tristan}{Done, thanks.} Such workflow engines usually describe
+WS-VLAM~\cite{wsvlam}.  Such workflow engines usually describe
 applications in a high-level language with specific data and control
@@
   the infrastructure or prepared for on-the-fly installation, which
-  may require specific privileges; \closednote{vladimir}{or bundled in
-    packages considering all needed dependencies than can be deployed
-    on the fly during the workflow execution}{Tristan}{added, thanks!} (\texttt{a$_2$}) the
+  may require specific privileges; (\texttt{a$_2$}) the
   workflow is configured in the science gateway so that it becomes
@@
 Figure~\ref{fig:architectures}) it might as well be delivered by the
-workflow engine directly to the user. \closednote{Sil}{about c1: should it
-  also show from engine to infra in the figure? this is what i
-  understand from the above sentence}{Tristan}{Actually c1 is either
-  coming from the science gatewray or from the workflow engine but
-  then it merges with c2. I don't think we can cover both cases in a
-  single architecture, the latter case would be another variation of
-  this architecture. I modified a bit the text to reflect this.}
+workflow engine directly to the user.
 
@@
 through workflow format conversion from a native format to the science
-gateway format. \closednote{vladimir}{Are we talking only about importing
-  workflows from other wf languages? I would suggest to modify the
-  title of the section to 'Workflow conversion and import' then, as
-  some gateways use the term 'import' to import their own native
-  workflows, like WS-PGRADE, and using import only in the context of
-  converting workflows can be confusing}{Tristan}{Good catch, the term
-  'import' is actually a bit vague. I guess 'conversion' is more
-  accurate. Replaced it everywhere in the text.} Workflow conversion
+gateway format. Workflow conversion
 is usually an offline process that is not involved in workflow
@@
 architecture. 
-\closednote{vladimir}{I believe, proper error handling through the different levels of wf engine/gateway must be especially mentioned even if not especially covered.
-One of serious issues we faced while working with gateways at the AMC, in particular with gUse/WS-PGRADE, was that when it failed sometimes it was completely not clear what was the problem exactly.
-So we might consider discussing not only the number of software components but how explicit and clear interactions are between them; how transparently we can monitor the execution through all the levels and thus detect and prevent possible malfunction}{Tristan}{In our current evaluation framework, this is supposed to be covered by O2.
-I still agree with you that O2 could be further refined.
-See Github issue #27}
 Robustness is measured here as a consequence of global
@@
   to distribute the load among several engines. \texttt{S$_1$}=0 means
-  \closednote{vladimir}{is this the right word here, 'to
-    means'?}{Tristan}{This sentence had been truncated, thanks.} that
+  that
   adding a new engine instance is inherently supported in the
@@
   amd \texttt{O$_1$}=1 means that they may be implemented with some
-  development effort. \closednote{si}{why here we do not have the
-    "possible to implement" situation?}{Tristan}{O$_1$=2 would never
-    be reached. Besides, that would make it difficult to sum O1 and O2
-    because they wouldn't measure the same type of thing. In the
-    current situation, we are only talking about the presence/absence
-    of a particular feature and this is why O1 and O2 can be added to
-    obtain a total number of 'missing features'.}
-  \closednote{vladimir}{I agree with Silvia, we should have the
-    intermediate values as well, since meta-workflows might be
-    straightforwardly supported by the gateway, or they are still
-    possible to implement but only with some workarounds. Like it was
-    with SHIWA: first 6 months coarsed-grained interoperability looked
-    more like a workaround till it was better integrated and supported
-    by WS-PGRADE itself}{Tristan}{Yes but in the situation you
-    describe, evolutions came from the wf engine implementation rather
-    than from the architecture to integrate it, which is our point
-    here.}
+  development effort.
 
@@
   \texttt{O$_2$}=1 means that it is obtained through one or more software
-  interactions. \closednote{vladimir}{otherwise - does it mean the
-    fine-grained info is available but hard to access or not available
-    it all? the latter happens as well}{Tristan}{Precised the text.}
+  interactions. 
 \end{itemize}
@@
 features in the architecture.
-
-\closednote{sil}{since we are dreaming, it seems to me that keeping provenance properly is also something to consider.
-the architecture can hide some details, and make it nearly impossible to keep provenance without hard coded hacks across the layers/components - i think}{Tristan}{Fixed the definition according to the discussion in PR #20.}
 
@@
 %, which is in general not possible
-\closednote{sil}{why not?
-i agree that S=2, but i do not agree that it would be impossible.
-you say how below}{Tristan}{Updated the definition of S$_1$ to explain that multiple engine instances have to be available in the same sg instance.}
 (\texttt{S$_1$}=2). 
@@
 may implement any kind of scheduling policy (\texttt{S$_3$}=0).
-\closednote{sil}{so same as the others = "easy"? i think this requires more explanation. maybe we should introduce a score for "dont know"?}{Tristan}{Fixed as discussed in PR #20}
 
@@
 gateway framework to execute different types of workflows and link
-them together.  \closednote{sil}{why not? if the engine is ws-pgrade
-  it does. i really think that meta-workflows (and nested workflows)
-  is a property exclusively of the engine. you can see that in fig 2.V
-  interaction *6}{Tristan}{If the engine is WS-PGRADE as used in SHIWA
-  CGI, then the architecture is nested workflows with service
-  invocation. In a tight integration, meta-workflows would be
-  expressed, e.g. as compositions of Liferay portlets that all
-  implement different types of engines.}  The science gateway can
+them together.  The science gateway can
 retrieve fine-grained debugging information from the workflow engine
@@
 engines is usually not possible unless the workflow engine has
-specific abilities (\texttt{S$_2$}=1).  \closednote{sil}{is this score for
-  VIP? i think that the score depends on the implementation of the
-  science gateway, which could adopt some elastic mechanism within. so
-  we should not say s_2=1 for the architecture}{Tristan}{Removed VIP example.
-Updated score definition according to the discussion in @SilviaOlabarriaga's PR (\#20).} The architecture does not add any particular complexity to the task scheduling problem
-since the workflow engine may implement any kind of scheduling
-policy (\texttt{S$_3$}=0).  \closednote{sil}{see comment in section 3.1}{Tristan}{Fixed the same way}
+specific abilities (\texttt{S$_2$}=1).
 
@@
 workflows to execute other workflows.  (\texttt{O$_1$}=1).
-\closednote{sil}{see comment in section 3.1}{Tristan}{same answer} The
+ The
 science gateway needs to invoke interaction \texttt{d} to retrieve
 dine-grained debugging information from the workflow service
-(\texttt{O$_2$}=1).  \closednote{sil}{this is really not the case. the
-  fact that the wf service is a separate component that hides
-  information is a big problem for debugging at the science gateway
-  level.  we had a paper about this:
-  @inproceedings{olabarriaga2014scientific, Author = {Olabarriaga,
-      Silvia and Pierantoni, Gabrielle and Taffoni, Giuliano and
-      Sciacca, Eva and Jaghoori, Mahdi and Korkhov, Vladimir and
-      Castelli, Giuliano and Vuerli, Claudio and Becciani, Ugo and
-      Carley, Eoin and others}, Booktitle = {e-Science (e-Science),
-      2014 IEEE 10th International Conference on}, Organization =
-    {IEEE}, Pages = {298--305}, Title = {Scientific Workflow
-      Management--For Whom?}, Year = {2014}} }{Tristan}{I agree. Fixed
-  it and precised the definition of O2.}
+(\texttt{O$_2$}=1).
 
@@
 (\texttt{R$_2$}=1).
-\closednote{vladimir}{there is a large piece of text in Scalability about reliability of the workflow engine task which on my opinion suits Robustness better}{Tristan}{I understand that that text may sound like robustness, but the robustness metrics are about the architecture, not the tasks, as explained in 3.1 (Errors coming from
-the infrastructure (e.g., unavailable data or termi-
-nated tasks) or workflows (e.g., wrong user input or
-application errors) are not covered).}
-\closednote{vladimir}{in addition, the claim of very good robustness of task encapsulation seems questionable to me as in this model you delegate the critical task of controlling workflow execution to, let's say, some 'random' node on the grid that you have no control of, not sure about its performance, connectivity and reliability and so on; however, considering the models and dependencies presented in fig.2 this evaluation of robustness metric may be correct}{Tristan}{You are right but the robustness issues that you are talking about are related to the infrastructure and not the architecture.
-This is really an issue related to task scheduling as the SG might decide to run the workflow tasks on reliable nodes of the infrastructure such as a dedicated cluster.
-So the issue that you mention is reflected in S3.}
 
@@
 task fails. The workflow task is also longer than all its sub-tasks,
-which increases its chances of failure.
-\closednote{vladimir}{isn't this about robustness?}{Tristan}{Nope, this is task scheduling, therefore scalability in our evaluation framework.
-See answer to your previous note.} In addition, task parameters,
+which increases its chances of failure. In addition, task parameters,
 for instance estimated walltime, are more difficult to estimate for
@@
 sub-tasks (interaction \texttt{e}), without any additional
-development. \closednote{sil}{see comment in section
-  3.1}{Tristan}{Fixed, actually updated the score of this architecture
-  now that the definition is more accurate.} Fine-grained debugging
+development. Fine-grained debugging
 information is obtained through interaction \texttt{b}
-(\texttt{O$_2$}=1). \closednote{sil}{mm, i think the case here is the
-  same as for service. or actually i would give this one a better
-  score, because there are less components to cross}{Tristan}{Fixed the text to better match the updated definition of O2}
+(\texttt{O$_2$}=1).
 
@@
 developments in the workflow engine to enable workflow submission
-(\texttt{O$_1$=1}).  \closednote{sil}{see comment in section 3.1}{Tristan}{Same answer.} Debugging information is accessed through interactions \texttt{d} and \texttt{f} (\texttt{O$_2$=1}).
-\closednote{sil}{i think the case here is the same as for service. or
-  actually i would give this one a worse score, because there is one
-  additional component - the pool - to cross}{Tristan}{Fixed using the updated definition of O2}
+(\texttt{O$_1$=1}). Debugging information is accessed through
+interactions \texttt{d} and \texttt{f} (\texttt{O$_2$=1}).
 
@@
 \texttt{d} (twice).
-\closednote{vladimir}{and if we have a meta-workflow with 3 and more sub-workflows the score will be even higher, is it a fair calculation for an architecture?
-we did not increase the scores for possible meta-workflows supported by wf engines in other cases}{Tristan}{Good point!
-You are totally right for meta-workflows, I added a sentence in the paragraph below.
-However, nested workflows are studied here as an architecture to integrate a workflow engine (and only one), not to describe meta-workflows.}
 
@@
 conversion, by connecting workflows in the language used in the
-science gateway (\texttt{O$_1$=0}).  \closednote{sil}{i dont
-  understand this about meta-workflows. translation does not imply
-  composition, right?  if the science gateway is allowed to implement
-  workflow composition, then this is probably the case in other
-  architectures as well?}{Tristan}{Well at this level, connecting two
-  workflows expressed in the same language is straightforward,
-  although I agree that it might be tricky in some cases.} Debugging information is accessible
+science gateway (\texttt{O$_1$=0}).  Debugging information is accessible
 as in the service invocation architecture (\texttt{O$_2$=1}).
-\closednote{sil}{see notes 3.2}{Tristan}{Fixed the same way}
 
@@
 science gateways with important workloads, which would favor task
-encapsulation and pool. \closednote{Naj}{I feel this paragraph is a bit
-  vague, and doesn't distill and contextualize the thorough work
-  described before.}{Tristan}{Rejected comment :)}
+encapsulation and pool.
 
@@
 workflows (with service invocation) and workflow conversion (with service
-invocation) are the only architectures that intrinsically support meta-workflows.
-\closednote{vladimir}{As mentioned earlier by Silvia, this might not be exactly the case, depending on the point of view}{Tristan}{The example given by Silvia (WS-PGRADE) actually comes down to the nested workflow architecture.
-I added 'intrinsically' to the sentence to better reflect that.} Besides,
+invocation) are the only architectures that intrinsically support meta-workflows. Besides,
 tight integration is the only architecture that allows
@@
 used. Nevertheless, it could be interesting to explore other types of
-instantiations.\closednote{Naj}{I think in the discussion, you may want to close the circle on a statement in introduction, which referred to your 7-year experience with CBRAIN and VIP}{Tristan}{Not sure how to do this without making the conclusion sound too specific to these systems.}
+instantiations.
 
@@
 integrate workflow engines. For instance, many workflow engines are
-already available as web services\closednote{Naj}{This is a good place to offer some examples}{Tristan}{Many of them are, precised the text, not sure adding explicit examples makes sense at this stage.}, which tends to favor service
+already available as web services, which tends to favor service
 invocation, and other science gateways may have strongly tested task
@@
 are not identified.
-
-\closednote{vladimir}{@silvia, does it make sense to mention workflow classification efforts based on workflow motifs and so on?}{Tristan}{I'm  not  sure. Which papers are you thinking of?}
 

### 2016-05-15 paper.tex
Fixed a few typos and inconsistencies

@@
 these tools also support advanced features such as provenance
-tracking, or  redundancy detection across
+tracking or  redundancy detection across
 analyses to avoid re-computation. A wide array of workflows
@@
 that may consist of files, database entries or simple parameter
-values. Although simplistic, this definition covers both e-Science workflow systems and
+values. Although simplistic, this definition covers both e-Science workflow engines and
 domain-specific pipeline systems.
@@
 characteristics, such as the presence of a shared file system between
-the computing nodes, the availability of a global task meta-scheduler,
-the presence of a file catalog, etc. In the analysis presented below,
+the computing nodes, the availability of a global task meta-scheduler, or
+the presence of a file catalog. In the analysis presented hereafter,
 such specific requirements are not discussed. Instead, we consider the
@@
 The interactions involved in the architectures are described below and
-labeled (\texttt{a},\texttt{b},\texttt{c}, \ldots) as in Figure~\ref{fig:architectures}.
+labeled from \texttt{a} to \texttt{g} as in Figure~\ref{fig:architectures}.
 
@@
   (\texttt{a$_1$}) the programs used in the workflow are installed on
-  the infrastructure or prepared for on-the-fly installation, which
+  the infrastructure or prepared for on-the-fly deployment, which
   may require specific privileges; (\texttt{a$_2$}) the
@@
 since it is used to authenticate users, add new workflows, transfer
-user data, etc., and
+user data, etc.
 \item Specific interactions (\texttt{R$_2$}): number of specific
@@
 scalability difficulties due to (partial) absence of a given feature in the architecture.  
-Features are evaluated using a 3-level metric: 0 means that
-the feature is very easy to enable, 1 means that it can be implemented
-but with some difficulty, and 2 means that the feature cannot
- be implemented in a realistic scenario. Three different features are identified:
+ Three different features are identified:
 \begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
@@
   implemented using the components and interactions already in place),
-  amd \texttt{O$_1$}=1 means that they may be implemented with some
+  and \texttt{O$_1$}=1 means that they may be implemented with some
   development effort.
@@
 science gateway needs to invoke interaction \texttt{d} to retrieve
-dine-grained debugging information from the workflow service
+fine-grained debugging information from the workflow service
 (\texttt{O$_2$}=1).
@@
 \paragraph{Scalability} By design, new engine instances only require
-new agents, which can be easily automated (\texttt{S$_1$=0}). For
-instance, auto-scaling rules can be implemented to start new agents
-when the workload in the science gateway exceeds a certain
-threshold~\cite{lorido2012auto}. Distributed engines are not available
+new agents, which can be easily automated (\texttt{S$_1$=0}). Distributed engines are not available
 by default (\texttt{S$_2$=1}) and the architecture does not add any
@@
 types involved in meta-workflows: for instance, a meta-workflow with
-child workflows executed by 2 different engine types would require an
+child workflows executed by 2 different types of engines would require an
 additional workflow service and the corresponding interactions.
@@
 tight integration is the only architecture that allows
-fine-grained debugging. 
-The availability of fine-grained debugging information may be
+fine-grained debugging, which may be
 critical to efficient user support.
@@
 architectures, based on the metrics in Table~\ref{table:evaluation}.
-Note that this analysis is only meant for illustration purposes, since
+Note that this analysis is only meant for illustration purposes since
 it assumes that all criteria have equal weights while real systems
@@
 conversion) is clearly more complex than interaction \texttt{d}
-(service invocation), and may challenge reliability more strongly. However, without entering into the details of a particular
+(service invocation). However, without entering into the details of a particular
 system, quantification of the robustness or the amount of development
@@
 We have systematically reviewed architectures used to integrate
-workflow engines in science gateways. These architectures are
+workflow engines in science gateways. These architectures were
 described in a system-independent framework suitable for comparison,
@@
 To the best of our knowledge, our work is the first to systematically
-review and compare software architectures used to integrate workflow
+review and compare software architectures  to integrate workflow
 engines into science gateways. So far, the literature on science gateways and
@@
 LABX-0063) of Universit\'e de Lyon, within the program
-``Investissements d’Avenir'' (ANR-11-IDEX-0007) operated by the French
+``Investissements d'Avenir'' (ANR-11-IDEX-0007) operated by the French
 National Research Agency (ANR). This work also falls into the scope of

### 2016-05-15 paper.tex
Fixed #25

@@
             Apache Airvata~\cite{marru2011apache}, Neuroscience Gateway~\cite{shahand2015data}$^{w,n}$, HubZero with Pegasus~\cite{CPE:CPE3257}, MoSGrid~\cite{kruger2014mosgrid}$^w$, System in~\cite{wu2010accelerating}, Vine Toolkit~\cite{DBLP:journals/scpe/SzejnfeldDKKKKLPTWDNW10}, Virtual Imaging Platform~\cite{GLAT-13}$^n$, WS-PGRADE/gUSE framework~\cite{Kacsuk2012},
-            Science gateways in~\cite{kacsuk2014science}$^w$
+            Science gateways in~\cite{kacsuk2014science}$^w$, GridSFEA~\cite{Elts2010}
             } \\
@@
 Virtual Imaging Platform, where the MOTEUR workflow
-engine~\cite{GLAT-08i} is invoked through a service.
+engine~\cite{GLAT-08i} is invoked through a service. GridSFEA~\cite{Elts2010} 
+uses WS-VLAM as a WS-RF service.
 

### 2016-05-15 paper.tex
minor sections 1, 2

--- 2016-05-15 paper.tex
@@
 \address[isi]{University of Southern California, Information Sciences Institute, Marina del Rey, CA, USA.}
-\address[amc]{Academic Medical Center of the University of Amsterdam, Department of Clinical Epidemiology, Biostatistics and Bioinformatics, Amsterdam, NL.}
+\address[amc]{Dept. Clinical Epidemiology, Biostatistics and Bioinformatics, Academic Medical Center, University of Amsterdam, NL.}
 \address[stpetersburg]{St. Petersburg State University, Russia.}
@@
 architectures can be adopted to integrate workflow engines in science
-gateways, with important consequences on the development effort
+gateways, with important consequences for the development effort
 required and resulting system.
@@
  by our experience in the development and
-sustained operation of the CBRAIN~\cite{SHER-14} and
-VIP~\cite{GLAT-13} science gateways for medical image analysis during
+sustained operation of the CBRAIN~\cite{SHER-14},
+VIP~\cite{GLAT-13} and NSG~\cite{shahand2015data,shahand:2012jgc} science gateways for medical image analysis during
 the past 7 years, as well as by lessons learned from several science
@@
 WS-VLAM~\cite{wsvlam}.  Such workflow engines usually describe
-applications in a high-level language with specific data and control
+applications using a high-level language with specific data and control
 flow constructs, parallelization operators, visual edition tools,
@@
 \emph{workflow engine} (also abbreviated \emph{engine}) is a piece of
-software that submits interdependent computing \emph{tasks} to an
+software that coordinates and submits interdependent computing \emph{tasks} to an
 \emph{infrastructure} (local server, cluster, grid or cloud) based on
@@
 
-Infrastructure consists of the computing and storage
+An infrastructure consists of the computing and storage
 resources involved in workflow execution, as well as the software
-services used to access these resources. Infrastructure can be
+services used to access these resources. The infrastructure can be
 composed of computing or file servers, databases, clusters, grids or
@@
 \item Task control: operations to manage tasks on the infrastructure,
-  including: authentication, submission, monitoring, termination,
-  deletion, etc. Controlling tasks requires to deal with the
+  including authentication, submission, monitoring, termination,
+  deletion, etc. Controlling tasks requires dealing with the
   heterogeneous batch managers and meta-schedulers that might be
   available on the infrastructure. When the infrastructure is a grid
-  or a cloud, it may for instance be achieved using libraries that
+  or a cloud, control may be achieved for instance using libraries that
   implement standards such as SAGA (Simple API for Grid
@@
 \item Data control: operations to manage data on the infrastructure,
-  such as: upload, download, deletion, browsing, replication, caching,
+  such as upload, download, deletion, browsing, replication, caching,
   etc. Data movements can be triggered by the user in the science
@@
   intermediate results) across the infrastructure so that tasks can
-  use it. The infrastructure might offer various data storage backends
+  use them. The infrastructure might offer various data storage backends
   with heterogeneous interfaces. 
@@
 \item Workflow control: operations to control workflow execution in an
-  engine, including: workflow submission, monitoring, termination,
+  engine, including workflow submission, monitoring, termination,
   etc. Workflow control can be coarse-grained (black box) or
@@
 data control are both initiated from the science gateway. Interactions
-\texttt{b} and \texttt{c$_2$} are initiated from the workflow engine
+\texttt{b} and \texttt{c$_2$} are initiated from the workflow engine,
 while \texttt{c$_1$} comes from other parts of the science gateway,

### 2016-05-15 paper.tex

@@
 other specific features. These are based on science gateway properties
-introduced in Section~\ref{sec:intro_sg}. Criteria break down to
+introduced in Section~\ref{sec:intro_sg}. Criteria break down into
 specific metrics where \emph{lower value indicates better
   performance}. Colors in Table~\ref{table:evaluation} represent
-normalized scores, as explained in the Table caption. For each
+normalized scores (see caption). For each
 criterion, a global score is computed by summing up the individual
@@
 Science gateway \emph{integration complexity} is measured through the number of software components and their interactions. It breaks
-down to the following 2 metrics:
+down into the following 2 metrics:
 \begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
@@
 parameters involved in this interaction are modified.  Extensibility
-breaks down to 4 metrics depending on the type of element (engine, workflow or infrastructure) that has to
+breaks down into 4 metrics depending on the type of element (engine, workflow or infrastructure) that has to
 be added or replaced:
@@
 difficulty to extend the architecture. Note that some extensions may
-involve several metrics in practice. For instance, adding a new type
+affect several metrics in practice. For instance, adding a new type
 of engine may help integrating new infrastructures when interactions
@@
   engine instances in the architecture. \texttt{S$_1$}=1 means that
-  multiple engine instances requires specific developments in the
+  supporting multiple engine instances requires specific developments in the
   science gateway or workflow engine. \texttt{S$_1$}=2 means that new
@@
 \item Task scheduling (\texttt{S$_3$}): measures the complexity
-  added by the architecture to task scheduling on
+  added by the architecture to implement task scheduling on
   the infrastructure. Task
@@
 fine-grained debugging, which may be
-critical to efficient user support.
+critical for efficient user support.
 
@@
 architectures, based on the metrics in Table~\ref{table:evaluation}.
-Note that this analysis is only meant for illustration purposes since
+Note that this analysis is only meant for illustration purposes, since
 it assumes that all criteria have equal weights while real systems
@@
 practice. Service invocation may also be combined with task
-encapsulation in some cases. However, the criteria discussed in this
-work would still apply to broadly compare and categorize such cases of
+encapsulation in some cases. Nevertheless, the criteria discussed in this
+work would still be applicable for broadly comparing and categorizing such cases of
 hybrid architectures.
 
-Moreover, all the interactions involved in the architectures were
+Note also that all the interactions involved in the architectures were
 treated equally, whereas some of them are obviously more complex than
@@
 could also be envisaged. We chose to limit ourselves to instantiations
-with service invocation because the resulting architectures are
+with service invocation because the resulting architectures have already been
 implemented in real systems, and because service invocation is largely
@@
 The particular technical or historical context of a science gateway
-project may obviously influence the choice of an architecture to
+project may also influence the choice of an architecture to
 integrate workflow engines. For instance, many workflow engines are
@@
 Section~\ref{sec:introduction-engines}. A few works described the
-architecture and features of workflow system at an abstract level. For
+architecture and features of workflow systems at an abstract level. For
 instance, the early work~\cite{yuTaxonomy} proposes a generic
@@
 analysis abstracts and evaluates architectural patterns independently
-from any particular system. It provides general insight to integrate
-science gateways and workflow engines, which we hope will be useful
-for software architects.
+from any particular system, providing general insight about ways to integrate
+science gateways and workflow engines. These insights will be valuable
+for software architects when considering alternatives for the architecture and evaluating the impact of their decisions.
 

### 2016-05-16 paper.tex
Updates following today's contributions and feedbacks

@@
 \begin{abstract}
-  Workflow engines are critical for the efficient and transparent
-  exploitation of distributed infrastructures in the ecosystem of
-  tools and services offered by science gateways.  We investigate six
-  software architectures commonly used to integrate workflow engines
-  into science gateways. In \emph{tight integration}, the workflow
-  engine shares software components with the science gateway. In
+  Science gateways often rely on workflow engines to execute
+  applications on distributed infrastructures. We investigate six software
+  architectures commonly used to integrate workflow engines into
+  science gateways. In \emph{tight integration}, the workflow engine
+  shares software components with the science gateway. In
   \emph{service invocation}, the engine is isolated and invoked
@@
 
-Workflow engines are critical for the efficient and transparent
-exploitation of distributed infrastructures in the ecosystem of tools
+Workflow engines are critical components in the ecosystem of tools
 and services offered by science gateways. Several software
@@
 architectures, based on system-independent representations of their
-main components and interactions. It is informed
- by our experience in the development and
-sustained operation of the CBRAIN~\cite{SHER-14},
-VIP~\cite{GLAT-13} and NSG~\cite{shahand2015data,shahand:2012jgc} science gateways for medical image analysis during
-the past 7 years, as well as by lessons learned from several science
-gateway and workflow projects such as
+main components and interactions. It is informed by our experience in
+the development and sustained operation of the CBRAIN~\cite{SHER-14},
+VIP~\cite{GLAT-13} and NSG~\cite{shahand2015data,shahand:2012jgc}
+science gateways for medical image analysis during the past 7 years,
+as well as by lessons learned from several science gateway and
+workflow projects such as
 SHIWA\footnote{\url{http://www.shiwa-workflow.eu}} and
@@
 science gateways and infrastructures. In
-Section~\ref{sec:architectures}, we describe six architectures
- that cover
-the main patterns that we encountered in the various systems that we
-studied during the last few years. The architectures are described
-within a consistent framework that underlines the functional
+Section~\ref{sec:architectures}, we describe six architectures that
+cover the main patterns that we encountered in the various systems
+that we studied during the last few years. The architectures are
+described within a consistent framework that underlines the functional
 interactions between their main software components. In
@@
 gateways. We finally compare the architectures and highlight
-recommendations for science gateway architects and 
-developers.
+recommendations for science gateway architects and developers.
 
@@
 This architecture is largely adopted in systems such as Apache
-Airvata~\cite{marru2011apache}, Vine
+Airvata~\cite{marru2011apache}, GridSFEA~\cite{Elts2010}, Vine
 Toolkit~\cite{DBLP:journals/scpe/SzejnfeldDKKKKLPTWDNW10}, Virtual
@@
 in~\cite{wu2010accelerating,shahand:2012jgc} and the WS-PGRADE/gUSE
-framework~\cite{Kacsuk2012}. The integration between the
-HubZero science gateway and the Pegasus workflow engine performed
+framework~\cite{Kacsuk2012}. The integration between the HubZero
+science gateway and the Pegasus workflow engine performed
 in~\cite{CPE:CPE3257} also uses a service architecture, where the
@@
 Virtual Imaging Platform, where the MOTEUR workflow
-engine~\cite{GLAT-08i} is invoked through a service. GridSFEA~\cite{Elts2010} 
-uses WS-VLAM as a WS-RF service.
+engine~\cite{GLAT-08i} is invoked through a service.
 

### 2016-05-16 paper.tex
Fixes from @poquirion

@@
 engines rely on general-purpose programming languages such as Bash,
-Octave and Python, which are much richer than scientific workflow
+Octave/Matlab and Python, which are much richer than scientific workflow
 languages.

### 2016-10-06 answers.tex
Template letter answering the reviewers' comments.

\documentclass[a4]{letter}
\usepackage{graphicx}
\usepackage{color}
\usepackage{hyperref}
\name{Tristan Glatard}
\signature{\includegraphics[width=4cm]{signature.png}}
\date{October 16th, 2016}

\newenvironment{review}%          environment name
{\textbf{Reviewer comment:}\begin{quote}}% begin code
{\end{quote}}%  
\newcommand{\answer}[1]{\textbf{Answer:} #1}

\newcommand{\revised}[1]{\color{blue} #1\color{black}}

\begin{document}

\begin{letter}{}

\opening{Dear Editors,}

We would like to thank you for offering us the opportunity to revise
and re-submit our paper entitled ``Software architectures to integrate
workflow engines in science gateways'' to Future Generation Computer
Systems.

The remainder of this letter contains our answer to the 3 reviews that
we received following the evaluation process. A revised version of the
manuscript follows, where corrected parts appear in blue for a better
identification.

We thank you again for your time and consideration, and we remain available for any further information you may need.

\vspace{0.5cm}

Sincerely yours,

\vspace{1cm}

\closing{Tristan Glatard, PhD.\\On behalf of the authors.}

\newpage

\textbf{Reviewer \#1}

\begin{review}
  The work presents a comprehensive overview on six of the most
  typical approaches that influenced the design of existing science
  gateways. The initial description of the individual architectures is
  indeed useful as background information leading to the realisation
  of a new gateway system. However, as the authors acknowledge, the
  different solutions may coexist in real implementations and the
  abstract perspectives provided are subject to a certain level of
  generality. In the opinion of the reviewer, the evaluation provided
  may result, in some cases, misleading for the the non-expert
  designer.
\end{review}

\answer{}

\begin{review}
  Section 2: In the description of the Task Encapsulation pattern, it
  is not really clear what is intended as the interaction 'e' that
  envisage the submission of task to the science gateway. Are these
  tasks pure notifications about the progressing of the sub-task, or
  the actual workflow engine, deployed in the infrastructure, requires
  the gateway to directly control its own sub-tasks?
\end{review}

\answer{}

\begin{review}
Section 3:
The evaluation is conducted by introducing a novel framework of metrics. It provides measurements based on the number of components of the gateway and the number of formal interactions required to satisfy four general characteristics: Robustness of Workflow execution, Extensibility, Scalability and Specific Features. Some aspects are a bit confusing. For instance, it is not immediately clear from the text, the difference between the Scalability metrics S1 and S2. Eventually they seem to differ, respectively, for the focus on dynamic elasticity of the former, versus the pre-hoc allocation of multiple workflow instances of the latter.
\end{review}

\begin{review}
The metric E3, measuring the complexity of integrating more workflows applications, seems to be more functional rather than architectural. The implications and impacts on the formal interactions required to support this feature across patterns are very similar. On the other hand, is opinion of the reviewer that integrating new workflow applications would rather affect more the Robustness of the system, especially in the case of the Workflow Conversion pattern. In this context, a formal conversion interaction (g), if not validated, could lead to logical or system errors.
\end{review}

\begin{review}
 Given these observations and the other measurements provided by the paper, the overall performance presented in Table 3, where the Workflow Conversion outperforms the Service Invocation pattern, is considered counter intuitive. The result is justified though by the evaluation of the Service-invocation pattern. Here, a high value (low score) is assigned to the Extensibility metrics E1, concerning the inclusion
of new workflow engines. This seems to be too negative, since a service invocation approach can rely on generic interfaces, re-purposing via a common service layer supporting multiple technologies. This is reflected also in the negative evaluation of the S1,S2 metrics for the same pattern. On the contrary, the choice of this architecture would intuitively suggest re-usable, scalable and decoupled solutions. Eventually, the arguments presented were not considered convincing.
\end{review}

\begin{review}
Again on the definition of Robustness. It introduces two metrics: Specific Components and Interactions. These are explained in terms of what they not include but no evidence or examples are provided to better understand what is intended specific. Moreover, some interactions which are mentioned as being not specific, become specific during the evaluation. For instance, interaction 'c', which is defined as the upload of input data and results (c1) and to transfer processed data across infrastructures (c2), are initially considered as an example of not specific interaction. Though, when measuring the Robustness of the Pool Model,  c2 suddenly appears in the count,  Does the definition of R2 means that non-specific interactions should be counted? The higher number the worse robustness?
\end{review}

\begin{review}
It is not always obvious that Scalability could be badly affected in a Tight Integration. The example of a liferay portlet is actually potentially demonstrating the opposite. Portlets can be independent web application that can be even remotely deployed (via WSRP). The integration of a workflow engine via ad-hoc portlet applications could instead guarantee quite of a level of decoupling from the rest of the gateway. Here, the limitations highlighted seems to be a bit superficial.
\end{review}

\begin{review}
The "Specific Feature", referring to the Fine-grained debugging does not specify what is relevant to be monitored and who is going to gain benefit from the monitoring. The scientists, the gateway developers or the administrators? What is a white box, the workflow engine or the workflow task? Why multiple software layers affect the accuracy of the tasks' internals?
\end{review}

\begin{review}
To conclude, the general feeling is that the definition of the metrics and the interaction patterns are not rigorous enough to provide an interpretation of the final evaluation which is entirely useful for the reader. Probably, some more information on how "Future Generation" systems should be designed, starting from the general overview provided, with more technical understanding of the system complexities, could represent a more valuable contribution to the main aim of this particular publication.
\end{review}

\begin{review}
To improve the reading, Dedicating separate sections to the introduction of the metrics and the evaluation of each pattern would be beneficial. For instance Section 3 Metrics, Section 4 Evaluation.
\end{review}

\textbf{Reviewer \#2}

\begin{review}
  This work describes and evaluates six software architectures
  commonly used to integrate workflow engines into science gateways
  with metrics for assessment of integration complexity, robustness,
  extensibility, scalability and support for meta-workflows and
  fine-grained debugging. The paper is well written in general, and
  the motivation is very nice, clearly comparing the different types
  of architectures that can be found at the back end of a science
  gateway. However, the work has a list of issues that could be
  addressed by the authors:

1. Apache Araviate and WS-PGRADE/gUSE are not science gateways. They
are open source science gateway framework that enables users to access
computing infrastructures frameworks. Therefore, in the Table 1, where
the classifications of science gateways are displayed, it should list
the science gateways that have used those architectures instead of the
frameworks. Otherwise, is quite confusing.
\end{review}

\begin{review}
  2. Tight Architecture -→ What happens if the workflow engine is
  built into the Science Gateway, but still need a submission
  framework (e.g. WS-PGRADE/gUSE or SAGA) for submitting the enteir
  workflow to the infrastructure.  It would be a tight architecture or
  a task architecture ? I can not see this type of architecture
  reflected in the paper.
\end{review}

\begin{review}
  3. Why there isn't a figure to illustrate an example of a real
  science gateway that corresponds to this architecture, like the
  other types of archtictures (e.g. Figure 3, Figure 4)
\end{review}

\begin{review}
  4. In my opininion Figures 3, 4, 5, 6, should use the iteractions
  explained in the subction 2.1, instead the steps/calls, to be more
  clear. Right now, they are difficult to read.
\end{review}

\begin{review}
  5.  Subsection 2.4, "see for instrance attribute depend of option -W
  torque"-→ repharase, and explain it better.
\end{review}

\begin{review}
6.  Figures 4, 5, 6 are very difficult to understand. Maybe will be a
good idea to create a subsection for explaining them better and reduce
the captions of those figures.
\end{review}

\begin{review}
  7.  Metrics: The metrics that have been used in this paper favour
  the simplest architectures. For example, looking at the Robustnes
  metrics, R1 and R2, they measure the probability that a workflow
  fails due to the errors of components and interactions that contains
  each architecture. But those metrics do not measure how "grave" are
  those failures and how "difficult" are those failures to repair. The
  same thing for the "Extensibility" metrics, some components can be
  much more difficult to "replace" than others.  Maybe would be a good
  idea to ponderete those metrics based on the number of components or
  interactions (e.g E.1---> dividing the number of interactions or
  components that need to be modified to integrate a new workflow
  between the total number of interactions or components. )
\end{review}

\begin{review}
8. In my opinion, if the workflow it is integrated in the science
gateway (tight architecture), then the science gateway should be
evaluated as a component, since it also could fail. The same thing for
the task encapsulation and the infrastructure, in that case, the
Infraestructure should be also considered as component. So, I would
modifiy the table 2, according to that ( e.g., R1=1 for Tight and Task
architecture).
\end{review}

\begin{review}
9. According with Figure 2.I, this workflow architecture has 3
interactions b, c1 and c2, since data movement can be done by the
users and workflows in this type of architecture (those interactions
are not exlcuyents) so R2=3
\end{review}

\begin{review}
  10. The text of Sections 3.2, 3.3, 3.4, 3.5, 3.6 and 3.7 could be
  reduced, since most of the values are obvious from Figures 2.II and
  Table 2.
\end{review}

\begin{review}
  11. Service invocation and Task encapsulation can also be the same
  thing
\end{review}

\begin{review}
12. No so sure that tight integration are the most easy for
integrating a workflow engine, since in case a the workflow engine
needs to be modified, the whole science gateway has to be change
(e.g. re-programe a portlet in Liferay). On the other hand, with
Service invocation, the replacement could be very much more easier.
\end{review}

\textbf{Reviewer \#3}

\begin{review}
This paper provides a comparison of 6 different software architectures commonly used to integrate workflow engines into science gateways. For this comparison, the authors define seven different types of interactions to outline the differences in approach. The combination of components and interactions for the 6 overview models in Figure 2 that form the underlying basis for the comparisons.

The authors then provide a set of Metrics: complexity, robustness, extensibility, scalability \& specific features to compare and contrast the different software architectures.

Complexity, for example, is measured by counting the number of components and interactions that were defined and annotated in Figure 2.  Extensibility measures the difficulty to replace or add elements to the architecture e.g. a workflow engine, an engine version update, add a new workflow or add new infrastructure.

I liked this paper but I also struggled a lot with it too because ultimately, whilst this metric structure provides some rudimentary structure to quantify the various software architectures, a many software architecture perspectives are missing, and the results are not normalized against the functionality a system provides.

For example, when you normally think of software architectures, a good design is often one that is layered and componentized. There are multiple benefits of such an an approach over monolithic systems but using the metrics defined in this paper, good software practices to expose functionality is not really factored in. So, the result is that the more layers to components in the system, the worse it scores. So, these aspects of good software engineering or providing enhanced functionality seem penalized in this paper, as it stands. 
\end{review}

\begin{review}
An example where the logic fails I believe is in the Extensibility. Let's take two cases initially:
-	Pool architecture
-	Nested.
For pool this score is 9 but for nested the score is 12.5. So, using these metrics the pool architecture is easier to extend.  But, if this is based on the same interface (e.g. SHIWA), then the difficulty to extend each system seems quite similar, because basically a workflow engine needs to implement the SHIWA course grained ORE interface to allow it to work. If it implements this interface as a client and a server then it can both consume other workflows and be embedded by them. For the pool architecture in the most generic sense, a workflow would need to do both (to digest ORE bundles and to pass them on), and for a nested architecture, a workflow engine would need to do one of the other depending on whether it was calling another workflow or being invoked by one.  However, as I understand, the case here assumes there is no connectivity (passing of bundles) in the pool case, so it seems less complex to extend. So, in the end, it is not about how complex the extension is
it is about what you want to do.  For example, for the nested case, you could always assume one master workflow, then it would be simpler (but less flexible). So in the end this table could look very different depending on how the different systems are defined in scope, and you could potentially have various versions of either of these cases. 
\end{review}

\begin{review}
Adding functionality e.g. in this case to add the ability to pass bundles to one system and another is a positive (and complex) thing to accomplish. There should be a way of counting this and having this score higher than Tight, Service or task, or normalizing this functionality into the metrics to provide a more balanced view. The "Scalability" touches on this but "Functionality" could be another aspect here to consider.
\end{review}

\begin{review}
Another example is for complexity.  Again, the less components and interactions the system has the less complex it is. But in practice, it is not so simple. For example, the tight integration is where the workflow system is integration directly into the science gateway and into the infrastructure. Whereas this architecture has the least number of components and interactions of them all it does not mean it is less complex in practice. Integrating a workflow into science gateway and directly on top of the infrastructure depends very much on the tools they sit upon and the complexity of that integration. Such tight integrated systems therefore could be massive monolithic pieces of code that are extremely difficult and complex to develop and maintain.  Whereas a task-based approach or service based approach through breaking code out to different services and tasks may be far simpler to develop and maintain. 
\end{review}

\begin{review}
I therefore think that the metrics defined only provide a very high level structure for an evaluation of the different software architectures and they should discuss other factors and scenarios that can affect development. Some of these limitations are discussed in section 4.2 but I think an extra section discuss these before the definition of the metrics would be very helpful to set the scene correctly for the evaluation. Otherwise, the reader is left asking a number of questions during the evaluation section. I also think that a few more metrics e.g. "functionality" could be added to provide a more balanced view of the different approaches. Another useful extension of the paper could be to normalize the results against the functionality each approach provides (perhaps in a separate table) - this may help to provide a more balanced table that factors in functionality.
\end{review}

\end{letter}
\end{document}

### 2016-10-10 answers.tex
Started answering comments from reviewer 1

@@
 
-\answer{}
+\answer{We thank the reviewer for the thorough review and useful comments.
+
+  The results and discussion reported in our paper are indeed aimed at
+  an expert audience of science gateway and workflow engine
+  developers. The evaluation metrics have been designed based on our
+  long experience with both science gateways and workflow engines, and
+  we completely acknowledge that non experts may misinterpret the
+  results.
+
+  The fact that the patterns may be mixed in real implementations is,
+  in our opinion, another motivation for our abstraction effort as it
+  helps ``untangling'' the various architectural concepts that may be
+  present in a real system. This is particularly true for the abstract
+  architectures (nested workflows and workflow conversion) that may be
+  instantiated in various ways. Our abstractions highlight that other
+  instantiation models may be worth designing, for instance nested
+  workflows with task encapsulation, as mentioned in Section 4.2. in
+  the Discussion.
+
+  We added the following paragraph to the Introduction to reflect this discussion:}
+
+\revised{The analysis presented in the paper is intended for
+experts in science gateway and workflow engine design. It is an
+abstraction effort to identify and evaluate the fundamental
+architectural patterns that are encountered while integrating workflow
+engines and science gateways. In real systems, such patterns are often
+intertwined due to considerations related to the historical and
+technical context of software projects. The following work helps
+identify the key patterns in systems and highlight potential
+improvements.}
 
@@
 
-\answer{}
+\answer{The actual workflow engine deployed in the infrastructure
+  requires the gateway to directly control its own sub-tasks. We
+  revised the wording of Section 2.4 to clarify this point.}
 
 \begin{review}
-Section 3:
-The evaluation is conducted by introducing a novel framework of metrics.
-It provides measurements based on the number of components of the gateway and the number of formal interactions required to satisfy four general characteristics: Robustness of Workflow execution, Extensibility, Scalability and Specific Features.
-Some aspects are a bit confusing.
-For instance, it is not immediately clear from the text, the difference between the Scalability metrics S1 and S2.
-Eventually they seem to differ, respectively, for the focus on dynamic elasticity of the former, versus the pre-hoc allocation of multiple workflow instances of the latter.
+  Section 3: The evaluation is conducted by introducing a novel
+  framework of metrics. It provides measurements based on the number
+  of components of the gateway and the number of formal interactions
+  required to satisfy four general characteristics: Robustness of
+  Workflow execution, Extensibility, Scalability and Specific
+  Features. Some aspects are a bit confusing. For instance, it is not
+  immediately clear from the text, the difference between the
+  Scalability metrics S1 and S2. Eventually they seem to differ,
+  respectively, for the focus on dynamic elasticity of the former,
+  versus the pre-hoc allocation of multiple workflow instances of the
+  latter.
 \end{review}
+
+\answer{S1 denotes the ability of the system to start multiple engine
+  instances to execute different workflows, while S2 denotes the
+  possibility to execute a single workflow with multiple engines. We
+  revised the wording in the description of S1 (page 11) to clarify
+  this point. We find the current description of S2 clear enough on
+  this point: ``[S2] measures the ability to distribute the execution of
+  \emph{a single workflow} among different engine instances.''}
 
@@
 
+\answer{We introduced E3 precisely to account for the additional
+  complexity introduced by interaction (g) when adding new
+  workflows.
+
+It should be understood here that workflow conversion is
+  presented as a pattern to integrate workflow \emph{engines} in a
+  science gateway, through workflow translation to a language already
+  supported by the gateway. We believe that integrating new workflow
+  engines (therefore new workflows) through workflow conversion is a
+  strong architectural design choice that may, depending on the
+  availability of a reliable interaction (g), be avoided by
+  integrating workflow engines using a different pattern. We added the following sentence at the beginning of Section 2.7 to explain that:
+
+\revised{Workflow conversion is presented here as a pattern to
+integrate workflow \emph{engines} in a science gateway, through
+workflow format conversion from a native format to the science gateway
+format.}
+}
+
+We definitely agree with the reviewer that interaction (g) is
+  a tricky one that may lead to critical errors. We added the
+  following sentence to the paragraph on interaction (g) in Section
+  4.2. of the discussion:
+
+\revised{If not properly validated, workflow language conversion
+  could introduce critical errors impacting the robustness and
+  correctness of the execution. }
+
+This sentence complements the note already made in section 3.7:
+``Note, however,
+that implementing interaction g can require sub-
+stantial work depending on the complexity of the
+workflow language used by the new engine.``
+
 \begin{review}
@@
 The result is justified though by the evaluation of the Service-invocation pattern.
-Here, a high value (low score) is assigned to the Extensibility metrics E1, concerning the inclusion
-of new workflow engines.
-This seems to be too negative, since a service invocation approach can rely on generic interfaces, re-purposing via a common service layer supporting multiple technologies.
+\end{review}
+
+\answer{As discussed previously, we share the reviewer's concerns
+  about the feasibility of workflow conversion (see comments on
+  interaction (g) in Section 4.2). The performance of workflow
+  conversion is strongly related to the availability of a validated
+  interaction (g), which is very dependent on the particular workflow
+  engines under consideration. We believe that the issue raised by the
+  reviewer basically comes down to the weight given to software
+  components and interactions in the analysis. However, using
+  different weights for different components or interactions would
+  open an infinite number of configurations for discussion. For
+  instance, in some cases, the workflow engine already comes with a
+  workflow service which therefore doesn't have to be included. To
+  address this issue, we shared a spreadsheet at
+  \url{https://frama.link/LrPRsCQp} that allows editing the following
+  weights:
+  \begin{itemize}
+  \item Weight given to software components and interactions in
+  ``integration'', ``extensibility'' and ``robustness'' criteria.
+  \item Weight given to metrics in a criterion (e.g., E$_1$ vs. E$_2$ in
+  ``extensibility'').
+  \item Weight given to criteria in overall evaluation of Table 3.
+\end{itemize}
+The paper references this spreadsheet wherever equal weights are
+assumed. The readers and reviewers are invited to rely on this
+spreadsheet for any issue related to the importance given to a
+specific component, interaction, metric or criterion in the
+analysis. In the paper, we assumed that all weights equal 1, which we
+believe is a reasonable but limited choice.
+}
+
+\begin{review}
+  Here, a high value (low score) is assigned to the Extensibility
+  metrics E1, concerning the inclusion of new workflow engines. This
+  seems to be too negative, since a service invocation approach can
+  rely on generic interfaces, re-purposing via a common service layer
+  supporting multiple technologies.
+\end{review}
+
+\answer{The reviewer is right that re-purposing software components
+  and interactions can lead to revisit the evaluation. We updated the following sentence in section 4.2. to better acknowledge this point:}
+
+The particular technical or historical context of a science gateway
+project may also influence the choice of an architecture to
+integrate workflow engines. For instance, many workflow engines are
+already available as web services, which tends to favor service
+invocation (\revised{in particular when interactions can be re-purposed}), and other science gateways may have strongly tested task
+and data control features (interactions \texttt{b} and \texttt{c}),
+which would favor task encapsulation. Similarly, adding a new type of engine
+may facilitate integration of a new infrastructure when interactions
+\texttt{b} and \texttt{c$_2$} are already available. The migration
+cost between architectures has been ignored as well.
+
+In the specific case of service invocation, we agree with the reviewer
+that, when adding a new workflow engine, the workflow service or interaction \texttt{d} can always be repurposed depending on the context.
+We decremented E1 and added the following explanation in section 3.3.:
+
+\revised{However, when a new type of workflow engine is added, it is always possible to reuse either an existing available workflow service or interaction \texttt{d} (when a new workflow service has to be developed).
+Therefore, \texttt{E$_1$}=3}.
+
+\begin{review}
  This is reflected also in the negative evaluation of the S1,S2 metrics for the same pattern.
 On the contrary, the choice of this architecture would intuitively suggest re-usable, scalable and decoupled solutions.
+\end{review}
+
+\answer{While we agree with the reviewer that re-usability should
+  definitely be taken into account in E$_1$, we do not see how it
+  could help S$_1$ and S$_2$.
+
+  For S$_1$, starting new instances of the workflow service still
+  requires that a specific mechanism is implemented in the science
+  gateway to distribute the load among these services, even if all
+  these services share the same interface. Again, the difficulty to
+  implement such mechanisms could be extensively discussed: some
+  systems may rely on a simple DNS-level load balancing, but this
+  would not cope with the fact that workflow executions have different
+  resource requirements and that workflow services may have different
+  availabilities (for instance, a workflow service may crash in case a
+  workflow execution filled up its disk); other systems may balance
+  the load based on resource usage (e.g. RAM, CPU and disk) in the
+  workflow services, but this requires an elaborated monitoring
+  infrastructure. On the contrary, adding new engine instances is
+  naturally supported by task encapsulation and pool, which justifies
+  that S$_1=0$ for these models while S$_1$=1 for service invocation.
+
+  For S$_2$, we cannot think of a practical reason why the service
+  invocation model would help distribute a single workflow execution
+  in multiple engine instances. In our opinion, this feature is
+  exclusive to the nested workflow model. }
+
+\begin{review}
  Eventually, the arguments presented were not considered convincing.
@@
 
+We hope that the update made in the extensibility scores and the
+explanations given for scalability convinced the reviewer.
+
 \begin{review}
-Again on the definition of Robustness.
-It introduces two metrics: Specific Components and Interactions.
-These are explained in terms of what they not include but no evidence or examples are provided to better understand what is intended specific.
-Moreover, some interactions which are mentioned as being not specific, become specific during the evaluation.
-For instance, interaction 'c', which is defined as the upload of input data and results (c1) and to transfer processed data across infrastructures (c2), are initially considered as an example of not specific interaction.
-Though, when measuring the Robustness of the Pool Model,  c2 suddenly appears in the count,  Does the definition of R2 means that non-specific interactions should be counted?
-The higher number the worse robustness?
+  Again on the definition of Robustness. It introduces two metrics:
+  Specific Components and Interactions. These are explained in terms
+  of what they not include but no evidence or examples are provided to
+  better understand what is intended specific. 
 \end{review}
+
+\answer{We updated the definition of R$_1$ and R$_2$ to include
+  positive examples as well as negative ones. In addition, the
+  specific components included in these metrics are represented in red
+  on Figure 2.}
+
+\begin{review}
+Moreover, some
+  interactions which are mentioned as being not specific, become
+  specific during the evaluation. For instance, interaction 'c', which
+  is defined as the upload of input data and results (c1) and to
+  transfer processed data across infrastructures (c2), are initially
+  considered as an example of not specific interaction. Though, when
+  measuring the Robustness of the Pool Model, c2 suddenly appears in
+  the count, 
+\end{review}
+
+\texttt{c}, \texttt{c$_1$} and \texttt{c$_2$} are three different
+interactions, which explains the confusion. \texttt{c$_2$} is always
+specific to workflow execution because it is defined as the
+interaction between the workflow engine and the infrastructure, used
+to transfer workflow data (inputs, outputs or intermediate
+results). \texttt{c$_1$} is not specific to workflow execution because
+it is used by the science gateway to transfer any data regardless of
+workflow executions. Letter \texttt{c} was used on the task
+encapsulation diagram in Figure 2 to represent that the science gateway transfers
+both user and workflow data. We now use \texttt{c$_1$} for all the
+data transfers initiated by the science gateway, and \texttt{c$_2$}
+for all the data transfers initiated by the workflow engine,
+regardless of the type of data transferred. Consequently, we now use
+\texttt{c$_1$} instead of \texttt{c} on the task encapsulation diagram in Figure 2.
+
+The ambiguity is now removed and the new definition of R$_1$ and R$_2$ reads as follows:
+
+Specific components (\texttt{R$_1$}): number of specific components involved in workflow executions. \revised{For instance, the
+  workflow service is a specific component because its main objective
+  is to execute workflows. On the contrary, the science gateway is
+  \emph{not} specific to workflow execution since it is used for a
+  variety of other functions such as user authentication, data
+  transfer, etc.}
+
+Specific interactions (\texttt{R$_2$}): number of specific
+  interactions involved in workflow executions. \revised{For
+    instance, the pool-agent interaction (interaction \texttt{f}) is
+    specific to workflow execution while data control (interaction
+    \texttt{c$_1$}) is \emph{not} because it is used by the science gateway to transfer
+    user data regardless of a particular workflow execution}.
+
+\begin{review}
+Does the definition of R2 means that non-specific
+  interactions should be counted?
+\end{review}
+
+\answer{No it doesn't. We hope that the new examples in the definition of
+R$_1$ and R$_2$ and the removal of any reference to interaction
+\texttt{c} clarifies this point.}
+
+\begin{review}
+The higher number the worse
+  robustness?
+\end{review}
+
+\answer{Yes, as for any other metric, high values indicate worse
+performance. This is explained at the beginning of section 3:
+``Criteria break down into specific met- rics where lower value
+indicates better performance.''}
 

### 2016-10-10 paper.tex
Updated affiliation
Started answering comments from reviewer 1

@@
 %\newcommand{\closedanswerednote}[6]{}
+\newcommand{\correction}[1]{\color{blue}#1\color{black}\xspace}
 
@@
 
-\author[mcgill,creatis]{Tristan Glatard}
+\author[concordia,mcgill,creatis]{Tristan Glatard}
 \author[mcgill]{Marc-\'Etienne Rousseau}
@@
 
+\address[concordia]{Department of Computer Science and Software Engineering, Concordia University, Montreal, Canada.}
 \address[mcgill]{McGill Centre for Integrative Neuroscience, Montreal Neurological Institute, McGill University, Canada.}
@@
 SHIWA\footnote{\url{http://www.shiwa-workflow.eu}} and
-ER-flow\footnote{\url{http://www.erflow.eu}}.  In this section we
-provide background information and definitions of workflow engines,
+ER-flow\footnote{\url{http://www.erflow.eu}}.
+
+\correction{The analysis presented in the paper is intended for
+experts in science gateway and workflow engine design. It is an
+abstraction effort to identify and evaluate the fundamental
+architectural patterns that are encountered while integrating workflow
+engines and science gateways. In real systems, such patterns are often
+intertwined due to considerations related to the historical and
+technical context of software projects. The following work helps
+identify the key patterns in systems and highlight potential
+improvements.}
+
+\correction{The reminder of this section provides}
+ background information and definitions of workflow engines,
 science gateways and infrastructures. In
@@
   such as upload, download, deletion, browsing, replication, caching,
-  etc. Data movements can be triggered by the user in the science
-  gateway (\texttt{$c_1$}), to upload input data or download processed
-  data. They can also be performed by the workflow engine
-  (\texttt{$c_2$}), to transfer workflow data (inputs, outputs, or
-  intermediate results) across the infrastructure so that tasks can
-  use them. The infrastructure might offer various data storage backends
-  with heterogeneous interfaces. 
-  Tools and services such as JSAGA~\cite{reynaud2010uniform} or Data Avenue~\cite{hajnal2014data}
+  etc. Data movements can be triggered \correction{by} the science
+  gateway (\texttt{$c_1$}), \correction{e.g.,} to upload input data or
+  download processed data. They can also be performed by the workflow
+  engine (\texttt{$c_2$}), to transfer workflow data (inputs, outputs,
+  or intermediate results) across the infrastructure so that tasks can
+  use them. The infrastructure might offer various data storage
+  backends with heterogeneous interfaces.  Tools and services such as
+  JSAGA~\cite{reynaud2010uniform} or Data Avenue~\cite{hajnal2014data}
   can be used to homogeneize these interfaces.
@@
 a particular task that can submit sub-tasks to the science gateway
-through interaction \texttt{e}. The workflow engine keeps track of the
-dependencies between the sub-tasks, but their execution is delegated
-to the science gateway that executes them on the infrastructure
-through interaction \texttt{b}. Although the science gateway has no
-global vision of the workflow, it can keep track of the sub-tasks
-submitted by a given task, for instance to cancel them when
-the task is canceled. The science gateway may also implement
+through interaction \texttt{e}. The workflow engine keeps track of
+the dependencies between the sub-tasks, but their execution is
+delegated to the science gateway that executes them on the
+infrastructure through interaction \texttt{b}. Although the science
+gateway has no global vision of the workflow, it completely \correction{controls}
+the sub-tasks. \correction{For instance, it may cancel} them when
+the \correction{workflow} is canceled. The science gateway may also implement
 mechanisms to facilitate the handling of task dependencies, for
@@
 The science gateway also transfers both user and workflow data across
-the infrastructure, so that interactions \texttt{c$_1$} and
-\texttt{c$_2$} are both covered by \texttt{c}.
+the infrastructure, through interaction \texttt{c$_1$}.
 
@@
 instantiated with the service invocation architecture for
-consistency. Workflow engines are integrated in the science gateway
-through workflow format conversion from a native format to the science
-gateway format. Workflow conversion
-is usually an offline process that is not involved in workflow
-execution. A few systems have implemented workflow conversion. In the
-SHIWA Simulation Platform, it is implemented through the IWIR
-language, which provides a common language for portability across grid
-workflow systems~\cite{plankensteiner-prodan-etal:2013} and allows
-conversion among $n$ workflow languages using $2n$ interactions
-instead of $n^2$.  Tavaxy~\cite{Abouelhoda2012} enables the import,
-merging and execution of Taverna~\cite{oinn2004taverna} and
-Galaxy~\cite{goecks2010galaxy} workflows; when some workflow parts
-cannot be imported, workflows are run by Tavaxy as nested workflows
-using their native engine. The work in~\cite{delaGarza2016} describes
-workflow conversion from KNIME~\cite{Berthold2008} to WS-PGRADE/gUSE
-and from Galaxy~\cite{goecks2010galaxy} to WS-PGRADE/gUSE.
+consistency. \correction{Workflow conversion is presented here as a pattern to
+integrate workflow \emph{engines} in a science gateway, through
+workflow format conversion from a native format to the science gateway
+format.} Workflow conversion is usually an offline process that is not
+involved in workflow execution. A few systems have implemented
+workflow conversion. In the SHIWA Simulation Platform, it is
+implemented through the IWIR language, which provides a common
+language for portability across grid workflow
+systems~\cite{plankensteiner-prodan-etal:2013} and allows conversion
+among $n$ workflow languages using $2n$ interactions instead of $n^2$.
+Tavaxy~\cite{Abouelhoda2012} enables the import, merging and execution
+of Taverna~\cite{oinn2004taverna} and Galaxy~\cite{goecks2010galaxy}
+workflows; when some workflow parts cannot be imported, workflows are
+run by Tavaxy as nested workflows using their native engine. The work
+in~\cite{delaGarza2016} describes workflow conversion from
+KNIME~\cite{Berthold2008} to WS-PGRADE/gUSE and from
+Galaxy~\cite{goecks2010galaxy} to WS-PGRADE/gUSE.
 
@@
   performance}. Colors in Table~\ref{table:evaluation} represent
-normalized scores (see caption). For each
-criterion, a global score is computed by summing up the individual
-scores for each metric. We ensure that the different metrics used to
-calculate the global score for a criterion are comparable, so that
-they can be combined meaningfully. The criteria and metrics are
+normalized scores (see caption). For each criterion, a global score is
+computed by summing up the individual scores for each metric. We
+ensure that the different metrics used to calculate the global score
+for a criterion are comparable, so that they can be combined
+meaningfully. \correction{All the metrics in a criterion contribute
+  equally to the global score. If this is not the case, the
+  reader is invited to adjust the weights in the public spreadsheet available
+  at \url{https://frama.link/LrPRsCQp} (column B of sheet ``Metrics'')
+  to generate custom versions of Table~\ref{table:evaluation}.}
+
+The criteria and metrics are
 explained hereafter.
@@
   New engine type -- \texttt{E$_1$}  & \cellcolor[HTML]{99C599}3
-                                     & \cellcolor[HTML]{99A899}4
+                                     & \cellcolor[HTML]{99A899}\correction{3}
                                      & \cellcolor[HTML]{99E299}2
@@
   \textbf{Total} (difficulty to extend) & \cellcolor[HTML]{99D299}\textbf{10}
-                                     & \cellcolor[HTML]{99D299}\textbf{10}
+                                     & \cellcolor[HTML]{99D299}\textbf{\correction{9}}
                                      & \cellcolor[HTML]{99FF99}\textbf{8}
@@
 infrastructure into the system usually requires some technical
-effort (e.g., account creation, software installation, APIs, etc.), therefore increasing complexity. The two metrics are summed to obtain a
+effort (e.g., account creation, software installation, APIs, etc.), therefore increasing complexity.
+\correction{We assume that software components contribute equally to $I_1$ and that interactions also contribute equally to $I_2$.
+In a real system, this may not be the case due to a range of project-specific practical reasons that cannot be covered in this paper.
+The weights given to components and interactions can be dynamically changed in the public spreadsheet (sheet ``Components and interactions'') to generate new versions of Table~\ref{table:evaluation} as required.} The two metrics \correction{I$_1$ and I$_2$} are summed to obtain a
 global score that measures the
@@
 \begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
-\item Specific components  (\texttt{R$_1$}): number of specific components
-involved in workflow executions. The Science
-gateway component, for instance, is \emph{not} specific to workflow execution
-since it is used to authenticate users, add new workflows, transfer
-user data, etc.
+\item Specific components (\texttt{R$_1$}): number of specific
+  components involved in workflow executions. \correction{For instance, the
+  workflow service is a specific component because its main objective
+  is to execute workflows. On the contrary, the science gateway is
+  \emph{not} specific to workflow execution since it is used for a
+  variety of other functions such as user authentication, data
+  transfer, etc.}
 \item Specific interactions (\texttt{R$_2$}): number of specific
-  interactions involved in workflow executions. Data control
-  (interaction \texttt{c}), for instance, is \emph{not} specific to
-  workflow execution since it is also used by the science gateway to
-  transfer user data. 
+  interactions involved in workflow executions. \correction{For
+    instance, the pool-agent interaction (interaction \texttt{f}) is
+    specific to workflow execution while data control (interaction
+    \texttt{c$_1$}) is \emph{not} because it is used by the science gateway to transfer
+    user data regardless of a particular workflow execution}.
 \end{itemize}
@@
 the total number of software pieces that are specifically involved in
-the workflow execution.
+the workflow execution. \correction{The weights given to software components in R$_1$ and interactions in R$_2$ can be adjusted in the public spreadsheet.}
 
@@
 The four metrics are summed to obtain a global score that measures the
-difficulty to extend the architecture. Note that some extensions may
+difficulty to extend the architecture. \correction{The weights given to software components and interactions in E$_{1-4}$ can be adjusted in the public spreadsheet.} 
+Note that some extensions may
 affect several metrics in practice. For instance, adding a new type
@@
 \item Multiple engine instances (\texttt{S$_1$}): measures the
-  potential for a single gateway instance to use more than one engine
-  instance simultaneously. Workflow engines may require important
-  amounts of resources when several workflows, or large workflows, are
-  executed. At some point, it may be required for the science gateway
-  to distribute the load among several engines. \texttt{S$_1$}=0 means
-  that
-  adding a new engine instance is inherently supported in the
-  architecture. In this case, elastic engines can be implemented
-  through some kind of auto-scaling mechanism to control the number of
-  engine instances in the architecture. \texttt{S$_1$}=1 means that
-  supporting multiple engine instances requires specific developments in the
-  science gateway or workflow engine. \texttt{S$_1$}=2 means that new
-  engine instances cannot be added.
-\item Distributed engines (\texttt{S$_2$}): measures the potential of distributing
+  \correction{ability} for a single gateway instance to use more than
+  one engine instance simultaneously, \correction{\emph{each workflow
+      being executed by a single engine instance}}.  Workflow engines
+  may require important amounts of resources when several workflows,
+  or large workflows, are executed. At some point, it may be required
+  for the science gateway to distribute the load among several
+  engines. \texttt{S$_1$}=0 means that adding a new engine instance is
+  inherently supported in the architecture. In this case, elastic
+  engines can be implemented through some kind of auto-scaling
+  mechanism to control the number of engine instances in the
+  architecture. \texttt{S$_1$}=1 means that supporting multiple engine
+  instances requires specific developments in the science gateway or
+  workflow engine. \texttt{S$_1$}=2 means that new engine instances
+  cannot be added.
+\item Distributed engines (\texttt{S$_2$}): measures the \correction{ability to distribute}
   the execution \emph{of a single workflow} among different engine
@@
 interaction \texttt{d}, and  implementing interactions \texttt{b} and
-\texttt{$c_2$} (\texttt{E$_1$}=4). New engine versions can be added by
+\texttt{$c_2$}.
+\correction{However, when a new type of workflow engine is added, it is always possible to reuse either an existing available workflow service or interaction \texttt{d} (when a new workflow service has to be developed).
+Therefore, \texttt{E$_1$}=3}.
+New engine versions can be added by
 updating the workflow service without modifying any interaction or
@@
 it assumes that all criteria have equal weights while real systems
-would favor some of them. Overall, task encapsulation performs a bit
+would favor some of them. \correction{Weights can be adjusted in the
+  public spreadsheet at \url{https://frama.link/LrPRsCQp} (column B of
+  sheet ``Metrics'') to produce a custom version of
+  Table~\ref{table:overall} if required.}
+
+ Overall, task encapsulation performs a bit
 better than the other architectures and nested workflows stand a bit
@@
   Extensibility (global score)  & \cellcolor[HTML]{99D299}{0.44}
-                                     & \cellcolor[HTML]{99D299}{0.44}
+                                     & \cellcolor[HTML]{99D299}{\correction{0.22}}
                                      & \cellcolor[HTML]{99FF99}{0.00}
@@
                                     & \cellcolor[HTML]{99DA99}\textbf{1.6}
-                                    & \cellcolor[HTML]{99B899}\textbf{2.6}
+                                    & \cellcolor[HTML]{99B899}\textbf{\correction{2.4}}
                                     & \cellcolor[HTML]{99FF99}\textbf{0.5}
@@
 engines rely on general-purpose programming languages such as Bash,
-Octave/Matlab and Python, which are much richer than scientific workflow
-languages.
+Octave/Matlab and Python, which are much richer than scientific
+workflow languages. \correction{If not properly validated, workflow
+  language conversion could introduce critical errors impacting the
+  robustness and correctness of the execution. }
 
@@
 already available as web services, which tends to favor service
-invocation, and other science gateways may have strongly tested task
+invocation (\correction{in particular when interactions can be re-purposed}), and other science gateways may have strongly tested task
 and data control features (interactions \texttt{b} and \texttt{c}),

### 2016-10-11 answers.tex
More answers to the reviewers.

@@
-\documentclass[a4]{letter}
+\documentclass[a4]{article}
 \usepackage{graphicx}
 \usepackage{color}
+\usepackage[x11names,dvipsnames,table]{xcolor} %for use in color links
+
 \usepackage{hyperref}
-\name{Tristan Glatard}
-\signature{\includegraphics[width=4cm]{signature.png}}
-\date{October 16th, 2016}
-
 \newenvironment{review}%          environment name
@@
 \newcommand{\answer}[1]{\textbf{Answer:} #1}
+\newcommand{\todo}[1]{\color{red}#1\color{black}}
 
@@
 
-\begin{letter}{}
-
-\opening{Dear Editors,}
-
-We would like to thank you for offering us the opportunity to revise
-and re-submit our paper entitled ``Software architectures to integrate
-workflow engines in science gateways'' to Future Generation Computer
-Systems.
-
-The remainder of this letter contains our answer to the 3 reviews that
-we received following the evaluation process. A revised version of the
-manuscript follows, where corrected parts appear in blue for a better
-identification.
-
-We thank you again for your time and consideration, and we remain available for any further information you may need.
-
-\vspace{0.5cm}
-
-Sincerely yours,
-
-\vspace{1cm}
-
-\closing{Tristan Glatard, PhD.\\On behalf of the authors.}
-
-\newpage
-
-\textbf{Reviewer \#1}
+\section{Reviewer \#1}
 
@@
   instantiation models may be worth designing, for instance nested
-  workflows with task encapsulation, as mentioned in Section 4.2. in
+  workflows with task encapsulation, as mentioned in Section 5.2. in
   the Discussion.
@@
   following sentence to the paragraph on interaction (g) in Section
-  4.2. of the discussion:
+  5.2. of the discussion:
 
@@
   about the feasibility of workflow conversion (see comments on
-  interaction (g) in Section 4.2). The performance of workflow
+  interaction (g) in Section 5.2). The performance of workflow
   conversion is strongly related to the availability of a validated
@@
 \answer{The reviewer is right that re-purposing software components
-  and interactions can lead to revisit the evaluation. We updated the following sentence in section 4.2. to better acknowledge this point:}
+  and interactions can lead to revisit the evaluation. We updated the following sentence in section 5.2. to better acknowledge this point:}
 
@@
 performance. This is explained at the beginning of section 3:
-``Criteria break down into specific met- rics where lower value
+``Criteria break down into specific metrics where lower value
 indicates better performance.''}
@@
 
+\todo{think of this}
+
 \begin{review}
@@
 
-\textbf{Reviewer \#2}
+\answer{This was done as suggested.}
+
+\section{Reviewer \#2}
 
@@
 
+\todo{just remove CSGF, Airvata and pgrade from the table, keeping the
+  exponents. But we don't have examples of gateways using Airvata,
+  still need to be found.}
+
 \begin{review}
@@
 \end{review}
+
+\answer{The task submission framework is part of the infrastructure
+  and its nature doesn't influence the architecture used to integrate
+  workflow engines in the science gateway. Formally, it is part of
+  interaction \texttt{b} (see definition in 2.1).
+
+  If the ``entire workflow'', i.e. the complete control and data flow,
+  is submitted to the infrastructure through interaction \texttt{b} as
+  a single task that will manage the execution and potentially submit
+  sub-tasks, then this is a task encapsulation architecture. In
+  particular, a dedicated engine instance can easily be started for
+  every workflow execution.
+
+  If only the workflow tasks are submitted to the infrastructure
+  through interaction \texttt{b}, and the main workflow logic is
+  controlled from within the science gateway, then this is clearly a
+  tight integration.
+
+  The reviewer's use case, however, suggest that the workflow engine
+  is split in two parts: (1) a high-level part integrated in the
+  science gateway, which submits ``entire workflows'' to the
+  infrastructure; (2) a lower-level part, executed on the
+  infrastructure, which executes such ``entire workflows''. It seems
+  to us that this matches the definition of a nested workflow model,
+  where the parent engine is integrated in the science gateway, and
+  the child engine runs on the infrastructure. This is another
+  instantiation of the nested workflow abstract model, which is not
+  evaluated in the paper as mentioned in section 5.2: "\emph{The
+    abstract nested workflows and workflow conversion patterns were
+    instantiated with service invocation so that they can be analyzed
+    in the same framework as the other architectures. Other types of
+    instantiation, for instance nested workflows with task
+    encapsulation, could also be envisaged.}". It sounds like the
+  architecture described by the reviewer is a nested workflow pattern
+  instantiated with a tight integration (parent engine) and task
+  encapsulation (child engine).
+
+  We found this scenario interesting and we evaluated it in our
+  framework. The analysis is reported in the new section 6 of the
+  paper.}
 
@@
 
+\todo{Don't divide by the total number of interactions because then a
+  given interaction could have a different value depending on the
+  architecture. Refer to spreadsheet where weights can be adjusted.}
+
 \begin{review}
@@
 
+\todo{OK but these components are not specific to workflow
+  execution. So it's out of the definition. Motivation for this definition: these components are here anyway. If the SG fails, then the
+  workflows can't even be started. If the infrastructure is down, same. }
+
 \begin{review}
@@
 
+\todo{no, because c1 is not red.}
+
 \begin{review}
@@
 
+\todo{I agree it's boring but it's also important to have it clearly written.}
+
 \begin{review}
@@
 \end{review}
+
+\todo{what???}
 
@@
 
-\textbf{Reviewer \#3}
+\section{Reviewer \#3}
 
@@
 
+\todo{answer metric by metric, focusing on E1 and E4. For E4 it's clear. For E1, }
+
 \begin{review}
@@
 
+\answer{complexity is used as a measure of reliability}
+
 \begin{review}
@@
 
-\end{letter}
 \end{document}

### 2016-10-11 letter.tex
More answers to the reviewers.

\documentclass[a4]{letter}
\usepackage{graphicx}
\name{Tristan Glatard}
\signature{\includegraphics[width=4cm]{signature.png}}
\date{October 16th, 2016}

\begin{document}

\begin{letter}{}

\opening{Dear Editors,}

We would like to thank you for offering us the opportunity to revise
and re-submit our paper entitled ``Software architectures to integrate
workflow engines in science gateways'' to Future Generation Computer
Systems.

The remainder of this letter contains our answer to the 3 reviews that
we received following the evaluation process. A revised version of the
manuscript follows, where corrected parts appear in blue for a better
identification.

We thank you again for your time and consideration, and we remain available for any further information you may need.

\vspace{0.5cm}

Sincerely yours,

\vspace{1cm}

\closing{Tristan Glatard, PhD.\\On behalf of the authors.}

\end{letter}
\end{document}

### 2016-10-11 paper.tex
More answers to the reviewers.

@@
 \usepackage{pdfcomment}
-
 \definecolor{ao(english)}{rgb}{0.0, 0.5, 0.0}
@@
 
-\correction{The reminder of this section provides}
- background information and definitions of workflow engines,
-science gateways and infrastructures. In
-Section~\ref{sec:architectures}, we describe six architectures that
-cover the main patterns that we encountered in the various systems
-that we studied during the last few years. The architectures are
-described within a consistent framework that underlines the functional
-interactions between their main software components. In
-Section~\ref{sec:evaluation}, the different architectures are
-evaluated with metrics measuring integration complexity, robustness,
-extensibility, scalability and other specific features. These metrics
-are specifically designed to measure the ability of the architectures
-to address workflow-related issues commonly encountered in science
-gateways. We finally compare the architectures and highlight
-recommendations for science gateway architects and developers.
+\correction{The reminder of this section provides} background
+information and definitions of workflow engines, science gateways and
+infrastructures. In Section~\ref{sec:architectures}, we describe six
+architectures that cover the main patterns that we encountered in the
+various systems that we studied during the last few years. The
+architectures are described within a consistent framework that
+underlines the functional interactions between their main software
+components. \correction{In Section~\ref{sec:metrics}, we introduce
+  metrics that measure integration complexity, robustness,
+  extensibility, scalability and other specific features. These
+  metrics are specifically designed to measure the ability of the
+  architectures to address workflow-related issues commonly
+  encountered in science gateways. In Section~\ref{sec:evaluation}, we
+  evaluate the different architectures based on the proposed metrics.}
+We finally compare the architectures and highlight recommendations for
+science gateway architects and developers.
 
@@
 
-\section{Evaluation}
-
-\label{sec:evaluation}
+\section{\correction{Metrics}}
+
+\label{sec:metrics}
 
@@
   to generate custom versions of Table~\ref{table:evaluation}.}
-
-The criteria and metrics are
-explained hereafter.
 
@@
 
-\subsection{Evaluation metrics}
-
 Science gateway \emph{integration complexity} is measured through the number of software components and their interactions. It breaks
@@
 
+\section{\correction{Evaluation}}
+
+\label{sec:evaluation}
+
 The architectures described in Figure~\ref{fig:architectures} are
-evaluated along these metrics in the remainder of this Section.
+evaluated \correction{hereafter with the proposed metrics}.
 
@@
 components (\texttt{I$_1$}=2).  It involves 5 interactions:
-\texttt{a$_1$}, \texttt{a$_2$}, \texttt{b}, \texttt{c} and
+\texttt{a$_1$}, \texttt{a$_2$}, \texttt{b}, \correction{\texttt{c$_1$}} and
 \texttt{e} (\texttt{I$_2$}=5).
@@
 infrastructure requires updating interactions \texttt{a$_1$},
-\texttt{b} and \texttt{c} in the science gateway (\texttt{E$_4$}=3).
+\texttt{b} and \correction{\texttt{c$_1$}} in the science gateway (\texttt{E$_4$}=3).
 
@@
 
+\correction{\section{Real systems}}
+
+\label{sec:real systems}
+
+\correction{\section{Application to systems design}}
+\label{sec:application}
+
+\correction{To illustrate how our evaluation framework can be applied to the
+design of new systems, let's consider a hypothetical system where the
+workflow engine is split in two parts: (1) a high-level part
+integrated in the science gateway, which submits sub-workflows to the
+infrastructure; (2) a lower-level part, executed on the
+infrastructure, which executes the sub-workflows. This scenario was
+suggested by Reviewer \#2 of the paper after the evaluation framework
+had been designed.
+
+We model such a system using a nested workflow architecture where the
+parent engine is tightly integrated in the science gateway, and the
+child engine is integrated through task encapsulation.
+Figure~\ref{fig:nested-3} shows the graphical representation of the
+system. From this representation we find that the sub-tasks submitted
+by the child engine through interaction \texttt{e} should be handled
+by the parent engine rather than by the science gateway. Indeed,
+handling the tasks directly in the science gateway would lead to
+create a new specific interaction, named, e.g., \texttt{b$_1$}, while
+\texttt{b} could be reused. We also find that interaction \texttt{e}
+allows running the sub-tasks on the infrastructure supported by the
+parent engine with no further development effort required in the child
+engine.
+
+The system evaluation is presented on
+Table~\ref{table:evaluation-2}. The detailed calculations are
+available in sheet ``Metrics with reviewer 2's architecture'' of the public spreadsheet.
+
+We find that the studied system supports nested workflows with a much
+better performance than the nested workflow model instantiated with
+service invocation: the integration score is improved by a factor of
+2, the robustness is improved by a factor of 2.6, the extensibility is
+substantially improved, the scalability is not penalized, and the
+specific features are slightly improved.}
+
+\begin{figure}
+  \centering
+  \def\svgwidth{0.5\columnwidth}
+  \input{figures/nested-3.pdf_tex}
+  \caption{\correction{Nested workflow architecture instantiated with tight integration (parent engine) and task encapsulation (child engine).}}
+  \label{fig:nested-3}
+\end{figure}
+
+\begin{table}
+\footnotesize
+\centering
+\begin{tabular}{rc}
+                                     & \textbf{Nested 1}\\
+\cellcolor[HTML]{EEEEEE}\textbf{Integration complexity}&\cellcolor[HTML]{EEEEEE}\\
+  Total components -- \texttt{I$_1$} & 2\\
+Total interactions -- \texttt{I$_2$} & 6\\
+\textbf{Total} (total software pieces) & \textbf{8}\\
+\cellcolor[HTML]{EEEEEE}\textbf{Robustness}&\cellcolor[HTML]{EEEEEE}\\
+Specific components -- \texttt{R$_1$} & 0\\
+  Specific interactions -- \texttt{R$_2$} & 3\\
+  \textbf{Total} (execution software pieces)& \textbf{3}\\
+\cellcolor[HTML]{EEEEEE}\textbf{Extensibility}&\cellcolor[HTML]{EEEEEE}\\
+  New engine type -- \texttt{E$_1$}  & 3\\
+New engine version -- \texttt{E$_2$} & 1\\
+  New workflow -- \texttt{E$_3$}     & 2\\
+New infrastructure -- \texttt{E$_4$} & 4\\
+  \textbf{Total} (difficulty to extend) & \textbf{10}\\
+\cellcolor[HTML]{EEEEEE}\textbf{Scalability}& \cellcolor[HTML]{EEEEEE}\\
+Multiple engine instances -- \texttt{S$_1$}& 1\\
+Distributed engines -- \texttt{S$_2$}& 0\\
+Task scheduling -- \texttt{S$_3$}    & 1\\
+\textbf{Total} (scalability issues)  & \textbf{2}\\
+\cellcolor[HTML]{EEEEEE}\textbf{Specific features}&\cellcolor[HTML]{EEEEEE}\\
+  Meta-workflow  -- \texttt{O$_1$}    & 0\\
+  Fine-grained debugging -- \texttt{O$_2$}  & 0.5\\
+  \textbf{Total} (missing features) & \textbf{0.5}\\
+\end{tabular}
+\caption{\correction{Evaluation of the nested workflow architecture instantiated with tight integration (parent engine) and task encapsulation (child engine).}}
+\label{table:evaluation-2}
+\end{table}
+
 \section{Discussion}
@@
 
-Table~\ref{table:overall} provides an overall comparison between
+Table~\ref{table:overall} \todo{update with new architecture?} provides an overall comparison between
 architectures, based on the metrics in Table~\ref{table:evaluation}.
@@
 invocation (\correction{in particular when interactions can be re-purposed}), and other science gateways may have strongly tested task
-and data control features (interactions \texttt{b} and \texttt{c}),
+and data control features (interactions \texttt{b} and \correction{\texttt{c$_1$}}),
 which would favor task encapsulation. Similarly, adding a new type of engine

### 2016-10-11 figures/export.sh

@@ -1,6 +1,6 @@
 #!/bin/bash
 
-for name in tight service sub-task agent nested-1 nested-2 import-1 import-2 notations VIP CBRAIN-PSOM CBRAIN-FSL
+for name in tight service sub-task agent nested-1 nested-2 nested-3 import-1 import-2 notations VIP CBRAIN-PSOM CBRAIN-FSL
 do
   echo "Exporting ${name}.svg"
   inkscape -D -z --file=figures/${name}.svg --export-pdf=figures/${name}.pdf --export-latex

### 2016-10-12 answers.tex
Finished new section 6

@@
 {\end{quote}}%  
-\newcommand{\answer}[1]{\textbf{Answer:} #1}
+\newenvironment{answer}%          environment name
+{\textbf{Answer:}\begin{small}\begin{quote}}% begin code
+{\end{quote}\end{small}}%
+
 \newcommand{\todo}[1]{\color{red}#1\color{black}}
@@
 
-\answer{We thank the reviewer for the thorough review and useful comments.
+\begin{answer}
+
+  We thank the reviewer for the thorough review and useful comments.
 
@@
 
-  We added the following paragraph to the Introduction to reflect this discussion:}
+  We added the following paragraph to the Introduction to reflect this discussion:
 
@@
 
+\end{answer}
+
 \begin{review}
@@
 
-\answer{The actual workflow engine deployed in the infrastructure
+\begin{answer}
+  The actual workflow engine deployed in the infrastructure
   requires the gateway to directly control its own sub-tasks. We
-  revised the wording of Section 2.4 to clarify this point.}
+  revised the wording of Section 2.4 to clarify this point.
+\end{answer}
 
@@
 
-\answer{S1 denotes the ability of the system to start multiple engine
+\begin{answer}
+  S1 denotes the ability of the system to start multiple engine
   instances to execute different workflows, while S2 denotes the
@@
   this point: ``[S2] measures the ability to distribute the execution of
-  \emph{a single workflow} among different engine instances.''}
+  \emph{a single workflow} among different engine instances.''
+\end{answer}
 
@@
 
-\answer{We introduced E3 precisely to account for the additional
+\begin{answer}
+  We introduced E3 precisely to account for the additional
   complexity introduced by interaction (g) when adding new
@@
 format.}
-}
 
@@
 workflow language used by the new engine.``
+\end{answer}
 
@@
 
-\answer{As discussed previously, we share the reviewer's concerns
+\begin{answer}
+
+  As discussed previously, we share the reviewer's concerns
   about the feasibility of workflow conversion (see comments on
@@
 believe is a reasonable but limited choice.
-}
+
+\end{answer}
 
@@
 
-\answer{The reviewer is right that re-purposing software components
-  and interactions can lead to revisit the evaluation. We updated the following sentence in section 5.2. to better acknowledge this point:}
+\begin{answer}
+
+The reviewer is right that re-purposing software components
+  and interactions can lead to revisit the evaluation. We updated the following sentence in section 5.2. to better acknowledge this point:
 
@@
 Therefore, \texttt{E$_1$}=3}.
+\end{answer}
 
@@
 
-\answer{While we agree with the reviewer that re-usability should
+\begin{answer}
+  While we agree with the reviewer that re-usability should
   definitely be taken into account in E$_1$, we do not see how it
@@
   in multiple engine instances. In our opinion, this feature is
-  exclusive to the nested workflow model. }
+  exclusive to the nested workflow model.
+\end{answer}
 
@@
 
+\begin{answer}
 We hope that the update made in the extensibility scores and the
 explanations given for scalability convinced the reviewer.
+\end{answer}
 
@@
 
-\answer{We updated the definition of R$_1$ and R$_2$ to include
+\begin{answer}
+We updated the definition of R$_1$ and R$_2$ to include
   positive examples as well as negative ones. In addition, the
   specific components included in these metrics are represented in red
-  on Figure 2.}
+  on Figure 2.
+\end{answer}
 
@@
 
+\begin{answer}
 \texttt{c}, \texttt{c$_1$} and \texttt{c$_2$} are three different
@@
 
-Specific components (\texttt{R$_1$}): number of specific components involved in workflow executions. \revised{For instance, the
+\begin{itemize}
+\item Specific components (\texttt{R$_1$}): number of specific components involved in workflow executions. \revised{For instance, the
   workflow service is a specific component because its main objective
@@
   transfer, etc.}
-
-Specific interactions (\texttt{R$_2$}): number of specific
+\item Specific interactions (\texttt{R$_2$}): number of specific
   interactions involved in workflow executions. \revised{For
@@
     user data regardless of a particular workflow execution}.
+\end{itemize}
+\end{answer}
 
@@
 
-\answer{No it doesn't. We hope that the new examples in the definition of
+\begin{answer}
+No it doesn't. We hope that the new examples in the definition of
 R$_1$ and R$_2$ and the removal of any reference to interaction
-\texttt{c} clarifies this point.}
+\texttt{c} clarifies this point.
+\end{answer}
 
@@
 
-\answer{Yes, as for any other metric, high values indicate worse
+\begin{answer}
+Yes, as for any other metric, high values indicate worse
 performance. This is explained at the beginning of section 3:
 ``Criteria break down into specific metrics where lower value
-indicates better performance.''}
+indicates better performance.''
+\end{answer}
 
@@
 
-\todo{think of this}
+\begin{answer}
+  We call ``white-box'' workflow a workflow
+  where detailed information about individual tasks is available,
+  for instance task statuses, execution logs and resource
+  consumption. On the contrary, a ``black-box'' workflow only
+  displays global information such as the workflow status and the
+  logs of the workflow engine. Fine-grained information about
+  workflow tasks helps workflow users (scientists)
+  identify errors, it allows gateway administrators to troubleshoot
+  executions, and it facilitates debugging by workflow
+  developers.
+
+  Multiple software layers affect the accuracy of the information
+  reported about the tasks' internals because each layer potentially
+  shields or modifies information. Task ids, for instance, may not
+  be properly linked to workflow activities if delivered through a
+  software layer that hasn't been instrumented for that. \todo{refer
+    to olabarriaga2014}.
+
+  The paragraph on fine-grained debugging has been edited accordingly, at the end of Section 3.
+\end{answer}
 
@@
 To conclude, the general feeling is that the definition of the metrics and the interaction patterns are not rigorous enough to provide an interpretation of the final evaluation which is entirely useful for the reader.
+\end{review}
+
+\begin{answer}
+We hope that the explanations provided in this letter and the updated text in the paper demonstrate that the metrics and definitions are rigorous.
+\end{answer}
+
+\begin{review}
  Probably, some more information on how "Future Generation" systems should be designed, starting from the general overview provided, with more technical understanding of the system complexities, could represent a more valuable contribution to the main aim of this particular publication.
@@
 
-\todo{think of this}
+\begin{answer}
+  We added a section in the paper (current section 6) to show
+  how our evaluation framework could be used in the design of new
+  systems.
+\end{answer}
 
@@
 
-\answer{This was done as suggested.}
+\begin{answer}
+This was done as suggested.
+\end{answer}
 
@@
 
-\answer{The task submission framework is part of the infrastructure
+\begin{answer}
+  The task submission framework is part of the infrastructure
   and its nature doesn't influence the architecture used to integrate
@@
   framework. The analysis is reported in the new section 6 of the
-  paper.}
+  paper.
+\end{answer}
 
@@
 
-\answer{complexity is used as a measure of reliability}
+\todo{complexity is used as a measure of reliability}
 

### 2016-10-12 paper.tex
Finished new section 6

@@
 }\\
-\subfloat[Nested workflows. Left: abstract model. Right: instantiation with service invocation.]{
+\subfloat[Nested workflows. Left: abstract model. Right: instantiation with service invocation (``Service+Service'').]{
 \def\svgwidth{0.9\columnwidth}
@@
 (Figure~\subref*{archi:nested}-Right) as this is the most used
-architecture. In the instantiation, we assume that the parent and
-child workflow engines are distinct pieces of software that require
-different workflow services invoked by distinct \texttt{d}
+architecture. \correction{We call this instantiation
+  ``Service+Service'' since both engines are integrated through
+  service invocation.} We assume that the parent
+and child workflow engines are distinct pieces of software that
+require different workflow services invoked by distinct \texttt{d}
 interactions. If this is not the case, then workflow services can be
@@
 itself. An example of such interaction is the use of hierarchical
-workflows in Pegasus~\cite{Deelman201517}.
-Workflow engines communicate with infrastructures using
-\texttt{b} and \texttt{c$_2$}. Science gateway transfers user data to
-infrastructures using \texttt{c$_1$} interactions.
+workflows in Pegasus~\cite{Deelman201517}.  Workflow engines
+communicate with infrastructures using \texttt{b} and
+\texttt{c$_2$}. Science gateway transfers user data to infrastructures
+using \texttt{c$_1$} interactions.
 
@@
                                      & \textbf{Conversion} \\
+&&&&& \textbf{\correction{Service+Service}}&\\
 \cellcolor[HTML]{EEEEEE}\textbf{Integration complexity}& \multicolumn{6}{l}{\cellcolor[HTML]{EEEEEE}}\\
@@
                                      & \cellcolor[HTML]{999999}1
-                                     & \cellcolor[HTML]{99FF99}0
+                                     & \cellcolor[HTML]{99FF99}\correction{1}
                                      & \cellcolor[HTML]{999999}1
@@
   fine-grained debugging information about workflow tasks (white-box
-  workflow).  Fine-grained information about workflow tasks is
-  required to properly troubleshoot workflow executions. Its
-  availability and accuracy usually depends on the number of software
-  layers between the science gateway and the workflow
-  engine~\cite{olabarriaga2014}. \texttt{O$_2$}=0 means that
-  the information is directly accessible in the science gateway,
-  \texttt{O$_2$}=1 means that it is obtained through one or more software
-  interactions. 
+  workflow). \correction{We call ``white-box'' workflow a workflow
+    where detailed information about individual tasks is available,
+    for instance task statuses, execution logs and resource
+    consumption. On the contrary, a ``black-box'' workflow only
+    displays global information such as the workflow status and the
+    logs of the workflow engine.} Fine-grained information about
+  workflow tasks is required to properly troubleshoot workflow
+  executions. \correction{It helps workflow users (scientists)
+    identify errors, it allows gateway administrators to troubleshoot
+    executions, and it facilitates debugging by workflow
+    developers}. \correction{The availability and accuracy of
+    fine-grained debugging information} usually depends on the number
+  of software layers between the science gateway and the workflow
+  engine~\cite{olabarriaga2014}. \texttt{O$_2$}=0 means that the
+  information is directly accessible in the science gateway,
+  \texttt{O$_2$}=1 means that it is obtained through one or more
+  software interactions.
 \end{itemize}
@@
 
-\paragraph{Specific features} Task encapsulation explicitly enables
-meta-workflows (\texttt{O$_1$}=0). Indeed, workflow engines may invoke
-other engines using the exact same interaction used to submit regular
-sub-tasks (interaction \texttt{e}), without any additional
-development. Fine-grained debugging
+\paragraph{Specific features} Task encapsulation \correction{does not enable meta-workflows by default (\texttt{O$_1$=1})}. Fine-grained debugging
 information is obtained through interaction \texttt{b}
@@
 
-\correction{To illustrate how our evaluation framework can be applied to the
-design of new systems, let's consider a hypothetical system where the
-workflow engine is split in two parts: (1) a high-level part
-integrated in the science gateway, which submits sub-workflows to the
-infrastructure; (2) a lower-level part, executed on the
-infrastructure, which executes the sub-workflows. This scenario was
-suggested by Reviewer \#2 of the paper after the evaluation framework
-had been designed.
-
-We model such a system using a nested workflow architecture where the
-parent engine is tightly integrated in the science gateway, and the
-child engine is integrated through task encapsulation.
-Figure~\ref{fig:nested-3} shows the graphical representation of the
-system. From this representation we find that the sub-tasks submitted
-by the child engine through interaction \texttt{e} should be handled
-by the parent engine rather than by the science gateway. Indeed,
-handling the tasks directly in the science gateway would lead to
-create a new specific interaction, named, e.g., \texttt{b$_1$}, while
-\texttt{b} could be reused. We also find that interaction \texttt{e}
-allows running the sub-tasks on the infrastructure supported by the
-parent engine with no further development effort required in the child
+\correction{To illustrate how our evaluation framework can be applied
+  to the design of new systems, let's consider a hypothetical system
+  where the workflow engine is split in two parts: (1) a high-level
+  part integrated in the science gateway, which submits sub-workflows
+  to the infrastructure; (2) a lower-level part, executed on the
+  infrastructure, which executes the sub-workflows. This scenario came
+  up in the discussion with Reviewer \#2 of this paper, after the
+  evaluation framework had already been designed.
+
+  We model such a system using a nested workflow architecture with a
+  ``Tight+Task'' instantiation, i.e., the parent engine is tightly
+  integrated in the science gateway and the child engine is
+  integrated through task encapsulation.  Figure~\ref{fig:nested-3}
+  shows the graphical representation of the system. From this
+  representation we find that the sub-tasks submitted by the child
+  engine through interaction \texttt{e} should be handled by the
+  parent engine rather than by the science gateway. Indeed, handling
+  the tasks directly in the science gateway would lead to create a new
+  specific interaction, named, e.g., \texttt{b$_1$}, while \texttt{b}
+  could be reused. We also find that interaction \texttt{e} allows
+  running the sub-tasks on the infrastructure supported by the parent
+  engine with no further development effort required in the child
   engine.
@@
 Table~\ref{table:evaluation-2}. The detailed calculations are
-available in sheet ``Metrics with reviewer 2's architecture'' of the public spreadsheet.
+available in sheet ``Metrics with reviewer 2's architecture'' of the public spreadsheet at \url{https://frama.link/LrPRsCQp} \todo{revise that sheet to merge it with the main one}.
 
@@
 better performance than the nested workflow model instantiated with
-service invocation: the integration score is improved by a factor of
-2, the robustness is improved by a factor of 2.6, the extensibility is
-substantially improved, the scalability is not penalized, and the
-specific features are slightly improved.}
+service invocation (evaluated on Table~\ref{table:evaluation}): the
+integration score is improved by a factor of 2, the robustness is
+improved by a factor of 2.6, the extensibility is substantially
+improved, the scalability and specific features are not penalized.
+
+Based on this system, we propose a new system design that instantiates
+the nested workflow model in a ``Task+Task'' way, i.e., with task
+encapsulation for both the parent and the child engine. The
+architecture of this system is represented in
+Figure~\ref{fig:nested-4}. The diagram is identical to the one in
+Figure~\ref{archi:sub-task}, except that interaction \texttt{e} is now
+used to submit sub-workflows instead of just sub-tasks. This is
+possible since, in the task encapsulation architecture, workflows are
+executed as tasks.
+
+The new architecture performs better similarly to task encapsulation,
+except that it now enables meta-workflows and distributed engines
+which were not available previously. It clearly outperforms the nested
+``Service+Service'' architecture.
+
+Based on this analysis, we plan to extend the PSOM-CBRAIN integration
+to support this architecture. Conceptually, the implementation is
+simple, as it only requires existing workflows to be declarable as
+tasks in the workflow engine. In practice, technical issues may delay
+the process. We also remain well aware of the scheduling issues raised
+by the task encapsulation architecture (S$_3$=1), which will be
+amplified in the ``Task+Task'' instantiation of nested workflows: for
+instance, properly setting resource requirements (e.g. walltime) is
+increasingly difficult when nested workflow engines are deployed on
+the infrastructure.  }
 
@@
   \centering
+\subfloat[Nested, Tight+Task]{
   \def\svgwidth{0.5\columnwidth}
   \input{figures/nested-3.pdf_tex}
-  \caption{\correction{Nested workflow architecture instantiated with tight integration (parent engine) and task encapsulation (child engine).}}
   \label{fig:nested-3}
+}
+\subfloat[Nested, Task+Task]{ 
+  \def\svgwidth{0.5\columnwidth}
+  \input{figures/sub-task.pdf_tex}
+  \label{fig:nested-4}
+}
+\caption{\correction{Nested workflow architecture instantiated with tight integration and task encapsulation (top) and task encapsulation only (bottom).}}
+\label{fig:new-systems}
 \end{figure}
@@
 \centering
-\begin{tabular}{rc}
-                                     & \textbf{Nested 1}\\
-\cellcolor[HTML]{EEEEEE}\textbf{Integration complexity}&\cellcolor[HTML]{EEEEEE}\\
-  Total components -- \texttt{I$_1$} & 2\\
-Total interactions -- \texttt{I$_2$} & 6\\
-\textbf{Total} (total software pieces) & \textbf{8}\\
-\cellcolor[HTML]{EEEEEE}\textbf{Robustness}&\cellcolor[HTML]{EEEEEE}\\
-Specific components -- \texttt{R$_1$} & 0\\
-  Specific interactions -- \texttt{R$_2$} & 3\\
-  \textbf{Total} (execution software pieces)& \textbf{3}\\
-\cellcolor[HTML]{EEEEEE}\textbf{Extensibility}&\cellcolor[HTML]{EEEEEE}\\
-  New engine type -- \texttt{E$_1$}  & 3\\
-New engine version -- \texttt{E$_2$} & 1\\
-  New workflow -- \texttt{E$_3$}     & 2\\
-New infrastructure -- \texttt{E$_4$} & 4\\
-  \textbf{Total} (difficulty to extend) & \textbf{10}\\
-\cellcolor[HTML]{EEEEEE}\textbf{Scalability}& \cellcolor[HTML]{EEEEEE}\\
-Multiple engine instances -- \texttt{S$_1$}& 1\\
-Distributed engines -- \texttt{S$_2$}& 0\\
-Task scheduling -- \texttt{S$_3$}    & 1\\
-\textbf{Total} (scalability issues)  & \textbf{2}\\
-\cellcolor[HTML]{EEEEEE}\textbf{Specific features}&\cellcolor[HTML]{EEEEEE}\\
-  Meta-workflow  -- \texttt{O$_1$}    & 0\\
-  Fine-grained debugging -- \texttt{O$_2$}  & 0.5\\
-  \textbf{Total} (missing features) & \textbf{0.5}\\
+\begin{tabular}{rcc}
+                                     & \textbf{Nested} & \textbf{Nested}\\
+                                     & \textbf{Tight+Task} & \textbf{Task+Task}\\
+\cellcolor[HTML]{EEEEEE}\textbf{Integration complexity}&\cellcolor[HTML]{EEEEEE}&\cellcolor[HTML]{EEEEEE}\\
+  Total components -- \texttt{I$_1$} & 2 & 2 \\
+Total interactions -- \texttt{I$_2$} & 6 & 5 \\
+\textbf{Total}& \textbf{8} & \textbf{7}\\
+\cellcolor[HTML]{EEEEEE}\textbf{Robustness}&\cellcolor[HTML]{EEEEEE}&\cellcolor[HTML]{EEEEEE}\\
+Specific components -- \texttt{R$_1$} & 0 & 0\\
+  Specific interactions -- \texttt{R$_2$} & 3 & 1\\
+  \textbf{Total}& \textbf{3} & \textbf{1}\\
+\cellcolor[HTML]{EEEEEE}\textbf{Extensibility}&\cellcolor[HTML]{EEEEEE}&\cellcolor[HTML]{EEEEEE}\\
+  New engine type -- \texttt{E$_1$}  & 3 & 2\\
+New engine version -- \texttt{E$_2$} & 1 & 1\\
+  New workflow -- \texttt{E$_3$}     & 2 & 2\\
+New infrastructure -- \texttt{E$_4$} & 4 & 3\\
+  \textbf{Total}& \textbf{10} & \textbf{8}\\
+\cellcolor[HTML]{EEEEEE}\textbf{Scalability}& \cellcolor[HTML]{EEEEEE}&\cellcolor[HTML]{EEEEEE}\\
+Multiple engine instances -- \texttt{S$_1$}& 1 & 0\\
+Distributed engines -- \texttt{S$_2$}& 0 & 0\\
+Task scheduling -- \texttt{S$_3$}    & 1 & 1\\
+\textbf{Total}& \textbf{2} & \textbf{1}\\
+\cellcolor[HTML]{EEEEEE}\textbf{Specific features}&\cellcolor[HTML]{EEEEEE}&\cellcolor[HTML]{EEEEEE}\\
+  Meta-workflow  -- \texttt{O$_1$}    & 0 & 0\\
+  Fine-grained debugging -- \texttt{O$_2$}  & 1 & 1\\
+  \textbf{Total}& \textbf{1} & \textbf{1}\\
 \end{tabular}
@@
                                     & \textbf{Conversion} \\
+                                    &&&&& \textbf{\correction{Service+Service}}&\\
   Integration (global score)    &

### 2016-10-12 answers.tex
Still more answers to the reviewers

@@
 analysis. In the paper, we assumed that all weights equal 1, which we
-believe is a reasonable but limited choice.
+believe is a reasonable choice although inherently limited.
 
@@
 
+\begin{answer}
+As suggested by the reviewer, the presentation of these Figures has been gathered in a separate section, the new section 3.
+\end{answer}
+
 \begin{review}
@@
 
-\todo{Don't divide by the total number of interactions because then a
-  given interaction could have a different value depending on the
-  architecture. Refer to spreadsheet where weights can be adjusted.}
+\begin{answer}
+  We agree with the reviewer that different weights could be given to
+  interactions and software components depending on the ``gravity'' of
+  the associated failures, the ``difficulty'' to repair them and even
+  the difficulty to implement such interactions and components in
+  practice (for instance, interaction \texttt{g} is clearly very
+  difficult to implement).
+
+  However, we believe that assigning such weights is specific to the
+  case of a particular system, i.e. a science gateway and workflow
+  engine(s). Very different ponderations could be obtained depending
+  on the technical context, for instance implementation language, type
+  of targeted infrastructure (dedicated cluster vs. grid vs. cloud)
+  and characteristics of the workflow engine(s) and language(s).
+
+  Instead, we shared a spreadsheet at
+  \url{https://frama.link/LrPRsCQp} that allows editing the following
+  weights (see also answer to Reviewer \#1):
+  \begin{itemize}
+  \item Weight given to software components and interactions in
+  ``integration'', ``extensibility'' and ``robustness'' criteria.
+  \item Weight given to metrics in a criterion (e.g., E$_1$ vs. E$_2$ in
+  ``extensibility'').
+  \item Weight given to criteria in overall evaluation of Table 3.
+\end{itemize}
+The paper references this spreadsheet wherever equal weights are
+assumed. The readers and reviewers are invited to rely on this
+spreadsheet for any issue related to the importance given to a
+specific component, interaction, metric or criterion in the
+analysis. In the paper, we assumed that all weights equal 1, which we
+believe is a reasonable choice although inherently limited.
+
+We find the particular ponderation method suggested by the reviewer
+(normalization by the number of components or interaction in the
+architecture) a bit unfortunate because it would give different
+ponderations to the same component or interaction depending on the
+architecture in which it is used, which we think would obfuscate the
+comparison.
+\end{answer}
 
@@
 
-\todo{OK but these components are not specific to workflow
-  execution. So it's out of the definition. Motivation for this definition: these components are here anyway. If the SG fails, then the
-  workflows can't even be started. If the infrastructure is down, same. }
+\begin{answer}
+  We believe that the robustness of a system for a particular function
+  should be evaluated w.r.t the set of components and interactions
+  that are \emph{specific} to this function. Indeed, the failure of
+  non-specific components is likely to impact many other functions
+  that are required by the particular function of interest. To take an
+  extreme example, a failure of the internet connection of the gateway
+  user would certainly prevent him/her to launch a new workflow, but
+  it does not really make sense to integrate the internet connection
+  in the robustness analysis because it is not specific to workflow
+  execution. Similarly, the failure of the science gateway would
+  prevent users to login in the first place, which would prevent
+  workflow execution in any case. Likewise, the failure of the
+  infrastructure would make any workflow execution worthless in any
+  architecture because infrastructure is used for other critical
+  functions such as task execution and data storage.
+
+  For these reasons, we disagree with the suggestion made by the
+  reviewer. In any case, science gateway and infrastructure are
+  present in all the architectures, which would only add 2 to all the
+  values of R1.
+\end{answer}
 
@@
 
-\todo{no, because c1 is not red.}
+\begin{answer}
+  In Figure 2.I, interaction \texttt{c$_1$} is not specific to
+  workflow execution since it is used to transfer any data to the
+  infrastructure. This is why we don't count it in R2, for the same
+  reason as explained previously.
+\end{answer}
 
@@
 
-\todo{I agree it's boring but it's also important to have it clearly written.}
+\begin{answer}
+  While we agree that the Sub-Sections in Section 3 look a bit
+  repetitive, we believe that the metrics evaluation reported in Table 2 needs to be
+  properly justified. 
+\end{answer}
 
@@
 
-\todo{what???}
+\begin{answer}
+  Service invocation and Task encapsulation are fundamentally different for the following reasons:
+  \begin{itemize}
+  \item In task encapsulation, the workflow engine runs on the infrastructure while in service invocation, it is hosted by a dedicated service.
+  \item In task encapsulation, the workflow engine is executed as a task of the infrastructure while in service invocation, it is invoked through a service call.
+That means, in particular, that in task encapsulation different workflow engines are executed through the same protocol (a task), while different types of service calls might be involved in service invocation (e.g.
+REST or SOAP Web-Service).
+  \item In task encapsulation, the workflow engine does not transfer data to the infrastructure while in service invocation it does (interactioni \texttt{c$_2$}.
+  \item In task encapuslation, the science gateway submits tasks to the infrastructure, while in service invocation the workflow engine does.
+  \end{itemize}
+  Therefore, while we acknowledge that task encapsulation and service
+  invocation share characteristics, we believe that their differences
+  can be clearly inferred from section 2 of the paper. 
+\end{answer}
 
@@
 \end{review}
+
+\todo{ponderation again}
 

### 2016-10-12 answers.tex
Even more answers to the reviewers

@@
 \usepackage{hyperref}
+\usepackage{xspace}
 \newenvironment{review}%          environment name
@@
 
-\newcommand{\revised}[1]{\color{blue} #1\color{black}}
+\newcommand{\revised}[1]{\color{blue} #1\color{black}\xspace}
 
@@
 
+\begin{answer}
+We thank the reviewer for the thorough review and useful comments.
+
 \todo{just remove CSGF, Airvata and pgrade from the table, keeping the
@@
   still need to be found.}
+\end{answer}
 
@@
 
-\todo{ponderation again}
+\begin{answer}
+  We assume that the reviewer is referring to the I$_1$ metric which
+  is, indeed, lower for the tight integration than for any other
+  architecture (it is equal for task encapsulation). We agree with the
+  reviewer that the difficulty to update a component depends on the
+  nature of this component. However, we believe that a quantification
+  of this difficulty is very much dependent on the particular system
+  under study. For instance, in some cases, tightly integrating a
+  workflow engine may be straightforward (e.g. when the science
+  gateway has the appropriates hooks) while developing a service may
+  require a lot of work if it has to be done from scratch.
+
+  We believe that this discussion comes down to the assignments of
+  different weights to the interactions and components in the
+  architectures. As discussed before, these weights can now be
+  adjusted using the spreadsheet at \url{https://frama.link/LrPRsCQp}
+  to better describe a particular system.
+\end{answer}
 
@@
 \begin{review}
-This paper provides a comparison of 6 different software architectures commonly used to integrate workflow engines into science gateways.
-For this comparison, the authors define seven different types of interactions to outline the differences in approach.
-The combination of components and interactions for the 6 overview models in Figure 2 that form the underlying basis for the comparisons.
+  This paper provides a comparison of 6 different software
+  architectures commonly used to integrate workflow engines into
+  science gateways. For this comparison, the authors define seven
+  different types of interactions to outline the differences in
+  approach. The combination of components and interactions for the 6
+  overview models in Figure 2 that form the underlying basis for the
+  comparisons.
 
-The authors then provide a set of Metrics: complexity, robustness, extensibility, scalability \& specific features to compare and contrast the different software architectures.
+  The authors then provide a set of Metrics: complexity, robustness,
+  extensibility, scalability \& specific features to compare and
+  contrast the different software architectures.
 
-Complexity, for example, is measured by counting the number of components and interactions that were defined and annotated in Figure 2.
-Extensibility measures the difficulty to replace or add elements to the architecture e.g.
-a workflow engine, an engine version update, add a new workflow or add new infrastructure.
+  Complexity, for example, is measured by counting the number of
+  components and interactions that were defined and annotated in
+  Figure 2.  Extensibility measures the difficulty to replace or add
+  elements to the architecture e.g. a workflow engine, an engine
+  version update, add a new workflow or add new infrastructure.
 
-I liked this paper but I also struggled a lot with it too because ultimately, whilst this metric structure provides some rudimentary structure to quantify the various software architectures, a many software architecture perspectives are missing, and the results are not normalized against the functionality a system provides.
+  I liked this paper but I also struggled a lot with it too because
+  ultimately, whilst this metric structure provides some rudimentary
+  structure to quantify the various software architectures, a many
+  software architecture perspectives are missing, and the results are
+  not normalized against the functionality a system provides.
 
-For example, when you normally think of software architectures, a good design is often one that is layered and componentized.
-There are multiple benefits of such an an approach over monolithic systems but using the metrics defined in this paper, good software practices to expose functionality is not really factored in.
-So, the result is that the more layers to components in the system, the worse it scores.
-So, these aspects of good software engineering or providing enhanced functionality seem penalized in this paper, as it stands.
+  For example, when you normally think of software architectures, a
+  good design is often one that is layered and componentized. There
+  are multiple benefits of such an an approach over monolithic systems
+  but using the metrics defined in this paper, good software practices
+  to expose functionality is not really factored in. So, the result is
+  that the more layers to components in the system, the worse it
+  scores. So, these aspects of good software engineering or providing
+  enhanced functionality seem penalized in this paper, as it stands.
 \end{review}
+
+\begin{answer}
+We thank the reviewer for the thorough review and useful comments.
+
+Regarding componentization: our architectures are described in terms
+of software components and interactions among these components (see
+Figure 1). So we believe that our approach fits in the context of
+component-based software engineering. The problem that we address is
+to determine what components should be defined and how they should
+interact to optimize the metrics. For instance, the comparison between
+the tightly integrated and service invocation architectures evaluates
+whether the workflow engine should be a specific component or if it
+should be integrated in the science gateway. Besides, the reviewer's
+comment touches upon the granularity at which components are
+defined. We extended the first paragraph of Section 2 to address this issue:
+\revised{Software
+  components are defined at a granularity such that two components may
+  potentially be operated by different teams on different hardware
+  systems. For instance, the science gateway and infrastructure are
+  usually distinct components. In real systems, software components
+  may exist at a finer granularity, to implement functions such as
+  authentication, data persistency or logging. Such fine-grained
+  components are not covered by our analysis since we focus on the
+  interactions between the science gateway, the infrastructure and the
+  workflow engine.}
+
+The discussion on software layers is more difficult. Roughly, our
+architectures could be described as multilayered software
+architectures. For instance, the tight integration has two layers
+(science gateway and infrastructure), the service invocation has
+three, the task encapsulation has two, etc. However, strictly
+speaking, these are not layers because it would mean that layer n-1
+can only communicate with layer n, which is not the case. For
+instance, in service invocation, layer 0 (science gateway)
+communicates with layer 2 (infrastructure) to transfer data. And in
+task encapsulation, layer 1 (infrastructure) communicates with layer 0
+(science gateway). Thus, we don't think that referring to multilayered
+architectures is accurate here.
+
+We don't think that our architectures lead to conclude systematically
+that ``the more layers to components in the system, the worse it
+scores'', as suggested by the reviewer. For instance, service
+invocation scores better than tight integration in E1 and S1 due to
+the additional ``layer'' used to host the workflow engine. Likewise,
+nested workflows score better than tight integration for S2. However,
+our metrics do imply that more layers leads to more integration
+complexity and reduced robustness, which we think is very commonly
+experienced in distributed systems, where  communication between
+``layers'' involve wide-area network communications between machines
+that may be operated by different people. We added a note in the definition of our robustness metrics to reflect this discussion:
+\begin{quote}
+ Robustness is measured here as a consequence of global
+complexity, since complex architectures tend to be more prone to
+failure, \revised{in particular in distributed systems where
+  interactions between components involve wide-area network
+  communications between machines that may be operated by different
+  people}.
+\end{quote}
+\end{answer}
 
@@
 
-\todo{answer metric by metric, focusing on E1 and E4. For E4 it's clear. For E1, }
+\begin{answer}
+  The issue raised by the reviewer needs to be discussed metric by
+  metric rather than on the global extensibility score. The pool and
+  nested architectures differ in metrics E$_1$ (adding a new engine
+  type) and E$_4$ (adding a new infrastructure). For E$_4$ (adding a
+  new infrastructure), it seems clear to us that the nested
+  architecture is more difficult to extend because it has two workflow
+  engines while the pool has only one. Therefore, interactions
+  \texttt{b} and \texttt{c$_2$} have to be counted twice, which
+  explains the difference in scores for E$_4$.
+
+  The situation highlighted by the reviewer is related to E$_1$ (new
+  engine type). If a new engine is added to the pool model, the
+  following interactions and components have to be updated:
+ \begin{enumerate}
+ \item The engine has to be wrapped in the SHIWA Agent, which includes
+   the reading/writing of ORE bundles so that the agent can get/pass
+   them to the pool through interaction \texttt{f}. In practice, the
+   agent component has to be modified to convert ORE bundles to the
+   input format supported by the engine, and to assemble ORE bundles
+   from the output format produced by the engine. As suggested by the
+   reviewer, another implementation possibility would be to modify the
+   engine to read/write ORE bundles so that the integration in the
+   agent is straightforward.
+ \item The engine has to be able to transfer data to the
+   infrastructure used in the science gateway, i.e. interaction
+   \texttt{c$_2$} has to be implemented in the engine.
+ \item The engine has to be able to submit tasks to the infrastructure
+   used in the science gateway, i.e. interaction \texttt{b} has to be
+   implemented in the engine.
+ \end{enumerate}
+This is why E$_1$=3 for the pool model.
+
+If a new child engine is added to the nested model then the following interactions an and components have to be updated:
+\begin{enumerate}
+\item A workflow service has to be created for the workflow engine.
+\item The parent engine has to be interfaced with the new workflow
+  service, i.e., the workflow service should be able to digest ORE
+  bundles, or the parent engine should convert ORE bundles to the
+  format supported by the workflow service. ORE bundles will be
+  exchanged through interaction \texttt{d}.
+ \item The child engine has to be able to transfer data to the
+   infrastructure used in the science gateway, i.e. interaction
+   \texttt{c$_2$} has to be implemented.
+ \item The child engine has to be able to submit tasks to the infrastructure
+   used in the science gateway, i.e. interaction \texttt{b} has to be
+   implemented,
+\end{enumerate}
+This is why E$_4$=4 for the nested model.
+
+The situation described by the reviewer focuses on item 1 of the pool
+extension and item 1 and 2 of the nested workflow extension
+(infrastructure is not a discriminant factor here). First, the
+reviewer suggest that the availability of ORE bundles in the SHIWA
+framework standardizes the communication with workflow engines so that
+it becomes possible to have a generic workflow service that can
+consume and produce ORE bundles regardless of the particular engine(s)
+it wraps. So indeed, item 1 of the nested case disappears. This is a
+typical case where, as we wrote in section 7.2, \emph{The particular
+technical or historical context of a science gateway project may also
+influence the \revised{evaluation} of an architecture.} In this
+case, the possible availability of a generic workflow service based on
+ORE bundles is contextual information that can hardly be integrated in
+the generic analysis. We believe that our framework helps evaluate
+the impact of technologies such as ORE bundles on architectures, as
+discussed here.
+
+Second, the
+reviewer suggests that the integration of the workflow engine in the
+agent (pool model) is easier than in the service because the engine
+has to produce and consume ORE bundles in the former case while it
+only has to consume them in the latter case. This is true again, but
+it comes down to the discussion on the weights of the different
+software components. In this particular case, it would make sense to
+assign a higher weight to the agent than to the service, which could
+be done through the public spreadsheet.  
+\end{answer}
 
 \begin{review}
-Adding functionality e.g.
-in this case to add the ability to pass bundles to one system and another is a positive (and complex) thing to accomplish.
-There should be a way of counting this and having this score higher than Tight, Service or task, or normalizing this functionality into the metrics to provide a more balanced view.
-The "Scalability" touches on this but "Functionality" could be another aspect here to consider.
+  Adding functionality e.g. in this case to add the ability to pass
+  bundles to one system and another is a positive (and complex) thing
+  to accomplish. There should be a way of counting this and having
+  this score higher than Tight, Service or task, or normalizing this
+  functionality into the metrics to provide a more balanced view. The
+  "Scalability" touches on this but "Functionality" could be another
+  aspect here to consider.
 \end{review}
+
+\begin{answer}
+  We agree with the reviewer that functionality is important. This was
+  exactly the goal of the ``Specific features'' evaluation
+  criterion. We renamed this criterion ``Functionality'' as we find it a much better name indeed. We also renamed O$_1$ to F$_1$ and O$_2$ to F$_2$.
+\end{answer}
 

### 2016-10-12 paper.tex
Still more answers to the reviewers
Even more answers to the reviewers

@@
   assessment of integration complexity, robustness, extensibility,
-  scalability and support for meta-workflows and fine-grained
-  debugging. Tight integration and task encapsulation are the easiest
+  scalability and \correction{functionality}. Tight integration and task encapsulation are the easiest
   to integrate and the most robust. Extensibility is equivalent in
@@
   metrics that measure integration complexity, robustness,
-  extensibility, scalability and other specific features. These
+  extensibility, scalability and \correction{functionality}. These
   metrics are specifically designed to measure the ability of the
@@
 
-% Software properties: integration effort, extensibility, robustness,
-% scalability, specific features
-
 Integration between workflow engines and science gateways varies
@@
 presented in Figure~\ref{fig:architectures} using the graphical
-notations defined in
-Figure~\ref{fig:notations}. Table~\ref{table:system-classification}
+notations defined in Figure~\ref{fig:notations}. \correction{Software
+  components are defined at a granularity such that two components may
+  potentially be operated by different teams on different hardware
+  systems. For instance, the science gateway and infrastructure are
+  usually distinct components. In real systems, software components
+  may exist at a finer granularity, to implement functions such as
+  authentication, data persistency or logging. Such fine-grained
+  components are not covered by our analysis since we focus on the
+  interactions between the science gateway, the infrastructure and the
+  workflow engine.}  Table~\ref{table:system-classification}
 summarizes the classification of a few systems by architecture.
@@
 
-\begin{figure*}
-\centering
-\def\svgwidth{1.5\columnwidth}
-\input{figures/VIP.pdf_tex}
-\caption{Architecture of the Virtual Imaging Platform (service
-  invocation).  Step \texttt{3} corresponds to interaction \texttt{d}
-  in Figure~\ref{fig:architectures}(II), step \texttt{2} corresponds to
-  interaction \texttt{c$_1$}, step \texttt{4} maps to interaction
-  \texttt{b}, and steps 7 and 9 map to \texttt{c$_2$} (in VIP, workflow data transfers are
-  performed by the tasks and embedded in their descriptions).}
-\label{fig:vip-architecture}
-\end{figure*}
-
 See Figure~\subref*{archi:service}. The workflow engine is available
@@
 command-line tools (e.g., pegasus-status, pegasus-plan,
-etc.). Figure~\ref{fig:vip-architecture} shows the architecture of the
-Virtual Imaging Platform, where the MOTEUR workflow
-engine~\cite{GLAT-08i} is invoked through a service.
+etc.).
 
@@
 the infrastructure, through interaction \texttt{c$_1$}.
+\subsection{Pool model}
+\label{sec:pool}
+See Figure~\subref*{archi:agent}. Workflows are submitted by the
+science gateway to a central pool through interaction
+\texttt{d}. Agents connect to the pool asynchronously to retrieve and
+execute workflows through interaction \texttt{f}. Agents may be
+started according to various policies, for instance to ensure load
+balancing. Workflow engine controls tasks and data on the
+infrastructure through interactions \texttt{b} and \texttt{c$_2$},
+science gateway transfers user data through interaction
+\texttt{c$_1$}, and administrator installs workflows through
+interaction \texttt{a$_1$} and \texttt{a$_2$}.
+
+\subsection{Nested workflows}
+
+See Figure~\subref*{archi:nested}. In nested workflows
+(Figure~\subref*{archi:nested}-Left), a workflow engine is integrated
+as a \emph{child} process of a \emph{parent} workflow engine. Parent
+and child engines might use different languages and might run on
+different infrastructures. A parent workflow is also a
+meta-workflow. The science gateway communicates with the parent engine
+through abstract interaction \texttt{*$_1$}. The science gateway also
+communicates with the infrastructure to transfer user data through
+abstract interactions \texttt{*$_2$} and \texttt{*$_3$}. Both workflow
+engines communicate with the infrastructure through abstract
+interactions \texttt{*$_4$} and \texttt{*$_5$}. The parent engine
+communicates with the child engine through abstract interaction
+\texttt{*$_6$}. Administrator installs workflows through interactions
+\texttt{a$_1$} and \texttt{a$_2$}.
+
+Nested workflows are abstract architectural patterns that can be
+instantiated in the various architectures described previously. We
+focus on instantiation with the service invocation model
+(Figure~\subref*{archi:nested}-Right) as this is the most used
+architecture. \correction{We call this instantiation
+  ``Service+Service'' since both engines are integrated through
+  service invocation.} We assume that the parent
+and child workflow engines are distinct pieces of software that
+require different workflow services invoked by distinct \texttt{d}
+interactions. If this is not the case, then workflow services can be
+collapsed into a single one with a \texttt{d} interaction with
+itself. An example of such interaction is the use of hierarchical
+workflows in Pegasus~\cite{Deelman201517}.  Workflow engines
+communicate with infrastructures using \texttt{b} and
+\texttt{c$_2$}. Science gateway transfers user data to infrastructures
+using \texttt{c$_1$} interactions.
+
+Nested workflows have long been available in workflow engines, for
+instance in the Taverna workbench~\cite{oinn2004taverna}. They are
+also used implicitly in several platforms where workflow engines are
+wrapped in workflow tasks as any other command-line tool. Nested
+workflows were notably used by the SHIWA Science Gateway to implement
+so-called Coarse-Grained workflow
+interoperability~\cite{terstyanszky2014enabling}, i.e. to integrate
+various workflow engines in a consistent
+platform.
+
+\subsection{Workflow conversion}
+
+See Figure~\subref*{archi:import}. This is an abstract model
+instantiated with the service invocation architecture for
+consistency. \correction{Workflow conversion is presented here as a pattern to
+integrate workflow \emph{engines} in a science gateway, through
+workflow format conversion from a native format to the science gateway
+format.} Workflow conversion is usually an offline process that is not
+involved in workflow execution. A few systems have implemented
+workflow conversion. In the SHIWA Simulation Platform, it is
+implemented through the IWIR language, which provides a common
+language for portability across grid workflow
+systems~\cite{plankensteiner-prodan-etal:2013} and allows conversion
+among $n$ workflow languages using $2n$ interactions instead of $n^2$.
+Tavaxy~\cite{Abouelhoda2012} enables the import, merging and execution
+of Taverna~\cite{oinn2004taverna} and Galaxy~\cite{goecks2010galaxy}
+workflows; when some workflow parts cannot be imported, workflows are
+run by Tavaxy as nested workflows using their native engine. The work
+in~\cite{delaGarza2016} describes workflow conversion from
+KNIME~\cite{Berthold2008} to WS-PGRADE/gUSE and from
+Galaxy~\cite{goecks2010galaxy} to WS-PGRADE/gUSE.
+
+%Furthermore, both systems provide web-portals allowing users to share and publish their workflows: These are the myExperiment portal for Taverna [23, 24] and the Public Pages for Galaxy [13].
+
+\correction{\section{Architectural patterns in real systems}}
+
+\label{sec:real systems}
+
+%\todo{Introduce this section and take the text out of the captions}
+
+Figure~\ref{fig:vip-architecture} shows the architecture of the
+Virtual Imaging Platform, where the MOTEUR workflow
+engine~\cite{GLAT-08i} is invoked through a service.
+
+\begin{figure*}
+\centering
+\def\svgwidth{1.5\columnwidth}
+\input{figures/VIP.pdf_tex}
+\caption{Architecture of the Virtual Imaging Platform (service
+  invocation).  Step \texttt{3} corresponds to interaction \texttt{d}
+  in Figure~\ref{fig:architectures}(II), step \texttt{2} corresponds to
+  interaction \texttt{c$_1$}, step \texttt{4} maps to interaction
+  \texttt{b}, and steps 7 and 9 map to \texttt{c$_2$} (in VIP, workflow data transfers are
+  performed by the tasks and embedded in their descriptions).}
+\label{fig:vip-architecture}
+\end{figure*}
 
@@
 
-\subsection{Pool model}
-\label{sec:pool}
-See Figure~\subref*{archi:agent}. Workflows are submitted by the
-science gateway to a central pool through interaction
-\texttt{d}. Agents connect to the pool asynchronously to retrieve and
-execute workflows through interaction \texttt{f}. Agents may be
-started according to various policies, for instance to ensure load
-balancing. Workflow engine controls tasks and data on the
-infrastructure through interactions \texttt{b} and \texttt{c$_2$},
-science gateway transfers user data through interaction
-\texttt{c$_1$}, and administrator installs workflows through
-interaction \texttt{a$_1$} and \texttt{a$_2$}.
-
 The pool model was implemented in the SHIWA pool~\cite{ROGE-13}
@@
 
-\subsection{Nested workflows}
-
-See Figure~\subref*{archi:nested}. In nested workflows
-(Figure~\subref*{archi:nested}-Left), a workflow engine is integrated
-as a \emph{child} process of a \emph{parent} workflow engine. Parent
-and child engines might use different languages and might run on
-different infrastructures. A parent workflow is also a
-meta-workflow. The science gateway communicates with the parent engine
-through abstract interaction \texttt{*$_1$}. The science gateway also
-communicates with the infrastructure to transfer user data through
-abstract interactions \texttt{*$_2$} and \texttt{*$_3$}. Both workflow
-engines communicate with the infrastructure through abstract
-interactions \texttt{*$_4$} and \texttt{*$_5$}. The parent engine
-communicates with the child engine through abstract interaction
-\texttt{*$_6$}. Administrator installs workflows through interactions
-\texttt{a$_1$} and \texttt{a$_2$}.
-
-Nested workflows are abstract architectural patterns that can be
-instantiated in the various architectures described previously. We
-focus on instantiation with the service invocation model
-(Figure~\subref*{archi:nested}-Right) as this is the most used
-architecture. \correction{We call this instantiation
-  ``Service+Service'' since both engines are integrated through
-  service invocation.} We assume that the parent
-and child workflow engines are distinct pieces of software that
-require different workflow services invoked by distinct \texttt{d}
-interactions. If this is not the case, then workflow services can be
-collapsed into a single one with a \texttt{d} interaction with
-itself. An example of such interaction is the use of hierarchical
-workflows in Pegasus~\cite{Deelman201517}.  Workflow engines
-communicate with infrastructures using \texttt{b} and
-\texttt{c$_2$}. Science gateway transfers user data to infrastructures
-using \texttt{c$_1$} interactions.
-
-Nested workflows have long been available in workflow engines, for
-instance in the Taverna workbench~\cite{oinn2004taverna}. They are
-also used implicitly in several platforms where workflow engines are
-wrapped in workflow tasks as any other command-line tool. Nested
-workflows were notably used by the SHIWA Science Gateway to implement
-so-called Coarse-Grained workflow
-interoperability~\cite{terstyanszky2014enabling}, i.e. to integrate
-various workflow engines in a consistent
-platform. Figure~\ref{fig:shiwa-architecture} shows the architecture
+Figure~\ref{fig:shiwa-architecture} shows the architecture
 used in the SHIWA Simulation Platform for nested workflow execution
@@
 
-\subsection{Workflow conversion}
-
-See Figure~\subref*{archi:import}. This is an abstract model
-instantiated with the service invocation architecture for
-consistency. \correction{Workflow conversion is presented here as a pattern to
-integrate workflow \emph{engines} in a science gateway, through
-workflow format conversion from a native format to the science gateway
-format.} Workflow conversion is usually an offline process that is not
-involved in workflow execution. A few systems have implemented
-workflow conversion. In the SHIWA Simulation Platform, it is
-implemented through the IWIR language, which provides a common
-language for portability across grid workflow
-systems~\cite{plankensteiner-prodan-etal:2013} and allows conversion
-among $n$ workflow languages using $2n$ interactions instead of $n^2$.
-Tavaxy~\cite{Abouelhoda2012} enables the import, merging and execution
-of Taverna~\cite{oinn2004taverna} and Galaxy~\cite{goecks2010galaxy}
-workflows; when some workflow parts cannot be imported, workflows are
-run by Tavaxy as nested workflows using their native engine. The work
-in~\cite{delaGarza2016} describes workflow conversion from
-KNIME~\cite{Berthold2008} to WS-PGRADE/gUSE and from
-Galaxy~\cite{goecks2010galaxy} to WS-PGRADE/gUSE.
-
-%Furthermore, both systems provide web-portals allowing users to share and publish their workflows: These are the myExperiment portal for Taverna [23, 24] and the Public Pages for Galaxy [13].
-
 \section{\correction{Metrics}}
@@
 integration complexity, robustness, extensibility, scalability and
-other specific features. These are based on science gateway properties
+\correction{functionality}. These are based on science gateway properties
 introduced in Section~\ref{sec:intro_sg}. Criteria break down into
@@
 meaningfully. \correction{All the metrics in a criterion contribute
-  equally to the global score. If this is not the case, the
-  reader is invited to adjust the weights in the public spreadsheet available
-  at \url{https://frama.link/LrPRsCQp} (column B of sheet ``Metrics'')
-  to generate custom versions of Table~\ref{table:evaluation}.}
+  equally to the global score. Should the metrics be weighted
+  differently in a particular system, the reader is invited to produce
+  his/her own version of the Table using the public spreadsheet
+  available at \url{https://frama.link/LrPRsCQp} (see column B of
+  sheet ``Metrics'').}
 
@@
                                      & \cellcolor[HTML]{99CC99}\textbf{2}\\
-\cellcolor[HTML]{EEEEEE}\textbf{Specific features}& \multicolumn{6}{l}{\cellcolor[HTML]{EEEEEE}}\\
-  Meta-workflow  -- \texttt{O$_1$}    & \cellcolor[HTML]{999999}1
+\cellcolor[HTML]{EEEEEE}\textbf{\correction{Functionality}}& \multicolumn{6}{l}{\cellcolor[HTML]{EEEEEE}}\\
+  Meta-workflow  -- \texttt{\correction{F}$_1$}    & \cellcolor[HTML]{999999}1
                                      & \cellcolor[HTML]{999999}1
@@
                                      & \cellcolor[HTML]{99FF99}0\\
-  Fine-grained debugging -- \texttt{O$_2$}   & \cellcolor[HTML]{99FF99}0
+  Fine-grained debugging -- \texttt{\correction{F}$_2$}   & \cellcolor[HTML]{99FF99}0
                                      & \cellcolor[HTML]{999999}1
@@
                                      & \cellcolor[HTML]{999999}\textbf{2}
-                                     & \cellcolor[HTML]{99FF99}\textbf{1}
+                                     & \cellcolor[HTML]{99FF99}\textbf{\correction{2}}
                                      & \cellcolor[HTML]{999999}\textbf{2}
@@
 One may wonder whether infrastructure should be counted in
-\texttt{I$_1$}, since it is usually an external entity that is not developed nor maintained by the groups who
-integrate workflow engines in science gateway. We support its inclusion here because integrating an
-infrastructure into the system usually requires some technical
-effort (e.g., account creation, software installation, APIs, etc.), therefore increasing complexity.
-\correction{We assume that software components contribute equally to $I_1$ and that interactions also contribute equally to $I_2$.
-In a real system, this may not be the case due to a range of project-specific practical reasons that cannot be covered in this paper.
-The weights given to components and interactions can be dynamically changed in the public spreadsheet (sheet ``Components and interactions'') to generate new versions of Table~\ref{table:evaluation} as required.} The two metrics \correction{I$_1$ and I$_2$} are summed to obtain a
-global score that measures the
-total number of software pieces affected by integration.
-
-\emph{Robustness of workflow execution} measures the likelihood that the
-workflow execution fails due to errors in the components or in the
-interactions of the software architecture. Errors coming
-from the infrastructure (e.g., unavailable data or terminated tasks)
-or workflows (e.g., wrong user input or application errors) are not
+\texttt{I$_1$}, since it is usually an external entity that is not
+developed nor maintained by the groups who integrate workflow engines
+in science gateway. We support its inclusion here because integrating
+an infrastructure into the system usually requires some technical
+effort (e.g., account creation, software installation, APIs, etc.),
+therefore increasing complexity. \correction{We assume that software
+  components contribute equally to $I_1$ and that interactions also
+  contribute equally to $I_2$. In a real system, this may not be the
+  case due to a range of project-specific practical reasons that
+  cannot be covered in this paper. The weights given to components and
+  interactions can be dynamically changed in the public spreadsheet
+  (sheet ``Components and interactions'') to generate new versions of
+  Table~\ref{table:evaluation} as required.} The two metrics
+\correction{I$_1$ and I$_2$} are summed to obtain a global score that
+measures the total number of software pieces affected by integration.
+
+\emph{Robustness of workflow execution} measures the likelihood that
+the workflow execution fails due to errors in the components or in the
+interactions of the software architecture. Errors coming from the
+infrastructure (e.g., unavailable data or terminated tasks) or
+workflows (e.g., wrong user input or application errors) are not
 covered, since they do not stem directly from the software
-architecture. 
-Robustness is measured here as a consequence of global
+architecture.  Robustness is measured here as a consequence of global
 complexity, since complex architectures tend to be more prone to
-failure. More precisely, robustness is determined by the number of
-software components and interactions represented in
-red in the architecture diagrams in
-Figure~\ref{fig:architectures}, which are specific to workflow
-execution. Two metrics are used:
+failure, \correction{in particular in distributed systems where
+  interactions between components involve wide-area network
+  communications between machines that may be operated by different
+  people}. More precisely, robustness is determined by the number of
+software components and interactions represented in red in the
+architecture diagrams in Figure~\ref{fig:architectures}, which are
+specific to workflow execution. Two metrics are used:
 \begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
@@
 
-\emph{Specific features} include:
+\correction{\emph{Functionality} includes}:
 \begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
-\item Meta-workflow (\texttt{O$_1$}): measures the ability to describe
+\item Meta-workflow (\texttt{\correction{F}$_1$}): measures the ability to describe
   meta-workflows from existing workflows. Meta-workflows offer an
   additional level of flexibility to build workflows from reusable
-  components. \texttt{O$_1$}=0 means that meta-workflows are
+  components. \texttt{\correction{F}$_1$}=0 means that meta-workflows are
   intrinsically enabled by the architecture (i.e. they can be
   implemented using the components and interactions already in place),
-  and \texttt{O$_1$}=1 means that they may be implemented with some
+  and \texttt{\correction{F}$_1$}=1 means that they may be implemented with some
   development effort.
 
-\item Fine-grained debugging (\texttt{O$_2$}): availability of
+\item Fine-grained debugging (\texttt{\correction{F}$_2$}): availability of
   fine-grained debugging information about workflow tasks (white-box
@@
   of software layers between the science gateway and the workflow
-  engine~\cite{olabarriaga2014}. \texttt{O$_2$}=0 means that the
+  engine~\cite{olabarriaga2014}. \texttt{\correction{F}$_2$}=0 means that the
   information is directly accessible in the science gateway,
-  \texttt{O$_2$}=1 means that it is obtained through one or more
+  \texttt{\correction{F}$_2$}=1 means that it is obtained through one or more
   software interactions.
 \end{itemize}
-\texttt{O$_1$} and \texttt{O$_2$} are summed to obtain a total number of missing specific
+\texttt{\correction{F}$_1$} and \texttt{\correction{F}$_2$} are summed to obtain a total number of missing 
 features in the architecture.
@@
 
-\paragraph{Specific features} Supporting meta-workflows requires
+\paragraph{\correction{Functionality}} Supporting meta-workflows requires
 specific developments in the workflow engine or science gateway to
-allow workflows to execute other workflows (\texttt{O$_1$}=1). For
+allow workflows to execute other workflows (\texttt{\correction{F}$_1$}=1). For
 instance, specific Java objects may be implemented in the science
@@
 retrieve fine-grained debugging information from the workflow engine
-directly (\texttt{O$_2$}=0).
+directly (\texttt{\correction{F}$_2$}=0).
 
@@
 
-\paragraph{Specific features} Meta-workflows require specific
+\paragraph{\correction{Functionality}} Meta-workflows require specific
 developments in the workflow engine or science gateway to allow
-workflows to execute other workflows.  (\texttt{O$_1$}=1).
+workflows to execute other workflows.  (\texttt{\correction{F}$_1$}=1).
  The
@@
 fine-grained debugging information from the workflow service
-(\texttt{O$_2$}=1).
+(\texttt{\correction{F}$_2$}=1).
 
@@
 
-\paragraph{Specific features} Task encapsulation \correction{does not enable meta-workflows by default (\texttt{O$_1$=1})}. Fine-grained debugging
-information is obtained through interaction \texttt{b}
-(\texttt{O$_2$}=1).
+\paragraph{\correction{Functionality}} Task encapsulation
+\correction{does not enable meta-workflows by default
+  (\texttt{\correction{F}$_1$=1})}. Fine-grained debugging information is obtained
+through interaction \texttt{b} (\texttt{\correction{F}$_2$}=1).
 
@@
 
-\paragraph{Specific features} Meta-workflows require specific
+\paragraph{\correction{Functionality}} Meta-workflows require specific
 developments in the workflow engine to enable workflow submission
-(\texttt{O$_1$=1}). Debugging information is accessed through
-interactions \texttt{d} and \texttt{f} (\texttt{O$_2$=1}).
+(\texttt{\correction{F}$_1$=1}). Debugging information is accessed through
+interactions \texttt{d} and \texttt{f} (\texttt{\correction{F}$_2$=1}).
 
@@
 
-\paragraph{Specific features}
+\paragraph{\correction{Functionality}}
 Meta-workflows can be implemented, which is one of the main interest
-of this architecture (\texttt{O$_1$=0}). Note, however, that the
+of this architecture (\texttt{\correction{F}$_1$=0}). Note, however, that the
 complexity of the architecture increases with the number of engine
 types involved in meta-workflows: for instance, a meta-workflow with
-child workflows executed by 2 different types of engines would require an
-additional workflow service and the corresponding interactions.
-Debugging information is accessed through interaction \texttt{d} (invoked twice) (\texttt{O$_2$=1}).
+child workflows executed by 2 different types of engines would require
+an additional workflow service and the corresponding interactions.
+Debugging information is accessed through interaction \texttt{d}
+(invoked twice) (\texttt{\correction{F}$_2$=1}).
 
@@
 
-\paragraph{Specific features} Meta-workflows are available after
+\paragraph{\correction{Functionality}} Meta-workflows are available after
 conversion, by connecting workflows in the language used in the
-science gateway (\texttt{O$_1$=0}).  Debugging information is accessible
-as in the service invocation architecture (\texttt{O$_2$=1}).
-
-\correction{\section{Real systems}}
-
-\label{sec:real systems}
+science gateway (\texttt{\correction{F}$_1$=0}).  Debugging information is accessible
+as in the service invocation architecture (\texttt{\correction{F}$_2$=1}).
 
@@
 service invocation (evaluated on Table~\ref{table:evaluation}): the
-integration score is improved by a factor of 2, the robustness is
-improved by a factor of 2.6, the extensibility is substantially
-improved, the scalability and specific features are not penalized.
+integration score is improved by a factor of 2, robustness is
+improved by a factor of 2.6, extensibility is substantially
+improved, scalability and functionality are not penalized.
 
@@
 \textbf{Total}& \textbf{2} & \textbf{1}\\
-\cellcolor[HTML]{EEEEEE}\textbf{Specific features}&\cellcolor[HTML]{EEEEEE}&\cellcolor[HTML]{EEEEEE}\\
-  Meta-workflow  -- \texttt{O$_1$}    & 0 & 0\\
-  Fine-grained debugging -- \texttt{O$_2$}  & 1 & 1\\
+\cellcolor[HTML]{EEEEEE}\textbf{\correction{Functionality}}&\cellcolor[HTML]{EEEEEE}&\cellcolor[HTML]{EEEEEE}\\
+  Meta-workflow  -- \texttt{\correction{F}$_1$}    & 0 & 0\\
+  Fine-grained debugging -- \texttt{\correction{F}$_2$}  & 1 & 1\\
   \textbf{Total}& \textbf{1} & \textbf{1}\\
@@
 
-Differences in other specific features should not be neglected. Task encapsulation, nested
-workflows (with service invocation) and workflow conversion (with service
-invocation) are the only architectures that intrinsically support meta-workflows. Besides,
-tight integration is the only architecture that allows
-fine-grained debugging, which may be
+Differences in \correction{functionality} should not be neglected. Task
+encapsulation, nested workflows (with service invocation) and workflow
+conversion (with service invocation) are the only architectures that
+intrinsically support meta-workflows. Besides, tight integration is
+the only architecture that allows fine-grained debugging, which may be
 critical for efficient user support.
@@
                                      & \cellcolor[HTML]{99CC99}{0.50}\\
-Specific features (global score) & \cellcolor[HTML]{99FF99}{0.00}
+\correction{Functionality} (global score) & \cellcolor[HTML]{99FF99}{0.00}
                                      & \cellcolor[HTML]{999999}{1.00}
@@
 conversion) is clearly more complex than interaction \texttt{d}
-(service invocation). However, without entering into the details of a particular
+(service invocation). However, without entering into the details of a particular \todo{refer to spreadsheet again}
 system, quantification of the robustness or the amount of development
@@
 The particular technical or historical context of a science gateway
-project may also influence the choice of an architecture to
-integrate workflow engines. For instance, many workflow engines are
-already available as web services, which tends to favor service
-invocation (\correction{in particular when interactions can be re-purposed}), and other science gateways may have strongly tested task
-and data control features (interactions \texttt{b} and \correction{\texttt{c$_1$}}),
-which would favor task encapsulation. Similarly, adding a new type of engine
-may facilitate integration of a new infrastructure when interactions
-\texttt{b} and \texttt{c$_2$} are already available. The migration
-cost between architectures has been ignored as well.
+project may also influence the \correction{evaluation} of an
+architecture to integrate workflow engines. For instance, many
+workflow engines are already available as web services, which tends to
+favor service invocation (\correction{in particular when interactions
+  can be re-purposed}), and other science gateways may have strongly
+tested task and data control features (interactions \texttt{b} and
+\correction{\texttt{c$_1$}}), which would favor task
+encapsulation. Similarly, adding a new type of engine may facilitate
+integration of a new infrastructure when interactions \texttt{b} and
+\texttt{c$_2$} are already available. The migration cost between
+architectures has been ignored as well.
 
@@
 
-This work has been made possible with the support of the Irving Ludmer
+\correction{We thank the anonymous reviewers for the thorough reviews and useful comments that greatly contributed to improve the quality of this paper.} This work has been made possible with the support of the Irving Ludmer
 Family Foundation and the Ludmer Centre for Neuroinformatics and

### 2016-10-12 answers.tex
Even more answers to the reviewers

@@
 \begin{review}
-It is not always obvious that Scalability could be badly affected in a Tight Integration.
-The example of a liferay portlet is actually potentially demonstrating the opposite.
-Portlets can be independent web application that can be even remotely deployed (via WSRP).
-The integration of a workflow engine via ad-hoc portlet applications could instead guarantee quite of a level of decoupling from the rest of the gateway.
-Here, the limitations highlighted seems to be a bit superficial.
+  It is not always obvious that Scalability could be badly affected in
+  a Tight Integration. The example of a liferay portlet is actually
+  potentially demonstrating the opposite. Portlets can be independent
+  web application that can be even remotely deployed (via WSRP). The
+  integration of a workflow engine via ad-hoc portlet applications
+  could instead guarantee quite of a level of decoupling from the rest
+  of the gateway. Here, the limitations highlighted seems to be a bit
+  superficial.
 \end{review}
 
-\todo{think of this}
+\begin{answer}
+  We agree with the reviewer that using WSRP (Web Services for Remote
+  Portlets) changes the architecture. When portlets are deployed on
+  different hosts, we think that the architecture becomes service
+  invocation as the portlets then behave as individual software
+  components according to our definitions. We added the following comment as a footnote in the presentation of tight integration in section 2.2:
+\revised{Note that the
+    particular case of portlets can be a bit ambiguous: when portlets
+    are deployed remotely, using a protocol such as Web Services for
+    Remote Portlets (WSRP), they become independent software
+    components according to our definition because they could be
+    operated by different teams on different hardware systems. In this
+    case, the system should be classified in the service invocation
+    architecture since the portlets actually become Web
+    Services. The same comment holds for remote object
+    invocation.}
+\end{answer}
 
 \begin{review}
-The "Specific Feature", referring to the Fine-grained debugging does not specify what is relevant to be monitored and who is going to gain benefit from the monitoring.
-The scientists, the gateway developers or the administrators?
-What is a white box, the workflow engine or the workflow task?
-Why multiple software layers affect the accuracy of the tasks' internals?
+  The "Specific Feature", referring to the Fine-grained debugging does
+  not specify what is relevant to be monitored and who is going to
+  gain benefit from the monitoring. The scientists, the gateway
+  developers or the administrators? What is a white box, the workflow
+  engine or the workflow task? Why multiple software layers affect the
+  accuracy of the tasks' internals?
 \end{review}
@@
 \begin{review}
-To conclude, the general feeling is that the definition of the metrics and the interaction patterns are not rigorous enough to provide an interpretation of the final evaluation which is entirely useful for the reader.
+  To conclude, the general feeling is that the definition of the
+  metrics and the interaction patterns are not rigorous enough to
+  provide an interpretation of the final evaluation which is entirely
+  useful for the reader.
 \end{review}
@@
 \begin{answer}
-We hope that the explanations provided in this letter and the updated text in the paper demonstrate that the metrics and definitions are rigorous.
+  We hope that the explanations provided in this letter and the
+  updated text in the paper demonstrate that the metrics and
+  definitions are rigorous.
 \end{answer}
@@
 \begin{review}
- Probably, some more information on how "Future Generation" systems should be designed, starting from the general overview provided, with more technical understanding of the system complexities, could represent a more valuable contribution to the main aim of this particular publication.
+  Probably, some more information on how "Future Generation" systems
+  should be designed, starting from the general overview provided,
+  with more technical understanding of the system complexities, could
+  represent a more valuable contribution to the main aim of this
+  particular publication.
 \end{review}
@@
 \begin{answer}
-  We added a section in the paper (current section 6) to show
-  how our evaluation framework could be used in the design of new
-  systems.
+  We added a section in the paper (current section 6) to show how our
+  evaluation framework could be used in the design of new systems.
 \end{answer}
@@
 \begin{review}
-To improve the reading, Dedicating separate sections to the introduction of the metrics and the evaluation of each pattern would be beneficial. For instance Section 3 Metrics, Section 4 Evaluation.
+  To improve the reading, Dedicating separate sections to the
+  introduction of the metrics and the evaluation of each pattern would
+  be beneficial. For instance Section 3 Metrics, Section 4 Evaluation.
 \end{review}
@@
 
-\todo{just remove CSGF, Airvata and pgrade from the table, keeping the
-  exponents. But we don't have examples of gateways using Airvata,
-  still need to be found.}
+We totally agree with the reviewer that Apache Airavata and
+WS-PGrade/gUSE are gateway frameworks. This is how they are presented
+in the text. Instead of providing an incomplete list of gateways that
+use these frameworks in Table 1, we now refer explicitly to ``gateways
+based on Apache Airvata'', ``gateways based on WS-PGRADE/gUSE'' and
+``gateways based on the Catania Science Gateway
+Framework''. Consistently, the table doesn't mention the individual
+gateways based on these frameworks any more.
 \end{answer}
@@
 \begin{answer}
-As suggested by the reviewer, the presentation of these Figures has been gathered in a separate section, the new section 3.
+  As suggested by the reviewer, the presentation of these Figures has
+  been gathered in a separate section which is now the new section 3.
 \end{answer}
@@
   exactly the goal of the ``Specific features'' evaluation
-  criterion. We renamed this criterion ``Functionality'' as we find it a much better name indeed. We also renamed O$_1$ to F$_1$ and O$_2$ to F$_2$.
+  criterion. We renamed this criterion ``Functionality'' as we find it
+  a much better name indeed. We also renamed O$_1$ to F$_1$ and O$_2$
+  to F$_2$.
+
+  However, we don't think that ``the ability to pass bundles'' should
+  be listed as a metric in the functionality criterion, since this
+  feature remains quite specific to the particular context mentioned
+  by the reviewer. 
 \end{answer}
@@
 \begin{review}
-Another example is for complexity.
-Again, the less components and interactions the system has the less complex it is.
-But in practice, it is not so simple.
-For example, the tight integration is where the workflow system is integration directly into the science gateway and into the infrastructure.
-Whereas this architecture has the least number of components and interactions of them all it does not mean it is less complex in practice.
-Integrating a workflow into science gateway and directly on top of the infrastructure depends very much on the tools they sit upon and the complexity of that integration.
-Such tight integrated systems therefore could be massive monolithic pieces of code that are extremely difficult and complex to develop and maintain.
-Whereas a task-based approach or service based approach through breaking code out to different services and tasks may be far simpler to develop and maintain.
+  Another example is for complexity.  Again, the less components and
+  interactions the system has the less complex it is. But in practice,
+  it is not so simple. For example, the tight integration is where the
+  workflow system is integration directly into the science gateway and
+  into the infrastructure. Whereas this architecture has the least
+  number of components and interactions of them all it does not mean
+  it is less complex in practice. Integrating a workflow into science
+  gateway and directly on top of the infrastructure depends very much
+  on the tools they sit upon and the complexity of that
+  integration. Such tight integrated systems therefore could be
+  massive monolithic pieces of code that are extremely difficult and
+  complex to develop and maintain.  Whereas a task-based approach or
+  service based approach through breaking code out to different
+  services and tasks may be far simpler to develop and maintain.
 \end{review}
 
-\todo{complexity is used as a measure of reliability}
+\begin{answer}
+  We totally agree with the reviewer. However, we think that this
+  issue cannot be addressed at the pure architectural level as it
+  requires accurate technical information about systems. In some
+  systems, the science gateway may become a massive piece of
+  unmaintainable code as described by the reviewer, and in some other
+  cases, it may have a neat internal architecture that allows for easy
+  integration of workflow engines. Once again, we believe that the
+  answer to this issue is about assigning weights to the different
+  components and interactions, which is now possible through the
+  public spreadsheet. For instance, if the science gateway code is
+  known to be ``untouchable'' due to a cumbersome, then a high weight
+  should be put on the science gateway component. 
+\end{answer}
 
 \begin{review}
-I therefore think that the metrics defined only provide a very high level structure for an evaluation of the different software architectures and they should discuss other factors and scenarios that can affect development.
-Some of these limitations are discussed in section 4.2 but I think an extra section discuss these before the definition of the metrics would be very helpful to set the scene correctly for the evaluation.
-Otherwise, the reader is left asking a number of questions during the evaluation section.
-I also think that a few more metrics e.g.
-"functionality" could be added to provide a more balanced view of the different approaches.
-Another useful extension of the paper could be to normalize the results against the functionality each approach provides (perhaps in a separate table) - this may help to provide a more balanced table that factors in functionality.
+  I therefore think that the metrics defined only provide a very high
+  level structure for an evaluation of the different software
+  architectures and they should discuss other factors and scenarios
+  that can affect development. 
 \end{review}
 
+\begin{answer}
+  The metrics provide an abstract framework for evaluating
+  architectures independently from any particular system. The goal is
+  to abstract architectural discussions from the particular technical
+  context of a real system. Some degree of system-specificity can now
+  be introduced by assigning weights to interactions and components
+  through our public spreadsheet, which we think provides a way to apply our
+  analysis to study real systems more precisely.
+\end{answer}
+
+\begin{review}
+Some of these limitations are discussed in section 4.2 but I think an
+extra section discuss these before the definition of the metrics would
+be very helpful to set the scene correctly for the
+evaluation.  Otherwise, the reader is left asking a number of questions
+during the evaluation section. 
+\end{review}
+
+\begin{answer}
+We added the following paragraphs to the introduction to section 3:
+
+\revised{Note that the metrics defined hereafter aim at providing a framework
+to evaluate architectures independently from any particular system,
+regardless of any technical constraint such as the use of a specific
+programming language or the internal architecture of the science
+gateway. The goal is to abstract architectural discussions from the
+particular technical context of a real system. Some degree of
+system-specificity can be introduced by assigning weights to
+interactions and components through a public spreadsheet available at
+\url{https://frama.link/LrPRsCQp}. In particular, sheet "Components
+and interactions" contains the weight given to these entities in the
+metrics and can be used to produce new versions of
+Table 2 that fit the particular technical context
+of a real system. For instance, a high weight could be assigned to the
+science gateway if its internal architecture is such that it can
+hardly be extended in practice.}
+
+\revised{In addition, all the metrics in a criterion contribute
+  equally to the global score used for the criterion. Column B of
+  sheet ``Metrics'' of the public spreadsheet may be used to assign
+  different weights to each metric to adjust the specific context of a
+  study.}
+\end{answer}
+
+\begin{review}
+I also think that a few more metrics
+e.g. "functionality" could be added to provide a more balanced view of
+the different approaches.
+
+ Another useful extension of the paper could
+be to normalize the results against the functionality each approach
+provides (perhaps in a separate table) - this may help to provide a
+more balanced table that factors in functionality.
+\end{review}
+
+\begin{answer}
+  As explained previously, the last evaluation criterion, now named
+  ``Functionality'' aims at this purpose. We hope it clarifies that
+  functionality is actually part of the analysis. 
+\end{answer}
+
 \end{document}

### 2016-10-12 paper.tex
Even more answers to the reviewers

@@
   Tight & \pbox{1.5\columnwidth}{
-          Catania Science Gateway Framework~\cite{ardizzone2012decide},
+          \correction{Gateways based on the} Catania Science Gateway Framework~\cite{ardizzone2012decide},
           Distributed application runtime environment (DARE~\cite{maddineni2012distributed}),
-          DECIDE~\cite{ardizzone2012decide}$^{c,n}$, LONI Pipeline Environment$^n$~\cite{dinov2009efficient}
+          LONI Pipeline Environment~\cite{dinov2009efficient}
           }
@@
   Service & \pbox{1.5\columnwidth}{
-            Apache Airvata~\cite{marru2011apache}, Neuroscience Gateway~\cite{shahand2015data}$^{w,n}$, HubZero with Pegasus~\cite{CPE:CPE3257}, MoSGrid~\cite{kruger2014mosgrid}$^w$, System in~\cite{wu2010accelerating}, Vine Toolkit~\cite{DBLP:journals/scpe/SzejnfeldDKKKKLPTWDNW10}, Virtual Imaging Platform~\cite{GLAT-13}$^n$, WS-PGRADE/gUSE framework~\cite{Kacsuk2012},
-            Science gateways in~\cite{kacsuk2014science}$^w$, GridSFEA~\cite{Elts2010}
+             \correction{Gateways based on} Apache Airvata~\cite{marru2011apache}, HubZero with Pegasus~\cite{CPE:CPE3257}, System in~\cite{wu2010accelerating}, Vine Toolkit~\cite{DBLP:journals/scpe/SzejnfeldDKKKKLPTWDNW10}, Virtual Imaging Platform~\cite{GLAT-13},           \correction{Gateways based on the} WS-PGRADE/gUSE framework~\cite{Kacsuk2012}, GridSFEA~\cite{Elts2010}
             } \\
-  Task & CBRAIN and PSOM~\cite{GLAT-16}$^n$, CBRAIN and FSL$^n$\\
+  Task & CBRAIN and PSOM~\cite{GLAT-16}, CBRAIN and FSL\\
   Pool & SHIWA pool~\cite{ROGE-13}\\
-  Nested & \pbox{1.5\columnwidth}{SHIWA Simulation Platform (Coarse-Grained Interoperability~\cite{terstyanszky2014enabling})$^w$, HubZero with Pegasus (via hierarchical workflows)~\cite{Deelman201517}, Tavaxy~\cite{Abouelhoda2012}.
+  Nested & \pbox{1.5\columnwidth}{SHIWA Simulation Platform (Coarse-Grained Interoperability~\cite{terstyanszky2014enabling}), HubZero with Pegasus (via hierarchical workflows)~\cite{Deelman201517}, Tavaxy~\cite{Abouelhoda2012}.
            }\\
-  Conversion & SHIWA Simulation Platform (Fine-Grained Interoperability)~\cite{plankensteiner-prodan-etal:2013}$^w$, Tavaxy~\cite{Abouelhoda2012}, system in~\cite{delaGarza2016}.
+  Conversion & SHIWA Simulation Platform (Fine-Grained Interoperability)~\cite{plankensteiner-prodan-etal:2013}, Tavaxy~\cite{Abouelhoda2012}, system in~\cite{delaGarza2016}.
 \end{tabular}
-\caption{Classification of science gateways based on the architecture used to integrate workflow engines.
-$^c$: based on the Catania Science Gateway Framework.
-$^w$: based on WS-PGRADE/gUSE.
-$^n$: used for neuroimaging.}
+\caption{Classification of science gateways based on the architecture used to integrate workflow engines.}
 \label{table:system-classification}
@@
 software components with the science gateway. For instance, the
-workflow engine might be a portlet in a Liferay portal or a model
-in a Ruby on Rails application. The workflow engine and the science
-gateway usually share a database where applications, users and other
-resources are stored.  In this model, task and
-data control are both initiated from the science gateway. Interactions
-\texttt{b} and \texttt{c$_2$} are initiated from the workflow engine,
-while \texttt{c$_1$} comes from other parts of the science gateway,
-for instance a data management interface. As in any other model, the
-installation of new workflows in the science gateway (\texttt{a$_2$})
-and infrastructure (\texttt{a$_1$}) is done by an administrator, for
-security reasons. This is the model adopted in the Catania Science
-Gateway Framework~\cite{ardizzone2012decide} (see specific documentation on
+workflow engine might be a portlet in a Liferay portal or a model in a
+Ruby on Rails application\footnote{\correction{Note that the
+    particular case of portlets can be a bit ambiguous: when portlets
+    are deployed remotely, using a protocol such as Web Services for
+    Remote Portlets (WSRP), they become independent software
+    components according to our definition because they could be
+    operated by different teams on different hardware systems. In this
+    case, the system should be classified in the service invocation
+    architecture since the portlets actually become Web
+    Services. The same comment holds for remote object
+    invocation.}}. The workflow engine and the science gateway usually
+share a database where applications, users and other resources are
+stored.  In this model, task and data control are both initiated from
+the science gateway. Interactions \texttt{b} and \texttt{c$_2$} are
+initiated from the workflow engine, while \texttt{c$_1$} comes from
+other parts of the science gateway, for instance a data management
+interface. As in any other model, the installation of new workflows in
+the science gateway (\texttt{a$_2$}) and infrastructure
+(\texttt{a$_1$}) is done by an administrator, for security
+reasons. This is the model adopted in the Catania Science Gateway
+Framework~\cite{ardizzone2012decide} (see specific documentation on
 workflows\footnote{\url{http://bit.ly/1oQrzvQ}}), in the Distributed
@@
 
+\correction{This section illustrates the architectural patterns on a
+  few real systems. The architecture diagrams presented are the
+  original ones as published by the system developers, which we
+  annotated using the interactions and components in
+  Figure~\ref{fig:architectures}. }
+
+\subsection{Service invocation}
+
 Figure~\ref{fig:vip-architecture} shows the architecture of the
@@
 \centering
-\def\svgwidth{1.5\columnwidth}
-\input{figures/VIP.pdf_tex}
+%\def\svgwidth{1.5\columnwidth}
+\includegraphics[width=0.9\textwidth]{figures/VIP}
 \caption{Architecture of the Virtual Imaging Platform (service
@@
 \end{figure*}
+
+\subsection{Task encapsulation}
 
@@
 for a criterion are comparable, so that they can be combined
-meaningfully. \correction{All the metrics in a criterion contribute
-  equally to the global score. Should the metrics be weighted
-  differently in a particular system, the reader is invited to produce
-  his/her own version of the Table using the public spreadsheet
-  available at \url{https://frama.link/LrPRsCQp} (see column B of
-  sheet ``Metrics'').}
+meaningfully.
+
+\correction{Note that the metrics defined hereafter aim at providing a framework
+to evaluate architectures independently from any particular system,
+regardless of any technical constraint such as the use of a specific
+programming language or the internal architecture of the science
+gateway. The goal is to abstract architectural discussions from the
+particular technical context of a real system. Some degree of
+system-specificity can be introduced by assigning weights to
+interactions and components through a public spreadsheet available at
+\url{https://frama.link/LrPRsCQp}. In particular, sheet "Components
+and interactions" contains the weight given to these entities in the
+metrics and can be used to produce new versions of
+Table~\ref{table:evaluation} that fit the particular technical context
+of a real system. For instance, a high weight could be assigned to the
+science gateway if its internal architecture is such that it can
+hardly be extended in practice.}
+
+\correction{In addition, all the metrics in a criterion contribute
+  equally to the global score used for the criterion. Column B of
+  sheet ``Metrics'' of the public spreadsheet may be used to assign
+  different weights to each metric to adjust the specific context of a
+  study.}
 

### 2016-10-12 figures/export.sh

@@ -1,6 +1,6 @@
 #!/bin/bash
 
-for name in tight service sub-task agent nested-1 nested-2 nested-3 import-1 import-2 notations VIP CBRAIN-PSOM CBRAIN-FSL
+for name in tight service sub-task agent nested-1 nested-2 nested-3 import-1 import-2 notations CBRAIN-PSOM CBRAIN-FSL
 do
   echo "Exporting ${name}.svg"
   inkscape -D -z --file=figures/${name}.svg --export-pdf=figures/${name}.pdf --export-latex

### 2016-10-12 figures/export.sh
Updates in figures

@@ -1,6 +1,6 @@
 #!/bin/bash
 
-for name in tight service sub-task agent nested-1 nested-2 nested-3 import-1 import-2 notations CBRAIN-PSOM CBRAIN-FSL
+for name in tight service sub-task agent nested-1 nested-2 nested-3 import-1 import-2 notations
 do
   echo "Exporting ${name}.svg"
   inkscape -D -z --file=figures/${name}.svg --export-pdf=figures/${name}.pdf --export-latex

### 2016-10-13 answers.tex
Still more answers to the reviewers

@@
 
+\begin{answer}
+  The revised manuscript now adds examples for tight integration
+  (Catania Science Gateway Framework) and workflow conversion
+  (Tavaxy).
+\end{answer}
+
 \begin{review}
@@
 \begin{review}
+  We agree with the reviewer that the Figures are difficult to
+  understand, however, we think it is important that the original
+  figures describing the system architectures are used. Indeed, a goal
+  of the examples is to illustrate the matching between our
+  architectures and the real system. As a compromise, we edited the
+  original figures with red color, to indicate where the software
+  components and interactions of our architectures fit. We hope the
+  figures are easier to read now.
+\end{review}
+
+\begin{review}
   5.  Subsection 2.4, "see for instrance attribute depend of option -W
@@
 \end{review}
+
+\begin{answer}
+We rephrased as follows:
+\emph{The science gateway may also implement mechanisms to facilitate the handling of task
+dependencies, for instance lists of tasks that must
+be completed before a particular task can be executed.}
+\end{answer}
 
@@
 \end{review}
+
+\begin{answer}
+  We followed the recommendation of the reviewer and created a new
+  Section 3 to present and discuss these examples. The caption of
+  Figures 5 and 6 is still long though, because we think that the
+  detailed references to the interaction numbers on the Figures should
+  be part of them, so that the reader doesn't have to constantly
+  navigate between the text and the Figures.
+\end{answer}
 

### 2016-10-13 paper.tex
Updates in figures
Still more answers to the reviewers

@@
 \usepackage{enumitem}
-\usepackage{hyperref}
+\usepackage[bookmarks]{hyperref}
 \usepackage{multirow}
@@
 underlines the functional interactions between their main software
-components. \correction{In Section~\ref{sec:metrics}, we introduce
-  metrics that measure integration complexity, robustness,
-  extensibility, scalability and \correction{functionality}. These
-  metrics are specifically designed to measure the ability of the
-  architectures to address workflow-related issues commonly
-  encountered in science gateways. In Section~\ref{sec:evaluation}, we
-  evaluate the different architectures based on the proposed metrics.}
-We finally compare the architectures and highlight recommendations for
-science gateway architects and developers.
+components. \correction{We illustrate the application of our
+  architectural patterns to real systems in Section~\ref{sec:real
+    systems}. In Section~\ref{sec:metrics}, we introduce metrics that
+  measure integration complexity, robustness, extensibility,
+  scalability and \correction{functionality}. These metrics are
+  specifically designed to measure the ability of the architectures to
+  address workflow-related issues commonly encountered in science
+  gateways. In Section~\ref{sec:evaluation}, we evaluate the different
+  architectures based on the proposed metrics.} \correction{Based on
+  this evaluation, we compare and discuss the architectures in
+  Section~\ref{sec:discussion} where we also highlight}
+recommendations for science gateway architects and
+developers. \correction{Finally, in Section~\ref{sec:application}, we
+  illustrate how our evaluation framework can be used to help design
+  new systems. Related work is presented in Section~\ref{sec:related}.}
 
@@
     particular case of portlets can be a bit ambiguous: when portlets
-    are deployed remotely, using a protocol such as Web Services for
+    are deployed remotely, using protocols such as Web Services for
     Remote Portlets (WSRP), they become independent software
@@
     case, the system should be classified in the service invocation
-    architecture since the portlets actually become Web
-    Services. The same comment holds for remote object
-    invocation.}}. The workflow engine and the science gateway usually
-share a database where applications, users and other resources are
-stored.  In this model, task and data control are both initiated from
-the science gateway. Interactions \texttt{b} and \texttt{c$_2$} are
-initiated from the workflow engine, while \texttt{c$_1$} comes from
-other parts of the science gateway, for instance a data management
-interface. As in any other model, the installation of new workflows in
-the science gateway (\texttt{a$_2$}) and infrastructure
-(\texttt{a$_1$}) is done by an administrator, for security
-reasons. This is the model adopted in the Catania Science Gateway
-Framework~\cite{ardizzone2012decide} (see specific documentation on
+    architecture since the portlets actually become Web Services. The
+    same comment holds for remote object invocation.}}. The workflow
+engine and the science gateway usually share a database where
+applications, users and other resources are stored.  In this model,
+task and data control are both initiated from the science
+gateway. Interactions \texttt{b} and \texttt{c$_2$} are initiated from
+the workflow engine, while \texttt{c$_1$} comes from other parts of
+the science gateway, for instance a data management interface. As in
+any other model, the installation of new workflows in the science
+gateway (\texttt{a$_2$}) and infrastructure (\texttt{a$_1$}) is done
+by an administrator, for security reasons. This is the model adopted
+in the Catania Science Gateway Framework~\cite{ardizzone2012decide}
+(see specific documentation on
 workflows\footnote{\url{http://bit.ly/1oQrzvQ}}), in the Distributed
@@
 a particular task that can submit sub-tasks to the science gateway
-through interaction \texttt{e}. The workflow engine keeps track of
-the dependencies between the sub-tasks, but their execution is
-delegated to the science gateway that executes them on the
-infrastructure through interaction \texttt{b}. Although the science
-gateway has no global vision of the workflow, it completely \correction{controls}
-the sub-tasks. \correction{For instance, it may cancel} them when
-the \correction{workflow} is canceled. The science gateway may also implement
-mechanisms to facilitate the handling of task dependencies, for
-instance basic dependency lists as available in most cluster managers
-(see for instance attribute \texttt{depend} of option \texttt{-W} in
-Torque\footnote{\url{http://docs.adaptivecomputing.com}}).
+through interaction \texttt{e}. The workflow engine keeps track of the
+dependencies between the sub-tasks, but their execution is delegated
+to the science gateway that executes them on the infrastructure
+through interaction \texttt{b}. Although the science gateway has no
+global vision of the workflow, it completely \correction{controls} the
+sub-tasks. \correction{For instance, it may cancel} them when the
+\correction{workflow} is canceled. The science gateway may also
+implement mechanisms to facilitate the handling of task dependencies,
+for instance \correction{lists of tasks that must be completed before a
+  particular task can be executed}.
 
@@
 
-\correction{\section{Architectural patterns in real systems}}
+\correction{\section{Architecture of real systems}}
 
@@
   annotated using the interactions and components in
-  Figure~\ref{fig:architectures}. }
+  Figure~\ref{fig:architectures}, in red color. Application porting
+  interactions \texttt{a$_1$} and \texttt{a$_2$} are not represented
+  on these diagrams.
+
+\subsection{Tight integration}
+
+The architecture of the Catania Science Gateway Framework (CSGF) is
+shown in Figure~\ref{fig:csgf}. In CSGF, workflows are integrated as
+Liferay portlets, i.e., as part of the Science Gateway when no
+specific protocol is used for remote invocation. A specific portlet
+called \texttt{mi-parallel-portlet} allows to create job collections,
+parametric jobs and split-and-merge types of workflows. The science
+gateway also includes a set of services to manage data and tasks
+(jobs) on a wide array of grid middleware systems through the SAGA
+API.
+\begin{figure}
+\centering
+\includegraphics[width=\columnwidth]{figures/CSGF}
+\caption{\correction{Architecture of the Catania Science Gateway Framework (tight
+  integration). Figure adapted from~\cite{ardizzone2012decide} with
+  permission of the first author} \todo{approval still pending}.}
+\label{fig:csgf}
+\end{figure}
 
@@
 Virtual Imaging Platform, where the MOTEUR workflow
-engine~\cite{GLAT-08i} is invoked through a service.
-
+engine~\cite{GLAT-08i} is invoked through a service. The interactions
+in the VIP architecture map to the ones defined by our framework in a
+straightforward manner, except for interaction \texttt{c$_2$}
+(management of workflow data) which, in VIP, is triggered by the
+workflow tasks instead of the workflow engine. However, this does not
+contradict the service invocation architecture since the tasks are
+generated by the workflow engine.
 \begin{figure*}
-\centering
 %\def\svgwidth{1.5\columnwidth}
-\includegraphics[width=0.9\textwidth]{figures/VIP}
+\includegraphics[width=0.9\textwidth]{figures/VIP.eps}
 \caption{Architecture of the Virtual Imaging Platform (service
   invocation).  Step \texttt{3} corresponds to interaction \texttt{d}
-  in Figure~\ref{fig:architectures}(II), step \texttt{2} corresponds to
-  interaction \texttt{c$_1$}, step \texttt{4} maps to interaction
-  \texttt{b}, and steps 7 and 9 map to \texttt{c$_2$} (in VIP, workflow data transfers are
-  performed by the tasks and embedded in their descriptions).}
+  in Figure~\ref{fig:architectures}(II), step \texttt{2} corresponds
+  to interaction \texttt{c$_1$}, step \texttt{4} maps to interaction
+  \texttt{b}, and steps 7 and 9 map to \texttt{c$_2$} (in VIP,
+  workflow data transfers are performed by the tasks and embedded in
+  their descriptions).}
 \label{fig:vip-architecture}
@@
 
-Task encapsulation is implemented in CBRAIN~\cite{SHER-14} where it is
-used to integrate the FSL toolkit~\cite{Jenkinson2012782} and the PSOM
-workflow engine~\cite{bellec2012pipeline}. The CBRAIN-FSL integration
-allows leveraging FSL pipelines written in low-level workflow
-languages (Linux executables and scripts) that submit tasks uniformly
-through a specific tool called fsl\_sub. This integration is shown in
-Figure~\subref*{fig:cbrain-fsl-architecture}, where the science gateway is
-represented by components \texttt{CBRAIN portal} and \texttt{CBRAIN
-  execution server} and the infrastructure consists of \texttt{Storage
-  servers} and \texttt{Computing cluster with shared file system}.
+Task encapsulation is used in the CBRAIN gateway~\cite{SHER-14} to
+integrate the FSL toolkit~\cite{Jenkinson2012782} and the PSOM
+workflow engine~\cite{bellec2012pipeline}. The CBRAIN-FSL integration,
+shown in Figure~\subref*{fig:cbrain-fsl-architecture}, allows
+leveraging FSL pipelines written in low-level workflow languages
+(Linux executables and scripts) that submit tasks uniformly through a
+specific tool called fsl\_sub. The science gateway is actually split
+in two services, the CBRAIN portal and the CBRAIN execution server,
+deployed on a head node of the infrastructure. The mapping of
+interactions \texttt{b}, \texttt{c$_1$} and \texttt{e} to the
+CBRAIN-FSL interactions is straightforward.
 
@@
 \subfloat[CBRAIN-FSL integration.]{
-\def\svgwidth{\columnwidth}
-\input{figures/CBRAIN-FSL.pdf_tex}
+\includegraphics[width=\columnwidth]{figures/CBRAIN-FSL.eps}
 \label{fig:cbrain-fsl-architecture}
 } \hfill \subfloat[CBRAIN-PSOM integration. Figure
-reproduced from~\cite{GLAT-16}.]{\def\svgwidth{\columnwidth}
-  \input{figures/CBRAIN-PSOM.pdf_tex
-}
+reproduced from~\cite{GLAT-16}.]{
+  \includegraphics[width=\columnwidth]{figures/CBRAIN-PSOM.eps}
 \label{fig:cbrain-psom-architecture}
@@
 
+\subsection{Pool model}
+
 The pool model was implemented in the SHIWA pool~\cite{ROGE-13}
@@
 can wrap different types of workflow engines to execute workflows
-expressed with different languages.
+expressed with different languages. The software components map
+directly to the ones in our architecture, except that the
+infrastructure is not represented. The interactions described in the
+SHIWA pool are much finer-grained than in our model, but they can be
+grouped to match interactions \texttt{d} and \texttt{f}, as explained
+in the Figure caption.
 
@@
   of the \texttt{agent} component and therefore have no corresponding
-  representation in Figure~\ref{fig:architectures}(IV). Figure reproduced
+  representation in Figure~\ref{fig:architectures}(IV). Figure adapted
   from~\cite{ROGE-13}.}
@@
 
-Figure~\ref{fig:shiwa-architecture} shows the architecture
-used in the SHIWA Simulation Platform for nested workflow execution
-with service invocation.
+\subsection{Nested workflows with service invocation}
+
+Figure~\ref{fig:shiwa-architecture} shows the architecture used in the
+SHIWA Simulation Platform for nested workflow execution with service
+invocation. The parent workflow engine is \texttt{WS-PGRADE/gUSE},
+invoked as a service in the Science Gateway (interaction
+\texttt{d}). Ten different child engines can be used by nested
+workflows, invoked through the \texttt{Submission service}
+(interaction \texttt{d}). Each of these engines can submit tasks and
+transfer data to a distributed computing infrastructure (\texttt{DCI}
+interactions \texttt{b} and \texttt{c$_2$}). User data interactions
+(\texttt{c$_1$}) and interactions between the parent engine and the
+infrastructure (\texttt{b} and \texttt{c$_2$}) are not represented.
 \begin{figure*}
 \centering
-\includegraphics[width=1.5\columnwidth]{figures/shiwa-science-gateway.pdf}
-\caption{Nested workflow execution through the SHIWA Science Gateway. The
-  parent workflow engine is \texttt{WS-PGRADE/gUSE}, invoked as a service
-  in the Science Gateway (step \texttt{1}, interaction \texttt{d} on
-  Figure~\ref{fig:architectures}(V)-Right). Ten different child engines can be used by nested
-  workflows, invoked through the \texttt{Submission service} (step
-  \texttt{2}, interaction \texttt{d}). Each of these engines can
-  submit tasks and transfer data to a distributed computing
-  infrastructure (\texttt{DCI}, step \texttt{3}, interactions
-  \texttt{b} and \texttt{c$_2$}). User data interactions (\texttt{c$_1$}) and
-  application porting ones (\texttt{a}) are not represented. Figure
-  reproduced from~\cite{terstyanszky2014enabling} with permission of
+\includegraphics[width=1.5\columnwidth]{figures/shiwa-science-gateway.png}
+\caption{Nested workflow execution through the SHIWA Science Gateway. Figure
+  adapted from~\cite{terstyanszky2014enabling} with permission of
   the first author.}
@@
 
-\section{\correction{Metrics}}
+\subsection{Workflow conversion with service invocation}
+
+Figure~\ref{fig:tavaxy} shows the architecture of Tavaxy where
+workflows in various formats (tSCUFL, SCUFL, t2flow and JSON) can be
+imported. As detailed in~\cite{Abouelhoda2012}, workflow conversion
+(interaction \texttt{g}) is implemented through an interactive
+authoring module where users can check and edit workflow imports, and
+a mapper module that performs some language substitutions. The
+workflow engine is a re-engineered version of the Galaxy engine that
+also includes some components of Taverna.
+\begin{figure}
+\centering
+\includegraphics[width=\columnwidth]{figures/Tavaxy.eps}
+\caption{\correction{Architecture of Tavaxy (workflow conversion). Figure adapted
+  from~\cite{Abouelhoda2012} with permission of the first author} \todo{approval is still pending}.}
+\label{fig:tavaxy}
+\end{figure}
+
+}
+
+\section{\correction{Evaluation metrics}}
 
@@
 
-Science gateway \emph{integration complexity} is measured through the number of software components and their interactions. It breaks
-down into the following 2 metrics:
+\subsection{\correction{Integration complexity}}
+
+\correction{Integration complexity} is measured through the number of
+software components and their interactions. It breaks down into the
+following 2 metrics:
 \begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
@@
 
-\emph{Robustness of workflow execution} measures the likelihood that
+\subsection{\correction{Robustness of workflow execution}}
+
+Robustness of workflow execution measures the likelihood that
 the workflow execution fails due to errors in the components or in the
@@
 
-\emph{Extensibility} measures the difficulty to replace or add
-elements to the architecture. It is determined by the number of interactions and
+\subsection{\correction{Extensibility}}
+
+Extensibility measures the difficulty to replace or add elements to
+the architecture. It is determined by the number of interactions and
 components that need to be modified when a new element is
@@
 parameters involved in this interaction are modified.  Extensibility
-breaks down into 4 metrics depending on the type of element (engine, workflow or infrastructure) that has to
-be added or replaced:
+breaks down into 4 metrics depending on the type of element (engine,
+workflow or infrastructure) that has to be added or replaced:
 \begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
@@
 
-\emph{Scalability} corresponds to the ability of an architecture to
+\subsection{\correction{Scalability}}
+
+Scalability corresponds to the ability of an architecture to
 cope with high workloads through the following mechanisms: starting multiple engine instances, distributing engines and task scheduling. 
@@
 
-\correction{\emph{Functionality} includes}:
+\subsection{\correction{Functionality}}
+
+\correction{Functionality includes}:
 \begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
@@
 
-\section{\correction{Evaluation}}
+\section{\correction{Architecture evaluation}}
 
@@
 
-\correction{\section{Application to systems design}}
-\label{sec:application}
-
-\correction{To illustrate how our evaluation framework can be applied
-  to the design of new systems, let's consider a hypothetical system
-  where the workflow engine is split in two parts: (1) a high-level
-  part integrated in the science gateway, which submits sub-workflows
-  to the infrastructure; (2) a lower-level part, executed on the
-  infrastructure, which executes the sub-workflows. This scenario came
-  up in the discussion with Reviewer \#2 of this paper, after the
-  evaluation framework had already been designed.
-
-  We model such a system using a nested workflow architecture with a
-  ``Tight+Task'' instantiation, i.e., the parent engine is tightly
-  integrated in the science gateway and the child engine is
-  integrated through task encapsulation.  Figure~\ref{fig:nested-3}
-  shows the graphical representation of the system. From this
-  representation we find that the sub-tasks submitted by the child
-  engine through interaction \texttt{e} should be handled by the
-  parent engine rather than by the science gateway. Indeed, handling
-  the tasks directly in the science gateway would lead to create a new
-  specific interaction, named, e.g., \texttt{b$_1$}, while \texttt{b}
-  could be reused. We also find that interaction \texttt{e} allows
-  running the sub-tasks on the infrastructure supported by the parent
-  engine with no further development effort required in the child
-  engine.
-
-The system evaluation is presented on
-Table~\ref{table:evaluation-2}. The detailed calculations are
-available in sheet ``Metrics with reviewer 2's architecture'' of the public spreadsheet at \url{https://frama.link/LrPRsCQp} \todo{revise that sheet to merge it with the main one}.
-
-We find that the studied system supports nested workflows with a much
-better performance than the nested workflow model instantiated with
-service invocation (evaluated on Table~\ref{table:evaluation}): the
-integration score is improved by a factor of 2, robustness is
-improved by a factor of 2.6, extensibility is substantially
-improved, scalability and functionality are not penalized.
-
-Based on this system, we propose a new system design that instantiates
-the nested workflow model in a ``Task+Task'' way, i.e., with task
-encapsulation for both the parent and the child engine. The
-architecture of this system is represented in
-Figure~\ref{fig:nested-4}. The diagram is identical to the one in
-Figure~\ref{archi:sub-task}, except that interaction \texttt{e} is now
-used to submit sub-workflows instead of just sub-tasks. This is
-possible since, in the task encapsulation architecture, workflows are
-executed as tasks.
-
-The new architecture performs better similarly to task encapsulation,
-except that it now enables meta-workflows and distributed engines
-which were not available previously. It clearly outperforms the nested
-``Service+Service'' architecture.
-
-Based on this analysis, we plan to extend the PSOM-CBRAIN integration
-to support this architecture. Conceptually, the implementation is
-simple, as it only requires existing workflows to be declarable as
-tasks in the workflow engine. In practice, technical issues may delay
-the process. We also remain well aware of the scheduling issues raised
-by the task encapsulation architecture (S$_3$=1), which will be
-amplified in the ``Task+Task'' instantiation of nested workflows: for
-instance, properly setting resource requirements (e.g. walltime) is
-increasingly difficult when nested workflow engines are deployed on
-the infrastructure.  }
-
-\begin{figure}
-  \centering
-\subfloat[Nested, Tight+Task]{
-  \def\svgwidth{0.5\columnwidth}
-  \input{figures/nested-3.pdf_tex}
-  \label{fig:nested-3}
-}
-\subfloat[Nested, Task+Task]{ 
-  \def\svgwidth{0.5\columnwidth}
-  \input{figures/sub-task.pdf_tex}
-  \label{fig:nested-4}
-}
-\caption{\correction{Nested workflow architecture instantiated with tight integration and task encapsulation (top) and task encapsulation only (bottom).}}
-\label{fig:new-systems}
-\end{figure}
-
-\begin{table}
-\footnotesize
-\centering
-\begin{tabular}{rcc}
-                                     & \textbf{Nested} & \textbf{Nested}\\
-                                     & \textbf{Tight+Task} & \textbf{Task+Task}\\
-\cellcolor[HTML]{EEEEEE}\textbf{Integration complexity}&\cellcolor[HTML]{EEEEEE}&\cellcolor[HTML]{EEEEEE}\\
-  Total components -- \texttt{I$_1$} & 2 & 2 \\
-Total interactions -- \texttt{I$_2$} & 6 & 5 \\
-\textbf{Total}& \textbf{8} & \textbf{7}\\
-\cellcolor[HTML]{EEEEEE}\textbf{Robustness}&\cellcolor[HTML]{EEEEEE}&\cellcolor[HTML]{EEEEEE}\\
-Specific components -- \texttt{R$_1$} & 0 & 0\\
-  Specific interactions -- \texttt{R$_2$} & 3 & 1\\
-  \textbf{Total}& \textbf{3} & \textbf{1}\\
-\cellcolor[HTML]{EEEEEE}\textbf{Extensibility}&\cellcolor[HTML]{EEEEEE}&\cellcolor[HTML]{EEEEEE}\\
-  New engine type -- \texttt{E$_1$}  & 3 & 2\\
-New engine version -- \texttt{E$_2$} & 1 & 1\\
-  New workflow -- \texttt{E$_3$}     & 2 & 2\\
-New infrastructure -- \texttt{E$_4$} & 4 & 3\\
-  \textbf{Total}& \textbf{10} & \textbf{8}\\
-\cellcolor[HTML]{EEEEEE}\textbf{Scalability}& \cellcolor[HTML]{EEEEEE}&\cellcolor[HTML]{EEEEEE}\\
-Multiple engine instances -- \texttt{S$_1$}& 1 & 0\\
-Distributed engines -- \texttt{S$_2$}& 0 & 0\\
-Task scheduling -- \texttt{S$_3$}    & 1 & 1\\
-\textbf{Total}& \textbf{2} & \textbf{1}\\
-\cellcolor[HTML]{EEEEEE}\textbf{\correction{Functionality}}&\cellcolor[HTML]{EEEEEE}&\cellcolor[HTML]{EEEEEE}\\
-  Meta-workflow  -- \texttt{\correction{F}$_1$}    & 0 & 0\\
-  Fine-grained debugging -- \texttt{\correction{F}$_2$}  & 1 & 1\\
-  \textbf{Total}& \textbf{1} & \textbf{1}\\
-\end{tabular}
-\caption{\correction{Evaluation of the nested workflow architecture instantiated with tight integration (parent engine) and task encapsulation (child engine).}}
-\label{table:evaluation-2}
-\end{table}
-
 \section{Discussion}
-
+\label{sec:discussion}
 \subsection{Comparison between architectures}
@@
 
+\correction{\section{Application to systems design}}
+\label{sec:application}
+
+\correction{To illustrate how our evaluation framework can be applied
+  to help design new systems, let's consider a hypothetical system
+  where the workflow engine is split in two parts: (1) a high-level
+  part integrated in the science gateway, which submits sub-workflows
+  to the infrastructure; (2) a lower-level part, executed on the
+  infrastructure, which executes the sub-workflows. This scenario came
+  up in the discussion with Reviewer \#2 of this paper, after the
+  evaluation framework had already been designed.
+
+  We model such a system using a nested workflow architecture with a
+  ``Tight+Task'' instantiation, i.e., the parent engine is tightly
+  integrated in the science gateway and the child engine is
+  integrated through task encapsulation.  Figure~\ref{fig:nested-3}
+  shows the graphical representation of the system. From this
+  representation we find that the sub-tasks submitted by the child
+  engine through interaction \texttt{e} should be handled by the
+  parent engine rather than by the science gateway. Indeed, handling
+  the tasks directly in the science gateway would lead to create a new
+  specific interaction, named, e.g., \texttt{b$_1$}, while \texttt{b}
+  could be reused. We also find that interaction \texttt{e} allows
+  running the sub-tasks on the infrastructure supported by the parent
+  engine with no further development effort required in the child
+  engine.
+
+  The system evaluation is presented on
+  Table~\ref{table:evaluation-2}. The detailed calculations are
+  available in our public spreadsheet at
+  \url{https://frama.link/LrPRsCQp}. We conclude that the studied system
+  supports nested workflows with a much better performance than the
+  nested workflow model instantiated with service invocation
+  (evaluated on Table~\ref{table:evaluation}): the integration score
+  is improved by a factor of 2, robustness is improved by a factor of
+  2.6, extensibility is substantially improved, scalability and
+  functionality are not penalized.  }
+
+\begin{figure}
+  \centering
+  \def\svgwidth{0.5\columnwidth}
+  \input{figures/nested-3.pdf_tex}
+\caption{\correction{Nested workflow architecture instantiated with tight integration and task encapsulation (Tight+Task).}}
+\label{fig:nested-3}
+\end{figure}
+
+\begin{table}
+\footnotesize
+\centering
+\begin{tabular}{rc}
+                                     & \textbf{Nested}\\
+                                     & \textbf{Tight+Task}\\
+\cellcolor[HTML]{EEEEEE}\textbf{Integration complexity}&\cellcolor[HTML]{EEEEEE}\\
+  Total components -- \texttt{I$_1$} & 2\\
+Total interactions -- \texttt{I$_2$} & 6\\
+\textbf{Total}& \textbf{8}\\
+\cellcolor[HTML]{EEEEEE}\textbf{Robustness}&\cellcolor[HTML]{EEEEEE}\\
+Specific components -- \texttt{R$_1$} & 0\\
+  Specific interactions -- \texttt{R$_2$} & 3\\
+  \textbf{Total}& \textbf{3}\\
+\cellcolor[HTML]{EEEEEE}\textbf{Extensibility}&\cellcolor[HTML]{EEEEEE}\\
+  New engine type -- \texttt{E$_1$}  & 3\\
+New engine version -- \texttt{E$_2$} & 1\\
+  New workflow -- \texttt{E$_3$}     & 2\\
+New infrastructure -- \texttt{E$_4$} & 4\\
+  \textbf{Total}& \textbf{10}\\
+\cellcolor[HTML]{EEEEEE}\textbf{Scalability}& \cellcolor[HTML]{EEEEEE}\\
+Multiple engine instances -- \texttt{S$_1$}& 1\\
+Distributed engines -- \texttt{S$_2$}& 0\\
+Task scheduling -- \texttt{S$_3$}    & 1\\
+\textbf{Total}& \textbf{2}\\
+\cellcolor[HTML]{EEEEEE}\textbf{\correction{Functionality}}&\cellcolor[HTML]{EEEEEE}\\
+  Meta-workflow  -- \texttt{\correction{F}$_1$}    & 0\\
+  Fine-grained debugging -- \texttt{\correction{F}$_2$}  & 1\\
+  \textbf{Total}& \textbf{1}\\
+\end{tabular}
+\caption{\correction{Evaluation of the nested workflow architecture instantiated with tight integration (parent engine) and task encapsulation (child engine).}}
+\label{table:evaluation-2}
+\end{table}
+
 \section{Related work}

### 2016-10-13 paper.tex
Fixed colors in tables

@@
   New engine type -- \texttt{E$_1$}  & \cellcolor[HTML]{99C599}3
-                                     & \cellcolor[HTML]{99A899}\correction{3}
+                                     & \cellcolor[HTML]{99C599}\correction{3}
                                      & \cellcolor[HTML]{99E299}2
@@
   \textbf{Total} (difficulty to extend) & \cellcolor[HTML]{99D299}\textbf{10}
-                                     & \cellcolor[HTML]{99D299}\textbf{\correction{9}}
+                                     & \cellcolor[HTML]{99E899}\textbf{\correction{9}}
                                      & \cellcolor[HTML]{99FF99}\textbf{8}
@@
                                      & \cellcolor[HTML]{999999}1
-                                     & \cellcolor[HTML]{99FF99}\correction{1}
+                                     & \cellcolor[HTML]{999999}\correction{1}
                                      & \cellcolor[HTML]{999999}1
@@
                                      & \cellcolor[HTML]{999999}\textbf{2}
-                                     & \cellcolor[HTML]{99FF99}\textbf{\correction{2}}
+                                     & \cellcolor[HTML]{999999}\textbf{\correction{2}}
                                      & \cellcolor[HTML]{999999}\textbf{2}
@@
   Extensibility (global score)  & \cellcolor[HTML]{99D299}{0.44}
-                                     & \cellcolor[HTML]{99D299}{\correction{0.22}}
+                                     & \cellcolor[HTML]{99E899}{\correction{0.22}}
                                      & \cellcolor[HTML]{99FF99}{0.00}
@@
                                      & \cellcolor[HTML]{999999}{1.00}
-                                     & \cellcolor[HTML]{99FF99}{0.00}
+                                     & \cellcolor[HTML]{999999}{\correction{1.00}}
                                      & \cellcolor[HTML]{999999}{1.00}
@@
                                      & \cellcolor[HTML]{99FF99}{0.00}\\
-                                    & \cellcolor[HTML]{99DA99}\textbf{1.6}
-                                    & \cellcolor[HTML]{99B899}\textbf{\correction{2.4}}
-                                    & \cellcolor[HTML]{99FF99}\textbf{0.5}
-                                    & \cellcolor[HTML]{99BF99}\textbf{2.4}
+                                    & \cellcolor[HTML]{99FA99}\textbf{1.6}
+                                    & \cellcolor[HTML]{99D299}\textbf{\correction{2.4}}
+                                    & \cellcolor[HTML]{99FE99}\textbf{\correction{1.5}}
+                                    & \cellcolor[HTML]{99D299}\textbf{2.4}
                                     & \cellcolor[HTML]{999999}\textbf{3.5}
-                                    & \cellcolor[HTML]{99DE99}\textbf{1.5}\\
+                                    & \cellcolor[HTML]{99FF99}\textbf{1.5}\\
 \end{tabular}

### 2016-10-13 paper.tex
Various fixes in paper

@@
 
-%% Title, authors and addresses
-
-%% use the tnoteref command within \title for footnotes;
-%% use the tnotetext command for theassociated footnotes;
-%% use the fnref command within \author or \address for footnotes;
-%% use the fntext command for theassociated footnote;
-%% use the corref command within \author for corresponding author footnotes;
-%% use the cortext command for theassociated footnote;
-%% use the ead command for the email address,
-%% and the form \ead[url] for the home page:
-%% \title{Title\tnoteref{label1}}
-%% \tnotetext[label1]{}
-%% \author{Name\corref{cor1}\fnref{label2}}
-%% \ead{email address}
-%% \ead[url]{home page}
-%% \fntext[label2]{}
-%% \cortext[cor1]{}
-%% \address{Address\fnref{label3}}
-%% \fntext[label3]{}
-
 \title{Software architectures to integrate workflow engines in science gateways}
-
-%% use optional labels to link authors explicitly to addresses:
-%% \author[label1,label2]{}
-%% \address[label1]{}
-%% \address[label2]{}
 
@@
   Science gateways often rely on workflow engines to execute
-  applications on distributed infrastructures. We investigate six software
-  architectures commonly used to integrate workflow engines into
-  science gateways. In \emph{tight integration}, the workflow engine
-  shares software components with the science gateway. In
+  applications on distributed infrastructures. We investigate six
+  software architectures commonly used to integrate workflow engines
+  into science gateways. In \emph{tight integration}, the workflow
+  engine shares software components with the science gateway. In
   \emph{service invocation}, the engine is isolated and invoked
@@
   assessment of integration complexity, robustness, extensibility,
-  scalability and \correction{functionality}. Tight integration and task encapsulation are the easiest
-  to integrate and the most robust. Extensibility is equivalent in
-  most architectures. The pool model is the most scalable one and
-  meta-workflows are only available in task encapsulation, nested
-  workflows and workflow conversion. These results provide insights
-  for science gateway architects and developers.
+  scalability and \correction{functionality}. Tight integration and
+  task encapsulation are the easiest to integrate and the most
+  robust. Extensibility is equivalent in most architectures. The pool
+  model is the most scalable one and meta-workflows are only available
+  in nested workflows and workflow conversion. These results provide
+  insights for science gateway architects and developers.
 \end{abstract}
@@
 
-Workflow engines are critical components in the ecosystem of tools
-and services offered by science gateways. Several software
-architectures can be adopted to integrate workflow engines in science
-gateways, with important consequences for the development effort
-required and resulting system.
-
-This paper describes, provides examples and compares such
+Several software architectures can be adopted to integrate workflow
+engines in \correction{the ecosystem of tools and services offered by}
+science gateways, with important consequences for the development
+effort required and resulting system.
+
+This paper describes, \correction{illustrates} and compares such
 architectures, based on system-independent representations of their
@@
 the development and sustained operation of the CBRAIN~\cite{SHER-14},
-VIP~\cite{GLAT-13} and NSG~\cite{shahand2015data,shahand:2012jgc}
-science gateways for medical image analysis during the past 7 years,
-as well as by lessons learned from several science gateway and
-workflow projects such as
+NSG~\cite{shahand2015data,shahand:2012jgc} and VIP~\cite{GLAT-13}
+science gateways during the past 7 years, as well as by lessons
+learned from several science gateway and workflow projects such as
 SHIWA\footnote{\url{http://www.shiwa-workflow.eu}} and
@@
 
-\correction{The analysis presented in the paper is intended for
-experts in science gateway and workflow engine design. It is an
-abstraction effort to identify and evaluate the fundamental
-architectural patterns that are encountered while integrating workflow
-engines and science gateways. In real systems, such patterns are often
-intertwined due to considerations related to the historical and
-technical context of software projects. The following work helps
-identify the key patterns in systems and highlight potential
-improvements.}
-
-\correction{The reminder of this section provides} background
+\correction{This analysis is intended for experts in science gateway
+  and workflow engine design. It is an abstraction effort to identify
+  and evaluate the fundamental architectural patterns that are
+  encountered while integrating workflow engines and science
+  gateways. In real systems, such patterns are often intertwined and
+  obfuscated due to the historical and
+  technical context of software projects.}
+
+\correction{The remainder of this section provides} background
 information and definitions of workflow engines, science gateways and
 infrastructures. In Section~\ref{sec:architectures}, we describe six
-architectures that cover the main patterns that we encountered in the
-various systems that we studied during the last few years. The
-architectures are described within a consistent framework that
-underlines the functional interactions between their main software
-components. \correction{We illustrate the application of our
-  architectural patterns to real systems in Section~\ref{sec:real
-    systems}. In Section~\ref{sec:metrics}, we introduce metrics that
-  measure integration complexity, robustness, extensibility,
-  scalability and \correction{functionality}. These metrics are
-  specifically designed to measure the ability of the architectures to
-  address workflow-related issues commonly encountered in science
-  gateways. In Section~\ref{sec:evaluation}, we evaluate the different
-  architectures based on the proposed metrics.} \correction{Based on
-  this evaluation, we compare and discuss the architectures in
-  Section~\ref{sec:discussion} where we also highlight}
-recommendations for science gateway architects and
-developers. \correction{Finally, in Section~\ref{sec:application}, we
-  illustrate how our evaluation framework can be used to help design
-  new systems. Related work is presented in Section~\ref{sec:related}.}
+architectures within a consistent framework that underlines the
+functional interactions between their main software
+components. \correction{We illustrate these architectures on real
+  systems in Section~\ref{sec:real systems}.} In
+  Section~\ref{sec:metrics}, we introduce metrics that measure
+  integration complexity, robustness, extensibility, scalability and
+  \correction{functionality}. The metrics are specifically designed to
+  measure the ability of the architectures to address workflow-related
+  issues commonly encountered in science gateways. \correction{We
+    evaluate the architectures individually in
+    Section~\ref{sec:evaluation}, and we compare and discuss them in
+    Section~\ref{sec:discussion}.} \correction{Finally, in
+    Section~\ref{sec:application}, we illustrate how our evaluation
+    framework can be used to help design new systems. Related work is
+    presented in Section~\ref{sec:related}.}
 
@@
   has a global view of the workflow execution. In a fine-grained
-  model, user is exposed to the workflow topology, i.e. to the outputs
+  model, user is exposed to the workflow topology, i.e., to the outputs
   of the individual tasks, their statuses, dependencies and so on.
@@
   sub-tasks may be terminated if the parent task is killed.
-\item Pool-agent: specific to the pool architecture described in
-  Section~\ref{sec:pool}. This is an interaction used when agents
-  retrieve work from a central pool. It covers agent registration and
-  de-registration to the pool, protocols to send work from the pool to
-  the agent, mechanisms to update work status, and so on. A similar
-  type of interaction is used in pilot-job
-  systems~\cite{turilli2015comprehensive} and other agent-based
-  computing models.
+\item Pool-agent: specific to the pool architecture. This is an
+  interaction used when agents retrieve work from a central pool. It
+  covers agent registration and de-registration to the pool, protocols
+  to send work from the pool to the agent, mechanisms to update work
+  status, and so on. A similar type of interaction is used in
+  pilot-job systems~\cite{turilli2015comprehensive} and other
+  agent-based computing models.
 \item Workflow conversion: translation from one workflow language into
@@
 gateway (\texttt{a$_2$}) and infrastructure (\texttt{a$_1$}) is done
-by an administrator, for security reasons. This is the model adopted
-in the Catania Science Gateway Framework~\cite{ardizzone2012decide}
-(see specific documentation on
-workflows\footnote{\url{http://bit.ly/1oQrzvQ}}), in the Distributed
-application runtime environment
+by an administrator, for security reasons. \correction{Tight
+  integration} is the model adopted in the Catania Science Gateway
+Framework~\cite{ardizzone2012decide} (see specific documentation on
+workflows\footnote{\url{https://frama.link/jfzgaEbj}}), in the
+Distributed application runtime environment
 (DARE)~\cite{maddineni2012distributed}, in
@@
 
-This architecture is largely adopted in systems such as Apache
-Airvata~\cite{marru2011apache}, GridSFEA~\cite{Elts2010}, Vine
+This architecture is largely adopted in systems such as
+\correction{the} Apache Airvata
+\correction{framework}~\cite{marru2011apache},
+GridSFEA~\cite{Elts2010}, Vine
 Toolkit~\cite{DBLP:journals/scpe/SzejnfeldDKKKKLPTWDNW10}, Virtual
@@
 \texttt{d} interaction is implemented as a set of calls to Pegasus'
-command-line tools (e.g., pegasus-status, pegasus-plan,
-etc.).
+command-line tools (e.g., pegasus-status, pegasus-plan, etc.).
 
@@
 \texttt{c$_2$}. Science gateway transfers user data to infrastructures
-using \texttt{c$_1$} interactions.
+using \texttt{c$_1$}.
 
@@
 so-called Coarse-Grained workflow
-interoperability~\cite{terstyanszky2014enabling}, i.e. to integrate
+interoperability~\cite{terstyanszky2014enabling}, i.e., to integrate
 various workflow engines in a consistent
@@
 
-\correction{\section{Architecture of real systems}}
+\correction{\section{Real systems}}
 
@@
 
-%\todo{Introduce this section and take the text out of the captions}
-
 \correction{This section illustrates the architectural patterns on a
-  few real systems. The architecture diagrams presented are the
-  original ones as published by the system developers, which we
-  annotated using the interactions and components in
-  Figure~\ref{fig:architectures}, in red color. Application porting
-  interactions \texttt{a$_1$} and \texttt{a$_2$} are not represented
-  on these diagrams.
+  few real systems, using the original architecture diagrams published
+  by the system developers. To improve readability, we annotated these
+  diagrams with the interactions and components used in
+  Figure~\ref{fig:architectures}.
 
@@
 shown in Figure~\ref{fig:csgf}. In CSGF, workflows are integrated as
-Liferay portlets, i.e., as part of the Science Gateway when no
+Liferay portlets, i.e., as part of the science gateway when no
 specific protocol is used for remote invocation. A specific portlet
@@
 Figure~\ref{fig:vip-architecture} shows the architecture of the
-Virtual Imaging Platform, where the MOTEUR workflow
+Virtual Imaging Platform (VIP), where the MOTEUR workflow
 engine~\cite{GLAT-08i} is invoked through a service. The interactions
@@
 (management of workflow data) which, in VIP, is triggered by the
-workflow tasks instead of the workflow engine. However, this does not
+workflow tasks instead of the workflow engine. This does not
 contradict the service invocation architecture since the tasks are
@@
 specific tool called fsl\_sub. The science gateway is actually split
-in two services, the CBRAIN portal and the CBRAIN execution server,
+in two services, the CBRAIN portal and the CBRAIN execution server
 deployed on a head node of the infrastructure. The mapping of
@@
 } \hfill \subfloat[CBRAIN-PSOM integration. Figure
-reproduced from~\cite{GLAT-16}.]{
+adapted from~\cite{GLAT-16}.]{
   \includegraphics[width=\columnwidth]{figures/CBRAIN-PSOM.eps}
@@
 \caption{\correction{Architecture of Tavaxy (workflow conversion). Figure adapted
-  from~\cite{Abouelhoda2012} with permission of the first author} \todo{approval is still pending}.}
+  from~\cite{Abouelhoda2012} with permission of the first author}.}
 \label{fig:tavaxy}
@@
 
-The architectures described in Section~\ref{sec:architectures} are
+The architectures are
 evaluated in Table~\ref{table:evaluation} using five main criteria:
 integration complexity, robustness, extensibility, scalability and
-\correction{functionality}. These are based on science gateway properties
-introduced in Section~\ref{sec:intro_sg}. Criteria break down into
+\correction{functionality}. Criteria break down into
 specific metrics where \emph{lower value indicates better
@@
 
-\correction{Note that the metrics defined hereafter aim at providing a framework
+\correction{The metrics defined hereafter aim at providing a framework
   to evaluate architectures independently from any particular system,
@@
   particular technical context of a real system. Some degree of
-system-specificity can be introduced by assigning weights to
-interactions and components through a public spreadsheet available at
-\url{https://frama.link/LrPRsCQp}. In particular, sheet "Components
-and interactions" contains the weight given to these entities in the
-metrics and can be used to produce new versions of
-Table~\ref{table:evaluation} that fit the particular technical context
-of a real system. For instance, a high weight could be assigned to the
-science gateway if its internal architecture is such that it can
-hardly be extended in practice.}
+  system-specificity can be introduced in the evaluation by assigning
+  weights to interactions and components through a public spreadsheet
+  available at \url{https://frama.link/LrPRsCQp}\footnote{The
+    spreadsheet may need to be downloaded to be edited.}. In
+  particular, sheet ``Components and interactions'' in this
+  spreadsheet contains the weight assigned to components and
+  interactions, which can be used to produce new versions of
+  Table~\ref{table:evaluation} that fit the particular technical
+  context of a real system. For instance, a high weight could be
+  assigned to the science gateway if its internal architecture is such
+  that it can hardly be extended in practice.}
 
@@
   different weights to each metric to adjust the specific context of a
-  study.}
+  study.
+
+  The remainder of this Section defines the metrics in each criterion.
+}
 
@@
 effort (e.g., account creation, software installation, APIs, etc.),
-therefore increasing complexity. \correction{We assume that software
-  components contribute equally to $I_1$ and that interactions also
-  contribute equally to $I_2$. In a real system, this may not be the
-  case due to a range of project-specific practical reasons that
-  cannot be covered in this paper. The weights given to components and
-  interactions can be dynamically changed in the public spreadsheet
-  (sheet ``Components and interactions'') to generate new versions of
-  Table~\ref{table:evaluation} as required.} The two metrics
+therefore increasing complexity. \correction{The weights given to
+  software components in I$_1$ and interactions in I$_2$ can be
+  adjusted in the public spreadsheet to cover the technical
+  peculiarities of a real system.}
+
+ The two metrics
 \correction{I$_1$ and I$_2$} are summed to obtain a global score that
@@
   equally regarding engine version updates. Components whose only
-  function is to host the workflow engine, i.e. workflow services and
+  function is to host the workflow engine, i.e., workflow services and
   agents, are not counted in E$_2$ because updates in such components
@@
   components that have other functions than wrapping the engine,
-  i.e. science gateway and infrastructure, are counted in E$_2$
-  because updates in these components require more effort, e.g. new
+  i.e., science gateway and infrastructure, are counted in E$_2$
+  because updates in these components require more effort, e.g., new
   release of the science gateway, gaining administrative privileges on
@@
   constrain data to remain in a particular network domain).
-%\item New science gateway: \todo{Not sure if this makes sense}.
 \end{itemize}
@@
   components. \texttt{\correction{F}$_1$}=0 means that meta-workflows are
-  intrinsically enabled by the architecture (i.e. they can be
+  intrinsically enabled by the architecture (i.e., they can be
   implemented using the components and interactions already in place),
@@
 
-\section{\correction{Architecture evaluation}}
+\section{\correction{Architecture evaluations}}
 
@@
 \texttt{b} and \correction{\texttt{c$_1$}} in the science gateway (\texttt{E$_4$}=3).
-
-%\todo{Integrating a workflow engine as a sub-task
-%only requires to implement interaction \texttt{e} (\texttt{I$_2$}=1)
-%and to install the workflow engine on the infrastructure
-%(\texttt{I$_3$}=1). A new task type is created in the science gateway
-%only when a new workflow has to be integrated (see M$_4$), but this is
-%not required to integrate the engine itself (\texttt{I$_1$}=0).}
 
@@
 
-Differences in \correction{functionality} should not be neglected. Task
-encapsulation, nested workflows (with service invocation) and workflow
+Differences in \correction{functionality} should not be
+neglected. Nested workflows (with service invocation) and workflow
 conversion (with service invocation) are the only architectures that
@@
 
-Table~\ref{table:overall} \todo{update with new architecture?} provides an overall comparison between
+Table~\ref{table:overall} provides an overall comparison between
 architectures, based on the metrics in Table~\ref{table:evaluation}.
-Note that this analysis is only meant for illustration purposes, since
+\correction{It should be noted} that this analysis is only meant for illustration purposes, since
 it assumes that all criteria have equal weights while real systems
@@
 
- Overall, task encapsulation performs a bit
+Overall, task encapsulation \correction{and workflow conversion} perform a bit
 better than the other architectures and nested workflows stand a bit
@@
 
-Note also that all the interactions involved in the architectures were
-treated equally, whereas some of them are obviously more complex than
-others. For example, interaction \texttt{g} (workflow language
-conversion) is clearly more complex than interaction \texttt{d}
-(service invocation). However, without entering into the details of a particular \todo{refer to spreadsheet again}
-system, quantification of the robustness or the amount of development
-associated with each interaction and software component will hardly be
-precise.
+\correction{It should also be noted} that all the interactions
+involved in the architectures were treated equally, whereas some of
+them are obviously more complex than others. For example, interaction
+\texttt{g} (workflow language conversion) is clearly more complex than
+interaction \texttt{d} (service invocation). However, without entering
+into the details of a particular system, quantification of the
+robustness or the amount of development associated with each
+interaction and software component will hardly be
+precise. \correction{If a particular system is evaluated, the public
+  spreadsheet can be used to adjust the weights given to specific
+  components and interactions based on the technical context.}
 
@@
   ``Tight+Task'' instantiation, i.e., the parent engine is tightly
-  integrated in the science gateway and the child engine is
-  integrated through task encapsulation.  Figure~\ref{fig:nested-3}
-  shows the graphical representation of the system. From this
-  representation we find that the sub-tasks submitted by the child
-  engine through interaction \texttt{e} should be handled by the
-  parent engine rather than by the science gateway. Indeed, handling
-  the tasks directly in the science gateway would lead to create a new
-  specific interaction, named, e.g., \texttt{b$_1$}, while \texttt{b}
-  could be reused. We also find that interaction \texttt{e} allows
-  running the sub-tasks on the infrastructure supported by the parent
-  engine with no further development effort required in the child
-  engine.
-
-  The system evaluation is presented on
+  integrated in the science gateway and the child engine is integrated
+  through task encapsulation.  Figure~\ref{fig:nested-3} shows the
+  graphical representation of the system. From this representation we
+  find that the sub-tasks submitted by the child engine through
+  interaction \texttt{e} should be handled by the parent engine rather
+  than by the science gateway. Indeed, handling the tasks directly in
+  the science gateway would lead to create a new specific interaction,
+  named, e.g., \texttt{b$_1$}, while \texttt{b} could be reused. We
+  also find that interaction \texttt{e} allows running the sub-tasks
+  on the infrastructure(s) supported by the parent engine with no
+  further development effort required in the child engine.
+
+  The system evaluation is presented in
   Table~\ref{table:evaluation-2}. The detailed calculations are
@@
   nested workflow model instantiated with service invocation
-  (evaluated on Table~\ref{table:evaluation}): the integration score
+  (evaluated in Table~\ref{table:evaluation}): the integration score
   is improved by a factor of 2, robustness is improved by a factor of
@@
 discussed the pros and cons of all the presented architectures based
-on these metrics.
+on these metrics, \correction{and we have shown how our evaluation
+  framework can be leveraged in the design of new systems}.
 

### 2016-10-13 answers.tex
Small fixes

@@
   instantiation models may be worth designing, for instance nested
-  workflows with task encapsulation, as mentioned in Section 5.2. in
-  the Discussion.
+  workflows with task encapsulation, as mentioned in Section 6.2. (Discussion).
 
@@
 
-\revised{The analysis presented in the paper is intended for
-experts in science gateway and workflow engine design. It is an
-abstraction effort to identify and evaluate the fundamental
-architectural patterns that are encountered while integrating workflow
-engines and science gateways. In real systems, such patterns are often
-intertwined due to considerations related to the historical and
-technical context of software projects. The following work helps
-identify the key patterns in systems and highlight potential
-improvements.}
+\revised{This analysis is intended for experts of science gateway
+  and workflow engine design. It is an abstraction effort to identify
+  and evaluate the fundamental architectural patterns that are
+  encountered while integrating workflow engines and science
+  gateways. In real systems, such patterns are often intertwined and
+  obfuscated due to the historical and
+  technical context of software projects.}
 
@@
   possibility to execute a single workflow with multiple engines. We
-  revised the wording in the description of S1 (page 11) to clarify
-  this point. We find the current description of S2 clear enough on
-  this point: ``[S2] measures the ability to distribute the execution of
-  \emph{a single workflow} among different engine instances.''
+  revised the wording in the description of S1 (Section 4.4) to clarify
+  this point.
 \end{answer}
@@
   following sentence to the paragraph on interaction (g) in Section
-  5.2. of the discussion:
+  6.2. of the discussion:
 
@@
 
-This sentence complements the note already made in section 3.7:
-``Note, however,
-that implementing interaction g can require sub-
-stantial work depending on the complexity of the
-workflow language used by the new engine.``
+This sentence complements the note already made in section 5.6:
+``Note, however, that implementing interaction g can require
+substantial work depending on the complexity of the workflow language
+used by the new engine.``
 \end{answer}
@@
 
-  As discussed previously, we share the reviewer's concerns
-  about the feasibility of workflow conversion (see comments on
-  interaction (g) in Section 5.2). The performance of workflow
-  conversion is strongly related to the availability of a validated
-  interaction (g), which is very dependent on the particular workflow
-  engines under consideration. We believe that the issue raised by the
-  reviewer basically comes down to the weight given to software
-  components and interactions in the analysis. However, using
-  different weights for different components or interactions would
-  open an infinite number of configurations for discussion. For
-  instance, in some cases, the workflow engine already comes with a
-  workflow service which therefore doesn't have to be included. To
-  address this issue, we shared a spreadsheet at
-  \url{https://frama.link/LrPRsCQp} that allows editing the following
-  weights:
+  As discussed previously, we share the reviewer's concerns about the
+  feasibility of workflow conversion (see comments on interaction (g)
+  in Section 6.2). The performance of workflow conversion is strongly
+  related to the availability of a validated interaction (g), which is
+  very dependent on the particular workflow engines under
+  consideration. We believe that the issue raised by the reviewer
+  basically comes down to the weight given to particular software
+  components and interactions in the analysis (interaction \texttt{g}
+  in this case). However, using different weights for different
+  components or interactions would open an infinite number of
+  configurations for discussion. For instance, in some cases, the
+  workflow engine already comes with a workflow service which
+  therefore doesn't have to be counted in the analysis. To address this issue, we
+  shared a spreadsheet at \url{https://frama.link/LrPRsCQp} that
+  allows editing the following weights:
   \begin{itemize}
@@
 The reviewer is right that re-purposing software components
-  and interactions can lead to revisit the evaluation. We updated the following sentence in section 5.2. to better acknowledge this point:
-
-The particular technical or historical context of a science gateway
+  and interactions can lead to revisit the evaluation. We updated the following sentence in section 6.2. to better acknowledge this point:
+
+``The particular technical or historical context of a science gateway
 project may also influence the choice of an architecture to
@@
 \texttt{b} and \texttt{c$_2$} are already available. The migration
-cost between architectures has been ignored as well.
+cost between architectures has been ignored as well.''
 
@@
 \begin{answer}
-We updated the definition of R$_1$ and R$_2$ to include
-  positive examples as well as negative ones. In addition, the
-  specific components included in these metrics are represented in red
-  on Figure 2.
+  We updated the definition of R$_1$ and R$_2$ to include positive
+  examples in addition to negative ones. In addition, the specific
+  components included in these metrics are represented in red on
+  Figure 2.
 \end{answer}
@@
 \begin{answer}
-\texttt{c}, \texttt{c$_1$} and \texttt{c$_2$} are three different
+\texttt{c}, \texttt{c$_1$} and \texttt{c$_2$} were three different
 interactions, which explains the confusion. \texttt{c$_2$} is always
@@
   performance. This is explained at the beginning of section 3:
-``Criteria break down into specific metrics where lower value
-indicates better performance.''
+  ``Criteria break down into specific metrics where \emph{lower value
+    indicates better performance}.''
 \end{answer}
@@
   Portlets) changes the architecture. When portlets are deployed on
-  different hosts, we think that the architecture becomes service
-  invocation as the portlets then behave as individual software
-  components according to our definitions. We added the following comment as a footnote in the presentation of tight integration in section 2.2:
-\revised{Note that the
-    particular case of portlets can be a bit ambiguous: when portlets
-    are deployed remotely, using a protocol such as Web Services for
-    Remote Portlets (WSRP), they become independent software
-    components according to our definition because they could be
-    operated by different teams on different hardware systems. In this
-    case, the system should be classified in the service invocation
-    architecture since the portlets actually become Web
-    Services. The same comment holds for remote object
-    invocation.}
+  different hosts, the architecture becomes service invocation as the
+  portlets then behave as individual software components according to
+  our definitions. We added the following comment as a footnote in the
+  presentation of tight integration in section 2.2: \revised{Note that
+    the particular case of portlets can be a bit ambiguous: when
+    portlets are deployed remotely, using a protocol such as Web
+    Services for Remote Portlets (WSRP), they become independent
+    software components according to our definition because they could
+    be operated by different teams on different hardware systems. In
+    this case, the system should be classified in the service
+    invocation architecture since the portlets actually become Web
+    Services. The same comment holds for remote object invocation.}
 \end{answer}
@@
 \begin{answer}
-  We call ``white-box'' workflow a workflow
+  We call ``white box'' workflow a workflow
   where detailed information about individual tasks is available,
   for instance task statuses, execution logs and resource
-  consumption. On the contrary, a ``black-box'' workflow only
+  consumption. On the contrary, a ``black box'' workflow only
   displays global information such as the workflow status and the
@@
   reported about the tasks' internals because each layer potentially
-  shields or modifies information. Task ids, for instance, may not
-  be properly linked to workflow activities if delivered through a
-  software layer that hasn't been instrumented for that. \todo{refer
-    to olabarriaga2014}.
-
-  The paragraph on fine-grained debugging has been edited accordingly, at the end of Section 3.
+  shields or modifies information. Task ids, for instance, may not be
+  properly linked to workflow activities if delivered through a
+  software layer that hasn't been instrumented for that. This issue is
+  also mentioned in the following paragraph extracted from [34]
+  (reference cited in Section 4.5):
+\begin{quote}
+  From the workflow system perspective, the operator needs interfaces
+  to facilitate DCI configuration, identification and solution of the
+  problems encountered during workflow execution. In practice this can
+  be challenging because the abstract layer introduced by the workflow
+  infrastructure may hamper debugging and troubleshooting, as it may
+  hide too much concrete level information.
+\end{quote}
+
+  The paragraph on fine-grained debugging has been edited accordingly, at the end of Section 4.5.
 \end{answer}
@@
 \begin{answer}
-  We added a section in the paper (current section 6) to show how our
+  We added a complete new section in the paper (Section 7) to show how our
   evaluation framework could be used in the design of new systems.
@@
 We totally agree with the reviewer that Apache Airavata and
-WS-PGrade/gUSE are gateway frameworks. This is how they are presented
-in the text. Instead of providing an incomplete list of gateways that
-use these frameworks in Table 1, we now refer explicitly to ``gateways
-based on Apache Airvata'', ``gateways based on WS-PGRADE/gUSE'' and
-``gateways based on the Catania Science Gateway
-Framework''. Consistently, the table doesn't mention the individual
-gateways based on these frameworks any more.
+WS-PGrade/gUSE are gateway frameworks. Table 1 now refers explicitly
+to ``gateways based on Apache Airvata'', ``gateways based on
+WS-PGRADE/gUSE'' and ``gateways based on the Catania Science Gateway
+Framework''. 
 \end{answer}
@@
 \begin{answer}
-  The task submission framework is part of the infrastructure
+  The task submission framework (e.g. SAGA) is part of the infrastructure
   and its nature doesn't influence the architecture used to integrate
@@
   instantiation of the nested workflow abstract model, which is not
-  evaluated in the paper as mentioned in section 5.2: "\emph{The
+  evaluated in the paper. Several other instantiations of the abstract
+  models could be envisaged, as mentioned in section 6.2: "\emph{The
     abstract nested workflows and workflow conversion patterns were
@@
     instantiation, for instance nested workflows with task
-    encapsulation, could also be envisaged.}". It sounds like the
-  architecture described by the reviewer is a nested workflow pattern
-  instantiated with a tight integration (parent engine) and task
-  encapsulation (child engine).
-
-  We found this scenario interesting and we evaluated it in our
-  framework. The analysis is reported in the new section 6 of the
+    encapsulation, could also be envisaged.}".
+
+  The architecture described by the reviewer is a nested workflow
+  pattern instantiated with a tight integration (parent engine) and
+  task encapsulation (child engine). We found this scenario
+  interesting as it illustrates how our framework can be used to help
+  designing new systems (although we realize that such a system may
+  already exist). We evaluated this hypothetical architecture in our
+  framework: the analysis is reported in the new section 7 of the
   paper.
@@
 \end{review}
-
-\begin{answer}
-  We followed the recommendation of the reviewer and created a new
-  Section 3 to present and discuss these examples. The caption of
-  Figures 5 and 6 is still long though, because we think that the
-  detailed references to the interaction numbers on the Figures should
-  be part of them, so that the reader doesn't have to constantly
-  navigate between the text and the Figures.
-\end{answer}
 
@@
 architecture) a bit unfortunate because it would give different
-ponderations to the same component or interaction depending on the
+values to the same component or interaction depending on the
 architecture in which it is used, which we think would obfuscate the
@@
 \begin{answer}
-  While we agree that the Sub-Sections in Section 3 look a bit
+  While we agree that the Sub-Sections in Section 5 (former Section 3) look a bit
   repetitive, we believe that the metrics evaluation reported in Table 2 needs to be
-  properly justified. 
+  properly justified. We contracted the text wherever possible.
 \end{answer}
@@
 REST or SOAP Web-Service).
-  \item In task encapsulation, the workflow engine does not transfer data to the infrastructure while in service invocation it does (interactioni \texttt{c$_2$}.
+  \item In task encapsulation, the workflow engine does not transfer data to the infrastructure while in service invocation it does (interaction \texttt{c$_2$}.
   \item In task encapuslation, the science gateway submits tasks to the infrastructure, while in service invocation the workflow engine does.
   \end{itemize}
-  Therefore, while we acknowledge that task encapsulation and service
-  invocation share characteristics, we believe that their differences
-  can be clearly inferred from section 2 of the paper. 
 \end{answer}
@@
 
-If a new child engine is added to the nested model then the following interactions an and components have to be updated:
+If a new child engine is added to the nested model then the following interactions an and components have to be updated (a similar analysis can be done when a parent engine is added):
 \begin{enumerate}
@@
 \end{enumerate}
-This is why E$_4$=4 for the nested model.
+This is why E$_4$=4 for the nested model when a child engine is added.
 
@@
 (infrastructure is not a discriminant factor here). First, the
-reviewer suggest that the availability of ORE bundles in the SHIWA
+reviewer suggests that the availability of ORE bundles in the SHIWA
 framework standardizes the communication with workflow engines so that
@@
 ORE bundles is contextual information that can hardly be integrated in
-the generic analysis. We believe that our framework helps evaluate
+the generic analysis. However, our framework helps evaluate
 the impact of technologies such as ORE bundles on architectures, as
@@
 
-Second, the
-reviewer suggests that the integration of the workflow engine in the
-agent (pool model) is easier than in the service because the engine
-has to produce and consume ORE bundles in the former case while it
-only has to consume them in the latter case. This is true again, but
-it comes down to the discussion on the weights of the different
-software components. In this particular case, it would make sense to
-assign a higher weight to the agent than to the service, which could
-be done through the public spreadsheet.  
+Second, the reviewer suggests that the integration of the workflow
+engine in the agent (pool model) is easier than in the service because
+the engine has to produce and consume ORE bundles in the former case
+while it only has to consume them in the latter case. This is true
+again, but we think that it comes down to a discussion on the weights
+of the different software components. In this particular case, it
+would make sense to assign a higher weight to the agent than to the
+service, because integrating a workflow engine in the agents requires
+more work than integrating it in a service.
+
+To address this issue, we
+  shared a spreadsheet at \url{https://frama.link/LrPRsCQp} that
+  allows editing the following weights (see also answers to Reviewers \#1 and \#2):
+  \begin{itemize}
+  \item Weight given to software components and interactions in
+  ``integration'', ``extensibility'' and ``robustness'' criteria.
+  \item Weight given to metrics in a criterion (e.g., E$_1$ vs. E$_2$ in
+  ``extensibility'').
+  \item Weight given to criteria in overall evaluation of Table 3.
+\end{itemize}
+The paper references this spreadsheet wherever equal weights are
+assumed. The readers and reviewers are invited to rely on this
+spreadsheet for any issue related to the importance given to a
+specific component, interaction, metric or criterion in the
+analysis. In the paper, we assumed that all weights equal 1, which we
+believe is a reasonable choice although inherently limited.
 \end{answer}
@@
   public spreadsheet. For instance, if the science gateway code is
-  known to be ``untouchable'' due to a cumbersome, then a high weight
+  known to be ``untouchable'' due to a bulky architecture, then a high weight
   should be put on the science gateway component. 
@@
 \begin{answer}
-We added the following paragraphs to the introduction to section 3:
-
-\revised{Note that the metrics defined hereafter aim at providing a framework
+We added the following paragraphs to the introduction to Section 4:
+
+\revised{The metrics defined hereafter aim at providing a framework
   to evaluate architectures independently from any particular system,
@@
   particular technical context of a real system. Some degree of
-system-specificity can be introduced by assigning weights to
-interactions and components through a public spreadsheet available at
-\url{https://frama.link/LrPRsCQp}. In particular, sheet "Components
-and interactions" contains the weight given to these entities in the
-metrics and can be used to produce new versions of
-Table 2 that fit the particular technical context
-of a real system. For instance, a high weight could be assigned to the
-science gateway if its internal architecture is such that it can
-hardly be extended in practice.}
-
-\revised{In addition, all the metrics in a criterion contribute
+  system-specificity can be introduced in the evaluation by assigning
+  weights to interactions and components through a public spreadsheet
+  available at \url{https://frama.link/LrPRsCQp}\footnote{The
+    spreadsheet may need to be downloaded to be edited.}. In
+  particular, sheet ``Components and interactions'' in this
+  spreadsheet contains the weight assigned to components and
+  interactions, which can be used to produce new versions of
+  Table 2 that fit the particular technical
+  context of a real system. For instance, a high weight could be
+  assigned to the science gateway if its internal architecture is such
+  that it can hardly be extended in practice.
+
+  In addition, all the metrics in a criterion contribute
   equally to the global score used for the criterion. Column B of
@@
   different weights to each metric to adjust the specific context of a
-  study.}
+  study.
+}
 \end{answer}

### 2016-10-13 paper.tex
Small fixes

@@
 
-\correction{This analysis is intended for experts in science gateway
+\correction{This analysis is intended for experts of science gateway
   and workflow engine design. It is an abstraction effort to identify
@@
   may exist at a finer granularity, to implement functions such as
-  authentication, data persistency or logging. Such fine-grained
+  authentication, data persistence or logging. Such fine-grained
   components are not covered by our analysis since we focus on the
@@
   JSAGA~\cite{reynaud2010uniform} or Data Avenue~\cite{hajnal2014data}
-  can be used to homogeneize these interfaces.
+  can be used to homogenize these interfaces.
 \item Workflow control: operations to control workflow execution in an
@@
 through interaction \texttt{b}. Although the science gateway has no
-global vision of the workflow, it completely \correction{controls} the
-sub-tasks. \correction{For instance, it may cancel} them when the
+global vision of the workflow, it \correction{submits the sub-tasks to the
+infrastructure} and completely \correction{controls them}. \correction{For instance, it may cancel} them when the
 \correction{workflow} is canceled. The science gateway may also
 implement mechanisms to facilitate the handling of task dependencies,
-for instance \correction{lists of tasks that must be completed before a
-  particular task can be executed}.
+for instance \correction{lists of tasks that must be completed before
+  a particular task can be executed}.
 
@@
 \item Fine-grained debugging (\texttt{\correction{F}$_2$}): availability of
-  fine-grained debugging information about workflow tasks (white-box
-  workflow). \correction{We call ``white-box'' workflow a workflow
+  fine-grained debugging information about workflow tasks (white box
+  workflow). \correction{We call ``white box'' workflow a workflow
     where detailed information about individual tasks is available,
     for instance task statuses, execution logs and resource
-    consumption. On the contrary, a ``black-box'' workflow only
+    consumption. On the contrary, a ``black box'' workflow only
     displays global information such as the workflow status and the
@@
 conversion) requires particular attention when implementing a real
-system since this interaction may not be easily generalizable to any
+system since this interaction may not be easily generalized to any
 workflow language. For instance, converting FSL, PSOM or Nipype

### 2016-10-17 answers.tex
Updates following Marc's comments. Submitted revision

@@
   long experience with both science gateways and workflow engines, and
-  we completely acknowledge that non experts may misinterpret the
-  results.
+  we completely acknowledge it may be of limited interest for non
+  expert users.
 
@@
   and evaluate the fundamental architectural patterns that are
-  encountered while integrating workflow engines and science
-  gateways. In real systems, such patterns are often intertwined and
-  obfuscated due to the historical and
-  technical context of software projects.}
+  encountered while integrating workflow engines and science gateways.
+  In real systems, such patterns sometimes coexist due to the
+  historical and technical context of software projects.}
 
@@
   updated text in the paper demonstrate that the metrics and
-  definitions are rigorous.
+  definitions are rigorous. We also hope that the flexibility provided
+  by the shared metrics sheet will bring the readers to consider this
+  paper to be a valuable tool to stimulate in depth discussions based
+  on an initial comparison between abstract architectural designs.
 \end{answer}

### 2016-10-17 paper.tex
Updates following Marc's comments. Submitted revision

@@
   integration). Figure adapted from~\cite{ardizzone2012decide} with
-  permission of the first author} \todo{approval still pending}.}
+  permission of the first author} \color{red}({approval still pending})\color{black}.}
 \label{fig:csgf}
@@
 \includegraphics[width=0.9\textwidth]{figures/VIP.eps}
-\caption{Architecture of the Virtual Imaging Platform (service
+\caption{\correction{Architecture of the Virtual Imaging Platform (service
   invocation).  Step \texttt{3} corresponds to interaction \texttt{d}
@@
   workflow data transfers are performed by the tasks and embedded in
-  their descriptions).}
+  their descriptions).}}
 \label{fig:vip-architecture}
@@
 }
-\caption{CBRAIN architecture for workflow engine integration
+\caption{\correction{CBRAIN architecture for workflow engine integration
   (task encapsulation).  1: User sends data and workflow execution request to
@@
   \texttt{4}, \texttt{5}, \texttt{7} and \texttt{8} (regular
-  interactions with batch manager). Interaction \texttt{c} is
+  interactions with batch manager). Interaction \texttt{$c_1$} is
   implemented by steps \texttt{3} and \texttt{9}. Interaction
@@
   \url{https://github.com/aces/cbrain-plugins-neuro}; for PSOM:
-  through a specific development in PSOM).}
+  through a specific development in PSOM).}}
 \label{fig:cbrain-sub-tasking}
@@
 \includegraphics[width=1.5\columnwidth]{figures/pool-interactions.pdf}
-\caption{Architecture of the SHIWA pool. Circle-terminated arrows
+\caption{\correction{Architecture of the SHIWA pool. Circle-terminated arrows
   indicate messages that are broadcast to all pool
@@
   representation in Figure~\ref{fig:architectures}(IV). Figure adapted
-  from~\cite{ROGE-13}.}
+  from~\cite{ROGE-13}.}}
 \label{fig:shiwa-pool-architecture}
@@
 \includegraphics[width=1.5\columnwidth]{figures/shiwa-science-gateway.png}
-\caption{Nested workflow execution through the SHIWA Science Gateway. Figure
+\caption{\correction{Nested workflow execution through the SHIWA Science Gateway. Figure
   adapted from~\cite{terstyanszky2014enabling} with permission of
-  the first author.}
+  the first author.}}
 \label{fig:shiwa-architecture}
@@
 
-\section{\correction{Architecture evaluations}}
+\section{\correction{Architecture evaluation}}
 
@@
   infrastructure, which executes the sub-workflows. This scenario came
-  up in the discussion with Reviewer \#2 of this paper, after the
-  evaluation framework had already been designed.
+  up after the evaluation framework had already been designed.
 
